### 2019-01-13 paper/paper.tex
Initial writing on paper

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from traditional SIGPLAN
%% proceedings format to PACMPL format must update the
%% '\documentclass' and topmatter commands above; see
%% 'acmart-pacmpl-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\usepackage{syntax}

\usepackage{semantic}

\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
  basewidth={.5em,.5em},
}
\newcommand{\ocaml}{\lstinline[language={[objective]caml}]}

%% symbols used throughout the paper

\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
\newcommand{\T}{\mathbb{T}} % Set of terminals

\begin{document}

%% Title information
\title{Resolvable Ambiguity}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Viktor Palmkvist}
\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{KTH Royal Institute of Technology}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{Stockholm}
  \state{State1}
  \postcode{Post-Code1}
  \country{Sweden}                    %% \country is recommended
}
\email{vipa@kth.se}          %% \email is recommended

%% Author with two affiliations and emails.
\author{First2 Last2}
\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2a}
  \department{Department2a}             %% \department is recommended
  \institution{Institution2a}           %% \institution is required
  \streetaddress{Street2a Address2a}
  \city{City2a}
  \state{State2a}
  \postcode{Post-Code2a}
  \country{Country2a}                   %% \country is recommended
}
\email{first2.last2@inst2a.com}         %% \email is recommended
\affiliation{
  \position{Position2b}
  \department{Department2b}             %% \department is recommended
  \institution{Institution2b}           %% \institution is required
  \streetaddress{Street3b Address2b}
  \city{City2b}
  \state{State2b}
  \postcode{Post-Code2b}
  \country{Country2b}                   %% \country is recommended
}
\email{first2.last2@inst2b.org}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Text of paper \ldots

\subsection{Motivating Ambiguity in Programming Languages}

% TODO: reference PADL paper for this example
Consider the following nested match expression in OCaml:

\begin{lstlisting}[language={[objective]caml}]
match 1 with
  | 1 -> match "one" with
         | str -> str
  | 2 -> "two"
\end{lstlisting}

\noindent The OCaml compiler, when presented with this code, will give a type error for the last line:

\begin{lstlisting}
Error: This pattern matches values of type int
       but a pattern was expected which matches
       values of type string
\end{lstlisting}

\noindent The compiler sees the last line as belonging to the inner \ocaml{match} rather than the outer, as was intended. The fix is simple; put parentheses around the inner match:

\begin{lstlisting}[language={[objective]caml}]
match 1 with
  | 1 -> (match "one" with
          | str -> str)
  | 2 -> "two"
\end{lstlisting}

\noindent The connection between the error message and the fix is not a clear one however; adding parentheses around an expression does not change the type of anything.

To come up with an alternative error to present in this case we look to the OCaml manual for inspiration. It contains an informal description of the syntax of the language\footnote{\url{https://caml.inria.fr/pub/docs/manual-ocaml/language.html}}, in the form of an EBNF-like grammar. Below is an excerpt of the productions for expressions, written in a more standard variant of EBNF:

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= 'match' <expr> 'with' <pattern-matching>

<pattern-matching> ::= ('|' <pattern> '->' <expr>)+
\end{grammar}

Note that \synt{pattern-matching} is slightly simplified, the original grammar supports \ocaml{when} guards and makes the first \lit{|} optional. If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match. The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended.

We instead argue that the grammar should be left ambiguous for this sort of corner cases that are likely to trip a user, allowing the compiler to present an ambiguity error, which lets the user select the intended alternative.

\subsection{Unresolvable Ambiguity}

Unfortunately, not all ambiguities can be resolved by adding parentheses. Again, looking to the informal OCaml grammar:

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <expr> ';' <expr>
  \alt '[' <expr> (';' <expr>)* ';'? ']'
  \alt <constant>
\end{grammar}

\noindent The first production is sequential composition, the second is lists (the empty list is under \synt{constant}). Now consider the following expression: ''\ocaml{[1; 2]}''.

We find that it is ambiguous with two alternatives:
\begin{enumerate}
  \item A list with two elements.
  \item A list with one element, namely a sequential composition.
\end{enumerate}

We can select the second option by putting parentheses around ''\ocaml{1; 2}'', but there is no way to select the first. If the user intended the first option we have a problem: we can present an accurate error message, but there is no way for an end-user to solve it; it requires changes to the grammar itself.

% TODO: ref for undecidable ambiguity checking
To prevent the possibility of an end-user encountering such an error we must ensure that the grammar cannot give rise to an unresolvable ambiguity. It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable. Unresolvable ambiguity, however, turns out to be decidable\footnote{cross your fingers}.

Note that, as for ambiguity, the shape of the grammar is important, since the property considers parse trees rather than merely words. For this paper, we consider context-free grammars with EBNF operators.

\subsection{Contributions}

\begin{itemize}
  % TODO: reference
  \item Building on PADL-PAPER, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
  \item A formalization of the unresolvable ambiguity property for context-free EBNF grammars.
  \item An algorithm for deciding if a grammar is unresolvably ambiguous or not.
\end{itemize}

\section{Preliminaries}

% TODO: the last line feels a bit odd, notationally
\begin{figure}
  \begin{tabular}{@{}ll@{}}
      Terminals & $t \in \T$ \\
      Non-terminals & $N \in \NT$ \\
      Regular expressions & $r ::= t \mid N \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
      Productions & $N -> r$ \\
  \end{tabular}
  \caption{Context-free EBNF grammars}
  \label{fig:grammar-definition}
\end{figure}

A context-free EBNF grammar is a tuple $(S, P, \NT, \T)$. $S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$ and $\T$ are disjoint sets of non-terminals and terminals, respectively.

The (word) language of a given non-terminal $N$ in a grammar $G = (S, P, \NT, \T)$ is given by $L_G(N)$:

$$
\begin{array}{r@{\;=\;}l}
  L_G(t) & \{t\} \\
  L_G(N) & \bigcup \{L_G(r) \mid (N -> r) \in P\} \\
  L_G(r_1 \cdot r_2) & \{w_1 \cdot w_2 \mid w_1 \in L_G(r_1), w_2 \in L_g(r_2)\} \\
  L_G(r_1 + r_2) & L_G(r_1) \cup L_G(r_2) \\
  L_G(\epsilon) & \{\epsilon\} \\
  L_G(r^{*}) & \{\epsilon\} \cup L_G(r) \cup L_G(r \cdot r) \cup \ldots \\
\end{array}
$$

The word language of a grammar $G$, written $L(G)$, is thus $L_G(S)$. We will omit the subscript whenever the intended grammar is clear from the context.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
\appendix
\section{Appendix}

Text of appendix \ldots

\end{document}

### 2019-01-16 vpl-ambiguity/src/Automaton/DFA.hs
Generate paren-nfa

@@ -2,12 +2,16 @@ module Automaton.DFA
 ( DFA(..)
 , renumber
 , minimize
+, asNFA
 ) where
 
 import Pre
 
 import qualified Data.HashMap.Lazy as M
 import qualified Data.HashSet as S
+
+import Automaton (FiniteAutomaton(..), EpsNFA(EpsNFA))
+import qualified Automaton.NFA as N
 
 import Util (flipMap)
 
@@ -81,3 +85,17 @@ repeatWhile :: Monad m => m [a] -> (a -> m b) -> m ()
 repeatWhile mlist action = mlist >>= \case
   a : _ -> action a >> repeatWhile mlist action
   [] -> return ()
+
+instance FiniteAutomaton DFA where
+  asEpsNFA DFA{initial, transitions, final} = EpsNFA initial (convert <$> transitions) final
+    where
+      convert = M.toList >>> fmap (Just *** S.singleton) >>> M.fromList
+  mapState convert DFA{initial, transitions, final} = DFA
+    { initial = convert initial
+    , transitions = M.toList transitions
+      & fmap (convert *** fmap convert)
+      & M.fromList
+    , final = S.map convert final }
+
+asNFA :: (Hashable s) => DFA s a -> N.NFA s a
+asNFA DFA{initial, transitions, final} = N.NFA initial (fmap S.singleton <$> transitions) final

### 2019-01-16 vpl-ambiguity/src/Automaton/EpsilonNFA.hs

@@ -1,9 +1,10 @@
 module Automaton.EpsilonNFA
-( EpsNFA(..)
-, edge
+( edge
 , addEdge
 , mergeTransitions
 , determinize
+, states
+, renumber
 ) where
 
 import Pre
@@ -13,13 +14,9 @@ import qualified Data.HashSet as S
 
 import Util (iterateInductivelyOptM, iterateInductively)
 
+import Automaton (EpsNFA(..))
 import Automaton.DFA (DFA(DFA))
 import qualified Automaton.DFA as D
-
-data EpsNFA state alphabet = EpsNFA
-  { initial :: state
-  , transitions :: HashMap state (HashMap (Maybe alphabet) (HashSet state))
-  , final :: state }
 
 edge :: (Hashable s, Hashable a) => s -> Maybe a -> s -> HashMap s (HashMap (Maybe a) (HashSet s))
 edge s a e = M.singleton s $ M.singleton a (S.singleton e)
@@ -31,17 +28,23 @@ addEdge start t end a@EpsNFA{transitions} = a
 mergeTransitions :: (Eq s, Hashable s, Eq a, Hashable a) => HashMap s (HashMap (Maybe a) (HashSet s)) -> HashMap s (HashMap (Maybe a) (HashSet s)) -> HashMap s (HashMap (Maybe a) (HashSet s))
 mergeTransitions = M.unionWith (M.unionWith S.union)
 
+states :: forall s a. (Eq s, Hashable s) => EpsNFA s a -> HashSet s
+states EpsNFA{initial, transitions, final} = S.insert initial $
+  final `S.union` (M.toList transitions & foldMap getStates)
+  where
+    getStates (s, tr) = S.insert s $ fold tr
+
 determinize :: forall s a. (Eq s, Hashable s, Eq a, Hashable a) => EpsNFA s a -> DFA (HashSet s) a
 determinize EpsNFA{initial, transitions, final} = DFA
   { D.initial = startState
   , D.transitions = transitions'
-  , D.final = S.filter (S.member final) states }
+  , D.final = S.filter (S.intersection final >>> S.null >>> not) newStates }
   where
     startState = epsClose $ S.singleton initial
     epsClose = iterateInductively (transition Nothing)
     transition tr s = M.lookup s transitions >>= M.lookup tr & fold
 
-    (states, transitions') = runState
+    (newStates, transitions') = runState
       (iterateInductivelyOptM findTransitions $ S.singleton startState)
       mempty
 
[…]

### 2019-01-16 vpl-ambiguity/src/GraphViz.hs

@@ -1,0 +1,47 @@
+module GraphViz where
+
+import Pre
+
+import qualified Data.Text as Text
+
+import qualified Data.HashMap.Lazy as M
+import qualified Data.HashSet as S
+
+import Automaton (EpsNFA(..), FiniteAutomaton(..))
+import qualified Automaton.EpsilonNFA as E
+
+-- TODO: escape labels
+toDotText :: (Eq s, Hashable s, Eq a, Hashable a, FiniteAutomaton fa)
+          => (s -> Text) -> (a -> Text) -> fa s a -> Text
+toDotText sToStrOriginal aToStrOriginal originalFA = "digraph {\n"
+  <> "  rankdir=LR;\n"
+  <> "  startState [shape=point, label=\"\"];\n"
+  <> "  startState -> " <> show initial <> ";\n"
+  <> foldMap stateDescription (E.states nfa)
+  <> foldMap edgeDescription (M.toList transitions)
+  <> "}\n"
+  where
+    aToStr = aToStrOriginal >>> escape
+    (nfa@EpsNFA{initial, transitions, final}, sToInt) = E.renumber $ asEpsNFA originalFA
+    intToS = M.toList sToInt & fmap swap & M.fromList
+    sToStr s = M.lookup s intToS
+      & compFromJust "GraphViz.toDot.sToStr" "unknown int state"
+      & sToStrOriginal
+      & escape
+
+    stateDescription s =
+      "  " <> show s <> " [label = \"" <> sToStr s <> "\", shape=" <> shape s <> "];\n"
+    shape s
+      | s `S.member` final = "doublecircle"
+      | otherwise = "circle"
+
+    edgeDescription (s, outgoing) = foldMap edge (M.toList outgoing)
+      where
+        edge (a, ss) = "  " <> show s <> " -> {" <> (S.toList ss & fmap show & intersperse ", " & mconcat)
+          <> "} [label=\"" <> maybe "eps" aToStr a <> "\"];\n"
+
+escape :: Text -> Text
+escape = Text.concatMap f
+  where
+    f '"' = "\\\""
+    f c = Text.singleton c

### 2019-01-16 vpl-ambiguity/src/Regex.hs

@@ -15,13 +15,12 @@ import Pre hiding (concat)
 import Data.String (fromString)
 
 import qualified Data.HashMap.Lazy as M
+import qualified Data.HashSet as S
 
 import Data.Functor.Foldable (cata)
 import Data.Functor.Foldable.TH (makeBaseFunctor)
 
-import Automaton.DFA (DFA)
-import qualified Automaton.DFA as D
-import Automaton.EpsilonNFA (EpsNFA(..))
+import qualified Automaton as FA
 import qualified Automaton.EpsilonNFA as E
 
 data Regex alphabet = Terminal alphabet
@@ -49,37 +48,46 @@ opt :: Regex a -> Regex a
 opt r = Choice Eps r
 
 type AutomataM a = State Int a
+data Automaton s a = Automaton
+  { initial :: s
+  , transitions :: HashMap s (HashMap (Maybe a) (HashSet s))
+  , final :: s }
+
+addEdge :: (Eq s, Hashable s, Eq a, Hashable a) => s -> Maybe a -> s -> Automaton s a -> Automaton s a
+addEdge start t end a@Automaton{transitions} = a
+  { transitions = transitions & M.insertWith (M.unionWith S.union) start (M.singleton t (S.singleton end)) }
 
 newState :: AutomataM Int
 newState = get <* modify (+1)
 
-toAutomaton :: (Eq a, Hashable a) => Regex a -> DFA Int a
-toAutomaton regex = evalState (cata alg regex) 1 & E.determinize & D.renumber & fst
+toAutomaton :: (Eq a, Hashable a) => Regex a -> FA.EpsNFA Int a
+toAutomaton regex = evalState (cata alg regex) 1 & toEpsNFA
   where
+    toEpsNFA Automaton{initial, transitions, final} = FA.EpsNFA initial transitions $ S.singleton final
     alg (TerminalF a) = do
       s <- newState
       e <- newState
-      return $ EpsNFA s (E.edge s (Just a) e) e
+      return $ Automaton s (E.edge s (Just a) e) e
     alg (ConcatF a b) = do
-      EpsNFA{initial = as, transitions = at, final = ae} <- a
-      EpsNFA{initial = bs, transitions = bt, final = be} <- b
+      Automaton{initial = as, transitions = at, final = ae} <- a
+      Automaton{initial = bs, transitions = bt, final = be} <- b
       let transitions = E.mergeTransitions at bt
-      return $ EpsNFA as transitions be
-        & E.addEdge ae Nothing bs
[…]

### 2019-02-25 paper/paper.tex
Description of the basic algorithm

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
@@
 
+\usepackage{tikz}
+\usetikzlibrary{arrows,automata}
+\usepackage{tikz-qtree} % Used for syntax trees
+\usepackage{tikz-cd} % Used for commutative diagrams
+
 \usepackage{syntax}
@@
 \usepackage{semantic}
+
+\usepackage{marginnote}
 
@@
 
-%% symbols used throughout the paper
+%% symbol definitions used throughout the paper
 
@@
 \newcommand{\T}{\mathbb{T}} % Set of terminals
+\newcommand{\I}{\mathbb{I}} % Set of identifiers
+\newcommand{\yield}{\mathit{yield}} % yield of a parse tree
+\newcommand{\semantic}{\mathit{semantic}} % remove semantically unimportant productions from a w' \in L(G')
+\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G') to a subset of L(T)
 
@@
 It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable.
-Unresolvable ambiguity, however, turns out to be decidable\footnote{cross your fingers}.
+Unresolvable ambiguity, however, turns out to be decidable\footnote{With some caveats, I'll talk more about this during the meeting.}.
 
@@
 \begin{itemize}
-  % TODO: reference
-  \item Building on PADL-PAPER, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
+  \item Building on \cite{palmkvistCreatingDomainSpecificLanguages2019}, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
   \item A formalization of the unresolvable ambiguity property for context-free EBNF grammars.
@@
 
+\begin{quote} % TODO: remove
+  \textbf{NOTE:} The text hereafter is written a bit later than the text before, and is ever so slightly inconsistent with it.
+\end{quote}
+
 % TODO: the last line feels a bit odd, notationally
@@
       Non-terminals & $N \in \NT$ \\
+      Identifiers & $i \in \I$ \\
       Regular expressions & $r ::= t \mid N \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
-      Productions & $N -> r$ \\
+      Productions & $i : N -> r$ \\
   \end{tabular}
@@
 
-A context-free EBNF grammar is a tuple $(S, P, \NT, \T)$.
-$S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$ and $\T$ are disjoint sets of non-terminals and terminals, respectively.
+A context-free EBNF grammar is a tuple $(S, P, \NT, \T, \I)$.
+$S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$, $\T$, and $\I$ are sets of non-terminals, terminals, and identifiers, respectively.
+Additionally, we require that $\NT \cap \T = \emptyset$, and that the identifiers uniquely identify each production, i.e., there are no two distinct productions in $P$ with the same identifier.
 
-The (word) language of a given non-terminal $N$ in a grammar $G = (S, P, \NT, \T)$ is given by $L_G(N)$:
+The (word) language of a given non-terminal $N$ in a grammar $G = (S, P, \NT, \T, \I)$ is given by $L_G(N)$:
 
@@
 
-The word language of a grammar $G$, written $L(G)$, is thus $L_G(S)$. We will omit the subscript whenever the intended grammar is clear from the context.
+\noindent The word language of a grammar $G$, written $L(G)$, is thus $L_G(S)$. We will omit the subscript whenever the intended grammar is clear from context.
+
+Note that the right hand side of a production is a regular expression, which could potentially be ambiguous in and of itself.
+For this work we assume the rhs regular expressions to either be unambigous, or that any remaining ambiguity is unimportant.
+This can be achieved in a number of ways, for example using the ambiguity checking of Brabrand and Thomsen \cite{brabrandTypedUnambiguousPattern2010}.
+
+As such, for a given grammar $G = (S, P, \NT, \T, \I)$ we construct a linear representation of its parse trees as another grammar $T_G = (S, P', \NT, \T', \I)$ where:
+
+$$
+\begin{array}{r@{\;=\;}l}
+  P' & \{i : N -> [_i r ]_i \mid (i : N -> r) \in P \} \\
+  \T' & \T \sqcup \bigcup_{(i : N -> r) \in P} \{[_i, ]_i \} \\
+\end{array}
+$$
+
+\noindent where $\sqcup$ denotes disjoint union.
+Intuitively, we surround the right hand side of each production with a unique pair of brackets, signifying the production that was used for the parse.
+Again, we will omit the subscript whenever the intended grammar is clear from context.
+
+The $\yield$ of a parse tree is the word it parsed, i.e., $\yield : L(T) -> L(G)$. Intuitively, $\yield$ removes the brackets introduced when constructing $T$.
+
+A given word $w \in L(G)$ is ambiguous iff:
+
+$$\exists t_1, t_2 \in L(T).\ \yield(t_1) = w \land \yield(t_2) = w \land t_1 \neq t_2$$
+
+\noindent A grammar is ambiguous iff it contains at least one ambiguous word.
+
+\section{Parse-time Disambiguation}
+
+We begin this section with some motivation, and then list our definition of \emph{resolvable ambiguity}.
+
+We can divide the productions present in a programming language grammar in two groups: those that are semantically important, and those that are semantically \emph{un}important.
+The former group covers most productions, while the latter contains, e.g., parentheses used for explicit grouping.
+If programs were written directly as syntax trees then the latter would be unnecessary; two syntax trees that differ only by parentheses are semantically the same.
+
+Thus we wish the output of parsing to be a syntax tree consisting entirely of semantically important productions.
+This distinction is useful to make, because it allows us to decouple \emph{what} we want to be expressible and \emph{how} it is to be expressed.
+As an example, in Figure~\ref{fig:example-grammar}, \ref{fig:example-grammar:ambig-grammar} is the \emph{what} and \ref{fig:example-grammar:unambig-grammar} is the \emph{how}.
+The latter grammar is a modification of the former that adds precedence, associativity, and parentheses, yielding an unambiguous grammar with at least one way to express each semantically distinct tree.
+
+\begin{figure}
+  \begin{subfigure}{\linewidth}
+    \setlength{\grammarindent}{5em}
+    \begin{grammar}
+      <expr> ::= Sum: <expr> '+' <expr>
+        \alt Product: <expr> '*' <expr>
+        \alt Number: <number>
+    \end{grammar}
+    \caption{The (ambiguous) intuitive grammar without parentheses.}
+    \label{fig:example-grammar:ambig-grammar}
+  \end{subfigure}
+
+  \begin{subfigure}{\linewidth}
+    \setlength{\grammarindent}{5em}
+    \begin{grammar}
+      <expr> ::= Sum: <product> '+' <expr>
+        \alt PassthroughE: <product>
+
+      <product> ::= Product: <atom> '*' <product>
+        \alt PassthroughP: <atom>
+
+      <atom> ::= Paren: '(' <expr> ')'
+        \alt Number: <number>
+    \end{grammar}
+    \caption{The (unambiguous) grammar with parentheses.}
+    \label{fig:example-grammar:unambig-grammar}
+  \end{subfigure}
+
+  \begin{subfigure}{\linewidth}
+    \begin{center}
+    \Tree [.PassthroughE
+      [.\underline{Product}
+        [.Paren
+          '('
+          [.\underline{Sum}
+            [.PassthroughP [.\underline{Number} 1 ] ]
+            '+'
+            [.PassthroughE [.PassthroughP [.\underline{Number} 2 ] ] ] ]
+          ')' ]
+        '*'
+        [.PassthroughP [.\underline{Number} 3 ] ] ] ]
+    \end{center}
+    \caption{Parse tree for $(1 + 2) * 3$. The underlined nodes are semantically important.}
+    \label{fig:example-grammar:tree}
+  \end{subfigure}
+
+  \begin{subfigure}{\linewidth}
+    \begin{center}
+    \Tree [.\underline{Product}
+      [.\underline{Sum} [.\underline{Number} 1 ] '+' [.\underline{Number} 2 ] ]
+      '*'
+      [.\underline{Number} 3 ] ]
+    \end{center}
+    \caption{The same parse tree, after removing the semantically unimportant nodes.}
+    \label{fig:example-grammar:sem-tree}
+  \end{subfigure}
+
+  \caption{A basic expression grammar in two variations, and two example syntax trees.}
+  \label{fig:example-grammar}
+\end{figure}
+
+Our definition thus refers to four grammars in total:
+
+\begin{itemize}
+  \item The semantic grammar $G$, containing only the semantically important productions (e.g., Figure~\ref{fig:example-grammar:ambig-grammar}).
+  \item $T_G$ (generally abbreviated as $T$), the parse trees of $G$, representing the trees that must be expressible (Figure~\ref{fig:example-grammar:sem-tree} is in this language).
+  \item The parse grammar $G'$, a modification of $G$ meant to actually be used for parsing (e.g., Figure~\ref{fig:example-grammar:unambig-grammar}).
+The details of these modifications will be covered later.
+% TODO: likely put a forward reference here
+  \item $T_{G'}$ (generally abbreviated as $T'$), the parse trees of $G'$ (Figure~\ref{fig:example-grammar:tree} is in this language).
+\end{itemize}
+
+\noindent We also require a function $\semantic : L(T') -> L(T)$ that removes the semantically unimportant productions from a parse tree.
+The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}.
+Finally, we define the function $\parse : L(G') -> 2^{L(T)}$:
+
+$$
+\parse(w') = \{ \semantic(t') \mid t' \in L(T') \land \yield(t') = w' \}
+$$
+
+\begin{figure}
+  \begin{tikzcd}
+    L(G) & L(T) \arrow[l, "\yield"] \\
+    L(G') & L(T') \arrow[u, "\semantic"'] \arrow[l, "\yield"]
+  \end{tikzcd}
+  \caption{The grammars considered, and their relation to each other.
+$G$ is provided by a user of the system, along with instructions how to modify $G$ to construct $G'$, while $T$ and $T'$ are automatically derived.}
+  \label{fig:grammar-square}
+\end{figure}
+
+%% \begin{definition}
+%%   A language defined by the semantic grammar $G$ and the parse grammar $G'$ is resolvable if:
+%%   $$
+%%   \begin{array}{l}
+%%   \forall t \in L(T).\\
+%%   \quad \exists w' \in L(G').\\
+%%   \qquad \forall t' \in L(T'). \yield(t') = w' -> \semantic(t') = t
+%%   \end{array}
+%%   $$
+%% \end{definition}
+
+\begin{definition}
+  A language defined by the semantic grammar $G$ and the parse grammar $G'$ is resolvable if:
+  $$
+  \begin{array}{l}
+  \forall t \in L(T).\\
+  \quad \exists w' \in L(G').\ \parse(w') = \{t\}
+  \end{array}
+  $$
+\end{definition}
+
+\noindent Intuitively, a language is resolvably ambiguous if there is at least one (unambiguous\footnote{For a slightly different meaning of unambiguous, it's here relating $G'$ to $T$, instead of $G'$ to $T'$.}) word for each semantically distinct tree.
+Note that neither $G$ nor $G'$ necessarily need to be unambiguous for this to hold.
+
+\section{Static Resolvability Check}
+
+This section describes a decision algorithm for detecting unresolvable ambiguities, starting with a version with several limitations, most of which are later lifted.
+
+\subsection{Basic Algorithm}
+
+Our initial limitations / assumptions are as follows:
+
+\begin{enumerate}
+  \item $G$ contains no parentheses, i.e., given $G = (S, P, \NT, \T, \I)$ we require $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.
+
+  \item $G'$ is constructed by adding parentheses to all non-terminals of $G$, i.e., $G' = (S, P', \NT, \T \cup \{\verb|'('|, \verb|')'|\}, \I \cup \NT)$ where $P' = P \cup \{ N: N -> \verb|'('|N\verb|')'| \mid N \in \NT \}$.
+
+  \item No production has a right-hand side that matches a single non-terminal, i.e., $\forall (i:N->r) \in P.\ \NT \cap R(r) = \emptyset$\footnote{$R$ is here the more traditional language of a regular expression, i.e., $R(r) \subseteq (\T \cup \NT)^{*}$.
+It should be written more explicitly somewhere, but I'm lazy at the moment.}.
+\end{enumerate}
+
+We are looking for a counterexample to the resolvable property, i.e., a $t \in L(T)$ for which $\lnot \exists w' \in L(G').\ \parse(w') = \{t\}$.
+By the construction of $G'$, we have $t \in \parse(\yield(t))$, thus there is at least one (potentialy ambiguous) word $w'$ that can be parsed as $t$.
+Our task is thus to find a tree $t$ such that $\forall w' \in L(G').\ t \in \parse(w') -> \{t\} \subset \parse(w')$, i.e., a tree that can only be parsed with ambiguous words.
+
+We begin by noting that parentheses cannot be added just anywhere; they must correspond to a node in the parse tree.
+For example, for the tree in Figure~\ref{fig:example-grammar:sem-tree} (which could be parsed from ''$1 + 2 * 3$'') it would be valid to add parentheses around ''$1 + 2$'' but not ''$2 * 3$''.
+As a consequence, adding parentheses restricts the possible parse trees, i.e., given two words $w'_1$ and $w'_2$ where the latter has added some parentheses we have $\parse(w'_1) \supseteq \parse(w'_2)$.
+We also note that adding double parentheses imposes no additional restriction, e.g., $\parse(\text{''}((1 + 2)) * 3\text{''})$ = $\parse(\text{''}(1 + 2) * 3\text{''})$.
+
+There is thus a ''most restrictive word'' for every tree.
+If this word is ambiguous, then we have our counterexample\footnote{There is one more thing to note here to motivate the pushdown automata: the only form of ambiguity we have is one where there is another tree $t_2$ such that if $t \in \parse(w')$ then $t_2 \in \parse(w')$.
+I am convinced this is true for the basic case, and I believe it is still true for the remaining cases, but it needs to be proved.
+As such, I'm unsure how to motivate it here, because I think it'll require more complicated reasoning in later cases, and it'd be nice if there could be some commonality between the explanations.}.
+
+We will now construct a pushdown automaton that recognizes these most restrictive words in such a way that there is a bijection between successful runs and trees in $L(T)$.
+As a running example, we will use the following (very simple) grammar $G$:
+
+\setlength{\grammarindent}{3.5em}
+\begin{grammar}
+  <N> ::= succ: 's' <N>
+    \alt zero: 'z'
+\end{grammar}
+
+\noindent We begin by constructing a DFA per production. This can be done in the standard way by constructing an NFA, then determinizing it, and optionally minimizing it.
+
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+
+  \node[state,initial] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+  \node[state,accepting] (3) [right of=2] {$3$};
+  \node[state,initial] (4) [below of=1, xshift=1cm] {$4$};
+  \node[state,accepting] (5) [right of=4] {$5$};
+
+  \path (1) edge              node {$s$} (2)
+  (2) edge              node {$N$} (3)
+  (4) edge              node {$z$} (5);
+
+\end{tikzpicture}
+\end{center}
+
+We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with ($+\gamma$ means ''push $\gamma$'', while $-\gamma$ means ''pop $\gamma$''):
+
+\begin{itemize}
+  \item An edge $p \xrightarrow{'(', +(p, q)} p'$ for every initial state $p'$ in some DFA belonging to non-terminal $N$.
+  \item An edge $q' \xrightarrow{')', -(p, q)} q$ for every final state $q'$ in some DFA belonging to non-terminal $N$.
+\end{itemize}
+
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+
+  \node[state,initial] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+  \node[state,accepting] (3) [right of=2] {$3$};
+  \node[state,initial] (4) [below of=1, xshift=1cm] {$4$};
+  \node[state,accepting] (5) [right of=4] {$5$};
+
+  \draw
+  (1) edge[above] node{$s$} (2)
+  (4) edge[above] node{$z$} (5)
+  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
+  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
+  (3) edge[loop above] node{$')', -(2, 3)$} (3)
+  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
+
+\end{tikzpicture}
+\end{center}
+
+\noindent Finally, we add a new initial state and a new final state, and connect them with the initial and final states belonging to the starting non-terminal:
+
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+
+  \node[state] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+  \node[state] (3) [right of=2] {$3$};
+  \node[state] (4) [below of=1, xshift=1cm] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+  \node[state,initial] (s) [left of=4] {};
+  \node[state,accepting] (e) [right of=5] {};
+
+  \draw
+  (s) edge[left] node{$'(', +s$} (1)
+  (s) edge[above] node{$'(', +s$} (4)
+  (3) edge[right] node{$')', -s$} (e)
+  (5) edge[above] node{$')', -s$} (e)
+  (1) edge[above] node{$s$} (2)
+  (4) edge[above] node{$z$} (5)
+  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
+  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
+  (3) edge[loop above] node{$')', -(2, 3)$} (3)
+  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
+
+\end{tikzpicture}
+\end{center}
+
+\noindent The resulting pushdown automaton has only a single source of non-determinism: the edges labelled \verb|'('|.
+Each one corresponds to one of the allowable child productions at that point in the parse tree.
+
+%% \noindent Generalizing this to a complete grammar, we get the following property: a grammar $G$ is resolvable iff:
+
+%% $$
+%% \begin{array}{l}
+%%   \forall t \in L(T).\\
+%%   \quad \exists w' \in L(G').\\
+%%   \qquad \forall t' \in L(T').\ \yield(t') = w' -> \unparen(t') = t
+%% \end{array}
+%% $$
 
@@
 
-%% Acknowledgments
-\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-  This material is based upon work supported by the
-  \grantsponsor{GS100000001}{National Science
-    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-  conclusions or recommendations expressed in this material are those
-  of the author and do not necessarily reflect the views of the
-  National Science Foundation.
-\end{acks}
+%% \begin{theorem}
+%%   If $\{(, )\} \cap \T = \emptyset$, then any unambiguous word $w \in L(G)$ is resolvable.
+%% \end{theorem}
+
+%% \noindent \textbf{Proof.} Since $w$ is unambiguous in $G$ there is exactly one parse tree $t \in L(T)$ such that $\yield(t) = w$.
+Since $L(G) \subseteq L(G')$ we have that $w \in L(G')$.
+Furthermore, $w$ is unambiguous in $L(G')$ as well, since no added productions can apply (they all contain parentheses, which cannot be present in $w$).
+Additionally, since $L(T) \subseteq L(T')$, we have that $t \in L(T')$, and since $\yield(t) = w$, it must be the only parse tree for $w$ in $L(T')$.
+Finally, $\unparen(t) = t$.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% %% Acknowledgments
+%% \begin{acks}                            %% acks environment is optional
+%%                                         %% contents suppressed with 'anonymous'
+%%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
+%%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
+%%   %% acknowledge financial support and will be used by metadata
+%%   %% extraction tools.
+%%   This material is based upon work supported by the
+%%   \grantsponsor{GS100000001}{National Science
+%%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
+%%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
+%%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
+%%   conclusions or recommendations expressed in this material are those
+%%   of the author and do not necessarily reflect the views of the
+%%   National Science Foundation.
+%% \end{acks}
 
 %% Bibliography
-%\bibliography{bibfile}
+\bibliography{All}
 
-%% Appendix
-\appendix
-\section{Appendix}
+%% %% Appendix
+%% \appendix
+%% \section{Appendix}
 
-Text of appendix \ldots
+%% Text of appendix \ldots
 

### 2019-02-26 paper/paper.tex
Described the superset automaton (though not with that name)

@@
 There is thus a ''most restrictive word'' for every tree.
-If this word is ambiguous, then we have our counterexample\footnote{There is one more thing to note here to motivate the pushdown automata: the only form of ambiguity we have is one where there is another tree $t_2$ such that if $t \in \parse(w')$ then $t_2 \in \parse(w')$.
-I am convinced this is true for the basic case, and I believe it is still true for the remaining cases, but it needs to be proved.
-As such, I'm unsure how to motivate it here, because I think it'll require more complicated reasoning in later cases, and it'd be nice if there could be some commonality between the explanations.}.
+If this word is ambiguous, then we have our counterexample.
+% TODO: motivate the other tree ''dominating'' or ''shadowing'' this one, to be able to use that term later
 
@@
 
-We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with ($+\gamma$ means ''push $\gamma$'', while $-\gamma$ means ''pop $\gamma$''):
+We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with:
 
 \begin{itemize}
-  \item An edge $p \xrightarrow{'(', +(p, q)} p'$ for every initial state $p'$ in some DFA belonging to non-terminal $N$.
-  \item An edge $q' \xrightarrow{')', -(p, q)} q$ for every final state $q'$ in some DFA belonging to non-terminal $N$.
+  \item an edge $p \xrightarrow{'(', +(p, q)} p'$ for every initial state $p'$ in some DFA belonging to non-terminal $N$, and
+  \item an edge $q' \xrightarrow{')', -(p, q)} q$ for every final state $q'$ in some DFA belonging to non-terminal $N$,
 \end{itemize}
+
+were $+\gamma$ means ''push $\gamma$'', while $-\gamma$ means ''pop $\gamma$''. Intuitively, parse a child node, but put parentheses around it.
 
@@
 
+We now have a pushdown automaton that recognizes the ''most restrictive word'' for each tree in $L(T)$, which we will call $A_($.
+Next we need to determine if it is a valid word for some other tree as well.
+To do this we create a copy of $A_($ with one modification: for every state $s$ in $A_($ (except the initial and final states), add two transitions $s \xrightarrow{'(', +p} s$ and $s \xrightarrow{')', -p}$.
+To avoid cluttering the graph too much, these transitions are shown unlabeled below:
+
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+
+  \node[state] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+  \node[state] (3) [right of=2] {$3$};
+  \node[state] (4) [below of=1, xshift=1cm] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+  \node[state,initial] (s) [left of=4] {};
+  \node[state,accepting] (e) [right of=5] {};
+
+  \draw
+  (1) edge[loop, in=80, out=100, looseness=7] node{} (1)
+  (1) edge[loop, in=70, out=110, looseness=6] node{} (1)
+  (2) edge[loop, in=80, out=100, looseness=7] node{} (2)
+  (2) edge[loop, in=70, out=110, looseness=6] node{} (2)
+  (3) edge[loop, in=-10, out=10, looseness=7] node{} (3)
+  (3) edge[loop, in=-20, out=20, looseness=6] node{} (3)
+  (4) edge[loop, in=-80, out=-100, looseness=7] node{} (4)
+  (4) edge[loop, in=-70, out=-110, looseness=6] node{} (4)
+  (5) edge[loop, in=-80, out=-100, looseness=7] node{} (5)
+  (5) edge[loop, in=-70, out=-110, looseness=6] node{} (5)
+  (s) edge[left] node{$'(', +s$} (1)
+  (s) edge[above] node{$'(', +s$} (4)
+  (3) edge[right] node{$')', -s$} (e)
+  (5) edge[above] node{$')', -s$} (e)
+  (1) edge[above] node{$s$} (2)
+  (4) edge[above] node{$z$} (5)
+  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
+  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
+  (3) edge[loop above] node{$')', -(2, 3)$} (3)
+  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
+
+\end{tikzpicture}
+\end{center}
+
+We will call this automaton $A'_($.
+Intuitively, this automaton recognizes a most restrictive word, but allows the addition of arbitrary well-nested parentheses.
+Successful runs in this automaton have a surjection to runs in $A_($ (ignore transitions along the newly added edges), and thus also have a surjection to parse trees in $L(T)$.
+We can also note that every sucessful run in $A_($ is also a successful run in $A'_($, since the latter has all states and transitions of the former.
+Furthermore, two distinct runs, one in $A_($ (call it $p$) and one in $A'_($ (call it $p'$), that both recognize the same word must represent different parse trees in $L(T)$.
+To see why, we consider two cases:
+
+\begin{enumerate}
+  \item $p'$ only uses transitions present in $A_($.
+This is a successful run in $A_($, and distinct from $p$.
+But there is a bijection between runs in $A_($ and parse trees in $L(T)$, thus $p'$ represents a different parse tree.
+  \item $p'$ uses at least one transition added in $A'_($.
+Using the surjection between runs in $A'_($ and $A_($ we find a new successful run that produces a different word (at least one fewer pair of parentheses).
+Since this run produces a different word, it must be distinct from $p$, and thus represent a different parse tree.
+\end{enumerate}
+
+% TODO: use ''dominating'' or ''shadowing'' here
+
 %% \noindent Generalizing this to a complete grammar, we get the following property: a grammar $G$ is resolvable iff:

### 2019-03-18 paper/paper.tex
Addressed Elias' feedback, incorporated lattice-think

@@
 \newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G') to a subset of L(T)
+\newcommand{\words}{\mathit{words}} % go from a t \in L(T) to a subset of L(G')
+\newcommand{\range}[2]{#1\!-\!#2}
 
@@
 \begin{quote} % TODO: remove
-  \textbf{NOTE:} The text hereafter is written a bit later than the text before, and is ever so slightly inconsistent with it.
+  \textbf{NOTE:} The text before this section is largely out of date.
 \end{quote}
+
+\subsection{Context-Free Grammars}
 
@@
       Regular expressions & $r ::= t \mid N \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
-      Productions & $i : N -> r$ \\
+      Productions & $N -> i : r$ \\
   \end{tabular}
@@
 $S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$, $\T$, and $\I$ are sets of non-terminals, terminals, and identifiers, respectively.
-Additionally, we require that $\NT \cap \T = \emptyset$, and that the identifiers uniquely identify each production, i.e., there are no two distinct productions in $P$ with the same identifier.
+Additionally, we require that $\NT \cap \T = \emptyset$.
+Most (though not all) grammars we will consider will have unique identifiers, i.e., there are no two distinct productions in $P$ with the same identifier.
+As an example, Figure~\ref{fig:example-grammar:ambig-grammar} contains a small grammar.
+The first production listed has the identifier ''Sum'' and consists of an expression, then a literal ''+'', then another expression.
 
@@
 
+\subsection{Automata}
+
+A nondeterministic finite automaton (NFA) is a tuple $(Q, \Sigma, \Delta, q_0, F)$:
+
+\begin{itemize}
+\item A finite set of states $Q$.
+\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
+\item A transition function $\delta: Q \times \Sigma -> 2^Q$.
+\item An initial state $q_0 \in Q$.
+\item A set of final states $F \subseteq Q$.
+\end{itemize}
+
+\noindent A successful run is a sequence of states $r_0, r_1, \ldots, r_n$ and a word $a_0a_1\ldots a_n$ such that:
+
+\begin{itemize}
+\item $r_0 = q_0$.
+\item $\forall_{i = 0}^{n-1} r_{i+1} \in \delta(r_i, a_i)$.
+\item $r_n \in F$.
+\end{itemize}
+
+\noindent We say that the automaton accepts the word $a_0a_1\ldots a_n$ iff there is such a succesful run.
+A particular state $q \in Q$ is \emph{reachable} if there is a (not necessarily successful) run starting in $q_0$ and ending in $q$.
+$q$ is \emph{coreachable} if there is a run starting in $q$ and ending in some $f \in F$.
+
+A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$.
+
+A pushdown automaton extends a finite automaton with a stack, and lets each transition push or pop symbols from it. Formally, a pushdown automaton is a tuple $(Q, \Sigma, \Gamma, \delta, q_0, F)$:
+
+\begin{itemize}
+\item A finite set of states $Q$.
+\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
+\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet.
+\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma*}$.
+\item An initial state $q_0 \in Q$.
+\item A set of final states $F \subseteq Q$.
+\end{itemize}
+
+\noindent A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma*$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma*$.
+Reachable and coreachable are now defined on configurations, rather than states.
+
+However, in this paper we will only consider pushdown automata with relatively limited stack manipulation, and will thus use some convenient shorthand:
+
+\begin{itemize}
+\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all.
+\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack.
+\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack (i.e., this transition cannot be taken if $g$ is not on top of the stack).
+\end{itemize}
+
+\subsection{Visibly Pushdown Languages}
+
+A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
+A visibly pushdown automaton is a pushdown automaton where the input alphabet $\Sigma$ can be partitioned into three disjoint sets $\Sigma_c$, $\Sigma_i$, and $\Sigma_r$, such that all transitions in the automaton has one of the following three forms:
+
+\begin{itemize}
+\item $p \xrightarrow{c, +s} q$, where $c \in \Sigma_c$.
+\item $p \xrightarrow{i} q$, where $i \in \Sigma_i$.
+\item $p \xrightarrow{r, -s} q$, where $r \in \Sigma_r$.
+\end{itemize}
+
+\noindent i.e., the terminal recognized by a transition fully determines the change to the stack height.
+
+This gives us some nice properties. Of particular relevance to this paper are the following two points:
+
+\begin{itemize}
+\item If two visibly pushdown automata have the same partitions $\Sigma_c$, $\Sigma_i$, and $\Sigma_r$, then we can construct a product automaton that simulates two simultaneous runs through both automata.
+This product automaton has the same input alphabet partitions, each state is a pair of states (one from each automaton), and each stack symbol is a pair of stack symbols (one from each automaton).
+\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified such that all reachable configurations are also coreachable.
+\end{itemize}
+
 \section{Parse-time Disambiguation}
 
-We begin this section with some motivation, and then list our definition of \emph{resolvable ambiguity}.
+We begin this section with some motivation, then list our definition of \emph{resolvable ambiguity} and what disambiguation we will consider, and finally an alternative definition of a word, which will be useful in later sections.
 
@@
 
-\begin{figure}
-  \begin{subfigure}{\linewidth}
+\begin{figure*}
+  \begin{subfigure}{.45\linewidth}
     \setlength{\grammarindent}{5em}
@@
   \end{subfigure}
-
-  \begin{subfigure}{\linewidth}
+  \begin{subfigure}{0.45\linewidth}
     \setlength{\grammarindent}{5em}
@@
 
-  \begin{subfigure}{\linewidth}
+  \begin{subfigure}{.55\linewidth}
     \begin{center}
@@
   \end{subfigure}
-
-  \begin{subfigure}{\linewidth}
+  \begin{subfigure}{.35\linewidth}
     \begin{center}
@@
   \label{fig:example-grammar}
-\end{figure}
+\end{figure*}
 
@@
   \item The parse grammar $G'$, a modification of $G$ meant to actually be used for parsing (e.g., Figure~\ref{fig:example-grammar:unambig-grammar}).
-The details of these modifications will be covered later.
-% TODO: likely put a forward reference here
+The identifiers in this grammar need not be unique, but should be a superset of the identifiers in $G$.
+The modifications we will consider in this paper are introduced in Section~\ref{sec:modifications}.
   \item $T_{G'}$ (generally abbreviated as $T'$), the parse trees of $G'$ (Figure~\ref{fig:example-grammar:tree} is in this language).
@@
 The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}.
-Finally, we define the function $\parse : L(G') -> 2^{L(T)}$:
+Finally, we define the function $\parse : L(G') -> 2^{L(T)}$, and its inverse $\words : L(T) -> 2^{L(G')}$:
 
 $$
-\parse(w') = \{ \semantic(t') \mid t' \in L(T') \land \yield(t') = w' \}
+\begin{array}{rcl}
+\parse(w') & = & \{ \semantic(t') \mid t' \in L(T') \land \yield(t') = w' \} \\
+\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
+\end{array}
 $$
+
+\noindent The latter will be useful later, while $\parse$ is used directly in the definition of resolvable ambiguity:
 
@@
 
-%% \begin{definition}
-%%   A language defined by the semantic grammar $G$ and the parse grammar $G'$ is resolvable if:
-%%   $$
-%%   \begin{array}{l}
-%%   \forall t \in L(T).\\
-%%   \quad \exists w' \in L(G').\\
-%%   \qquad \forall t' \in L(T'). \yield(t') = w' -> \semantic(t') = t
-%%   \end{array}
-%%   $$
-%% \end{definition}
-
 \begin{definition}
@@
   \begin{array}{l}
-  \forall t \in L(T).\\
+  \forall t \in L(T_G).\\
   \quad \exists w' \in L(G').\ \parse(w') = \{t\}
@@
 
+\subsection{Modifications in G'} \label{sec:modifications} % TODO: can't typeset G' in mathmode for some reason
+
+For this paper we consider two possible modifications:
+
+\begin{itemize}
+\item Adding parentheses for explicit grouping.
+\item Forbidding non-terminals in certain positions from expanding using certain productions.
+\end{itemize}
+
+\noindent The latter requires more explanation. As an example, in the following grammar we are forbidding the second \synt{<expr>} in the first production from expanding using the ''Sum'' production:
+
+\setlength{\grammarindent}{5em}
+\begin{grammar}
+  <expr> ::= Sum: <expr> '+' <expr>$_{Sum}$
+    \alt Number: <number>
+\end{grammar}
+
+\noindent Transforming this into a normal grammar, and adding parentheses, we get the following:
+
+\setlength{\grammarindent}{6em}
+\begin{grammar}
+  <expr> ::= Sum: <expr> '+' <expr>$_{Sum}$
+  \alt Number: <number>
+  \alt Paren: '(' <expr> ')'
+
+  <expr$_{Sum}$> ::= Number: <number>
+  \alt Paren: '(' <expr> ')'
+\end{grammar}
+
+\noindent In other words, we produce a grammar where addition is left-associative\footnote{I'm having some issues typesetting this appropriately, and I'm not all that happy with the explanation.}, but we can use explicit grouping via parentheses to override that.
+
+\subsection{An Alternative Word View} \label{sec:word-view}
+
+This section introduces an alternative (isomorphic) definition of a word that will be used to motivate the correctness of later algorithms.
+The modifications permitted by the previous section significantly limit how $G'$ can differ from $G$.
+In particular, no new terminals are introduced, except '(' and ')', and they are always introduced in a well-balanced fashion.
+
+If we thus delimit ourselves to only consider languages where words have no unbalanced parentheses\footnote{I.e., the vast, vast majority of programming languages currently in use.} we can give the following alternative definition of a word: a word is a two-tuple containing a sequence of non-parenthesis terminals and a bag (or multiset) of ranges covered by parentheses.
+For example, the word ''$(1 + 2) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{1}{3}\})$, while ''$((1 + 2)) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{1}{3}, \range{1}{3}\})$.
+The first component will be referred to as a \emph{basic word}, the second as a \emph{range bag}
+
+We will now note a few things about the words $w' \in \words(t)$ for some given $t \in L(T)$:
+
+\begin{itemize}
+\item All $w'$ have the same basic word.
+Limiting the modifications to those in the previous section means that they can only differ in parentheses.
+This also implies that two trees that share a word (i.e., that can be ambiguous) must also share a basic word.
+\item Some parentheses are required, i.e., some ranges must be present in all $w'$.
+For example, removing the parentheses in ''$(1 + 2) * 3$'' changes the tree produced, since multiplication has higher precedence than addition.
+\item There is a finite set of possible ranges in the range bags of $w'$.
+Grouping parentheses can only be added if they exactly cover a node in the parse tree, and other parentheses can only be added where $G$ allows them.
+\item Duplicated grouping parentheses do not matter.
+For example, $((1 + 2)) * 3$ permits the same syntax trees as $(1 + 2) * 3$.
+If there are no parentheses present in $G$ then all parentheses are grouping parentheses, thus we can consider the range bag as a \emph{set} instead.
+\end{itemize}
+
 \section{Static Resolvability Check}
 
-This section describes a decision algorithm for detecting unresolvable ambiguities, starting with a version with several limitations, most of which are later lifted.
+This section describes a decision algorithm for detecting unresolvable ambiguities in a pair of grammars $G$ and $G'$, starting with a version with several limitations, most of which are later lifted.
+The overarching goal is to find a tree $t \in L(T)$ such that there is no $w' \in L(G')$ for which $\parse(w') = \{t\}$.
+Alternatiely, find a tree that has no unambiguous words.
 
@@
   \item $G$ contains no parentheses, i.e., given $G = (S, P, \NT, \T, \I)$ we require $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.
+This implies that all parentheses in $G'$ are grouping parentheses, thus we only need to consider words $w' \in L(G')$ whose range bag is a set.
 
-  \item $G'$ is constructed by adding parentheses to all non-terminals of $G$, i.e., $G' = (S, P', \NT, \T \cup \{\verb|'('|, \verb|')'|\}, \I \cup \NT)$ where $P' = P \cup \{ N: N -> \verb|'('|N\verb|')'| \mid N \in \NT \}$.
+  \item $G'$ is constructed by adding parentheses to all non-terminals of $G$, with no other modifications, i.e., $G' = (S, P', \NT, \T \cup \{\verb|'('|, \verb|')'|\}, \I \cup \NT)$ where $P' = P \cup \{ N: N -> \verb|'('|N\verb|')'| \mid N \in \NT \}$.
+This implies that there are no required grouping parentheses.
 
@@
 
-We are looking for a counterexample to the resolvable property, i.e., a $t \in L(T)$ for which $\lnot \exists w' \in L(G').\ \parse(w') = \{t\}$.
-By the construction of $G'$, we have $t \in \parse(\yield(t))$, thus there is at least one (potentialy ambiguous) word $w'$ that can be parsed as $t$.
-Our task is thus to find a tree $t$ such that $\forall w' \in L(G').\ t \in \parse(w') -> \{t\} \subset \parse(w')$, i.e., a tree that can only be parsed with ambiguous words.
+\noindent The first two limitations allow us to place all words in $\words(t)$ for some given $t$ in a lattice, whose structure is formed by the subset ordering of the range sets.
+For example, the tree in Figure~\ref{fig:example-grammar:sem-tree} (but where $G'$ has no associativity or precedence) has the lattice of words seen in Figure~\ref{fig:lattice} (omitting the pair of parentheses around the entire word, since they are always permissible).
 
-We begin by noting that parentheses cannot be added just anywhere; they must correspond to a node in the parse tree.
-For example, for the tree in Figure~\ref{fig:example-grammar:sem-tree} (which could be parsed from ''$1 + 2 * 3$'') it would be valid to add parentheses around ''$1 + 2$'' but not ''$2 * 3$''.
-As a consequence, adding parentheses restricts the possible parse trees, i.e., given two words $w'_1$ and $w'_2$ where the latter has added some parentheses we have $\parse(w'_1) \supseteq \parse(w'_2)$.
-We also note that adding double parentheses imposes no additional restriction, e.g., $\parse(\text{''}((1 + 2)) * 3\text{''})$ = $\parse(\text{''}(1 + 2) * 3\text{''})$.
+\begin{figure*}[t]
+\begin{tikzpicture}%[scale = 0.7,font=\tiny]
+  \node (a) at (0,5) {$((1) + (2)) * (3)$};
 
-There is thus a ''most restrictive word'' for every tree.
-If this word is ambiguous, then we have our counterexample.
-% TODO: motivate the other tree ''dominating'' or ''shadowing'' this one, to be able to use that term later
+  \node (b1) at (-3.6,4.2) {$(1) + (2) * (3)$};
+  \node (b2) at (-1.2,4.2) {$((1) + (2)) * 3$};
+  \node (b3) at ( 1.2,4.2) {$((1) + 2) * (3)$};
+  \node (b4) at ( 3.6,4.2) {$(1 + (2)) * (3)$};
 
-We will now construct a pushdown automaton that recognizes these most restrictive words in such a way that there is a bijection between successful runs and trees in $L(T)$.
+  \node (c1) at (-5.5,3) {$(1) + (2) * 3$};
+  \node (c2) at (-3.3,3) {$(1) + 2 * (3)$};
+  \node (c3) at (-1.1,3) {$((1) + 2) * 3$};
+  \node (c4) at (1.1,3) {$1 + (2) * (3)$};
+  \node (c5) at (3.3,3) {$(1 + (2)) * 3$};
+  \node (c6) at (5.5,3) {$(1 + 2) * (3)$};
+
+  \node (d1) at (-3.6,1.8) {$(1) + 2 * 3$};
+  \node (d2) at (-1.2,1.8) {$1 + (2) * 3$};
+  \node (d3) at ( 1.2,1.8) {$1 + 2 * (3)$};
+  \node (d4) at ( 3.6,1.8) {$(1 + 2) * 3$};
+
+  \node (e) at (0, 1) {$1 + 2 * 3$};
+
+  \draw (e) -- (d1);
+  \draw (e) -- (d2);
+  \draw (e) -- (d3);
+  \draw (e) -- (d4);
+
+  \draw (d1) -- (c1);
+  \draw (d1) -- (c2);
+  \draw (d1) -- (c3);
+  \draw (d2) -- (c1);
+  \draw (d2) -- (c4);
+  \draw (d2) -- (c5);
+  \draw (d3) -- (c2);
+  \draw (d3) -- (c4);
+  \draw (d3) -- (c6);
+  \draw (d4) -- (c3);
+  \draw (d4) -- (c5);
+  \draw (d4) -- (c6);
+
+  \draw (c1) -- (b1);
+  \draw (c1) -- (b2);
+  \draw (c2) -- (b1);
+  \draw (c2) -- (b3);
+  \draw (c3) -- (b2);
+  \draw (c3) -- (b3);
+  \draw (c4) -- (b1);
+  \draw (c4) -- (b4);
+  \draw (c5) -- (b2);
+  \draw (c5) -- (b4);
+  \draw (c6) -- (b3);
+  \draw (c6) -- (b4);
+
+  \draw (b1) -- (a);
+  \draw (b2) -- (a);
+  \draw (b3) -- (a);
+  \draw (b4) -- (a);
+%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
+\end{tikzpicture}
+\caption{The lattice of words for the tree in Figure~\ref{fig:example-grammar:sem-tree}, assuming no precedence in $G'$.}
+\label{fig:lattice}
+\end{figure*}
+
+The bottom word corresponds to the empty range set, while the top word corresponds to the set of possible parentheses ranges.
+All trees that can be ambiguous must thus share the same bottom word, while the top may differ.
+
+Of particular use is to examine whether the top word (call it $w'$) is ambiguous. There are two cases:
+
+\begin{itemize}
+\item The top word is unambiguous, then this is not a tree we are looking for.
+\item It is ambiguous.
+That means that there is some other lattice for some other tree that also contains the word.
+Call the top word of this other lattice $w'_2$.
+For $w'$ to be in the lattice whose top is $w'_2$ its rangeset must be a subset of the rangeset of the latter.
+But this is true also for all the other words in the lattice, thus there are no unambiguous words in the lattice, i.e., this tree is what we are looking for.
+\end{itemize}
+
+The algorithm thus looks for two trees, where the set of possible parentheses for one is a superset of the set of possible parentheses for the other.
+To do this, we use a linear representation of each lattice: namely the top word.
+If two trees have the same top word, that implies that their sets of possible parentheses are equal.
+If we can add parentheses to one word and get the other word, that implies that the rangeset of the former is a subset of the latter.
+that term later
+
+We will now construct a pushdown automaton that recognizes these top words in such a way that there is a bijection between successful runs and trees in $L(T)$.
 As a running example, we will use the following (very simple) grammar $G$:
@@
 
-We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with:
+\noindent We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with:
 
@@
 
-were $+\gamma$ means ''push $\gamma$'', while $-\gamma$ means ''pop $\gamma$''. Intuitively, parse a child node, but put parentheses around it.
+\noindent Intuitively, we parse a child node, but put parentheses around it. This is where we use assumption 3, without it we might introduce double parentheses here.
 
@@
 
-We now have a pushdown automaton that recognizes the ''most restrictive word'' for each tree in $L(T)$, which we will call $A_($.
-Next we need to determine if it is a valid word for some other tree as well.
-To do this we create a copy of $A_($ with one modification: for every state $s$ in $A_($ (except the initial and final states), add two transitions $s \xrightarrow{'(', +p} s$ and $s \xrightarrow{')', -p}$.
+We now have a pushdown automaton (call it $A_{()}$) that recognizes the top word for each tree in $L(T)$.
+Next we need to be able to add arbitrary parentheses, to produce a word with a rangeset superset.
+To do this we create a copy of $A_{()}$ with one modification: for every state $s$ in $A_{()}$ (except the initial and final states), add two transitions $s \xrightarrow{'(', +p} s$ and $s \xrightarrow{')', -p} s$.
 To avoid cluttering the graph too much, these transitions are shown unlabeled below:
@@
 
-We will call this automaton $A'_($.
-Intuitively, this automaton recognizes a most restrictive word, but allows the addition of arbitrary well-nested parentheses.
+We will call this automaton $A'_{()}$.
 Successful runs in this automaton have a surjection to runs in $A_($ (ignore transitions along the newly added edges), and thus also have a surjection to parse trees in $L(T)$.
-We can also note that every sucessful run in $A_($ is also a successful run in $A'_($, since the latter has all states and transitions of the former.
-Furthermore, two distinct runs, one in $A_($ (call it $p$) and one in $A'_($ (call it $p'$), that both recognize the same word must represent different parse trees in $L(T)$.
+We can also note that every sucessful run in $A_{()}$ is also a successful run in $A'_{()}$, since the latter has all states and transitions of the former.
+Furthermore, two distinct runs, one in $A_{()}$ (call it $p$) and one in $A'_{()}$ (call it $p'$), that both recognize the same word must represent different parse trees in $L(T)$.
 To see why, we consider two cases:
@@
 \begin{enumerate}
-  \item $p'$ only uses transitions present in $A_($.
-This is a successful run in $A_($, and distinct from $p$.
-But there is a bijection between runs in $A_($ and parse trees in $L(T)$, thus $p'$ represents a different parse tree.
-  \item $p'$ uses at least one transition added in $A'_($.
-Using the surjection between runs in $A'_($ and $A_($ we find a new successful run that produces a different word (at least one fewer pair of parentheses).
+  \item $p'$ only uses transitions present in $A_{()}$.
+This is a successful run in $A_{()}$, and distinct from $p$.
+But there is a bijection between runs in $A_{()}$ and parse trees in $L(T)$, thus $p'$ represents a different parse tree.
+  \item $p'$ uses at least one transition added in $A'_{()}$.
+Using the surjection between runs in $A'_{()}$ and $A_{()}$ we find a new successful run that produces a different word (at least one fewer pair of parentheses).
 Since this run produces a different word, it must be distinct from $p$, and thus represent a different parse tree.
@@
 
-% TODO: use ''dominating'' or ''shadowing'' here
+Two distinct successful runs that accept the same word thus represent two trees where one permits a superset rangeset of the other.
+To find such runs we construct a product automaton and trim it.
+We can construct a product automaton since both $A_{()}$ and $A'_{()}$ are visibly pushdown, with the same partition of the input alphabet (push on open parenthesis, pop on close parenthesis, do nothing otherwise).
+We can trim the product since it retains the same partitioning and thus is also visibly pushdown.
+
+In this product automaton, if any transition pushes a stack symbol $(a, b)$ where $a \neq b$, or transitions to a state $(p, q)$ where $p \neq q$, then there is a successful run that corresponds to two distinct runs through $A_{()}$ and $A'_{()}$ (since the automaton is trim).
+
+\subsection{Forbidden Children}
+
+\begin{quote}
+  \textbf{NOTE:} I had some text here about version 2 of the problem, but I haven't had time to rewrite it to align properly with the lattice formulation, so I've removed it.
+I'll talk about it during the meeting.
+Also, I probably need a better title here, this is probably a smidge too dramatic.
+\end{quote}
+
+\section{Parsetime Ambiguity Reporting} \label{sec:parse-time-reporting}
 

### 2019-03-26 paper/paper.tex
New figures, lightly edited later sections

@@
 This section introduces an alternative (isomorphic) definition of a word that will be used to motivate the correctness of later algorithms.
-The modifications permitted by the previous section significantly limit how $G'$ can differ from $G$.
+The method by which we generate $G_w$ and $G'_w$ limits the possible differences between them significantly.
 In particular, no new terminals are introduced, except '(' and ')', and they are always introduced in a well-balanced fashion.
@@
 
-We will now note a few things about the words $w' \in \words(t)$ for some given $t \in L(T)$:
+We will now note a few things about the words $w' \in \words(t)$ for some given $t \in L(G_t)$:
 
@@
 \item All $w'$ have the same basic word.
-Limiting the modifications to those in the previous section means that they can only differ in parentheses.
+$G'_w$ is constructed in such a way that the only possible difference is one of parentheses.
 This also implies that two trees that share a word (i.e., that can be ambiguous) must also share a basic word.
@@
 \item There is a finite set of possible ranges in the range bags of $w'$.
-Grouping parentheses can only be added if they exactly cover a node in the parse tree, and other parentheses can only be added where $G$ allows them.
+Grouping parentheses can only be added if they exactly cover a node in the parse tree, and other parentheses can only be added where $G_w$ allows them.
 \item Duplicated grouping parentheses do not matter.
 For example, $((1 + 2)) * 3$ permits the same syntax trees as $(1 + 2) * 3$.
-If there are no parentheses present in $G$ then all parentheses are grouping parentheses, thus we can consider the range bag as a \emph{set} instead.
+If there are no parentheses present in $G_w$ then all parentheses are grouping parentheses, thus we can consider the range bag as a \emph{set} instead.
 \end{itemize}
@@
 
-This section describes a decision algorithm for detecting unresolvable ambiguities in a pair of grammars $G$ and $G'$, starting with a version with several limitations, most of which are later lifted.
-The overarching goal is to find a tree $t \in L(T)$ such that there is no $w' \in L(G')$ for which $\parse(w') = \{t\}$.
-Alternatiely, find a tree that has no unambiguous words.
+This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted.
+The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$.
+Alternatively, find a tree that has no unambiguous words.
 
@@
 \begin{enumerate}
-  \item $G$ contains no parentheses, i.e., given $G = (S, P, \NT, \T, \I)$ we require $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.
-This implies that all parentheses in $G'$ are grouping parentheses, thus we only need to consider words $w' \in L(G')$ whose range bag is a set.
-
-  \item $G'$ is constructed by adding parentheses to all non-terminals of $G$, with no other modifications, i.e., $G' = (S, P', \NT, \T \cup \{\verb|'('|, \verb|')'|\}, \I \cup \NT)$ where $P' = P \cup \{ N: N -> \verb|'('|N\verb|')'| \mid N \in \NT \}$.
-This implies that there are no required grouping parentheses.
-
-  \item No production has a right-hand side that matches a single non-terminal, i.e., $\forall (i:N->r) \in P.\ \NT \cap R(r) = \emptyset$\footnote{$R$ is here the more traditional language of a regular expression, i.e., $R(r) \subseteq (\T \cup \NT)^{*}$.
-It should be written more explicitly somewhere, but I'm lazy at the moment.}.
+  \item The input language definition contains no parentheses.
+This implies that all parentheses in $G'_w$ are grouping parentheses, thus we only need to consider words $w' \in L(G'_w)$ whose range bag is a set.
+
+  \item No non-terminals in the input language definition are marked, i.e., there are no forbidden children. This implies that there are no required grouping parentheses.
+
+  \item No production has a right-hand side that matches a single non-terminal, i.e., $\forall (N->i:r) \in D.\ \NT \cap L(r) = \emptyset$\footnote{I haven't really written a proper description for the language definition, but I'm here using $D$ for it, and considering it essentially as a set of ''productions''.}.
 \end{enumerate}
@@
 \noindent The first two limitations allow us to place all words in $\words(t)$ for some given $t$ in a lattice, whose structure is formed by the subset ordering of the range sets.
-For example, the tree in Figure~\ref{fig:example-grammar:sem-tree} (but where $G'$ has no associativity or precedence) has the lattice of words seen in Figure~\ref{fig:lattice} (omitting the pair of parentheses around the entire word, since they are always permissible).
-
-\begin{figure*}[t]
-\begin{tikzpicture}%[scale = 0.7,font=\tiny]
-  \node (a) at (0,5) {$((1) + (2)) * (3)$};
-
-  \node (b1) at (-3.6,4.2) {$(1) + (2) * (3)$};
-  \node (b2) at (-1.2,4.2) {$((1) + (2)) * 3$};
-  \node (b3) at ( 1.2,4.2) {$((1) + 2) * (3)$};
-  \node (b4) at ( 3.6,4.2) {$(1 + (2)) * (3)$};
-
-  \node (c1) at (-5.5,3) {$(1) + (2) * 3$};
-  \node (c2) at (-3.3,3) {$(1) + 2 * (3)$};
-  \node (c3) at (-1.1,3) {$((1) + 2) * 3$};
-  \node (c4) at (1.1,3) {$1 + (2) * (3)$};
-  \node (c5) at (3.3,3) {$(1 + (2)) * 3$};
-  \node (c6) at (5.5,3) {$(1 + 2) * (3)$};
-
-  \node (d1) at (-3.6,1.8) {$(1) + 2 * 3$};
-  \node (d2) at (-1.2,1.8) {$1 + (2) * 3$};
-  \node (d3) at ( 1.2,1.8) {$1 + 2 * (3)$};
-  \node (d4) at ( 3.6,1.8) {$(1 + 2) * 3$};
-
-  \node (e) at (0, 1) {$1 + 2 * 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1) -- (c1);
-  \draw (d1) -- (c2);
-  \draw (d1) -- (c3);
-  \draw (d2) -- (c1);
-  \draw (d2) -- (c4);
-  \draw (d2) -- (c5);
-  \draw (d3) -- (c2);
-  \draw (d3) -- (c4);
-  \draw (d3) -- (c6);
-  \draw (d4) -- (c3);
-  \draw (d4) -- (c5);
-  \draw (d4) -- (c6);
-
-  \draw (c1) -- (b1);
-  \draw (c1) -- (b2);
-  \draw (c2) -- (b1);
-  \draw (c2) -- (b3);
-  \draw (c3) -- (b2);
-  \draw (c3) -- (b3);
-  \draw (c4) -- (b1);
-  \draw (c4) -- (b4);
-  \draw (c5) -- (b2);
-  \draw (c5) -- (b4);
-  \draw (c6) -- (b3);
-  \draw (c6) -- (b4);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-\end{tikzpicture}
-\caption{The lattice of words for the tree in Figure~\ref{fig:example-grammar:sem-tree}, assuming no precedence in $G'$.}
-\label{fig:lattice}
-\end{figure*}
-
-The bottom word corresponds to the empty range set, while the top word corresponds to the set of possible parentheses ranges.
+For example, the two trees in Figure~\ref{fig:ambig-example-square} have the lattices of words seen in Figures~\ref{fig:lattice1} and \ref{fig:lattice2} respectively.
+
+The bottom word corresponds to the set of required parentheses (i.e., the empty set, in this sub-problem), while the top word corresponds to the set of possible parentheses ranges.
 All trees that can be ambiguous must thus share the same bottom word, while the top may differ.
@@
 \begin{itemize}
-\item The top word is unambiguous, then this is not a tree we are looking for.
+\item The top word is unambiguous, then this is not a tree we are looking for (since it has at least one unambiguous word).
 \item It is ambiguous.
@@
 
-The algorithm thus looks for two trees, where the set of possible parentheses for one is a superset of the set of possible parentheses for the other.
+\noindent The algorithm thus looks for two trees, where the set of possible parentheses for one is a superset of the other.
 To do this, we use a linear representation of each lattice: namely the top word.
@@
 If we can add parentheses to one word and get the other word, that implies that the rangeset of the former is a subset of the latter.
-that term later
-
-We will now construct a pushdown automaton that recognizes these top words in such a way that there is a bijection between successful runs and trees in $L(T)$.
-As a running example, we will use the following (very simple) grammar $G$:
-
-\setlength{\grammarindent}{3.5em}
-\begin{grammar}
-  <N> ::= succ: 's' <N>
-    \alt zero: 'z'
-\end{grammar}
+
+We will now construct a pushdown automaton that recognizes these top words in such a way that there is a bijection between successful runs and trees in $L(G_t)$.
+As a running example, we will use the following (very simple) language definition $D$:
+
+\begin{center}
+\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+  \synt{N} & Succ & 's' \synt{N}\\
+  \synt{N} & Zero & 'z' \\
+\end{tabular}
+\end{center}
 
@@
 
-\noindent Intuitively, we parse a child node, but put parentheses around it. This is where we use assumption 3, without it we might introduce double parentheses here.
+\noindent Intuitively, we parse a child node, but put parentheses around it, then return. This is where we use assumption 3, without it we might introduce double parentheses here.
 
@@
 
-We now have a pushdown automaton (call it $A_{()}$) that recognizes the top word for each tree in $L(T)$.
+We now have a pushdown automaton (call it $A_{()}$) that recognizes the top word for each tree in $L(G_t)$.
 Next we need to be able to add arbitrary parentheses, to produce a word with a rangeset superset.
@@
 
-We will call this automaton $A'_{()}$.
-Successful runs in this automaton have a surjection to runs in $A_($ (ignore transitions along the newly added edges), and thus also have a surjection to parse trees in $L(T)$.
+\noindent We will call this automaton $A'_{()}$.
+Successful runs in this automaton have a surjection to runs in $A_($ (ignore transitions along the newly added edges), and thus also have a surjection to parse trees in $L(G_t)$.
 We can also note that every sucessful run in $A_{()}$ is also a successful run in $A'_{()}$, since the latter has all states and transitions of the former.
-Furthermore, two distinct runs, one in $A_{()}$ (call it $p$) and one in $A'_{()}$ (call it $p'$), that both recognize the same word must represent different parse trees in $L(T)$.
+Furthermore, two distinct runs, one in $A_{()}$ (call it $p$) and one in $A'_{()}$ (call it $p'$), that both recognize the same word must represent different parse trees in $L(G_t)$.
 To see why, we consider two cases:
@@
 
-Two distinct successful runs that accept the same word thus represent two trees where one permits a superset rangeset of the other.
+\noindent Two distinct successful runs that accept the same word thus represent two trees where one permits a superset rangeset of the other.
 To find such runs we construct a product automaton and trim it.
@@
 \end{quote}
+
+% TODO: this section should be moved to be in the appropriate position {
+\newpage
+
+\begin{figure*}[ht]
+  \begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+    \synt{Expr} & Sum & \synt{Expr} \verb|'+'| \synt{Expr} \\
+    \synt{Expr} & Product & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ \\
+    \synt{Expr} & Number & \synt{Number} \\
+  \end{tabular}
+  \caption{The input language definition used as a running example, a simple expression language with addition and multiplication, with precedence defined, but not associativity.
+Assumes that \synt{Number} matches a numeric terminal.}
+\end{figure*}
+
+\begin{figure*}[ht]
+  \begin{tikzcd}
+    L(G_w) & L(G_t) \arrow[l, "\yield"] \\
+    L(G'_w) & L(G'_t) \arrow[u, "\semantic"'] \arrow[l, "\yield"]
+  \end{tikzcd}
+  \caption{The generated grammars and their relation to each other.}
+\end{figure*}
+
+\begin{figure*}[ht]
+  \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      \toprule
+      \synt{Expr} & \synt{Expr} \verb|'+'| \synt{Expr} \\
+      \synt{Expr} & \synt{Expr} \verb|'*'| \synt{Expr} \\
+      \synt{Expr} & \synt{Number} \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G_w$, the generated abstract syntax.}
+  \end{subfigure}%
+%
+  \begin{subfigure}[t]{.54\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
+      \toprule
+      \synt{Expr} & $[_{Sum}$ & \synt{Expr} \verb|'+'| \synt{Expr} & $]_{Sum}$ \\
+      \synt{Expr} & $[_{Product}$ & \synt{Expr} \verb|'*'| \synt{Expr} & $]_{Product}$ \\
+      \synt{Expr} & $[_{Number}$ & \synt{Number} & $]_{Number}$ \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G_t$, a language describing a linear encoding of the parse trees of $G_w$.}
+  \end{subfigure}
+
+  \begin{subfigure}{.45\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      \toprule
+      \synt{Expr} & \synt{Expr} \verb|'+'| \synt{Expr} \\
+      \synt{Expr} & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ \\
+      \synt{Expr} & \synt{Number} \\
+      \synt{Expr} & \verb|'('| \synt{Expr} \verb|')'| \\
+      \midrule
+      \synt{Expr}$_{Sum}$ & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ \\
+      \synt{Expr}$_{Sum}$ & \synt{Number} \\
+      \synt{Expr}$_{Sum}$ & \verb|'('| \synt{Expr} \verb|')'| \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G'_w$, the generated concrete syntax.}
+  \end{subfigure}%
+%
+  \begin{subfigure}{.54\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
+      \toprule
+      \synt{Expr} & $[_{Sum}$ & \synt{Expr} \verb|'+'| \synt{Expr} & $]_{Sum}$ \\
+      \synt{Expr} & $[_{Product}$ & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ & $]_{Product}$ \\
+      \synt{Expr} & $[_{Number}$ & \synt{Number} & $]_{Number}$ \\
+      \synt{Expr} & $[_{Grouping}$ & \verb|'('| \synt{Expr} \verb|')'| & $]_{Grouping}$ \\
+      \midrule
+      \synt{Expr}$_{Sum}$ & $[_{Sum2}$ & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ & $]_{Sum2}$ \\
+      \synt{Expr}$_{Sum}$ & $[_{Number2}$ & \synt{Number} & $]_{Number2}$ \\
+      \synt{Expr}$_{Sum}$ & $[_{Grouping2}$ & \verb|'('| \synt{Expr} \verb|')'| & $]_{Grouping2}$ \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G'_t$, a language describing a linear encoding of the parse trees of $G'_w$.}
+  \end{subfigure}%
+  \caption{The generated grammars.}
+\end{figure*}
+
+\begin{figure*}[ht]
+  \begin{subfigure}[b]{.45\linewidth}
+    \[1 + 2 * 3\]
+    \caption{Example word in $L(G_w)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.45\linewidth}
+    \begin{center}
+    \Tree [.Product
+        [.Sum
+          [.Number 1 ]
+          '+'
+          [.Number 2 ] ]
+        '*'
+        [.Number 3 ] ]
+    \end{center}
+    \[[_P [_S [_N 1 ]_N + [_N 2 ]_N ]_S * [_N 3 ]_N ]_P\]
+    \caption{Example word in $L(G_T)$ and visualized tree.}
+  \end{subfigure}
+
+  \begin{subfigure}[b]{.45\linewidth}
+    \[(1 + 2) * 3\]
+    \caption{Example word in $L(G'_w)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.45\linewidth}
+    \begin{center}
+    \Tree [.Product
+        [.Grouping2
+          '('
+          [.Sum
+            [.Number 1 ]
+            '+'
+            [.Number 2 ] ]
+          ')' ]
+        '*'
+        [.Number2 3 ] ]
+    \end{center}
+    \[[_P [_{G2} ( [_S [_N 1 ]_N + [_N 2 ]_N ]_S ) ]_{G2} * [_{N2} 3 ]_{N2} ]_P\]
+    \caption{Example word in $L(G'_T)$ and visualized tree.}
+  \end{subfigure}
+  \caption{Example with words from each grammar that correspond to each other.
+The subscripts are abbreviated as follows: $S = \text{Sum}$ , $P = \text{Product}$ , $N = \text{Number}$ , and $G = \text{Grouping}$ .}
+\end{figure*}
+
+\begin{figure*}[ht]
+  \begin{subfigure}[b]{.45\linewidth}
+    \begin{align*}
+      1 + 2 + 3
+    \end{align*}
+    \caption{Example word in $L(G_w)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.45\linewidth}
+      \begin{align*}
+        [_S           [_S  &[_N 1 ]_N + \hphantom{[_S} [_N 2 ]_N           ]_S  + [_N 3 ]_N \hphantom{]_S} ]_S \\
+        [_S \hphantom{[_S} &[_N 1 ]_N +           [_S  [_N 2 ]_N \hphantom{]_S} + [_N 3 ]_N           ]_S  ]_S
+      \end{align*}
+    \caption{Example words in $L(G_T)$.}
+  \end{subfigure}
+
+  \begin{subfigure}[b]{.45\linewidth}
+    \begin{align*}
+      1 + 2 + 3
+    \end{align*}
+    \caption{Example word in $L(G'_w)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.45\linewidth}
+      \begin{align*}
+        [_S            [_S &[_N 1 ]_N + \hphantom{[_S} [_N 2 ]_N            ]_S + [_N 3 ]_N \hphantom{]_S}]_S \\
+        [_S \hphantom{[_S} &[_N 1 ]_N +            [_S [_N 2 ]_N \hphantom{]_S} + [_N 3 ]_N ]_S ]_S
+      \end{align*}
+    \caption{Example words in $L(G'_T)$.}
+  \end{subfigure}
+  \caption{Example with an ambiguous word in $L(G'_w)$ with corresponding words from the other grammars.}
+  \label{fig:ambig-example-square}
+\end{figure*}
+
+\begin{figure*}[ht]
+  \begin{tikzpicture}[every node/.style={rectangle,draw}]
+    \node (a) [draw=white] at (0,5) {$((1) + (2)) + (3)$};
+
+    \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
+    \node (b2) [draw=white] at (-1.2,4.2) {$((1) + (2)) + 3$};
+    \node (b3) [draw=white] at ( 1.2,4.2) {$((1) + 2) + (3)$};
+    \node (b4) [draw=white] at ( 3.6,4.2) {$(1 + (2)) + (3)$};
+
+    \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
+    \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
+    \node (c3) [draw=white] at (-1.1,3) {$((1) + 2) + 3$};
+    \node (c4) at (1.1,3) {$1 + (2) + (3)$};
+    \node (c5) [draw=white] at (3.3,3) {$(1 + (2)) + 3$};
+    \node (c6) [draw=white] at (5.5,3) {$(1 + 2) + (3)$};
+
+    \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
+    \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
+    \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
+    \node (d4) [draw=white] at ( 3.6,1.8) {$(1 + 2) + 3$};
+
+    \node (e) at (0, 1) {$1 + 2 + 3$};
+
+  \draw (e) -- (d1);
+  \draw (e) -- (d2);
+  \draw (e) -- (d3);
+  \draw (e) -- (d4);
+
+  \draw (d1.90) -- (c1.270);
+  \draw (d1.90) -- (c2.270);
+  \draw (d1.90) -- (c3.270);
+  \draw (d2.90) -- (c1.270);
+  \draw (d2.90) -- (c4.270);
+  \draw (d2.90) -- (c5.270);
+  \draw (d3.90) -- (c2.270);
+  \draw (d3.90) -- (c4.270);
+  \draw (d3.90) -- (c6.270);
+  \draw (d4.90) -- (c3.270);
+  \draw (d4.90) -- (c5.270);
+  \draw (d4.90) -- (c6.270);
+
+  \draw (c1.90) -- (b1.270);
+  \draw (c1.90) -- (b2.270);
+  \draw (c2.90) -- (b1.270);
+  \draw (c2.90) -- (b3.270);
+  \draw (c3.90) -- (b2.270);
+  \draw (c3.90) -- (b3.270);
+  \draw (c4.90) -- (b1.270);
+  \draw (c4.90) -- (b4.270);
+  \draw (c5.90) -- (b2.270);
+  \draw (c5.90) -- (b4.270);
+  \draw (c6.90) -- (b3.270);
+  \draw (c6.90) -- (b4.270);
+
+  \draw (b1) -- (a);
+  \draw (b2) -- (a);
+  \draw (b3) -- (a);
+  \draw (b4) -- (a);
+    %%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
+  \end{tikzpicture}
+  \caption{The lattice of words for the tree $[_S [_S [_N 1 ]_N + [_N 2 ]_N ]_S + [_N 3 ]_N ]_S$ (in $L(G_t)$). The boxed words are shared with Figure~\ref{fig:lattice2}.}
+  \label{fig:lattice1}
+\end{figure*}
+
+\begin{figure*}[t]
+
+\begin{tikzpicture}[every node/.style={rectangle,draw}]
+  \node (a) [draw=white] at (0,5) {$(1) + ((2) + (3))$};
+
+  \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
+  \node (b2) [draw=white] at (-1.2,4.2) {$(1) + ((2) + 3)$};
+  \node (b3) [draw=white] at ( 1.2,4.2) {$(1) + (2 + (3))$};
+  \node (b4) [draw=white] at ( 3.6,4.2) {$1 + ((2) + (3))$};
+
+  \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
+  \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
+  \node (c3) [draw=white] at (-1.1,3) {$(1) + (2 + 3)$};
+  \node (c4) at (1.1,3) {$1 + (2) + (3)$};
+  \node (c5) [draw=white] at (3.3,3) {$1 + ((2) + 3)$};
+  \node (c6) [draw=white] at (5.5,3) {$1 + (2 + (3))$};
+
+  \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
+  \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
+  \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
+  \node (d4) [draw=white] at ( 3.6,1.8) {$1 + (2 + 3)$};
+
+  \node (e) at (0, 1) {$1 + 2 + 3$};
+
+  \draw (e) -- (d1);
+  \draw (e) -- (d2);
+  \draw (e) -- (d3);
+  \draw (e) -- (d4);
+
+  \draw (d1.90) -- (c1.270);
+  \draw (d1.90) -- (c2.270);
+  \draw (d1.90) -- (c3.270);
+  \draw (d2.90) -- (c1.270);
+  \draw (d2.90) -- (c4.270);
+  \draw (d2.90) -- (c5.270);
+  \draw (d3.90) -- (c2.270);
+  \draw (d3.90) -- (c4.270);
+  \draw (d3.90) -- (c6.270);
+  \draw (d4.90) -- (c3.270);
+  \draw (d4.90) -- (c5.270);
+  \draw (d4.90) -- (c6.270);
+
+  \draw (c1.90) -- (b1.270);
+  \draw (c1.90) -- (b2.270);
+  \draw (c2.90) -- (b1.270);
+  \draw (c2.90) -- (b3.270);
+  \draw (c3.90) -- (b2.270);
+  \draw (c3.90) -- (b3.270);
+  \draw (c4.90) -- (b1.270);
+  \draw (c4.90) -- (b4.270);
+  \draw (c5.90) -- (b2.270);
+  \draw (c5.90) -- (b4.270);
+  \draw (c6.90) -- (b3.270);
+  \draw (c6.90) -- (b4.270);
+
+  \draw (b1) -- (a);
+  \draw (b2) -- (a);
+  \draw (b3) -- (a);
+  \draw (b4) -- (a);
+%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
+\end{tikzpicture}
+\caption{The lattice of words for the tree $[_S [_N 1 ]_N + [_S [_N 2 ]_N + [_N 3 ]_N ]_S ]_S$ (in $L(G_t)$). The boxed words are shared with Figure~\ref{fig:lattice1}.}
+\label{fig:lattice2}
+\end{figure*}
+
+\clearpage
+
+% }
 

### 2019-04-09 paper/paper.tex
Progress + switch to new popl template

@@
 %% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[sigplan]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
+%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
 %% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
 %% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
+%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
 %% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[sigplan]{acmart}\settopmatter{}
-
-%% Conference information
+%\documentclass[acmsmall]{acmart}\settopmatter{}
+
+%% Journal information
 %% Supplied to authors by publisher for camera-ready submission;
 %% use defaults for review submission.
-\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
+\acmJournal{PACMPL}
+\acmVolume{1}
+\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
+\acmArticle{1}
 \acmYear{2018}
-\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
+\acmMonth{1}
 \acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
@@
 %% Citation style
-%\citestyle{acmauthoryear}  %% For author/year citations
+\citestyle{acmauthoryear}  %% For author/year citations
 %\citestyle{acmnumeric}     %% For numeric citations
@@
 \newcommand{\NT}{\mathbb{N}} % Set of nonterminals
-\newcommand{\T}{\mathbb{T}} % Set of terminals
-\newcommand{\I}{\mathbb{I}} % Set of identifiers
+\newcommand{\T}{\Sigma} % Set of terminals
+\newcommand{\Labels}{\mathbb{L}} % Set of labels
 \newcommand{\yield}{\mathit{yield}} % yield of a parse tree
@@
 
+% Submission may be up to 25 pages, excluding references
+
 %% Title information
@@
 %% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
+%% Each author should be introduced by \author, followe by
 %% \authornote (optional), \orcid (optional), \affiliation, and
@@
 
-% TODO: ref for undecidable ambiguity checking
 To prevent the possibility of an end-user encountering such an error we must ensure that the grammar cannot give rise to an unresolvable ambiguity.
-It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable.
+It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable \cite{cantorAmbiguityProblemBackus1962}.
 Unresolvable ambiguity, however, turns out to be decidable\footnote{With some caveats, I'll talk more about this during the meeting.}.
@@
 \begin{itemize}
-  \item Building on \cite{palmkvistCreatingDomainSpecificLanguages2019}, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
+  \item Building on the work by \citet{palmkvistCreatingDomainSpecificLanguages2019}, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
   \item A formalization of the unresolvable ambiguity property for context-free EBNF grammars.
@@
 \end{quote}
+% TODO: note that the syntax tree definition is a little bit non-standard
 
@@
 
-% TODO: the last line feels a bit odd, notationally
-\begin{figure}
-  \begin{tabular}{@{}ll@{}}
-      Terminals & $t \in \T$ \\
-      Non-terminals & $N \in \NT$ \\
-      Identifiers & $i \in \I$ \\
-      Regular expressions & $r ::= t \mid N \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
-      Productions & $N -> i : r$ \\
-  \end{tabular}
-  \caption{Context-free EBNF grammars}
-  \label{fig:grammar-definition}
-\end{figure}
-
-A context-free EBNF grammar is a tuple $(S, P, \NT, \T, \I)$.
-$S \in \NT$ is the starting symbol, $P$ is a set of productions, as given in Figure~\ref{fig:grammar-definition}, $\NT$, $\T$, and $\I$ are sets of non-terminals, terminals, and identifiers, respectively.
-Additionally, we require that $\NT \cap \T = \emptyset$.
-Most (though not all) grammars we will consider will have unique identifiers, i.e., there are no two distinct productions in $P$ with the same identifier.
-As an example, Figure~\ref{fig:example-grammar:ambig-grammar} contains a small grammar.
-The first production listed has the identifier ''Sum'' and consists of an expression, then a literal ''+'', then another expression.
-
-The (word) language of a given non-terminal $N$ in a grammar $G = (S, P, \NT, \T, \I)$ is given by $L_G(N)$:
-
-$$
-\begin{array}{r@{\;=\;}l}
-  L_G(t) & \{t\} \\
-  L_G(N) & \bigcup \{L_G(r) \mid (N -> r) \in P\} \\
-  L_G(r_1 \cdot r_2) & \{w_1 \cdot w_2 \mid w_1 \in L_G(r_1), w_2 \in L_g(r_2)\} \\
-  L_G(r_1 + r_2) & L_G(r_1) \cup L_G(r_2) \\
-  L_G(\epsilon) & \{\epsilon\} \\
-  L_G(r^{*}) & \{\epsilon\} \cup L_G(r) \cup L_G(r \cdot r) \cup \ldots \\
-\end{array}
-$$
-
-\noindent The word language of a grammar $G$, written $L(G)$, is thus $L_G(S)$. We will omit the subscript whenever the intended grammar is clear from context.
-
-Note that the right hand side of a production is a regular expression, which could potentially be ambiguous in and of itself.
-For this work we assume the rhs regular expressions to either be unambigous, or that any remaining ambiguity is unimportant.
-This can be achieved in a number of ways, for example using the ambiguity checking of Brabrand and Thomsen \cite{brabrandTypedUnambiguousPattern2010}.
-
-As such, for a given grammar $G = (S, P, \NT, \T, \I)$ we construct a linear representation of its parse trees as another grammar $T_G = (S, P', \NT, \T', \I)$ where:
-
-$$
-\begin{array}{r@{\;=\;}l}
-  P' & \{i : N -> [_i r ]_i \mid (i : N -> r) \in P \} \\
-  \T' & \T \sqcup \bigcup_{(i : N -> r) \in P} \{[_i, ]_i \} \\
-\end{array}
-$$
-
-\noindent where $\sqcup$ denotes disjoint union.
-Intuitively, we surround the right hand side of each production with a unique pair of brackets, signifying the production that was used for the parse.
-Again, we will omit the subscript whenever the intended grammar is clear from context.
-
-The $\yield$ of a parse tree is the word it parsed, i.e., $\yield : L(T) -> L(G)$. Intuitively, $\yield$ removes the brackets introduced when constructing $T$.
-
-A given word $w \in L(G)$ is ambiguous iff:
-
-$$\exists t_1, t_2 \in L(T).\ \yield(t_1) = w \land \yield(t_2) = w \land t_1 \neq t_2$$
-
-\noindent A grammar is ambiguous iff it contains at least one ambiguous word.
+A context-free grammar $G$ is a tuple $(\NT, \T, P, S)$ where:
+
+\begin{itemize}
+\item $\NT$ is a set of non-terminals.
+\item $\T$, disjoint from $\NT$, is a set of terminals.
+\item $P$, a subset of $\NT \times (\NT \cup \T)^{*}$, is a set of productions.
+\item $S \in \NT$ is the starting non-terminal.
+\end{itemize}
+
+\noindent A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
+The set of words recognized by $G$ is written $L(G)$, and is said to be the word language of $G$.
+
+\subsection{Syntax Trees}
+
+A syntax tree takes the terminals of a word and arranges them into a tree, the shape of which is determined by how (i.e., using which productions) the word was parsed / recognized.
+For this paper we will use a linear encoding of syntax trees, letting us use context-free grammars to express the set of syntax trees for some given word language.
+For example, the parse trees of a context-free grammar $G_w = (\NT, \T, P, S)$ are described by the context-free grammar $G_t = (\NT, \T \cup \T_{[]}, P', S)$ where:
+
+\begin{itemize}
+\item $T_{[]} = \bigcup_{p \in P} \{ [_p, ]_p \}$, i.e., a unique pair of brackets per production.
+\item $P' = \{ N -> [_p \alpha ]_p \mid p \in P \land p = N -> \alpha \}$, i.e., each production is wrapped with a unique pair of brackets.
+\end{itemize}
+
+\noindent Such a grammar will (slightly erroneously) be referred to as a tree grammar, and the recognized language as a tree language.
+We will also use syntax trees that are not directly derived from the parse trees of a context-free grammar.
+The precise construction of these will be described at the point of their introduction, but the basic process is the same; surround each production with a unique pair of brackets.
+
+% TODO: reference a book on tree automata and languages
+As such we can define a function $\yield$ (as seen, e.g., in ) that takes a tree and flattens it, producing the word it was constructed from.
+Intuitively, this function merely removes the added brackets.
+
+\subsection{Ambiguity}
+
+The standard definition of ambiguity, using the above formulation of syntax trees, is as follows:
+
+\begin{definition}
+A word $w \in L(G_w)$ is ambiguous relative its parse tree $G_t$ if:
+  \[ \exists t_1, t_2 \in L(G_t).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2)\]
+\end{definition}
+
+\noindent For this paper, we widen the definition slightly to the following:
+
+\begin{definition}
+  A word $w \in L(G_w)$ is ambiguous relative a tree grammar $G_t$ and a function $\parse : L(G_w) -> 2^{L(G_t)}$ if:
+  \[ |\parse(w)| > 1 \]
+\end{definition}
+
+\noindent where $2^S$ is the powerset of $S$ and $|S|$ is the cardinality of $S$. The standard ambiguity definition is regained by choosing
+
+\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(G_t) \}\ \]
 
@@
 
-A nondeterministic finite automaton (NFA) is a tuple $(Q, \Sigma, \Delta, q_0, F)$:
+A nondeterministic finite automaton (NFA) is a tuple $(Q, \Sigma, \delta, q_0, F)$:
 
@@
 \item $r_0 = q_0$.
-\item $\forall_{i = 0}^{n-1} r_{i+1} \in \delta(r_i, a_i)$.
+\item $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$.
 \item $r_n \in F$.
@@
 \noindent We say that the automaton accepts the word $a_0a_1\ldots a_n$ iff there is such a succesful run.
-A particular state $q \in Q$ is \emph{reachable} if there is a (not necessarily successful) run starting in $q_0$ and ending in $q$.
-$q$ is \emph{coreachable} if there is a run starting in $q$ and ending in some $f \in F$.
-
-A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$.
-
-A pushdown automaton extends a finite automaton with a stack, and lets each transition push or pop symbols from it. Formally, a pushdown automaton is a tuple $(Q, \Sigma, \Gamma, \delta, q_0, F)$:
+
+A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to.
+
+A pushdown automaton extends a finite automaton with a stack, and lets each transition push or pop symbols from it.
+Formally, a nondeterministic pushdown automaton is a tuple $(Q, \Sigma, \Gamma, \delta, q_0, F)$:
 
@@
 \item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet.
-\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma*}$.
+\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma^{*}}$.
 \item An initial state $q_0 \in Q$.
@@
 
-\noindent A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma*$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma*$.
-Reachable and coreachable are now defined on configurations, rather than states.
+\noindent A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma*$.
 
@@
 This product automaton has the same input alphabet partitions, each state is a pair of states (one from each automaton), and each stack symbol is a pair of stack symbols (one from each automaton).
-\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified such that all reachable configurations are also coreachable.
+\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run, none are redundant.
 \end{itemize}
@@
 This distinction is useful to make, because it allows us to decouple \emph{what} we want to be expressible and \emph{how} it is to be expressed.
+
+\begin{figure}
+  \begin{tabular}{@{}ll@{}}
+      Terminals & $t \in \T$ \\
+      Non-terminals & $N \in \NT$ \\
+      Labels & $l \in \Labels$ \\
+      Marks & $m \subseteq \Labels$ \\
+      Regular expressions & $r ::= t \mid N_m \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
+      Labelled productions & $N -> l : r$ \\
+  \end{tabular}
+  \caption{The abstract syntax of a language definition.}
+  \label{fig:input-language-definition}
+\end{figure}
+
+\begin{figure}
+  \begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+    $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
+    $E$ & $a$ & $E$ \verb|'+'| $E$ \\
+    $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
+    $E$ & $n$ & $N$ \\
+  \end{tabular}
+  \caption{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity.
+Assumes that $N$ matches a numeric terminal.}
+  \label{fig:running-example-definition}
+\end{figure}
+
+With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}.
+Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label.
+A \emph{mark} is used for disambiguation, and forbids certain productions from taking the place of the marked non-terminal.
+To lessen clutter, we will write $E_\emptyset$ as $E$.
+As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example.
+In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a child of a multiplication, enforcing conventional precedence.
+
+From $D$ we then generate four grammars: $G_w$, $G_t$, $G'_w$, and $G'_t$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:
+
+\begin{itemize}
+\item $G_w$ represents a word language describing all semantically distinct programs.
+\item $G_t$ represents a tree language describing the syntax trees obtained by parsing a word in $L(G_w)$.
+Note that the added brackets are not present in the productions of $E_{l1}$ and $E_{l2}$, since those are all part of the same labelled production in $D$ in Figure~\ref{fig:running-example-definition}.
+\item $G'_w$ is essentially a modified version of $G_w$, e.g., adding parentheses and other forms of disambiguation.
+\item $G'_t$ represents a tree language describing the syntax trees obtained by parsing a word in $L(G'_w)$.
+\end{itemize}
+
+\begin{figure}
+  \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      \toprule
+      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
+      $E$ & $E$ \verb|'+'| $E$ \\
+      $E$ & $E$ \verb|'*'| $E$ \\
+      $E$ & $N$ \\
+      \midrule
+      $E_{l1}$ & $\epsilon$ \\
+      $E_{l1}$ & $E$ $E_{l2}$ \\
+      \midrule
+      $E_{l2}$ & $\epsilon$ \\
+      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G_w$, the generated abstract syntax.}
+    \label{fig:running-example-generated:w}
+  \end{subfigure}%
+%
+  \begin{subfigure}[t]{.54\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
+      \toprule
+      $E$ & $[_{l}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l}$ \\
+      $E$ & $[_{a}$ & $E$ \verb|'+'| $E$ & $]_{a}$ \\
+      $E$ & $[_{m}$ & $E$ \verb|'*'| $E$ & $]_{m}$ \\
+      $E$ & $[_{n}$ & $N$ & $]_{n}$ \\
+      \midrule
+      $E_{l1}$ & & $\epsilon$ & \\
+      $E_{l1}$ & & $E$ $E_{l2}$ & \\
+      \midrule
+      $E_{l2}$ & & $\epsilon$ & \\
+      $E_{l2}$ & & \verb|';'| $E$ $E_{l2}$ & \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G_t$, a language describing a linear encoding of the parse trees of $G_w$.}
+    \label{fig:running-example-generated:t}
+  \end{subfigure}
+
+  \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      \toprule
+      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
+      $E$ & $E$ \verb|'+'| $E$ \\
+      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
+      $E$ & $N$ \\
+      $E$ & \verb|'('| $E$ \verb|')'| \\
+      \midrule
+      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
+      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
+      $E_{\{a\}}$ & $N$ \\
+      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
+      \midrule
+      $E_{l1}$ & $\epsilon$ \\
+      $E_{l1}$ & $E$ $E_{l2}$ \\
+      \midrule
+      $E_{l2}$ & $\epsilon$ \\
+      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G'_w$, the generated concrete syntax.}
+    \label{fig:running-example-generated:w-prime}
+  \end{subfigure}%
+%
+  \begin{subfigure}[t]{.54\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
+      \toprule
+      $E$ & $[_{l, \emptyset}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l, \emptyset}$ \\
+      $E$ & $[_{a, \emptyset}$ & $E$ \verb|'+'| $E$ & $]_{a, \emptyset}$ \\
+      $E$ & $[_{m, \emptyset}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ & $]_{m, \emptyset}$ \\
+      $E$ & $[_{n, \emptyset}$ & $N$ & $]_{n, \emptyset}$ \\
+      $E$ & $[_{g, \emptyset}$ & \verb|'('| $E$ \verb|')'| & $]_{g, \emptyset}$ \\
+      \midrule
+      $E_{\{a\}}$ & $[_{l, \{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l, \{a\}}$ \\
+      $E_{\{a\}}$ & $[_{m, \{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ & $]_{m, \{a\}}$ \\
+      $E_{\{a\}}$ & $[_{n, \{a\}}$ & $N$ & $]_{n, \{a\}}$ \\
+      $E_{\{a\}}$ & $[_{g, \{a\}}$ & \verb|'('| $E$ \verb|')'| & $]_{g, \{a\}}$ \\
+      \midrule
+      $E_{l1}$ & & $\epsilon$ & \\
+      $E_{l1}$ & & $E$ $E_{l2}$ & \\
+      \midrule
+      $E_{l2}$ & & $\epsilon$ & \\
+      $E_{l2}$ & & \verb|';'| $E$ $E_{l2}$ & \\
+      \bottomrule
+    \end{tabular}
+    \caption{$G'_t$, a language describing a linear encoding of the parse trees of $G'_w$.}
+    \label{fig:running-example-generated:t-prime}
+  \end{subfigure}%
+  \caption{The generated grammars.}
+  \label{fig:running-example-generated}
+\end{figure}
+
+\noindent Examples of words in each of these four languages can be seen in Figure~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
+
+{
+\newcommand{\terminal}[1]{\ \underline{#1}\ }
+
+\begin{figure}
+  \begin{subfigure}[b]{.40\linewidth}
+    \[1 + 2 * 3\]
+    \caption{Example word in $L(G_w)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.55\linewidth}
+    \begin{center}
+    \Tree [.{$m$}
+        [.{$a$}
+          [.{$n$} 1 ]
+          +
+          [.{$n$} 2 ] ]
+        *
+        [.{$n$} 3 ] ]
+    \end{center}
+    \[[_m [_a [_n \terminal{1} ]_n \terminal{+} [_n \terminal{2} ]_n ]_a \terminal{*} [_n \terminal{3} ]_n ]_m\]
+    \caption{Example word in $L(G_T)$ and visualized tree.}
+  \end{subfigure}
+
+  \begin{subfigure}[b]{.40\linewidth}
+    \[(1 + 2) * 3\]
+    \caption{Example word in $L(G'_w)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.55\linewidth}
+    \begin{center}
+    \Tree [.{$m, \emptyset$}
+        [.{$g, \{a\}$}
+          (
+          [.{$a, \emptyset$}
+            [.{$n, \emptyset$} 1 ]
+            +
+            [.{$n, \emptyset$} 2 ] ]
+          ) ]
+        *
+        [.{$n, \{a\}$} 3 ] ]
+    \end{center}
+    \[[_{m} [_{g, \{a\}} \terminal{(} [_{a} [_{n} \terminal{1} ]_{n} \terminal{+} [_{n} \terminal{2} ]_{n} ]_{a} \terminal{)} ]_{g, \{a\}} \terminal{*} [_{n, \{a\}} \terminal{3} ]_{n, \{a\}} ]_{m}\]
+    \caption{Example word in $L(G'_T)$ and visualized tree.}
+  \end{subfigure}
+  \caption{Example with words from each grammar that correspond to each other.
+We abbreviate $[_{n, \emptyset}$ as $[_n$.
+The non-bracket terminals in the tree languages appear underlined for increased clarity.}
+  \label{fig:example-words-in-square}
+\end{figure}
+}
+
+At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses.
+For example, $G_w$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
+However, starting with that language definition would allow the expression ''$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ''$;2$'' is not a valid expression.
+
 As an example, in Figure~\ref{fig:example-grammar}, \ref{fig:example-grammar:ambig-grammar} is the \emph{what} and \ref{fig:example-grammar:unambig-grammar} is the \emph{how}.
@@
 \begin{figure*}[ht]
-  \begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-    \synt{Expr} & Sum & \synt{Expr} \verb|'+'| \synt{Expr} \\
-    \synt{Expr} & Product & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ \\
-    \synt{Expr} & Number & \synt{Number} \\
-  \end{tabular}
-  \caption{The input language definition used as a running example, a simple expression language with addition and multiplication, with precedence defined, but not associativity.
-Assumes that \synt{Number} matches a numeric terminal.}
-\end{figure*}
-
-\begin{figure*}[ht]
   \begin{tikzcd}
@@
   \caption{The generated grammars and their relation to each other.}
-\end{figure*}
-
-\begin{figure*}[ht]
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
-      \toprule
-      \synt{Expr} & \synt{Expr} \verb|'+'| \synt{Expr} \\
-      \synt{Expr} & \synt{Expr} \verb|'*'| \synt{Expr} \\
-      \synt{Expr} & \synt{Number} \\
-      \bottomrule
-    \end{tabular}
-    \caption{$G_w$, the generated abstract syntax.}
-  \end{subfigure}%
-%
-  \begin{subfigure}[t]{.54\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
-      \toprule
-      \synt{Expr} & $[_{Sum}$ & \synt{Expr} \verb|'+'| \synt{Expr} & $]_{Sum}$ \\
-      \synt{Expr} & $[_{Product}$ & \synt{Expr} \verb|'*'| \synt{Expr} & $]_{Product}$ \\
-      \synt{Expr} & $[_{Number}$ & \synt{Number} & $]_{Number}$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$G_t$, a language describing a linear encoding of the parse trees of $G_w$.}
-  \end{subfigure}
-
-  \begin{subfigure}{.45\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
-      \toprule
-      \synt{Expr} & \synt{Expr} \verb|'+'| \synt{Expr} \\
-      \synt{Expr} & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ \\
-      \synt{Expr} & \synt{Number} \\
-      \synt{Expr} & \verb|'('| \synt{Expr} \verb|')'| \\
-      \midrule
-      \synt{Expr}$_{Sum}$ & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ \\
-      \synt{Expr}$_{Sum}$ & \synt{Number} \\
-      \synt{Expr}$_{Sum}$ & \verb|'('| \synt{Expr} \verb|')'| \\
-      \bottomrule
-    \end{tabular}
-    \caption{$G'_w$, the generated concrete syntax.}
-  \end{subfigure}%
-%
-  \begin{subfigure}{.54\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
-      \toprule
-      \synt{Expr} & $[_{Sum}$ & \synt{Expr} \verb|'+'| \synt{Expr} & $]_{Sum}$ \\
-      \synt{Expr} & $[_{Product}$ & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ & $]_{Product}$ \\
-      \synt{Expr} & $[_{Number}$ & \synt{Number} & $]_{Number}$ \\
-      \synt{Expr} & $[_{Grouping}$ & \verb|'('| \synt{Expr} \verb|')'| & $]_{Grouping}$ \\
-      \midrule
-      \synt{Expr}$_{Sum}$ & $[_{Sum2}$ & \synt{Expr}$_{Sum}$ \verb|'*'| \synt{Expr}$_{Sum}$ & $]_{Sum2}$ \\
-      \synt{Expr}$_{Sum}$ & $[_{Number2}$ & \synt{Number} & $]_{Number2}$ \\
-      \synt{Expr}$_{Sum}$ & $[_{Grouping2}$ & \verb|'('| \synt{Expr} \verb|')'| & $]_{Grouping2}$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$G'_t$, a language describing a linear encoding of the parse trees of $G'_w$.}
-  \end{subfigure}%
-  \caption{The generated grammars.}
-\end{figure*}
-
-\begin{figure*}[ht]
-  \begin{subfigure}[b]{.45\linewidth}
-    \[1 + 2 * 3\]
-    \caption{Example word in $L(G_w)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{center}
-    \Tree [.Product
-        [.Sum
-          [.Number 1 ]
-          '+'
-          [.Number 2 ] ]
-        '*'
-        [.Number 3 ] ]
-    \end{center}
-    \[[_P [_S [_N 1 ]_N + [_N 2 ]_N ]_S * [_N 3 ]_N ]_P\]
-    \caption{Example word in $L(G_T)$ and visualized tree.}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{.45\linewidth}
-    \[(1 + 2) * 3\]
-    \caption{Example word in $L(G'_w)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{center}
-    \Tree [.Product
-        [.Grouping2
-          '('
-          [.Sum
-            [.Number 1 ]
-            '+'
-            [.Number 2 ] ]
-          ')' ]
-        '*'
-        [.Number2 3 ] ]
-    \end{center}
-    \[[_P [_{G2} ( [_S [_N 1 ]_N + [_N 2 ]_N ]_S ) ]_{G2} * [_{N2} 3 ]_{N2} ]_P\]
-    \caption{Example word in $L(G'_T)$ and visualized tree.}
-  \end{subfigure}
-  \caption{Example with words from each grammar that correspond to each other.
-The subscripts are abbreviated as follows: $S = \text{Sum}$ , $P = \text{Product}$ , $N = \text{Number}$ , and $G = \text{Grouping}$ .}
 \end{figure*}

### 2019-04-10 syncon-parser/src/Main.hs
Initial commit for parser implementation

@@ -1,0 +1,7 @@
+module Main where
+
+import Pre
+
+
+main :: IO ()
+main = putStrLn @Text "Hello world"

### 2019-04-10 syncon-parser/src/P1Lexing/Types.hs

@@ -1,0 +1,34 @@
+module P1Lexing.Types where
+
+import Pre
+
+data Range = Nowhere | Range Position Position
+data Position = Position { line :: Int, column :: Int } deriving (Eq, Ord)
+
+instance Semigroup Range where
+  Range s1 e1 <> Range s2 e2 = Range (min s1 s2) (max e1 e2)
+  Nowhere <> r = r
+  r <> Nowhere = r
+instance Monoid Range where
+  mempty = Nowhere
+  mappend = (<>)
+
+data Token
+  = IntTok Range Int
+  | FloatTok Range Double
+  | StringTok Range Text
+  | IdentTok Range Text
+  | OtherTok Range Text
+
+class Ranged a where
+  range :: a -> Range
+
+instance Ranged Range where
+  range = identity
+
+instance Ranged Token where
+  range (IntTok r _) = r
+  range (FloatTok r _) = r
+  range (StringTok r _) = r
+  range (IdentTok r _) = r
+  range (OtherTok r _) = r

### 2019-04-10 syncon-parser/src/P2LanguageDefinition/Types.hs

@@ -1,0 +1,81 @@
+{-# LANGUAGE TemplateHaskell #-}
+
+-- | This module describes a surface level language definition, i.e., what a user encounters,
+--   as opposed to what later algorithms use.
+module P2LanguageDefinition.Types where
+
+import Pre
+
+import Data.Functor.Foldable.TH (makeBaseFunctor)
+
+import P1Lexing.Types (Range, Ranged, range)
+import qualified P1Lexing.Types as P1
+
+-- TODO: make this something more fancy once we start namespacing things
+newtype Name = Name Text
+
+-- |
+-- = Syncons
+
+-- | A single syncon definition. Note that we will have some special handling for
+--   operators (infix, prefix, postfix), where certain special names are used in
+--   the syntax description in proper positions.
+data Syncon = Syncon
+  { s_name :: Name
+  , s_syntaxType :: Name
+  , s_syntaxDescription :: SyntaxDescription
+  , s_range :: Range
+  }
+
+-- |
+-- == Syntax Descriptions
+
+-- | A syntax description present in a syncon (or an operator)
+data SyntaxDescription
+  = SDSeq Range (Seq SyntaxDescription)
+  | SDRep Range Repetition SyntaxDescription
+  | SDNamed Range SDName SyntaxDescription
+  | SDToken P1.Token -- ^ A literal token, i.e., something written as a quoted string in a .syncon file
+
+data Repetition = RepStar | RepQuestion | RepPlus
+
+-- | Names for subsections of syntax descriptions. These include special names
+--   for operator positions
+data SDName
+  = SDBinaryLeft | SDBinaryRight
+  | SDUnary
+  | SDName Text
+
+-- |
+-- = Disambiguation
+
+-- | In the syncon with the first name, forbid the syntax-type appearing in the
+--   syntax description with the second name, from parsing as the syncon with
+--   the third name (without parens surrounding it). This is the most basic
+--   disambiguation tool, which all other disambiguations eventually translate to
+data Forbid = Forbid Range Name SDName Name
+
+-- | Make the (operator) syncons appearing earlier in the list have higher precedence
+--   than the latter (i.e., total precedence among the mentioned syncons).
+--
+--   Syncons with the same precedence that have the same associativity will
+--   associate together, while those of differing associativity will have
+--   undefined relative precedence.
[…]

### 2019-04-10 syncon-parser/src/Pre.hs

@@ -1,0 +1,67 @@
+{-
+Welcome to your custom Prelude
+Export here everything that should always be in your library scope
+For more info on what is exported by Protolude check:
+https://github.com/sdiehl/protolude/blob/master/Symbols.md
+-}
+module Pre
+( module Exports
+, (<>)
+
+, (>>>)
+, (<<<)
+, (&&&)
+, (***)
+
+, coerce
+
+, compErr
+, compFromJust
+
+, Equal
+, equal
+, isEqual
+
+, Compose(..)
+
+, HashMap
+, HashSet
+) where
+
+import Protolude as Exports hiding ((<>), reader, sourceLine, sourceColumn)
+
+import Control.Arrow ((***), (>>>), (<<<), (&&&))
+import Data.Semigroup ((<>))
+import Data.Coerce (coerce)
+
+import Data.HashMap.Lazy (HashMap)
+import Data.HashSet (HashSet)
+
+import Data.Functor.Compose (Compose(..))
+
+compErr :: Text -> Text -> a
+compErr section = panic . mappend ("Compiler error[" <> section <> "]")
+
+compFromJust :: Text -> Text -> Maybe a -> a
+compFromJust _ _ (Just a) = a
+compFromJust section message _ = compErr section message
+
+data Equal a = Equal (Maybe a) | NotEqual deriving (Eq, Show)
+
+equal :: a -> Equal a
+equal = Just >>> Equal
+
+isEqual :: Equal a -> Bool
+isEqual Equal{} = True
+isEqual _ = False
+
+instance Eq a => Semigroup (Equal a) where
+  NotEqual <> _ = NotEqual
+  _ <> NotEqual = NotEqual
+  Equal Nothing <> b = b
+  a <> Equal Nothing = a
+  e@(Equal (Just a)) <> Equal (Just b) = if a == b then e else NotEqual
+
+instance Eq a => Monoid (Equal a) where
+  mempty = Equal Nothing
+  mappend = (<>)

### 2019-05-20 syncon-parser/src/Data/Automaton/NVA.hs
Only look for resolutions that are at most 10 terminals longer than the original (approximately)

@@ -1,4 +1,5 @@
 {-# LANGUAGE RecordWildCards #-}
+{-# LANGUAGE ViewPatterns #-}
 
 module Data.Automaton.NVA where
 
@@ -7,6 +8,8 @@ import qualified Pre
 
 import qualified Data.HashMap.Strict as M
 import qualified Data.HashSet as S
+import Data.Vector (Vector)
+import qualified Data.Vector as Vec
 
 import qualified Data.Automaton.NFA as N
 import Data.Automaton.EpsilonNVA (EpsNVA(EpsNVA), TaggedTerminal(..))
@@ -26,7 +29,8 @@ data NVA s sta i o c = NVA
   , openTransitions :: !(HashMap s (HashMap o (HashSet (sta, s))))
   , closeTransitions :: !(HashMap s (HashMap c (HashSet (sta, s))))
   , final :: !(HashSet s) }
-  deriving (Show)
+  deriving (Show, Eq, Generic)
+instance (Hashable s, Hashable sta, Hashable i, Hashable o, Hashable c) => Hashable (NVA s sta i o c)
 
 -- | Retrieve all the states mentioned in an NVA
 states :: (Eq s, Hashable s) => NVA s sta i o c -> HashSet s
@@ -507,6 +511,105 @@ shortestWord NVA{..} = recur initialConfigs
     isFinal :: forall x. (s, [sta], x) -> Bool
     isFinal (s, [], _) = S.member s final
     isFinal _ = False
+
+shortestUniqueWord :: forall s sta i o c.
+                      ( Eq s, Hashable s
+                      , Eq sta, Hashable sta
+                      , Eq i, Hashable i
+                      , Eq o, Hashable o
+                      , Eq c, Hashable c )
+                   => Int  -- ^ Max word length to consider (inclusive)
+                   -> HashSet (NVA s sta i o c)  -- ^ NVAs to find unique words for
+                   -> HashMap (NVA s sta i o c) [TaggedTerminal i o c]  -- ^ Produced shortest unique words, per NVA. If no unique word was found for an NVA, then there is no entry for that NVA here
+shortestUniqueWord maxLength (toList >>> Vec.fromList -> nvas) =
+  recur maxLength (getUniques initials) initials
+  <&> Pre.reverse
+  where
+    asNVA idx = Vec.unsafeIndex nvas idx
+    initialConfigs NVA{initial} = S.map (, []) initial
+    initials = nvas
+      <&> initialConfigs
+      & M.singleton []
+
[…]

### 2019-05-20 syncon-parser/src/P4Parsing/AmbiguityReporter.hs

@@ -15,14 +15,14 @@ import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
 import qualified Data.Sequence as Seq
 import Data.Sequence (pattern Empty, pattern (:<|), pattern (:|>))
+import Data.Semigroup (Max(..))
 
 import Data.Generics.Uniplate.Data (universe)
 import Data.Functor.Foldable (project)
 
-import Data.Automaton.NVA (fromEpsNVA, shortestWord, trim)
+import Data.Automaton.NVA (fromEpsNVA, NVA, reduce, shortestUniqueWord, shortestWord, renumber)
 import Data.Automaton.EpsilonNVA (EpsNVA(..), untag, fromNFA, renumberStack, TaggedTerminal(..))
 import qualified Data.Automaton.EpsilonNVA as EpsNVA
-import Data.Automaton.DVA (DVA(..), determinize, renumber, difference, asNVA)
 import Data.Automaton.Regex (Regex(..))
 import qualified Data.Automaton.Regex as Regex
 
@@ -84,12 +84,12 @@ localizeAmbiguities forest
 -- |
 -- = Parse-time resolvability
 
-type ResLang = DVA Int Int Token Token Token
+type ResLang = NVA Int Int Token Token Token
 
 resolvability :: DefinitionFile -> Range -> HashSet Node -> Error
-resolvability df r nodes = M.toList unambigLanguages
+resolvability df r nodes = M.toList languages
   <&> (\(node, lang) ->
-         case asNVA lang & trim & shortestWord of
+         case M.lookup lang shortest of
            Nothing -> Left node
            Just w -> untag identity identity identity <$> w
              & fmap textualToken
@@ -137,19 +137,14 @@ resolvability df r nodes = M.toList unambigLanguages
     languages = S.toMap nodes
       & M.mapWithKey (\node _ -> mkLanguage isForbidden (bracketKind df) getSyTy getGrouping info node
                        & fromEpsNVA
-                       & determinize
+                       & reduce
                        & renumber)
 
-    unambigLanguages :: HashMap Node ResLang
-    unambigLanguages = M.mapWithKey mkUnambig languages
-    mkUnambig node lang = S.delete node nodes
-      & toList
-      & fmap (\n -> M.lookup n languages
-               & compFromJust "P4Parsing.AmbiguityReporter.resolvability.mkUnambig" "Missed computing a language for a node")
-      & foldl' langDiff lang
[…]

### 2019-05-21 paper/paper.tex
EOD

@@
 \newcommand{\semantic}{\mathit{semantic}} % remove semantically unimportant productions from a w' \in L(G')
-\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G') to a subset of L(T)
-\newcommand{\words}{\mathit{words}} % go from a t \in L(T) to a subset of L(G')
+\newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
+\newcommand{\words}{\mathit{words}} % go from a t \in L(G_t) to a subset of L(G'_w)
+\newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
+\newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
+\newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
+\newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
+\newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
 \newcommand{\range}[2]{#1\!-\!#2}
@@
 
+\section{Properties We May Want to Prove}
+
+\subsection{Static Stuff}
+
+\begin{lemma}
+There is an injective function ($\alt$) from well-balanced words to tuples of basic words and range bags.
+\end{lemma}
+
+\begin{lemma}
+$\forall t \in L(G_t), w_1, w_2 \in L(G'_w).\ \{w_1, w_2\} \subseteq \words(t) -> \basic(w_1) = \basic(w_2)$
+\label{lemma:tree-single-basic}
+\end{lemma}
+
+\begin{lemma}
+$\forall t \in L(G_t).\ \words(t) \neq \emptyset$
+\label{lemma:words-nonempty}
+\end{lemma}
+
+\begin{lemma}
+$\forall t_1, t_2 \in L(G_t).\ \words(t_1) \cap \words(t_2) \neq \emptyset -> \basic(t_1) = \basic(t_2)$ \\
+(We're using \ref{lemma:tree-single-basic} and \ref{lemma:words-nonempty} to (ab)use ''$\basic$'' on trees as well)
+\end{lemma}
+
+\begin{definition}
+A language definition is in version 1 or 2 if $\T \cap \mathit{par} = \emptyset$.
+\end{definition}
+
+\begin{theorem}
+In version 1 and 2: $\forall w_1, w_2 \in L(G'_w).\ \altset(w_1) = \altset(w_2) -> \parse(w_1) = \parse(w_2)$, where $\altset(w)$ is the same as $\alt(w)$, but with the second component (the rangebag) replaced with a corresponding set.
+\end{theorem}
+
+\begin{lemma}
+In version 1 and 2, given a $t \in L(G_t)$, the words in $\words(t)$ can be partitioned into a lattice.
+Each partition contains the words that are equal by $\altset$, and ordering is by subset on the rangeset.
+Furthermore, this lattice is bounded, and uniquely described by its bottom and top elements; it contains all elements between bottom and top.
+We denote this lattice as $\lattice(t)$.
+\label{lemma:bounded-lattice}
+\end{lemma}
+
+\begin{definition}
+A language definition is in version 1 if it contains no marked non-terminals and $\T \cap \mathit{par} = \emptyset$.
+\end{definition}
+
+\begin{lemma}
+In version 1: the bottom element of the lattice for a tree has a rangeset that is the empty set.
+\end{lemma}
+
+\begin{theorem}
+Given a tree $t \in L(G_t)$ in version 1, $\exists w \in L(G'_w).\ \parse(w) = \{t\} <-> \neg \exists t' \in L(G_t).\ \lattice(t) \subseteq \lattice(t')$, i.e., a tree is resolvable iff there is no other tree that entirely covers its lattice.
+\end{theorem}
+
+\begin{lemma}
+We can construct a VPDA that recognizes a linear encoding of the lattices of trees in $L(G_t)$, such that there is a bijection between successful runs and trees.
+\label{lemma:linear-lattices}
+\end{lemma}
+
+\begin{lemma}
+Using \ref{lemma:linear-lattices}, we can construct a VPDA that reconizes a linear encoding of ''superlattices'' of trees in $L(G_t)$ (i.e., lattices that completely cover the original lattices), where each successful run corresponds to exactly one tree.
+\label{lemma:linear-super-lattices}
+\end{lemma}
+
+\begin{theorem}
+Using \ref{lemma:linear-lattices}, \ref{lemma:linear-super-lattices}, and their product automaton, we can construct a VPDA such that there is a bijection between successful runs and pairs of trees in $L(G_t)$ such that the lattice of one is entirely covered by the other (this holds in both version 1 and 2).
+\end{theorem}
+
+\begin{theorem}
+We can construct a sound, decidable algorithm that takes a language definition in version 1 or 2 and answers ''resolvably ambiguous'', ''unresolvably ambiguous'', or ''unknown''.
+Additionally, in version 1, this algorithm will never answer ''unknown'', i.e., it is then complete.
+\label{lemma:static-algorithm}
+\end{theorem}
+
+\begin{definition}
+Version 2.5 allows a language definition to use parentheses, but restricts it to never produce double parentheses, and changes $G'_w$ to also never recognize double parentheses.
+\end{definition}
+
+\begin{theorem}
+\ref{lemma:bounded-lattice} holds also for version 2.5, thus we can extend \ref{lemma:static-algorithm} to cover version 2.5 as well.
+\end{theorem}
+
+\subsection{Dynamic Stuff}
+
+This section assumes that we at no point produce an infinite amount of parse trees.
+My intuition (corroborated by some quick googling) suggests that this can only happen if there is a productive, accessible nonterminal $N$ such that $N =>^{+} N$.
+This is not checked at the moment.
+
+\begin{lemma}
+$\forall t \in L(G_t).\ \words(t)$ is a visibly pushdown language.
+\end{lemma}
+
+\begin{theorem}
+  Given a set of trees in $L(G_t)$, we can construct a corresponding set of VPDAs that recognize the words that are unique to each tree (i.e., isn't in $\words(t)$ for some other tree $t$ in the set).
+\end{theorem}
+
+\begin{theorem}
+  We can construct an algorithm $\localize(F)$ that takes a finite parse forest $F \subseteq L(G_t)$ and selects a set of subforests\footnote{This is a term that I have introduced, and so it will need to be described.
+Basically a parse forest for a part of a word, where each tree is a subtree of the original parse forest.} such that:
+  \begin{itemize}
+  \item The unselected parts are identical across the different syntax trees, including range (i.e., we report all ambiguities).
+  \item No selected subforest is contained in another (i.e., we only report one ambiguity per range).
+  \item No selected subforest can be replaced by zero or more of its children while still satisfying the points above (i.e., the selected subforests are minimal\footnote{By covered range, not by count.}).
+  \end{itemize}
+\end{theorem}
+
+\begin{theorem}
+  Given an ambiguous word $w \in L(G'_w)$, a localized ambiguity $F' \in \localize(\parse(w))$ covering the subword $w' \in L_{G'_w}(N)$, and a resolution $w'' \in L_{G'_w}(N)$ such that $\parse_N(w'') = \{t\} \subset \parse_N(w')$ for some $t \in L_{G_t}(N)$, the following holds:
+  \begin{itemize}
+  \item $\localize(\parse(w[w'/w''])) = \localize(\parse(w)) \setminus F'$
+  \end{itemize}
+  where $w[w'/w'']$ denotes replacing a subword $w'$ with $w''$.
+\end{theorem}
+
 %% \noindent Generalizing this to a complete grammar, we get the following property: a grammar $G$ is resolvable iff:

### 2019-05-22 syncon-parser/src/Data/Automaton/EpsilonNVA.hs
Fix ambiguity in syncon definition file syntax
fixes #4

@@ -23,6 +23,7 @@ data EpsNVA s sta i o c = EpsNVA
   , openTransitions :: HashMap s (HashMap o (HashSet (sta, s)))
   , closeTransitions :: HashMap s (HashMap c (HashSet (sta, s)))
   , final :: HashSet s }
+  deriving (Show)
 
 data TaggedTerminal i o c
   = Open o
@@ -130,6 +131,24 @@ renumberStack nva = mapSta oldToNew nva
     newSymbols = evalState computeNewSymbols 0
     computeNewSymbols = stackSymbols nva & S.toMap & traverse (\_ -> get <* modify (+1))
 
+renumberStackKeepLeft :: ( Eq s, Hashable s
+                         , Eq stal, Hashable stal
+                         , Eq star, Hashable star )
+                      => EpsNVA s (Either stal star) i o c -> EpsNVA s (Either stal Int) i o c
+renumberStackKeepLeft nva = mapSta (second oldToNew) nva
+  where
+    oldToNew sta = M.lookup sta newSymbols
+      & compFromJust "Data.Automaton.EpsilonNVA.renumberStackKeepLeft" "missing symbol"
+    newSymbols = evalState computeNewSymbols 0
+    computeNewSymbols = stackSymbols nva
+      & toList
+      & mapMaybe asRight
+      & S.fromList
+      & S.toMap
+      & traverse (\_ -> get <* modify (+1))
+    asRight Left{} = Nothing
+    asRight (Right a) = Just a
+
 renumber :: ( Eq s, Hashable s
             , Eq sta, Hashable sta
             , Eq i, Hashable i
@@ -168,6 +187,33 @@ instance
   , Eq o, Hashable o
   , Eq c, Hashable c )
   => Monoid (EpsNVA Int Int i o c) where
+  mempty = EpsNVA
+    { initial = S.singleton 0
+    , innerTransitions = M.empty
+    , openTransitions = M.empty
+    , closeTransitions = M.empty
+    , final = S.singleton 0 }
+  mappend = (<>)
+
+instance
+  ( Eq i, Hashable i
+  , Eq o, Hashable o
+  , Eq c, Hashable c
+  , Eq sta, Hashable sta)
+  => Semigroup (EpsNVA Int (Either sta Int) i o c) where
+  a <> b = concat a b & renumberStates & mapSta (either (shuffle Left) (shuffle Right))
+    & renumberStackKeepLeft
+    where
[…]

### 2019-05-22 syncon-parser/src/Data/Automaton/GraphViz.hs

@@ -54,5 +54,6 @@ debugWriteDotFile path sToStr aToStr fa b = unsafePerformIO $ do
 escape :: Text -> Text
 escape = Text.concatMap f
   where
+    f '\\' = "\\\\"
     f '"' = "\\\""
     f c = Text.singleton c

### 2019-05-22 syncon-parser/src/Data/Automaton/NVA.hs

@@ -512,6 +512,7 @@ shortestWord NVA{..} = recur initialConfigs
     isFinal (s, [], _) = S.member s final
     isFinal _ = False
 
+-- TODO: make a timeouting version of this in IO
 shortestUniqueWord :: forall s sta i o c.
                       ( Eq s, Hashable s
                       , Eq sta, Hashable sta
@@ -521,6 +522,8 @@ shortestUniqueWord :: forall s sta i o c.
                    => Int  -- ^ Max word length to consider (inclusive)
                    -> HashSet (NVA s sta i o c)  -- ^ NVAs to find unique words for
                    -> HashMap (NVA s sta i o c) [TaggedTerminal i o c]  -- ^ Produced shortest unique words, per NVA. If no unique word was found for an NVA, then there is no entry for that NVA here
+shortestUniqueWord maxLength _
+  | maxLength < 0 = compErr "Data.Automaton.NVA.shortestUniqueWord" $ "Got a negative maxLength: " <> show maxLength
 shortestUniqueWord maxLength (toList >>> Vec.fromList -> nvas) =
   recur maxLength (getUniques initials) initials
   <&> Pre.reverse

### 2019-05-22 syncon-parser/src/Main.hs

@@ -156,7 +156,7 @@ testReduce = do
     post = NVA.reduce nva
 
 test :: IO ()
-test = parseToHTMLDebug "examples/ambig.syncon" "examples/ambig.test" "out.html"
+test = parseToHTMLDebug "examples/bootstrap.syncon" "examples/bootstrap.syncon" "out.html"
 
 main :: IO ()
 main = getArgs >>= \case

### 2019-05-22 syncon-parser/src/P2LanguageDefinition/Parser.hs

@@ -248,7 +248,7 @@ syntaxDescription :: Grammar r (Prod r SyntaxDescription)
 syntaxDescription = mdo
   let sdsequence = do
         start <- lit "("
-        descrs <- fmap Seq.fromList $ (:) <$> alted <*> some alted
+        descrs <- fmap Seq.fromList $ (:) <$> repped <*> some repped
         end <- lit ")"
         return $ SDSeq (range start <> range end) descrs
   let sdlit = uncurry SDToken <$> string
@@ -262,7 +262,7 @@ syntaxDescription = mdo
   let rep = ((,) <$> lit "*" <*> pure RepStar)
             <|> ((,) <$> lit "+" <*> pure RepPlus)
             <|> ((,) <$> lit "?" <*> pure RepQuestion)
-  let repped = (mkRep <$> (named <|> atom) <*> rep) <|> named <|> atom
+  repped <- rule $ (mkRep <$> (named <|> atom) <*> rep) <|> named <|> atom
   let alts = mkAlt <$> some repped <*> some (lit "|" *> some repped)
   alted <- rule $ repped <|> alts <?> "syntax description"
   return repped

### 2019-05-22 syncon-parser/src/P4Parsing/AmbiguityReporter.hs

@@ -21,7 +21,7 @@ import Data.Generics.Uniplate.Data (universe)
 import Data.Functor.Foldable (project)
 
 import Data.Automaton.NVA (fromEpsNVA, NVA, reduce, shortestUniqueWord, shortestWord, renumber)
-import Data.Automaton.EpsilonNVA (EpsNVA(..), untag, fromNFA, renumberStack, TaggedTerminal(..))
+import Data.Automaton.EpsilonNVA (EpsNVA(..), untag, fromNFA, TaggedTerminal(..))
 import qualified Data.Automaton.EpsilonNVA as EpsNVA
 import Data.Automaton.Regex (Regex(..))
 import qualified Data.Automaton.Regex as Regex
@@ -49,7 +49,7 @@ instance FormatError Error where
     "Unresolvable ambiguity error with " <> show (length trees + length resolvable) <> " alternatives.\n" <>
     formattedResolvable <> "\n" <>
     "Unresolvable alternatives:\n" <>
-    foldMap ((n_name &&& (project >>> toList)) >>> format) trees
+    fold unresolvables
     where
       format (name, children) = "  " <> coerce name <> formatChildren children <> "\n"
       formatChildren = sortBy (compare `on` range) >>> foldMap formatNode
@@ -58,6 +58,10 @@ instance FormatError Error where
       formattedResolvable
         | null resolvable = ""
         | otherwise = "\nResolvable alternatives:\n" <> foldMap (\t -> "  " <> t <> "\n") resolvable
+      unresolvables = trees
+        <&> (n_name &&& (project >>> toList))
+        <&> format
+        & S.fromList
   formatError (Ambiguity r resolvable) = simpleErrorMessage r $
     "Ambiguity error with " <> show (length resolvable) <> " alternatives:\n\n" <>
     foldMap (\t -> "  " <> t <> "\n") resolvable
@@ -192,7 +196,7 @@ mkLanguage :: (Name -> SDName -> Name -> Bool)  -- ^ True if forbidden
                            , HashMap (SavedPoint ()) Path
                            , HashSet Path )
            -> Node
-           -> EpsNVA Int Int Token Token Token
+           -> EpsNVA Int (Either () Int) Token Token Token
 mkLanguage isForbidden bracketKind getSyTy getGroupings syncons n@Node{n_name} =
   zipWith genSegment (toList points) (tail $ fst <$> toList points)
[…]

### 2019-05-22 syncon-parser/src/FileAnnotation.hs
Include html template in executable

@@ -1,4 +1,4 @@
-module FileAnnotation (annotate, putInTemplate) where
+module FileAnnotation (annotate, putInTextTemplate, putInTemplate) where
 
 import Pre
 
@@ -40,6 +40,9 @@ annotate source annotations = flip evalState initState $ do
 putInTemplate :: FilePath -> Text -> IO Text
 putInTemplate templatePath html =
   readFile templatePath <&> Text.replace "$prettyprint$" html
+
+putInTextTemplate :: Text -> Text -> Text
+putInTextTemplate template html = Text.replace "$prettyprint$" html template
 
 step :: Char -> AnnoM Builder
 step c = do

### 2019-05-22 syncon-parser/src/Main.hs

@@ -1,3 +1,4 @@
+{-# LANGUAGE TemplateHaskell #-}
 {-# OPTIONS_GHC -fno-warn-unused-imports #-}
 
 module Main where
@@ -5,12 +6,13 @@ module Main where
 import Pre
 import Result (Result(..))
 
-import FileAnnotation (annotate, putInTemplate)
+import FileAnnotation (annotate, putInTextTemplate)
 import ErrorMessage (FormatError, formatErrors, formatError)
 
 import Data.Data (Data)
 import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
+import Data.FileEmbed (embedFile)
 
 import Data.Generics.Uniplate.Data (universe, universeBi)
 import Data.Functor.Foldable (project, cata)
@@ -107,8 +109,8 @@ parseToHTMLDebug defFile sourceFile outFile = do
   case Parser.report df setOfNodes of
     Data node -> universe node >>= nodeAnnotation
       & annotate source
-      & putInTemplate "resources/htmlTemplate.html"
-      & (>>= writeFile outFile)
+      & putInTextTemplate (toS $(embedFile "resources/htmlTemplate.html"))
+      & writeFile outFile
       & (>> putStrLn @Text "Done")
     Error errs -> formatError <$> errs
       & formatErrors fileSource

### 2019-05-22 syncon-parser/src/P2LanguageDefinition/BasicChecker.hs
Report undefined syntax types mentioned in a syntax description

@@ -169,6 +169,9 @@ checkSyncon :: HashMap TypeName (Either SyntaxType TokenType) -> (Either Text Ty
 checkSyncon types tokKind Syncon{..} = do
   checkIsSyTy types s_syntaxType
   findDuplicates fst [(t, r) | SDNamed r (SDName t) _ <- universe s_syntaxDescription]
+  forM_ [(r, tyn) | SDSyTy r tyn <- universe s_syntaxDescription] $ \(r, tyn) ->
+    unless (tyn `M.member` types) $
+      Error [Undefined (coerce tyn) r]
   join . (`cataA` s_syntaxDescription) $ \case
     SDSyTyF r tyn | isSyTy tyn  -> pure $ Error [UnnamedSyntaxTypeOccurrence r]
     SDRecF r _ -> pure $ Error [UnnamedSyntaxTypeOccurrence r]

### 2019-05-22 syncon-parser/src/P4Parsing/Parser.hs
Only print tokens as expected, not syntax types or syncons

@@ -148,7 +148,7 @@ generateGrammar DefinitionFile{..}
           syncons'
           & (`M.intersection` (S.toMap $ M.lookupDefault S.empty tyn syTyToSyncon))
           & (`M.difference` S.toMap excludes)
-          & toList & (lookupEmpty tyn parens : ) & asum & (<?> coerce tyn)
+          & toList & (lookupEmpty tyn parens : ) & asum
       M.lookup (TypeName "Top", S.empty) nts'
         & compFromJust "P4Parsing.Parser.generateGrammar" "Top somehow vanished during generation"
         & fmap fst
@@ -177,7 +177,6 @@ generateSyncon :: HashMap (Name, Either Rec SDName) (TypeName, HashSet Name)
 generateSyncon markings isSyTy nts Syncon{s_name = n, s_syntaxDescription, s_syntaxType = (_, syty)} =
   para alg s_syntaxDescription
   & fmap (\(internals, r) -> (Node n (asStruct internals) r, r))
-  & (<?> ("syncon " <> coerce n))
   where
     alg (SDNamedF _ sdname (SDSyTy _ tyn, _))
       | isSyTy tyn = M.lookupDefault (tyn, S.empty) (n, Right sdname) markings
@@ -211,7 +210,7 @@ lit t = (<?> ("literal " <> t)) . satisfy $ \case
 
 -- | Parse a specific kind of token.
 othertok :: TypeName -> Prod r l (Tok l)
-othertok tyn = (<?> ("othertok " <> coerce tyn)) . satisfy $ \case
+othertok tyn = (<?> ("token " <> coerce tyn)) . satisfy $ \case
   OtherTok _ _ tyn' _ | tyn == tyn' -> True
   _ -> False
 

### 2019-05-22 syncon-parser/src/Data/Automaton/NVA.hs
Timeout in dynamic resolvability algorithm

@@ -10,6 +10,8 @@ import qualified Data.HashMap.Strict as M
 import qualified Data.HashSet as S
 import Data.Vector (Vector)
 import qualified Data.Vector as Vec
+import Data.IORef (IORef, newIORef, readIORef, atomicWriteIORef)
+import System.Timeout (timeout)
 
 import qualified Data.Automaton.NFA as N
 import Data.Automaton.EpsilonNVA (EpsNVA(EpsNVA), TaggedTerminal(..))
@@ -519,14 +521,16 @@ shortestUniqueWord :: forall s sta i o c.
                       , Eq i, Hashable i
                       , Eq o, Hashable o
                       , Eq c, Hashable c )
-                   => Int  -- ^ Max word length to consider (inclusive)
+                   => Int  -- ^ Timeout in microseconds. Negative to never timeout
+                   -> Int  -- ^ Max word length to consider (inclusive)
                    -> HashSet (NVA s sta i o c)  -- ^ NVAs to find unique words for
-                   -> HashMap (NVA s sta i o c) [TaggedTerminal i o c]  -- ^ Produced shortest unique words, per NVA. If no unique word was found for an NVA, then there is no entry for that NVA here
-shortestUniqueWord maxLength _
+                   -> IO (HashMap (NVA s sta i o c) [TaggedTerminal i o c])  -- ^ Produced shortest unique words, per NVA. If no unique word was found for an NVA, then there is no entry for that NVA here
+shortestUniqueWord _ maxLength _
   | maxLength < 0 = compErr "Data.Automaton.NVA.shortestUniqueWord" $ "Got a negative maxLength: " <> show maxLength
-shortestUniqueWord maxLength (toList >>> Vec.fromList -> nvas) =
-  recur maxLength (getUniques initials) initials
-  <&> Pre.reverse
+shortestUniqueWord timeoutDuration maxLength (toList >>> Vec.fromList -> nvas) = do
+  res <- newIORef $ getUniques initials
+  _ <- timeout timeoutDuration $ recur maxLength res initials
+  readIORef res <&> (M.toList >>> fmap (asNVA *** Pre.reverse) >>> M.fromList)
   where
     asNVA idx = Vec.unsafeIndex nvas idx
     initialConfigs NVA{initial} = S.map (, []) initial
@@ -535,22 +539,22 @@ shortestUniqueWord maxLength (toList >>> Vec.fromList -> nvas) =
       & M.singleton []
 
     recur :: Int  -- ^ max amount of further steps to take
[…]

### 2019-05-22 syncon-parser/src/P4Parsing/AmbiguityReporter.hs

@@ -16,6 +16,7 @@ import qualified Data.HashMap.Lazy as M
 import qualified Data.Sequence as Seq
 import Data.Sequence (pattern Empty, pattern (:<|), pattern (:|>))
 import Data.Semigroup (Max(..))
+import System.IO.Unsafe (unsafePerformIO)
 
 import Data.Generics.Uniplate.Data (universe)
 import Data.Functor.Foldable (project)
@@ -145,7 +146,7 @@ resolvability df r nodes = M.toList languages
                        & renumber)
 
     shortest :: HashMap ResLang [TaggedTerminal Token Token Token]
-    shortest = shortestUniqueWord (len + 10) nvas
+    shortest = unsafePerformIO $ shortestUniqueWord 1_000_000 (len + 10) nvas
       where
         Max len = foldMap (shortestWord >>> fmap (length >>> Max) >>> fold) nvas
         nvas = foldMap S.singleton languages

### 2019-05-22 syncon-parser/src/Main.hs
Ensure that all syncons are non-nullable

@@ -158,7 +158,7 @@ testReduce = do
     post = NVA.reduce nva
 
 test :: IO ()
-test = parseToHTMLDebug "examples/bootstrap.syncon" "examples/bootstrap.syncon" "out.html"
+test = parseToHTMLDebug "case-studies/ocaml.syncon" "case-studies/fizzbuzz.ml" "out.html"
 
 main :: IO ()
 main = getArgs >>= \case

### 2019-05-22 syncon-parser/src/P2LanguageDefinition/BasicChecker.hs

@@ -17,7 +17,7 @@ import qualified Data.HashSet as S
 
 import ErrorMessage (ErrorMessage(..), FormatError(..), simpleErrorMessage)
 
-import Data.Functor.Foldable (cataA, para)
+import Data.Functor.Foldable (cataA, cata, para)
 import Data.Generics.Uniplate.Data (universe)
 
 import P1Lexing.Types (Range, range, Ranged)
@@ -39,6 +39,7 @@ data Error
   | InconsistentBracketKinds (Either Text TypeName) [(BracketKind, HashSet Range)]
   | UnequalAltDescriptions Range [((Int, Int), HashSet Range)]
   | UnbalancedDescription Range [Range] [Range]  -- ^ unpaired closing bracket locations, unpaired opening bracket locations
+  | NullableSyntaxDescription Range
   deriving (Show, Eq)
 
 instance FormatError Error where
@@ -113,6 +114,8 @@ instance FormatError Error where
       & sortBy (compare `on` fst)
       & ((r, "") :)
     }
+  formatError (NullableSyntaxDescription sd) = simpleErrorMessage (range sd) $
+    "A syncon may not match nothing (it may not be nullable, in case that is more familiar terminology)."
 
 -- | Add the implicitly defined things.
 addImplicits :: [Top] -> [Top]
@@ -189,6 +192,8 @@ checkSyncon types tokKind Syncon{..} = do
         balances -> Error [UnequalAltDescriptions r $ second (fmap snd >>> S.fromList) <$> balances]
     SDRepF _ _ (original, sd) -> checkBalanced original sd
     other -> foldMap snd other
+  when (isNullable s_syntaxDescription) $
+    Error [NullableSyntaxDescription $ range s_syntaxDescription]
   pure ()
   where
     isSyTy tyn = M.lookup tyn types & maybe False isLeft
@@ -203,6 +208,14 @@ checkSyncon types tokKind Syncon{..} = do
       OpenBracket -> BalancedDescription Seq.empty (Seq.singleton r)
       NonBracket -> mempty
       CloseBracket -> BalancedDescription (Seq.singleton r) Seq.empty
+    isNullable = cata $ \case
+      SDSeqF _ sds -> and sds
+      SDAltF _ sds -> or sds
+      SDRepF _ RepStar _ -> True
+      SDRepF _ RepQuestion _ -> True
+      SDRepF _ RepPlus sd -> sd
+      SDNamedF _ _ sd -> sd
[…]

### 2019-05-22 syncon-parser/src/P2LanguageDefinition/Types.hs

@@ -95,14 +95,14 @@ data SyntaxDescription
   | SDSyTy Range TypeName
   | SDRec Range Rec
   | SDToken Range Text -- ^ A literal token, i.e., something written as a quoted string in a .syncon file
-  deriving (Show, Data, Typeable)
+  deriving (Eq, Show, Data, Typeable)
 
 -- | Recursion to the same syntax type, behaves exactly the same as just writing that syntax type,
 -- except in precedence lists, where forbids are generated for recs.
 data Rec = LRec | RRec | Rec deriving (Show, Data, Typeable, Eq, Generic)
 instance Hashable Rec
 
-data Repetition = RepStar | RepQuestion | RepPlus deriving (Show, Data, Typeable)
+data Repetition = RepStar | RepQuestion | RepPlus deriving (Eq, Show, Data, Typeable)
 
 -- |
 -- = Disambiguation

### 2019-05-22 paper/paper.tex
EOD

@@
 \begin{theorem}
-Given a tree $t \in L(G_t)$ in version 1, $\exists w \in L(G'_w).\ \parse(w) = \{t\} <-> \neg \exists t' \in L(G_t).\ \lattice(t) \subseteq \lattice(t')$, i.e., a tree is resolvable iff there is no other tree that entirely covers its lattice.
+Given a tree $t \in L(G_t)$ in version 1, $\exists w \in L(G'_w).\ \parse(w) = \{t\} <-> \neg \exists t' \in L(G_t).\ t \neq t' \land \lattice(t) \subseteq \lattice(t')$, i.e., a tree is resolvable iff there is no other tree that entirely covers its lattice.
 \end{theorem}
@@
 \begin{lemma}
-We can construct a VPDA that recognizes a linear encoding of the lattices of trees in $L(G_t)$, such that there is a bijection between successful runs and trees.
+(Version 1 and 2) We can construct a VPDA that recognizes a linear encoding of the lattices of trees in $L(G_t)$, such that there is a bijection between successful runs and trees.
 \label{lemma:linear-lattices}
@@
 \begin{lemma}
-Using \ref{lemma:linear-lattices}, we can construct a VPDA that reconizes a linear encoding of ''superlattices'' of trees in $L(G_t)$ (i.e., lattices that completely cover the original lattices), where each successful run corresponds to exactly one tree.
+(Version 1 and 2) Using \ref{lemma:linear-lattices}, we can construct a VPDA that reconizes a linear encoding of ''superlattices'' of trees in $L(G_t)$ (i.e., lattices that completely cover the original lattices), where each successful run corresponds to exactly one tree.
 \label{lemma:linear-super-lattices}
@@
 \begin{theorem}
-Using \ref{lemma:linear-lattices}, \ref{lemma:linear-super-lattices}, and their product automaton, we can construct a VPDA such that there is a bijection between successful runs and pairs of trees in $L(G_t)$ such that the lattice of one is entirely covered by the other (this holds in both version 1 and 2).
+(Version 1 and 2) Using \ref{lemma:linear-lattices}, \ref{lemma:linear-super-lattices}, and their product automaton, we can construct a VPDA such that there is a bijection between successful runs and pairs of trees in $L(G_t)$ such that the lattice of one is entirely covered by the other.
 \end{theorem}
@@
 \begin{theorem}
-  Given an ambiguous word $w \in L(G'_w)$, a localized ambiguity $F' \in \localize(\parse(w))$ covering the subword $w' \in L_{G'_w}(N)$, and a resolution $w'' \in L_{G'_w}(N)$ such that $\parse_N(w'') = \{t\} \subset \parse_N(w')$ for some $t \in L_{G_t}(N)$, the following holds:
+  Given an ambiguous word $w \in L(G'_w)$, a localized ambiguity $F' \in \localize(\parse(w))$ covering the subword $w' \in L_{G'_w}(N)$ (i.e., $\parse_N(w') = F'$), and a resolution $w'' \in L_{G'_w}(N)$ such that $\parse_N(w'') = \{t\} \subset \parse_N(w')$ for some $t \in L_{G_t}(N)$, the following holds:
   \begin{itemize}

### 2019-05-23 syncon-parser/src/FileAnnotation.hs
Fixed bug in FileAnnotation, opens and closes at the same position no longer mix improperly

@@ -30,9 +30,10 @@ annotate source annotations = flip evalState initState $ do
   where
     initState = AnnotationState { position = firstPosition, commands }
     commands :: [(Position, Maybe Text)]
-    commands = sortBy (compare `on` fst) $ do
+    commands = sortBy comp $ do
       (Range start end, annotation) <- annotations
       [(start, Just annotation), (end, Nothing)]
+    comp (p1, command1) (p2, command2) = compare p1 p2 <> comparing void command1 command2
 
 -- | Read the provided file, replace the text "$prettyprint$" with the second argument, then return
 -- the result. Presumably, the template should contain whatever styling and other wrapping is required

### 2019-05-23 syncon-parser/src/Data/Automaton/DVA.hs
Add forcing non-operators in precedence lists and non-productive cycle checking

@@ -11,7 +11,7 @@ import qualified Data.HashSet as S
 import Data.Align (align, padZip)
 import Data.These (fromThese)
 
-import Util (iterateInductivelyOptM, repeatUntilStable)
+import Util (iterateInductivelyOptM, repeatUntilStableM)
 
 import Data.Automaton.NVA (NVA(NVA))
 import qualified Data.Automaton.NVA as NVA
@@ -89,7 +89,7 @@ determinize NVA
       { final = S.filter (M.keys >>> S.fromList >>> S.intersection fin >>> S.null >>> not) (states dva) }
 
     allEdges = execState discoverAll mempty
-    discoverAll = discoverOnce >> gets stackSyms & repeatUntilStable
+    discoverAll = discoverOnce >> gets stackSyms & repeatUntilStableM
     discoverOnce = gets dstates >>= \currStates -> iterateInductivelyOptM discoverSuccessors
       (if S.null currStates
       then S.singleton startState

### 2019-05-23 syncon-parser/src/P2LanguageDefinition/BasicChecker.hs

@@ -8,6 +8,7 @@ module P2LanguageDefinition.BasicChecker
 
 import Pre hiding (check)
 import Result (Result(..), errorIfNonEmpty)
+import Util (repeatUntilStableBy)
 
 import qualified Data.Text as Text
 import Data.Bitraversable (bisequence)
@@ -17,7 +18,7 @@ import qualified Data.HashSet as S
 
 import ErrorMessage (ErrorMessage(..), FormatError(..), simpleErrorMessage)
 
-import Data.Functor.Foldable (cataA, cata, para)
+import Data.Functor.Foldable (cataA, cata, para, zygo)
 import Data.Generics.Uniplate.Data (universe)
 
 import P1Lexing.Types (Range, range, Ranged)
@@ -27,7 +28,7 @@ type Res = Result [Error]
 data Error
   = DuplicateDefinition Text [Range]
   | Undefined Text Range
-  | NotAnOperator Name Range
+  | ExpectedOther Name OpOrNot Bool Range  -- ^ name is opOrNot, expected opposite, bool = is in main list
   | InconsistentPrecedence (Name, Name) [(Ordering, HashSet Range)]
   | NonEqSelfPrecedence Name (HashSet Range)
   | WrongSyntaxType Name SDName TypeName Name TypeName Range  -- ^ name.sdname : syntaxtype, but name : typename
@@ -40,6 +41,7 @@ data Error
   | UnequalAltDescriptions Range [((Int, Int), HashSet Range)]
   | UnbalancedDescription Range [Range] [Range]  -- ^ unpaired closing bracket locations, unpaired opening bracket locations
   | NullableSyntaxDescription Range
+  | NonProductiveCycle (Seq (Name, Range, TypeName))
   deriving (Show, Eq)
 
 instance FormatError Error where
@@ -51,8 +53,11 @@ instance FormatError Error where
     }
   formatError (Undefined defname r) = simpleErrorMessage r $
     defname <> " is undefined"
-  formatError (NotAnOperator defname r) = simpleErrorMessage r $
-    coerce defname <> " is not an operator (it was defined with 'syncon', not 'infix', 'prefix', or 'suffix')."
+  formatError (ExpectedOther defname Op _ r) = simpleErrorMessage r $
+    coerce defname <> " is an operator, but the '!' prefix suggests that it is not."
[…]

### 2019-05-23 syncon-parser/src/P2LanguageDefinition/Parser.hs

@@ -75,7 +75,7 @@ synconTokens = Lexer.LanguageTokens
   -- Literal tokens
   [ "token", "=", "syncon", ":", "{", ";", "}", "prefix", "postfix", "infix"
   , "(", ")", "*", "+", "?", ".", "comment", "left", "right", "precedence", "except"
-  , "type", "builtin", "forbid", "|", "rec", "grouping" ]
+  , "type", "builtin", "forbid", "|", "rec", "grouping", "!" ]
   -- Regex tokens
   [ (NameTok, (Nowhere, "[[:lower:]][[:word:]]*"))
   , (TypeNameTok, (Nowhere, "[[:upper:]][[:word:]]*"))
@@ -132,10 +132,12 @@ precedenceDef :: Grammar r (Prod r PrecedenceList)
 precedenceDef = rule . (<?> "precedence disambiguation list") $ do
   start <- lit "precedence" <* lit "{"
   pList <- innerList
+    $ (NonOp,) <$ lit "!"
+    <|> pure (Op,)
   end <- lit "}"
   mExcept <- optional $ do
     lit "except" *> lit "{"
-    eList <- innerList
+    eList <- innerList $ pure identity
     end' <- lit "}"
     pure (end', eList)
   pure $ PrecedenceList
@@ -143,7 +145,7 @@ precedenceDef = rule . (<?> "precedence disambiguation list") $ do
     pList
     (foldMap snd mExcept)
   where
-    innerList = Seq.fromList <$> some (Seq.fromList <$> some (snd <$> name) <* lit ";")
+    innerList f = Seq.fromList <$> some (Seq.fromList <$> some (f <*> (snd <$> name)) <* lit ";")
 
 groupingDef :: Grammar r (Prod r Grouping)
 groupingDef = rule . (<?> "grouping disambiguation") $ do

### 2019-05-23 syncon-parser/src/P2LanguageDefinition/Types.hs

@@ -125,8 +125,10 @@ data Forbid
 --
 --   The last member is an exception list, the operators appearing in the same list
 --   there will *not* have their relative precedence defined by this precedence list.
-data PrecedenceList = PrecedenceList !Range !(Seq (Seq Name)) !(Seq (Seq Name))
+data PrecedenceList = PrecedenceList !Range !(Seq (Seq (OpOrNot, Name))) !(Seq (Seq Name))
   deriving (Show, Data, Typeable)
+data OpOrNot = Op | NonOp deriving (Eq, Show, Data, Typeable, Generic)
+instance Hashable OpOrNot
 
 -- | A (sparse) matrix of precedences. Should only ever contain references to syncons
 -- defined as operators. Internally is a map from (min a b, max a b) names to orderings

### 2019-05-23 syncon-parser/src/Util.hs

@@ -1,6 +1,8 @@
+{-# LANGUAGE ViewPatterns #-}
+
 module Util where
 
-import Pre hiding (all)
+import Pre hiding (all, check)
 
 import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
@@ -34,10 +36,19 @@ iterateInductivelyM f s = recur S.empty s
 iterateInductively :: (Eq s, Hashable s) => (s -> HashSet s) -> HashSet s -> HashSet s
 iterateInductively f s = runIdentity $ iterateInductivelyOptM (f >>> return) s
 
-repeatUntilStable :: (Monad m, Eq a) => m a -> m a
-repeatUntilStable action = action >>= recur
+repeatUntilStableM :: (Monad m, Eq a) => m a -> m a
+repeatUntilStableM action = action >>= recur
   where
     recur prev = action >>= \new -> if prev == new then return new else recur new
+
+repeatUntilStableBy :: Eq b => (a -> b) -> (a -> a) -> a -> a
+repeatUntilStableBy check step a = recur (check a) a
+  where
+    recur prev (step -> a')
+      | prev == next = a'
+      | otherwise = recur next a'
+      where
+        next = check a'
 
 flipMap :: forall a b c. (Eq a, Hashable a, Eq b, Hashable b, Eq c, Hashable c)
         => HashMap a (HashMap b c)

### 2019-05-24 syncon-parser/src/ErrorMessage.hs
In progress

@@ -13,6 +13,7 @@ import Text.Printf (printf)
 import qualified Data.Sequence as Seq
 import Data.Text (split, splitAt)
 import qualified Data.Text as Text
+import qualified Data.HashMap.Lazy as M
 
 import System.Console.Pretty (style, Style(Bold), color, Color(Red))
 import Text.Show.Pretty (ppShow)
@@ -39,15 +40,18 @@ simpleErrorMessage r t = ErrorMessage { e_message = t, e_range = r, e_ranges = [
 -- TODO: omit parts of the code if it's a large number of lines
 -- | Produce (colored) output suitable for printing the errors. Will extract source code and highlight the
 -- relevant section.
-formatErrors :: Foldable f => Text -> f ErrorMessage -> Text
-formatErrors source messages = sortedMessages <&> formatSingle & Text.intercalate "\n--------------------\n\n"
+formatErrors :: Foldable f => HashMap Text Text -> f ErrorMessage -> Text
+formatErrors sources messages = sortedMessages <&> formatSingle & Text.intercalate "\n--------------------\n\n"
   where
     sortedMessages :: [ErrorMessage]
     sortedMessages = toList messages & sortBy (compare `on` range)
     firstLine :: Int
     firstLine = line firstPosition
-    lines :: Seq (Int, Text)
-    lines = [firstLine..] `zip` split (== '\n') source & Seq.fromList
+    lines :: HashMap Text (Seq (Int, Text))
+    lines = sources <&> \source -> [firstLine..] `zip` split (== '\n') source & Seq.fromList
+    getLines :: Text -> Seq (Int, Text)
+    getLines path = M.lookup path lines
+      & errJust "missing file"
 
     -- Format a single ErrorMessage
     formatSingle :: ErrorMessage -> Text
@@ -61,12 +65,12 @@ formatErrors source messages = sortedMessages <&> formatSingle & Text.intercalat
 
     extractRange :: Range -> Text
     extractRange Nowhere = ""
-    extractRange pos@(Range (Position l1 _) (Position l2 _))
-      | l1 == l2 = Seq.lookup (l1-firstLine) lines
+    extractRange pos@(Range f (Position l1 _) (Position l2 _))
+      | l1 == l2 = Seq.lookup (l1-firstLine) (getLines f)
         & errJust "line out of bounds"
         & formatLine pos
[…]

### 2019-05-24 syncon-parser/src/FileAnnotation.hs

@@ -1,4 +1,4 @@
-module FileAnnotation (annotate, putInTextTemplate, putInTemplate) where
+module FileAnnotation (annotate, annotateOne, putInTextTemplate, putInTemplate) where
 
 import Pre
 
@@ -6,6 +6,7 @@ import Data.List (span)
 import Data.Text.Lazy.Builder (Builder)
 import qualified Data.Text.Lazy.Builder as Builder
 import qualified Data.Text as Text
+import qualified Data.HashMap.Strict as M
 
 import P1Lexing.Types (Range(..), Position(..), firstPosition, stepPosition)
 
@@ -16,13 +17,28 @@ data AnnotationState = AnnotationState
 
 type AnnoM = State AnnotationState
 
+-- | Takes a bunch of source files (map from filepath to contents) and annotations,
+-- and creates one piece of HTML, annotated as with 'annotateOne', but with a <h1>
+-- header for each file, and a little bit of vertical whitespace inbetween.
+annotate :: HashMap Text Text -> [(Range, Text)] -> Text
+annotate sources annotations = M.intersectionWith annotateOne sources splitAnnotations
+  & M.toList
+  & sortBy (comparing fst)
+  & foldMap mkFileSection
+  where
+    splitAnnotations = mapFromFoldable (fst >>> getFilename) annotations
+    getFilename (Range f _ _) = f
+    getFilename Nowhere = ""
+    mkFileSection (path, annotated) = "<h1>" <> (escape path & Builder.toLazyText & toS) <> "</h1>\n" <> annotated <> "\n\n"
+
 -- TODO: assumes all ranges are well-nested, either check for this, or (ideally), implement support for non-nested annotations
--- | Takes some source code and a list of (well-nested) annotations and produces well-formatted HTML.
+-- | Takes some source code and a list of (well-nested) annotations
+-- and produces well-formatted HTML.
 -- Each annotated section is formatted as follows:
 -- <span class="tagged"><div class="tag">annotation-content</div>section-content</span>
 -- All text is properly escaped, and whitespace is preserved.
-annotate :: Text -> [(Range, Text)] -> Text
-annotate source annotations = flip evalState initState $ do
+annotateOne :: Text -> [(Range, Text)] -> Text
+annotateOne source annotations = flip evalState initState $ do
   start <- dumpCurrentCommands
   source' <- Text.unpack source & traverse step & fmap mconcat
   extras <- dumpRemainingCommands
[…]

### 2019-05-24 syncon-parser/src/Main.hs

@@ -12,6 +12,7 @@ import ErrorMessage (FormatError, formatErrors, formatError)
 import Data.Data (Data)
 import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
+import qualified Data.Sequence as Seq
 import Data.FileEmbed (embedFile)
 
 import Data.Generics.Uniplate.Data (universe, universeBi)
@@ -95,20 +96,20 @@ parse4Test = do
 --       -- Parser.Ambiguity r alts -> (r, fmap (project >>> fmap (project >>> void)) $ toList alts)
 --       Parser.Ambiguity r alts -> (r, fmap (project >>> void) $ toList alts)
 
-parseToHTMLDebug :: FilePath -> FilePath -> FilePath -> IO ()
-parseToHTMLDebug defFile sourceFile outFile = do
-  defSource <- readFile defFile
+parseToHTMLDebug :: [FilePath] -> FilePath -> FilePath -> IO ()
+parseToHTMLDebug defFiles sourceFile outFile = do
+  defSource :: HashMap Text Text <- defFiles <&> toS & S.fromList & S.toMap & M.traverseWithKey (\path _ -> readFile $ toS path)
   putStrLn @Text "Parsing definition file"
-  tops <- LD.parseFile defFile >>= dataOrError defSource
+  tops <- M.traverseWithKey (\defFile _ -> LD.parseFile $ toS defFile) defSource
+          >>= (fold >>> dataOrError defSource)
   df <- LD.mkDefinitionFile tops & dataOrError defSource
   parseFile <- Parser.parseSingleLanguage df & dataOrError defSource
-  fileSource <- readFile sourceFile
+  fileSource <- readFile sourceFile <&> M.singleton (toS sourceFile)
   putStrLn @Text "Parsing source file"
   setOfNodes <- parseFile sourceFile >>= dataOrError fileSource
-  source <- readFile sourceFile
   case Parser.report df setOfNodes of
     Data node -> universe node >>= nodeAnnotation
-      & annotate source
+      & annotate fileSource
       & putInTextTemplate (toS $(embedFile "resources/htmlTemplate.html"))
       & writeFile outFile
       & (>> putStrLn @Text "Done")
@@ -125,7 +126,7 @@ nodeAnnotation n = (range n, Parser.n_name n & coerce)
[…]

### 2019-05-24 syncon-parser/src/P1Lexing/Lexer.hs

@@ -35,21 +35,21 @@ data LanguageTokens n = LanguageTokens
   [(Range, Text)] -- ^ Regexes for comments (range for the string, the regex itself)
 
 data Error l n
-  = RegexError l Position Text
+  = RegexError l Text Position Text
   | UnicodeError UnicodeException
   | OverlappingLex l [Token l n]
-  | NoLex Position
+  | NoLex Text Position
   deriving (Show)
 
 instance (Show l, Show n) => FormatError (Error l n) where
-  formatError (RegexError _ pos t) = simpleErrorMessage (Range pos pos) $
+  formatError (RegexError _ f pos t) = simpleErrorMessage (Range f pos pos) $
     "Regex error: " <> t
   formatError (UnicodeError ex) = simpleErrorMessage Nowhere $
     "The file must be UTF8 encoded, got an error: " <> show ex
   formatError (OverlappingLex _ toks) = simpleErrorMessage (foldMap range toks) $
     "Lexing needs to produce a unique longest token, but there were multiple:\n"
     <> foldMap (show >>> (<> "\n")) toks
-  formatError (NoLex pos) = simpleErrorMessage (Range pos (Position (line pos + 4) maxBound)) $
+  formatError (NoLex f pos) = simpleErrorMessage (Range f pos (Position (line pos + 4) maxBound)) $
     "There is no valid token starting here."
 
 data LanguageInternal n = LanguageInternal
@@ -60,6 +60,7 @@ data LanguageInternal n = LanguageInternal
 
 data Lexer l n = Lexer
   { remainingSource :: !ByteString
+  , file :: !Text
   , position :: !Position
   , languages :: !(HashMap l (LanguageInternal n))
   }
@@ -85,8 +86,8 @@ compileRegex t = unsafePerformIO (compile compOpts execOpts (encodeUtf8 ("(*UCP)
 compileRegex' :: l -> Range -> Text -> Result [Error l n] Regex
 compileRegex' l r t = case compileRegex t of
   Left (errColumn, message) -> case r of
-    Range (Position line column) _ -> Error [RegexError l (Position line (column + errColumn)) (toS message)]
-    Nowhere -> Error [RegexError l (Position (-1) errColumn) (toS message)]
+    Range f (Position line column) _ -> Error [RegexError l f (Position line (column + errColumn)) (toS message)]
+    Nowhere -> Error [RegexError l "" (Position (-1) errColumn) (toS message)]
   Right regex -> Data regex
 
[…]

### 2019-05-24 syncon-parser/src/P1Lexing/Types.hs

@@ -4,7 +4,7 @@ import Pre
 
 import Data.Data (Data)
 
-data Range = Nowhere | Range !Position !Position deriving (Show, Eq, Ord, Data, Typeable, Generic)
+data Range = Nowhere | Range !Text !Position !Position deriving (Show, Eq, Ord, Data, Typeable, Generic)
 data Position = Position { line :: !Int, column :: !Int } deriving (Show, Eq, Ord, Data, Typeable, Generic)
 
 firstPosition :: Position
@@ -16,13 +16,15 @@ stepPosition (Position l c) _ = Position l (c+1)
 
 textualRange :: Range -> Text
 textualRange Nowhere = "(nowhere"
-textualRange (Range (Position l1 c1) (Position l2 c2))
-  | l1 == l2 && c1 == c2 = show l1 <> ":" <> show c1
-  | l1 == l2 = show l1 <> ":" <> show c1 <> "-" <> show c2
-  | otherwise = show l1 <> ":" <> show c1 <> "-" <> show l2 <> ":" <> show c2
+textualRange (Range f (Position l1 c1) (Position l2 c2))
+  | l1 == l2 && c1 == c2 = f <> ":" <> show l1 <> ":" <> show c1
+  | l1 == l2 = f <> ":" <> show l1 <> ":" <> show c1 <> "-" <> show c2
+  | otherwise = f <> ":" <> show l1 <> ":" <> show c1 <> "-" <> show l2 <> ":" <> show c2
 
 instance Semigroup Range where
-  Range s1 e1 <> Range s2 e2 = Range (min s1 s2) (max e1 e2)
+  Range f1 s1 e1 <> Range f2 s2 e2
+    | f1 == f2 = Range f1 (min s1 s2) (max e1 e2)
+    | otherwise = compErr "P1Lexing.Types.<>" $ "mappending ranges from different files: " <> f1 <> " and " <> f2
   Nowhere <> r = r
   r <> Nowhere = r
 instance Monoid Range where

### 2019-05-24 syncon-parser/src/ErrorMessage.hs
Parse multiple definition files, treat them as one

@@ -60,8 +60,11 @@ formatErrors sources messages = sortedMessages <&> formatSingle & Text.intercala
     -- Format code extracts
     formatRange :: (Range, Text) -> Text
     formatRange (Nowhere, "") = ""
-    formatRange (r, "") = "\n" <> extractRange r
-    formatRange (r, t) = "\n" <> ensureLn t <> extractRange r
+    formatRange (r, "") = "\n" <> formatPath r <> extractRange r
+    formatRange (r, t) = "\n" <> ensureLn t <> formatPath r <> extractRange r
+
+    formatPath (Range f _ _) = " " <> f <> ":\n"
+    formatPath Nowhere = ""
 
     extractRange :: Range -> Text
     extractRange Nowhere = ""

### 2019-05-24 syncon-parser/src/Main.hs

@@ -99,7 +99,7 @@ parse4Test = do
 parseToHTMLDebug :: [FilePath] -> FilePath -> FilePath -> IO ()
 parseToHTMLDebug defFiles sourceFile outFile = do
   defSource :: HashMap Text Text <- defFiles <&> toS & S.fromList & S.toMap & M.traverseWithKey (\path _ -> readFile $ toS path)
-  putStrLn @Text "Parsing definition file"
+  putStrLn @Text "Parsing definition file(s)"
   tops <- M.traverseWithKey (\defFile _ -> LD.parseFile $ toS defFile) defSource
           >>= (fold >>> dataOrError defSource)
   df <- LD.mkDefinitionFile tops & dataOrError defSource
@@ -159,7 +159,7 @@ testReduce = do
     post = NVA.reduce nva
 
 test :: IO ()
-test = parseToHTMLDebug ["case-studies/ocaml.syncon"] "case-studies/fizzbuzz.ml" "out.html"
+test = parseToHTMLDebug ["examples/split1.syncon", "examples/split2.syncon"] "examples/split.test" "out.html"
 
 getArgsSeq :: IO (Seq [Char])
 getArgsSeq = getArgs <&> Seq.fromList

### 2019-05-24 syncon-parser/src/P1Lexing/Types.hs

@@ -22,9 +22,10 @@ textualRange (Range f (Position l1 c1) (Position l2 c2))
   | otherwise = f <> ":" <> show l1 <> ":" <> show c1 <> "-" <> show l2 <> ":" <> show c2
 
 instance Semigroup Range where
-  Range f1 s1 e1 <> Range f2 s2 e2
+  r1@(Range f1 s1 e1) <> r2@(Range f2 s2 e2)
     | f1 == f2 = Range f1 (min s1 s2) (max e1 e2)
-    | otherwise = compErr "P1Lexing.Types.<>" $ "mappending ranges from different files: " <> f1 <> " and " <> f2
+    | f1 > f2 = r1
+    | otherwise = r2
   Nowhere <> r = r
   r <> Nowhere = r
 instance Monoid Range where

### 2019-05-24 syncon-parser/src/P2LanguageDefinition/BasicChecker.hs

@@ -345,7 +345,9 @@ checkPrecedences names tops = consistentTypes *> matrix
           & fmap (uncurry $ mkEntry EQ)
           & foldl' mergePreMatrices M.empty
         exceptions :: HashMap (Name, Name) ()
-        exceptions = foldMap orderedPairs eList & S.fromList & S.toMap
+        exceptions = foldMap orderedPairs eList <&> uncurry (mkEntry EQ)
+          & foldl' mergePreMatrices M.empty
+          & void
         mkEntry :: Ordering -> Name -> Name -> HashMap (Name, Name) (HashMap Ordering (HashSet Range))
         mkEntry ordering n1@(Name t1) n2@(Name t2)
           | t1 <= t2 = M.singleton (n1, n2) $ M.singleton ordering $ S.singleton r

### 2019-05-24 syncon-parser/src/Main.hs
Minor formatting

@@ -98,7 +98,11 @@ parse4Test = do
 
 parseToHTMLDebug :: [FilePath] -> FilePath -> FilePath -> IO ()
 parseToHTMLDebug defFiles sourceFile outFile = do
-  defSource :: HashMap Text Text <- defFiles <&> toS & S.fromList & S.toMap & M.traverseWithKey (\path _ -> readFile $ toS path)
+  defSource :: HashMap Text Text <- defFiles
+    <&> toS
+    & S.fromList
+    & S.toMap
+    & M.traverseWithKey (\path _ -> readFile $ toS path)
   putStrLn @Text "Parsing definition file(s)"
   tops <- M.traverseWithKey (\defFile _ -> LD.parseFile $ toS defFile) defSource
           >>= (fold >>> dataOrError defSource)

### 2019-05-24 syncon-parser/src/Pre.hs
Fixed bug in FileAnnotation, annotations starting in the same position were in reverse order

@@ -63,7 +63,7 @@ equalBy f = foldMap (f >>> equal) >>> isEqual
 
 -- | Make a 'HashMap' given a function to generate a key and a collection of values
 mapFromFoldable :: (Eq b, Hashable b, Foldable t) => (a -> b) -> t a -> HashMap b [a]
-mapFromFoldable getName = toList >>> fmap (getName &&& pure) >>> M.fromListWith (<>)
+mapFromFoldable getName = toList >>> fmap (getName &&& pure) >>> M.fromListWith (flip (<>))
 
 instance Eq a => Semigroup (Equal a) where
   NotEqual <> _ = NotEqual

### 2019-05-24 syncon-parser/src/Main.hs
Factored out language generation from ambiguity reporting

@@ -30,7 +30,8 @@ import qualified P2LanguageDefinition.Elaborator as LD
 
 import qualified P4Parsing.Types as Parser
 import qualified P4Parsing.Parser as Parser
-import qualified P4Parsing.AmbiguityReporter as Parser
+
+import qualified P5DynamicAmbiguity.AmbiguityReporter as DynAmb
 
 import qualified Data.Automaton.NVA as NVA
 import qualified Data.Automaton.GraphViz as GraphViz
@@ -111,7 +112,7 @@ parseToHTMLDebug defFiles sourceFile outFile = do
   fileSource <- readFile sourceFile <&> M.singleton (toS sourceFile)
   putStrLn @Text "Parsing source file"
   setOfNodes <- parseFile sourceFile >>= dataOrError fileSource
-  case Parser.report df setOfNodes of
+  case DynAmb.report df setOfNodes of
     Data node -> universe node >>= nodeAnnotation
       & annotate fileSource
       & putInTextTemplate (toS $(embedFile "resources/htmlTemplate.html"))

### 2019-05-24 syncon-parser/src/P4Parsing/Parser.hs

@@ -5,6 +5,7 @@ module P4Parsing.Parser
 ( Error(..)
 , SingleLanguage(..)
 , parseSingleLanguage
+, mkParser
 ) where
 
 import Pre
@@ -72,6 +73,16 @@ parseSingleLanguage df = do
       (res, _) -> return $ S.fromList res
   where
     lexer = Lexer.allOneLanguage SingleLanguage (dfToLanguageTokens df) & first (fmap LexingError)
+
+-- | Given a 'DefinitionFile', produce a parser function for a single language. This function does
+-- is to be applied directly to a list of tokens, as opposed to a file.
+mkParser :: DefinitionFile -> Res SL ([Tok SL] -> Res SL (HashSet (Node SL TypeName)))
+mkParser df = do
+  parser <- generateGrammar df
+  pure $ Earley.fullParses (Earley.parser $ unquant parser) >>> \case
+    ([], Report{expected, unconsumed = next : _}) -> Error [UnexpectedToken (S.fromList expected) next]
+    ([], Report{expected, unconsumed = []}) -> Error [UnexpectedEOF $ S.fromList expected]
+    (res, _) -> return $ S.fromList res
 
 -- | Look up keywords, comment syntax, etc., to produce the parameters for the lexer.
 dfToLanguageTokens :: DefinitionFile -> LanguageTokens TypeName

### 2019-05-24 syncon-parser/src/P5DynamicAmbiguity/AmbiguityReporter.hs

@@ -1,0 +1,112 @@
+{-# LANGUAGE RecordWildCards #-}
+{-# LANGUAGE ViewPatterns #-}
+
+module P5DynamicAmbiguity.AmbiguityReporter
+( report
+, Error(..)
+) where
+
+import Pre hiding (reduce)
+import Result (Result(..))
+
+import Text.Printf (printf)
+import qualified Data.Text as Text
+import qualified Data.HashSet as S
+import qualified Data.HashMap.Lazy as M
+import Data.Semigroup (Max(..))
+import System.IO.Unsafe (unsafePerformIO)
+
+import Data.Functor.Foldable (project)
+
+import Data.Automaton.NVA (NVA, shortestUniqueWord, shortestWord)
+import Data.Automaton.EpsilonNVA (untag, TaggedTerminal(..))
+
+import ErrorMessage (FormatError(..), simpleErrorMessage)
+
+import P1Lexing.Types (Range, range, textualRange)
+import P2LanguageDefinition.Types (Name(..), TypeName(..), DefinitionFile(..))
+import P4Parsing.Types (pattern Node, n_name)
+import qualified P4Parsing.Types as P4
+import P4Parsing.Parser (SingleLanguage(..))
+import P5DynamicAmbiguity.TreeLanguage (treeLanguage)
+import P5DynamicAmbiguity.Types
+
+data Error
+  = Ambiguity Range [Text]
+  | Unresolvable Range [Text] [Node]  -- ^ Resolvable alternatives, and unresolvable alternatives
+  deriving (Show)
+
+type Node = P4.Node SingleLanguage TypeName
+
+-- TODO: pluralize the text below appropriately
+instance FormatError Error where
+  formatError (Unresolvable r resolvable trees) = simpleErrorMessage r $
+    "Unresolvable ambiguity error with " <> show (length trees + length resolvable) <> " alternatives.\n" <>
+    formattedResolvable <> "\n" <>
+    "Unresolvable alternatives:\n" <>
+    fold unresolvables
+    where
+      format (name, children) = "  " <> coerce name <> formatChildren children <> "\n"
+      formatChildren = sortBy (compare `on` range) >>> foldMap formatNode
+      formatNode n@Node{n_name = Name n'} =
+        printf "\n   - %- 20s %s" n' (textualRange $ range n) & Text.pack
+      formattedResolvable
+        | null resolvable = ""
+        | otherwise = "\nResolvable alternatives:\n" <> foldMap (\t -> "  " <> t <> "\n") resolvable
[…]

### 2019-05-24 syncon-parser/src/P5DynamicAmbiguity/TreeLanguage.hs

@@ -1,117 +1,53 @@
 {-# LANGUAGE RecordWildCards #-}
 {-# LANGUAGE ViewPatterns #-}
 
-module P4Parsing.AmbiguityReporter
-( report
-, Error(..)
-) where
+module P5DynamicAmbiguity.TreeLanguage (treeLanguage) where
 
 import Pre hiding (reduce)
-import Result (Result(..))
 
-import Text.Printf (printf)
-import qualified Data.Text as Text
 import qualified Data.HashSet as S
-import qualified Data.HashMap.Lazy as M
+import qualified Data.HashMap.Strict as M
 import qualified Data.Sequence as Seq
 import Data.Sequence (pattern Empty, pattern (:<|), pattern (:|>))
-import Data.Semigroup (Max(..))
-import System.IO.Unsafe (unsafePerformIO)
 
 import Data.Generics.Uniplate.Data (universe)
-import Data.Functor.Foldable (project)
 
-import Data.Automaton.NVA (fromEpsNVA, NVA, reduce, shortestUniqueWord, shortestWord, renumber)
-import Data.Automaton.EpsilonNVA (EpsNVA(..), untag, fromNFA, TaggedTerminal(..))
+import Data.Automaton.EpsilonNVA (EpsNVA(..), fromNFA, TaggedTerminal(..))
 import qualified Data.Automaton.EpsilonNVA as EpsNVA
+import Data.Automaton.NVA (fromEpsNVA, NVA, reduce, renumber)
 import Data.Automaton.Regex (Regex(..))
 import qualified Data.Automaton.Regex as Regex
 
-import ErrorMessage (FormatError(..), simpleErrorMessage)
-
-import P1Lexing.Types (Range, range, textualRange)
+import P1Lexing.Types (range)
 import qualified P1Lexing.Types as P1
-import P2LanguageDefinition.Types (Name(..), SDName(..), TypeName(..), Syncon(..), SyntaxDescription(..), BracketKind(..), DefinitionFile(..), Repetition(..))
 import P2LanguageDefinition.Elaborator (elaborate)
+import P2LanguageDefinition.Types (Name(..), SDName(..), TypeName(..), Syncon(..), SyntaxDescription(..), BracketKind(..), DefinitionFile(..), Repetition(..))
+import P4Parsing.Parser (SingleLanguage(..))
 import P4Parsing.Types (pattern Node, n_name, n_contents, NodeInternals(..))
[…]

### 2019-05-24 syncon-parser/src/P5DynamicAmbiguity/Types.hs

@@ -1,0 +1,30 @@
+module P5DynamicAmbiguity.Types where
+
+import Pre
+
+import qualified P2LanguageDefinition.Types as P2
+
+data Token
+  = LitTok Text
+  | OtherTokInstance P2.TypeName Text
+  | OtherTok P2.TypeName
+  deriving (Show)
+
+eitherRepr :: Token -> Either Text P2.TypeName
+eitherRepr (LitTok t) = Left t
+eitherRepr (OtherTokInstance n _) = Right n
+eitherRepr (OtherTok n) = Right n
+
+textualToken :: Token -> Text
+textualToken (LitTok t) = t
+textualToken (OtherTokInstance _ t) = t
+textualToken (OtherTok tyn) = coerce tyn  -- TODO: make this more distinct somehow
+
+mkTok :: Either Text P2.TypeName -> Token
+mkTok (Left t) = LitTok t
+mkTok (Right n) = OtherTok n
+
+instance Eq Token where
+  (==) = (==) `on` eitherRepr
+instance Hashable Token where
+  hashWithSalt = hashUsing eitherRepr

### 2019-05-30 paper/paper.tex
EOD

@@
 \newcommand{\yield}{\mathit{yield}} % yield of a parse tree
-\newcommand{\semantic}{\mathit{semantic}} % remove semantically unimportant productions from a w' \in L(G')
+\newcommand{\semantic}{\mathit{unparen}} % remove semantically unimportant productions from a w' \in L(G')
 \newcommand{\parse}{\mathit{parse}} % go from a w' \in L(G'_w) to a subset of L(G_t)
@@
 
-\begin{quote} % TODO: remove
-  \textbf{NOTE:} The text before this section is largely out of date.
-\end{quote}
-% TODO: note that the syntax tree definition is a little bit non-standard
-
-\subsection{Context-Free Grammars}
+This section briefly describes the theoretical foundations we build upon.
+Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation.
+Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}.
+Finally, Sections~\ref{sec:preliminaries-trees} and \ref{sec:preliminaries-ambiguity} describe trees and ambiguity, the former with some non-standard notation and the latter with a slightly wider definition than normal.
+
+\subsection{Context-Free Grammars} \label{sec:preliminaries-cfgs}
 
@@
 
-\subsection{Syntax Trees}
-
-A syntax tree takes the terminals of a word and arranges them into a tree, the shape of which is determined by how (i.e., using which productions) the word was parsed / recognized.
-For this paper we will use a linear encoding of syntax trees, letting us use context-free grammars to express the set of syntax trees for some given word language.
-For example, the parse trees of a context-free grammar $G_w = (\NT, \T, P, S)$ are described by the context-free grammar $G_t = (\NT, \T \cup \T_{[]}, P', S)$ where:
-
-\begin{itemize}
-\item $T_{[]} = \bigcup_{p \in P} \{ [_p, ]_p \}$, i.e., a unique pair of brackets per production.
-\item $P' = \{ N -> [_p \alpha ]_p \mid p \in P \land p = N -> \alpha \}$, i.e., each production is wrapped with a unique pair of brackets.
-\end{itemize}
-
-\noindent Such a grammar will (slightly erroneously) be referred to as a tree grammar, and the recognized language as a tree language.
-We will also use syntax trees that are not directly derived from the parse trees of a context-free grammar.
-The precise construction of these will be described at the point of their introduction, but the basic process is the same; surround each production with a unique pair of brackets.
-
-% TODO: reference a book on tree automata and languages
-As such we can define a function $\yield$ (as seen, e.g., in ) that takes a tree and flattens it, producing the word it was constructed from.
-Intuitively, this function merely removes the added brackets.
-
-\subsection{Ambiguity}
-
-The standard definition of ambiguity, using the above formulation of syntax trees, is as follows:
-
-\begin{definition}
-A word $w \in L(G_w)$ is ambiguous relative its parse tree $G_t$ if:
-  \[ \exists t_1, t_2 \in L(G_t).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2)\]
-\end{definition}
-
-\noindent For this paper, we widen the definition slightly to the following:
-
-\begin{definition}
-  A word $w \in L(G_w)$ is ambiguous relative a tree grammar $G_t$ and a function $\parse : L(G_w) -> 2^{L(G_t)}$ if:
-  \[ |\parse(w)| > 1 \]
-\end{definition}
-
-\noindent where $2^S$ is the powerset of $S$ and $|S|$ is the cardinality of $S$. The standard ambiguity definition is regained by choosing
-
-\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(G_t) \}\ \]
-
-\subsection{Automata}
+\subsection{Automata} \label{sec:preliminaries-automata}
 
@@
 \item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
-\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet.
+\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet, including a special bottom of stack symbol $\bot$.
 \item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma^{*}}$.
@@
 
-\subsection{Visibly Pushdown Languages}
+\subsection{Visibly Pushdown Languages} \label{sec:preliminaries-vpls}
 
@@
 \end{itemize}
+
+\subsection{Unranked Regular Tree Grammars} \label{sec:preliminaries-trees}
+
+Trees generalize words by allowing each terminal to have more than one successor.
+The successors have an associated ordering, i.e., in the presence of multiple successors, one is considered to be first, one second, etc.
+Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors.
+This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed.
+In this case, the sequence of successors to a single terminal tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.
+
+Our later results and properties are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
+We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves, to give a more natural definition of $\yield$.
+
+An unranked tree grammar $T$ is a tuple $(\NT, \T, X, P, S)$ where:
+
+\begin{itemize}
+\item $\NT$ is a set of (zero-arity) non-terminals.
+\item $\T$ is a set of zero-arity terminals, used as leaves.
+\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
+\item $P$ is a set of productions, where a production has the form $N -> x(r)$ for $N \in \NT$, $x \in X$, and $r$ a regular expression over $\NT$ and $\T$.
+\end{itemize}
+
+\noindent A tree $t$ is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
+The set of trees recognized by $T$ is written $L(T)$\footnote{All word grammars will be named $G$ (possibly with a subscript), while all tree grammars will be named $T$ (possibly with a subscript), lessening the risk of confusion between the two cases.}, and is said to be the tree language of $T$.
+
+$\yield$ then, is the sequence $w \in \T^{*}$ of terminals $t \in \T$ obtained by a left-to-right traversal of a tree.
+Informally, it is the flattening of a tree after all internal nodes have been removed.
+
+\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
+
+The standard definition of ambiguity, given a grammar $G$ and its parse trees $T_G$, is as follows:
+
+\begin{definition}
+A word $w \in L(G)$ is ambiguous if:
+  \[ \exists t_1, t_2 \in L(T_G).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2) \]
+\end{definition}
+
+\noindent For this paper, we widen the definition slightly to the following:
+
+\begin{definition}
+  A word $w \in L(G)$ is ambiguous relative a tree grammar $T$ and a function $\parse : L(G) -> 2^{L(T)}$ if:
+  \[ |\parse(w)| > 1 \]
+\end{definition}
+
+\noindent where $2^S$ is the powerset of $S$ and $|S|$ is the cardinality of $S$. The standard ambiguity definition is regained by choosing
+
+\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(T_G) \}\ \]
 
@@
 \begin{figure}
-  \begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
     $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
@@
 
-From $D$ we then generate four grammars: $G_w$, $G_t$, $G'_w$, and $G'_t$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:
-
-\begin{itemize}
-\item $G_w$ represents a word language describing all semantically distinct programs.
-\item $G_t$ represents a tree language describing the syntax trees obtained by parsing a word in $L(G_w)$.
-Note that the added brackets are not present in the productions of $E_{l1}$ and $E_{l2}$, since those are all part of the same labelled production in $D$ in Figure~\ref{fig:running-example-definition}.
-\item $G'_w$ is essentially a modified version of $G_w$, e.g., adding parentheses and other forms of disambiguation.
-\item $G'_t$ represents a tree language describing the syntax trees obtained by parsing a word in $L(G'_w)$.
+From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:
+
+\begin{itemize}
+\item $G_D$ represents a word language describing all semantically distinct programs.
+\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
+\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation.
+\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
 \end{itemize}
@@
     \end{tabular}
-    \caption{$G_w$, the generated abstract syntax.}
+    \caption{$G_D$, the generated abstract syntax.}
     \label{fig:running-example-generated:w}
@@
     \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
+    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
       \toprule
-      $E$ & $[_{l}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l}$ \\
-      $E$ & $[_{a}$ & $E$ \verb|'+'| $E$ & $]_{a}$ \\
-      $E$ & $[_{m}$ & $E$ \verb|'*'| $E$ & $]_{m}$ \\
-      $E$ & $[_{n}$ & $N$ & $]_{n}$ \\
-      \midrule
-      $E_{l1}$ & & $\epsilon$ & \\
-      $E_{l1}$ & & $E$ $E_{l2}$ & \\
-      \midrule
-      $E_{l2}$ & & $\epsilon$ & \\
-      $E_{l2}$ & & \verb|';'| $E$ $E_{l2}$ & \\
+      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
+      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
+      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
+      $E$ & $n$ & $N$ $)$ \\
       \bottomrule
     \end{tabular}
-    \caption{$G_t$, a language describing a linear encoding of the parse trees of $G_w$.}
+    \caption{$T_D$, the parse trees of $G_D$.}
     \label{fig:running-example-generated:t}
@@
     \end{tabular}
-    \caption{$G'_w$, the generated concrete syntax.}
+    \caption{$G'_D$, the generated concrete syntax.}
     \label{fig:running-example-generated:w-prime}
@@
     \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}lll@{}}
+    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
       \toprule
-      $E$ & $[_{l, \emptyset}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l, \emptyset}$ \\
-      $E$ & $[_{a, \emptyset}$ & $E$ \verb|'+'| $E$ & $]_{a, \emptyset}$ \\
-      $E$ & $[_{m, \emptyset}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ & $]_{m, \emptyset}$ \\
-      $E$ & $[_{n, \emptyset}$ & $N$ & $]_{n, \emptyset}$ \\
-      $E$ & $[_{g, \emptyset}$ & \verb|'('| $E$ \verb|')'| & $]_{g, \emptyset}$ \\
+      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
+      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
+      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
+      $E$ & $n$ & $N$ $)$ \\
+      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
       \midrule
-      $E_{\{a\}}$ & $[_{l, \{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| & $]_{l, \{a\}}$ \\
-      $E_{\{a\}}$ & $[_{m, \{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ & $]_{m, \{a\}}$ \\
-      $E_{\{a\}}$ & $[_{n, \{a\}}$ & $N$ & $]_{n, \{a\}}$ \\
-      $E_{\{a\}}$ & $[_{g, \{a\}}$ & \verb|'('| $E$ \verb|')'| & $]_{g, \{a\}}$ \\
-      \midrule
-      $E_{l1}$ & & $\epsilon$ & \\
-      $E_{l1}$ & & $E$ $E_{l2}$ & \\
-      \midrule
-      $E_{l2}$ & & $\epsilon$ & \\
-      $E_{l2}$ & & \verb|';'| $E$ $E_{l2}$ & \\
+      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
+      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
+      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
+      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
       \bottomrule
     \end{tabular}
-    \caption{$G'_t$, a language describing a linear encoding of the parse trees of $G'_w$.}
+    \caption{$T'_D$, the parse trees of $G'_D$.}
     \label{fig:running-example-generated:t-prime}
@@
     \[1 + 2 * 3\]
-    \caption{Example word in $L(G_w)$.}
+    \caption{Example word in $L(G_D)$.}
   \end{subfigure}
@@
         [.{$a$}
-          [.{$n$} 1 ]
-          +
-          [.{$n$} 2 ] ]
-        *
-        [.{$n$} 3 ] ]
+          [.{$n$} \terminal{1} ]
+          \terminal{+}
+          [.{$n$} \terminal{2} ] ]
+        \terminal{*}
+        [.{$n$} \terminal{3} ] ]
     \end{center}
-    \[[_m [_a [_n \terminal{1} ]_n \terminal{+} [_n \terminal{2} ]_n ]_a \terminal{*} [_n \terminal{3} ]_n ]_m\]
-    \caption{Example word in $L(G_T)$ and visualized tree.}
+    \[m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))\]
+    \caption{Example tree in $L(T_D)$ and visualization.}
   \end{subfigure}
@@
     \[(1 + 2) * 3\]
-    \caption{Example word in $L(G'_w)$.}
+    \caption{Example word in $L(G'_D)$.}
   \end{subfigure}
@@
     \begin{center}
-    \Tree [.{$m, \emptyset$}
-        [.{$g, \{a\}$}
-          (
-          [.{$a, \emptyset$}
-            [.{$n, \emptyset$} 1 ]
-            +
-            [.{$n, \emptyset$} 2 ] ]
-          ) ]
-        *
-        [.{$n, \{a\}$} 3 ] ]
+    \Tree [.{$m$}
+        [.{$g$}
+          \terminal{(}
+          [.{$a$}
+            [.{$n$} \terminal{1} ]
+            \terminal{+}
+            [.{$n$} \terminal{2} ] ]
+          \terminal{)} ]
+        \terminal{*}
+        [.{$n$} \terminal{3} ] ]
     \end{center}
-    \[[_{m} [_{g, \{a\}} \terminal{(} [_{a} [_{n} \terminal{1} ]_{n} \terminal{+} [_{n} \terminal{2} ]_{n} ]_{a} \terminal{)} ]_{g, \{a\}} \terminal{*} [_{n, \{a\}} \terminal{3} ]_{n, \{a\}} ]_{m}\]
-    \caption{Example word in $L(G'_T)$ and visualized tree.}
+    \[m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))\]
+    \caption{Example tree in $L(T'_D)$ and visualization.}
   \end{subfigure}
-  \caption{Example with words from each grammar that correspond to each other.
-We abbreviate $[_{n, \emptyset}$ as $[_n$.
-The non-bracket terminals in the tree languages appear underlined for increased clarity.}
+  \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
   \label{fig:example-words-in-square}
@@
 At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses.
-For example, $G_w$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
+For example, $G_D$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
 However, starting with that language definition would allow the expression ''$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ''$;2$'' is not a valid expression.
 
-As an example, in Figure~\ref{fig:example-grammar}, \ref{fig:example-grammar:ambig-grammar} is the \emph{what} and \ref{fig:example-grammar:unambig-grammar} is the \emph{how}.
-The latter grammar is a modification of the former that adds precedence, associativity, and parentheses, yielding an unambiguous grammar with at least one way to express each semantically distinct tree.
-
-\begin{figure*}
-  \begin{subfigure}{.45\linewidth}
-    \setlength{\grammarindent}{5em}
-    \begin{grammar}
-      <expr> ::= Sum: <expr> '+' <expr>
-        \alt Product: <expr> '*' <expr>
-        \alt Number: <number>
-    \end{grammar}
-    \caption{The (ambiguous) intuitive grammar without parentheses.}
-    \label{fig:example-grammar:ambig-grammar}
-  \end{subfigure}
-  \begin{subfigure}{0.45\linewidth}
-    \setlength{\grammarindent}{5em}
-    \begin{grammar}
-      <expr> ::= Sum: <product> '+' <expr>
-        \alt PassthroughE: <product>
-
-      <product> ::= Product: <atom> '*' <product>
-        \alt PassthroughP: <atom>
-
-      <atom> ::= Paren: '(' <expr> ')'
-        \alt Number: <number>
-    \end{grammar}
-    \caption{The (unambiguous) grammar with parentheses.}
-    \label{fig:example-grammar:unambig-grammar}
-  \end{subfigure}
-
-  \begin{subfigure}{.55\linewidth}
-    \begin{center}
-    \Tree [.PassthroughE
-      [.\underline{Product}
-        [.Paren
-          '('
-          [.\underline{Sum}
-            [.PassthroughP [.\underline{Number} 1 ] ]
-            '+'
-            [.PassthroughE [.PassthroughP [.\underline{Number} 2 ] ] ] ]
-          ')' ]
-        '*'
-        [.PassthroughP [.\underline{Number} 3 ] ] ] ]
-    \end{center}
-    \caption{Parse tree for $(1 + 2) * 3$. The underlined nodes are semantically important.}
-    \label{fig:example-grammar:tree}
-  \end{subfigure}
-  \begin{subfigure}{.35\linewidth}
-    \begin{center}
-    \Tree [.\underline{Product}
-      [.\underline{Sum} [.\underline{Number} 1 ] '+' [.\underline{Number} 2 ] ]
-      '*'
-      [.\underline{Number} 3 ] ]
-    \end{center}
-    \caption{The same parse tree, after removing the semantically unimportant nodes.}
-    \label{fig:example-grammar:sem-tree}
-  \end{subfigure}
-
-  \caption{A basic expression grammar in two variations, and two example syntax trees.}
-  \label{fig:example-grammar}
-\end{figure*}
-
-Our definition thus refers to four grammars in total:
-
-\begin{itemize}
-  \item The semantic grammar $G$, containing only the semantically important productions (e.g., Figure~\ref{fig:example-grammar:ambig-grammar}).
-  \item $T_G$ (generally abbreviated as $T$), the parse trees of $G$, representing the trees that must be expressible (Figure~\ref{fig:example-grammar:sem-tree} is in this language).
-  \item The parse grammar $G'$, a modification of $G$ meant to actually be used for parsing (e.g., Figure~\ref{fig:example-grammar:unambig-grammar}).
-The identifiers in this grammar need not be unique, but should be a superset of the identifiers in $G$.
-The modifications we will consider in this paper are introduced in Section~\ref{sec:modifications}.
-  \item $T_{G'}$ (generally abbreviated as $T'$), the parse trees of $G'$ (Figure~\ref{fig:example-grammar:tree} is in this language).
-\end{itemize}
-
-\noindent We also require a function $\semantic : L(T') -> L(T)$ that removes the semantically unimportant productions from a parse tree.
+We also require a function $\semantic : L(T'_D) -> L(T_D)$ that removes the semantically unimportant productions from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
 The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}.
-Finally, we define the function $\parse : L(G') -> 2^{L(T)}$, and its inverse $\words : L(T) -> 2^{L(G')}$:
+Finally, we define the function $\parse : L(G'_D) -> 2^{L(T_D)}$, and its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
 
@@
 \begin{array}{rcl}
-\parse(w') & = & \{ \semantic(t') \mid t' \in L(T') \land \yield(t') = w' \} \\
+\parse(w') & = & \{ \semantic(t') \mid t' \in L(T'_D) \land \yield(t') = w' \} \\
 \words(t) & = & \{ w' \mid t \in \parse(w') \} \\
@@
 
-\noindent The latter will be useful later, while $\parse$ is used directly in the definition of resolvable ambiguity:
+\noindent The latter will be useful later, while $\parse$ is used directly in the definition of \emph{resolvable ambiguity}:
 
@@
   \begin{tikzcd}
-    L(G) & L(T) \arrow[l, "\yield"] \\
-    L(G') & L(T') \arrow[u, "\semantic"'] \arrow[l, "\yield"]
+    L(G_D) & L(T_D) \arrow[l, "\yield"] \\
+    L(G'_D) & L(T'_D) \arrow[u, "\semantic"'] \arrow[l, "\yield"]
   \end{tikzcd}
-  \caption{The grammars considered, and their relation to each other.
-$G$ is provided by a user of the system, along with instructions how to modify $G$ to construct $G'$, while $T$ and $T'$ are automatically derived.}
+  \caption{The grammars considered, and their relation to each other.}
   \label{fig:grammar-square}
@@
 \begin{definition}
-  A language defined by the semantic grammar $G$ and the parse grammar $G'$ is resolvable if:
+  A word $w \in L(G'_D)$ defined by the language definition $D$ is resolvably ambiguous if:
   $$
   \begin{array}{l}
-  \forall t \in L(T_G).\\
-  \quad \exists w' \in L(G').\ \parse(w') = \{t\}
+  \forall t \in \parse(w).\\
+  \quad \exists w' \in L(G'_D).\ \parse(w') = \{t\}
   \end{array}
@@
 
-\noindent Intuitively, a language is resolvably ambiguous if there is at least one (unambiguous\footnote{For a slightly different meaning of unambiguous, it's here relating $G'$ to $T$, instead of $G'$ to $T'$.}) word for each semantically distinct tree.
-Note that neither $G$ nor $G'$ necessarily need to be unambiguous for this to hold.
-
-\subsection{Modifications in G'} \label{sec:modifications} % TODO: can't typeset G' in mathmode for some reason
-
-For this paper we consider two possible modifications:
-
-\begin{itemize}
-\item Adding parentheses for explicit grouping.
-\item Forbidding non-terminals in certain positions from expanding using certain productions.
-\end{itemize}
-
-\noindent The latter requires more explanation. As an example, in the following grammar we are forbidding the second \synt{<expr>} in the first production from expanding using the ''Sum'' production:
-
-\setlength{\grammarindent}{5em}
-\begin{grammar}
-  <expr> ::= Sum: <expr> '+' <expr>$_{Sum}$
-    \alt Number: <number>
-\end{grammar}
-
-\noindent Transforming this into a normal grammar, and adding parentheses, we get the following:
-
-\setlength{\grammarindent}{6em}
-\begin{grammar}
-  <expr> ::= Sum: <expr> '+' <expr>$_{Sum}$
-  \alt Number: <number>
-  \alt Paren: '(' <expr> ')'
-
-  <expr$_{Sum}$> ::= Number: <number>
-  \alt Paren: '(' <expr> ')'
-\end{grammar}
-
-\noindent In other words, we produce a grammar where addition is left-associative\footnote{I'm having some issues typesetting this appropriately, and I'm not all that happy with the explanation.}, but we can use explicit grouping via parentheses to override that.
+\noindent As an example, in the language defined in Figure~\ref{fig:running-example-definition}, the word \verb|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
+
+\begin{center}
+  \begin{tabular}{l}
+    $t_1 = a($ \hphantom{$a($} $n($ \verb|'1'| $)$ \verb|'+'| $a($ $n($ \verb|'2'| $)$ \hphantom{$)$} \verb|'+'| $n($ \verb|'3'| $)$ $)$ $)$ \\
+    $t_2 = a($ $a($ $n($ \verb|'1'| $)$ \verb|'+'| \hphantom{$a($} $n($ \verb|'2'| $)$ $)$ \verb|'+'| $n($ \verb|'3'| $)$ \hphantom{$)$} $)$ \\
+  \end{tabular}
+\end{center}
+
+\noindent This ambiguity can be resolved however, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
+To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
+
+\begin{center}
+  \begin{tabular}{l}
+    $t_3 = l($ \verb|'['| \hphantom{$s($} $n($ \verb|'1'| $)$ \verb|';'| $n($ \verb|'2'| $)$ \hphantom{$)$} \verb|']'| $)$ \\
+    $t_4 = l($ \verb|'['| $s($ $n($ \verb|'1'| $)$ \verb|';'| $n($ \verb|'2'| $)$ $)$ \verb|']'| $)$ \\
+  \end{tabular}
+\end{center}
+
+\noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not.
+The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_1\}$.
+
+We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, alternatively:
+
+\begin{definition}
+  A language defined by the language definition $D$ is resolvably ambiguous if:
+  $$
+  \begin{array}{l}
+  \forall t \in L(T_D).\\
+  \quad \exists w' \in L(G'_D).\ \parse(w') = \{t\}
+  \end{array}
+  $$
+\end{definition}
+
+\noindent Intuitively, a language is resolvably ambiguous if there is at least one (unambiguous\footnote{Using $\parse$ and $T_D$, i.e., relating $G'_D$ and $T_D$, instead of $G'_D$ and its parse trees.}) word for each semantically distinct tree.
+Note that neither $G_D$ nor $G'_D$ necessarily need to be unambiguous for this to hold.
+
+We can now state the two central problems we wish to consider:
+
+\begin{itemize}
+\item Static resolvability analysis: determine if a given language definition $D$ is resolvably ambiguous, otherwise produce an unresolvably ambiguous word $w \in L(G'_D)$.
+Section~\ref{sec:static} provides a partial solution to this problem.
+\item Dynamic resolvability analysis: given a language definition $D$ and a word $w \in L(G'_D)$, determine if it is resolvably ambiguous, and if so, present an unambiguous word for every tree $t \in \parse(w)$.
+Section~\ref{sec:dynamic} provides a full solution with one caveat: it only considers languages with balanced parentheses.
+\end{itemize}
+
+\noindent Additionally, on the dynamic side, there is another property of great practical relevance; we want ambiguities to be minimal and independent.
+In a practical setting, it is rarely helpful to merely state that an entire program is ambiguous.
+Instead, we wish to minimize the reported ambiguous portion of the program to, e.g., a sigle expression.
+Section~\ref{sec:local} addresses this.
+
+\section{Static Resolvability Analysis} \label{sec:static}
+
+This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted.
+The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$.
+Alternatively, find a tree that has no unambiguous words.
 
@@
 This section introduces an alternative (isomorphic) definition of a word that will be used to motivate the correctness of later algorithms.
-The method by which we generate $G_w$ and $G'_w$ limits the possible differences between them significantly.
+The method by which we generate $G_D$ and $G'_D$ limits the possible differences between them significantly.
 In particular, no new terminals are introduced, except '(' and ')', and they are always introduced in a well-balanced fashion.
@@
 \end{itemize}
-
-\section{Static Resolvability Check}
-
-This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted.
-The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$.
-Alternatively, find a tree that has no unambiguous words.
 
@@
 
-\subsection{Forbidden Children}
-
-\begin{quote}
-  \textbf{NOTE:} I had some text here about version 2 of the problem, but I haven't had time to rewrite it to align properly with the lattice formulation, so I've removed it.
-I'll talk about it during the meeting.
-Also, I probably need a better title here, this is probably a smidge too dramatic.
-\end{quote}
+\section{Dynamic Resolvability Analysis} \label{sec:dynamic}
+
+\section{Local Ambiguities} \label{sec:local}
 

### 2019-05-30 syncon-parser/src/Main.hs
More flexible/nice CLI

@@ -9,15 +9,22 @@ import Result (Result(..))
 import FileAnnotation (annotate, putInTextTemplate)
 import ErrorMessage (FormatError, formatErrors, formatError)
 
+import System.Environment (withArgs)
+import System.FilePath ((</>))
+
 import Data.Data (Data)
+import Data.List (partition)
 import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
 import qualified Data.Sequence as Seq
+import qualified Data.Text as Text
 import Data.FileEmbed (embedFile)
 
 import Data.Generics.Uniplate.Data (universe, universeBi)
 import Data.Functor.Foldable (project, cata)
 import Text.Show.Pretty (pPrint)
+
+import qualified Options.Applicative as Opt
 
 import P1Lexing.Types (Range(..), range, textualRange)
 import qualified P1Lexing.Types as Lexer
@@ -37,98 +44,6 @@ import qualified P5DynamicAmbiguity.TreeLanguage as DynAmb
 
 import qualified Data.Automaton.NVA as NVA
 import qualified Data.Automaton.GraphViz as GraphViz
-
-synconTokens :: Lexer.LanguageTokens LD.TokenKind
-synconTokens = Lexer.LanguageTokens
-  -- Literal tokens
-  [ "token", "=", "syncon", ":", "{", ";", "}", "prefix", "postfix", "infix"
-  , "(", ")", "*", "+", "?", ".", "comment", "left", "right", "precedence", "except"
-  , "type", "builtin", "forbid", "|", "rec", "grouping", "!" ]
-  -- Regex tokens
-  [ (LD.NameTok, (Nowhere, "[[:lower:]][[:word:]]*"))
-  , (LD.TypeNameTok, (Nowhere, "[[:upper:]][[:word:]]*"))
-  , (LD.StringTok, (Nowhere, "\"(\\\\.|[^\"\\\\])*\"")) ]
-  -- Comment regex
-  [((Nowhere, "//[^\\n]*(\\n|$)"), (Nowhere, "^"))]
-
-lexTest :: IO ()
-lexTest = do
-  res <- Lexer.allOneLanguage' @Text "SynconDef" synconTokens "examples/bootstrap.syncon"
-  pPrint $ fmap Lexer.textualToken <$> res
-
-parseTest :: IO ()
-parseTest = do
-  res <- LD.parseFile "examples/bootstrap.syncon"
-  pPrint res
-
[…]

### 2019-05-30 syncon-parser/src/P5DynamicAmbiguity/AmbiguityReporter.hs

@@ -24,11 +24,11 @@ import Data.Automaton.EpsilonNVA (untag, TaggedTerminal(..))
 import ErrorMessage (FormatError(..), simpleErrorMessage)
 
 import P1Lexing.Types (Range, range, textualRange)
-import P2LanguageDefinition.Types (Name(..), TypeName(..), DefinitionFile(..))
+import P2LanguageDefinition.Types (Name(..), TypeName(..))
 import P4Parsing.Types (pattern Node, n_name)
 import qualified P4Parsing.Types as P4
 import P4Parsing.Parser (SingleLanguage(..))
-import P5DynamicAmbiguity.TreeLanguage (treeLanguage, PreLanguage, precompute)
+import P5DynamicAmbiguity.TreeLanguage (treeLanguage, PreLanguage)
 import P5DynamicAmbiguity.Types
 
 data Error
@@ -63,16 +63,14 @@ instance FormatError Error where
 
 -- | If the argument is a singleton set, return the element, otherwise produce
 -- one or more localized ambiguity errors.
-report :: DefinitionFile -> HashSet Node -> Result [Error] Node
-report df = toList >>> \case
+report :: PreLanguage -> HashSet Node -> Result [Error] Node
+report pl = toList >>> \case
   [] -> compErr "P4Parsing.AmbiguityReporter.report" "Expected one or more parse trees, got zero."
   [top] -> Data top
   forest -> localizeAmbiguities forest
     <&> ((head >>> foldMap range) &&& S.fromList)
     <&> uncurry (resolvability pl)
     & Error
-  where
-    pl = precompute df
 
 localizeAmbiguities :: (Eq l, Eq n) => [P4.Node l n] -> [[P4.Node l n]]
 localizeAmbiguities forest

### 2019-05-30 paper/paper.tex
EOD

@@
 \begin{figure}
+\centering
+\begin{minipage}{.47\textwidth}
+  \centering
   \begin{tabular}{@{}ll@{}}
@@
       Marks & $m \subseteq \Labels$ \\
-      Regular expressions & $r ::= t \mid N_m \mid r \cdot r \mid r + r \mid \epsilon \mid r^{*}$ \\
+      Regular expressions & $r ::= t \mid N_m \mid r \cdot r $ \\
+      & \hphantom{$r ::=$}$\mid r + r \mid \epsilon \mid r^{*}$ \\
       Labelled productions & $N -> l : r$ \\
   \end{tabular}
-  \caption{The abstract syntax of a language definition.}
+  \captionof{figure}{The abstract syntax of a language definition.}
   \label{fig:input-language-definition}
-\end{figure}
-
-\begin{figure}
+\end{minipage}\quad
+\begin{minipage}{.45\textwidth}
+  \centering
   \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
@@
   \end{tabular}
-  \caption{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity.
+  \captionof{figure}{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity.
 Assumes that $N$ matches a numeric terminal.}
   \label{fig:running-example-definition}
+\end{minipage}
 \end{figure}
@@
 \begin{figure}
+
+  \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
+      \toprule
+      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
+      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
+      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
+      $E$ & $n$ & $N$ $)$ \\
+      \bottomrule
+    \end{tabular}
+    \caption{$T_D$, the parse trees of $G_D$.}
+    \label{fig:running-example-generated:t}
+  \end{subfigure}%
+%
+  \begin{subfigure}[t]{.45\linewidth}
+    \centering
+    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
+      \toprule
+      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
+      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
+      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
+      $E$ & $n$ & $N$ $)$ \\
+      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
+      \midrule
+      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
+      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
+      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
+      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
+      \bottomrule
+    \end{tabular}
+    \caption{$T'_D$, the parse trees of $G'_D$.}
+    \label{fig:running-example-generated:t-prime}
+  \end{subfigure}
+
   \begin{subfigure}[t]{.45\linewidth}
@@
 %
-  \begin{subfigure}[t]{.54\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
-      \toprule
-      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
-      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
-      $E$ & $n$ & $N$ $)$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$T_D$, the parse trees of $G_D$.}
-    \label{fig:running-example-generated:t}
-  \end{subfigure}
-
   \begin{subfigure}[t]{.45\linewidth}
@@
     \label{fig:running-example-generated:w-prime}
-  \end{subfigure}%
-%
-  \begin{subfigure}[t]{.54\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
-      \toprule
-      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
-      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
-      $E$ & $n$ & $N$ $)$ \\
-      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
-      \midrule
-      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
-      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
-      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$T'_D$, the parse trees of $G'_D$.}
-    \label{fig:running-example-generated:t-prime}
-  \end{subfigure}%
+  \end{subfigure}
   \caption{The generated grammars.}
@@
 \begin{figure}
-  \begin{subfigure}[b]{.40\linewidth}
-    \[1 + 2 * 3\]
-    \caption{Example word in $L(G_D)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.55\linewidth}
+  \begin{subfigure}[b]{.45\linewidth}
     \begin{center}
@@
   \end{subfigure}
-
-  \begin{subfigure}[b]{.40\linewidth}
-    \[(1 + 2) * 3\]
-    \caption{Example word in $L(G'_D)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.55\linewidth}
+  \begin{subfigure}[b]{.45\linewidth}
     \begin{center}
@@
   \end{subfigure}
+
+  \begin{subfigure}[b]{.40\linewidth}
+    \[1 + 2 * 3\]
+    \caption{Example word in $L(G_D)$.}
+  \end{subfigure}
+  \begin{subfigure}[b]{.40\linewidth}
+    \[(1 + 2) * 3\]
+    \caption{Example word in $L(G'_D)$.}
+  \end{subfigure}
   \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
@@
   \begin{tikzcd}
-    L(G_D) & L(T_D) \arrow[l, "\yield"] \\
-    L(G'_D) & L(T'_D) \arrow[u, "\semantic"'] \arrow[l, "\yield"]
+    L(T_D) \arrow[d, "\yield"] & L(T'_D) \arrow[l, "\semantic"'] \arrow[d, "\yield"] \\
+    L(G_D) & L(G'_D)
   \end{tikzcd}
@@
 
-\noindent This ambiguity can be resolved however, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
+\noindent This ambiguity can be resolved, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
 To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
@@
 \noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not.
-The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_1\}$.
-
-We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, alternatively:
+The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.
+
+We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, or more concisely:
 
@@
 In a practical setting, it is rarely helpful to merely state that an entire program is ambiguous.
-Instead, we wish to minimize the reported ambiguous portion of the program to, e.g., a sigle expression.
+Instead, we wish to minimize the reported ambiguous portion of the program to, e.g., a single expression.
 Section~\ref{sec:local} addresses this.
@@
 This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted.
-The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$.
+The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists.
 Alternatively, find a tree that has no unambiguous words.

### 2019-05-31 paper/paper.tex
Static intro

@@
 
-Trees generalize words by allowing each terminal to have more than one successor.
-The successors have an associated ordering, i.e., in the presence of multiple successors, one is considered to be first, one second, etc.
+Trees generalize words by allowing each terminal to have multiple ordered successors, instead of just zero or one.
 Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors.
@@
 
-\noindent A tree $t$ is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
+\noindent A tree $t$ (containing only terminals from $\T$ and $X$) is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
 The set of trees recognized by $T$ is written $L(T)$\footnote{All word grammars will be named $G$ (possibly with a subscript), while all tree grammars will be named $T$ (possibly with a subscript), lessening the risk of confusion between the two cases.}, and is said to be the tree language of $T$.
@@
 \begin{figure}
-
   \begin{subfigure}[t]{.45\linewidth}
@@
 \noindent Examples of words in each of these four languages can be seen in Figure~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
+As a memory aid, the prime versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the straightforward translations from $D$.
 
@@
 
-This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted.
-The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists.
-Alternatively, find a tree that has no unambiguous words.
+To determine if a given language definition $D$ is resolvably ambiguous we are tasked with the following: find a tree $t \in L(T_D)$ such that there is no $w \in L(G'_D)$ for which $\parse(w) = \{t\}$, or prove that no such tree exists.
+Or, more briefly put: find a tree that has no unambiguous words or show that none exist.
+
+Before an actual algorithm, we split the problem into three different versions of increasing difficulty.
+We then lay the groundwork for our correctness proofs, eventually culminating in the algorithm, which we show to be sound and complete for version 1 and sound for version 2.
+
+The three versions are as follows:
+
+\begin{description}
+\item[Version 1] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$ and no non-terminals appear marked, i.e., for all non-terminals $N_m$ appearing on the right-hand side of the labelled productions in $D$, we have $m = \emptyset$.
+\item[Version 2] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$.
+\item[Version 3] There are no constraints on $D$.
+\end{description}
+
+\noindent The first restriction in version 1 states that $D$ cannot contain parentheses.
+This implies that all parentheses present in $G'_D$ are grouping parentheses.
+The second restriction implies that no parentheses are required (e.g., assuming normal precedence, the parentheses in ''$(1 + 2) * 3$'' are required, removing them would produce a semantically different tree).
+
+The reason for this split stems from the following insight: double grouping parentheses do not matter, in the sense that, e.g., $\parse($\verb|'((1 + 2)) * 3'|$) = \parse($\verb|'(1 + 2) * 3'|$)$.
+Version 1 and 2 have only grouping parentheses, meaning that no double parentheses matter.
+What follows is a brief outline of the remainder of this section, building from that observation.
+Note that this outline is unlikely to be easy to understand initially, but is rather there to serve as a context to place the various steps in later\footnote{The intent here is to place references to subsections dealing with each of these points.
+The number of referenced sections will likely not be exactly equal to the number of bulletpoints.}.
+
+\begin{itemize}
+\item We can consider an alternate representation of words that split non-parenthesis terminals from the ranges covered by parentheses, the former as a word, the latter as a multiset or a bag.
+\item In version 1 and 2 the rangebag can be treated as a set, since double parentheses do not matter.
+\item This gives rise to a partitioning of $\words(t)$ for any given $t \in L(T_D)$, where the partitions form a lattice.
+\item That lattice is bounded and uniquely determined by its top and bottom elements.
+\item We can construct a linear representation (i.e., a word) of these lattices.
+\item We can construct a visibly pushdown automaton that recognizes these linear representations, where each successful run corresponds exactly to the tree that generates that lattice.
+\item We can modify the VPDA to recognize larger lattices, i.e., where the top element is greater and/or the bottom element is lesser, while still retaining the connection to the generating trees.
+\item By examining the product automaton of these two automata we can determine whether there exists a pair of distinct trees, such that the lattice of one entirely covers the lattice of the other.
+This is sufficient for completeness in version 1, but not in version 2.
+\end{itemize}
 

### 2019-06-02 syncon-parser/src/P2LanguageDefinition/Elaborator.hs
Fix #10

@@ -16,6 +16,10 @@ import P2LanguageDefinition.Types
 -- associativities on the same levels merely have to stay undefined, there is no good definition.
 data Assoc = AssocLeft | AssocRight
 
+-- | The forbids are generated differently depending on what kind of operator it is, e.g.,
+-- a prefix operator will never be forbidden in RRec, since there's no ambiguity there.
+data OpKind = InfixOp | PrefixOp | PostfixOp | OtherOp
+
 -- | Produce a single structure containing all disambiguation. Note that there won't be a mapping
 -- for the (qualified) sdnames that have no marks, i.e., conceptually the total map is obtained
 -- through `M.lookupDefault S.empty (name, sdname) (elaborate ...)`.
@@ -50,6 +54,16 @@ elaborate syncons forbids (PrecedenceMatrix mat) = elaboratedForbids : assocEntr
     hasAssocForbid :: Name -> Rec -> Bool
     hasAssocForbid name r = M.lookup (name, Left r) elaboratedForbids & fold & S.member name
 
+    opKinds :: HashMap Name OpKind
+    opKinds = recs & M.mapMaybe getKind
+    getKind :: HashSet Rec -> Maybe OpKind
+    getKind recs' = case (LRec `S.member` recs', Rec `S.member` recs', RRec `S.member` recs') of
+      (_, True, _) -> Just OtherOp
+      (True, _, True) -> Just InfixOp
+      (True, _, False) -> Just PostfixOp
+      (False, _, True) -> Just PrefixOp
+      (False, _, False) -> Nothing
+
     -- NOTE: (<>) is ok here because we know that their keySets are disjoint
     processEntry :: (Name, Name) -> Ordering -> Elaboration
     processEntry (n1, n2) EQ = case (M.lookup n1 assocs, M.lookup n2 assocs) of
@@ -63,9 +77,16 @@ elaborate syncons forbids (PrecedenceMatrix mat) = elaboratedForbids : assocEntr
     assocEntries = mat <&> fst & M.mapWithKey processEntry & toList
 
     forbid :: Name -> Rec -> Name -> Elaboration
-    forbid n1 r n2 = if M.lookup n1 recs & fold & S.member r
-      then M.singleton (n1, Left r) (S.singleton n2)
-      else M.empty
+    forbid n1 r n2
+      | LRec <- r
+      , Just PostfixOp <- mKind = M.empty
+      | RRec <- r
[…]

### 2019-06-02 syncon-parser/src/P5DynamicAmbiguity/TreeLanguage.hs
Fix bug when constructing a tree language with a capture inside a '?'

@@ -359,7 +359,7 @@ regexIntermission fullSd start end (S.delete end -> others) = case (recur fullSd
         result = recur sd path Start
     recur (SDRep _ RepQuestion sd) path target@Path{} = recur sd path target
     recur (SDRep _ RepQuestion sd) path Start = recur sd path Start
-      & mapResult (Choice Eps)
+      & mapPartial (Choice Eps)
       & asNonFail
 
     recur sd path target = compErr "P4Parsing.AmbiguityReporter.RegexIntermission.recur" $
@@ -374,3 +374,5 @@ regexIntermission fullSd start end (S.delete end -> others) = case (recur fullSd
     mapResult f (Done a) = Done $ f a
     mapResult f (Partial a) = Partial $ f a
     mapResult _ Fail = Fail
+    mapPartial f (Partial a) = Partial $ f a
+    mapPartial _ a = a

### 2019-06-16 paper/paper.tex
EOD

--- 2019-05-30 paper/paper.tex
@@
 
-\newcommand{\NT}{\mathbb{N}} % Set of nonterminals
+\newcommand{\support}{\mathit{Supp}}
+\newcommand{\NT}{V} % Set of nonterminals
 \newcommand{\T}{\Sigma} % Set of terminals
-\newcommand{\Labels}{\mathbb{L}} % Set of labels
+\newcommand{\Labels}{L} % Set of labels
 \newcommand{\yield}{\mathit{yield}} % yield of a parse tree
@@
 \newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
+\newcommand{\rangebag}{\mathit{rangebag}} % go from a w \in L(G'_w) to the first element of \alt(w).
+\newcommand{\rangeset}{\mathit{rangeset}} % go from a w \in L(G'_w) to the first element of \alt(w).
 \newcommand{\altset}{\mathit{altset}} % go from a w \in L(G'_w) to \alt(w), except the second element is replaced by a set (that contains an element if the bag contained at least one of that element).
@@
 \newcommand{\range}[2]{#1\!-\!#2}
+\newcommand{\regex}{\mathit{Reg}}
 
@@
 This section briefly describes the theoretical foundations we build upon.
-Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation.
+Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation to make later sections easier to read.
 Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}.
@@
 
+\subsection{Bags} \label{sec:bags}
+
+A \emph{bag} (alternatively known as a \emph{multiset}) is a generalization of a set, it is an unordered collection where each element may appear multiple times.
+The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$.
+A bag $A$ is included in another bag $B$, written $A \subseteq B$, iff $\forall x.\ m_A(x) \leq m_B(x)$.
+The underlying set of elements of a bag $B$ is called the \emph{support} of $B$, and is given by $\support(B) = \{ x \mid m_B(x) > 0 \}$.
+
+\subsection{Regular Expressions} \label{sec:preliminaries-regexes}
+
+A regular expression $r$ over alphabet $\T$ is defined inductively:
+
+$$r ::= \epsilon \mid a \mid r \cdot r \mid r + r \mid r^{*}$$
+
+\noindent where $a \in \T$. The language of a regular expression $r$ is given by $L(r)$:
+
+$$
+\begin{array}{r@{\ =\ }l}
+  L(t) & \{a\} \\
+  L(\epsilon) & \{\epsilon\} \\
+  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \\
+  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
+  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
+\end{array}
+$$
+
+\noindent We will denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
+
 \subsection{Context-Free Grammars} \label{sec:preliminaries-cfgs}
 
-A context-free grammar $G$ is a tuple $(\NT, \T, P, S)$ where:
-
-\begin{itemize}
-\item $\NT$ is a set of non-terminals.
-\item $\T$, disjoint from $\NT$, is a set of terminals.
-\item $P$, a subset of $\NT \times (\NT \cup \T)^{*}$, is a set of productions.
-\item $S \in \NT$ is the starting non-terminal.
-\end{itemize}
-
-\noindent A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
-The set of words recognized by $G$ is written $L(G)$, and is said to be the word language of $G$.
+A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is Kleene-star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
+
+ A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
+Such a sequence is called a \emph{derivation}.
+The set of words recognized by $G$ is written $L(G)$\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.
+
+\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
+
+The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{left-most derivations}.
+A left-most derivation is a derivation where the non-terminal being replaced is always the left-most one.
+
+\begin{definition}
+A word $w \in L(G)$ is ambiguous if there are two distinct left-most derivations of that word.
+\end{definition}
 
@@
 
-A nondeterministic finite automaton (NFA) is a tuple $(Q, \Sigma, \delta, q_0, F)$:
-
-\begin{itemize}
-\item A finite set of states $Q$.
-\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
-\item A transition function $\delta: Q \times \Sigma -> 2^Q$.
-\item An initial state $q_0 \in Q$.
-\item A set of final states $F \subseteq Q$.
-\end{itemize}
-
-\noindent A successful run is a sequence of states $r_0, r_1, \ldots, r_n$ and a word $a_0a_1\ldots a_n$ such that:
+A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$ ; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
+
+A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that:
 
@@
 A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to.
-
-A pushdown automaton extends a finite automaton with a stack, and lets each transition push or pop symbols from it.
-Formally, a nondeterministic pushdown automaton is a tuple $(Q, \Sigma, \Gamma, \delta, q_0, F)$:
-
-\begin{itemize}
-\item A finite set of states $Q$.
-\item A finite set of input symbols $\Sigma$, i.e., an input alphabet.
-\item A finite set of stack symbols $\Gamma$, i.e., a stack alphabet, including a special bottom of stack symbol $\bot$.
-\item A transition function $\delta: Q \times \Sigma \times \Gamma -> 2^{Q \times \Gamma^{*}}$.
-\item An initial state $q_0 \in Q$.
-\item A set of final states $F \subseteq Q$.
-\end{itemize}
-
-\noindent A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma*$.
+NFAs and DFAs have the same expressive power as regular expressions, i.e., for every regular expression there is an NFA and a DFA both reconizing the same language, and vice-versa.
+% TODO: probably reference for this
+
+A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols.
+Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
+$\lambda$ essentially means ''ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).
+
+A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma^{*}$.
 
@@
 \begin{itemize}
-\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all.
-\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack.
-\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack (i.e., this transition cannot be taken if $g$ is not on top of the stack).
+\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all, i.e., $\delta(p, a, \lambda) \supseteq \{(q, \lambda)\}$.
+\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack, i.e., $\delta(p, a, \lambda) \supseteq \{(q, g)\}$.
+\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack, i.e., $\delta(p, a, g) \supseteq \{(q, \lambda)\}$.
 \end{itemize}
@@
 A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
-A visibly pushdown automaton is a pushdown automaton where the input alphabet $\Sigma$ can be partitioned into three disjoint sets $\Sigma_c$, $\Sigma_i$, and $\Sigma_r$, such that all transitions in the automaton has one of the following three forms:
-
-\begin{itemize}
-\item $p \xrightarrow{c, +s} q$, where $c \in \Sigma_c$.
-\item $p \xrightarrow{i} q$, where $i \in \Sigma_i$.
-\item $p \xrightarrow{r, -s} q$, where $r \in \Sigma_r$.
+A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton has one of the following three forms:
+
+\begin{itemize}
+\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$.
+\item $p \xrightarrow{i} q$, where $i \in \T_i$.
+\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$.
 \end{itemize}
@@
 
-This gives us some nice properties. Of particular relevance to this paper are the following two points:
-
-\begin{itemize}
-\item If two visibly pushdown automata have the same partitions $\Sigma_c$, $\Sigma_i$, and $\Sigma_r$, then we can construct a product automaton that simulates two simultaneous runs through both automata.
-This product automaton has the same input alphabet partitions, each state is a pair of states (one from each automaton), and each stack symbol is a pair of stack symbols (one from each automaton).
+The partition names stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
+We will instead primarily use $\T_c$ and $\T_r$ for balanced parentheses.
+
+This partitioning gives us some useful properties. Of particular relevance to this paper are the following two points:
+
+\begin{itemize}
+\item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}.
+Intersection in particular is given by a product automaton, i.e., given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
+  $$
+  \begin{array}{r@{\ =\ }ll}
+    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \text{where } c \in \T_c \\
+    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \text{where } i \in \T_i \\
+    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \text{where } r \in \T_r \\
+  \end{array}
+  $$
+
 \item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run, none are redundant.
+Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
 \end{itemize}
@@
 
-Trees generalize words by allowing each terminal to have more than one successor.
-The successors have an associated ordering, i.e., in the presence of multiple successors, one is considered to be first, one second, etc.
+Trees generalize words by allowing each terminal to have multiple ordered successors, instead of just zero or one.
 Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors.
 This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed.
-In this case, the sequence of successors to a single terminal tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.
-
-Our later results and properties are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
-We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves, to give a more natural definition of $\yield$.
+The sequence of successors to a single terminal in an unranked tree tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.
+
+The results and properties presented in this paper are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
+We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves.
+Since we will use unranked trees to represent parse trees, the former will represent terminals from the parsed word, while the latter represent terminals introduced as internal nodes.
 
@@
 \item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
-\item $P$ is a set of productions, where a production has the form $N -> x(r)$ for $N \in \NT$, $x \in X$, and $r$ a regular expression over $\NT$ and $\T$.
-\end{itemize}
-
-\noindent A tree $t$ is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
-The set of trees recognized by $T$ is written $L(T)$\footnote{All word grammars will be named $G$ (possibly with a subscript), while all tree grammars will be named $T$ (possibly with a subscript), lessening the risk of confusion between the two cases.}, and is said to be the tree language of $T$.
-
-$\yield$ then, is the sequence $w \in \T^{*}$ of terminals $t \in \T$ obtained by a left-to-right traversal of a tree.
+\item $P$ is a set of productions, a finite subset of $\NT \times X \times \regex(\T \cup X)$. We will write a production $(N, x, r)$ as $N -> x(r)$.
+\end{itemize}
+
+\noindent A tree $t$ (containing only terminals from $\T$ and $X$) is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
+The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.
+
+Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree.
 Informally, it is the flattening of a tree after all internal nodes have been removed.
 
-\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
-
-The standard definition of ambiguity, given a grammar $G$ and its parse trees $T_G$, is as follows:
+\section{Resolvable Ambiguity} \label{sec:resolvable-definition}
+
+This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to some more standard concepts in formal languages.
+
+Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense.
+We will call this meaning an \emph{interpretation} of a word.
+Note that a single word may have multiple interpretations.
+As such, we will define a language as:
+
+\begin{itemize}
+\item An alphabet $\T$.
+\item A set of interpretations $T$.
+\item A function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
+\end{itemize}
+
+\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees.
+In such a language, $\parse(1 + 2 * 3)$ would produce a set containing the two ASTs in Figure~\ref{fig:arith-example}:
+
+\begin{figure}[t]
+\begin{tabular}{cc}
+  \Tree [.{$+$}
+    1
+    [.{$*$}
+      2
+      3 ] ] &
+  \Tree [.{$*$}
+    [.{$+$}
+      1
+      2 ]
+    3 ] \\
+\end{tabular}
+\caption{The two allowable interpretations of $1 + 2 * 3$ in an arithmetic language without precedence.}
+\label{fig:arith-example}
+\end{figure}
+
+\noindent We further classify each word $w \in \T^{*}$ depending on the result of $\parse$:
 
 \begin{definition}
-A word $w \in L(G)$ is ambiguous if:
-  \[ \exists t_1, t_2 \in L(T_G).\ t_1 \neq t_2 \land \yield(t_1) = w = \yield(t_2) \]
+  A word $w \in \T^{*}$ is:
+  \begin{itemize}
+  \item not in the language if $\parse(w) = \emptyset$.
+  \item unambiguous if $\exists t \in T.\ \parse(w) = \{t\}$.
+  \item ambiguous if $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$.
+  \end{itemize}
 \end{definition}
 
-\noindent For this paper, we widen the definition slightly to the following:
+\noindent We can connect these cases to the classical definition of a language:
+
+\begin{itemize}
+\item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
+\item If we select a $\parse$ that relates words to their left-most derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
+\end{itemize}
+
+\noindent A \emph{resolvably} ambiguous word is a word where all its interpretations can be written in an unambiguous way, or more formally:
 
 \begin{definition}
-  A word $w \in L(G)$ is ambiguous relative a tree grammar $T$ and a function $\parse : L(G) -> 2^{L(T)}$ if:
-  \[ |\parse(w)| > 1 \]
+  A word $w \in \T^{*}$ in a language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-word}
 \end{definition}
 
-\noindent where $2^S$ is the powerset of $S$ and $|S|$ is the cardinality of $S$. The standard ambiguity definition is regained by choosing
-
-\[ \parse(w) = \{ t \mid w = \yield(t) \land t \in L(T_G) \}\ \]
-
-\section{Parse-time Disambiguation}
-
-We begin this section with some motivation, then list our definition of \emph{resolvable ambiguity} and what disambiguation we will consider, and finally an alternative definition of a word, which will be useful in later sections.
-
-We can divide the productions present in a programming language grammar in two groups: those that are semantically important, and those that are semantically \emph{un}important.
-The former group covers most productions, while the latter contains, e.g., parentheses used for explicit grouping.
-If programs were written directly as syntax trees then the latter would be unnecessary; two syntax trees that differ only by parentheses are semantically the same.
-
-Thus we wish the output of parsing to be a syntax tree consisting entirely of semantically important productions.
-This distinction is useful to make, because it allows us to decouple \emph{what} we want to be expressible and \emph{how} it is to be expressed.
+\noindent The example in Figure~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way.
+However, if we add grouping parentheses to the language we find that $1 + (2 * 3)$ and $(1 + 2) * 3$ are unambiguously interpreted as the first and second tree, respectively.
+
+Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, or more formally:
+
+\begin{definition}
+  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall w \in \T^{*}.\ \forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-language}
+\end{definition}
+
+\noindent We can now state a few things:
+
+\begin{itemize}
+\item An unambigous word $w$ is trivially resolvably ambiguous, since its only interpretation $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$).
+The set of resolvably ambiguous words is thus a superset of the unambiguous words.
+\item If a given interpretation $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous.
+In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
+\end{itemize}
+
+\noindent The second point suggests that resolvable ambiguity is only an interesting property if an element of $T$ does not uniquely identify an element of $\T^{*}$.
+Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual interpretation.
+Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace.
+In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the interpretation.
+
+We thus propose to loosen the common ''no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous.
+However, merely having an arbitrary function $\parse$ gives us very little to work with, and no way to decide whether the language it defines is resolvably ambiguous or not.
+The remainder of this paper will thus consider $\parse$ functions defined with a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, that gives us some decidable properties.
+
+Before introducing this formalism however, we introduce the two central problems we consider in this paper:
+
+\begin{description}
+\item[Static resolvability.] Given a language defined by $\parse$, determine if it is resolvably ambiguous.
+\item[Dynamic resolvability.] Given a language defined by $\parse$ and a word $w$ such that $\parse(w) \neq \emptyset$, determine if $w$ is resolvably ambiguous.
+\end{description}
+
+\noindent Our main concern is producing decision procedures for these problems. As such, we define soundness and completeness for the static problem:
+
+\begin{definition}
+  A decision procedure $f$ solving the static resolvability problem is:
+  \begin{itemize}
+  \item Sound if $f(\parse) = \mathit{resolvable}$ implies that the language given by $\parse$ is resolvably ambiguous, and $f(\parse) = \mathit{unresolvable}$ implies that the language given by $\parse$ is not resolvably ambiguous.
+  \item Complete if $f$ terminates and produces a result for all possible inputs $\parse$.
+  \end{itemize}
+  \label{def:static-procedure}
+\end{definition}
+
+\noindent Similarly, for the dynamic problem:
+
+\begin{definition}
+  A decision procedure $f$ solving the dynamic resolvability problem is:
+  \begin{itemize}
+  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
+  \item Complete if $f$ terminates and produces a result for all possible inputs $\parse$ and $w$.
+  \end{itemize}
+  \label{def:dynamic-procedure}
+\end{definition}
+
+\section{Parse-time Disambiguation} \label{sec:parse-time-disambiguation}
+
+This section describes our chosen language definition formalism, and motivates its design.
+
+The primary purpose of this formalism is, as described in the previous section, to produce a $\parse$ function, i.e., to describe a word language and assign one or more interpretations to each word.
+The interpretations will be unranked trees, intended to be somewhat reminiscent of the abstract syntax trees used in most compilers.
+Section~\ref{sec:resolvable-definition} suggests that $\parse$ discard some information, to enable the resolution of some ambiguities; we here choose to discard grouping parentheses.
+
+With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}.
+Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label.
+Also note that the right-hand side of a production is here a regular expression, rather than the theoretically simpler sequence used in a context-free grammar.
+Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal.
+To lessen clutter, we will write $E_\emptyset$ as $E$.
+As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example.
+In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication.
+By ''direct child'' we mean ''without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.
 
@@
 
-With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}.
-Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label.
-A \emph{mark} is used for disambiguation, and forbids certain productions from taking the place of the marked non-terminal.
-To lessen clutter, we will write $E_\emptyset$ as $E$.
-As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example.
-In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a child of a multiplication, enforcing conventional precedence.
-
-From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$. The first two represent the \emph{what}, while the latter two represent the \emph{how}:
+From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$.
+Technically, only $G'_D$ and $T_D$ are required, $G'_D$ is used as the defined word language and $T_D$ as the interpretations, but the remaining two grammars help the presentation.
 
@@
 \item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
-\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation.
+\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation (i.e., the result of marks).
 \item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
@@
 
+\noindent Figure~\ref{fig:running-example-generated} contains the four grammars generated from our running example in Figure~\ref{fig:running-example-definition}.
+The context-free grammars are produced by a rather standard translation from regular expressions to CFGs, while the primed grammars get a new non-terminal per distinctly marked non-terminal in $D$, where each new non-terminal only has the productions whose label is not in the mark.
+For example, the non-terminal $E_{\{a\}}$ in Figure~\ref{fig:running-example-generated:t-prime} has no production corresponding to the $a$ production in Figure~\ref{fig:running-example-definition}.
+
 \begin{figure}
-
   \begin{subfigure}[t]{.45\linewidth}
@@
 
-\noindent Examples of words in each of these four languages can be seen in Figure~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
+Examples of elements in each of these four languages can be seen in Figure~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
+Each element corresponds to the word ''$(1 + 2) * 3$'' in $L(G'_D)$.
+Note that the word in $L(G_D)$ as ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ''$((1 + 2)) * 3$'' and ''$(1 + 2) * (3)$''.
+As a memory aid, the prime versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations from $D$.
+We will generally refer to elements of $L(G_D)$ as $w$, $L(G'_D)$ as $w'$, $L(T_D)$ as $t$, and $L(T'_D)$ as $t'$.
 
@@
 
-At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses.
+At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses; they are allowed exactly where they would surround a complete production.
 For example, $G_D$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
@@
 
-We also require a function $\semantic : L(T'_D) -> L(T_D)$ that removes the semantically unimportant productions from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
-The relation between the four grammars can be seen in Figure~\ref{fig:grammar-square}.
-Finally, we define the function $\parse : L(G'_D) -> 2^{L(T_D)}$, and its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
-
-$$
-\begin{array}{rcl}
-\parse(w') & = & \{ \semantic(t') \mid t' \in L(T'_D) \land \yield(t') = w' \} \\
-\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
-\end{array}
-$$
-
-\noindent The latter will be useful later, while $\parse$ is used directly in the definition of \emph{resolvable ambiguity}:
+Finally, we require a function $\semantic : L(T'_D) -> L(T_D)$ that removes grouping parentheses from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
+The relation between the four grammars in terms of $\yield$ and $\semantic$ can be seen in Figure~\ref{fig:grammar-square}.
+With this we can define $\parse : L(G'_D) -> 2^{L(T_D)}$, along with its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
 
@@
 
-\begin{definition}
-  A word $w \in L(G'_D)$ defined by the language definition $D$ is resolvably ambiguous if:
 $$
-  \begin{array}{l}
-  \forall t \in \parse(w).\\
-  \quad \exists w' \in L(G'_D).\ \parse(w') = \{t\}
+\begin{array}{rcl}
+\parse(w') & = & \{ \semantic(t') \mid t' \in L(T'_D) \land \yield(t') = w' \} \\
+\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
 \end{array}
 $$
-\end{definition}
-
-\noindent As an example, in the language defined in Figure~\ref{fig:running-example-definition}, the word \verb|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
+
+\noindent The latter is mostly useful in later sections, but $\parse$ allow us to consider some concrete examples of resolvable and unresolvable ambiguities.
+For example, in our running example (Figure~\ref{fig:running-example-definition}), the word \verb|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
 
@@
 
-\noindent This ambiguity can be resolved, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
+\noindent This is a resolvable ambiguity, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
 To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
@@
 
-We can expand this to an entire language, stating that a language is resolvably ambiguous iff all its words are resolvably ambiguous, or more concisely:
-
-\begin{definition}
-  A language defined by the language definition $D$ is resolvably ambiguous if:
-  $$
-  \begin{array}{l}
-  \forall t \in L(T_D).\\
-  \quad \exists w' \in L(G'_D).\ \parse(w') = \{t\}
-  \end{array}
-  $$
-\end{definition}
-
-\noindent Intuitively, a language is resolvably ambiguous if there is at least one (unambiguous\footnote{Using $\parse$ and $T_D$, i.e., relating $G'_D$ and $T_D$, instead of $G'_D$ and its parse trees.}) word for each semantically distinct tree.
-Note that neither $G_D$ nor $G'_D$ necessarily need to be unambiguous for this to hold.
-
-We can now state the two central problems we wish to consider:
-
-\begin{itemize}
-\item Static resolvability analysis: determine if a given language definition $D$ is resolvably ambiguous, otherwise produce an unresolvably ambiguous word $w \in L(G'_D)$.
-Section~\ref{sec:static} provides a partial solution to this problem.
-\item Dynamic resolvability analysis: given a language definition $D$ and a word $w \in L(G'_D)$, determine if it is resolvably ambiguous, and if so, present an unambiguous word for every tree $t \in \parse(w)$.
-Section~\ref{sec:dynamic} provides a full solution with one caveat: it only considers languages with balanced parentheses.
-\end{itemize}
-
-\noindent Additionally, on the dynamic side, there is another property of great practical relevance; we want ambiguities to be minimal and independent.
-In a practical setting, it is rarely helpful to merely state that an entire program is ambiguous.
-Instead, we wish to minimize the reported ambiguous portion of the program to, e.g., a single expression.
-Section~\ref{sec:local} addresses this.
+We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure} and \ref{def:dynamic-procedure}.
+Section~\ref{sec:static} gives a partial solution to the former problem, while Section~\ref{sec:dynamic} fully solves the latter with one caveat: we only consider languages with balanced parentheses.
 
@@
 
-This section describes a decision algorithm for detecting unresolvable ambiguities in a language definition, starting with a version with several limitations, most of which are later lifted.
-The overarching goal is to find a tree $t \in L(G_t)$ such that there is no $w' \in L(G'_w)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists.
-Alternatively, find a tree that has no unambiguous words.
+For this section, we will use an alternative formulation of resolvable ambiguity for languages, stated in terms of interpretations instead of words:
+
+\begin{theorem}
+  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
+\end{theorem}
+
+\begin{proof}
+  The quantifier $\forall t \in T$ is equivalent with $\forall w \in \T^{*}. \forall t \in \parse(w)$, since $T = \bigcup_{w \in \T^{*}} \parse(w)$ (by definition).
+\end{proof}
+
+\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w' \in L(G'_D)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists.
+Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exist.
+
+Before an actual algorithm, we split the problem into three different versions of increasing difficulty.
+We then lay the groundwork for our correctness proofs, eventually culminating in the algorithm, which we show to be sound and complete for version 1 and sound for version 2.
+
+The three versions are as follows:
+
+\begin{description}
+\item[Version 1] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$ and no non-terminals appear marked, i.e., for all non-terminals $N_m$ appearing on the right-hand side of the labelled productions in $D$, we have $m = \emptyset$.
+\item[Version 2] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$.
+\item[Version 3] There are no constraints on $D$.
+\end{description}
+
+\noindent The first restriction in version 1 states that $D$ cannot contain parentheses.
+This implies that all parentheses present in $G'_D$ are grouping parentheses.
+The second restriction implies that no parentheses are required (as a counterexample, assuming normal precedence, the parentheses in ''$(1 + 2) * 3$'' are required and removing them would produce a different interpretation).
+
+The reason for this split stems from the following insight: double grouping parentheses do not matter, in the sense that they do not change the interpretation of the word, e.g., $\parse($\verb|'((1 + 2)) * 3'|$) = \parse($\verb|'(1 + 2) * 3'|$)$.
+Versions~1 and 2 have only grouping parentheses, meaning that no double parentheses matter.
+What follows is a brief outline of the remainder of this section, which details our static analysis for version 1 and 2, building from that observation.
+
+\begin{itemize}
+\item We can consider an alternate representation of words where parentheses are not explicitly part of the string of terminals, but are represented by an accompanying bag of ranges denoting which terminals are covered by parentheses.
+(Section~\ref{sec:word-view}).
+\item This alternate representation gives rise to a lattice per tree in $L(T_D)$, where a tree has only ambiguous words iff its lattice is entirely covered by the lattices of other trees (Section~\ref{sec:lattice}).
+\item These lattices can be encoded as words, leading to the construction of a visibly pushdown automaton that we can examine to determine the existence of a tree whose lattice is covered by the lattice of a \emph{single} other tree, which is sufficient for completeness in version 1 (Section~\ref{sec:lattice-vpl}).
+\end{itemize}
 
@@
 
-This section introduces an alternative (isomorphic) definition of a word that will be used to motivate the correctness of later algorithms.
+This section introduces an alternative (isomorphic) definition of a word, heavily used in sections~\ref{sec:lattice} and \ref{sec:lattice-vpl}.
 The method by which we generate $G_D$ and $G'_D$ limits the possible differences between them significantly.
@@
 For example, the word ''$(1 + 2) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{1}{3}\})$, while ''$((1 + 2)) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{1}{3}, \range{1}{3}\})$.
-The first component will be referred to as a \emph{basic word}, the second as a \emph{range bag}
-
-We will now note a few things about the words $w' \in \words(t)$ for some given $t \in L(G_t)$:
-
-\begin{itemize}
-\item All $w'$ have the same basic word.
-$G'_w$ is constructed in such a way that the only possible difference is one of parentheses.
-This also implies that two trees that share a word (i.e., that can be ambiguous) must also share a basic word.
-\item Some parentheses are required, i.e., some ranges must be present in all $w'$.
-For example, removing the parentheses in ''$(1 + 2) * 3$'' changes the tree produced, since multiplication has higher precedence than addition.
-\item There is a finite set of possible ranges in the range bags of $w'$.
-Grouping parentheses can only be added if they exactly cover a node in the parse tree, and other parentheses can only be added where $G_w$ allows them.
-\item Duplicated grouping parentheses do not matter.
-For example, $((1 + 2)) * 3$ permits the same syntax trees as $(1 + 2) * 3$.
-If there are no parentheses present in $G_w$ then all parentheses are grouping parentheses, thus we can consider the range bag as a \emph{set} instead.
-\end{itemize}
-
-\subsection{Basic Algorithm}
+We will refer to the first component of the tuple as the \emph{basic word}, the second as the \emph{range bag}.
+
+We denote the function producing the alternative representation by $\alt$, while $\basic$ and $\rangebag$ directly produce the basic word and rangebag, respectively.
+
+The next four lemmas, all concerning the words in $\words(t)$ for some given $t \in L(G_t)$, form the basis of the lattices mentioned at the end of the previous section.
+
+\begin{lemma}
+  $\forall w'_1, w'_2 \in \words(t).\ \basic(w'_1) = \basic(w'_2)$, i.e., all $w'$ have the same basic word.
+  \label{lemma:same-basic}
+\end{lemma}
+\noindent Intuitively, $\words$ first applies the ''inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
+The only terminals changed in this process are parentheses, i.e., different $w'$ can only differ in their range bags.
+This also implies that two trees that share a word (i.e., two trees $t_1$ and $t_2$ that have a word $w'$ such that $\parse(w') \supseteq \{t_1, t_2\}$) must also share a basic word.
+
+\begin{lemma}
+  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\ R \subseteq \support(\rangebag(w'))$, i.e., some parentheses are required.
+  \label{lemma:required-parentheses}
+\end{lemma}
+\noindent For example, removing the parentheses in ''$(1 + 2) * 3$'' changes the interpretations produced.
+Each required range is a direct consequence of a mark on a non-terminal in $D$.
+We will write the word implied by the first quantified expression as $w'_\bot$.
+It is unique, since its basic word is fixed by $t$, and its rangebag is a set.
+
+\begin{lemma}
+  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\\ \support(\rangebag(w')) \subseteq R$, i.e., there is a finite set of possible parentheses.
+  \label{lemma:possible-parentheses}
+\end{lemma}
+\noindent As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes.
+We will write the word implied by the first quantified expression $w'_\top$.
+
+\begin{lemma}
+  $\forall w'_1, w'_2 \in \words(t).\ \support(\rangebag(w'_1)) = \support(\rangebag(w'_2)) => \parse(w'_1) = \parse(w'_2)$, i.e., duplicated parentheses do not matter.
+  \label{lemma:rangeset-equality}
+\end{lemma}
+\noindent We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses.
+Each pair of parentheses thus corresponds to a single grouping node $g$ in a tree in $L(T'_D)$.
+Duplicated grouping parentheses correspond to nested grouping nodes, all of which are removed by $\semantic$, thus producing identical sets of trees in $L(T_D)$.
+For example, $((1 + 2)) * 3$ has the same interpretations as $(1 + 2) * 3$.
+
+\subsection{A Lattice of Word Partitions} \label{sec:lattice}
+
+\begin{figure}[t]
+  \begin{tikzpicture}[every node/.style={rectangle,draw}]
+    \node (a) [draw=white] at (0,5) {$((1) + (2)) + (3)$};
+
+    \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
+    \node (b2) [draw=white] at (-1.2,4.2) {$((1) + (2)) + 3$};
+    \node (b3) [draw=white] at ( 1.2,4.2) {$((1) + 2) + (3)$};
+    \node (b4) [draw=white] at ( 3.6,4.2) {$(1 + (2)) + (3)$};
+
+    \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
+    \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
+    \node (c3) [draw=white] at (-1.1,3) {$((1) + 2) + 3$};
+    \node (c4) at (1.1,3) {$1 + (2) + (3)$};
+    \node (c5) [draw=white] at (3.3,3) {$(1 + (2)) + 3$};
+    \node (c6) [draw=white] at (5.5,3) {$(1 + 2) + (3)$};
+
+    \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
+    \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
+    \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
+    \node (d4) [draw=white] at ( 3.6,1.8) {$(1 + 2) + 3$};
+
+    \node (e) at (0, 1) {$1 + 2 + 3$};
+
+  \draw (e) -- (d1);
+  \draw (e) -- (d2);
+  \draw (e) -- (d3);
+  \draw (e) -- (d4);
+
+  \draw (d1.90) -- (c1.270);
+  \draw (d1.90) -- (c2.270);
+  \draw (d1.90) -- (c3.270);
+  \draw (d2.90) -- (c1.270);
+  \draw (d2.90) -- (c4.270);
+  \draw (d2.90) -- (c5.270);
+  \draw (d3.90) -- (c2.270);
+  \draw (d3.90) -- (c4.270);
+  \draw (d3.90) -- (c6.270);
+  \draw (d4.90) -- (c3.270);
+  \draw (d4.90) -- (c5.270);
+  \draw (d4.90) -- (c6.270);
+
+  \draw (c1.90) -- (b1.270);
+  \draw (c1.90) -- (b2.270);
+  \draw (c2.90) -- (b1.270);
+  \draw (c2.90) -- (b3.270);
+  \draw (c3.90) -- (b2.270);
+  \draw (c3.90) -- (b3.270);
+  \draw (c4.90) -- (b1.270);
+  \draw (c4.90) -- (b4.270);
+  \draw (c5.90) -- (b2.270);
+  \draw (c5.90) -- (b4.270);
+  \draw (c6.90) -- (b3.270);
+  \draw (c6.90) -- (b4.270);
+
+  \draw (b1) -- (a);
+  \draw (b2) -- (a);
+  \draw (b3) -- (a);
+  \draw (b4) -- (a);
+    %%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
+  \end{tikzpicture}
+  \caption{The lattice of words for the tree $a(a(n(1)+n(2))+n(3))$. The boxed words are shared with the lattice in Figure~\ref{fig:lattice2}.}
+  \label{fig:lattice1}
+\end{figure}
+
+Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w'$ by their \emph{rangeset}, where $\rangeset(w') = \support(\rangebag(w'))$.
+These partitions can be partially ordered by subset on the rangeset, resulting in a lattice of word partitions per tree.
+This lattice is bounded, with top and bottom elements given by lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} respectively.
+Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
+For example, Figure~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Figure~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}).
+Each partition is represented by the word whose rangebag is a set.
+To reduce clutter, we do not draw the partitions that have grouping parentheses around the entire word.
+This outermost possible pair is uninteresting since it is always allowed, and would double the size of the figure if it was included.
+
+\begin{figure}[t]
+
+\begin{tikzpicture}[every node/.style={rectangle,draw}]
+  \node (a) [draw=white] at (0,5) {$(1) + ((2) + (3))$};
+
+  \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
+  \node (b2) [draw=white] at (-1.2,4.2) {$(1) + ((2) + 3)$};
+  \node (b3) [draw=white] at ( 1.2,4.2) {$(1) + (2 + (3))$};
+  \node (b4) [draw=white] at ( 3.6,4.2) {$1 + ((2) + (3))$};
+
+  \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
+  \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
+  \node (c3) [draw=white] at (-1.1,3) {$(1) + (2 + 3)$};
+  \node (c4) at (1.1,3) {$1 + (2) + (3)$};
+  \node (c5) [draw=white] at (3.3,3) {$1 + ((2) + 3)$};
+  \node (c6) [draw=white] at (5.5,3) {$1 + (2 + (3))$};
+
+  \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
+  \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
+  \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
+  \node (d4) [draw=white] at ( 3.6,1.8) {$1 + (2 + 3)$};
+
+  \node (e) at (0, 1) {$1 + 2 + 3$};
+
+  \draw (e) -- (d1);
+  \draw (e) -- (d2);
+  \draw (e) -- (d3);
+  \draw (e) -- (d4);
+
+  \draw (d1.90) -- (c1.270);
+  \draw (d1.90) -- (c2.270);
+  \draw (d1.90) -- (c3.270);
+  \draw (d2.90) -- (c1.270);
+  \draw (d2.90) -- (c4.270);
+  \draw (d2.90) -- (c5.270);
+  \draw (d3.90) -- (c2.270);
+  \draw (d3.90) -- (c4.270);
+  \draw (d3.90) -- (c6.270);
+  \draw (d4.90) -- (c3.270);
+  \draw (d4.90) -- (c5.270);
+  \draw (d4.90) -- (c6.270);
+
+  \draw (c1.90) -- (b1.270);
+  \draw (c1.90) -- (b2.270);
+  \draw (c2.90) -- (b1.270);
+  \draw (c2.90) -- (b3.270);
+  \draw (c3.90) -- (b2.270);
+  \draw (c3.90) -- (b3.270);
+  \draw (c4.90) -- (b1.270);
+  \draw (c4.90) -- (b4.270);
+  \draw (c5.90) -- (b2.270);
+  \draw (c5.90) -- (b4.270);
+  \draw (c6.90) -- (b3.270);
+  \draw (c6.90) -- (b4.270);
+
+  \draw (b1) -- (a);
+  \draw (b2) -- (a);
+  \draw (b3) -- (a);
+  \draw (b4) -- (a);
+%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
+\end{tikzpicture}
+\caption{The lattice of words for the tree $a(n(1)+a(n(2)+n(3)))$. The boxed words are shared with the lattice in Figure~\ref{fig:lattice1}.}
+\label{fig:lattice2}
+\end{figure}
+
+To show the connection between resolvable ambiguity and these lattices, consider the word ''$1 + 2 + 3$''.
+It has two interpretations in our running example, $a(a(n(1)+n(2))+n(3))$, which we would normally write as $(1 + 2) + 3$, and $a(n(1)+a(n(2)+n(3)))$, which we would normally write $1 + (2 + 3)$.
+The lattices for these two trees are given in Figures~\ref{fig:lattice1} and \ref{fig:lattice2} respectively.
+The partitions that appear in both lattices are represented as boxed words, the others are unboxed.
+These shared partitions represent words that are ambiguous between these particular trees.
+Finding an unambiguous word is thus the same as finding a partition that is not shared with any other tree.
+In this particular case, there is no ambiguity with any other tree at all, and so the unboxed words are all valid resolutions of the ambiguity.
+
+At this point it is clear that a given tree has an unambiguous word iff its lattice has at least one partition that is not shared with any other tree.
+
+Next, we note that each lattice is uniquely determined by its top and bottom elements, it contains all elements between them:
+
+\begin{lemma}
+  Given $t \in L(T_D)$, $w'_\top, w'_\bot \in \words(t)$ such that $\forall w'.\ w' \in \words(t) => \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top)$ the following holds:\\
+  $\forall w'.\ \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top) => w' \in \words(t)$.
+\end{lemma}
+
+\noindent Lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} guarantee the existence of two such words $w'_\top$ and $w'_\bot$.
+
+We now have the tools we require to motivate our approach.
+
+\begin{theorem}
+  In version 1, for a given $t \in L(T_D)$, $\exists w'.\ \parse(w') = \{t\}$ iff $\parse(w'_\top) = \{t\}$ (where $w'_\top$ is given by Lemma~\ref{lemma:possible-parentheses}).
+\end{theorem}
+
+\subsection{A Lattice as a Word, and an Algorithm} \label{sec:lattice-vpl}
 
@@
 
-\section{Local Ambiguities} \label{sec:local}
+The dynamic resolvability problem is as follows: for a given $w' \in L(G'_D)$ determine whether $\forall t \in \parse(w').\ \exists w'_2.\ \parse(w'_2) = \{t\}$.
+Furthermore, for practical reasons, if the word is resolvably ambiguous we wish to produce a (minimal) witness for each tree.
+Additionally, while Section~\ref{sec:static} requires a language definition $D$ to not contain parentheses, this section merely requires parentheses to be balanced.
+
+Our approach centers around around $\words(t)$, which, being a set of words, can be considered a language.
+Each such language turns out to be a visibly pushdown language.
+Given a set of trees in $L(T_D)$ we can construct a corresponding set of non-overlapping visibly pushdown automata (i.e., each automaton only accepts words not accepted by any other automaton) since VPLs are closed under complement and intersection \cite{alurVisiblyPushdownLanguages2004}.
+These automata can then be examined to determine if they accept the empty language (which implies that the corresponding tree only has ambiguous words), or otherwise producing a witness.
+
+\begin{theorem}
+  Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
+\end{theorem}
+
+\section{Evaluation} \label{sec:evaluation}
+
+\subsection{OCaml} \label{sec:evaluation-ocaml}
+
+% TODO: talk about the ''low precedence unary operator'' issue.
+
+\section{Related Work}
+
+\cite{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
+However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
 
@@
 
-\begin{figure*}[ht]
-  \begin{tikzpicture}[every node/.style={rectangle,draw}]
-    \node (a) [draw=white] at (0,5) {$((1) + (2)) + (3)$};
-
-    \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
-    \node (b2) [draw=white] at (-1.2,4.2) {$((1) + (2)) + 3$};
-    \node (b3) [draw=white] at ( 1.2,4.2) {$((1) + 2) + (3)$};
-    \node (b4) [draw=white] at ( 3.6,4.2) {$(1 + (2)) + (3)$};
-
-    \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
-    \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
-    \node (c3) [draw=white] at (-1.1,3) {$((1) + 2) + 3$};
-    \node (c4) at (1.1,3) {$1 + (2) + (3)$};
-    \node (c5) [draw=white] at (3.3,3) {$(1 + (2)) + 3$};
-    \node (c6) [draw=white] at (5.5,3) {$(1 + 2) + (3)$};
-
-    \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
-    \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
-    \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
-    \node (d4) [draw=white] at ( 3.6,1.8) {$(1 + 2) + 3$};
-
-    \node (e) at (0, 1) {$1 + 2 + 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1.90) -- (c1.270);
-  \draw (d1.90) -- (c2.270);
-  \draw (d1.90) -- (c3.270);
-  \draw (d2.90) -- (c1.270);
-  \draw (d2.90) -- (c4.270);
-  \draw (d2.90) -- (c5.270);
-  \draw (d3.90) -- (c2.270);
-  \draw (d3.90) -- (c4.270);
-  \draw (d3.90) -- (c6.270);
-  \draw (d4.90) -- (c3.270);
-  \draw (d4.90) -- (c5.270);
-  \draw (d4.90) -- (c6.270);
-
-  \draw (c1.90) -- (b1.270);
-  \draw (c1.90) -- (b2.270);
-  \draw (c2.90) -- (b1.270);
-  \draw (c2.90) -- (b3.270);
-  \draw (c3.90) -- (b2.270);
-  \draw (c3.90) -- (b3.270);
-  \draw (c4.90) -- (b1.270);
-  \draw (c4.90) -- (b4.270);
-  \draw (c5.90) -- (b2.270);
-  \draw (c5.90) -- (b4.270);
-  \draw (c6.90) -- (b3.270);
-  \draw (c6.90) -- (b4.270);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-    %%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-  \end{tikzpicture}
-  \caption{The lattice of words for the tree $[_S [_S [_N 1 ]_N + [_N 2 ]_N ]_S + [_N 3 ]_N ]_S$ (in $L(G_t)$). The boxed words are shared with Figure~\ref{fig:lattice2}.}
-  \label{fig:lattice1}
-\end{figure*}
-
-\begin{figure*}[t]
-
-\begin{tikzpicture}[every node/.style={rectangle,draw}]
-  \node (a) [draw=white] at (0,5) {$(1) + ((2) + (3))$};
-
-  \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
-  \node (b2) [draw=white] at (-1.2,4.2) {$(1) + ((2) + 3)$};
-  \node (b3) [draw=white] at ( 1.2,4.2) {$(1) + (2 + (3))$};
-  \node (b4) [draw=white] at ( 3.6,4.2) {$1 + ((2) + (3))$};
-
-  \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
-  \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
-  \node (c3) [draw=white] at (-1.1,3) {$(1) + (2 + 3)$};
-  \node (c4) at (1.1,3) {$1 + (2) + (3)$};
-  \node (c5) [draw=white] at (3.3,3) {$1 + ((2) + 3)$};
-  \node (c6) [draw=white] at (5.5,3) {$1 + (2 + (3))$};
-
-  \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
-  \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
-  \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
-  \node (d4) [draw=white] at ( 3.6,1.8) {$1 + (2 + 3)$};
-
-  \node (e) at (0, 1) {$1 + 2 + 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1.90) -- (c1.270);
-  \draw (d1.90) -- (c2.270);
-  \draw (d1.90) -- (c3.270);
-  \draw (d2.90) -- (c1.270);
-  \draw (d2.90) -- (c4.270);
-  \draw (d2.90) -- (c5.270);
-  \draw (d3.90) -- (c2.270);
-  \draw (d3.90) -- (c4.270);
-  \draw (d3.90) -- (c6.270);
-  \draw (d4.90) -- (c3.270);
-  \draw (d4.90) -- (c5.270);
-  \draw (d4.90) -- (c6.270);
-
-  \draw (c1.90) -- (b1.270);
-  \draw (c1.90) -- (b2.270);
-  \draw (c2.90) -- (b1.270);
-  \draw (c2.90) -- (b3.270);
-  \draw (c3.90) -- (b2.270);
-  \draw (c3.90) -- (b3.270);
-  \draw (c4.90) -- (b1.270);
-  \draw (c4.90) -- (b4.270);
-  \draw (c5.90) -- (b2.270);
-  \draw (c5.90) -- (b4.270);
-  \draw (c6.90) -- (b3.270);
-  \draw (c6.90) -- (b4.270);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-\end{tikzpicture}
-\caption{The lattice of words for the tree $[_S [_N 1 ]_N + [_S [_N 2 ]_N + [_N 3 ]_N ]_S ]_S$ (in $L(G_t)$). The boxed words are shared with Figure~\ref{fig:lattice1}.}
-\label{fig:lattice2}
-\end{figure*}
-
 \clearpage
@@
 \end{theorem}
-
-%% \noindent Generalizing this to a complete grammar, we get the following property: a grammar $G$ is resolvable iff:
-
-%% $$
-%% \begin{array}{l}
-%%   \forall t \in L(T).\\
-%%   \quad \exists w' \in L(G').\\
-%%   \qquad \forall t' \in L(T').\ \yield(t') = w' -> \unparen(t') = t
-%% \end{array}
-%% $$
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% \begin{theorem}
-%%   If $\{(, )\} \cap \T = \emptyset$, then any unambiguous word $w \in L(G)$ is resolvable.
-%% \end{theorem}
-
-%% \noindent \textbf{Proof.} Since $w$ is unambiguous in $G$ there is exactly one parse tree $t \in L(T)$ such that $\yield(t) = w$.
-Since $L(G) \subseteq L(G')$ we have that $w \in L(G')$.
-Furthermore, $w$ is unambiguous in $L(G')$ as well, since no added productions can apply (they all contain parentheses, which cannot be present in $w$).
-Additionally, since $L(T) \subseteq L(T')$, we have that $t \in L(T')$, and since $\yield(t) = w$, it must be the only parse tree for $w$ in $L(T')$.
-Finally, $\unparen(t) = t$.
 

### 2019-06-18 paper/paper.tex
EOD

@@
 
-Text of paper \ldots
-
-\subsection{Motivating Ambiguity in Programming Languages}
-
-% TODO: reference PADL paper for this example
-Consider the following nested match expression in OCaml:
-
-\begin{lstlisting}[language={[objective]caml}]
+Formal languages have long had the concept of ambiguity, of words that can be constructed in multiple ways, for example using multiple distinct derivations of a context-free grammar.
+
+However, when using these formal methods to specify the syntax for programming languages, ambiguity becomes a strictly undesirable property.
+We tend to want to give the compiler a single interpretation of any given program, and so we work hard to make programming language grammars entirely unambiguous.
+This view, that eliminating ambiguity is \emph{the} correct thing to do, is highly prevalent (e.g.
+\cite{sudkampLanguagesMachinesIntroduction1997}).
+
+Some exceptions do exist however.
+Closest to our approach are \citet{danielssonParsingMixfixOperators2011}, who ''feel that it is overly restrictive to require the grammar to be unambiguous'', and suggest that only ambiguous \emph{parses} should be rejected, not ambiguous grammars.
+This seems to stem largely from the fact that their grammars are mostly constructed from user-defined operators, and so cannot be expected to compose without ambiguity.
+Instead, in case of an unambiguous parse, they present an error that includes the possible parses, to assist in debugging.
+
+We propose that the this idea---treating ambiguity as a parse-time error---can be beneficial even when the entire grammar is constructed by a single language designer, if done in a more rigorous fashion.
+Consider the following nested match expression in OCaml (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
+
+\begin{tabular}{l|l}
+\begin{lstlisting}[language={[objective]caml},numbers=left]
 match 1 with
@@
   | 2 -> "two"
+\end{lstlisting} &
+
+\begin{lstlisting}
+File "./nestmatch.ml", line 4, characters 4-5:
+Error: This pattern matches values of type int
+       but a pattern was expected which
+       matches values of type string
+\end{lstlisting} \\
+\end{tabular}
+
+\noindent The compiler sees the last line as belonging to the inner \ocaml{match} rather than the outer, as was intended. The solution is simple; we put parentheses around the inner match:
+
+\begin{tabular}{l}
+\begin{lstlisting}[language={[objective]caml},numbers=left]
+match 1 with
+  | 1 -> (match "one" with
+          | str -> str)
+  | 2 -> "two"
 \end{lstlisting}
-
-\noindent The OCaml compiler, when presented with this code, will give a type error for the last line:
-
-\begin{lstlisting}
-Error: This pattern matches values of type int
-       but a pattern was expected which matches
-       values of type string
-\end{lstlisting}
-
-\noindent The compiler sees the last line as belonging to the inner \ocaml{match} rather than the outer, as was intended. The fix is simple; put parentheses around the inner match:
-
+\end{tabular}
+
+\noindent However, the connection between the error message and the solution is not particularly clear; surrounding an expression with parentheses does not change its type.
+
+Instead, we look to the OCaml manual for inspiration.
+It contains an informal description of the syntax of the language\footnote{\url{https://caml.inria.fr/pub/docs/manual-ocaml/language.html}}, in the form of an EBNF-like grammar.
+Below is an excerpt of the productions for expressions, written in a more standard variant of EBNF:
+
+\setlength{\grammarindent}{9.5em}
+\begin{grammar}
+<expr> ::= 'match' <expr> 'with' <pattern-matching>
+
+<pattern-matching> ::= '|'? <pattern> ('when' <expr>)? \\
+('|' <pattern> ('when' <expr>)? '->' <expr>)*
+\end{grammar}
+
+\noindent If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match.
+The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended.
+Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user.
+Instead, we can present the alternatives in the form of modified codethat parses unambiguously as the corresponding alternative:
+
+\begin{center}
+\begin{tabular}{l|l}
 \begin{lstlisting}[language={[objective]caml}]
@@
   | 2 -> "two"
-\end{lstlisting}
-
-\noindent The connection between the error message and the fix is not a clear one however; adding parentheses around an expression does not change the type of anything.
-
-To come up with an alternative error to present in this case we look to the OCaml manual for inspiration.
-It contains an informal description of the syntax of the language\footnote{\url{https://caml.inria.fr/pub/docs/manual-ocaml/language.html}}, in the form of an EBNF-like grammar.
-Below is an excerpt of the productions for expressions, written in a more standard variant of EBNF:
-
-\setlength{\grammarindent}{5em}
-\begin{grammar}
-<expr> ::= 'match' <expr> 'with' <pattern-matching>
-
-<pattern-matching> ::= ('|' <pattern> '->' <expr>)+
-\end{grammar}
-
-Note that \synt{pattern-matching} is slightly simplified, the original grammar supports \ocaml{when} guards and makes the first \lit{|} optional.
-If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match.
-The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended.
-
-We instead argue that the grammar should be left ambiguous for this sort of corner cases that are likely to trip a user, allowing the compiler to present an ambiguity error, which lets the user select the intended alternative.
-
-\subsection{Unresolvable Ambiguity}
-
-Unfortunately, not all ambiguities can be resolved by adding parentheses. Again, looking to the informal OCaml grammar:
+\end{lstlisting} &
+\begin{lstlisting}[language={[objective]caml}]
+match 1 with
+  | 1 -> (match "one" with
+          | str -> str
+  | 2 -> "two")
+\end{lstlisting} \\
+\end{tabular}
+\end{center}
+
+\noindent Unfortunately, not all ambiguities can be resolved in this way. Again, looking at an extract from the informal OCaml grammar:
 
@@
 
-\noindent The first production is sequential composition, the second is lists (the empty list is under \synt{constant}). Now consider the following expression: ''\ocaml{[1; 2]}''.
-
+\noindent The first production is sequential composition, the second is lists (the empty list is under \synt{constant}).
+Now consider the expression ''\ocaml{[1; 2]}''.
 We find that it is ambiguous with two alternatives:
@@
 
-We can select the second option by putting parentheses around ''\ocaml{1; 2}'', but there is no way to select the first.
-If the user intended the first option we have a problem: we can present an accurate error message, but there is no way for an end-user to solve it; it requires changes to the grammar itself.
-
-To prevent the possibility of an end-user encountering such an error we must ensure that the grammar cannot give rise to an unresolvable ambiguity.
-It is worth mentioning here that statically checking if a context-free grammar is ambiguous has long been known to be undecidable \cite{cantorAmbiguityProblemBackus1962}.
-Unresolvable ambiguity, however, turns out to be decidable\footnote{With some caveats, I'll talk more about this during the meeting.}.
-
-Note that, as for ambiguity, the shape of the grammar is important, since the property considers parse trees rather than merely words.
-For this paper, we consider context-free grammars with EBNF operators.
-
-\subsection{Contributions}
-
-\begin{itemize}
-  \item Building on the work by \citet{palmkvistCreatingDomainSpecificLanguages2019}, which merely isolates ambiguities, an algorithm that suggests solutions to ambiguity errors.
-  \item A formalization of the unresolvable ambiguity property for context-free EBNF grammars.
-  \item An algorithm for deciding if a grammar is unresolvably ambiguous or not.
+\noindent We can write the latter option unambiguously by putting parentheses around ''\ocaml{1; 2}'', but there is no way to unambiguously write the former.
+OCaml itself solves this by unambiguously parsing the expression as a two-element list.
+In this case, the ''arbitrary'' choice seems reasonable; there is no other way to write a two-element list, ambiguously or not\footnote{Ignoring the cons operator, since this syntax is supposed to be syntactic sugar for it.}, hence we would rather like ''\ocaml{[1; 2]}'' to be interpreted as such.
+
+The possibility of an unresolvable ambiguity is troubling however, since it represents an error that a user might encounter but only a language designer can solve; it requires a change to the grammar.
+As such, we propose to allow ambiguous programming language grammars, as long as all possible ambiguities can be solved, and only require \emph{programs} to be unambigous.
+We call this property \emph{resolvable ambiguity}.
+
+We claim the following contributions:
+
+\begin{itemize}
+\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} (and \emph{dynamic}) problem, i.e., determine if a grammar (and word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
+\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
+  \begin{itemize}
+  \item A formulation of three versions of the static problem of increasing difficulty, and a decidable algorithm that is sound for the first two versions, and sound for the first (Section~\ref{sec:static}).
+  \item A decidable algorithm that solves the dynamic problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
+  \end{itemize}
+\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
+  \begin{itemize}
+  \item A large subset of OCaml where some ambiguities have been reintroduced.
+  \item A domain-specific language for cyber-physical systems.
+  \end{itemize}
 \end{itemize}
@@
 
-Next, we note that each lattice is uniquely determined by its top and bottom elements, it contains all elements between them:
+Next, we note that each lattice is uniquely determined by its top and bottom elements; it contains all elements between them:
 
@@
 \noindent Lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} guarantee the existence of two such words $w'_\top$ and $w'_\bot$.
+
+Finally, we show the difference between versions~1 and 2 in the lattice setting: in version~1, the rangeset of the bottom word is the empty set:
+
+\begin{lemma}
+  In version~1, $rangeset(w'_\bot) = \emptyset$ for all trees, where $w'_\bot$ is given by Lemma~\ref{lemma:required-parentheses}.
+\end{lemma}
+
+\noindent Since version~1 has no markings there are no required parentheses, thus the bottom word has no parentheses.
 
@@
 
-Our approach centers around around $\words(t)$, which, being a set of words, can be considered a language.
+Our approach centers around around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
 Each such language turns out to be a visibly pushdown language.
 Given a set of trees in $L(T_D)$ we can construct a corresponding set of non-overlapping visibly pushdown automata (i.e., each automaton only accepts words not accepted by any other automaton) since VPLs are closed under complement and intersection \cite{alurVisiblyPushdownLanguages2004}.
-These automata can then be examined to determine if they accept the empty language (which implies that the corresponding tree only has ambiguous words), or otherwise producing a witness.
+These automata can then be examined to determine if they accept the empty language (which implies that the corresponding tree only has ambiguous words), or otherwise produce a witness, a word accepted only by that automaton.
 
@@
 
+\subsection{Cyphym} \label{sec:evaluation-cyphym}
+
 \section{Related Work}
@@
 However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
-
-% TODO: this section should be moved to be in the appropriate position {
-\newpage
-
-\begin{figure*}[ht]
-  \begin{tikzcd}
-    L(G_w) & L(G_t) \arrow[l, "\yield"] \\
-    L(G'_w) & L(G'_t) \arrow[u, "\semantic"'] \arrow[l, "\yield"]
-  \end{tikzcd}
-  \caption{The generated grammars and their relation to each other.}
-\end{figure*}
-
-\begin{figure*}[ht]
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{align*}
-      1 + 2 + 3
-    \end{align*}
-    \caption{Example word in $L(G_w)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.45\linewidth}
-      \begin{align*}
-        [_S           [_S  &[_N 1 ]_N + \hphantom{[_S} [_N 2 ]_N           ]_S  + [_N 3 ]_N \hphantom{]_S} ]_S \\
-        [_S \hphantom{[_S} &[_N 1 ]_N +           [_S  [_N 2 ]_N \hphantom{]_S} + [_N 3 ]_N           ]_S  ]_S
-      \end{align*}
-    \caption{Example words in $L(G_T)$.}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{align*}
-      1 + 2 + 3
-    \end{align*}
-    \caption{Example word in $L(G'_w)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.45\linewidth}
-      \begin{align*}
-        [_S            [_S &[_N 1 ]_N + \hphantom{[_S} [_N 2 ]_N            ]_S + [_N 3 ]_N \hphantom{]_S}]_S \\
-        [_S \hphantom{[_S} &[_N 1 ]_N +            [_S [_N 2 ]_N \hphantom{]_S} + [_N 3 ]_N ]_S ]_S
-      \end{align*}
-    \caption{Example words in $L(G'_T)$.}
-  \end{subfigure}
-  \caption{Example with an ambiguous word in $L(G'_w)$ with corresponding words from the other grammars.}
-  \label{fig:ambig-example-square}
-\end{figure*}
-
-\clearpage
-
-% }
-
-\section{Parsetime Ambiguity Reporting} \label{sec:parse-time-reporting}
 

### 2019-06-20 paper/paper.tex
EOD

@@
 
+% I really don't like this paragraph. I'm trying to say that much work has been done on ambiguity, and that work in other fields include it, while PL tends to be more narrow-minded.
 Formal languages have long had the concept of ambiguity, of words that can be constructed in multiple ways, for example using multiple distinct derivations of a context-free grammar.
+In linguistics, this is important, since human languages tend to be highly ambiguous, and as such much work has been done on the subject.
+Examples include general parsing algorithms \cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}, ambiguity detection (most importantly its undecidability \cite{cantorAmbiguityProblemBackus1962}, but also various pragmatic approaches \cite{brabrandAnalyzingAmbiguityContextFree2007,axelssonAnalyzingContextFreeGrammars2008,bastenAmbiguityDetectionProgramming2011})
 
@@
 This view, that eliminating ambiguity is \emph{the} correct thing to do, is highly prevalent (e.g.
-\cite{sudkampLanguagesMachinesIntroduction1997}).
+\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).
 
@@
 
-We propose that the this idea---treating ambiguity as a parse-time error---can be beneficial even when the entire grammar is constructed by a single language designer, if done in a more rigorous fashion.
+We propose that this idea---treating ambiguity as a parse-time error---can be beneficial even when the entire grammar is constructed by a single language designer, if done in a more rigorous fashion.
 Consider the following nested match expression in OCaml (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
@@
 
-\noindent However, the connection between the error message and the solution is not particularly clear; surrounding an expression with parentheses does not change its type.
+\noindent However, the connection between error message and solution is not particularly clear; surrounding an expression with parentheses does not change its type.
 
@@
 Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user.
-Instead, we can present the alternatives in the form of modified codethat parses unambiguously as the corresponding alternative:
+Instead, we can present the alternatives in the form of modified code that parses unambiguously as the corresponding alternative:
 
@@
 The possibility of an unresolvable ambiguity is troubling however, since it represents an error that a user might encounter but only a language designer can solve; it requires a change to the grammar.
-As such, we propose to allow ambiguous programming language grammars, as long as all possible ambiguities can be solved, and only require \emph{programs} to be unambigous.
+As such, we propose to allow ambiguous programming language grammars, as long as all possible ambiguities can be resolved, and only require \emph{programs} to be unambigous.
 We call this property \emph{resolvable ambiguity}.
@@
 \begin{itemize}
-\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} (and \emph{dynamic}) problem, i.e., determine if a grammar (and word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
+\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp.
+word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
 \item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
   \begin{itemize}
-  \item A formulation of three versions of the static problem of increasing difficulty, and a decidable algorithm that is sound for the first two versions, and sound for the first (Section~\ref{sec:static}).
-  \item A decidable algorithm that solves the dynamic problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
+  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound for the first two versions, and complete for the first (Section~\ref{sec:static}).
+  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
   \end{itemize}
@@
   \begin{itemize}
-  \item A large subset of OCaml where some ambiguities have been reintroduced.
-  \item A domain-specific language for cyber-physical systems.
+  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
+  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
   \end{itemize}

### 2019-06-23 paper/paper.tex
EOD

@@
 \newcommand{\regex}{\mathit{Reg}}
+\newcommand{\reqp}[1]{(#1)}
+\newcommand{\posp}[1]{[#1]}
 
@@
 \noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w' \in L(G'_D)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists.
-Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exist.
+Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.
 
@@
 If we thus delimit ourselves to only consider languages where words have no unbalanced parentheses\footnote{I.e., the vast, vast majority of programming languages currently in use.} we can give the following alternative definition of a word: a word is a two-tuple containing a sequence of non-parenthesis terminals and a bag (or multiset) of ranges covered by parentheses.
-For example, the word ''$(1 + 2) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{1}{3}\})$, while ''$((1 + 2)) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{1}{3}, \range{1}{3}\})$.
-We will refer to the first component of the tuple as the \emph{basic word}, the second as the \emph{range bag}.
+For example, the word ''$(1 + 2) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{0}{3}\})$, while ''$((1 + 2)) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
+We will refer to the first component of the tuple as the \emph{basic word} and the second as the \emph{range bag}.
+
+However, there is a case where this alternate representation is not fully isomorphic; namely when a pair of parentheses would yield a zero-length range.
+As an example, consider the two words ''$(())$'' and ''$()()$''.
+Both of these have an alternate representation of $(\text{''}\text{''}, \{\range{0}{0}, \range{0}{0}\})$.
+In versions~1 and 2 this corresponds to grouping parentheses surrounding zero-length productions.
+We can avoid this case by finding every production $N -> l : r$ where $\epsilon \in L(r)$, replacing it with $N -> l : r'$ where $L(r') = L(r) \setminus \{\epsilon\}$, and replacing every use of $N$ with $N + \epsilon$.
+The one exception is if the start symbol has a nullable production, but that we can trivially special case.
+As such, going forward, we will assume no zero-length grouping parentheses.
 
@@
   Given $t \in L(T_D)$, $w'_\top, w'_\bot \in \words(t)$ such that $\forall w'.\ w' \in \words(t) => \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top)$ the following holds:\\
-  $\forall w'.\ \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top) => w' \in \words(t)$.
+  $\forall w'.\ \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top) => w' \in \words(t)$. \label{lemma:top-bottom-determine}
 \end{lemma}
@@
 \begin{lemma}
-  In version~1, $rangeset(w'_\bot) = \emptyset$ for all trees, where $w'_\bot$ is given by Lemma~\ref{lemma:required-parentheses}.
+  In version~1, $rangeset(w'_\bot) = \emptyset$ for all trees, where $w'_\bot$ is given by Lemma~\ref{lemma:required-parentheses}. \label{lemma:version1-bot}
 \end{lemma}
@@
 
-We now have the tools we require to motivate our approach.
+Our approach centers around finding a pair of lattices, such that one is entirely contained in the other.
+Since the former shares \emph{all} partitions with the latter, it has only ambiguous words, thus any of those words will be unresolvably ambiguous.
 
 \begin{theorem}
-  In version 1, for a given $t \in L(T_D)$, $\exists w'.\ \parse(w') = \{t\}$ iff $\parse(w'_\top) = \{t\}$ (where $w'_\top$ is given by Lemma~\ref{lemma:possible-parentheses}).
+  In versions~1 and 2, given a $t \in L(T_D)$ with a lattice determined by $w_\bot$ and $w_\top$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w_\bot) \subseteq \rangeset(w'_\bot) \land \rangeset(w'_\top) \subseteq \rangeset(w_\top) => \neg \exists w.\ \parse(w) = \{t\}$.
+\label{lemma:lattice-containment-implies-unresolvable}
 \end{theorem}
 
+\noindent For version~1, \emph{every} unresolvable ambiguity has this form:
+
+\begin{theorem}
+  In version 1, given a $t \in L(T_D)$ with a lattice determined by $w_\bot$ and $w_\top$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w'_\bot) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(w'_\top) <=> \neg \exists w.\ \parse(w) = \{t\}$.
+\end{theorem}
+
+\begin{proof}
+  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(w'_\bot)$ implies that $w_\bot = w'_\bot$.
+We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w'_\top) \subseteq \rangeset(w_\top)$.
+
+  If all words in $\words(t)$ are ambiguous, then $\exists t' \in L(T_D).\ t' \neq t \land t' \in \parse(w_\top)$.
+Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(w'_\bot)$.
+The definition of $w'_\top$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(w'_\top)$, but since the rangebag of $w'_\top$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(w'_\top)$.
+\end{proof}
+
 \subsection{A Lattice as a Word, and an Algorithm} \label{sec:lattice-vpl}
 
-Our initial limitations / assumptions are as follows:
-
-\begin{enumerate}
-  \item The input language definition contains no parentheses.
-This implies that all parentheses in $G'_w$ are grouping parentheses, thus we only need to consider words $w' \in L(G'_w)$ whose range bag is a set.
-
-  \item No non-terminals in the input language definition are marked, i.e., there are no forbidden children. This implies that there are no required grouping parentheses.
-
-  \item No production has a right-hand side that matches a single non-terminal, i.e., $\forall (N->i:r) \in D.\ \NT \cap L(r) = \emptyset$\footnote{I haven't really written a proper description for the language definition, but I'm here using $D$ for it, and considering it essentially as a set of ''productions''.}.
-\end{enumerate}
-
-\noindent The first two limitations allow us to place all words in $\words(t)$ for some given $t$ in a lattice, whose structure is formed by the subset ordering of the range sets.
-For example, the two trees in Figure~\ref{fig:ambig-example-square} have the lattices of words seen in Figures~\ref{fig:lattice1} and \ref{fig:lattice2} respectively.
-
-The bottom word corresponds to the set of required parentheses (i.e., the empty set, in this sub-problem), while the top word corresponds to the set of possible parentheses ranges.
-All trees that can be ambiguous must thus share the same bottom word, while the top may differ.
-
-Of particular use is to examine whether the top word (call it $w'$) is ambiguous. There are two cases:
-
-\begin{itemize}
-\item The top word is unambiguous, then this is not a tree we are looking for (since it has at least one unambiguous word).
-\item It is ambiguous.
-That means that there is some other lattice for some other tree that also contains the word.
-Call the top word of this other lattice $w'_2$.
-For $w'$ to be in the lattice whose top is $w'_2$ its rangeset must be a subset of the rangeset of the latter.
-But this is true also for all the other words in the lattice, thus there are no unambiguous words in the lattice, i.e., this tree is what we are looking for.
-\end{itemize}
-
-\noindent The algorithm thus looks for two trees, where the set of possible parentheses for one is a superset of the other.
-To do this, we use a linear representation of each lattice: namely the top word.
-If two trees have the same top word, that implies that their sets of possible parentheses are equal.
-If we can add parentheses to one word and get the other word, that implies that the rangeset of the former is a subset of the latter.
-
-We will now construct a pushdown automaton that recognizes these top words in such a way that there is a bijection between successful runs and trees in $L(G_t)$.
-As a running example, we will use the following (very simple) language definition $D$:
+This section introduces a linear encoding of the lattices of the previous chapter as words.
+Lemmas~\ref{lemma:same-basic} and \ref{lemma:top-bottom-determine} imply that a lattice encoding requires three things: a basic word, a set of \emph{required} parentheses, and a set of \emph{possible} parentheses.
+We encode required parentheses with ''$\reqp{}$'' and possible parentheses with ''$\posp{}$''.
+For example, the lattice in Figure~\ref{fig:lattice2} is represented by ''$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ''$(1 + 2) * 3$'' has a lattice represented by ''$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
+With this encoding we can apply the formidable body of knowledge available for word languages, albeit with some extra care; the linear encoding of a lattice is not unique.
+
+For example, ''$\posp{\posp{1}}$'' and ''$\posp{1}$'' represent the same lattice, as do ''$\reqp{\posp{2}}$'' and ''$\reqp{2}$''.
+To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over optional parentheses, e.g., ''$\posp{\posp{1}}$'' is uniquely represented as ''$\posp{1}$'' while ''$\reqp{\posp{2}}$'' is uniquely represented as ''$\reqp{2}$''.
+
+Lattice equality is now equvialent with equality of the linear encoding.
+To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ''up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ''down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
+For example, ''$\reqp{1}2$'' is entirely contained in ''$\reqp{1}\posp{2}$'', which is entirely contained in ''$\posp{1}\posp{2}$''.
+
+The centerpiece of our algorithm is a visibly pushdown automaton constructed in such a way that:
+
+\begin{itemize}
+\item There is a bijection between trees in $L(T_D)$ and successful runs.
+\item The word recognized by a successful run is the linear encoding of the corresponding tree's lattice.
+\end{itemize}
+
+\noindent Two distinct successful runs through this automaton that recognize the same word thus imply two distinct trees that have exactly the same lattice.
+To detect lattices contained in each other, we create a modified copy, where the copy may add arbitrary (balanced, well-nested) possible parentheses, and replace any required pair of parentheses with a possible pair.
+The copy maintains much of the structure of the original, and in particular, keeps the connection to trees in $L(T_D)$ (though it is no longer a bijection, there are multiple successful runs per tree since there are multiple larger lattices).
+
+We will now walk through the construction of this automaton, using the following language definition:
 
@@
 \begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-  \synt{N} & Succ & 's' \synt{N}\\
-  \synt{N} & Zero & 'z' \\
+  \toprule
+  $E$ & $l$ & \verb|'['| ($E$ \verb|';'|)* \verb|']'| \\
+  $E$ & $s$ & $E_{\{s\}}$ \verb|';'| $E$ \\
+  $E$ & $c$ & $C$ \\
+  \midrule
+  $C$ & $v$ & \verb|'e'| \\
+  \bottomrule
 \end{tabular}
@@
   \node[state] (2) [right of=1] {$2$};
-  \node[state,accepting] (3) [right of=2] {$3$};
-  \node[state,initial] (4) [below of=1, xshift=1cm] {$4$};
-  \node[state,accepting] (5) [right of=4] {$5$};
-
-  \path (1) edge              node {$s$} (2)
-  (2) edge              node {$N$} (3)
-  (4) edge              node {$z$} (5);
+  \node[state] (3) [below of=2, yshift=5mm] {$3$};
+  \node[state,accepting] (4) [right of=2] {$4$};
+
+  \node[state,initial] (5) [below of=1, yshift=-10mm] {$5$};
+  \node[state] (6) [right of=5] {$6$};
+  \node[state] (7) [right of=6] {$7$};
+  \node[state,accepting] (8) [right of=7] {$8$};
+
+  \path (1) edge node {$'['$} (2)
+  (2) edge[bend right,left] node {$E$} (3)
+  (3) edge[right] node {$';'$} (2)
+  (2) edge node {$']'$} (4);
+  \path (5) edge node {$E_{\{s\}}$} (6)
+  (6) edge node {$';'$} (7)
+  (7) edge node {$E$} (8);
 

### 2019-06-27 paper/paper.tex
EOD

@@
 \newcommand{\regex}{\mathit{Reg}}
-\newcommand{\reqp}[1]{(#1)}
-\newcommand{\posp}[1]{[#1]}
+\newcommand{\reqpl}{(}
+\newcommand{\reqpr}{)}
+\newcommand{\reqp}[1]{\reqpl#1\reqpr}
+\newcommand{\pospl}{[}
+\newcommand{\pospr}{]}
+\newcommand{\posp}[1]{\pospl#1\pospr}
 
@@
 
+Outline of introduction, one bullet per paragraph, roughly:
+
+\begin{itemize}
+\item Creating languages, particularly DSLs, it's nice to create pieces and compose.
+Composing syntax in particular quickly creates ambiguous grammars.
+Two broad approaches: forbid/prevent ambiguities at all (statically), or handle at parse-time (dynamically).
+Former more well-explored than the latter, which we focus on.
+The former is essentially the ambiguity problem, but the latter has not been formally defined.
+(To my knowledge we're first in presenting resolutions as opposed to interpretations (ASTs), which means that what we're talking about here is a little bit different from what previous approaches have done)
+
+\item Motivated by issues in \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011} ''we introduce a formal definition for resolvable ambiguity.'' (They both merely present ASTs, which are unlikely to be nice for an end-user.
+Probably want better motivation than ''nice.'') \ldots It turns out that this is not always possible, prompting an additional problem: can we statically determine if a language is resolvably ambiguous?
+We call the former the dynamic resolvability problem, and the latter the static resolvability problem.
+
+\item The dynamic problem is ''easy'' (though we haven't found any previously presented solution), but the static problem is not.
+We present a subclass of languages for which we have a decidable, sound, and complete decision procedure for the static problem, but also two larger classes of languages for which the algorithm is merely decidable and sound, or not applicable at all, respectively.
+\end{itemize}
+
 % I really don't like this paragraph. I'm trying to say that much work has been done on ambiguity, and that work in other fields include it, while PL tends to be more narrow-minded.
@@
       Labels & $l \in \Labels$ \\
+      \multicolumn{2}{@{}l@{}}{$\T$, $\NT$, and $\Labels$ disjoint} \\
+      \addlinespace
       Marks & $m \subseteq \Labels$ \\
@@
   \toprule
-  $E$ & $l$ & \verb|'['| ($E$ \verb|';'|)* \verb|']'| \\
   $E$ & $s$ & $E_{\{s\}}$ \verb|';'| $E$ \\
-  $E$ & $c$ & $C$ \\
-  \midrule
-  $C$ & $v$ & \verb|'e'| \\
+  $E$ & $v$ & \verb|'e'| \\
   \bottomrule
@@
 
-\noindent We begin by constructing a DFA per production. This can be done in the standard way by constructing an NFA, then determinizing it, and optionally minimizing it.
+\noindent We first present a simplified construction that assumes no production can match a single non-terminal, i.e., for all right-hand side regular expressions $r$, $\neg \exists N_m.\ N_m \in L(r)$, to make the base idea easier to follow.
+Such cases introduce duplicated parentheses if handled in a naive way, and the extra book-keeping required to correctly handle them complicate this presentation, and will so be introduce after the more naive method.
+
+Conceptually, each non-terminal represents a choice of which child production should replace it, and whether that production requires parentheses around it.
+We now make this explicit, to ensure that standard operations on finite automata produce the correct result.
+We replace each $N_m$ with a sum of the labels of productions in $N$, where each label in $m$ is distinguished by a prime, e.g., $E_{\{s\}}$ is replaced with $s' + v$.
+
+\begin{center}
+\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+  \toprule
+  $E$ & $s$ & $(s' + v)$ \verb|';'| $(s + v)$ \\
+  $E$ & $v$ & \verb|'e'| \\
+  \bottomrule
+\end{tabular}
+\end{center}
+
+\noindent Next we construct a DFA per production. This can be done in the standard way by constructing an NFA, then determinizing it, and optionally minimizing it.
 
@@
   \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [below of=2, yshift=5mm] {$3$};
-  \node[state,accepting] (4) [right of=2] {$4$};
-
-  \node[state,initial] (5) [below of=1, yshift=-10mm] {$5$};
-  \node[state] (6) [right of=5] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-  \node[state,accepting] (8) [right of=7] {$8$};
-
-  \path (1) edge node {$'['$} (2)
-  (2) edge[bend right,left] node {$E$} (3)
-  (3) edge[right] node {$';'$} (2)
-  (2) edge node {$']'$} (4);
-  \path (5) edge node {$E_{\{s\}}$} (6)
-  (6) edge node {$';'$} (7)
-  (7) edge node {$E$} (8);
+  \node[state] (3) [right of=2] {$3$};
+  \node[state,accepting] (4) [right of=3] {$4$};
+
+  \node[state,initial] (5) [below of=2] {$5$};
+  \node[state,accepting] (6) [right of=5] {$6$};
+
+  \path (1) edge[bend left] node {$s'$} (2)
+  (1) edge[bend right] node {$v$} (2)
+  (2) edge node {$';'$} (3)
+  (3) edge[bend left] node {$s$} (4)
+  (3) edge[bend right] node {$v$} (4);
+  \path (5) edge node {$'e'$} (6);
 
@@
 
-\noindent We then combine them into a single pushdown automata by replacing each edge with a non-terminal label $p \xrightarrow{N} q$ with:
-
-\begin{itemize}
-  \item an edge $p \xrightarrow{'(', +(p, q)} p'$ for every initial state $p'$ in some DFA belonging to non-terminal $N$, and
-  \item an edge $q' \xrightarrow{')', -(p, q)} q$ for every final state $q'$ in some DFA belonging to non-terminal $N$,
-\end{itemize}
-
-\noindent Intuitively, we parse a child node, but put parentheses around it, then return. This is where we use assumption 3, without it we might introduce double parentheses here.
+\noindent We then combine them into a single visibly pushdown automaton.
+Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we go back.
+We use $Q \times Q \times \{\text{''}\reqp{}\text{''}, \text{''}\posp{}\text{''}\}$ as our stack alphabet, where the first two components are states in the parent production automaton and the last component records whether the parenthesis pair is required or merely possible.
+We replace every edge $p \xrightarrow{l} q$, where $l \in \Labels$, with:
+
+\begin{itemize}
+  \item an edge $p \xrightarrow{'\pospl', +(p, q, \text{''}\posp{}\text{''})} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
+  \item an edge $q' \xrightarrow{'\pospr', -(p, q, \text{''}\posp{}\text{''})} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
+\end{itemize}
+
+\noindent Similarly, for every edge $p \xrightarrow{l'} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\reqpl', +(p, q, \text{''}\reqp{}\text{''})} p'$ and $q' \xrightarrow{'\reqpr', -(p, q, \text{''}\reqp{}\text{''})} q$.
+
+Intuitively, we parse a child node with parentheses around it, then return.
+This is where our simplification is used, without it we might introduce double parentheses here.
+To lessen clutter in the figure below we omit the third component of the stack symbol, since it is fully determined by the input symbol, e.g., if an edge reads $'\pospl'$ then the third component of the stack symbol pushed must be ''$\posp{}$''.
 
@@
   \node[state] (2) [right of=1] {$2$};
-  \node[state,accepting] (3) [right of=2] {$3$};
-  \node[state,initial] (4) [below of=1, xshift=1cm] {$4$};
-  \node[state,accepting] (5) [right of=4] {$5$};
-
-  \draw
-  (1) edge[above] node{$s$} (2)
-  (4) edge[above] node{$z$} (5)
-  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
-  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
-  (3) edge[loop above] node{$')', -(2, 3)$} (3)
-  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
+  \node[state] (3) [right of=2] {$3$};
+  \node[state,accepting] (4) [right of=3] {$4$};
+
+  \node[state,initial] (5) [below of=2] {$5$};
+  \node[state,accepting] (6) [right of=5] {$6$};
+
+  \path (2) edge[below] node {$';'$} (3);
+  \path (5) edge[below] node {$'e'$} (6);
+
+  \path (1) edge[loop above] node{$'\reqpl', +(1, 2)$} (1)
+  (4) edge[bend right, above] node{$'\reqpr', -(1, 2)$} (2);
+  \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
+  (6) edge[left, in=270, out=90] node[xshift=-9.8mm,yshift=3mm]{$'\pospr', -(1, 2)$} (2);
+  \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
+  (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
+  \path (3) edge[in=90, out=270, right] node[xshift=9mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
+  (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
 
@@
 
+  \node[state,initial] (7) [left of=5, xshift=-1.5cm] {$7$};
+  \node[state,accepting] (8) [right of=6, xshift=1.5cm] {$8$};
+
   \node[state] (1) {$1$};
@@
   \node[state] (3) [right of=2] {$3$};
-  \node[state] (4) [below of=1, xshift=1cm] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-  \node[state,initial] (s) [left of=4] {};
-  \node[state,accepting] (e) [right of=5] {};
-
-  \draw
-  (s) edge[left] node{$'(', +s$} (1)
-  (s) edge[above] node{$'(', +s$} (4)
-  (3) edge[right] node{$')', -s$} (e)
-  (5) edge[above] node{$')', -s$} (e)
-  (1) edge[above] node{$s$} (2)
-  (4) edge[above] node{$z$} (5)
-  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
-  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
-  (3) edge[loop above] node{$')', -(2, 3)$} (3)
-  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
+  \node[state] (4) [right of=3] {$4$};
+
+  \node[state] (5) [below of=2] {$5$};
+  \node[state] (6) [right of=5] {$6$};
+
+  \path (2) edge[below] node {$';'$} (3);
+  \path (5) edge[below] node {$'e'$} (6);
+
+  \path (1) edge[loop above] node{$'\reqpl', +(1, 2)$} (1)
+  (4) edge[bend right, above] node{$'\reqpr', -(1, 2)$} (2);
+  \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
+  (6) edge[left, in=270, out=90] node[xshift=-9.8mm,yshift=3mm]{$'\pospr', -(1, 2)$} (2);
+  \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
+  (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
+  \path (3) edge[in=90, out=270, right] node[xshift=9mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
+  (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
+
+  \path (7) edge[bend left, left] node{$'\pospl', +(7, 8)$} (1)
+  (7) edge[below] node{$'\pospl', +(7, 8)$} (5);
+  \path (4) edge[bend left, right] node{$'\pospr', -(7, 8)$} (8)
+  (6) edge[below] node{$'\pospr', -(7, 8)$} (8);
 
@@
 
-\noindent The resulting pushdown automaton has only a single source of non-determinism: the edges labelled \verb|'('|.
-Each one corresponds to one of the allowable child productions at that point in the parse tree.
+\noindent The resulting pushdown automaton has only two sources of non-determinism: the edges labelled $'\pospl'$ and $'\reqpl'$.
+Each one corresponds to one of the allowable child productions at that point in the tree, thus there is one run per tree.
+STOP HERE, THE TEXT BEYOND IS NOT UPDATED.
+
+%% TODO: Continue here. Also figure out how to handle (E + E_m).
 

### 2019-06-27 syncon-parser/src/Main.hs
New timeout and continue-after-error options

@@ -11,8 +11,10 @@ import ErrorMessage (FormatError, formatErrors, formatError, ErrorOpts)
 
 import System.Environment (withArgs)
 import System.FilePath ((</>))
+import System.Timeout (timeout)
 
 import Data.Data (Data)
+import Data.IORef (newIORef, modifyIORef', readIORef)
 import Data.List (partition)
 import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
@@ -59,11 +61,10 @@ dataOrError _ _ (Data a) = return a
 dataOrError source opts (Error e) =
   die $ formatErrors source $ formatError opts <$> e
 
-dataOrError' :: Show e => Result e a -> IO a
-dataOrError' (Data a) = return a
-dataOrError' (Error e) = do
-  pPrint e
-  compErr "Main.dataOrError" "Got error"
+dataOrError' :: (Functor f, Foldable f, FormatError e) => HashMap Text Text -> ErrorOpts e -> Result (f e) a -> IO a
+dataOrError' _ _ (Data a) = return a
+dataOrError' source opts (Error e) =
+  die' $ formatErrors source $ formatError opts <$> e
 
 testReduce :: IO ()
 testReduce = do
@@ -93,6 +94,13 @@ test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--two-level"] 
 getArgsSeq :: IO (Seq [Char])
 getArgsSeq = getArgs <&> Seq.fromList
 
+newtype SourceFileException = SourceFileException Text deriving (Show)
+instance Exception SourceFileException
+
+die' :: Text -> IO a
+die' t = do
+  throwIO $ SourceFileException t
+
 common :: Opt.Parser (IO ())
 common = do
   html <- optional $ Opt.strOption
@@ -110,6 +118,14 @@ common = do
   showTwoLevel <- Opt.switch
     $ Opt.long "two-level"
     <> Opt.help "Always show the two level representation, even if some alternatives are resolvable."
+  sourceTimeout <- fmap (*1_000_000) $ Opt.option Opt.auto
+    $ Opt.long "timeout"
+    <> Opt.metavar "S"
+    <> Opt.help "Timeout for attempting to parse a single source file, in seconds. A negative value means 'wait forever'."
+    <> Opt.value (-1)
+  continueAfterError <- Opt.switch
+    $ Opt.long "continue-after-error"
+    <> Opt.help "Don't abort after the first source file that gives errors."
   files <- some $ Opt.argument Opt.str $
[…]

### 2019-06-28 paper/paper.tex
EOD

@@
 
+\noindent\hrulefill
+
+When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments.
+However, how to make such compositions well-behaved is not always obvious.
+In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous.
+Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous words.
+i.e., ambiguous programs.
+
+\noindent\hrulefill
+
 % I really don't like this paragraph. I'm trying to say that much work has been done on ambiguity, and that work in other fields include it, while PL tends to be more narrow-minded.
@@
 \begin{definition}
-  A decision procedure $f$ solving the static resolvability problem is:
+  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unresolvable}$) solving the static resolvability problem is:
   \begin{itemize}
   \item Sound if $f(\parse) = \mathit{resolvable}$ implies that the language given by $\parse$ is resolvably ambiguous, and $f(\parse) = \mathit{unresolvable}$ implies that the language given by $\parse$ is not resolvably ambiguous.
-  \item Complete if $f$ terminates and produces a result for all possible inputs $\parse$.
+  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$.
   \end{itemize}
@@
 \begin{definition}
-  A decision procedure $f$ solving the dynamic resolvability problem is:
+  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unresolvable}$) solving the dynamic resolvability problem is:
   \begin{itemize}
   \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
-  \item Complete if $f$ terminates and produces a result for all possible inputs $\parse$ and $w$.
+  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
   \end{itemize}
@@
 Also note that the right-hand side of a production is here a regular expression, rather than the theoretically simpler sequence used in a context-free grammar.
+For proof technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense.
+This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}.
+In practice, this is likely desireable regardless, since a user is likely to attach different semantic meaning to each occurrence of a non-terminal.
+
 Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal.
@@
 Each element corresponds to the word ''$(1 + 2) * 3$'' in $L(G'_D)$.
-Note that the word in $L(G_D)$ as ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ''$((1 + 2)) * 3$'' and ''$(1 + 2) * (3)$''.
+Note that the word in $L(G_D)$ is ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ''$((1 + 2)) * 3$'' and ''$(1 + 2) * (3)$''.
 As a memory aid, the prime versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations from $D$.
@@
   \end{tikzcd}
-  \caption{The grammars considered, and their relation to each other.}
+  \caption{The generated grammars, and their relation to each other.}
   \label{fig:grammar-square}
@@
 Versions~1 and 2 have only grouping parentheses, meaning that no double parentheses matter.
-What follows is a brief outline of the remainder of this section, which details our static analysis for version 1 and 2, building from that observation.
+What follows is a brief outline of the remainder of this section, which details our static analysis for versions~1 and 2, building from that observation.
 
@@
 We can avoid this case by finding every production $N -> l : r$ where $\epsilon \in L(r)$, replacing it with $N -> l : r'$ where $L(r') = L(r) \setminus \{\epsilon\}$, and replacing every use of $N$ with $N + \epsilon$.
-The one exception is if the start symbol has a nullable production, but that we can trivially special case.
+The one exception is if the start symbol has a nullable production, but that can be trivially special cased.
 As such, going forward, we will assume no zero-length grouping parentheses.
 
-We denote the function producing the alternative representation by $\alt$, while $\basic$ and $\rangebag$ directly produce the basic word and rangebag, respectively.
+We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., $\basic(\text{''}((1 + 2)) * 3\text{''}) = \text{''}1 + 2 * 3\text{''}$ and $\rangebag(\text{''}((1 + 2)) * 3\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
 
@@
 For example, ''$\posp{\posp{1}}$'' and ''$\posp{1}$'' represent the same lattice, as do ''$\reqp{\posp{2}}$'' and ''$\reqp{2}$''.
-To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over optional parentheses, e.g., ''$\posp{\posp{1}}$'' is uniquely represented as ''$\posp{1}$'' while ''$\reqp{\posp{2}}$'' is uniquely represented as ''$\reqp{2}$''.
+To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses, e.g., ''$\posp{\posp{1}}$'' is uniquely represented as ''$\posp{1}$'' while ''$\reqp{\posp{2}}$'' is uniquely represented as ''$\reqp{2}$''.
 
@@
 The copy maintains much of the structure of the original, and in particular, keeps the connection to trees in $L(T_D)$ (though it is no longer a bijection, there are multiple successful runs per tree since there are multiple larger lattices).
+Two distinct successful runs, one in each automaton, that recognize the same word then imply two distinct trees where the lattice of one is entirely contained in the other.
 
@@
 \noindent We first present a simplified construction that assumes no production can match a single non-terminal, i.e., for all right-hand side regular expressions $r$, $\neg \exists N_m.\ N_m \in L(r)$, to make the base idea easier to follow.
-Such cases introduce duplicated parentheses if handled in a naive way, and the extra book-keeping required to correctly handle them complicate this presentation, and will so be introduce after the more naive method.
+Such cases introduce duplicated parentheses if handled in a naive way, and the extra book-keeping required to correctly handle them complicate this presentation, and will so be introduced after the more naive method.
 
@@
 
-% TODO: talk about the ''low precedence unary operator'' issue.
+% TODO: talk about the ''low precedence unary operator/deep precedence'' issue.
 
@@
 However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
-
-\section{Properties We May Want to Prove}
-
-\subsection{Static Stuff}
-
-\begin{lemma}
-There is an injective function ($\alt$) from well-balanced words to tuples of basic words and range bags.
-\end{lemma}
-
-\begin{lemma}
-$\forall t \in L(G_t), w_1, w_2 \in L(G'_w).\ \{w_1, w_2\} \subseteq \words(t) -> \basic(w_1) = \basic(w_2)$
-\label{lemma:tree-single-basic}
-\end{lemma}
-
-\begin{lemma}
-$\forall t \in L(G_t).\ \words(t) \neq \emptyset$
-\label{lemma:words-nonempty}
-\end{lemma}
-
-\begin{lemma}
-$\forall t_1, t_2 \in L(G_t).\ \words(t_1) \cap \words(t_2) \neq \emptyset -> \basic(t_1) = \basic(t_2)$ \\
-(We're using \ref{lemma:tree-single-basic} and \ref{lemma:words-nonempty} to (ab)use ''$\basic$'' on trees as well)
-\end{lemma}
-
-\begin{definition}
-A language definition is in version 1 or 2 if $\T \cap \mathit{par} = \emptyset$.
-\end{definition}
-
-\begin{theorem}
-In version 1 and 2: $\forall w_1, w_2 \in L(G'_w).\ \altset(w_1) = \altset(w_2) -> \parse(w_1) = \parse(w_2)$, where $\altset(w)$ is the same as $\alt(w)$, but with the second component (the rangebag) replaced with a corresponding set.
-\end{theorem}
-
-\begin{lemma}
-In version 1 and 2, given a $t \in L(G_t)$, the words in $\words(t)$ can be partitioned into a lattice.
-Each partition contains the words that are equal by $\altset$, and ordering is by subset on the rangeset.
-Furthermore, this lattice is bounded, and uniquely described by its bottom and top elements; it contains all elements between bottom and top.
-We denote this lattice as $\lattice(t)$.
-\label{lemma:bounded-lattice}
-\end{lemma}
-
-\begin{definition}
-A language definition is in version 1 if it contains no marked non-terminals and $\T \cap \mathit{par} = \emptyset$.
-\end{definition}
-
-\begin{lemma}
-In version 1: the bottom element of the lattice for a tree has a rangeset that is the empty set.
-\end{lemma}
-
-\begin{theorem}
-Given a tree $t \in L(G_t)$ in version 1, $\exists w \in L(G'_w).\ \parse(w) = \{t\} <-> \neg \exists t' \in L(G_t).\ t \neq t' \land \lattice(t) \subseteq \lattice(t')$, i.e., a tree is resolvable iff there is no other tree that entirely covers its lattice.
-\end{theorem}
-
-\begin{lemma}
-(Version 1 and 2) We can construct a VPDA that recognizes a linear encoding of the lattices of trees in $L(G_t)$, such that there is a bijection between successful runs and trees.
-\label{lemma:linear-lattices}
-\end{lemma}
-
-\begin{lemma}
-(Version 1 and 2) Using \ref{lemma:linear-lattices}, we can construct a VPDA that reconizes a linear encoding of ''superlattices'' of trees in $L(G_t)$ (i.e., lattices that completely cover the original lattices), where each successful run corresponds to exactly one tree.
-\label{lemma:linear-super-lattices}
-\end{lemma}
-
-\begin{theorem}
-(Version 1 and 2) Using \ref{lemma:linear-lattices}, \ref{lemma:linear-super-lattices}, and their product automaton, we can construct a VPDA such that there is a bijection between successful runs and pairs of trees in $L(G_t)$ such that the lattice of one is entirely covered by the other.
-\end{theorem}
-
-\begin{theorem}
-We can construct a sound, decidable algorithm that takes a language definition in version 1 or 2 and answers ''resolvably ambiguous'', ''unresolvably ambiguous'', or ''unknown''.
-Additionally, in version 1, this algorithm will never answer ''unknown'', i.e., it is then complete.
-\label{lemma:static-algorithm}
-\end{theorem}
-
-\begin{definition}
-Version 2.5 allows a language definition to use parentheses, but restricts it to never produce double parentheses, and changes $G'_w$ to also never recognize double parentheses.
-\end{definition}
-
-\begin{theorem}
-\ref{lemma:bounded-lattice} holds also for version 2.5, thus we can extend \ref{lemma:static-algorithm} to cover version 2.5 as well.
-\end{theorem}
-
-\subsection{Dynamic Stuff}
-
-This section assumes that we at no point produce an infinite amount of parse trees.
-My intuition (corroborated by some quick googling) suggests that this can only happen if there is a productive, accessible nonterminal $N$ such that $N =>^{+} N$.
-This is not checked at the moment.
-
-\begin{lemma}
-$\forall t \in L(G_t).\ \words(t)$ is a visibly pushdown language.
-\end{lemma}
-
-\begin{theorem}
-  Given a set of trees in $L(G_t)$, we can construct a corresponding set of VPDAs that recognize the words that are unique to each tree (i.e., isn't in $\words(t)$ for some other tree $t$ in the set).
-\end{theorem}
-
-\begin{theorem}
-  We can construct an algorithm $\localize(F)$ that takes a finite parse forest $F \subseteq L(G_t)$ and selects a set of subforests\footnote{This is a term that I have introduced, and so it will need to be described.
-Basically a parse forest for a part of a word, where each tree is a subtree of the original parse forest.} such that:
-  \begin{itemize}
-  \item The unselected parts are identical across the different syntax trees, including range (i.e., we report all ambiguities).
-  \item No selected subforest is contained in another (i.e., we only report one ambiguity per range).
-  \item No selected subforest can be replaced by zero or more of its children while still satisfying the points above (i.e., the selected subforests are minimal\footnote{By covered range, not by count.}).
-  \end{itemize}
-\end{theorem}
-
-\begin{theorem}
-  Given an ambiguous word $w \in L(G'_w)$, a localized ambiguity $F' \in \localize(\parse(w))$ covering the subword $w' \in L_{G'_w}(N)$ (i.e., $\parse_N(w') = F'$), and a resolution $w'' \in L_{G'_w}(N)$ such that $\parse_N(w'') = \{t\} \subset \parse_N(w')$ for some $t \in L_{G_t}(N)$, the following holds:
-  \begin{itemize}
-  \item $\localize(\parse(w[w'/w''])) = \localize(\parse(w)) \setminus F'$
-  \end{itemize}
-  where $w[w'/w'']$ denotes replacing a subword $w'$ with $w''$.
-\end{theorem}
 

### 2019-06-29 paper/paper.tex
EOD

@@
 i.e., ambiguous programs.
+The former is well explored in the form of heuristics for ambiguity detection (e.g.
+\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g.
+\cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny.
+Here, previous work is limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
+However, an abstract syntax tree is essentially an implementation detail, and may not provide any useful clues for an end-user to resolve the ambiguity.
+Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019}, is to automatically generate several unambiguous programs, each of which produces the same abstract syntax tree as one of the posssible interpretations, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation.
+For example, in a language with out precedence, ''$1 + 2 * 3$'' is ambiguous, but the two possible interpretations can be written unambiguously as ''$(1 + 2) * 3$'' and ''$1 + (2 * 3)$'' respectively.
+
+In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program is exists.
+This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
+We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot.
+We call the former \emph{resolvable ambiguities}, and the latter \emph{unresolvable ambiguities}.
 

### 2019-07-04 paper/paper.tex
Before splitting work

--- 2019-06-28 paper/paper.tex
@@
 \newcommand{\posp}[1]{\pospl#1\pospr}
+\newcommand{\dfa}{\mathit{dfa}} % Go from label to its corresponding dfa
+\newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
+\newcommand{\nt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs
 
@@
 
-Outline of introduction, one bullet per paragraph, roughly:
-
-\begin{itemize}
-\item Creating languages, particularly DSLs, it's nice to create pieces and compose.
-Composing syntax in particular quickly creates ambiguous grammars.
-Two broad approaches: forbid/prevent ambiguities at all (statically), or handle at parse-time (dynamically).
-Former more well-explored than the latter, which we focus on.
-The former is essentially the ambiguity problem, but the latter has not been formally defined.
-(To my knowledge we're first in presenting resolutions as opposed to interpretations (ASTs), which means that what we're talking about here is a little bit different from what previous approaches have done)
-
-\item Motivated by issues in \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011} ''we introduce a formal definition for resolvable ambiguity.'' (They both merely present ASTs, which are unlikely to be nice for an end-user.
-Probably want better motivation than ''nice.'') \ldots It turns out that this is not always possible, prompting an additional problem: can we statically determine if a language is resolvably ambiguous?
-We call the former the dynamic resolvability problem, and the latter the static resolvability problem.
-
-\item The dynamic problem is ''easy'' (though we haven't found any previously presented solution), but the static problem is not.
-We present a subclass of languages for which we have a decidable, sound, and complete decision procedure for the static problem, but also two larger classes of languages for which the algorithm is merely decidable and sound, or not applicable at all, respectively.
-\end{itemize}
-
-\noindent\hrulefill
-
 When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments.
@@
 In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous.
-Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous words.
-i.e., ambiguous programs.
-
-\noindent\hrulefill
-
-% I really don't like this paragraph. I'm trying to say that much work has been done on ambiguity, and that work in other fields include it, while PL tends to be more narrow-minded.
-Formal languages have long had the concept of ambiguity, of words that can be constructed in multiple ways, for example using multiple distinct derivations of a context-free grammar.
-In linguistics, this is important, since human languages tend to be highly ambiguous, and as such much work has been done on the subject.
-Examples include general parsing algorithms \cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}, ambiguity detection (most importantly its undecidability \cite{cantorAmbiguityProblemBackus1962}, but also various pragmatic approaches \cite{brabrandAnalyzingAmbiguityContextFree2007,axelssonAnalyzingContextFreeGrammars2008,bastenAmbiguityDetectionProgramming2011})
-
-However, when using these formal methods to specify the syntax for programming languages, ambiguity becomes a strictly undesirable property.
-We tend to want to give the compiler a single interpretation of any given program, and so we work hard to make programming language grammars entirely unambiguous.
-This view, that eliminating ambiguity is \emph{the} correct thing to do, is highly prevalent (e.g.
+Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs.
+The former is well explored in the form of heuristics for ambiguity detection (e.g.
+\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g.
+\cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny.
+This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g.
 \cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).
 
-Some exceptions do exist however.
-Closest to our approach are \citet{danielssonParsingMixfixOperators2011}, who ''feel that it is overly restrictive to require the grammar to be unambiguous'', and suggest that only ambiguous \emph{parses} should be rejected, not ambiguous grammars.
-This seems to stem largely from the fact that their grammars are mostly constructed from user-defined operators, and so cannot be expected to compose without ambiguity.
-Instead, in case of an unambiguous parse, they present an error that includes the possible parses, to assist in debugging.
-
-We propose that this idea---treating ambiguity as a parse-time error---can be beneficial even when the entire grammar is constructed by a single language designer, if done in a more rigorous fashion.
-Consider the following nested match expression in OCaml (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
+Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
+\cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
+However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity.
+Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation.
+For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.
+
+In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists.
+For example, in a language with semicolon for both sequential composition and element separation in a list (e.g.
+OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$.
+The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.
+This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
+We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot.
+We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}.
+This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.
+
+% TODO: I feel like there should be a clear transition between the more pracctical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.
+
+% TODO: resolvable ambiguity cannot be defined solely in terms of a grammar, we require interpretations of words as well, thus we have to introduce a new formalism that does this, the formalism is an intuitive extension of EBNF grammars.
+this is probably written as a contrast with classical ambiguity, which should be described briefly for contrast
+
+Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)?
+For classical ambiguity in context-free grammars, the former problem is solved by any general parser, while the latter is undecidable \cite{cantorAmbiguityProblemBackus1962}.
+For resolvable ambiguity, in a formalism that associates trees with EBNF-grammars, the dynamic problem is similarly solvable, and the static problem is similarly difficult.
+We do not yet know if the latter is undecidable in general, but we have found a subclass of languages for which it is decidable.
+
+More specifically, this paper makes the following contributions:
+
+\begin{itemize}
+\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp.
+word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
+\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
+  \begin{itemize}
+  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
+  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
+  \end{itemize}
+\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
+  \begin{itemize}
+  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
+  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
+  \end{itemize}
+  % Briefly say something about the tool, that it lies ``on top of'' what's presented herein
+\end{itemize}
+
+\section{Motivating Example}
+
+This section walks through an example from OCaml where an ambiguity error can be clearer than what the actual compiler produces, suggesting that (resolvable) ambiguity has merit even in the context of a general purpose programming language.
+
+Consider the following nested match expression (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
 
@@
 \begin{grammar}
-<expr> ::= 'match' <expr> 'with' <pattern-matching>
-
-<pattern-matching> ::= '|'? <pattern> ('when' <expr>)? \\
-('|' <pattern> ('when' <expr>)? '->' <expr>)*
+<expr> ::= \verb|'match'| <expr> \verb|'with'| <pattern-matching>
+
+<pattern-matching> ::= \verb{'|'{$^{?}$ <pattern> (\verb|'when'| <expr>)$^{?}$ \\
+(\verb{'|'{ <pattern> (\verb|'when'| <expr>)$^{?}$ \verb|'->'| <expr>)*
 \end{grammar}
@@
 Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user.
-Instead, we can present the alternatives in the form of modified code that parses unambiguously as the corresponding alternative:
+Instead, we can present the alternatives in the form of modified code that parses unambiguously as each corresponding alternative:
 
@@
 
-\noindent Unfortunately, not all ambiguities can be resolved in this way. Again, looking at an extract from the informal OCaml grammar:
-
-\setlength{\grammarindent}{5em}
-\begin{grammar}
-<expr> ::= <expr> ';' <expr>
-  \alt '[' <expr> (';' <expr>)* ';'? ']'
-  \alt <constant>
-\end{grammar}
-
-\noindent The first production is sequential composition, the second is lists (the empty list is under \synt{constant}).
-Now consider the expression ''\ocaml{[1; 2]}''.
-We find that it is ambiguous with two alternatives:
-\begin{enumerate}
-  \item A list with two elements.
-  \item A list with one element, namely a sequential composition.
-\end{enumerate}
-
-\noindent We can write the latter option unambiguously by putting parentheses around ''\ocaml{1; 2}'', but there is no way to unambiguously write the former.
-OCaml itself solves this by unambiguously parsing the expression as a two-element list.
-In this case, the ''arbitrary'' choice seems reasonable; there is no other way to write a two-element list, ambiguously or not\footnote{Ignoring the cons operator, since this syntax is supposed to be syntactic sugar for it.}, hence we would rather like ''\ocaml{[1; 2]}'' to be interpreted as such.
-
-The possibility of an unresolvable ambiguity is troubling however, since it represents an error that a user might encounter but only a language designer can solve; it requires a change to the grammar.
-As such, we propose to allow ambiguous programming language grammars, as long as all possible ambiguities can be resolved, and only require \emph{programs} to be unambigous.
-We call this property \emph{resolvable ambiguity}.
-
-We claim the following contributions:
-
-\begin{itemize}
-\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp.
-word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
-\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
-  \begin{itemize}
-  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound for the first two versions, and complete for the first (Section~\ref{sec:static}).
-  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
-  \end{itemize}
-\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
-  \begin{itemize}
-  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
-  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
-  \end{itemize}
-\end{itemize}
-
 \section{Preliminaries}
@@
 
-A \emph{bag} (alternatively known as a \emph{multiset}) is a generalization of a set, it is an unordered collection where each element may appear multiple times.
+A \emph{bag}, or \emph{multiset}, is a generalization of a set; it is an unordered collection where each element may appear multiple times.
 The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$.
@@
   L(\epsilon) & \{\epsilon\} \\
-  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \\
+  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \} \\
   L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
@@
 
-\noindent We will denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
+\noindent We denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
 
@@
 
-A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is Kleene-star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
+A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is the Kleene star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
 
@@
 
-The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{left-most derivations}.
-A left-most derivation is a derivation where the non-terminal being replaced is always the left-most one.
+The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{leftmost derivations}.
+A leftmost derivation is a derivation where the non-terminal being replaced is always the leftmost one.
 
 \begin{definition}
-A word $w \in L(G)$ is ambiguous if there are two distinct left-most derivations of that word.
+A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of that word.
 \end{definition}
@@
 Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
-$\lambda$ essentially means ''ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).
+$\lambda$ essentially means ``ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).
 
@@
 \begin{itemize}
-\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$.
-\item $p \xrightarrow{i} q$, where $i \in \T_i$.
-\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$.
+\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$; or
+\item $p \xrightarrow{i} q$, where $i \in \T_i$; or
+\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$,
 \end{itemize}
@@
 
-The partition names stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
-We will instead primarily use $\T_c$ and $\T_r$ for balanced parentheses.
+The names of the partitions stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
 
@@
 
-\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2013}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run, none are redundant.
+\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
 Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
@@
 Note that a single word may have multiple interpretations.
-As such, we will define a language as:
+As such, we define a language as:
 
@@
 \noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees.
-In such a language, $\parse(1 + 2 * 3)$ would produce a set containing the two ASTs in Figure~\ref{fig:arith-example}:
+In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs in Fig.~\ref{fig:arith-example}:
 
@@
     1
-    [.{$*$}
+    [.{$\cdot$}
       2
       3 ] ] &
-  \Tree [.{$*$}
+  \Tree [.{$\cdot$}
     [.{$+$}
@@
 \end{tabular}
-\caption{The two allowable interpretations of $1 + 2 * 3$ in an arithmetic language without precedence.}
+\caption{The two allowable interpretations of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
 \label{fig:arith-example}
@@
 \item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
-\item If we select a $\parse$ that relates words to their left-most derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
+\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
 \end{itemize}
@@
 
-\noindent The example in Figure~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way.
-However, if we add grouping parentheses to the language we find that $1 + (2 * 3)$ and $(1 + 2) * 3$ are unambiguously interpreted as the first and second tree, respectively.
+\noindent The example in Fig.~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way.
+However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.
 
@@
 
-We thus propose to loosen the common ''no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous.
+We thus propose to loosen the common ``no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous.
 However, merely having an arbitrary function $\parse$ gives us very little to work with, and no way to decide whether the language it defines is resolvably ambiguous or not.
-The remainder of this paper will thus consider $\parse$ functions defined with a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, that gives us some decidable properties.
+The remainder of this paper will thus consider $\parse$ functions defined using a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, which gives us some decidable properties.
 
@@
 \begin{definition}
-  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unresolvable}$) solving the static resolvability problem is:
+  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the static resolvability problem is:
   \begin{itemize}
@@
 \begin{definition}
-  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unresolvable}$) solving the dynamic resolvability problem is:
+  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
   \begin{itemize}
@@
 
-With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Figure~\ref{fig:input-language-definition}.
+With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Fig.~\ref{fig:input-language-definition}.
 Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label.
-Also note that the right-hand side of a production is here a regular expression, rather than the theoretically simpler sequence used in a context-free grammar.
-For proof technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense.
+Also note that the right-hand side of a production is a regular expression, rather than the theoretically simpler sequence used in a context-free grammar.
+For proof-technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense, when marks are removed.
 This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}.
@@
 To lessen clutter, we will write $E_\emptyset$ as $E$.
-As an example, consider the language definition in Figure~\ref{fig:running-example-definition}, which will be used as a running example.
+As an example, consider the language definition in Fig.~\ref{fig:running-example-definition}, which will be used as a running example.
 In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication.
-By ''direct child'' we mean ''without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.
+By ``direct child'' we mean ``without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.
 
@@
 \item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
-\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding parentheses and other forms of disambiguation (i.e., the result of marks).
+\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding grouping parentheses and other forms of disambiguation (i.e., the result of marks).
 \item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
@@
 
-\noindent Figure~\ref{fig:running-example-generated} contains the four grammars generated from our running example in Figure~\ref{fig:running-example-definition}.
+\noindent Fig.~\ref{fig:running-example-generated} shows the four grammars generated from our running example in Fig.~\ref{fig:running-example-definition}.
 The context-free grammars are produced by a rather standard translation from regular expressions to CFGs, while the primed grammars get a new non-terminal per distinctly marked non-terminal in $D$, where each new non-terminal only has the productions whose label is not in the mark.
-For example, the non-terminal $E_{\{a\}}$ in Figure~\ref{fig:running-example-generated:t-prime} has no production corresponding to the $a$ production in Figure~\ref{fig:running-example-definition}.
+For example, the non-terminal $E_{\{a\}}$ in Fig.~\ref{fig:running-example-generated:t-prime} has no production corresponding to the $a$ production in Fig.~\ref{fig:running-example-definition}.
 
@@
 
-Examples of elements in each of these four languages can be seen in Figure~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
-Each element corresponds to the word ''$(1 + 2) * 3$'' in $L(G'_D)$.
-Note that the word in $L(G_D)$ is ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ''$((1 + 2)) * 3$'' and ''$(1 + 2) * (3)$''.
-As a memory aid, the prime versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations from $D$.
+Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
+Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$.
+Note that the word in $L(G_D)$ is ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ``$((1 + 2)) * 3$'' and ``$(1 + 2) * (3)$''.
+As a memory aid, the primed versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations (i.e., ignoring marks) from $D$.
 We will generally refer to elements of $L(G_D)$ as $w$, $L(G'_D)$ as $w'$, $L(T_D)$ as $t$, and $L(T'_D)$ as $t'$.
@@
 At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses; they are allowed exactly where they would surround a complete production.
-For example, $G_D$ in Figure~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
-However, starting with that language definition would allow the expression ''$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ''$;2$'' is not a valid expression.
+For example, $G_D$ in Fig.~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
+However, starting with that language definition would allow the expression ``$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ``$;2$'' is not a valid expression.
 
 Finally, we require a function $\semantic : L(T'_D) -> L(T_D)$ that removes grouping parentheses from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
-The relation between the four grammars in terms of $\yield$ and $\semantic$ can be seen in Figure~\ref{fig:grammar-square}.
+The relation between the four grammars in terms of $\yield$ and $\semantic$ can be seen in Fig.~\ref{fig:grammar-square}.
 With this we can define $\parse : L(G'_D) -> 2^{L(T_D)}$, along with its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
@@
 \noindent The latter is mostly useful in later sections, but $\parse$ allow us to consider some concrete examples of resolvable and unresolvable ambiguities.
-For example, in our running example (Figure~\ref{fig:running-example-definition}), the word \verb|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
+For example, in our running example (Fig.~\ref{fig:running-example-definition}), the word \texttt|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
 
@@
 \noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not.
-The solution is to modify the language definition in Figure~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.
+The solution is to forbid list elements from being sequences by modify the language definition in Fig.~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.
 
@@
 This implies that all parentheses present in $G'_D$ are grouping parentheses.
-The second restriction implies that no parentheses are required (as a counterexample, assuming normal precedence, the parentheses in ''$(1 + 2) * 3$'' are required and removing them would produce a different interpretation).
+The second restriction implies that no parentheses are required (as a counterexample, assuming normal precedence, the parentheses in ``$(1 + 2) * 3$'' are required and removing them would produce a different interpretation).
 
@@
 If we thus delimit ourselves to only consider languages where words have no unbalanced parentheses\footnote{I.e., the vast, vast majority of programming languages currently in use.} we can give the following alternative definition of a word: a word is a two-tuple containing a sequence of non-parenthesis terminals and a bag (or multiset) of ranges covered by parentheses.
-For example, the word ''$(1 + 2) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{0}{3}\})$, while ''$((1 + 2)) * 3$'' is equivalent to $(\text{''}1 + 2 * 3\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
+For example, the word ``$(1 + 2) * 3$'' is equivalent to $(\text{``}1 + 2 * 3\text{''}, \{\range{0}{3}\})$, while ``$((1 + 2)) * 3$'' is equivalent to $(\text{``}1 + 2 * 3\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
 We will refer to the first component of the tuple as the \emph{basic word} and the second as the \emph{range bag}.
@@
 However, there is a case where this alternate representation is not fully isomorphic; namely when a pair of parentheses would yield a zero-length range.
-As an example, consider the two words ''$(())$'' and ''$()()$''.
-Both of these have an alternate representation of $(\text{''}\text{''}, \{\range{0}{0}, \range{0}{0}\})$.
+As an example, consider the two words ``$(())$'' and ``$()()$''.
+Both of these have an alternate representation of $(\text{``}\text{''}, \{\range{0}{0}, \range{0}{0}\})$.
 In versions~1 and 2 this corresponds to grouping parentheses surrounding zero-length productions.
@@
 
-We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., $\basic(\text{''}((1 + 2)) * 3\text{''}) = \text{''}1 + 2 * 3\text{''}$ and $\rangebag(\text{''}((1 + 2)) * 3\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
+We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., $\basic(\text{``}((1 + 2)) * 3\text{''}) = \text{``}1 + 2 * 3\text{''}$ and $\rangebag(\text{``}((1 + 2)) * 3\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
 
@@
 \end{lemma}
-\noindent Intuitively, $\words$ first applies the ''inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
+\noindent Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
 The only terminals changed in this process are parentheses, i.e., different $w'$ can only differ in their range bags.
@@
 \begin{lemma}
-  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\ R \subseteq \support(\rangebag(w'))$, i.e., some parentheses are required.
+  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\\ R \subseteq \support(\rangebag(w'))$, i.e., some parentheses are required.
   \label{lemma:required-parentheses}
 \end{lemma}
-\noindent For example, removing the parentheses in ''$(1 + 2) * 3$'' changes the interpretations produced.
+\noindent For example, removing the parentheses in ``$(1 + 2) * 3$'' changes the interpretations produced.
 Each required range is a direct consequence of a mark on a non-terminal in $D$.
-We will write the word implied by the first quantified expression as $w'_\bot$.
+We will write $w'_\bot$ for the word implied by the first quantified expression.
 It is unique, since its basic word is fixed by $t$, and its rangebag is a set.
@@
 \noindent As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes.
-We will write the word implied by the first quantified expression $w'_\top$.
+We will write $w'_\top$ for the word implied by the first quantified expression, which is also unique.
 
@@
   \end{tikzpicture}
-  \caption{The lattice of words for the tree $a(a(n(1)+n(2))+n(3))$. The boxed words are shared with the lattice in Figure~\ref{fig:lattice2}.}
+  \caption{The lattice of words for the tree $a(a(n(1)+n(2))+n(3))$. The boxed words are shared with the lattice in Fig.~\ref{fig:lattice2}.}
   \label{fig:lattice1}
@@
 Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
-For example, Figure~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Figure~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}).
+For example, Fig.~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}).
 Each partition is represented by the word whose rangebag is a set.
@@
 \end{tikzpicture}
-\caption{The lattice of words for the tree $a(n(1)+a(n(2)+n(3)))$. The boxed words are shared with the lattice in Figure~\ref{fig:lattice1}.}
+\caption{The lattice of words for the tree $a(n(1)+a(n(2)+n(3)))$. The boxed words are shared with the lattice in Fig.~\ref{fig:lattice1}.}
 \label{fig:lattice2}
@@
 
-To show the connection between resolvable ambiguity and these lattices, consider the word ''$1 + 2 + 3$''.
+To show the connection between resolvable ambiguity and these lattices, consider the word ``$1 + 2 + 3$''.
 It has two interpretations in our running example, $a(a(n(1)+n(2))+n(3))$, which we would normally write as $(1 + 2) + 3$, and $a(n(1)+a(n(2)+n(3)))$, which we would normally write $1 + (2 + 3)$.
@@
 Lemmas~\ref{lemma:same-basic} and \ref{lemma:top-bottom-determine} imply that a lattice encoding requires three things: a basic word, a set of \emph{required} parentheses, and a set of \emph{possible} parentheses.
-We encode required parentheses with ''$\reqp{}$'' and possible parentheses with ''$\posp{}$''.
-For example, the lattice in Figure~\ref{fig:lattice2} is represented by ''$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ''$(1 + 2) * 3$'' has a lattice represented by ''$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
+We encode required parentheses with ``$\reqp{}$'' and possible parentheses with ``$\posp{}$''.
+For example, the lattice in Fig.~\ref{fig:lattice2} is represented by ``$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ``$(1 + 2) * 3$'' has a lattice represented by ``$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
 With this encoding we can apply the formidable body of knowledge available for word languages, albeit with some extra care; the linear encoding of a lattice is not unique.
 
-For example, ''$\posp{\posp{1}}$'' and ''$\posp{1}$'' represent the same lattice, as do ''$\reqp{\posp{2}}$'' and ''$\reqp{2}$''.
-To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses, e.g., ''$\posp{\posp{1}}$'' is uniquely represented as ''$\posp{1}$'' while ''$\reqp{\posp{2}}$'' is uniquely represented as ''$\reqp{2}$''.
+For example, ``$\posp{\posp{1}}$'' and ``$\posp{1}$'' represent the same lattice, as do ``$\reqp{\posp{2}}$'' and ``$\reqp{2}$''.
+To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses, e.g., ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
 
 Lattice equality is now equvialent with equality of the linear encoding.
-To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ''up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ''down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
-For example, ''$\reqp{1}2$'' is entirely contained in ''$\reqp{1}\posp{2}$'', which is entirely contained in ''$\posp{1}\posp{2}$''.
+To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
+For example, ``$\reqp{1}2$'' is entirely contained in ``$\reqp{1}\posp{2}$'', which is entirely contained in ``$\posp{1}\posp{2}$''.
 
@@
 To detect lattices contained in each other, we create a modified copy, where the copy may add arbitrary (balanced, well-nested) possible parentheses, and replace any required pair of parentheses with a possible pair.
-The copy maintains much of the structure of the original, and in particular, keeps the connection to trees in $L(T_D)$ (though it is no longer a bijection, there are multiple successful runs per tree since there are multiple larger lattices).
+The copy maintains much of the structure of the original, and in particular, keeps the connection to trees in $L(T_D)$ (though it is no longer a bijection: there are multiple successful runs per tree since there are multiple larger lattices).
 Two distinct successful runs, one in each automaton, that recognize the same word then imply two distinct trees where the lattice of one is entirely contained in the other.
@@
 Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we go back.
-We use $Q \times Q \times \{\text{''}\reqp{}\text{''}, \text{''}\posp{}\text{''}\}$ as our stack alphabet, where the first two components are states in the parent production automaton and the last component records whether the parenthesis pair is required or merely possible.
+We use $Q \times Q \times \{\text{``}\reqp{}\text{''}, \text{``}\posp{}\text{''}\}$ as our stack alphabet, where the first two components are states in the parent production automaton and the last component records whether the parenthesis pair is required or merely possible.
 We replace every edge $p \xrightarrow{l} q$, where $l \in \Labels$, with:
@@
 \begin{itemize}
-  \item an edge $p \xrightarrow{'\pospl', +(p, q, \text{''}\posp{}\text{''})} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
-  \item an edge $q' \xrightarrow{'\pospr', -(p, q, \text{''}\posp{}\text{''})} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
-\end{itemize}
-
-\noindent Similarly, for every edge $p \xrightarrow{l'} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\reqpl', +(p, q, \text{''}\reqp{}\text{''})} p'$ and $q' \xrightarrow{'\reqpr', -(p, q, \text{''}\reqp{}\text{''})} q$.
+  \item an edge $p \xrightarrow{'\pospl', +(p, q, \text{``}\posp{}\text{''})} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
+  \item an edge $q' \xrightarrow{'\pospr', -(p, q, \text{``}\posp{}\text{''})} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
+\end{itemize}
+
+\noindent Similarly, for every edge $p \xrightarrow{l'} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\reqpl', +(p, q, \text{``}\reqp{}\text{''})} p'$ and $q' \xrightarrow{'\reqpr', -(p, q, \text{``}\reqp{}\text{''})} q$.
 
@@
 This is where our simplification is used, without it we might introduce double parentheses here.
-To lessen clutter in the figure below we omit the third component of the stack symbol, since it is fully determined by the input symbol, e.g., if an edge reads $'\pospl'$ then the third component of the stack symbol pushed must be ''$\posp{}$''.
+To lessen clutter in the figure below we omit the third component of the stack symbol, since it is fully determined by the input symbol, e.g., if an edge reads $'\pospl'$ then the third component of the stack symbol pushed must be ``$\posp{}$''.
 
@@
 Each one corresponds to one of the allowable child productions at that point in the tree, thus there is one run per tree.
-STOP HERE, THE TEXT BEYOND IS NOT UPDATED.
+
+UNFINISHED
 
@@
 
-We now have a pushdown automaton (call it $A_{()}$) that recognizes the top word for each tree in $L(G_t)$.
+%% We now have a pushdown automaton (call it $A_{()}$) that recognizes the top word for each tree in $L(G_t)$.
 Next we need to be able to add arbitrary parentheses, to produce a word with a rangeset superset.
@@
 
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-  \node[state] (4) [below of=1, xshift=1cm] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-  \node[state,initial] (s) [left of=4] {};
-  \node[state,accepting] (e) [right of=5] {};
-
-  \draw
-  (1) edge[loop, in=80, out=100, looseness=7] node{} (1)
-  (1) edge[loop, in=70, out=110, looseness=6] node{} (1)
-  (2) edge[loop, in=80, out=100, looseness=7] node{} (2)
-  (2) edge[loop, in=70, out=110, looseness=6] node{} (2)
-  (3) edge[loop, in=-10, out=10, looseness=7] node{} (3)
-  (3) edge[loop, in=-20, out=20, looseness=6] node{} (3)
-  (4) edge[loop, in=-80, out=-100, looseness=7] node{} (4)
-  (4) edge[loop, in=-70, out=-110, looseness=6] node{} (4)
-  (5) edge[loop, in=-80, out=-100, looseness=7] node{} (5)
-  (5) edge[loop, in=-70, out=-110, looseness=6] node{} (5)
-  (s) edge[left] node{$'(', +s$} (1)
-  (s) edge[above] node{$'(', +s$} (4)
-  (3) edge[right] node{$')', -s$} (e)
-  (5) edge[above] node{$')', -s$} (e)
-  (1) edge[above] node{$s$} (2)
-  (4) edge[above] node{$z$} (5)
-  (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
-  (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
-  (3) edge[loop above] node{$')', -(2, 3)$} (3)
-  (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
-
-\end{tikzpicture}
-\end{center}
-
-\noindent We will call this automaton $A'_{()}$.
+%% \begin{center}
+%% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+%%     scale = 1,transform shape]
+
+%%   \node[state] (1) {$1$};
+%%   \node[state] (2) [right of=1] {$2$};
+%%   \node[state] (3) [right of=2] {$3$};
+%%   \node[state] (4) [below of=1, xshift=1cm] {$4$};
+%%   \node[state] (5) [right of=4] {$5$};
+%%   \node[state,initial] (s) [left of=4] {};
+%%   \node[state,accepting] (e) [right of=5] {};
+
+%%   \draw
+%%   (1) edge[loop, in=80, out=100, looseness=7] node{} (1)
+%%   (1) edge[loop, in=70, out=110, looseness=6] node{} (1)
+%%   (2) edge[loop, in=80, out=100, looseness=7] node{} (2)
+%%   (2) edge[loop, in=70, out=110, looseness=6] node{} (2)
+%%   (3) edge[loop, in=-10, out=10, looseness=7] node{} (3)
+%%   (3) edge[loop, in=-20, out=20, looseness=6] node{} (3)
+%%   (4) edge[loop, in=-80, out=-100, looseness=7] node{} (4)
+%%   (4) edge[loop, in=-70, out=-110, looseness=6] node{} (4)
+%%   (5) edge[loop, in=-80, out=-100, looseness=7] node{} (5)
+%%   (5) edge[loop, in=-70, out=-110, looseness=6] node{} (5)
+%%   (s) edge[left] node{$'(', +s$} (1)
+%%   (s) edge[above] node{$'(', +s$} (4)
+%%   (3) edge[right] node{$')', -s$} (e)
+%%   (5) edge[above] node{$')', -s$} (e)
+%%   (1) edge[above] node{$s$} (2)
+%%   (4) edge[above] node{$z$} (5)
+%%   (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
+%%   (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
+%%   (3) edge[loop above] node{$')', -(2, 3)$} (3)
+%%   (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
+
+%% \end{tikzpicture}
+%% \end{center}
+
+%% \noindent We will call this automaton $A'_{()}$.
 Successful runs in this automaton have a surjection to runs in $A_($ (ignore transitions along the newly added edges), and thus also have a surjection to parse trees in $L(G_t)$.
@@
 
-\begin{enumerate}
-  \item $p'$ only uses transitions present in $A_{()}$.
+%% \begin{enumerate}
+%%   \item $p'$ only uses transitions present in $A_{()}$.
 This is a successful run in $A_{()}$, and distinct from $p$.
 But there is a bijection between runs in $A_{()}$ and parse trees in $L(T)$, thus $p'$ represents a different parse tree.
-  \item $p'$ uses at least one transition added in $A'_{()}$.
+%%   \item $p'$ uses at least one transition added in $A'_{()}$.
 Using the surjection between runs in $A'_{()}$ and $A_{()}$ we find a new successful run that produces a different word (at least one fewer pair of parentheses).
 Since this run produces a different word, it must be distinct from $p$, and thus represent a different parse tree.
-\end{enumerate}
-
-\noindent Two distinct successful runs that accept the same word thus represent two trees where one permits a superset rangeset of the other.
+%% \end{enumerate}
+
+%% \noindent Two distinct successful runs that accept the same word thus represent two trees where one permits a superset rangeset of the other.
 To find such runs we construct a product automaton and trim it.
@@
 
-In this product automaton, if any transition pushes a stack symbol $(a, b)$ where $a \neq b$, or transitions to a state $(p, q)$ where $p \neq q$, then there is a successful run that corresponds to two distinct runs through $A_{()}$ and $A'_{()}$ (since the automaton is trim).
+%% In this product automaton, if any transition pushes a stack symbol $(a, b)$ where $a \neq b$, or transitions to a state $(p, q)$ where $p \neq q$, then there is a successful run that corresponds to two distinct runs through $A_{()}$ and $A'_{()}$ (since the automaton is trim).
+
+\begin{theorem}
+  There is a bijection between successful runs in $A_{()}$ and trees $t \in L(T_D)$.
+\end{theorem}
+
+\begin{proof}
+  $A_{()}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
+The two are distinct, and $f$ is reachable only through transitions that pop a stack symbol with $s$ and $f$ as the two first components.
+Transitions that push such symbols originate in $s$, and no other transitions originate in $s$.
+Thus all successful runs start in configuration $(s, \epsilon)$ and end in $(f, \epsilon)$, whereby every pushing transition in the run can be paired with a corresponding popping transition.
+We will now construct a tree $t \in L(T_D)$ in a bottom-up fashion.
+
+  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, g, w)} (p', (p, q, g, w) \cdot s) \; R_2 \; (q', (p, q, g, w) \cdot s \xrightarrow{b, -(p, q, g, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
+The only transitions that leave the stack unchanged in $A_{()}$ come from DFAs generated from productions in $D$.
+There is no way to transition between two such DFAs without using a pushing or popping transition, thus $R_2$ must be a successful run in the DFA for some labelled production $N -> l : r$ in $D$.
+$T_D$ must contain a corresponding production $N -> l(r)$.
+Now consider the sequence of input symbols $w'$ along the transitions in $R_2$.
+We have $w' \in L(r)$, since $R_2$ is a successful run in a DFA constructed from $r$.
+Thus $l(w') \in L_N(T_D)$.
+Now add parent nodes to this subtree for every label in $w$, e.g., if $w = l_1 \cdot l_2 \cdot l_3$ construct $l_1(l_2(l_3(l(w'))))$.
+We can then replace the subsequence of $R$ with this subtree, and then repeat the process until we have a tree with root $l'$, where $D$ has a labelled production $S -> l': r'$ for some $r'$ and the starting symbol $S$.
+
+  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{()}$ starting from $(s, \epsilon)$.
+If the next terminal is in $\T$, i.e., a strictly zero-arity terminal, we take a transition originating in a DFA, which must be unique.
+If the next terminal is in $\Labels$, i.e., an unranked terminal, we take one of the added edges.
+However, this choice of edge might not be unique, there may be multiple edges from the current state to an appropriate DFA initial state.
+These stem from the original DFA having two transitions from the same state with the same label, except one is primed, e.g., $p \xrightarrow{l} q$ and $p \xrightarrow{l'} q'$.
+However, if both of these transitions were valid choices that would imply that the DFA for that level can recognize two distinct words that differ only in whether some terminals are primed.
+But primed and non-primed terminals must originate from distinct occurances of marked non-terminals (since a primed terminal corresponds to a mark, and an unprimed terminal corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
+
+  Proof sketch: to go from a successful run to a tree $t \in L(T_D)$: by construction, all successful runs must begin and end with an empty stack.
+Every pushing transition can thus be paired with a popping transition in the run.
+We can construct a tree in a bottom-up fashion by replacing every such pair with a tree $l_1(l_2(\ldots l_n(x)\ldots))$, where $x$ is the sequence of terminals and/or trees between the pair, the terminals $l_1, l_2, \ldots, l_{n-1}$ stem from the fourth component of the pushed stack symbol ($l_1 \cdots l_{n-1}$), and $l_n$ is the label of the production whose DFA the pushing transition ends in.
+
+  For the inverse direction we construct a run by traversing the tree in a preorder\footnote{Where we also keep track of when we ``go back up'' the tree.}.
+Terminals in $\T$ cause transitions internally in a DFA, while terminals in $\Labels$ cause a pushing transition (and later a corresponding, unique popping transition) to the initial state of the DFA corresponding to the production with that label.
+Note that a sequence of terminals in $\Labels$ that have only a single child (which is also in $\Labels$) cause only one transition, all terminals but the last end up in the fourth component of the pushed stack symbol.
+The choice of pushing transition is \emph{not} unique however; there may be multiple available transitions that differ in the recognized input symbol and the second and third components of the pushed stack symbol, none of which are necessarily determined by the shape of the tree.
+This non-determinism does not introduce multiple successful runs however, since two such runs would imply the existence of two distinct runs through the DFA corresponding to some labelled production that recognize the same word, except for whether some terminals are primed.
+But two otherwise equal terminals where one is primed must originate from different occurrences of marked non-terminals in the original regular expression (since a primed terminal represents a mark, and an unprimed terminal represents the absence of a mark), thus that regular expression must be ambiguous after all marks are removed.
+But Section~\ref{sec:parse-time-disambiguation} explicitly ruled this out, thus only one of the pushing transitions can be a valid choice.
+
+% TODO: technically, we should perhaps mention that these two constructions are inverses of each other, but that's relatively easy to see once you understand them.
+\end{proof}
+
+\subsubsection{Formal Stuff}
+
+Given a language definition $D$, with terminals $\T$, labels $\Labels$, and starting non-terminal $S$.
+Construct two functions $\dfa : \Labels -> \mathit{DFA}(\T \cup (\Labels \times \{\top, \bot\})$ and $\passthrough : \Labels -> 2^{\Labels \times \Labels \times \{\top, \bot\} \times \Labels^{*}}$:
+
+\begin{description}
+\item[$\dfa(l)$:] Given the production $N -> l : r$, compute the regular language $L' := L(r) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
+Then replace every marked non-terminal $N'_m$ with the regular expression $(l_1, b_1) + (l_2, b_2) + \ldots + (l_n, b_n)$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, and $b_i = l_i \in m$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l + a' + m + n$.
+Then construct a DFA for this language.
+
+\item[$\passthrough(l)$:] Given a production $N -> l : r$, this is the set $\{(l, l, \bot, \epsilon)\} \cup \{(l, l', l' \in m, l) \mid \nt(l') = N', N'_m \in L(r)\}$.
+\end{description}
+
+\noindent Compute the set $P$ as the closure of $\bigcup_l \passthrough(l)$ using the following operation:
+
+$$ (l_1, l_2, r_1, w_1) \cdot (l_2, l_3, r_2, w_2) = (l_1, l_3, r_1 \lor r_2, w_1 \cdot w_2) $$
+
+\noindent $P$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
+Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for that tree are unresolvably ambiguous.
+
+Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$, construct a VPDA $(Q, \T', \Gamma, \delta, s, \{f\})$ where:
+
+% TODO: add transitions to/from final/initial states
+
+\begin{itemize}
+\item $s$ and $f$ are two new distinct states.
+\item $Q = \{s, f\} \cup \bigcup_l Q_l$.
+\item $\T' = \T \cup \{\reqpl, \reqpr, \pospl, \pospr\}$.
+\item $\Gamma = Q \times Q \times \{\reqp{}, \posp{}\} \times \Labels^{*}$.
+\item $\delta$ is defined by:
+  $$
+  \begin{array}{r@{\;=\;}l@{\;\mid\;}l}
+    \delta(p, a, \lambda)
+    & \multicolumn{2}{@{}l@{}}{\{(\delta_l(p, a), \lambda)\} \qquad \text{ where } p \in Q_l \text{ and } a \in \T} \\
+
+    \delta(p, \text{``}\reqpl\text{''}, \lambda)
+    & \{ (s_{l_3}, (p, q, \reqp{}, w))
+    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, r_1 \lor r_2 \} \\
+
+    \delta(p, \text{``}\pospl\text{''}, \lambda)
+    & \{ (s_{l_3}, (p, q, \posp{}, w))
+    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, \neg (r_1 \lor r_2) \} \\
+
+    \delta(f_{l_3}, \text{``}\reqpr\text{''}, (p, q, \reqp{}, w))
+    & \{ (q, \lambda)
+    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, r_1 \lor r_2, f_{l_3} \in F_{l_3} \} \\
+
+    \delta(f_{l_3}, \text{``}\pospr\text{''}, (p, q, \posp{}, w))
+    & \{ (q, \lambda)
+    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, \neg (r_1 \lor r_2), f_{l_3} \in F_{l_3} \} \\
+  \end{array}
+  $$
+\end{itemize}
 
@@
 
+UNFINISHED
+
 \section{Evaluation} \label{sec:evaluation}
@@
 
-% TODO: talk about the ''low precedence unary operator/deep precedence'' issue.
+% TODO: talk about the ``low precedence unary operator/deep precedence'' issue.
 
@@
 However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
+
+UNFINISHED
+
+\section{Conclusion}
 

### 2019-07-04 paper/paper.tex
Minipage hack to fix compilation bug
More related work

@@
 
-\begin{tabular}{l|l}
+\begin{minipage}{.35\textwidth}
 \begin{lstlisting}[language={[objective]caml},numbers=left]
@@
   | 2 -> "two"
-\end{lstlisting} &
-
+\end{lstlisting}
+\end{minipage}
+\vrule
+\begin{minipage}{\textwidth}
 \begin{lstlisting}
@@
          matches values of type string
-\end{lstlisting} \\
-\end{tabular}
+\end{lstlisting}
+\end{minipage}
 
@@
 
-\begin{tabular}{l}
+\begin{minipage}{\textwidth}
 \begin{lstlisting}[language={[objective]caml},numbers=left]
@@
 \end{lstlisting}
-\end{tabular}
+\end{minipage}
 
@@
 
-\section{Evaluation} \label{sec:evaluation}
+\section{Case Studies} \label{sec:evaluation}
 
@@
 
-\cite{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
+\citet{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
 However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
+This approach thus suggests an interesting direction for future work: to integrate it into a system designed around resolvable ambiguity, and extending the algorithms presented in this paper to cover it.
+
+\citet{danielssonParsingMixfixOperators2011} give a method for specifying grammars for expressions containing mixfix operators.
+They allow non-transitive, non-total precedence, and ``feel that it is overly restrictive to require the grammar to be unambiguous.'' Similar to our approach, they do not reject ambiguous grammars, only ambiguous parses.
+They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must be related by precedence (children must have higher precedence than parents).
+This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'-'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.
+
+Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
+For this paper, only the syntactic component \cite{schwerdfegerVerifiableCompositionDeterministic2009} of this analysis is relevant.
+This analysis guarantees that the composition of a base language and any number of extensions that have passed it will compose to a grammar in LALR(1).
+This is a far more restricted class of languages than unambiguous languages, not to mention resolvably ambiguous languages, and thus places far more restrictions on language designers than our approach.
+
+Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language \cite{heeringSyntaxDefinitionFormalism1989}.
+These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but also assume that the composed language is unambiguous, which in practice precludes the composition of languages constructed independently of each other.
 

### 2019-07-05 paper/paper.tex
Add rigor to section 4
Adds definition of interpreted language. Streamlines the definition
of an ambiguous word, as well as the definitions of resolvably
ambiguous words and languages.

@@
 \newcommand{\alt}{\mathit{alt}} % go from a w \in L(G'_w) to a tuple of a basic word and a range bag.
+\newcommand{\amb}{\mathit{amb}}
+
 \newcommand{\basic}{\mathit{basic}} % go from a w \in L(G'_w) to the first element of \alt(w).
@@
 Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense.
-We will call this meaning an \emph{interpretation} of a word.
+We call this meaning an \emph{interpretation} of a word.
 Note that a single word may have multiple interpretations.
-As such, we define a language as:
-
-\begin{itemize}
-\item An alphabet $\T$.
-\item A set of interpretations $T$.
-\item A function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
-\end{itemize}
+As such, we define an \emph{interpreted language} as:
+
+\begin{definition}
+  An interpreted language, written $L(\T, T, \parse)$, consists of
+\begin{itemize}
+\item an alphabet $\T$,
+\item a set of interpretations $T$, and
+\item a function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
+\end{itemize}
+
+  A word $w \in \T^{*}$ is in the language $L(\T, T, \parse)$, written
+  $w \in L(\T, T, \parse)$ iff $\parse(w) \neq \emptyset$.
+\end{definition}
 
 \noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees.
-In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs in Fig.~\ref{fig:arith-example}:
+In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs shown in Fig.~\ref{fig:arith-example}.
 
@@
 
-\noindent We further classify each word $w \in \T^{*}$ depending on the result of $\parse$:
+The ambiguity of a word $w \in \T^{*}$ is defined in terms of $\parse$:
 
 \begin{definition}
-  A word $w \in \T^{*}$ is:
-  \begin{itemize}
-  \item not in the language if $\parse(w) = \emptyset$.
-  \item unambiguous if $\exists t \in T.\ \parse(w) = \{t\}$.
-  \item ambiguous if $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$.
-  \end{itemize}
+  A word $w \in L(\T, T, \parse)$ is ambiguous, written $\amb(w)$, iff
+
+  $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$
 \end{definition}
 
-\noindent We can connect these cases to the classical definition of a language:
-
-\begin{itemize}
-\item Given a language defined by $\parse$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
+\noindent
+Note that the above definition implies that a word $w \in L(\T, T,
+\parse)$ is not ambiguous (or unambiguous) if $\exists t \in
+T.\ \parse(w) = \{t\}$.
+
+We can connect the above definition of an interpreted language to the
+classical definition of a language as follows:
+
+\begin{itemize}
+\item Given an interpreted language $L(\T, T, \parse)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
 \item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
@@
 
-\noindent A \emph{resolvably} ambiguous word is a word where all its interpretations can be written in an unambiguous way, or more formally:
-
-\begin{definition}
-  A word $w \in \T^{*}$ in a language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-word}
+\noindent A \emph{resolvably ambiguous} word is a word where all its interpretations can be written in an unambiguous way, formally:
+
+\begin{definition}\label{def:resolvable-word}
+  A word $w \in L(\T, T, \parse)$ is resolvably ambiguous, written $\rho_{\Sigma}(w, \parse)$, iff
+
+  $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
 \end{definition}
 
-\noindent The example in Fig.~\ref{fig:arith-example} is unresolvably ambiguous, since neither tree can be written in any other way.
+For clarity we abbreviate $\rho_{\T}(w, \parse)$ as $\rho(w)$ if $\T$
+and $\parse$ are clear from the context.
+
+The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way.
 However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.
 
-Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, or more formally:
-
-\begin{definition}
-  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous if $\forall w \in \T^{*}.\ \forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$. \label{def:resolvable-language}
+Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, formally:
+
+\begin{definition}\label{def:resolvable-language}
+  An interpreted language $L(\T, T, \parse)$ is resolvably ambiguous, written $\rho(L)$, iff
+
+  $\forall w \in L(\T, T, \parse).\ \rho_{\T}(w, \parse)$.
 \end{definition}

### 2019-07-05 paper/paper.tex
Some more related work

@@
 
+The detection of classical ambiguity in context-free grammars is undecidable in general \cite{cantorAmbiguityProblemBackus1962}, yet numerous heuristic approaches exist.
+Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using sat-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of \citet{bastenAmbiguityDetectionProgramming2011}.
+
 Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language \cite{heeringSyntaxDefinitionFormalism1989}.

### 2019-07-05 paper/paper.tex
Clean up and make resolvability problems more formal

@@
 \begin{definition}
-  A word $w \in L(\T, T, \parse)$ is ambiguous, written $\amb(w)$, iff
+  A word $w \in L(\T, T, \parse)$ is ambiguous, written $\amb_L(w)$, iff
 
@@
 \begin{definition}\label{def:resolvable-word}
-  A word $w \in L(\T, T, \parse)$ is resolvably ambiguous, written $\rho_{\Sigma}(w, \parse)$, iff
+  A word $w \in L(\T, T, \parse)$ is resolvably ambiguous, written $\rho_L(w)$, iff
 
@@
 
-For clarity we abbreviate $\rho_{\T}(w, \parse)$ as $\rho(w)$ if $\T$
-and $\parse$ are clear from the context.
-
 The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way.
@@
 
-  $\forall w \in L(\T, T, \parse).\ \rho_{\T}(w, \parse)$.
+  $\forall w \in L(\T, T, \parse).\ \rho_L(w)$.
 \end{definition}
@@
 \begin{description}
-\item[Static resolvability.] Given a language defined by $\parse$, determine if it is resolvably ambiguous.
-\item[Dynamic resolvability.] Given a language defined by $\parse$ and a word $w$ such that $\parse(w) \neq \emptyset$, determine if $w$ is resolvably ambiguous.
+\item[Static resolvability.] Given a language $L(\T, T, \parse)$, determine whether $\rho(L)$.
+\item[Dynamic resolvability.] Given a word $w \in L(\T, T, \parse)$, determine whether $\rho_L(w)$.
 \end{description}

### 2019-07-05 paper/paper.tex
Case studies introduction

@@
 }
+\lstdefinelanguage{syncon}
+  {morekeywords={syncon,infix,postfix,prefix,left,right,comment,precedence,forbid,except,type,builtin,rec,grouping},
+   sensitive=true,
+   morecomment=[l]{//},
+   morecomment=[s]{/*}{*/},
+   morecomment=[s][<.][.>],
+   morestring=[b]",
+  }
 \newcommand{\ocaml}{\lstinline[language={[objective]caml}]}
+\newcommand{\syncon}{\lstinline[language=syncon]}
 
@@
 
+We have implemented a tool and a small DSL for syntax definition that generates a language definition in the style of Section~\ref{sec:parse-time-disambiguation}, which the tool then uses to construct a parser and do a dynamic resolvability analysis.
+The tool is implemented in Haskell and uses an off-the-shelf implementation\footnote{\url{http://hackage.haskell.org/package/Earley}} of the Earley parsing algorithm \cite{earleyEfficientContextfreeParsing1970}.
+The purpose of these case studies is to showcase possible use cases of resolvable ambiguity in the context of a pre-existing general purpose programming language (OCaml, Section~\ref{sec:evaluation-ocaml}), a pre-existing DSL (ORC, Section~\ref{sec:evaluation-orc}), and a new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).
+
+The syntax definition DSL is a continuation of the ``syncons'' introduced by \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each production is defined separately from each other (one \syncon{syncon} each), and marks are primarily introduced through precedence lists.
+For example, the running example used in Section~\ref{sec:parse-time-disambiguation} (Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}) can be written as follows:
+
+\begin{tabular}{ll}
+\begin{lstlisting}[language=syncon,boxpos=t]
+type Exp
+grouping "(" Exp ")"
+precedence {
+  mul; // higher in precedence list
+  add; // means higher precedence
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[language=syncon,boxpos=t]
+syncon list: Exp =
+  "[" (head:Exp (";" tail:Exp))? "]"
+infix add: Exp = "+"
+infix mul: Exp = "*"
+syncon literal: Exp = n:Integer
+\end{lstlisting}
+\end{tabular}
+
+\noindent The design of the DSL diverges from the formalism in Section~\ref{sec:parse-time-disambiguation} partially for familiarity (precedence instead of marks), but also for composability reasons: the DSL allows marks to be introduced without altering the original definition of a syncon.
+If marks are needed in non-precedence related situations, e.g., adding sequential composition to the running example, they can be introduced with ``forbids'':
+
+\begin{lstlisting}[language=syncon]
+infix seqComp: Exp = ";"
+forbid list.head = seqComp
+forbid list.tail = seqComp
+\end{lstlisting}
+
 \subsection{OCaml} \label{sec:evaluation-ocaml}

### 2019-07-07 paper/paper.tex
EOD

@@
 \newcommand{\regex}{\mathit{Reg}}
-\newcommand{\reqpl}{(}
+\newcommand{\reqpl}{\underline{(}}
 \newcommand{\reqpr}{)}
 \newcommand{\reqp}[1]{\reqpl#1\reqpr}
-\newcommand{\pospl}{[}
-\newcommand{\pospr}{]}
+\newcommand{\pospl}{(}
+\newcommand{\pospr}{)}
 \newcommand{\posp}[1]{\pospl#1\pospr}
@@
 \begin{theorem}
-  There is a bijection between successful runs in $A_{()}$ and trees $t \in L(T_D)$.
+  There is a bijection between successful runs in $A_{\posp{}}$ and trees $t \in L(T_D)$.
 \end{theorem}
@@
 \begin{proof}
-  $A_{()}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
+  $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
 The two are distinct, and $f$ is reachable only through transitions that pop a stack symbol with $s$ and $f$ as the two first components.
@@
   Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, g, w)} (p', (p, q, g, w) \cdot s) \; R_2 \; (q', (p, q, g, w) \cdot s \xrightarrow{b, -(p, q, g, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
-The only transitions that leave the stack unchanged in $A_{()}$ come from DFAs generated from productions in $D$.
+The only transitions that leave the stack unchanged in $A_{\posp{}}$ come from DFAs generated from productions in $D$.
 There is no way to transition between two such DFAs without using a pushing or popping transition, thus $R_2$ must be a successful run in the DFA for some labelled production $N -> l : r$ in $D$.
@@
 
-  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{()}$ starting from $(s, \epsilon)$.
+  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{\posp{}}$ starting from $(s, \epsilon)$.
 If the next terminal is in $\T$, i.e., a strictly zero-arity terminal, we take a transition originating in a DFA, which must be unique.
@@
 
-Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$, construct a VPDA $(Q, \T', \Gamma, \delta, s, \{f\})$ where:
+Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$, construct a VPDA $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
 
@@
 \item $Q = \{s, f\} \cup \bigcup_l Q_l$.
-\item $\T' = \T \cup \{\reqpl, \reqpr, \pospl, \pospr\}$.
-\item $\Gamma = Q \times Q \times \{\reqp{}, \posp{}\} \times \Labels^{*}$.
+\item $\T' = \T \cup \{\reqpl, \pospl, \pospr\}$.
+\item $\Gamma = Q \times Q \times \Labels^{*}$.
 \item $\delta$ is defined by:
@@
     \delta(p, \text{``}\reqpl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, \reqp{}, w))
+    & \{ (s_{l_3}, (p, q, w))
     & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, r_1 \lor r_2 \} \\
@@
     \delta(p, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, \posp{}, w))
+    & \{ (s_{l_3}, (p, q, w))
     & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, \neg (r_1 \lor r_2) \} \\
 
-    \delta(f_{l_3}, \text{``}\reqpr\text{''}, (p, q, \reqp{}, w))
+    \delta(f_{l_3}, \text{``}\reqpr\text{''}, (p, q, w))
     & \{ (q, \lambda)
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, r_1 \lor r_2, f_{l_3} \in F_{l_3} \} \\
-
-    \delta(f_{l_3}, \text{``}\pospr\text{''}, (p, q, \posp{}, w))
-    & \{ (q, \lambda)
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, \neg (r_1 \lor r_2), f_{l_3} \in F_{l_3} \} \\
+    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, f_{l_3} \in F_{l_3} \} \\
+
+    \delta(s, \text{``}\reqpl\text{''}, \lambda)
+    & \{ (s_{l_2}, (s, f, w))
+    & \nt(l_1) = S, (l_1, l_2, \top, w) \in P \} \\
+
+    \delta(s, \text{``}\pospl\text{''}, \lambda)
+    & \{ (s_{l_2}, (s, f, w))
+    & \nt(l_1) = S, (l_1, l_2, \bot, w) \in P \} \\
+
+    \delta(f_{l_2}, \text{``}\reqpr\text{''}, (s, f, w))
+    & \{ (f, \lambda)
+    & nt(l_1) = S, (l_1, l_2, r, w) \in P, f_{l_2} \in F_{l_2}\} \\
   \end{array}
@@
 \end{itemize}
+
+\noindent We also construct a modified VPDA $A'_{\posp{}} = (Q, \T', \Gamma \cup \{\gamma\}, \delta', s, \{f\})$ where:
+
+\begin{itemize}
+\item $\gamma$ is a new distinct stack symbol.
+\item $\delta'(p, a, g) = \delta(p, a, g) \cup \delta''(p, a, g)$, where $\delta''$ is given by:
+  $$
+  \begin{array}{r@{\;=\;}l@{}}
+    \delta''(p, \text{``}\pospl\text{''}, \lambda)
+    & \{ (p, \gamma) \} \\
+
+    \delta''(p, \text{``}\pospr\text{''}, \gamma)
+    & \{ (p, \lambda) \} \\
+
+    \delta''(p, \text{``}\pospl\text{''}, \lambda)
+    & \delta(p, \text{``}\reqpl\text{''}, \lambda) \\
+  \end{array}
+  $$
+\end{itemize}
+
+\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is given in Section~\ref{sec:preliminaries-vpls} (using the partitions $\T_c = \{ \reqpl, \pospl \}$, $\T_i = \T$, and $\T_r = \{\reqpr\}$).
+If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$, then that run corresponds to two distinct runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
+We check for the existence of such a run by trimming the product automaton and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
+
+Performing this procedure directly on a language definition $D$ gives a sound decision procedure where the presence of two distinct runs implies unresolvably ambiguous.
+If we instead do it on a modified $D$ with no marks on the right hand side of any production (i.e., all non-terminals have the form $N_\emptyset$) we get a sound decision procedure where the absence of two distinct runs implies resolvably ambiguous.
+
+There are thus cases where we can answer resolvably ambiguous or unresolvably ambiguous with certainty, and cases where we cannot answer either way with certainty.
 
@@
 
+For this section, we assume the presence of a single language definition $D$ with terminals in $\T$, labels in $\Labels$, and non-terminals in $\NT$.
+
 \begin{theorem}
@@
 \end{theorem}
+
+\noindent We construct this automaton in a bottom-up fashion.
+A sequence of terminals in $\T$ is easily recognized by a sequence of transitions that do not interact with the stack.
+Each subtree allows surrounding grouping parentheses, which can be represented by transitions that push and pop stack symbols (one unique stack symbol per subtree), to ensure that they are balanced.
+With that, there are two remaining complications: a subtree $l(w)$ recognized from a marked non-terminal $N_m$ where $l \in m$ must have one or more surrounding parentheses instead of zero or more, and we must maintain the visibly pushdown property, i.e., all transitions labelled ``$($'' or ``$)$'' must push or pop stack symbols, respectively.
+The former can be solved by either tracking the subtree's origin when parsing, or reparsing with the right hand side regular expression, while the latter is solved by introducing a new stack symbol $\gamma$ that is pushed and popped by all non-grouping parenthesis terminals.
+Writing the transition function $\delta : Q \times \T \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ as a subset of $Q \times \T \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, and using $\sqcup$ for disjoint union:
+
+$$
+\begin{array}{r@{\;}l}
+  f(a)
+  & = \begin{cases}
+    (\{p, q\}, \T, \emptyset, \{(p, a, \lambda, q, \lambda)\}, p, \{q\})
+    & \text{if } a \notin \{\text{``}(\text{''}, \text{``})\text{''}\} \\
+
+    (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``}(\text{''}, \lambda, q, \gamma)\}, p, \{q\})
+    & \text{if } a = \text{``}(\text{''} \\
+
+    (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``})\text{''}, \lambda, q, \lambda))\}, p, \{q\})
+    & \text{if } $a = \text{``})\text{''}$ \\
+  \end{cases} \\
+
+  f(w_1 \cdot w_2)
+  & = (Q_1 \sqcup Q_2, \T, \Gamma', \delta_1 \cup \delta_2 \cup \{(f_1, \lambda, \lambda, s_2, \lambda)\}, s_1, \{f_2\}) \\
+  & \text{ where } f(w_i) = (Q_i, \T, \Gamma_i, \delta_i, s_i, \{f_i\}) \\
+  & \hphantom{\text{ where }} \Gamma' = (\Gamma_1 \setminus \{\gamma\}) \sqcup (\Gamma_2 \setminus \{\gamma\}) \cup \{\gamma\} \\
+
+  f(l(w))
+  & = \begin{cases}
+    (Q \sqcup \{p, q\}, \T, \Gamma', \delta \cup \delta_1, p, \{q\}) & \text{if $l$ is marked in the introducing non-terminal.} \\
+    (Q, \T, \Gamma', \delta \cup \delta_2, s, \{f\}) & \text{if $l$ is unmarked in the introducing non-terminal.} \\
+  \end{cases} \\
+  & \text{ where } f(w) = (Q, \T, \Gamma, \delta, s, \{f\}) \\
+  & \hphantom{\text{ where }} \Gamma' = (\Gamma \setminus \{\gamma\}) \sqcup \{g, \gamma\} \\
+  & \hphantom{\text{ where }} \delta_1 = \{(p, \text{``}(\text{''}, \lambda, p, g), (p, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, q, \lambda), (q, \text{``})\text{''}, g, q, \lambda)\} \\
+  & \hphantom{\text{ where }} \delta_2 = \{(s, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, f, \lambda)\} \\
+\end{array}
+$$
 

### 2019-07-07 paper/paper.tex
First lines of the abstract.

@@
 \begin{abstract}
-Text of abstract \ldots.
+% Why?
+
+A traditional view when designing programming languages is that syntax
+definitions should be unambiguous: only unambiguous grammars are used to construct
+language specific parsers that either rejects a program, or accepts it and
+produces a unique parse tree. However,
+
+% Why composing languages
+-
+
+% Problem we address
+Resolvable ambiguity
+Two sub-problems. Static and dynamic resolvability
+
+% Our solution
+Prove static resolvable for
+
+% Three case studies
+
 \end{abstract}
@@
 %% comma separated list
-\keywords{keyword1, keyword2, keyword3}  %% \keywords are mandatory in final camera-ready submission
+\keywords{Language Composition, Context Free Grammars, Domain-Specific Languages}  %% \keywords are mandatory in final camera-ready submission
 

### 2019-07-07 paper/paper.tex
Updated the abstract.

--- 2019-07-07 paper/paper.tex
@@
 \begin{abstract}
-Text of abstract \ldots.
+% Why?
+
+Ever since the early 60s, the problem of deciding if a context-free
+grammar is unambiguous or not has been known to be undecidable. As a
+consequence, a large number of restricted grammars have been developed
+to guarantee that a language definition is unambiguous. This
+traditional view of only allowing unambiguous grammars has until
+recently been taken for granted as the only truth: \emph{the} way of
+how the syntax of a language must be defined. However, up front
+unambiguous grammars can force language designers to make arbitrary
+choice to disambiguate the language, although the natural choice may be to
+postpone the decision to the programmer. Moreover, an important
+objective of domain-specific language development is to construct
+languages by composing and extending existing
+languages. Unfortunately, the composition of restricted unambiguous
+grammars easily results in ambiguous grammars, which again are
+undecidable. In this paper, we depart from the traditional view of
+unambiguous grammar design, and allow ambiguities to be delayed until
+compile time, allowing the user to perform the disambiguation. A
+natural decision problems follows: given a language definition, can a
+user always disambiguate an ambiguous program?  We introduce and
+formalize this fundamental problem---called the \emph{resolvable
+  ambiguity problem}---and divide it into separate static and dynamic
+resolvability problems. We prove soundness and completeness for a
+restricted language in the static case, and soundness in the general
+case for dynamic resolvability. The approach is evaluated through
+three separate case studies, covering both a large
+existing programming language, and the composability of
+domain-specific languages.
+%A traditional view when designing programming languages is that syntax
+%definitions should be unambiguous: only unambiguous grammars are used
+%to construct language specific parsers that either rejects a program,
+%or accepts it and produces a unique parse tree. However,
 \end{abstract}

### 2019-07-08 paper/paper.tex
Introduce the notion of an abstract parser

@@
 
-This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to some more standard concepts in formal languages.
-
-Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense.
-We call this meaning an \emph{interpretation} of a word.
-Note that a single word may have multiple interpretations.
-As such, we define an \emph{interpreted language} as:
+This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to standard concepts in formal languages.
+
+A formal language is defined as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$.
+For defining \emph{resolvable ambiguity} we additionally have to consider the results of parsing words.
+In order to stay as general as possible, we first define the notion of an \emph{abstract parser}.
 
 \begin{definition}
-  An interpreted language, written $L(\T, T, \parse)$, consists of
+  An abstract parser $P$ is a triple $(\T, T, \parse)$ consisting of
 \begin{itemize}
 \item an alphabet $\T$,
-\item a set of interpretations $T$, and
-\item a function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
-\end{itemize}
-
-  A word $w \in \T^{*}$ is in the language $L(\T, T, \parse)$, written
-  $w \in L(\T, T, \parse)$ iff $\parse(w) \neq \emptyset$.
+\item a set of parse trees $T$, and
+\item a function $\parse : \T^{*} -> 2^T$ that relates words to their parse trees, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
+\end{itemize}
 \end{definition}
 
-\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees.
-In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs shown in Fig.~\ref{fig:arith-example}.
+This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}
+
+\begin{definition}
+  Given a word $w \in \T^{*}$ and an abstract parser $P = (\T, T,
+  \parse)$, the set of words contained in the \emph{parse language}
+  $L(P)$ is defined as follows:
+
+  $w \in L(P)$ iff $\parse(w) \neq \emptyset$.
+\end{definition}
+
+\noindent For example, consider a simple arithmetic language without precedence and parentheses.
+In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two parse trees shown in Fig.~\ref{fig:arith-example}.
 
@@
 \end{tabular}
-\caption{The two allowable interpretations of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
+\caption{The two allowable parse trees of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
 \label{fig:arith-example}
@@
 \begin{definition}
-  A word $w \in L(\T, T, \parse)$ is ambiguous, written $\amb_L(w)$, iff
+  Given an abstract parser $P = (\T, T, \parse)$, a word $w \in L(P)$
+  is ambiguous, written $\amb_P(w)$, iff
 
@@
 \noindent
-Note that the above definition implies that a word $w \in L(\T, T,
-\parse)$ is not ambiguous (or unambiguous) if $\exists t \in
-T.\ \parse(w) = \{t\}$.
-
-We can connect the above definition of an interpreted language to the
-classical definition of a language as follows:
-
-\begin{itemize}
-\item Given an interpreted language $L(\T, T, \parse)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
+Note that the above definition implies that a word $w \in L(P)$, where
+$P = (\T, T, \parse)$, is not ambiguous, or \emph{unambiguous}, if
+$\exists t \in T.\ \parse(w) = \{t\}$.
+
+We can connect the above definition of a parse language to the
+classical definition of a (formal) language as follows:
+
+\begin{itemize}
+\item Given a parse language $L(P)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
 \item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
@@
 
-\noindent A \emph{resolvably ambiguous} word is a word where all its interpretations can be written in an unambiguous way, formally:
+\noindent A \emph{resolvably ambiguous} word is a word where all its parse trees can be written in an unambiguous way, formally:
 
 \begin{definition}\label{def:resolvable-word}
-  A word $w \in L(\T, T, \parse)$ is resolvably ambiguous, written $\rho_L(w)$, iff
+  Given an abstract parser $P = (\T, T, \parse)$, a word $w \in L(P)$ is \emph{resolvably ambiguous}, written $\rho_P(w)$, iff
 
@@
 The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way.
-However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.
-
-Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, formally:
+However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously parsed as the first and second tree, respectively.
+
+Additionally, we define an \emph{abstract parser} to be resolvably ambiguous if all words in its corresponding parse language $L(P)$ are resolvably ambiguous, formally:
 
 \begin{definition}\label{def:resolvable-language}
-  An interpreted language $L(\T, T, \parse)$ is resolvably ambiguous, written $\rho(L)$, iff
-
-  $\forall w \in L(\T, T, \parse).\ \rho_L(w)$.
+  An abstract parser $P$ is resolvably ambiguous, written $\rho(P)$, iff
+
+  $\forall w \in L(P).\ \rho_P(w)$.
 \end{definition}
@@
 \begin{itemize}
-\item An unambigous word $w$ is trivially resolvably ambiguous, since its only interpretation $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$).
+\item An unambigous word $w$ is trivially resolvably ambiguous, since its only parse tree $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$).
 The set of resolvably ambiguous words is thus a superset of the unambiguous words.
-\item If a given interpretation $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous.
+\item If a given parse tree $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous.
 In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
@@
 \noindent The second point suggests that resolvable ambiguity is only an interesting property if an element of $T$ does not uniquely identify an element of $\T^{*}$.
-Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual interpretation.
+Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual parse tree.
 Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace.
-In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the interpretation.
+In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the parse tree.
 
@@
 \begin{description}
-\item[Static resolvability.] Given a language $L(\T, T, \parse)$, determine whether $\rho(L)$.
-\item[Dynamic resolvability.] Given a word $w \in L(\T, T, \parse)$, determine whether $\rho_L(w)$.
+\item[Static resolvability.] Given an abstract parser $P$, determine whether $\rho(P)$.
+\item[Dynamic resolvability.] Given a word $w \in L(P)$ for some abstract parser $P$, determine whether $\rho_P(w)$.
 \end{description}

### 2019-07-08 paper/paper.tex
Completeness of static resolvability
Filled some paragraphs

@@
 
-\noindent Our main concern is producing decision procedures for these problems. As such, we define soundness and completeness for the static problem:
-
-\begin{definition}
-  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the static resolvability problem is:
-  \begin{itemize}
-  \item Sound if $f(\parse) = \mathit{resolvable}$ implies that the language given by $\parse$ is resolvably ambiguous, and $f(\parse) = \mathit{unresolvable}$ implies that the language given by $\parse$ is not resolvably ambiguous.
-  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$.
-  \end{itemize}
-  \label{def:static-procedure}
+\noindent Our main concern is producing decision procedures for these problems. First, we define soundness and completeness for the static problem.
+
+% todo: improve label (where used?)
+\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure}
+  A decision procedure $f$ solving the static resolvability problem is sound iff
+
+  $f(P) \implies \rho(P) \text{ for all abstract parsers } P$
+\end{definition}
+
+In practice, we may additionally be interested in a decision procedure
+$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
+  parsers } P$. It turns out, we can give $g$ for the parse languages
+studied in following sections.
+
+\begin{definition}[Completeness of Static Resolvability]
+  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff
+
+  $\rho(P) \implies f(P) \text{ for all abstract parsers } P$
 \end{definition}
@@
 If we thus delimit ourselves to only consider languages where words have no unbalanced parentheses\footnote{I.e., the vast, vast majority of programming languages currently in use.} we can give the following alternative definition of a word: a word is a two-tuple containing a sequence of non-parenthesis terminals and a bag (or multiset) of ranges covered by parentheses.
-For example, the word ``$(1 + 2) * 3$'' is equivalent to $(\text{``}1 + 2 * 3\text{''}, \{\range{0}{3}\})$, while ``$((1 + 2)) * 3$'' is equivalent to $(\text{``}1 + 2 * 3\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
+For example, the word ``\verb|(1 + 2) * 3|'' is equivalent to $(\text{``}\verb|1 + 2 * 3|\text{''}, \{\range{0}{3}\})$, while ``\verb|((1 + 2)) * 3|'' is equivalent to $(\text{``}\verb|1 + 2 * 3|\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
 We will refer to the first component of the tuple as the \emph{basic word} and the second as the \emph{range bag}.
@@
 However, there is a case where this alternate representation is not fully isomorphic; namely when a pair of parentheses would yield a zero-length range.
-As an example, consider the two words ``$(())$'' and ``$()()$''.
+As an example, consider the two words ``\verb|(())|'' and ``\verb|()()|''.
 Both of these have an alternate representation of $(\text{``}\text{''}, \{\range{0}{0}, \range{0}{0}\})$.
@@
 
-We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., $\basic(\text{``}((1 + 2)) * 3\text{''}) = \text{``}1 + 2 * 3\text{''}$ and $\rangebag(\text{``}((1 + 2)) * 3\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
+We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., \\$\basic(\text{``}\verb|((1 + 2)) * 3|\text{''}) = \text{``}\verb|1 + 2 * 3|\text{''}$ and $\rangebag(\text{``}\verb|((1 + 2)) * 3|\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
 
@@
 \end{lemma}
-\noindent For example, removing the parentheses in ``$(1 + 2) * 3$'' changes the interpretations produced.
+\noindent For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the interpretations produced.
 Each required range is a direct consequence of a mark on a non-terminal in $D$.
@@
 Duplicated grouping parentheses correspond to nested grouping nodes, all of which are removed by $\semantic$, thus producing identical sets of trees in $L(T_D)$.
-For example, $((1 + 2)) * 3$ has the same interpretations as $(1 + 2) * 3$.
+For example, \verb|((1 + 2)) * 3| has the same interpretations as \verb|(1 + 2) * 3|.
 
@@
 
-To show the connection between resolvable ambiguity and these lattices, consider the word ``$1 + 2 + 3$''.
-It has two interpretations in our running example, $a(a(n(1)+n(2))+n(3))$, which we would normally write as $(1 + 2) + 3$, and $a(n(1)+a(n(2)+n(3)))$, which we would normally write $1 + (2 + 3)$.
+To show the connection between resolvable ambiguity and these lattices, consider the word ``\verb|1 + 2 + 3|''.
+It has two interpretations in our running example, $a(a(n(1)+n(2))+n(3))$, which we would normally write as \verb|(1 + 2) + 3|, and $a(n(1)+a(n(2)+n(3)))$, which we would normally write \verb|1 + (2 + 3)|.
 The lattices for these two trees are given in Figures~\ref{fig:lattice1} and \ref{fig:lattice2} respectively.
@@
 We now make this explicit, to ensure that standard operations on finite automata produce the correct result.
-We replace each $N_m$ with a sum of the labels of productions in $N$, where each label in $m$ is distinguished by a prime, e.g., $E_{\{s\}}$ is replaced with $s' + v$.
+We replace each $N_m$ with a sum of the labels of productions in $N$, where each label has a subscript $\top$ or $\bot$ if parentheses are required or possible, respectively.
+For example, $E_{\{s\}}$ is replaced with $s_\top + v_\bot$.
 
@@
   \toprule
-  $E$ & $s$ & $(s' + v)$ \verb|';'| $(s + v)$ \\
+  $E$ & $s$ & $(s_\top + v_\bot)$ \verb|';'| $(s_\bot + v_\bot)$ \\
   $E$ & $v$ & \verb|'e'| \\
@@
 
-  \path (1) edge[bend left] node {$s'$} (2)
-  (1) edge[bend right] node {$v$} (2)
+  \path (1) edge[bend left] node {$s_\top$} (2)
+  (1) edge[bend right] node {$v_\bot$} (2)
   (2) edge node {$';'$} (3)
-  (3) edge[bend left] node {$s$} (4)
-  (3) edge[bend right] node {$v$} (4);
+  (3) edge[bend left] node {$s_\bot$} (4)
+  (3) edge[bend right] node {$v_\bot$} (4);
   \path (5) edge node {$'e'$} (6);
@@
 Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we go back.
-We use $Q \times Q \times \{\text{``}\reqp{}\text{''}, \text{``}\posp{}\text{''}\}$ as our stack alphabet, where the first two components are states in the parent production automaton and the last component records whether the parenthesis pair is required or merely possible.
-We replace every edge $p \xrightarrow{l} q$, where $l \in \Labels$, with:
+We use $Q \times Q$ as our stack alphabet, to record the transition we are replacing in the parent.
+We thus replace every edge $p \xrightarrow{l_\top} q$, where $l \in \Labels$, with:
 
 \begin{itemize}
-  \item an edge $p \xrightarrow{'\pospl', +(p, q, \text{``}\posp{}\text{''})} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
-  \item an edge $q' \xrightarrow{'\pospr', -(p, q, \text{``}\posp{}\text{''})} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
+  \item an edge $p \xrightarrow{'\reqpl', +(p, q)} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
+  \item an edge $q' \xrightarrow{'\reqpr', -(p, q)} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
 \end{itemize}
 
-\noindent Similarly, for every edge $p \xrightarrow{l'} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\reqpl', +(p, q, \text{``}\reqp{}\text{''})} p'$ and $q' \xrightarrow{'\reqpr', -(p, q, \text{``}\reqp{}\text{''})} q$.
-
-Intuitively, we parse a child node with parentheses around it, then return.
-This is where our simplification is used, without it we might introduce double parentheses here.
-To lessen clutter in the figure below we omit the third component of the stack symbol, since it is fully determined by the input symbol, e.g., if an edge reads $'\pospl'$ then the third component of the stack symbol pushed must be ``$\posp{}$''.
+\noindent Similarly, for every edge $p \xrightarrow{l_\bot} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\pospl', +(p, q)} p'$ and $q' \xrightarrow{'\pospr', -(p, q)} q$.
+
+Intuitively, we parse a child node with parentheses around it, then return. This is where our simplification is used, without it we might introduce double parentheses here.
 
@@
 \begin{proof}
+  (These are two attempts at showing the same thing, next attempt starts at ``Proof sketch''. I would strongly suggest reading ``Formal stuff'' first.)
   $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
@@
 
-  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, g, w)} (p', (p, q, g, w) \cdot s) \; R_2 \; (q', (p, q, g, w) \cdot s \xrightarrow{b, -(p, q, g, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
+  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, w)} (p', (p, q, w) \cdot s) \; R_2 \; (q', (p, q, w) \cdot s \xrightarrow{b, -(p, q, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
 The only transitions that leave the stack unchanged in $A_{\posp{}}$ come from DFAs generated from productions in $D$.
@@
 If the next terminal is in $\Labels$, i.e., an unranked terminal, we take one of the added edges.
-However, this choice of edge might not be unique, there may be multiple edges from the current state to an appropriate DFA initial state.
-These stem from the original DFA having two transitions from the same state with the same label, except one is primed, e.g., $p \xrightarrow{l} q$ and $p \xrightarrow{l'} q'$.
-However, if both of these transitions were valid choices that would imply that the DFA for that level can recognize two distinct words that differ only in whether some terminals are primed.
-But primed and non-primed terminals must originate from distinct occurances of marked non-terminals (since a primed terminal corresponds to a mark, and an unprimed terminal corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
+However, this choice of edge might not be unique: there may be multiple edges from the current state to an appropriate DFA initial state.
+These stem from the original DFA having two transitions from the same state with the same label but different subscripts, e.g., $p \xrightarrow{l_\bot} q$ and $p \xrightarrow{l_\top} q'$.
+However, if both of these transitions were valid choices for the current tree that would imply that the DFA for that level can recognize two distinct words that differ only in subscripts.
+But distinctly subscripted terminals must originate from distinct occurances of marked non-terminals (since a $\top$ subscript corresponds to a mark, and a $\bot$ subscript corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
 
@@
 Every pushing transition can thus be paired with a popping transition in the run.
-We can construct a tree in a bottom-up fashion by replacing every such pair with a tree $l_1(l_2(\ldots l_n(x)\ldots))$, where $x$ is the sequence of terminals and/or trees between the pair, the terminals $l_1, l_2, \ldots, l_{n-1}$ stem from the fourth component of the pushed stack symbol ($l_1 \cdots l_{n-1}$), and $l_n$ is the label of the production whose DFA the pushing transition ends in.
+We can construct a tree in a bottom-up fashion by replacing every such pair with a tree $l_1(l_2(\ldots l_n(x)\ldots))$, where $x$ is the sequence of terminals and/or trees between the pair, the terminals $l_1, l_2, \ldots, l_{n-1}$ stem from the third component of the pushed stack symbol ($l_1 \cdots l_{n-1}$), and $l_n$ is the label of the production whose DFA the pushing transition ends in.
 
@@
 Terminals in $\T$ cause transitions internally in a DFA, while terminals in $\Labels$ cause a pushing transition (and later a corresponding, unique popping transition) to the initial state of the DFA corresponding to the production with that label.
-Note that a sequence of terminals in $\Labels$ that have only a single child (which is also in $\Labels$) cause only one transition, all terminals but the last end up in the fourth component of the pushed stack symbol.
-The choice of pushing transition is \emph{not} unique however; there may be multiple available transitions that differ in the recognized input symbol and the second and third components of the pushed stack symbol, none of which are necessarily determined by the shape of the tree.
-This non-determinism does not introduce multiple successful runs however, since two such runs would imply the existence of two distinct runs through the DFA corresponding to some labelled production that recognize the same word, except for whether some terminals are primed.
-But two otherwise equal terminals where one is primed must originate from different occurrences of marked non-terminals in the original regular expression (since a primed terminal represents a mark, and an unprimed terminal represents the absence of a mark), thus that regular expression must be ambiguous after all marks are removed.
+Note that a sequence of terminals in $\Labels$ that have only a single child (which is also in $\Labels$) cause only one transition, all terminals but the last end up in the third component of the pushed stack symbol.
+The choice of pushing transition is \emph{not} unique however; there may be multiple available transitions that differ in the recognized input symbol and the second component of the pushed stack symbol, neither of which are necessarily determined by the shape of the tree.
+This non-determinism does not introduce multiple successful runs however, since two such runs would imply the existence of two distinct runs through the DFA corresponding to some labelled production that recognize the same word, except with different subscripts.
+But two otherwise equal terminals with different subscripts must originate from different occurrences of marked non-terminals in the original regular expression (since a $\top$ subscript represents a mark, and a $\bot$ subscript represents the absence of a mark), thus that regular expression must be ambiguous after all marks are removed.
 But Section~\ref{sec:parse-time-disambiguation} explicitly ruled this out, thus only one of the pushing transitions can be a valid choice.
@@
 \item[$\dfa(l)$:] Given the production $N -> l : r$, compute the regular language $L' := L(r) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
-Then replace every marked non-terminal $N'_m$ with the regular expression $(l_1, b_1) + (l_2, b_2) + \ldots + (l_n, b_n)$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, and $b_i = l_i \in m$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l + a' + m + n$.
+Then replace every marked non-terminal $N'_m$ with the regular expression $l_{1l_1 \in m} + l_{2l_2 \in m} + \ldots + l_{nl_n \in m}$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l_\bot + a_\top + m_\bot + n_\bot$.
 Then construct a DFA for this language.
@@
 \noindent $P$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
-Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for that tree are unresolvably ambiguous.
-
-Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$, construct a VPDA $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
-
-% TODO: add transitions to/from final/initial states
+Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
+
+Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$ for all $l \in Labels$, and $\nt(l) = N$ iff $(N -> l : r) \in D$, construct a VPDA $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
 
@@
 \item $s$ and $f$ are two new distinct states.
-\item $Q = \{s, f\} \cup \bigcup_l Q_l$.
+\item $Q = \{s, f\} \cup \bigsqcup_l Q_l$.
 \item $\T' = \T \cup \{\reqpl, \pospl, \pospr\}$.
 \item $\Gamma = Q \times Q \times \Labels^{*}$.
-\item $\delta$ is defined by:
+\item $\delta$ is defined by (in hindsight this should probably be written as constructing a set $\delta$, i.e., instead of a function $\delta: Q \times \T' \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ we have a set $\delta \subseteq Q \times \T' \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, would probably be easier to read):
   $$
@@
 
-We have implemented a tool and a small DSL for syntax definition that generates a language definition in the style of Section~\ref{sec:parse-time-disambiguation}, which the tool then uses to construct a parser and do a dynamic resolvability analysis.
-The tool is implemented in Haskell and uses an off-the-shelf implementation\footnote{\url{http://hackage.haskell.org/package/Earley}} of the Earley parsing algorithm \cite{earleyEfficientContextfreeParsing1970}.
-The purpose of these case studies is to showcase possible use cases of resolvable ambiguity in the context of a pre-existing general purpose programming language (OCaml, Section~\ref{sec:evaluation-ocaml}), a pre-existing DSL (ORC, Section~\ref{sec:evaluation-orc}), and a new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).
-
-The syntax definition DSL is a continuation of the ``syncons'' introduced by \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each production is defined separately from each other (one \syncon{syncon} each), and marks are primarily introduced through precedence lists.
-For example, the running example used in Section~\ref{sec:parse-time-disambiguation} (Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}) can be written as follows:
+We have implemented a tool and a small DSL for syntax definition
+that generates a language definition in the style of
+Section~\ref{sec:parse-time-disambiguation}, which the tool then
+uses to construct a parser and do a dynamic resolvability
+analysis. The tool is implemented in Haskell and uses an
+off-the-shelf
+implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
+of the Earley parsing algorithm
+\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
+case studies is to showcase possible use cases of resolvable
+ambiguity in the context of a pre-existing general purpose
+programming language (OCaml, Section~\ref{sec:evaluation-ocaml}),
+a pre-existing DSL (Orc, Section~\ref{sec:evaluation-orc}), and a
+new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).
+
+The syntax definition DSL is a continuation of the ``syncons''
+introduced by
+\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
+production is defined separately from each other (one
+\syncon{syncon} each), and marks are primarily introduced through
+precedence lists. For example, the running example used in
+Section~\ref{sec:parse-time-disambiguation}
+(Fig.~\ref{fig:running-example-definition} on
+page~\pageref{fig:running-example-definition}) can be written as
+follows:
 
@@
 
-\noindent The design of the DSL diverges from the formalism in Section~\ref{sec:parse-time-disambiguation} partially for familiarity (precedence instead of marks), but also for composability reasons: the DSL allows marks to be introduced without altering the original definition of a syncon.
-If marks are needed in non-precedence related situations, e.g., adding sequential composition to the running example, they can be introduced with ``forbids'':
+\noindent The design of the DSL diverges from the formalism in
+Section~\ref{sec:parse-time-disambiguation} partially for
+familiarity (precedence instead of marks), but also for
+composability reasons: the DSL allows marks to be introduced
+without altering the original definition of a syncon. If marks are
+needed in non-precedence related situations, e.g., adding
+sequential composition to the running example, they can be
+introduced with ``forbids'':
 

### 2019-07-08 paper/paper.tex
Consistent syntax pass through section 6

--- 2019-07-08 paper/paper.tex
@@
 \item[$\dfa(l)$:] Given the production $N -> l : r$, compute the regular language $L' := L(r) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
-Then replace every marked non-terminal $N'_m$ with the regular expression $l_{1l_1 \in m} + l_{2l_2 \in m} + \ldots + l_{nl_n \in m}$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l_\bot + a_\top + m_\bot + n_\bot$.
+Then replace every marked non-terminal $N'_m$ with the regular expression $l_1_{l_1 \in m} + l_2_{l_2 \in m} + \ldots + l_n_{l_n \in m}$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l_\bot + a_\top + m_\bot + n_\bot$.
 Then construct a DFA for this language.
@@
 
-We have implemented a tool and a small DSL for syntax definition
-that generates a language definition in the style of
-Section~\ref{sec:parse-time-disambiguation}, which the tool then
-uses to construct a parser and do a dynamic resolvability
-analysis. The tool is implemented in Haskell and uses an
-off-the-shelf
-implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
-of the Earley parsing algorithm
-\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
-case studies is to showcase possible use cases of resolvable
-ambiguity in the context of a pre-existing general purpose
-programming language (OCaml, Section~\ref{sec:evaluation-ocaml}),
-a pre-existing DSL (Orc, Section~\ref{sec:evaluation-orc}), and a
-new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).
-
-The syntax definition DSL is a continuation of the ``syncons''
-introduced by
-\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
-production is defined separately from each other (one
-\syncon{syncon} each), and marks are primarily introduced through
-precedence lists. For example, the running example used in
-Section~\ref{sec:parse-time-disambiguation}
-(Fig.~\ref{fig:running-example-definition} on
-page~\pageref{fig:running-example-definition}) can be written as
-follows:
+We have implemented a tool and a small DSL for syntax definition that generates a language definition in the style of Section~\ref{sec:parse-time-disambiguation}, which the tool then uses to construct a parser and do a dynamic resolvability analysis.
+The tool is implemented in Haskell and uses an off-the-shelf implementation\footnote{\url{http://hackage.haskell.org/package/Earley}} of the Earley parsing algorithm \cite{earleyEfficientContextfreeParsing1970}.
+The purpose of these case studies is to showcase possible use cases of resolvable ambiguity in the context of a pre-existing general purpose programming language (OCaml, Section~\ref{sec:evaluation-ocaml}), a pre-existing DSL (ORC, Section~\ref{sec:evaluation-orc}), and a new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).
+
+The syntax definition DSL is a continuation of the ``syncons'' introduced by \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each production is defined separately from each other (one \syncon{syncon} each), and marks are primarily introduced through precedence lists.
+For example, the running example used in Section~\ref{sec:parse-time-disambiguation} (Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}) can be written as follows:
 
@@
 
-\noindent The design of the DSL diverges from the formalism in
-Section~\ref{sec:parse-time-disambiguation} partially for
-familiarity (precedence instead of marks), but also for
-composability reasons: the DSL allows marks to be introduced
-without altering the original definition of a syncon. If marks are
-needed in non-precedence related situations, e.g., adding
-sequential composition to the running example, they can be
-introduced with ``forbids'':
+\noindent The design of the DSL diverges from the formalism in Section~\ref{sec:parse-time-disambiguation} partially for familiarity (precedence instead of marks), but also for composability reasons: the DSL allows marks to be introduced without altering the original definition of a syncon.
+If marks are needed in non-precedence related situations, e.g., adding sequential composition to the running example, they can be introduced with ``forbids'':
 

### 2019-07-08 paper/paper.tex
Fix double subscript

@@
 \item[$\dfa(l)$:] Given the production $N -> l : r$, compute the regular language $L' := L(r) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
-Then replace every marked non-terminal $N'_m$ with the regular expression $l_1_{l_1 \in m} + l_2_{l_2 \in m} + \ldots + l_n_{l_n \in m}$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l_\bot + a_\top + m_\bot + n_\bot$.
+Then replace every marked non-terminal $N'_m$ with the regular expression $l_{1l_1 \in m} + l_{2l_2 \in m} + \ldots + l_{nl_n \in m}$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l_\bot + a_\top + m_\bot + n_\bot$.
 Then construct a DFA for this language.

### 2019-07-08 paper/paper.tex
Almost finished first draft of Orc case study

@@
 \lstdefinelanguage{syncon}
-  {morekeywords={syncon,infix,postfix,prefix,left,right,comment,precedence,forbid,except,type,builtin,rec,grouping},
+  {morekeywords={syncon,infix,postfix,prefix,left,right,comment,precedence,forbid,except,type,builtin,rec,grouping,token},
    sensitive=true,
@@
 
+\synctex=1
 \begin{document}
@@
   \end{itemize}
-\item Two case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
+\item Three case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
   \begin{itemize}
   \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
+  \item A domain-specific language for orchestrating parallel computations (Section~\ref{sec:evaluation-orc}).
   \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
@@
 
-We have implemented a tool and a small DSL for syntax definition that generates a language definition in the style of Section~\ref{sec:parse-time-disambiguation}, which the tool then uses to construct a parser and do a dynamic resolvability analysis.
-The tool is implemented in Haskell and uses an off-the-shelf implementation\footnote{\url{http://hackage.haskell.org/package/Earley}} of the Earley parsing algorithm \cite{earleyEfficientContextfreeParsing1970}.
-The purpose of these case studies is to showcase possible use cases of resolvable ambiguity in the context of a pre-existing general purpose programming language (OCaml, Section~\ref{sec:evaluation-ocaml}), a pre-existing DSL (ORC, Section~\ref{sec:evaluation-orc}), and a new DSL (Cyphym, Section~\ref{sec:evaluation-cyphym}).
-
-The syntax definition DSL is a continuation of the ``syncons'' introduced by \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each production is defined separately from each other (one \syncon{syncon} each), and marks are primarily introduced through precedence lists.
-For example, the running example used in Section~\ref{sec:parse-time-disambiguation} (Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}) can be written as follows:
+We have implemented a tool and a small DSL for syntax definition
+that generates a language definition in the style of
+Section~\ref{sec:parse-time-disambiguation}, which the tool then
+uses to construct a parser and do a dynamic resolvability
+analysis. The tool is implemented in Haskell and uses an
+off-the-shelf
+implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
+of the Earley parsing algorithm
+\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
+case studies is to showcase possible use cases of resolvable
+ambiguity in the context of a pre-existing general purpose
+programming language (OCaml, Section~\ref{sec:evaluation-ocaml}),
+a pre-existing DSL (Orc, Section~\ref{sec:evaluation-orc}), and a
+new DSL that is under development (Cyphym, Section~\ref{sec:evaluation-cyphym}).
+
+The syntax definition DSL builds on the the ``syncons''
+introduced by
+\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
+production is defined separately from each other (one
+\syncon{syncon} each), and marks are primarily introduced through
+precedence lists. For example, the running example used in
+Section~\ref{sec:parse-time-disambiguation}
+(Fig.~\ref{fig:running-example-definition} on
+page~\pageref{fig:running-example-definition}) can be written as
+follows:
 
@@
 \begin{lstlisting}[language=syncon,boxpos=t]
+token Integer = "[0-9]+"
 syncon list: Exp =
@@
 \end{lstlisting}
-\end{tabular}
-
-\noindent The design of the DSL diverges from the formalism in Section~\ref{sec:parse-time-disambiguation} partially for familiarity (precedence instead of marks), but also for composability reasons: the DSL allows marks to be introduced without altering the original definition of a syncon.
-If marks are needed in non-precedence related situations, e.g., adding sequential composition to the running example, they can be introduced with ``forbids'':
+\end{tabular}\smallskip
+
+\noindent The design of the DSL diverges from the formalism in
+Section~\ref{sec:parse-time-disambiguation} partially for
+familiarity (precedence instead of marks), but also for
+composability reasons: the DSL allows marks to be introduced
+without altering the original definition of a syncon. If marks are
+needed in non-precedence related situations, e.g., adding
+sequential composition to the running example, they can be
+introduced with ``forbids'':
 
@@
 % TODO: talk about the ``low precedence unary operator/deep precedence'' issue.
+
+\subsection{Orc} \label{sec:evaluation-orc}
+
+Orc~\cite{kitchinOrc2009} is a functional programming language
+which includes a number of special-purpose combinators for
+coordinating concurrent workflows. In this section, we consider
+the use case of adding these concurrency combinators to an already
+defined programming language.
+
+In Orc, every expression can ``publish'' zero or more values. Orc
+defines four combinators for orchestrating these published values:
+the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
+pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
+combinators. The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
+$e_2$ in parallel, publishing any value published by either of
+them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
+value published by $e_1$ to $x$ and executes $e_2$, building a
+concurrent pipeline. The expression $e_1$ \ocaml{<}$x$\ocaml{<}
+$e_2$ binds the \emph{first} value published by $e_2$ to $x$ and
+executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if
+running $e_1$ does not publish any values.
+
+The syncon definition of the Orc combinators is very simple (the
+precedence rules follow the original
+definition~\cite{kitchinOrc2009}):
+
+\begin{tabular}{ll}
+\begin{lstlisting}[language=syncon,boxpos=t]
+type Exp
+grouping "(" Exp ")"
+precedence {
+  seq; par; prune; otherwise;
+}
+\end{lstlisting}
+&
+\begin{lstlisting}[language=syncon,boxpos=t]
+token Ident = "[[:lower:]][[:word:]]*"
+infix par:Exp = "|"
+infix seq:Exp = ">" x:Ident ">"
+infix prune:Exp = "<" x:Ident "<"
+infix otherwise:Exp = ";"
+\end{lstlisting}
+\end{tabular}\smallskip
+
+\noindent
+Na\"{i}vely composing this definition with another, separately
+defined language is likely to introduce ambiguities. For example,
+if that language also contains infix operators, the precedence
+between these and the Orc combinators will be undefined. With
+support for resolvable ambiguity however, we can allow this
+ambiguity and let programmers disambiguate expressions using
+parentheses. After composing the above definition with a simple
+language supporting addition, variables and function calls
+(omitted for brevity), parsing the expression
+%
+``\ocaml{1 + 2 >x> f(x)}''
+%
+results in the following error message from our tool:
+
+\begin{lstlisting}[language={[objective]caml}]
+Ambiguity error with 2 alternatives:
+  ( 1 + 2 ) > x > f ( x )
+  1 + ( 2 > x > f ( x ) )
+\end{lstlisting}
+
+\noindent
+In this case, it is likely that the preferred semantics is to have
+all operators in the base language bind tighter than the Orc
+combinators, and these precedence rules can be added after the
+composition without changing the original definitions.
+
+% TODO: Clashes involving ">" or ";"?
 

### 2019-07-08 paper/paper.tex
Polish text in 6.1

@@
 The one exception is if the start symbol has a nullable production, but that can be trivially special cased.
-As such, going forward, we will assume no zero-length grouping parentheses.
+Thus, we will assume no zero-length grouping parentheses in the following.
 
@@
 The only terminals changed in this process are parentheses, i.e., different $w'$ can only differ in their range bags.
-This also implies that two trees that share a word (i.e., two trees $t_1$ and $t_2$ that have a word $w'$ such that $\parse(w') \supseteq \{t_1, t_2\}$) must also share a basic word.
+This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w'$ such that $\parse(w') \supseteq \{t_1, t_2\}$) must also share a basic word.
 
@@
 \end{lemma}
-\noindent For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the interpretations produced.
+\noindent For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees.
 Each required range is a direct consequence of a mark on a non-terminal in $D$.

### 2019-07-09 paper/paper.tex
Finished first draft of Orc case study

@@
 them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
-value published by $e_1$ to $x$ and executes $e_2$, building a
-concurrent pipeline. The expression $e_1$ \ocaml{<}$x$\ocaml{<}
-$e_2$ binds the \emph{first} value published by $e_2$ to $x$ and
-executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if
-running $e_1$ does not publish any values.
+value published by $e_1$ to $x$ and executes $e_2$ for each such
+binding, building a concurrent pipeline. The expression $e_1$
+\ocaml{<}$x$\ocaml{<} $e_2$ binds the \emph{first} value published
+by $e_2$ to $x$ and executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$
+runs $e_2$ only if running $e_1$ does not publish any values.
 
@@
 parentheses. After composing the above definition with a simple
-language supporting addition, variables and function calls
-(omitted for brevity), parsing the expression
+language supporting addition, variables, and function calls
+(language definition omitted for brevity), parsing the expression
 %
@@
 \noindent
+Because there is no precedence specified between \ocaml{+} and
+\ocaml{>x>}, disambiguation is required to specify the order of
+operations.
+%
 In this case, it is likely that the preferred semantics is to have
@@
 combinators, and these precedence rules can be added after the
-composition without changing the original definitions.
-
-% TODO: Clashes involving ">" or ";"?
+composition without changing the original definitions. Importantly
+though, we are not \emph{required} to resolve this ambiguity at
+time of composition.
+
+Because of how we defined the syntax of the combinators, and since
+parsing with the tool is currently whitespace insensitive, the
+multi-character combinators (sequence and pruning) are parsed as
+three separate lexical tokens (this is visible in how whitespace
+is inserted in the error message above). This means we will also
+run into \emph{unresolvable} ambiguities, for example if our base
+language includes comparison of numbers with \ocaml{<} and
+\ocaml{>}. With such a base language (assuming no associativity
+for \ocaml{>}), parsing the expression
+%
+``\ocaml{42 >x> f(x)}''
+%
+will result in the following error message:
+
+\begin{lstlisting}[language={[objective]caml}]
+Unresolvable ambiguity error with 2 alternatives.
+\end{lstlisting}
+
+\begin{minipage}{.3\textwidth}
+\begin{lstlisting}[language={[objective]caml}]
+Resolvable alternatives:
+  ( 42 > x ) > f ( x )
+  42 > ( x > f ( x ) )
+\end{lstlisting}
+\end{minipage}
+\hfill
+\begin{minipage}{.5\textwidth}
+\begin{lstlisting}[language={[objective]caml}]
+Unresolvable alternatives:
+  seq
+   - int                  gt.orc:1:1-3
+   - call                 gt.orc:1:8-12
+\end{lstlisting}
+\end{minipage}
+
+\noindent
+By adding parentheses, a programmer can disambiguate the
+expression as two comparisons. However, there is no way for a
+programmer to specify that what they want is a sequential
+combinator with left and right children being an integer and a
+function call respectively. In this case we have at least three
+choices to make as language designers:
+
+\begin{itemize}
+\item We could decide to forbid the case where two comparisons are
+  right next to each other, e.g., \syncon{forbid gt.left = gt} and
+  \syncon{forbid gt.right = gt} (where \syncon{gt} is the syncon
+  for the greater-than operator).
+\item We could change the definition of the parallel combinators
+  and define separate lexical tokens for the combinators, e.g.,
+  \syncon{token Seq = ">[[:lower:]][[:word:]]*>"}.
+\item We could change the syntax of the combinators completely to
+  avoid clashes with other operators.
+\end{itemize}
+
+The change in the first alternative can be done after composition,
+and would allow us to reuse both language definitions without
+modifications. In the latter two alternatives, we lose reuse of
+the definitions of Orc combinators, but place no additional
+restrictions on the base language. The preferred variant is likely
+to differ between different compositions and different languages.
 

### 2019-07-09 paper/paper.tex
Small addendum to Orc

@@
   and define separate lexical tokens for the combinators, e.g.,
-  \syncon{token Seq = ">[[:lower:]][[:word:]]*>"}.
+  \syncon{token Seq = ">[[:lower:]][[:word:]]*>"}. This would add
+  just enough whitespace sensitivity to allow separating the
+  different cases.
 \item We could change the syntax of the combinators completely to
@@
 the definitions of Orc combinators, but place no additional
-restrictions on the base language. The preferred variant is likely
-to differ between different compositions and different languages.
+restrictions on the base language. The preferred resolution
+strategy is likely to differ between different compositions and
+different languages.
+%
+For example, another unresolvable ambiguity is going to show up if
+the base language uses semi-colons, e.g. for sequencing
+expressions (this would clash with the ``otherwise'' combinator).
+In this case, there is no reasonable way to disambiguate
+``\ocaml{e1; e2}'' without changing the syntax of one of the
+operations.
 

### 2019-07-09 paper/paper.tex
Info dump-ish on OCaml case study

@@
 
-% TODO: talk about the ``low precedence unary operator/deep precedence'' issue.
+The syncon definition of the OCaml language includes most syntax described in chapters 7 and 8 in the OCaml reference manual (the base language and extensions, respectively) and is sufficient to parse roughly 75\% (1012 out of 1334 files) of the \verb|.ml| files present in the OCaml compiler itself.
+The definition is just under 700 lines, excluding comments and empty lines (just over 1000 lines otherwise).
+It has been written to be as close to the grammar presented in the reference manual as possible, but with additions to conform to how the canonical compiler behaves on cases that are underspecified in the manual.
+
+The failing cases stem from three different causes:
+
+\begin{itemize}
+\item As of yet unspecified syntax. Due to a lack of time there are still parts of the language that we have not yet implemented. These appear solvable, but we can of course not be sure until we do.
+\item Our system does not support specifying ``longest match'' on a production, which is required to handle the pattern matching constructs correctly.
+We can handle some of these cases with marks, but not all.
+\item Different definitions of precedence.
+Our translation from precedence to marks is shallow, it only considers direct children, but OCaml has deep precedence.
+For example, addition binds stronger than \ocaml{if} (which syntactically functions as a prefix operator in OCaml), thus
+  \begin{lstlisting}[language={[objective]caml}]
+  1 + let x = 1 in x + 2
+  \end{lstlisting}
+  should be parsed as
+  \begin{lstlisting}[language={[objective]caml}]
+  1 + (let x = 1 in (x + 2))
+  \end{lstlisting}
+  while our tool presents the following ambiguity error message:
+  \begin{lstlisting}[language={[objective]caml}]
+Ambiguity error with 2 alternatives:
+
+  1 + ( let x = 1 in x ) + 2
+  1 + let x = 1 in ( x + 2 )
+  \end{lstlisting}
+  The former alternative is perhaps more easily understood as
+  \begin{lstlisting}[language={[objective]caml}]
+  (1 + (let x = 1 in x)) + 2
+  \end{lstlisting}
+  Looking at this tree, two levels at a time, we see ``\ocaml{(_ + _) + 2}'', ``\ocaml{1 + let _ in _}'', and ``\ocaml{let x = 1 in x}'', all of which are individually precedence correct.
+\end{itemize}
+
+The various definitions in the syncon OCaml definition are organized by the chapter in the reference manual that introduces them, plus an appendix with additions and changes required to mimic the behavior of the canonical compiler.
+Many of these are strict additions, requiring no changes to previous syncons, but not all.
+For example, OCaml allows placing ``attributes'' on most syntactical constructs in the language, extra information visible to tools but otherwise without semantic meaning.
+These can be written with an ``infix'' notation right after the keyword that introduces the syntactical construct, e.g., \ocaml{let[@foo] x = 1 in x}.
+This syntax is not fully specified; the manual merely contains seven examples, while the testsuite uses this syntax in roughly 50 different constructs.
+All these cases require changes to the original syncon definitions, making them slightly more complicated than the manual suggests them to be upon their introduction.
 

### 2019-07-10 paper/paper.tex
Can we? Will we?
Polishing Orc case study (including David's comments)

@@
 \begin{document}
+% TODO: Decide on British or American English.
 
@@
 
+% TODO: Make sure that forbids are explained
+
 We have implemented a tool and a small DSL for syntax definition
@@
 \subsection{Orc} \label{sec:evaluation-orc}
-
-Orc~\cite{kitchinOrc2009} is a functional programming language
-which includes a number of special-purpose combinators for
-coordinating concurrent workflows. In this section, we consider
-the use case of adding these concurrency combinators to an already
-defined programming language.
+%TODO Title should be descriptive
+
+In this section, we consider the use case of defining a language
+and composing it with another previously defined language, showing
+how we can deal with the resulting ambiguities that arise from the
+composition.
+%
+The language being defined is a subset of
+Orc~\cite{kitchinOrc2009}, a functional programming language which
+includes a number of special-purpose combinators for coordinating
+concurrent workflows.
 
@@
 pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
-combinators. The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
+combinators.
+%
+The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
 $e_2$ in parallel, publishing any value published by either of
-them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
-value published by $e_1$ to $x$ and executes $e_2$ for each such
-binding, building a concurrent pipeline. The expression $e_1$
-\ocaml{<}$x$\ocaml{<} $e_2$ binds the \emph{first} value published
-by $e_2$ to $x$ and executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$
-runs $e_2$ only if running $e_1$ does not publish any values.
+them.
+%
+The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ executes
+$[x\mapsto v]e_2$ for \emph{each} value $v$ published by $e_1$,
+building a concurrent pipeline.
+%
+The expression $e_1$ \ocaml{<}$x$\ocaml{<} $e_2$ executes
+$[x\mapsto v]e_1$ for \emph{first} value $v$ published by $e_2$
+(discarding any remaining values published by $e_2$).
+%
+Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if running $e_1$ does
+not publish any values.
 
@@
 parsing with the tool is currently whitespace insensitive, the
-multi-character combinators (sequence and pruning) are parsed as
+multi-character combinators (sequencing and pruning) are parsed as
 three separate lexical tokens (this is visible in how whitespace
-is inserted in the error message above). This means we will also
+is inserted in the error message above). This means we can also
 run into \emph{unresolvable} ambiguities, for example if our base
@@
 
-The change in the first alternative can be done after composition,
+\noindent
+The change in the first alternative is somewhat ad-hoc, but can be done after composition
 and would allow us to reuse both language definitions without
@@
 operations.
+
+The main takeaway from this case study is that resolvable
+ambiguity makes composition of languages less restrictive than if
+all ambiguity is completely banned. The approach is strictly more
+general since dynamic resolvability analysis allows deferring
+disambiguation to the programmer, while removing ambiguities from
+the resulting grammar is also possible.
 

### 2019-07-10 paper/paper.tex
Pass over case studies

@@
 Section~\ref{sec:parse-time-disambiguation}, which the tool then
-uses to construct a parser and do a dynamic resolvability
-analysis. The tool is implemented in Haskell and uses an
-off-the-shelf
+uses to construct a parser and do dynamic resolvability analysis.
+The tool is implemented in Haskell and uses an off-the-shelf
 implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
@@
 case studies is to showcase possible use cases of resolvable
-ambiguity in the context of a pre-existing general purpose
-programming language (OCaml, Section~\ref{sec:evaluation-ocaml}),
-a pre-existing DSL (Orc, Section~\ref{sec:evaluation-orc}), and a
-new DSL that is under development (Cyphym, Section~\ref{sec:evaluation-cyphym}).
-
-The syntax definition DSL builds on the the ``syncons''
-introduced by
+ambiguity in the context of composing separately defined DSLs (a
+subset of Orc, Section~\ref{sec:evaluation-orc}), and of defining
+a pre-existing general purpose programming language (OCaml,
+Section~\ref{sec:evaluation-ocaml}.
+
+The syntax definition DSL builds on the ``syncons'' introduced by
 \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
 production is defined separately from each other (one
-\syncon{syncon} each), and marks are primarily introduced through
-precedence lists. For example, the running example used in
+\syncon{syncon} each).
+%
+The main difference between the DSL and the formalism in
+Section~\ref{sec:parse-time-disambiguation} is that marks are
+introduced as separate \syncon{forbid} declarations rather than
+inline in productions. This allows supporting high-level
+convenience constructs, e.g., specifying precedence between
+previously defined operators instead of manually inserting marks,
+but is also important for composability as ambiguities can be
+statically resolved without changing the original definitions.
+Other convenience constructs include defining prefix, infix and
+postfix operators with a given associativity.
+
+The running example used in
 Section~\ref{sec:parse-time-disambiguation}
 (Fig.~\ref{fig:running-example-definition} on
-page~\pageref{fig:running-example-definition}) can be written as
+page~\pageref{fig:running-example-definition}) can be defined as
 follows:
@@
 token Integer = "[0-9]+"
+syncon literal: Exp = n:Integer
 syncon list: Exp =
-  "[" (head:Exp (";" tail:Exp))? "]"
+  "[" (head:Exp (";" tail:Exp)*)? "]"
 infix add: Exp = "+"
 infix mul: Exp = "*"
-syncon literal: Exp = n:Integer
 \end{lstlisting}
@@
 
-\noindent The design of the DSL diverges from the formalism in
-Section~\ref{sec:parse-time-disambiguation} partially for
-familiarity (precedence instead of marks), but also for
-composability reasons: the DSL allows marks to be introduced
-without altering the original definition of a syncon. If marks are
-needed in non-precedence related situations, e.g., adding
-sequential composition to the running example, they can be
-introduced with ``forbids'':
-
-\begin{lstlisting}[language=syncon]
-infix seqComp: Exp = ";"
-forbid list.head = seqComp
-forbid list.tail = seqComp
-\end{lstlisting}
-
-\subsection{OCaml} \label{sec:evaluation-ocaml}
-
-The syncon definition of the OCaml language includes most syntax described in chapters 7 and 8 in the OCaml reference manual (the base language and extensions, respectively) and is sufficient to parse roughly 75\% (1012 out of 1334 files) of the \verb|.ml| files present in the OCaml compiler itself.
-The definition is just under 700 lines, excluding comments and empty lines (just over 1000 lines otherwise).
-It has been written to be as close to the grammar presented in the reference manual as possible, but with additions to conform to how the canonical compiler behaves on cases that are underspecified in the manual.
-
-The failing cases stem from three different causes:
-
-\begin{itemize}
-\item As of yet unspecified syntax. Due to a lack of time there are still parts of the language that we have not yet implemented. These appear solvable, but we can of course not be sure until we do.
-\item Our system does not support specifying ``longest match'' on a production, which is required to handle the pattern matching constructs correctly.
-We can handle some of these cases with marks, but not all.
-\item Different definitions of precedence.
-Our translation from precedence to marks is shallow, it only considers direct children, but OCaml has deep precedence.
-For example, addition binds stronger than \ocaml{if} (which syntactically functions as a prefix operator in OCaml), thus
-  \begin{lstlisting}[language={[objective]caml}]
-  1 + let x = 1 in x + 2
-  \end{lstlisting}
-  should be parsed as
-  \begin{lstlisting}[language={[objective]caml}]
-  1 + (let x = 1 in (x + 2))
-  \end{lstlisting}
-  while our tool presents the following ambiguity error message:
-  \begin{lstlisting}[language={[objective]caml}]
-Ambiguity error with 2 alternatives:
-
-  1 + ( let x = 1 in x ) + 2
-  1 + let x = 1 in ( x + 2 )
-  \end{lstlisting}
-  The former alternative is perhaps more easily understood as
-  \begin{lstlisting}[language={[objective]caml}]
-  (1 + (let x = 1 in x)) + 2
-  \end{lstlisting}
-  Looking at this tree, two levels at a time, we see ``\ocaml{(_ + _) + 2}'', ``\ocaml{1 + let _ in _}'', and ``\ocaml{let x = 1 in x}'', all of which are individually precedence correct.
-\end{itemize}
-
-The various definitions in the syncon OCaml definition are organized by the chapter in the reference manual that introduces them, plus an appendix with additions and changes required to mimic the behavior of the canonical compiler.
-Many of these are strict additions, requiring no changes to previous syncons, but not all.
-For example, OCaml allows placing ``attributes'' on most syntactical constructs in the language, extra information visible to tools but otherwise without semantic meaning.
-These can be written with an ``infix'' notation right after the keyword that introduces the syntactical construct, e.g., \ocaml{let[@foo] x = 1 in x}.
-This syntax is not fully specified; the manual merely contains seven examples, while the testsuite uses this syntax in roughly 50 different constructs.
-All these cases require changes to the original syncon definitions, making them slightly more complicated than the manual suggests them to be upon their introduction.
-
-\subsection{Orc} \label{sec:evaluation-orc}
-%TODO Title should be descriptive
+\noindent Here, we define a syntax type \syncon{Exp} for
+expressions, and declare that parentheses can be used to group
+expressions. We define a lexical token for integers, and a
+\syncon{syncon} for integer literals using this token. A list is
+defined as a sequence of zero or more expressions separated by
+semi-colons wrapped in square braces. Finally, addition and
+multiplication is defined as infix syncons with the expected
+precedence rules. Note that we could just as well have replaced
+the precedence list with explicit \syncon{forbid} declarations,
+``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right
+  = add}'' (cf. the mark $\{a\}$ in the production of $m$ in
+Fig.~\ref{fig:running-example-definition}).
+
+% If marks are needed in non-precedence related
+% situations, e.g., adding sequential composition to the running
+% example, they can be introduced with ``forbids'':
+
+% \begin{lstlisting}[language=syncon]
+% infix seqComp: Exp = ";"
+% forbid list.head = seqComp
+% forbid list.tail = seqComp
+% \end{lstlisting}
+
+\subsection{Composing Language Definitions} \label{sec:evaluation-orc}
 
@@
 includes a number of special-purpose combinators for coordinating
-concurrent workflows.
+concurrent workflows. On their own, these combinators act as a DSL
+for concurrency.
 
@@
 
-\subsection{Cyphym} \label{sec:evaluation-cyphym}
+\subsection{Applying Resolvable Ambiguity to a Full Language} \label{sec:evaluation-ocaml}
+
+The syncon definition of the OCaml language includes most syntax described in chapters 7 and 8 in the OCaml reference manual (the base language and extensions, respectively) and is sufficient to parse roughly 75\% (1012 out of 1334 files) of the \verb|.ml| files present in the OCaml compiler itself.
+The definition is just under 700 lines, excluding comments and empty lines (just over 1000 lines otherwise).
+It has been written to be as close to the grammar presented in the reference manual as possible, but with additions to conform to how the canonical compiler behaves on cases that are underspecified in the manual.
+
+The failing cases stem from three different causes:
+
+\begin{itemize}
+\item As of yet unspecified syntax. Due to a lack of time there are still parts of the language that we have not yet implemented. These appear solvable, but we can of course not be sure until we do.
+\item Our system does not support specifying ``longest match'' on a production, which is required to handle the pattern matching constructs correctly.
+We can handle some of these cases with marks, but not all.
+\item Different definitions of precedence.
+Our translation from precedence to marks is shallow, it only considers direct children, but OCaml has deep precedence.
+For example, addition binds stronger than \ocaml{if} (which syntactically functions as a prefix operator in OCaml), thus
+  \begin{lstlisting}[language={[objective]caml}]
+  1 + let x = 1 in x + 2
+  \end{lstlisting}
+  should be parsed as
+  \begin{lstlisting}[language={[objective]caml}]
+  1 + (let x = 1 in (x + 2))
+  \end{lstlisting}
+  while our tool presents the following ambiguity error message:
+  \begin{lstlisting}[language={[objective]caml}]
+Ambiguity error with 2 alternatives:
+
+  1 + ( let x = 1 in x ) + 2
+  1 + let x = 1 in ( x + 2 )
+  \end{lstlisting}
+  The former alternative is perhaps more easily understood as
+  \begin{lstlisting}[language={[objective]caml}]
+  (1 + (let x = 1 in x)) + 2
+  \end{lstlisting}
+  Looking at this tree, two levels at a time, we see ``\ocaml{(_ + _) + 2}'', ``\ocaml{1 + let _ in _}'', and ``\ocaml{let x = 1 in x}'', all of which are individually precedence correct.
+\end{itemize}
+
+The various definitions in the syncon OCaml definition are organized by the chapter in the reference manual that introduces them, plus an appendix with additions and changes required to mimic the behavior of the canonical compiler.
+Many of these are strict additions, requiring no changes to previous syncons, but not all.
+For example, OCaml allows placing ``attributes'' on most syntactical constructs in the language, extra information visible to tools but otherwise without semantic meaning.
+These can be written with an ``infix'' notation right after the keyword that introduces the syntactical construct, e.g., \ocaml{let[@foo] x = 1 in x}.
+This syntax is not fully specified; the manual merely contains seven examples, while the testsuite uses this syntax in roughly 50 different constructs.
+All these cases require changes to the original syncon definitions, making them slightly more complicated than the manual suggests them to be upon their introduction.
+
+% \subsection{Cyphym} \label{sec:evaluation-cyphym}
 

### 2019-07-10 paper/paper.tex
Fix build
Polish in Sec 4
First changes of intro

@@
 \newcommand{\lattice}{\mathit{lattice}} % go from a t \in L(G_t) to its lattice, partitioned by equality on \altset and ordered by subset on the rangeset.
+\newcommand{\topw}{\top_w} % go from a tree to its top word
+\newcommand{\botw}{\bot_w} % go from a tree to its bottom word
 \newcommand{\localize}{\mathit{localize}} % go from a F \subseteq L(G_t) to a set of localized ambiguities, i.e., a set of subforests.
@@
 \newcommand{\passthrough}{\mathit{passthrough}} % Go from label to a set of (label x bool), where each element denotes a child production that can be an only child (i.e., no other terminals whatsoever), and the bool states whether it requires parentheses
-\newcommand{\nt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs
+\newcommand{\labelnt}{\mathit{nt}} % Go from label to the non-terminal to which that labelled production belongs
+\newcommand{\labelregex}{\mathit{regex}} % Go from label the regex of the production with that label
 
@@
 \begin{abstract}
-% Why?
-
-Ever since the early 60s, the problem of deciding if a context-free
-grammar is unambiguous or not has been known to be undecidable. As a
-consequence, a large number of restricted grammars have been developed
-to guarantee that a language definition is unambiguous. This
-traditional view of only allowing unambiguous grammars has until
-recently been taken for granted as the only truth: \emph{the} way of
-how the syntax of a language must be defined. However, up front
-unambiguous grammars can force language designers to make arbitrary
-choice to disambiguate the language, although the natural choice may be to
-postpone the decision to the programmer. Moreover, an important
-objective of domain-specific language development is to construct
-languages by composing and extending existing
-languages. Unfortunately, the composition of restricted unambiguous
-grammars easily results in ambiguous grammars, which again are
-undecidable. In this paper, we depart from the traditional view of
+A common point of view when designing grammars for parsing is that
+the grammar definition has to be unambiguous.  However, up front unambiguous grammars
+can force language designers to make arbitrary choice to disambiguate
+the language. In this paper, we depart from the traditional view of
 unambiguous grammar design, and allow ambiguities to be delayed until
 compile time, allowing the user to perform the disambiguation. A
-natural decision problems follows: given a language definition, can a
+natural decision problem follows: given a language definition, can a
 user always disambiguate an ambiguous program?  We introduce and
@@
 case for dynamic resolvability. The approach is evaluated through
-three separate case studies, covering both a large
-existing programming language, and the composability of
-domain-specific languages.
-%A traditional view when designing programming languages is that syntax
-%definitions should be unambiguous: only unambiguous grammars are used
-%to construct language specific parsers that either rejects a program,
-%or accepts it and produces a unique parse tree. However,
+two separate case studies, covering both a large existing
+programming language, and the composability of domain-specific
+languages.
 \end{abstract}
@@
 
-When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments.
-However, how to make such compositions well-behaved is not always obvious.
-In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous.
-Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs.
-The former is well explored in the form of heuristics for ambiguity detection (e.g.
-\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g.
-\cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny.
-This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g.
-\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).
-
-Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
-\cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
+% Motivate why unambiguous parsing is a good (and important) idea.
+Ever since the early 60s, the problem of deciding if a context-free
+grammar is ambiguous has been known to be
+undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
+a large number of restricted grammars have been developed to guarantee
+that a language definition is unambiguous. This traditional view of
+only allowing unambiguous grammars has been taken for granted as the
+only truth: \emph{the} way of how the syntax of a language must be
+defined
+%This seems to stem
+%from the view that ambiguity has no place in programming language
+%syntax, which is highly prevalent
+~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
+%
+However, recent work by
+~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
+carefully designed ambiguous syntax definitions are in some cases
+preferable compared to completely unambiguous
+definitions. Specifically, instead of forcing a language designers to
+make arbitrary choice to disambiguate the language, a more natural
+choice may be to postpone the decision to the programmer.
+
+% Motivation II, compising DSLs
+Another area---where ambiguous grammars naturally arises---is
+domain-specific language development.  An important
+objective of domain-specific language development is to construct
+languages by composing and extending existing
+languages.
+Unfortunately, the composition of restricted
+%grammars easily results in ambiguous grammars, which again are
+%undecidable.
+%
+%constructing domain-specific programming languages there is often
+%a desire to reuse work, to compose smaller language
+%fragments.
+%However, how to make such compositions well-behaved is not
+%always obvious.
+%In particular, when constructing new domain-specific languages from
+%existing languages, by composing multiple syntax definitions,
+grammars easily produces
+an ambiguous grammar, even if all of the original grammars are
+unambiguous. Approaches for solving this problem can be broadly split
+into two categories: those that handle ambiguities on the
+grammar-level (detection, prevention, etc.), and those that work on
+particular examples of ambiguous programs. The former is well explored
+in the form of heuristics for ambiguity detection
+(e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007})
+or restrictions on the grammars to be composed
+(e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the
+latter has received relatively little attention.
+
+%Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
+\cite{earleyEfficientCo%ntextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
 However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity.
@@
 
-In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists.
+%In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists.
 For example, in a language with semicolon for both sequential composition and element separation in a list (e.g.
@@
 The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.
+
+% The problem
+
+This paper focuses on the problems that arises if a grammar definition
+is not guaranteed to be unambiguous. Specifically, if parsing a
+program results in several parse trees, the grammar was obviously
+ambiguous. The programmer then needs to disambiguate the program by,
+for instance, inserting extra parenthesis. However, is it always
+possible to resolve an ambiguous input program by augmenting the
+program? We say that a program has a \emph{resolvable ambiguity} if
+the ambiguity can be solved by augmenting the program.
+The \emph{resolvable ambiguity problem} can then be divided into several decision problems: (i) the dynamic decidability problem: given a program (or a
+
+dynamic resolvable ambiguity problem of deciding if a context-free grammar is
+
 This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
@@
 
-\noindent We say that the automaton accepts the word $a_0a_1\ldots a_n$ iff there is such a succesful run.
+\noindent We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a succesful run.
 
@@
 \item[Static resolvability.] Given an abstract parser $P$, determine whether $\rho(P)$.
-\item[Dynamic resolvability.] Given a word $w \in L(P)$ for some abstract parser $P$, determine whether $\rho_P(w)$.
+\item[Dynamic resolvability.] Given an abstract parser $P$ and a word $w \in L(P)$, determine whether $\rho_P(w)$.
 \end{description}
@@
 
-% todo: improve label (where used?)
-\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure}
+\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure-sound}
   A decision procedure $f$ solving the static resolvability problem is sound iff
@@
   parsers } P$. It turns out, we can give $g$ for the parse languages
-studied in following sections.
-
-\begin{definition}[Completeness of Static Resolvability]
+studied in the following sections.
+
+\begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
   A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff
@@
 
-\noindent Similarly, for the dynamic problem:
-
-\begin{definition}
-  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
-  \begin{itemize}
-  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
-  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
-  \end{itemize}
-  \label{def:dynamic-procedure}
+\noindent Similarly, we define soundness and completeness for the dynamic problem.
+
+\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure-sound}
+  A decision procedure $f$ solving the dynamic resolvability problem is sound iff
+
+  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
+\end{definition}
+
+Analogous to the static case, for the parse languages studied in the
+following sections we can also provide a decision procedure $g$ such
+that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
+\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.
+
+\begin{definition}[Completeness of Dynamic Resolvability]\label{def:dynamic-procedure-complete}
+  A decision procedure $f$ solving the dynamic resolvability problem is complete iff
+
+  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
 \end{definition}
@@
 As a memory aid, the primed versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations (i.e., ignoring marks) from $D$.
-We will generally refer to elements of $L(G_D)$ as $w$, $L(G'_D)$ as $w'$, $L(T_D)$ as $t$, and $L(T'_D)$ as $t'$.
 
@@
 \begin{array}{rcl}
-\parse(w') & = & \{ \semantic(t') \mid t' \in L(T'_D) \land \yield(t') = w' \} \\
-\words(t) & = & \{ w' \mid t \in \parse(w') \} \\
+\parse(w) & = & \{ \semantic(t) \mid t \in L(T'_D) \land \yield(t) = w \} \\
+\words(t) & = & \{ w \mid t \in \parse(w) \} \\
 \end{array}
@@
 
-We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure} and \ref{def:dynamic-procedure}.
-Section~\ref{sec:static} gives a partial solution to the former problem, while Section~\ref{sec:dynamic} fully solves the latter with one caveat: we only consider languages with balanced parentheses.
+We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}.
+Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem with one caveat: we only consider languages with balanced parentheses.
 
@@
 \begin{theorem}
-  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
+  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
 \end{theorem}
@@
 
-\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w' \in L(G'_D)$ for which $\parse(w') = \{t\}$, or prove that no such tree exists.
+\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w \in L(G'_D)$ for which $\parse(w) = \{t\}$, or prove that no such tree exists.
 Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.
@@
 \begin{lemma}
-  $\forall w'_1, w'_2 \in \words(t).\ \basic(w'_1) = \basic(w'_2)$, i.e., all $w'$ have the same basic word.
+  $\forall w_1, w_2 \in \words(t).\ \basic(w_1) = \basic(w_2)$.
   \label{lemma:same-basic}
 \end{lemma}
-\noindent Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
-The only terminals changed in this process are parentheses, i.e., different $w'$ can only differ in their range bags.
-This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w'$ such that $\parse(w') \supseteq \{t_1, t_2\}$) must also share a basic word.
+\noindent More informally, all $w \in \words(t)$ have the same basic word.
+Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
+The only terminals changed in this process are parentheses, i.e., different $w$ can only differ in their range bags.
+This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w$ such that $\parse(w) \supseteq \{t_1, t_2\}$) must also share a basic word.
 
 \begin{lemma}
-  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\\ R \subseteq \support(\rangebag(w'))$, i.e., some parentheses are required.
+  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ R \subseteq \support(\rangebag(w))$
   \label{lemma:required-parentheses}
 \end{lemma}
-\noindent For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees.
+\noindent Recall that $\support(B)$ is the support of the bag $B$, i.e., the underlying set of elements in $B$.
+More informally, some parentheses are required.
+For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees.
 Each required range is a direct consequence of a mark on a non-terminal in $D$.
-We will write $w'_\bot$ for the word implied by the first quantified expression.
+We will write $\botw(t)$ for the word implied by the first quantified expression.
 It is unique, since its basic word is fixed by $t$, and its rangebag is a set.
@@
 \begin{lemma}
-  $\exists R$ such that $\exists w' \in \words(t).\ \rangebag(w') = R$ and $\forall w' \in \words(t).\\ \support(\rangebag(w')) \subseteq R$, i.e., there is a finite set of possible parentheses.
+  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ \support(\rangebag(w)) \subseteq R$.
   \label{lemma:possible-parentheses}
 \end{lemma}
-\noindent As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes.
-We will write $w'_\top$ for the word implied by the first quantified expression, which is also unique.
+\noindent More informally, there is a finite set of possible parentheses.
+As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes.
+We will write $\topw(t)$ for the word implied by the first quantified expression, which is also unique.
 
 \begin{lemma}
-  $\forall w'_1, w'_2 \in \words(t).\ \support(\rangebag(w'_1)) = \support(\rangebag(w'_2)) => \parse(w'_1) = \parse(w'_2)$, i.e., duplicated parentheses do not matter.
+  $\forall w_1, w_2 \in \words(t).\ \support(\rangebag(w_1)) = \support(\rangebag(w_2)) => \parse(w_1) = \parse(w_2)$.
   \label{lemma:rangeset-equality}
 \end{lemma}
-\noindent We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses.
+\noindent More informally, duplicated parentheses do not matter.
+We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses.
 Each pair of parentheses thus corresponds to a single grouping node $g$ in a tree in $L(T'_D)$.
@@
 
-Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w'$ by their \emph{rangeset}, where $\rangeset(w') = \support(\rangebag(w'))$.
+Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w$ by their \emph{rangeset}, where $\rangeset(w) = \support(\rangebag(w))$.
 These partitions can be partially ordered by subset on the rangeset, resulting in a lattice of word partitions per tree.
-This lattice is bounded, with top and bottom elements given by lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} respectively.
+This lattice is bounded, with top and bottom elements $\topw(t)$ and $\botw(t)$.
 Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
@@
 \begin{lemma}
-  Given $t \in L(T_D)$, $w'_\top, w'_\bot \in \words(t)$ such that $\forall w'.\ w' \in \words(t) => \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top)$ the following holds:\\
-  $\forall w'.\ \rangebag(w'_\bot) \subseteq \rangeset(w') \subseteq \rangebag(w'_\top) => w' \in \words(t)$. \label{lemma:top-bottom-determine}
+  Given $t \in L(T_D)$, we have that:\\
+  $\forall w.\ \rangebag(\botw(t)) \subseteq \rangeset(w) \subseteq \rangebag(\topw(t)) => w \in \words(t)$. \label{lemma:top-bottom-determine}
 \end{lemma}
 
-\noindent Lemmas~\ref{lemma:possible-parentheses} and \ref{lemma:required-parentheses} guarantee the existence of two such words $w'_\top$ and $w'_\bot$.
-
+\noindent Recall that $\rangebag = \rangeset$ for $\botw(t)$ and $\topw(t)$.
 Finally, we show the difference between versions~1 and 2 in the lattice setting: in version~1, the rangeset of the bottom word is the empty set:
@@
 \begin{lemma}
-  In version~1, $rangeset(w'_\bot) = \emptyset$ for all trees, where $w'_\bot$ is given by Lemma~\ref{lemma:required-parentheses}. \label{lemma:version1-bot}
+  In version~1, $\forall t \in L(T_D).\ \rangeset(\botw(t)) = \emptyset$. \label{lemma:version1-bot}
 \end{lemma}
@@
 \begin{theorem}
-  In versions~1 and 2, given a $t \in L(T_D)$ with a lattice determined by $w_\bot$ and $w_\top$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w_\bot) \subseteq \rangeset(w'_\bot) \land \rangeset(w'_\top) \subseteq \rangeset(w_\top) => \neg \exists w.\ \parse(w) = \{t\}$.
+  In versions~1 and 2, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, the existence of a tree $t' \in L(T_D)$ such that $\basic(w_\bot) = \basic(\botw(t')) \land \rangeset(w_\bot) \subseteq \rangeset(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$ implies $\neg \exists w.\ \parse(w) = \{t\}$.
 \label{lemma:lattice-containment-implies-unresolvable}
@@
 \begin{theorem}
-  In version 1, given a $t \in L(T_D)$ with a lattice determined by $w_\bot$ and $w_\top$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w'_\bot) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(w'_\top) <=> \neg \exists w.\ \parse(w) = \{t\}$.
+  In version 1, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\botw(t')) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(\topw(t')) <=> \neg \exists w.\ \parse(w) = \{t\}$.
 \end{theorem}
@@
 \begin{proof}
-  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(w'_\bot)$ implies that $w_\bot = w'_\bot$.
-We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(w'_\bot) \land \rangeset(w'_\top) \subseteq \rangeset(w_\top)$.
+  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(\botw(t'))$ implies that $w_\bot = \botw(t')$.
+We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$.
 
   If all words in $\words(t)$ are ambiguous, then $\exists t' \in L(T_D).\ t' \neq t \land t' \in \parse(w_\top)$.
-Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(w'_\bot)$.
-The definition of $w'_\top$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(w'_\top)$, but since the rangebag of $w'_\top$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(w'_\top)$.
+Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(\botw(t'))$.
+The definition of $\topw(t')$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(\topw(t'))$, but since the rangebag of $\topw(t')$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(\topw(t'))$.
 \end{proof}
@@
 
-Conceptually, each non-terminal represents a choice of which child production should replace it, and whether that production requires parentheses around it.
+Conceptually, each non-terminal represents a choice of which production should replace it, and whether that production requires parentheses around it.
 We now make this explicit, to ensure that standard operations on finite automata produce the correct result.
@@
 \noindent Similarly, for every edge $p \xrightarrow{l_\bot} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\pospl', +(p, q)} p'$ and $q' \xrightarrow{'\pospr', -(p, q)} q$.
+Note the difference: for $l_\top$ we add a required parenthesis ``$\reqpl$'', while $l_\bot$ adds a possible parenthesis ``$\pospl$''.
 
@@
   \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
-  (6) edge[left, in=270, out=90] node[xshift=-9.8mm,yshift=3mm]{$'\pospr', -(1, 2)$} (2);
+  (6) edge[right] node[xshift=5mm,yshift=-3mm]{$'\pospr', -(1, 2)$} (2);
   \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
   (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
-  \path (3) edge[in=90, out=270, right] node[xshift=9mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
+  \path (3) edge[right] node[xshift=5mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
   (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
@@
 
-\noindent The resulting pushdown automaton has only two sources of non-determinism: the edges labelled $'\pospl'$ and $'\reqpl'$.
-Each one corresponds to one of the allowable child productions at that point in the tree, thus there is one run per tree.
+\noindent For example, the linearly encoded lattice $\posp{\posp{e};\posp{e}}$ is recognized by the run
+
+\begin{minipage}{.7\textwidth}
+$$
+\begin{array}{l@{\enspace}l@{\;}r@{\enspace}l@{\;}r@{\enspace}l@{\;}r}
+  (7, \epsilon) &\xrightarrow{\pospl} (1, &(7,8)) &\xrightarrow{\pospl} (5, &(1,2)\cdot(7,8)) &\xrightarrow{e} (6, &(1,2)\cdot(7,8)) \\
+  &\xrightarrow{\pospr} (2, &(7,8)) &\xrightarrow{;} (3, &(7,8)) &\xrightarrow{\pospl} (5, &(3,4)\cdot(7,8)) \\
+  &\xrightarrow{e} (6, &(3,4)\cdot(7,8)) &\xrightarrow{\pospr} (4, &(7, 8)) &\xrightarrow{\pospr} (8, &\epsilon) \\
+\end{array}
+$$
+\end{minipage}
+\begin{minipage}{.3\textwidth}
+\begin{center}
+{
+\newcommand{\terminal}[1]{\ \underline{#1}\ }
+\Tree [.{$s$}
+    [.{$v$} \terminal{e} ]
+    \terminal{;}
+    [.{$v$} \terminal{e} ] ]
+}
+\end{center}
+\end{minipage}
+
+\noindent and corresponds to the tree on the right.
+
+We now illustrate the issue that our previous simplification avoids, and then modify the construction such that the simplification is no longer required. Consider the following language definition:
+
+\begin{center}
+\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+  \toprule
+  $A$ & $a$ & $B$ \\
+  $B$ & $b$ & $C$ $(\verb|';'| + \epsilon)$ \\
+  $C$ & $c$ & \verb|'e'| \\
+  \bottomrule
+\end{tabular}
+\end{center}
+
+\noindent Following the simplified construction we produce the following automaton:
+
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+
+  \node[state,initial] (8) [left of=1, xshift=-1cm] {$8$};
+  \node[state,accepting] (9) [right of=2, xshift=1cm] {$9$};
+
+  \node[state] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+
+  \node[state] (3) [below of=1,yshift=3mm] {$3$};
+  \node[state] (4) [right of=3] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+
+  \node[state] (6) [below of=3,yshift=3mm] {$6$};
+  \node[state] (7) [right of=6] {$7$};
+
+  \path (4) edge[below] node {$';'$} (5);
+  \path (6) edge[below] node {$'e'$} (7);
+
+  \path (1) edge[left] node {$'\pospl', +(1,2)$} (3)
+  (4) edge node {$'\pospr', -(1,2)$} (2)
+  (5) edge[right] node {$'\pospr', -(1,2)$} (2);
+  \path (3) edge[left] node {$'\pospl', +(3, 4)$} (6)
+  (7) edge[left] node {$'\pospr', -(3, 4)$} (4);
+
+  \path (8) edge node {$'\pospl', +(8, 9)$} (1)
+  (2) edge node {$'\pospr', -(8, 9)$} (9);
+
+\end{tikzpicture}
+\end{center}
+
+\noindent Unfortunately, this automaton recognizes the word $(((e)))$, which is not the unique linear encoding we require; it should be $(e)$.
+
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+
+  \node[state,initial] (8) [left of=1, xshift=-1cm] {$8$};
+  \node[state,accepting] (9) [right of=2, xshift=1cm] {$9$};
+
+  \node[state] (1) {$1$};
+
+  \node[state] (3) [below of=1,yshift=3mm] {$3$};
+  \node[state] (4) [right of=3] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+
+  \node[state] (6) [below of=3,yshift=3mm] {$6$};
+  \node[state] (7) [right of=6] {$7$};
+
+  \path (4) edge[below] node {$';'$} (5);
+  \path (6) edge[below] node {$'e'$} (7);
+
+  \path (3) edge[left] node {$'\pospl', +(3, 4, \epsilon)$} (6)
+  (7) edge[left] node {$'\pospr', -(3, 4, \epsilon)$} (4);
+
+  \path (8) edge node {$'\pospl', +(8, 9, \epsilon)$} (1)
+  (8) edge[right] node[xshift=1mm,yshift=1mm] {$'\pospl', +(8, 9, a)$} (3)
+  (5) edge[left] node {$'\pospr', -(8,9,a)$} (9)
+  (8) edge[left,out=270,in=180] node {$'\pospl', +(8, 9, ab)$} (6)
+  (7) edge[right,out=0,in=270] node {$'\pospr', -(8, 9, ab)$} (9);
+
+\end{tikzpicture}
+\end{center}
 
@@
 
-%% TODO: Continue here. Also figure out how to handle (E + E_m).
-
-%% We now have a pushdown automaton (call it $A_{()}$) that recognizes the top word for each tree in $L(G_t)$.
-Next we need to be able to add arbitrary parentheses, to produce a word with a rangeset superset.
-To do this we create a copy of $A_{()}$ with one modification: for every state $s$ in $A_{()}$ (except the initial and final states), add two transitions $s \xrightarrow{'(', +p} s$ and $s \xrightarrow{')', -p} s$.
-To avoid cluttering the graph too much, these transitions are shown unlabeled below:
-
-%% \begin{center}
-%% \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-%%     scale = 1,transform shape]
-
-%%   \node[state] (1) {$1$};
-%%   \node[state] (2) [right of=1] {$2$};
-%%   \node[state] (3) [right of=2] {$3$};
-%%   \node[state] (4) [below of=1, xshift=1cm] {$4$};
-%%   \node[state] (5) [right of=4] {$5$};
-%%   \node[state,initial] (s) [left of=4] {};
-%%   \node[state,accepting] (e) [right of=5] {};
-
-%%   \draw
-%%   (1) edge[loop, in=80, out=100, looseness=7] node{} (1)
-%%   (1) edge[loop, in=70, out=110, looseness=6] node{} (1)
-%%   (2) edge[loop, in=80, out=100, looseness=7] node{} (2)
-%%   (2) edge[loop, in=70, out=110, looseness=6] node{} (2)
-%%   (3) edge[loop, in=-10, out=10, looseness=7] node{} (3)
-%%   (3) edge[loop, in=-20, out=20, looseness=6] node{} (3)
-%%   (4) edge[loop, in=-80, out=-100, looseness=7] node{} (4)
-%%   (4) edge[loop, in=-70, out=-110, looseness=6] node{} (4)
-%%   (5) edge[loop, in=-80, out=-100, looseness=7] node{} (5)
-%%   (5) edge[loop, in=-70, out=-110, looseness=6] node{} (5)
-%%   (s) edge[left] node{$'(', +s$} (1)
-%%   (s) edge[above] node{$'(', +s$} (4)
-%%   (3) edge[right] node{$')', -s$} (e)
-%%   (5) edge[above] node{$')', -s$} (e)
-%%   (1) edge[above] node{$s$} (2)
-%%   (4) edge[above] node{$z$} (5)
-%%   (2) edge[bend left, below] node{$'(', +(2, 3)$} (1)
-%%   (2) edge[bend left, left] node{$'(', +(2, 3)$} (4)
-%%   (3) edge[loop above] node{$')', -(2, 3)$} (3)
-%%   (5) edge[left, bend right] node{$')', -(2, 3)$} (3);
-
-%% \end{tikzpicture}
-%% \end{center}
-
-%% \noindent We will call this automaton $A'_{()}$.
-Successful runs in this automaton have a surjection to runs in $A_($ (ignore transitions along the newly added edges), and thus also have a surjection to parse trees in $L(G_t)$.
-We can also note that every sucessful run in $A_{()}$ is also a successful run in $A'_{()}$, since the latter has all states and transitions of the former.
-Furthermore, two distinct runs, one in $A_{()}$ (call it $p$) and one in $A'_{()}$ (call it $p'$), that both recognize the same word must represent different parse trees in $L(G_t)$.
-To see why, we consider two cases:
-
-%% \begin{enumerate}
-%%   \item $p'$ only uses transitions present in $A_{()}$.
-This is a successful run in $A_{()}$, and distinct from $p$.
-But there is a bijection between runs in $A_{()}$ and parse trees in $L(T)$, thus $p'$ represents a different parse tree.
-%%   \item $p'$ uses at least one transition added in $A'_{()}$.
-Using the surjection between runs in $A'_{()}$ and $A_{()}$ we find a new successful run that produces a different word (at least one fewer pair of parentheses).
-Since this run produces a different word, it must be distinct from $p$, and thus represent a different parse tree.
-%% \end{enumerate}
-
-%% \noindent Two distinct successful runs that accept the same word thus represent two trees where one permits a superset rangeset of the other.
-To find such runs we construct a product automaton and trim it.
-We can construct a product automaton since both $A_{()}$ and $A'_{()}$ are visibly pushdown, with the same partition of the input alphabet (push on open parenthesis, pop on close parenthesis, do nothing otherwise).
-We can trim the product since it retains the same partitioning and thus is also visibly pushdown.
-
-%% In this product automaton, if any transition pushes a stack symbol $(a, b)$ where $a \neq b$, or transitions to a state $(p, q)$ where $p \neq q$, then there is a successful run that corresponds to two distinct runs through $A_{()}$ and $A'_{()}$ (since the automaton is trim).
-
-\begin{theorem}
-  There is a bijection between successful runs in $A_{\posp{}}$ and trees $t \in L(T_D)$.
-\end{theorem}
-
-\begin{proof}
-  (These are two attempts at showing the same thing, next attempt starts at ``Proof sketch''. I would strongly suggest reading ``Formal stuff'' first.)
-  $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
-The two are distinct, and $f$ is reachable only through transitions that pop a stack symbol with $s$ and $f$ as the two first components.
-Transitions that push such symbols originate in $s$, and no other transitions originate in $s$.
-Thus all successful runs start in configuration $(s, \epsilon)$ and end in $(f, \epsilon)$, whereby every pushing transition in the run can be paired with a corresponding popping transition.
-We will now construct a tree $t \in L(T_D)$ in a bottom-up fashion.
-
-  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, w)} (p', (p, q, w) \cdot s) \; R_2 \; (q', (p, q, w) \cdot s \xrightarrow{b, -(p, q, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
-The only transitions that leave the stack unchanged in $A_{\posp{}}$ come from DFAs generated from productions in $D$.
-There is no way to transition between two such DFAs without using a pushing or popping transition, thus $R_2$ must be a successful run in the DFA for some labelled production $N -> l : r$ in $D$.
-$T_D$ must contain a corresponding production $N -> l(r)$.
-Now consider the sequence of input symbols $w'$ along the transitions in $R_2$.
-We have $w' \in L(r)$, since $R_2$ is a successful run in a DFA constructed from $r$.
-Thus $l(w') \in L_N(T_D)$.
-Now add parent nodes to this subtree for every label in $w$, e.g., if $w = l_1 \cdot l_2 \cdot l_3$ construct $l_1(l_2(l_3(l(w'))))$.
-We can then replace the subsequence of $R$ with this subtree, and then repeat the process until we have a tree with root $l'$, where $D$ has a labelled production $S -> l': r'$ for some $r'$ and the starting symbol $S$.
-
-  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{\posp{}}$ starting from $(s, \epsilon)$.
-If the next terminal is in $\T$, i.e., a strictly zero-arity terminal, we take a transition originating in a DFA, which must be unique.
-If the next terminal is in $\Labels$, i.e., an unranked terminal, we take one of the added edges.
-However, this choice of edge might not be unique: there may be multiple edges from the current state to an appropriate DFA initial state.
-These stem from the original DFA having two transitions from the same state with the same label but different subscripts, e.g., $p \xrightarrow{l_\bot} q$ and $p \xrightarrow{l_\top} q'$.
-However, if both of these transitions were valid choices for the current tree that would imply that the DFA for that level can recognize two distinct words that differ only in subscripts.
-But distinctly subscripted terminals must originate from distinct occurances of marked non-terminals (since a $\top$ subscript corresponds to a mark, and a $\bot$ subscript corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
-
-  Proof sketch: to go from a successful run to a tree $t \in L(T_D)$: by construction, all successful runs must begin and end with an empty stack.
-Every pushing transition can thus be paired with a popping transition in the run.
-We can construct a tree in a bottom-up fashion by replacing every such pair with a tree $l_1(l_2(\ldots l_n(x)\ldots))$, where $x$ is the sequence of terminals and/or trees between the pair, the terminals $l_1, l_2, \ldots, l_{n-1}$ stem from the third component of the pushed stack symbol ($l_1 \cdots l_{n-1}$), and $l_n$ is the label of the production whose DFA the pushing transition ends in.
-
-  For the inverse direction we construct a run by traversing the tree in a preorder\footnote{Where we also keep track of when we ``go back up'' the tree.}.
-Terminals in $\T$ cause transitions internally in a DFA, while terminals in $\Labels$ cause a pushing transition (and later a corresponding, unique popping transition) to the initial state of the DFA corresponding to the production with that label.
-Note that a sequence of terminals in $\Labels$ that have only a single child (which is also in $\Labels$) cause only one transition, all terminals but the last end up in the third component of the pushed stack symbol.
-The choice of pushing transition is \emph{not} unique however; there may be multiple available transitions that differ in the recognized input symbol and the second component of the pushed stack symbol, neither of which are necessarily determined by the shape of the tree.
-This non-determinism does not introduce multiple successful runs however, since two such runs would imply the existence of two distinct runs through the DFA corresponding to some labelled production that recognize the same word, except with different subscripts.
-But two otherwise equal terminals with different subscripts must originate from different occurrences of marked non-terminals in the original regular expression (since a $\top$ subscript represents a mark, and a $\bot$ subscript represents the absence of a mark), thus that regular expression must be ambiguous after all marks are removed.
-But Section~\ref{sec:parse-time-disambiguation} explicitly ruled this out, thus only one of the pushing transitions can be a valid choice.
-
-% TODO: technically, we should perhaps mention that these two constructions are inverses of each other, but that's relatively easy to see once you understand them.
-\end{proof}
-
-\subsubsection{Formal Stuff}
-
-Given a language definition $D$, with terminals $\T$, labels $\Labels$, and starting non-terminal $S$.
-Construct two functions $\dfa : \Labels -> \mathit{DFA}(\T \cup (\Labels \times \{\top, \bot\})$ and $\passthrough : \Labels -> 2^{\Labels \times \Labels \times \{\top, \bot\} \times \Labels^{*}}$:
+%%%%%%%%%%%%%%%%%%%%%%%%
+% FORMAL STUFF BARRIER %
+%%%%%%%%%%%%%%%%%%%%%%%%
+
+Given a language definition $D$, with terminals $\T$, non-terminals $\NT$, labels $\Labels$, and starting non-terminal $S$.
+We use $\labelnt(l)$ and $\labelregex(l)$ as the non-terminal and regex, respectively, of the production labelled $l$.
+We define the set $T : \Labels \times \Labels \times \{\top, \bot\} \times \Labels^{*}$ of transitions between productions inductively through:
+
+\begin{itemize}
+\item $(l, l, \bot, \epsilon) \in T$ for all $l \in \Labels$.
+\item $(l, l', l' \in m, l) \in T$ for all $N_m \in L(\labelregex(l))$ where $\labelnt(l') = N$.
+\item $(l_1, l_3, r_1 \lor r_2, w_1 \cdot w_2) \in T$ for all $(l_1, l_2, r_1, w_1), (l_2, l_3, r_2, w_2) \in T$.
+\end{itemize}
+
+\noindent The third component of each element denotes whether that transition introduces a required pair of parentheses ($\top$) or a possible pair ($\bot$).
+$T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
+Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
+
+We define the translation from production to DFA through a function $\dfa$ from a label to a DFA with the alphabet $\T \cup (\Labels \times \{\top, \bot\})$:
 
 \begin{description}
-\item[$\dfa(l)$:] Given the production $N -> l : r$, compute the regular language $L' := L(r) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
-Then replace every marked non-terminal $N'_m$ with the regular expression $l_{1l_1 \in m} + l_{2l_2 \in m} + \ldots + l_{nl_n \in m}$, where $l_i$ are all the labels of productions $N' -> l_i : r_i$, e.g., $E_{\{a\}}$ in the running example would be replaced by $l_\bot + a_\top + m_\bot + n_\bot$.
+\item[$\dfa(l)$:] Compute the regular language $L' := L(\labelregex(l)) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
+Then replace every marked non-terminal $N_m$ with the regular expression $(l_1, l_1 \in m) + (l_2, l_2 \in m) + \ldots + (l_n, l_n \in m)$, where $l_i$ are all the labels of productions $N -> l_i : r_i$.
+For example, $E_{\{a\}}$ in the running example would be replaced by $(l, \bot) + (a, \top) + (m, \bot) + (n, \bot)$.
 Then construct a DFA for this language.
-
-\item[$\passthrough(l)$:] Given a production $N -> l : r$, this is the set $\{(l, l, \bot, \epsilon)\} \cup \{(l, l', l' \in m, l) \mid \nt(l') = N', N'_m \in L(r)\}$.
 \end{description}
 
-\noindent Compute the set $P$ as the closure of $\bigcup_l \passthrough(l)$ using the following operation:
-
-$$ (l_1, l_2, r_1, w_1) \cdot (l_2, l_3, r_2, w_2) = (l_1, l_3, r_1 \lor r_2, w_1 \cdot w_2) $$
-
-\noindent $P$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
-Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
-
-Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$ for all $l \in Labels$, and $\nt(l) = N$ iff $(N -> l : r) \in D$, construct a VPDA $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
+\noindent We now define the visibly pushdown automaton $A_{\posp{}}$.
+Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
 
@@
     & \{ (s_{l_2}, (s, f, w))
-    & \nt(l_1) = S, (l_1, l_2, \top, w) \in P \} \\
+    & \labelnt(l_1) = S, (l_1, l_2, \top, w) \in P \} \\
 
@@
     & \{ (s_{l_2}, (s, f, w))
-    & \nt(l_1) = S, (l_1, l_2, \bot, w) \in P \} \\
+    & \labelnt(l_1) = S, (l_1, l_2, \bot, w) \in P \} \\
 
@@
 \noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is given in Section~\ref{sec:preliminaries-vpls} (using the partitions $\T_c = \{ \reqpl, \pospl \}$, $\T_i = \T$, and $\T_r = \{\reqpr\}$).
-If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$, then that run corresponds to two distinct runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
+If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
 We check for the existence of such a run by trimming the product automaton and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
@@
 There are thus cases where we can answer resolvably ambiguous or unresolvably ambiguous with certainty, and cases where we cannot answer either way with certainty.
+
+\begin{lemma}
+  There is a bijection between successful runs in $A_{\posp{}}$ and trees $t \in L(T_D)$.
+\end{lemma}
+
+\begin{proof}
+  (These are two attempts at showing the same thing, next attempt starts at ``Proof sketch''. I would strongly suggest reading ``Formal stuff'' first.)
+  $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
+The two are distinct, and $f$ is reachable only through transitions that pop a stack symbol with $s$ and $f$ as the two first components.
+Transitions that push such symbols originate in $s$, and no other transitions originate in $s$.
+Thus all successful runs start in configuration $(s, \epsilon)$ and end in $(f, \epsilon)$, whereby every pushing transition in the run can be paired with a corresponding popping transition.
+We will now construct a tree $t \in L(T_D)$ in a bottom-up fashion.
+
+  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, w)} (p', (p, q, w) \cdot s) \; R_2 \; (q', (p, q, w) \cdot s \xrightarrow{b, -(p, q, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
+The only transitions that leave the stack unchanged in $A_{\posp{}}$ come from DFAs generated from productions in $D$.
+There is no way to transition between two such DFAs without using a pushing or popping transition, thus $R_2$ must be a successful run in the DFA for some labelled production $N -> l : r$ in $D$.
+$T_D$ must contain a corresponding production $N -> l(r)$.
+Now consider the sequence of input symbols $w'$ along the transitions in $R_2$.
+We have $w' \in L(r)$, since $R_2$ is a successful run in a DFA constructed from $r$.
+Thus $l(w') \in L_N(T_D)$.
+Now add parent nodes to this subtree for every label in $w$, e.g., if $w = l_1 \cdot l_2 \cdot l_3$ construct $l_1(l_2(l_3(l(w'))))$.
+We can then replace the subsequence of $R$ with this subtree, and then repeat the process until we have a tree with root $l'$, where $D$ has a labelled production $S -> l': r'$ for some $r'$ and the starting symbol $S$.
+
+  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{\posp{}}$ starting from $(s, \epsilon)$.
+If the next terminal is in $\T$, i.e., a strictly zero-arity terminal, we take a transition originating in a DFA, which must be unique.
+If the next terminal is in $\Labels$, i.e., an unranked terminal, we take one of the added edges.
+However, this choice of edge might not be unique: there may be multiple edges from the current state to an appropriate DFA initial state.
+These stem from the original DFA having two transitions from the same state with the same label but different subscripts, e.g., $p \xrightarrow{l_\bot} q$ and $p \xrightarrow{l_\top} q'$.
+However, if both of these transitions were valid choices for the current tree that would imply that the DFA for that level can recognize two distinct words that differ only in subscripts.
+But distinctly subscripted terminals must originate from distinct occurances of marked non-terminals (since a $\top$ subscript corresponds to a mark, and a $\bot$ subscript corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
+
+% TODO: technically, we should perhaps mention that these two constructions are inverses of each other, but that's relatively easy to see once you understand them.
+\end{proof}
+
+\begin{lemma}
+  The word recognized by a successful run in $A_{\posp{}}$ is the unique linear encoding of the lattice for the corresponding $t \in L(T_D)$.
+\end{lemma}
+
+\begin{lemma}
+  For every successful run $R'$ in $A'_{\posp{}}$ there is one (or more) successful runs $R$ in $A_{\posp{}}$ such that the lattice encoded in the word recognized by $R'$ is larger than the lattice for the tree corresponding to $R$.
+\end{lemma}
+
+\begin{theorem}
+  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
+\end{theorem}
 
@@
 Each such language turns out to be a visibly pushdown language.
-Given a set of trees in $L(T_D)$ we can construct a corresponding set of non-overlapping visibly pushdown automata (i.e., each automaton only accepts words not accepted by any other automaton) since VPLs are closed under complement and intersection \cite{alurVisiblyPushdownLanguages2004}.
-These automata can then be examined to determine if they accept the empty language (which implies that the corresponding tree only has ambiguous words), or otherwise produce a witness, a word accepted only by that automaton.
+VPLs are closed under difference, which enables us to create a visibly pushdown automaton for the language $\words(t_1) \setminus \words(t_2)$ for any pair of trees $t_1, t_2 \in L(T_D)$.
+This is useful, since $\words(t_1)$ is the set of possible resolutions for $t_1$ in an ambiguity, and $\words(t_1) \cap \words(t_2)$ is the set of ambiguous words that can parse as both $t_1$ and $t_2$, i.e., words that \emph{cannot} resolve an ambiguity.
+
+Intuitively we then need to, for each tree $t \in \parse(w)$, determine if $\words(t) \setminus \bigcup_{t' \in L(T_D) \land t' \neq t} \words(t)$ is the empty set.
+Unfortunately, $L(T_D)$ tends to be infinite.
+However, for any particular $t$, most trees $t' \in L(T_D)$ are irrelevant: their words do not overlap at all; $\words(t) \cap \words(t') = \emptyset$.
+Hence we construct a (finite) subset of $L(T_D)$ such that all other trees are irrelevant.
+As a bonus, this approach constructs a VPDA per tree that recognizes valid resolutions for that tree; a minimal witness is thus a shortest word recognized by a VPDA, and the problem of finding such a word is a known one.
 
@@
 
-\begin{theorem}
+\begin{lemma}
   Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
-\end{theorem}
+\end{lemma}
 
@@
 The former can be solved by either tracking the subtree's origin when parsing, or reparsing with the right hand side regular expression, while the latter is solved by introducing a new stack symbol $\gamma$ that is pushed and popped by all non-grouping parenthesis terminals.
+We now present the function $f : L(T_D) -> VPDA$ that performs this construction.
 Writing the transition function $\delta : Q \times \T \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ as a subset of $Q \times \T \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, and using $\sqcup$ for disjoint union:
@@
 $$
+
+\begin{lemma}
+  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
+\end{lemma}
 
@@
 the precedence list with explicit \syncon{forbid} declarations,
-``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right
-  = add}'' (cf. the mark $\{a\}$ in the production of $m$ in
+``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right = add}''
+(cf. the mark $\{a\}$ in the production of $m$ in
 Fig.~\ref{fig:running-example-definition}).

### 2019-07-10 paper/paper.tex
Formalism update

@@
 \begin{abstract}
-A common point of view when designing grammars for parsing is that
-the grammar definition has to be unambiguous.  However, up front unambiguous grammars
-can force language designers to make arbitrary choice to disambiguate
-the language. In this paper, we depart from the traditional view of
+% Why?
+
+Ever since the early 60s, the problem of deciding if a context-free
+grammar is unambiguous or not has been known to be undecidable. As a
+consequence, a large number of restricted grammars have been developed
+to guarantee that a language definition is unambiguous. This
+traditional view of only allowing unambiguous grammars has until
+recently been taken for granted as the only truth: \emph{the} way of
+how the syntax of a language must be defined. However, up front
+unambiguous grammars can force language designers to make arbitrary
+choice to disambiguate the language, although the natural choice may be to
+postpone the decision to the programmer. Moreover, an important
+objective of domain-specific language development is to construct
+languages by composing and extending existing
+languages. Unfortunately, the composition of restricted unambiguous
+grammars easily results in ambiguous grammars, which again are
+undecidable. In this paper, we depart from the traditional view of
 unambiguous grammar design, and allow ambiguities to be delayed until
 compile time, allowing the user to perform the disambiguation. A
-natural decision problem follows: given a language definition, can a
+natural decision problems follows: given a language definition, can a
 user always disambiguate an ambiguous program?  We introduce and
@@
 case for dynamic resolvability. The approach is evaluated through
-two separate case studies, covering both a large existing
-programming language, and the composability of domain-specific
-languages.
+three separate case studies, covering both a large
+existing programming language, and the composability of
+domain-specific languages.
+%A traditional view when designing programming languages is that syntax
+%definitions should be unambiguous: only unambiguous grammars are used
+%to construct language specific parsers that either rejects a program,
+%or accepts it and produces a unique parse tree. However,
 \end{abstract}
@@
 
-% Motivate why unambiguous parsing is a good (and important) idea.
-Ever since the early 60s, the problem of deciding if a context-free
-grammar is ambiguous has been known to be
-undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
-a large number of restricted grammars have been developed to guarantee
-that a language definition is unambiguous. This traditional view of
-only allowing unambiguous grammars has been taken for granted as the
-only truth: \emph{the} way of how the syntax of a language must be
-defined
-%This seems to stem
-%from the view that ambiguity has no place in programming language
-%syntax, which is highly prevalent
-~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
-%
-However, recent work by
-~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
-carefully designed ambiguous syntax definitions are in some cases
-preferable compared to completely unambiguous
-definitions. Specifically, instead of forcing a language designers to
-make arbitrary choice to disambiguate the language, a more natural
-choice may be to postpone the decision to the programmer.
-
-% Motivation II, compising DSLs
-Another area---where ambiguous grammars naturally arises---is
-domain-specific language development.  An important
-objective of domain-specific language development is to construct
-languages by composing and extending existing
-languages.
-Unfortunately, the composition of restricted
-%grammars easily results in ambiguous grammars, which again are
-%undecidable.
-%
-%constructing domain-specific programming languages there is often
-%a desire to reuse work, to compose smaller language
-%fragments.
-%However, how to make such compositions well-behaved is not
-%always obvious.
-%In particular, when constructing new domain-specific languages from
-%existing languages, by composing multiple syntax definitions,
-grammars easily produces
-an ambiguous grammar, even if all of the original grammars are
-unambiguous. Approaches for solving this problem can be broadly split
-into two categories: those that handle ambiguities on the
-grammar-level (detection, prevention, etc.), and those that work on
-particular examples of ambiguous programs. The former is well explored
-in the form of heuristics for ambiguity detection
-(e.g. \cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007})
-or restrictions on the grammars to be composed
-(e.g. \cite{kaminskiModularWellDefinednessAnalysis2013}), while the
-latter has received relatively little attention.
-
-%Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
-\cite{earleyEfficientCo%ntextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
+When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments.
+However, how to make such compositions well-behaved is not always obvious.
+In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous.
+Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs.
+The former is well explored in the form of heuristics for ambiguity detection (e.g.
+\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g.
+\cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny.
+This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g.
+\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).
+
+Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
+\cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
 However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity.
@@
 
-%In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists.
+In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists.
 For example, in a language with semicolon for both sequential composition and element separation in a list (e.g.
@@
 The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.
-
-% The problem
-
-This paper focuses on the problems that arises if a grammar definition
-is not guaranteed to be unambiguous. Specifically, if parsing a
-program results in several parse trees, the grammar was obviously
-ambiguous. The programmer then needs to disambiguate the program by,
-for instance, inserting extra parenthesis. However, is it always
-possible to resolve an ambiguous input program by augmenting the
-program? We say that a program has a \emph{resolvable ambiguity} if
-the ambiguity can be solved by augmenting the program.
-The \emph{resolvable ambiguity problem} can then be divided into several decision problems: (i) the dynamic decidability problem: given a program (or a
-
-dynamic resolvable ambiguity problem of deciding if a context-free grammar is
-
 This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
@@
 
-\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure-sound}
+% todo: improve label (where used?)
+\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure}
   A decision procedure $f$ solving the static resolvability problem is sound iff
@@
   parsers } P$. It turns out, we can give $g$ for the parse languages
-studied in the following sections.
-
-\begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
+studied in following sections.
+
+\begin{definition}[Completeness of Static Resolvability]
   A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff
@@
 
-\noindent Similarly, we define soundness and completeness for the dynamic problem.
-
-\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure-sound}
-  A decision procedure $f$ solving the dynamic resolvability problem is sound iff
-
-  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
-\end{definition}
-
-Analogous to the static case, for the parse languages studied in the
-following sections we can also provide a decision procedure $g$ such
-that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
-\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.
-
-\begin{definition}[Completeness of Dynamic Resolvability]\label{def:dynamic-procedure-complete}
-  A decision procedure $f$ solving the dynamic resolvability problem is complete iff
-
-  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
+\noindent Similarly, for the dynamic problem:
+
+\begin{definition}
+  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
+  \begin{itemize}
+  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
+  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
+  \end{itemize}
+  \label{def:dynamic-procedure}
 \end{definition}
@@
 
-We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}.
-Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem with one caveat: we only consider languages with balanced parentheses.
+We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure} and \ref{def:dynamic-procedure}.
+Section~\ref{sec:static} gives a partial solution to the former problem, while Section~\ref{sec:dynamic} fully solves the latter with one caveat: we only consider languages with balanced parentheses.
 

### 2019-07-10 paper/paper.tex
Soundness and completeness of dynamic resolvability

--- 2019-07-10 paper/paper.tex
@@
   parsers } P$. It turns out, we can give $g$ for the parse languages
-studied in following sections.
+studied in the following sections.
 
@@
 
-\noindent Similarly, for the dynamic problem:
-
-\begin{definition}
-  A decision procedure $f$ (with possible outputs $\mathit{resolvable}$, $\mathit{unresolvable}$, and $\mathit{unknown}$) solving the dynamic resolvability problem is:
-  \begin{itemize}
-  \item Sound if $f(\parse, w) = \mathit{resolvable}$ implies that $w$ is resolvably ambiguous, and $f(\parse, w) = \mathit{unresolvable}$ implies $w$ is not resolvably ambiguous.
-  \item Complete if $f(\parse) \neq \mathit{unknown}$ for all possible inputs $\parse$ and $w$.
-  \end{itemize}
-  \label{def:dynamic-procedure}
+\noindent Similarly, we define soundness and completeness for the dynamic problem.
+
+% todo: improve label (where used?)
+\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure}
+  A decision procedure $f$ solving the dynamic resolvability problem is sound iff
+
+  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
+\end{definition}
+
+Analogous to the static case, for the parse languages studied in the
+following sections we can also provide a decision procedure $g$ such
+that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
+\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.
+
+\begin{definition}[Completeness of Dynamic Resolvability]
+  A decision procedure $f$ solving the dynamic resolvability problem is complete iff
+
+  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
 \end{definition}

### 2019-07-10 paper/paper.tex
Fix labels of static and dynamic resolvability definitions

@@
 
-\begin{definition}[Completeness of Static Resolvability]
+\begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
   A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff
@@
 
-% todo: improve label (where used?)
-\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure}
+\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure-sound}
   A decision procedure $f$ solving the dynamic resolvability problem is sound iff
@@
 
-\begin{definition}[Completeness of Dynamic Resolvability]
+\begin{definition}[Completeness of Dynamic Resolvability]\label{def:dynamic-procedure-complete}
   A decision procedure $f$ solving the dynamic resolvability problem is complete iff

### 2019-07-10 paper/paper.tex
EOD

@@
 \newcommand{\labelregex}{\mathit{regex}} % Go from label the regex of the production with that label
+\newcommand{\treetorun}{\mathit{run}}
+\newcommand{\runtotree}{\mathit{tree}}
+\newcommand{\runtoword}{\mathit{word}}
+\newcommand{\runprimetorun}{\mathit{normalrun}}
 
@@
 \begin{abstract}
-% Why?
-
-Ever since the early 60s, the problem of deciding if a context-free
-grammar is unambiguous or not has been known to be undecidable. As a
-consequence, a large number of restricted grammars have been developed
-to guarantee that a language definition is unambiguous. This
-traditional view of only allowing unambiguous grammars has until
-recently been taken for granted as the only truth: \emph{the} way of
-how the syntax of a language must be defined. However, up front
-unambiguous grammars can force language designers to make arbitrary
-choice to disambiguate the language, although the natural choice may be to
-postpone the decision to the programmer. Moreover, an important
-objective of domain-specific language development is to construct
-languages by composing and extending existing
-languages. Unfortunately, the composition of restricted unambiguous
-grammars easily results in ambiguous grammars, which again are
-undecidable. In this paper, we depart from the traditional view of
+A common standpoint when designing the syntax of programming languages is that
+the grammar definition has to be unambiguous.  However, up front unambiguous grammars
+can force language designers to make arbitrary choice to disambiguate
+the language. In this paper, we depart from the traditional view of
 unambiguous grammar design, and allow ambiguities to be delayed until
 compile time, allowing the user to perform the disambiguation. A
-natural decision problems follows: given a language definition, can a
+natural decision problem follows: given a language definition, can a
 user always disambiguate an ambiguous program?  We introduce and
@@
 
-When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments.
-However, how to make such compositions well-behaved is not always obvious.
-In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous.
-Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs.
-The former is well explored in the form of heuristics for ambiguity detection (e.g.
-\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g.
-\cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny.
-This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g.
-\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).
-
-Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
-\cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
+% Motivate why unambiguous parsing is a good (and important) idea.
+Ever since the early 60s, it is known to be undecidable to determine whether a context-free
+grammar is ambiguous~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
+a large number of restricted grammars have been developed to guarantee
+that a language definition is unambiguous. This traditional view of
+only allowing unambiguous grammars has been taken for granted as the
+only truth: \emph{the} way of how the syntax of a language must be
+defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
+%
+However, recent work by
+~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
+carefully designed ambiguous syntax definitions are in some cases
+preferable compared to completely unambiguous
+definitions. Specifically, instead of forcing language designers to
+make arbitrary choice to disambiguate the language, a more natural
+choice may be to postpone the decision to the programmer.
+
+% Motivation II, compising DSLs
+Another area---where ambiguous grammars naturally arises---is
+domain-specific language development.  An important
+objective when designing domain-specific languages is to construct
+languages by composing and extending existing
+languages.
+Unfortunately, the composition of restricted
+%grammars easily results in ambiguous grammars, which again are
+%undecidable.
+%
+%constructing domain-specific programming languages there is often
+%a desire to reuse work, to compose smaller language
+%fragments.
+%However, how to make such compositions well-behaved is not
+%always obvious.
+%In particular, when constructing new domain-specific languages from
+%existing languages, by composing multiple syntax definitions,
+grammars easily produces
+an ambiguous grammar, even if all of the original grammars are
+unambiguous. Approaches for solving this problem can be broadly split
+into two categories: those that handle ambiguities on the
+grammar-level (detection, prevention, etc.), and those that work on
+particular examples of ambiguous programs. The former is well explored
+in the form of heuristics for ambiguity detection
+\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}
+or restrictions on the composed grammars
+\cite{kaminskiModularWellDefinednessAnalysis2013}, while the
+latter has received relatively little attention.
+
+%Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
+\cite{earleyEfficientCo%ntextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
 However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity.
@@
 The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.
-This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
+
+% The problem
+
+This paper focuses on the problems that arise if a grammar definition
+is not guaranteed to be unambiguous. Concretely, if parsing a program
+results in several parse trees, the grammar is obviously
+ambiguous. The programmer then needs to disambiguate the program by,
+for instance, inserting extra parenthesis. We say that a program has a
+\emph{resolvable ambiguity} if the ambiguity can be solved by
+modifying the program, such that each of the possible parse trees can
+be selected using different modifications.
+We can divide the problem
+related to resolvable ambiguity into several different decision
+problems.
+The \emph{dynamic resolvable ambiguity problem} is as follows: given a parse function for a language, and a specific program that is parsed into a set of parse trees, decided whether it possible to rewrite the program, such that each of the parse trees can be generated unambiguously.
+This problem can be generalized into the \emph{static resolvable ambiguity problem}: decide whether all possible parsable programs are resolvably ambiguous.
+Static resolvablity means that a language designer can be confident that the end-user (the program designer) always can resolve an ambiguity by modifying the program.
+By contrast, dynamic resolvability means that a language designed knows if a \emph{specific} program is resolvable.
+Hence, static resolvability gives formally verified guarantees on the syntax definition, whereas dynamic resolvability enables the language designer to test the resolvability of the syntax definition, program by program.
+
+%This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
 We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot.
@@
 
-Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)?
+%Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)?
 For classical ambiguity in context-free grammars, the former problem is solved by any general parser, while the latter is undecidable \cite{cantorAmbiguityProblemBackus1962}.
@@
 
-More specifically, this paper makes the following contributions:
+In this paper, we solve the static resolvability problems for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when designing language syntaxes.
+More specifically, we make the following contributions:
 
 \begin{itemize}
-\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp.
-word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
-\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
-  \begin{itemize}
-  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
-  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
+\item We formally define the term \emph{resolvable ambiguity}, as well as giving precise formalizations of the static and dynamic \emph{resolvable ambiguity problems} (Section~\ref{sec:resolvable-definition}).
+\item We formulate the \emph{static resolvability problem} of increasing difficulty, by introducing two different restricted language versions.
+We formalize a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
+\item We define a decidable algorithm that solves the \emph{dynamic resolvability problem} for general context-free languages with well-balanced parentheses (Section~\ref{sec:dynamic}).
+\item The dynamic resolvability methodology is implemented as a tool and a small DSL for defining syntax definitions.
+Two different cases studies are evaluated using this methodology and tool (Section~\ref{sec:evaluation}): (i) an implementation and evaluation of a large subset of OCaml's syntax, where we study the effect of implementing an underspecified language definition, and (ii) a domain-specific language for orchestrating parallel computations, where we investigate the effects of composing language fragments.
 \end{itemize}
-\item Three case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
-  \begin{itemize}
-  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
-  \item A domain-specific language for orchestrating parallel computations (Section~\ref{sec:evaluation-orc}).
-  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
-  \end{itemize}
-  % Briefly say something about the tool, that it lies ``on top of'' what's presented herein
-\end{itemize}
+
+\noindent As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF), where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
+Before the above contributions are introduced, the paper starts with motivating examples (Section~\ref{sec:motivation}), and a section with preliminaries (Section~\ref{sec:prel}).
 
 \section{Motivating Example}
-
+\label{sec:motivation}
 This section walks through an example from OCaml where an ambiguity error can be clearer than what the actual compiler produces, suggesting that (resolvable) ambiguity has merit even in the context of a general purpose programming language.
@@
 \section{Preliminaries}
+\label{sec:prel}
 
@@
 \begin{proof}
-  (These are two attempts at showing the same thing, next attempt starts at ``Proof sketch''. I would strongly suggest reading ``Formal stuff'' first.)
-  $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
-The two are distinct, and $f$ is reachable only through transitions that pop a stack symbol with $s$ and $f$ as the two first components.
-Transitions that push such symbols originate in $s$, and no other transitions originate in $s$.
-Thus all successful runs start in configuration $(s, \epsilon)$ and end in $(f, \epsilon)$, whereby every pushing transition in the run can be paired with a corresponding popping transition.
-We will now construct a tree $t \in L(T_D)$ in a bottom-up fashion.
-
-  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, w)} (p', (p, q, w) \cdot s) \; R_2 \; (q', (p, q, w) \cdot s \xrightarrow{b, -(p, q, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
-The only transitions that leave the stack unchanged in $A_{\posp{}}$ come from DFAs generated from productions in $D$.
-There is no way to transition between two such DFAs without using a pushing or popping transition, thus $R_2$ must be a successful run in the DFA for some labelled production $N -> l : r$ in $D$.
-$T_D$ must contain a corresponding production $N -> l(r)$.
-Now consider the sequence of input symbols $w'$ along the transitions in $R_2$.
-We have $w' \in L(r)$, since $R_2$ is a successful run in a DFA constructed from $r$.
-Thus $l(w') \in L_N(T_D)$.
-Now add parent nodes to this subtree for every label in $w$, e.g., if $w = l_1 \cdot l_2 \cdot l_3$ construct $l_1(l_2(l_3(l(w'))))$.
-We can then replace the subsequence of $R$ with this subtree, and then repeat the process until we have a tree with root $l'$, where $D$ has a labelled production $S -> l': r'$ for some $r'$ and the starting symbol $S$.
-
-  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{\posp{}}$ starting from $(s, \epsilon)$.
-If the next terminal is in $\T$, i.e., a strictly zero-arity terminal, we take a transition originating in a DFA, which must be unique.
-If the next terminal is in $\Labels$, i.e., an unranked terminal, we take one of the added edges.
-However, this choice of edge might not be unique: there may be multiple edges from the current state to an appropriate DFA initial state.
-These stem from the original DFA having two transitions from the same state with the same label but different subscripts, e.g., $p \xrightarrow{l_\bot} q$ and $p \xrightarrow{l_\top} q'$.
-However, if both of these transitions were valid choices for the current tree that would imply that the DFA for that level can recognize two distinct words that differ only in subscripts.
-But distinctly subscripted terminals must originate from distinct occurances of marked non-terminals (since a $\top$ subscript corresponds to a mark, and a $\bot$ subscript corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
-
-% TODO: technically, we should perhaps mention that these two constructions are inverses of each other, but that's relatively easy to see once you understand them.
+  By defining two functions $\treetorun$ and $\runtotree$, the former from a tree to a run, the latter from run to tree, then showing these two functions to be inverses of each other.
 \end{proof}
 
+TODO: define a \emph{well-formed} linear encoding as a linear encoding without double parens. Ensure $\runtoword$ and $\lattice$ are defined.
+
 \begin{lemma}
-  The word recognized by a successful run in $A_{\posp{}}$ is the unique linear encoding of the lattice for the corresponding $t \in L(T_D)$.
+  All words $w \in L(A_{\posp{}})$ are well-formed linear encodings.
 \end{lemma}
@@
 \begin{lemma}
-  For every successful run $R'$ in $A'_{\posp{}}$ there is one (or more) successful runs $R$ in $A_{\posp{}}$ such that the lattice encoded in the word recognized by $R'$ is larger than the lattice for the tree corresponding to $R$.
+  $\forall t \in L(T_D).\ \lattice(\runtoword(\treetorun(t))) = \lattice(t)$.
 \end{lemma}
 
+\begin{lemma}
+  For all $t \in L(T_D)$ and every well-formed lattice encoding $w$ such that $\lattice(t) \subseteq \lattice(w)$, $w \in L(A'_{\posp{}})$.
+\end{lemma}
+
+\begin{lemma}
+  We can define a function $\runprimetorun$ from runs in $A'_{\posp{}}$ to runs in $A_{\posp{}}$ such that $\forall R.\ \lattice(\runtoword(\runprimetorun(R))) \subseteq \lattice(\runtoword(R))$.
+\end{lemma}
+
+\begin{proof}
+  By removing every transition that pushes or pops $\gamma$ and replacing every used transition $p \xrightarrow{'\pospl', +g} q$ that does not exist in $A_{\posp{}}$ with a transition $p \xrightarrow{'\reqpl', +g} q$.
+\end{proof}
+
 \begin{theorem}
-  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
+  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, and $\runtoword(R) = \runtoword(R')$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
 \end{theorem}
+
+\begin{proof}
+  Assume $\runprimetorun(R') = R$.
+$\runprimetorun$ either leaves its argument unchanged, or changes the recognized word.
+If $R'$ and $R$ recognize the same word, and $\runprimetorun(R') = R$, then $R' = R$, but that is a contradiction.
+
+  Since $\runprimetorun(R') \neq R$ we have two distinct trees $t_1 = \runtotree(\runprimetorun(R'))$ and $t_2 = \runtotree(R)$.
+Since $\lattice(\runtoword(\runprimetorun(R'))) \subseteq \lattice(\runtoword(R'))$ and $\lattice(\runtoword(R')) = \lattice(\runtoword(R))$ we have that $\lattice(t_1) \subseteq \lattice(t_2)$.
+\end{proof}
 
@@
   Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
+  \label{lemma:vpda-from-t}
 \end{lemma}
@@
   Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
+\end{lemma}
+
+\begin{proof}
+  Follows trivially from Lemma~\ref{lemma:vpda-from-t} and VPDAs being closed under union and difference.
+\end{proof}
+
+\begin{lemma}
 \end{lemma}

### 2019-07-10 paper/supplementary.tex
EOD

--- 2019-07-10 paper/paper.tex
@@
 
-\section{Introduction}
-
-When constructing domain-specific programming languages there is often a desire to reuse work, to compose smaller language fragments.
-However, how to make such compositions well-behaved is not always obvious.
-In particular, composing multiple syntaxes is likely to produce an ambiguous grammar, even if all of the original grammars are unambiguous.
-Approaches for solving this problem can be broadly split into two categories: those that handle ambiguities on the grammar-level (detection, prevention, etc.), and those that work on particular examples of ambiguous programs.
-The former is well explored in the form of heuristics for ambiguity detection (e.g.
-\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}) or restrictions on the grammars to be composed (e.g.
-\cite{kaminskiModularWellDefinednessAnalysis2013}), while the latter has received relatively little scrutiny.
-This seems to stem from the view that ambiguity has no place in programming language syntax, which is highly prevalent (e.g.
-\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}).
-
-Handling ambiguity at parse-time on the other hand, beyond just using a general parsing algorithm (e.g.
-\cite{earleyEfficientContextfreeParsing1970,scottGLLParsing2010,youngerRecognitionParsingContextfree1967}), seems limited to merely presenting each possible interpretation as an abstract syntax tree of some form \cite{palmkvistCreatingDomainSpecificLanguages2019,danielssonParsingMixfixOperators2011}.
-However, an abstract syntax tree is essentially an implementation detail, and might not provide any useful clues for an end-user to resolve the ambiguity.
-Another possibility, mentioned by \citet{palmkvistCreatingDomainSpecificLanguages2019} but not yet explored, is to automatically generate several unambiguous programs, one for each possible interpretation, thereby directly showing a user how to resolve the ambiguity, regardless of their desired interpretation.
-For example, in a language without precedence, \verb|'1 + 2 * 3'| is ambiguous, but the two possible interpretations can be written unambiguously as \verb|'(1 + 2) * 3'| and \verb|'1 + (2 * 3)'| respectively.
-
-In exploring this approach, we find that this is not always possible; some ambiguities have one or more interpretations for which no unambiguous program exists.
-For example, in a language with semicolon for both sequential composition and element separation in a list (e.g.
-OCaml), ``$[1; 2]$'' is ambiguous: it is either a list of two elements, or a list of one element, namely the sequential composition of $1$ and $2$.
-The latter can be written unambiguously as ``$[(1; 2)]$'', but the former has no unambiguous alternative.
-This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
-We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot.
-We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}.
-This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.
-
-% TODO: I feel like there should be a clear transition between the more pracctical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.
-
-% TODO: resolvable ambiguity cannot be defined solely in terms of a grammar, we require interpretations of words as well, thus we have to introduce a new formalism that does this, the formalism is an intuitive extension of EBNF grammars.
-this is probably written as a contrast with classical ambiguity, which should be described briefly for contrast
-
-Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)?
-For classical ambiguity in context-free grammars, the former problem is solved by any general parser, while the latter is undecidable \cite{cantorAmbiguityProblemBackus1962}.
-For resolvable ambiguity, in a formalism that associates trees with EBNF-grammars, the dynamic problem is similarly solvable, and the static problem is similarly difficult.
-We do not yet know if the latter is undecidable in general, but we have found a subclass of languages for which it is decidable.
-
-More specifically, this paper makes the following contributions:
-
-\begin{itemize}
-\item A formal definition of \emph{resolvable ambiguity} in terms of languages where words have associated interpretations, as well as formulations of the \emph{static} and \emph{dynamic} resolvability problems, i.e., determine if a grammar (resp.
-word) is resolvably ambiguous (Section~\ref{sec:resolvable-definition}).
-\item A syntax definition formalism based on EBNF, where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}). For this formalism we also provide:
-  \begin{itemize}
-  \item A formulation of three versions of the static resolvability problem of increasing difficulty, and a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
-  \item A decidable algorithm that solves the dynamic resolvability problem for languages that have only well-balanced parentheses (Section~\ref{sec:dynamic}).
-  \end{itemize}
-\item Three case studies of programming languages and the errors produced when we allow (resolvable) ambiguity:
-  \begin{itemize}
-  \item A large subset of OCaml where some ambiguities have been reintroduced (Section~\ref{sec:evaluation-ocaml}).
-  \item A domain-specific language for orchestrating parallel computations (Section~\ref{sec:evaluation-orc}).
-  \item A domain-specific language for cyber-physical systems (Section~\ref{sec:evaluation-cyphym}).
-  \end{itemize}
-  % Briefly say something about the tool, that it lies ``on top of'' what's presented herein
-\end{itemize}
-
-\section{Motivating Example}
-
-This section walks through an example from OCaml where an ambiguity error can be clearer than what the actual compiler produces, suggesting that (resolvable) ambiguity has merit even in the context of a general purpose programming language.
-
-Consider the following nested match expression (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
-
-\begin{minipage}{.35\textwidth}
-\begin{lstlisting}[language={[objective]caml},numbers=left]
-match 1 with
-  | 1 -> match "one" with
-         | str -> str
-  | 2 -> "two"
-\end{lstlisting}
-\end{minipage}
-\vrule
-\begin{minipage}{\textwidth}
-\begin{lstlisting}
-  File "./nestmatch.ml", line 4, characters 4-5:
-  Error: This pattern matches values of type int
-         but a pattern was expected which
-         matches values of type string
-\end{lstlisting}
-\end{minipage}
-
-\noindent The compiler sees the last line as belonging to the inner \ocaml{match} rather than the outer, as was intended. The solution is simple; we put parentheses around the inner match:
-
-\begin{minipage}{\textwidth}
-\begin{lstlisting}[language={[objective]caml},numbers=left]
-match 1 with
-  | 1 -> (match "one" with
-          | str -> str)
-  | 2 -> "two"
-\end{lstlisting}
-\end{minipage}
-
-\noindent However, the connection between error message and solution is not particularly clear; surrounding an expression with parentheses does not change its type.
-
-Instead, we look to the OCaml manual for inspiration.
-It contains an informal description of the syntax of the language\footnote{\url{https://caml.inria.fr/pub/docs/manual-ocaml/language.html}}, in the form of an EBNF-like grammar.
-Below is an excerpt of the productions for expressions, written in a more standard variant of EBNF:
-
-\setlength{\grammarindent}{9.5em}
-\begin{grammar}
-<expr> ::= \verb|'match'| <expr> \verb|'with'| <pattern-matching>
-
-<pattern-matching> ::= \verb{'|'{$^{?}$ <pattern> (\verb|'when'| <expr>)$^{?}$ \\
-(\verb{'|'{ <pattern> (\verb|'when'| <expr>)$^{?}$ \verb|'->'| <expr>)*
-\end{grammar}
-
-\noindent If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match.
-The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended.
-Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user.
-Instead, we can present the alternatives in the form of modified code that parses unambiguously as each corresponding alternative:
-
-\begin{center}
-\begin{tabular}{l|l}
-\begin{lstlisting}[language={[objective]caml}]
-match 1 with
-  | 1 -> (match "one" with
-          | str -> str)
-  | 2 -> "two"
-\end{lstlisting} &
-\begin{lstlisting}[language={[objective]caml}]
-match 1 with
-  | 1 -> (match "one" with
-          | str -> str
-  | 2 -> "two")
-\end{lstlisting} \\
-\end{tabular}
-\end{center}
-
-\section{Preliminaries}
-
-This section briefly describes the theoretical foundations we build upon.
-Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation to make later sections easier to read.
-Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}.
-Finally, Sections~\ref{sec:preliminaries-trees} and \ref{sec:preliminaries-ambiguity} describe trees and ambiguity, the former with some non-standard notation and the latter with a slightly wider definition than normal.
-
-\subsection{Bags} \label{sec:bags}
-
-A \emph{bag}, or \emph{multiset}, is a generalization of a set; it is an unordered collection where each element may appear multiple times.
-The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$.
-A bag $A$ is included in another bag $B$, written $A \subseteq B$, iff $\forall x.\ m_A(x) \leq m_B(x)$.
-The underlying set of elements of a bag $B$ is called the \emph{support} of $B$, and is given by $\support(B) = \{ x \mid m_B(x) > 0 \}$.
-
-\subsection{Regular Expressions} \label{sec:preliminaries-regexes}
-
-A regular expression $r$ over alphabet $\T$ is defined inductively:
-
-$$r ::= \epsilon \mid a \mid r \cdot r \mid r + r \mid r^{*}$$
-
-\noindent where $a \in \T$. The language of a regular expression $r$ is given by $L(r)$:
-
-$$
-\begin{array}{r@{\ =\ }l}
-  L(t) & \{a\} \\
-  L(\epsilon) & \{\epsilon\} \\
-  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \} \\
-  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
-  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
-\end{array}
-$$
-
-\noindent We denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
-
-\subsection{Context-Free Grammars} \label{sec:preliminaries-cfgs}
-
-A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is the Kleene star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
-
- A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
-Such a sequence is called a \emph{derivation}.
-The set of words recognized by $G$ is written $L(G)$\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.
-
-\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
-
-The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{leftmost derivations}.
-A leftmost derivation is a derivation where the non-terminal being replaced is always the leftmost one.
-
-\begin{definition}
-A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of that word.
-\end{definition}
-
-\subsection{Automata} \label{sec:preliminaries-automata}
-
-A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$ ; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
-
-A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that:
-
-\begin{itemize}
-\item $r_0 = q_0$.
-\item $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$.
-\item $r_n \in F$.
-\end{itemize}
-
-\noindent We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a succesful run.
-
-A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to.
-NFAs and DFAs have the same expressive power as regular expressions, i.e., for every regular expression there is an NFA and a DFA both reconizing the same language, and vice-versa.
-% TODO: probably reference for this
-
-A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols.
-Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
-$\lambda$ essentially means ``ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).
-
-A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma^{*}$.
-
-However, in this paper we will only consider pushdown automata with relatively limited stack manipulation, and will thus use some convenient shorthand:
-
-\begin{itemize}
-\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all, i.e., $\delta(p, a, \lambda) \supseteq \{(q, \lambda)\}$.
-\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack, i.e., $\delta(p, a, \lambda) \supseteq \{(q, g)\}$.
-\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack, i.e., $\delta(p, a, g) \supseteq \{(q, \lambda)\}$.
-\end{itemize}
-
-\subsection{Visibly Pushdown Languages} \label{sec:preliminaries-vpls}
-
-A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
-A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton has one of the following three forms:
-
-\begin{itemize}
-\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$; or
-\item $p \xrightarrow{i} q$, where $i \in \T_i$; or
-\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$,
-\end{itemize}
-
-\noindent i.e., the terminal recognized by a transition fully determines the change to the stack height.
-
-The names of the partitions stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
-
-This partitioning gives us some useful properties. Of particular relevance to this paper are the following two points:
-
-\begin{itemize}
-\item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}.
-Intersection in particular is given by a product automaton, i.e., given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
-  $$
-  \begin{array}{r@{\ =\ }ll}
-    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \text{where } c \in \T_c \\
-    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \text{where } i \in \T_i \\
-    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \text{where } r \in \T_r \\
-  \end{array}
-  $$
-
-\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
-Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
-\end{itemize}
-
-\subsection{Unranked Regular Tree Grammars} \label{sec:preliminaries-trees}
-
-Trees generalize words by allowing each terminal to have multiple ordered successors, instead of just zero or one.
-Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors.
-This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed.
-The sequence of successors to a single terminal in an unranked tree tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.
-
-The results and properties presented in this paper are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
-We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves.
-Since we will use unranked trees to represent parse trees, the former will represent terminals from the parsed word, while the latter represent terminals introduced as internal nodes.
-
-An unranked tree grammar $T$ is a tuple $(\NT, \T, X, P, S)$ where:
-
-\begin{itemize}
-\item $\NT$ is a set of (zero-arity) non-terminals.
-\item $\T$ is a set of zero-arity terminals, used as leaves.
-\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
-\item $P$ is a set of productions, a finite subset of $\NT \times X \times \regex(\T \cup X)$. We will write a production $(N, x, r)$ as $N -> x(r)$.
-\end{itemize}
-
-\noindent A tree $t$ (containing only terminals from $\T$ and $X$) is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
-The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.
-
-Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree.
-Informally, it is the flattening of a tree after all internal nodes have been removed.
-
-\section{Resolvable Ambiguity} \label{sec:resolvable-definition}
-
-This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to standard concepts in formal languages.
-
-A formal language is defined as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$.
-For defining \emph{resolvable ambiguity} we additionally have to consider the results of parsing words.
-In order to stay as general as possible, we first define the notion of an \emph{abstract parser}.
-
-\begin{definition}
-  An abstract parser $P$ is a triple $(\T, T, \parse)$ consisting of
-\begin{itemize}
-\item an alphabet $\T$,
-\item a set of parse trees $T$, and
-\item a function $\parse : \T^{*} -> 2^T$ that relates words to their parse trees, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
-\end{itemize}
-\end{definition}
-
-This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}
-
-\begin{definition}
-  Given a word $w \in \T^{*}$ and an abstract parser $P = (\T, T,
-  \parse)$, the set of words contained in the \emph{parse language}
-  $L(P)$ is defined as follows:
-
-  $w \in L(P)$ iff $\parse(w) \neq \emptyset$.
-\end{definition}
-
-\noindent For example, consider a simple arithmetic language without precedence and parentheses.
-In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two parse trees shown in Fig.~\ref{fig:arith-example}.
-
-\begin{figure}[t]
-\begin{tabular}{cc}
-  \Tree [.{$+$}
-    1
-    [.{$\cdot$}
-      2
-      3 ] ] &
-  \Tree [.{$\cdot$}
-    [.{$+$}
-      1
-      2 ]
-    3 ] \\
-\end{tabular}
-\caption{The two allowable parse trees of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
-\label{fig:arith-example}
-\end{figure}
-
-The ambiguity of a word $w \in \T^{*}$ is defined in terms of $\parse$:
-
-\begin{definition}
-  Given an abstract parser $P = (\T, T, \parse)$, a word $w \in L(P)$
-  is ambiguous, written $\amb_P(w)$, iff
-
-  $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$
-\end{definition}
-
-\noindent
-Note that the above definition implies that a word $w \in L(P)$, where
-$P = (\T, T, \parse)$, is not ambiguous, or \emph{unambiguous}, if
-$\exists t \in T.\ \parse(w) = \{t\}$.
-
-We can connect the above definition of a parse language to the
-classical definition of a (formal) language as follows:
-
-\begin{itemize}
-\item Given a parse language $L(P)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
-\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
-\end{itemize}
-
-\noindent A \emph{resolvably ambiguous} word is a word where all its parse trees can be written in an unambiguous way, formally:
-
-\begin{definition}\label{def:resolvable-word}
-  Given an abstract parser $P = (\T, T, \parse)$, a word $w \in L(P)$ is \emph{resolvably ambiguous}, written $\rho_P(w)$, iff
-
-  $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
-\end{definition}
-
-The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way.
-However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously parsed as the first and second tree, respectively.
-
-Additionally, we define an \emph{abstract parser} to be resolvably ambiguous if all words in its corresponding parse language $L(P)$ are resolvably ambiguous, formally:
-
-\begin{definition}\label{def:resolvable-language}
-  An abstract parser $P$ is resolvably ambiguous, written $\rho(P)$, iff
-
-  $\forall w \in L(P).\ \rho_P(w)$.
-\end{definition}
-
-\noindent We can now state a few things:
-
-\begin{itemize}
-\item An unambigous word $w$ is trivially resolvably ambiguous, since its only parse tree $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$).
-The set of resolvably ambiguous words is thus a superset of the unambiguous words.
-\item If a given parse tree $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous.
-In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
-\end{itemize}
-
-\noindent The second point suggests that resolvable ambiguity is only an interesting property if an element of $T$ does not uniquely identify an element of $\T^{*}$.
-Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual parse tree.
-Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace.
-In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the parse tree.
-
-We thus propose to loosen the common ``no ambiguity'' restriction on programming language grammars, and instead only require them to be resolvably ambiguous.
-However, merely having an arbitrary function $\parse$ gives us very little to work with, and no way to decide whether the language it defines is resolvably ambiguous or not.
-The remainder of this paper will thus consider $\parse$ functions defined using a particular formalism, introduced in Section~\ref{sec:parse-time-disambiguation}, which gives us some decidable properties.
-
-Before introducing this formalism however, we introduce the two central problems we consider in this paper:
-
-\begin{description}
-\item[Static resolvability.] Given an abstract parser $P$, determine whether $\rho(P)$.
-\item[Dynamic resolvability.] Given an abstract parser $P$ and a word $w \in L(P)$, determine whether $\rho_P(w)$.
-\end{description}
-
-\noindent Our main concern is producing decision procedures for these problems. First, we define soundness and completeness for the static problem.
-
-% todo: improve label (where used?)
-\begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure}
-  A decision procedure $f$ solving the static resolvability problem is sound iff
-
-  $f(P) \implies \rho(P) \text{ for all abstract parsers } P$
-\end{definition}
-
-In practice, we may additionally be interested in a decision procedure
-$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
-  parsers } P$. It turns out, we can give $g$ for the parse languages
-studied in the following sections.
-
-\begin{definition}[Completeness of Static Resolvability]
-  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff
-
-  $\rho(P) \implies f(P) \text{ for all abstract parsers } P$
-\end{definition}
-
-\noindent Similarly, we define soundness and completeness for the dynamic problem.
-
-% todo: improve label (where used?)
-\begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure}
-  A decision procedure $f$ solving the dynamic resolvability problem is sound iff
-
-  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
-\end{definition}
-
-Analogous to the static case, for the parse languages studied in the
-following sections we can also provide a decision procedure $g$ such
-that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
-\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.
-
-\begin{definition}[Completeness of Dynamic Resolvability]
-  A decision procedure $f$ solving the dynamic resolvability problem is complete iff
-
-  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
-\end{definition}
-
-\section{Parse-time Disambiguation} \label{sec:parse-time-disambiguation}
-
-This section describes our chosen language definition formalism, and motivates its design.
-
-The primary purpose of this formalism is, as described in the previous section, to produce a $\parse$ function, i.e., to describe a word language and assign one or more interpretations to each word.
-The interpretations will be unranked trees, intended to be somewhat reminiscent of the abstract syntax trees used in most compilers.
-Section~\ref{sec:resolvable-definition} suggests that $\parse$ discard some information, to enable the resolution of some ambiguities; we here choose to discard grouping parentheses.
-
-With that in mind, we define a language definition $D$ as a set of labelled productions, as described in Fig.~\ref{fig:input-language-definition}.
-Note that we require the labels to uniquely identify the production, i.e., there can be no two distinct productions in $D$ that share the same label.
-Also note that the right-hand side of a production is a regular expression, rather than the theoretically simpler sequence used in a context-free grammar.
-For proof-technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense, when marks are removed.
-This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}.
-In practice, this is likely desireable regardless, since a user is likely to attach different semantic meaning to each occurrence of a non-terminal.
-
-Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal.
-To lessen clutter, we will write $E_\emptyset$ as $E$.
-As an example, consider the language definition in Fig.~\ref{fig:running-example-definition}, which will be used as a running example.
-In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication.
-By ``direct child'' we mean ``without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.
-
-\begin{figure}
-\centering
-\begin{minipage}{.47\textwidth}
-  \centering
-  \begin{tabular}{@{}ll@{}}
-      Terminals & $t \in \T$ \\
-      Non-terminals & $N \in \NT$ \\
-      Labels & $l \in \Labels$ \\
-      \multicolumn{2}{@{}l@{}}{$\T$, $\NT$, and $\Labels$ disjoint} \\
-      \addlinespace
-      Marks & $m \subseteq \Labels$ \\
-      Regular expressions & $r ::= t \mid N_m \mid r \cdot r $ \\
-      & \hphantom{$r ::=$}$\mid r + r \mid \epsilon \mid r^{*}$ \\
-      Labelled productions & $N -> l : r$ \\
-  \end{tabular}
-  \captionof{figure}{The abstract syntax of a language definition.}
-  \label{fig:input-language-definition}
-\end{minipage}\quad
-\begin{minipage}{.45\textwidth}
-  \centering
-  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
-    $E$ & $l$ & \verb|'['| ($E$ (\verb|';'| $E$)$^{*}$ $+$ $\epsilon$) \verb|']'| \\
-    $E$ & $a$ & $E$ \verb|'+'| $E$ \\
-    $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
-    $E$ & $n$ & $N$ \\
-  \end{tabular}
-  \captionof{figure}{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity.
-Assumes that $N$ matches a numeric terminal.}
-  \label{fig:running-example-definition}
-\end{minipage}
-\end{figure}
-
-From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$.
-Technically, only $G'_D$ and $T_D$ are required, $G'_D$ is used as the defined word language and $T_D$ as the interpretations, but the remaining two grammars help the presentation.
-
-\begin{itemize}
-\item $G_D$ represents a word language describing all semantically distinct programs.
-\item $T_D$ represents a tree language describing the parse trees of words in $L(G_D)$.
-\item $G'_D$ is essentially a modified version of $G_D$, e.g., adding grouping parentheses and other forms of disambiguation (i.e., the result of marks).
-\item $T'_D$ represents a tree language describing the parse trees of words in $L(G'_D)$.
-\end{itemize}
-
-\noindent Fig.~\ref{fig:running-example-generated} shows the four grammars generated from our running example in Fig.~\ref{fig:running-example-definition}.
-The context-free grammars are produced by a rather standard translation from regular expressions to CFGs, while the primed grammars get a new non-terminal per distinctly marked non-terminal in $D$, where each new non-terminal only has the productions whose label is not in the mark.
-For example, the non-terminal $E_{\{a\}}$ in Fig.~\ref{fig:running-example-generated:t-prime} has no production corresponding to the $a$ production in Fig.~\ref{fig:running-example-definition}.
-
-\begin{figure}
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
-      \toprule
-      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
-      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
-      $E$ & $n$ & $N$ $)$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$T_D$, the parse trees of $G_D$.}
-    \label{fig:running-example-generated:t}
-  \end{subfigure}%
-%
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
-      \toprule
-      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
-      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
-      $E$ & $n$ & $N$ $)$ \\
-      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
-      \midrule
-      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
-      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
-      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$T'_D$, the parse trees of $G'_D$.}
-    \label{fig:running-example-generated:t-prime}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
-      \toprule
-      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
-      $E$ & $E$ \verb|'+'| $E$ \\
-      $E$ & $E$ \verb|'*'| $E$ \\
-      $E$ & $N$ \\
-      \midrule
-      $E_{l1}$ & $\epsilon$ \\
-      $E_{l1}$ & $E$ $E_{l2}$ \\
-      \midrule
-      $E_{l2}$ & $\epsilon$ \\
-      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$G_D$, the generated abstract syntax.}
-    \label{fig:running-example-generated:w}
-  \end{subfigure}%
-%
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
-    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
-      \toprule
-      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
-      $E$ & $E$ \verb|'+'| $E$ \\
-      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
-      $E$ & $N$ \\
-      $E$ & \verb|'('| $E$ \verb|')'| \\
-      \midrule
-      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
-      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
-      $E_{\{a\}}$ & $N$ \\
-      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
-      \midrule
-      $E_{l1}$ & $\epsilon$ \\
-      $E_{l1}$ & $E$ $E_{l2}$ \\
-      \midrule
-      $E_{l2}$ & $\epsilon$ \\
-      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
-      \bottomrule
-    \end{tabular}
-    \caption{$G'_D$, the generated concrete syntax.}
-    \label{fig:running-example-generated:w-prime}
-  \end{subfigure}
-  \caption{The generated grammars.}
-  \label{fig:running-example-generated}
-\end{figure}
-
-Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
-Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$.
-Note that the word in $L(G_D)$ is ambiguous, and that there are other words in $L(G'_D)$ that correspond to the same element in $L(T_D)$, e.g., ``$((1 + 2)) * 3$'' and ``$(1 + 2) * (3)$''.
-As a memory aid, the primed versions ($G'_D$ and $T'_D$) contain disambiguation (grouping parentheses, precedence, associativity, etc.) while the unprimed versions ($G_D$ and $T_D$) are the (likely ambiguous) straightforward translations (i.e., ignoring marks) from $D$.
-
-{
-\newcommand{\terminal}[1]{\ \underline{#1}\ }
-
-\begin{figure}
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{center}
-    \Tree [.{$m$}
-        [.{$a$}
-          [.{$n$} \terminal{1} ]
-          \terminal{+}
-          [.{$n$} \terminal{2} ] ]
-        \terminal{*}
-        [.{$n$} \terminal{3} ] ]
-    \end{center}
-    \[m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))\]
-    \caption{Example tree in $L(T_D)$ and visualization.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{center}
-    \Tree [.{$m$}
-        [.{$g$}
-          \terminal{(}
-          [.{$a$}
-            [.{$n$} \terminal{1} ]
-            \terminal{+}
-            [.{$n$} \terminal{2} ] ]
-          \terminal{)} ]
-        \terminal{*}
-        [.{$n$} \terminal{3} ] ]
-    \end{center}
-    \[m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))\]
-    \caption{Example tree in $L(T'_D)$ and visualization.}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{.40\linewidth}
-    \[1 + 2 * 3\]
-    \caption{Example word in $L(G_D)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.40\linewidth}
-    \[(1 + 2) * 3\]
-    \caption{Example word in $L(G'_D)$.}
-  \end{subfigure}
-  \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
-  \label{fig:example-words-in-square}
-\end{figure}
-}
-
-At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses; they are allowed exactly where they would surround a complete production.
-For example, $G_D$ in Fig.~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
-However, starting with that language definition would allow the expression ``$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ``$;2$'' is not a valid expression.
-
-Finally, we require a function $\semantic : L(T'_D) -> L(T_D)$ that removes grouping parentheses from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
-The relation between the four grammars in terms of $\yield$ and $\semantic$ can be seen in Fig.~\ref{fig:grammar-square}.
-With this we can define $\parse : L(G'_D) -> 2^{L(T_D)}$, along with its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
-
-\begin{figure}
-  \begin{tikzcd}
-    L(T_D) \arrow[d, "\yield"] & L(T'_D) \arrow[l, "\semantic"'] \arrow[d, "\yield"] \\
-    L(G_D) & L(G'_D)
-  \end{tikzcd}
-  \caption{The generated grammars, and their relation to each other.}
-  \label{fig:grammar-square}
-\end{figure}
-
-$$
-\begin{array}{rcl}
-\parse(w) & = & \{ \semantic(t) \mid t \in L(T'_D) \land \yield(t) = w \} \\
-\words(t) & = & \{ w \mid t \in \parse(w) \} \\
-\end{array}
-$$
-
-\noindent The latter is mostly useful in later sections, but $\parse$ allow us to consider some concrete examples of resolvable and unresolvable ambiguities.
-For example, in our running example (Fig.~\ref{fig:running-example-definition}), the word \texttt|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
-
-\begin{center}
-  \begin{tabular}{l}
-    $t_1 = a($ \hphantom{$a($} $n($ \verb|'1'| $)$ \verb|'+'| $a($ $n($ \verb|'2'| $)$ \hphantom{$)$} \verb|'+'| $n($ \verb|'3'| $)$ $)$ $)$ \\
-    $t_2 = a($ $a($ $n($ \verb|'1'| $)$ \verb|'+'| \hphantom{$a($} $n($ \verb|'2'| $)$ $)$ \verb|'+'| $n($ \verb|'3'| $)$ \hphantom{$)$} $)$ \\
-  \end{tabular}
-\end{center}
-
-\noindent This is a resolvable ambiguity, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
-To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
-
-\begin{center}
-  \begin{tabular}{l}
-    $t_3 = l($ \verb|'['| \hphantom{$s($} $n($ \verb|'1'| $)$ \verb|';'| $n($ \verb|'2'| $)$ \hphantom{$)$} \verb|']'| $)$ \\
-    $t_4 = l($ \verb|'['| $s($ $n($ \verb|'1'| $)$ \verb|';'| $n($ \verb|'2'| $)$ $)$ \verb|']'| $)$ \\
-  \end{tabular}
-\end{center}
-
-\noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not.
-The solution is to forbid list elements from being sequences by modify the language definition in Fig.~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.
-
-We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure} and \ref{def:dynamic-procedure}.
-Section~\ref{sec:static} gives a partial solution to the former problem, while Section~\ref{sec:dynamic} fully solves the latter with one caveat: we only consider languages with balanced parentheses.
-
-\section{Static Resolvability Analysis} \label{sec:static}
-
-For this section, we will use an alternative formulation of resolvable ambiguity for languages, stated in terms of interpretations instead of words:
-
-\begin{theorem}
-  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
-\end{theorem}
-
-\begin{proof}
-  The quantifier $\forall t \in T$ is equivalent with $\forall w \in \T^{*}. \forall t \in \parse(w)$, since $T = \bigcup_{w \in \T^{*}} \parse(w)$ (by definition).
-\end{proof}
-
-\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w \in L(G'_D)$ for which $\parse(w) = \{t\}$, or prove that no such tree exists.
-Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.
-
-Before an actual algorithm, we split the problem into three different versions of increasing difficulty.
-We then lay the groundwork for our correctness proofs, eventually culminating in the algorithm, which we show to be sound and complete for version 1 and sound for version 2.
-
-The three versions are as follows:
-
-\begin{description}
-\item[Version 1] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$ and no non-terminals appear marked, i.e., for all non-terminals $N_m$ appearing on the right-hand side of the labelled productions in $D$, we have $m = \emptyset$.
-\item[Version 2] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$.
-\item[Version 3] There are no constraints on $D$.
-\end{description}
-
-\noindent The first restriction in version 1 states that $D$ cannot contain parentheses.
-This implies that all parentheses present in $G'_D$ are grouping parentheses.
-The second restriction implies that no parentheses are required (as a counterexample, assuming normal precedence, the parentheses in ``$(1 + 2) * 3$'' are required and removing them would produce a different interpretation).
-
-The reason for this split stems from the following insight: double grouping parentheses do not matter, in the sense that they do not change the interpretation of the word, e.g., $\parse($\verb|'((1 + 2)) * 3'|$) = \parse($\verb|'(1 + 2) * 3'|$)$.
-Versions~1 and 2 have only grouping parentheses, meaning that no double parentheses matter.
-What follows is a brief outline of the remainder of this section, which details our static analysis for versions~1 and 2, building from that observation.
-
-\begin{itemize}
-\item We can consider an alternate representation of words where parentheses are not explicitly part of the string of terminals, but are represented by an accompanying bag of ranges denoting which terminals are covered by parentheses.
-(Section~\ref{sec:word-view}).
-\item This alternate representation gives rise to a lattice per tree in $L(T_D)$, where a tree has only ambiguous words iff its lattice is entirely covered by the lattices of other trees (Section~\ref{sec:lattice}).
-\item These lattices can be encoded as words, leading to the construction of a visibly pushdown automaton that we can examine to determine the existence of a tree whose lattice is covered by the lattice of a \emph{single} other tree, which is sufficient for completeness in version 1 (Section~\ref{sec:lattice-vpl}).
-\end{itemize}
-
-\subsection{An Alternative Word View} \label{sec:word-view}
-
-This section introduces an alternative (isomorphic) definition of a word, heavily used in sections~\ref{sec:lattice} and \ref{sec:lattice-vpl}.
-The method by which we generate $G_D$ and $G'_D$ limits the possible differences between them significantly.
-In particular, no new terminals are introduced, except '(' and ')', and they are always introduced in a well-balanced fashion.
-
-If we thus delimit ourselves to only consider languages where words have no unbalanced parentheses\footnote{I.e., the vast, vast majority of programming languages currently in use.} we can give the following alternative definition of a word: a word is a two-tuple containing a sequence of non-parenthesis terminals and a bag (or multiset) of ranges covered by parentheses.
-For example, the word ``\verb|(1 + 2) * 3|'' is equivalent to $(\text{``}\verb|1 + 2 * 3|\text{''}, \{\range{0}{3}\})$, while ``\verb|((1 + 2)) * 3|'' is equivalent to $(\text{``}\verb|1 + 2 * 3|\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
-We will refer to the first component of the tuple as the \emph{basic word} and the second as the \emph{range bag}.
-
-However, there is a case where this alternate representation is not fully isomorphic; namely when a pair of parentheses would yield a zero-length range.
-As an example, consider the two words ``\verb|(())|'' and ``\verb|()()|''.
-Both of these have an alternate representation of $(\text{``}\text{''}, \{\range{0}{0}, \range{0}{0}\})$.
-In versions~1 and 2 this corresponds to grouping parentheses surrounding zero-length productions.
-We can avoid this case by finding every production $N -> l : r$ where $\epsilon \in L(r)$, replacing it with $N -> l : r'$ where $L(r') = L(r) \setminus \{\epsilon\}$, and replacing every use of $N$ with $N + \epsilon$.
-The one exception is if the start symbol has a nullable production, but that can be trivially special cased.
-Thus, we will assume no zero-length grouping parentheses in the following.
-
-We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., \\$\basic(\text{``}\verb|((1 + 2)) * 3|\text{''}) = \text{``}\verb|1 + 2 * 3|\text{''}$ and $\rangebag(\text{``}\verb|((1 + 2)) * 3|\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
-
-The next four lemmas, all concerning the words in $\words(t)$ for some given $t \in L(G_t)$, form the basis of the lattices mentioned at the end of the previous section.
-
-\begin{lemma}
-  $\forall w_1, w_2 \in \words(t).\ \basic(w_1) = \basic(w_2)$.
-  \label{lemma:same-basic}
-\end{lemma}
-\noindent More informally, all $w \in \words(t)$ have the same basic word.
-Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
-The only terminals changed in this process are parentheses, i.e., different $w$ can only differ in their range bags.
-This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w$ such that $\parse(w) \supseteq \{t_1, t_2\}$) must also share a basic word.
-
-\begin{lemma}
-  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ R \subseteq \support(\rangebag(w))$
-  \label{lemma:required-parentheses}
-\end{lemma}
-\noindent Recall that $\support(B)$ is the support of the bag $B$, i.e., the underlying set of elements in $B$.
-More informally, some parentheses are required.
-For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees.
-Each required range is a direct consequence of a mark on a non-terminal in $D$.
-We will write $\botw(t)$ for the word implied by the first quantified expression.
-It is unique, since its basic word is fixed by $t$, and its rangebag is a set.
-
-\begin{lemma}
-  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ \support(\rangebag(w)) \subseteq R$.
-  \label{lemma:possible-parentheses}
-\end{lemma}
-\noindent More informally, there is a finite set of possible parentheses.
-As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes.
-We will write $\topw(t)$ for the word implied by the first quantified expression, which is also unique.
-
-\begin{lemma}
-  $\forall w_1, w_2 \in \words(t).\ \support(\rangebag(w_1)) = \support(\rangebag(w_2)) => \parse(w_1) = \parse(w_2)$.
-  \label{lemma:rangeset-equality}
-\end{lemma}
-\noindent More informally, duplicated parentheses do not matter.
-We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses.
-Each pair of parentheses thus corresponds to a single grouping node $g$ in a tree in $L(T'_D)$.
-Duplicated grouping parentheses correspond to nested grouping nodes, all of which are removed by $\semantic$, thus producing identical sets of trees in $L(T_D)$.
-For example, \verb|((1 + 2)) * 3| has the same interpretations as \verb|(1 + 2) * 3|.
-
-\subsection{A Lattice of Word Partitions} \label{sec:lattice}
-
-\begin{figure}[t]
-  \begin{tikzpicture}[every node/.style={rectangle,draw}]
-    \node (a) [draw=white] at (0,5) {$((1) + (2)) + (3)$};
-
-    \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
-    \node (b2) [draw=white] at (-1.2,4.2) {$((1) + (2)) + 3$};
-    \node (b3) [draw=white] at ( 1.2,4.2) {$((1) + 2) + (3)$};
-    \node (b4) [draw=white] at ( 3.6,4.2) {$(1 + (2)) + (3)$};
-
-    \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
-    \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
-    \node (c3) [draw=white] at (-1.1,3) {$((1) + 2) + 3$};
-    \node (c4) at (1.1,3) {$1 + (2) + (3)$};
-    \node (c5) [draw=white] at (3.3,3) {$(1 + (2)) + 3$};
-    \node (c6) [draw=white] at (5.5,3) {$(1 + 2) + (3)$};
-
-    \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
-    \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
-    \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
-    \node (d4) [draw=white] at ( 3.6,1.8) {$(1 + 2) + 3$};
-
-    \node (e) at (0, 1) {$1 + 2 + 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1.90) -- (c1.270);
-  \draw (d1.90) -- (c2.270);
-  \draw (d1.90) -- (c3.270);
-  \draw (d2.90) -- (c1.270);
-  \draw (d2.90) -- (c4.270);
-  \draw (d2.90) -- (c5.270);
-  \draw (d3.90) -- (c2.270);
-  \draw (d3.90) -- (c4.270);
-  \draw (d3.90) -- (c6.270);
-  \draw (d4.90) -- (c3.270);
-  \draw (d4.90) -- (c5.270);
-  \draw (d4.90) -- (c6.270);
-
-  \draw (c1.90) -- (b1.270);
-  \draw (c1.90) -- (b2.270);
-  \draw (c2.90) -- (b1.270);
-  \draw (c2.90) -- (b3.270);
-  \draw (c3.90) -- (b2.270);
-  \draw (c3.90) -- (b3.270);
-  \draw (c4.90) -- (b1.270);
-  \draw (c4.90) -- (b4.270);
-  \draw (c5.90) -- (b2.270);
-  \draw (c5.90) -- (b4.270);
-  \draw (c6.90) -- (b3.270);
-  \draw (c6.90) -- (b4.270);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-    %%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-  \end{tikzpicture}
-  \caption{The lattice of words for the tree $a(a(n(1)+n(2))+n(3))$. The boxed words are shared with the lattice in Fig.~\ref{fig:lattice2}.}
-  \label{fig:lattice1}
-\end{figure}
-
-Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w$ by their \emph{rangeset}, where $\rangeset(w) = \support(\rangebag(w))$.
-These partitions can be partially ordered by subset on the rangeset, resulting in a lattice of word partitions per tree.
-This lattice is bounded, with top and bottom elements $\topw(t)$ and $\botw(t)$.
-Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
-For example, Fig.~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}).
-Each partition is represented by the word whose rangebag is a set.
-To reduce clutter, we do not draw the partitions that have grouping parentheses around the entire word.
-This outermost possible pair is uninteresting since it is always allowed, and would double the size of the figure if it was included.
-
-\begin{figure}[t]
-
-\begin{tikzpicture}[every node/.style={rectangle,draw}]
-  \node (a) [draw=white] at (0,5) {$(1) + ((2) + (3))$};
-
-  \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
-  \node (b2) [draw=white] at (-1.2,4.2) {$(1) + ((2) + 3)$};
-  \node (b3) [draw=white] at ( 1.2,4.2) {$(1) + (2 + (3))$};
-  \node (b4) [draw=white] at ( 3.6,4.2) {$1 + ((2) + (3))$};
-
-  \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
-  \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
-  \node (c3) [draw=white] at (-1.1,3) {$(1) + (2 + 3)$};
-  \node (c4) at (1.1,3) {$1 + (2) + (3)$};
-  \node (c5) [draw=white] at (3.3,3) {$1 + ((2) + 3)$};
-  \node (c6) [draw=white] at (5.5,3) {$1 + (2 + (3))$};
-
-  \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
-  \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
-  \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
-  \node (d4) [draw=white] at ( 3.6,1.8) {$1 + (2 + 3)$};
-
-  \node (e) at (0, 1) {$1 + 2 + 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1.90) -- (c1.270);
-  \draw (d1.90) -- (c2.270);
-  \draw (d1.90) -- (c3.270);
-  \draw (d2.90) -- (c1.270);
-  \draw (d2.90) -- (c4.270);
-  \draw (d2.90) -- (c5.270);
-  \draw (d3.90) -- (c2.270);
-  \draw (d3.90) -- (c4.270);
-  \draw (d3.90) -- (c6.270);
-  \draw (d4.90) -- (c3.270);
-  \draw (d4.90) -- (c5.270);
-  \draw (d4.90) -- (c6.270);
-
-  \draw (c1.90) -- (b1.270);
-  \draw (c1.90) -- (b2.270);
-  \draw (c2.90) -- (b1.270);
-  \draw (c2.90) -- (b3.270);
-  \draw (c3.90) -- (b2.270);
-  \draw (c3.90) -- (b3.270);
-  \draw (c4.90) -- (b1.270);
-  \draw (c4.90) -- (b4.270);
-  \draw (c5.90) -- (b2.270);
-  \draw (c5.90) -- (b4.270);
-  \draw (c6.90) -- (b3.270);
-  \draw (c6.90) -- (b4.270);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-\end{tikzpicture}
-\caption{The lattice of words for the tree $a(n(1)+a(n(2)+n(3)))$. The boxed words are shared with the lattice in Fig.~\ref{fig:lattice1}.}
-\label{fig:lattice2}
-\end{figure}
-
-To show the connection between resolvable ambiguity and these lattices, consider the word ``\verb|1 + 2 + 3|''.
-It has two interpretations in our running example, $a(a(n(1)+n(2))+n(3))$, which we would normally write as \verb|(1 + 2) + 3|, and $a(n(1)+a(n(2)+n(3)))$, which we would normally write \verb|1 + (2 + 3)|.
-The lattices for these two trees are given in Figures~\ref{fig:lattice1} and \ref{fig:lattice2} respectively.
-The partitions that appear in both lattices are represented as boxed words, the others are unboxed.
-These shared partitions represent words that are ambiguous between these particular trees.
-Finding an unambiguous word is thus the same as finding a partition that is not shared with any other tree.
-In this particular case, there is no ambiguity with any other tree at all, and so the unboxed words are all valid resolutions of the ambiguity.
-
-At this point it is clear that a given tree has an unambiguous word iff its lattice has at least one partition that is not shared with any other tree.
-
-Next, we note that each lattice is uniquely determined by its top and bottom elements; it contains all elements between them:
-
-\begin{lemma}
-  Given $t \in L(T_D)$, we have that:\\
-  $\forall w.\ \rangebag(\botw(t)) \subseteq \rangeset(w) \subseteq \rangebag(\topw(t)) => w \in \words(t)$. \label{lemma:top-bottom-determine}
-\end{lemma}
-
-\noindent Recall that $\rangebag = \rangeset$ for $\botw(t)$ and $\topw(t)$.
-Finally, we show the difference between versions~1 and 2 in the lattice setting: in version~1, the rangeset of the bottom word is the empty set:
-
-\begin{lemma}
-  In version~1, $\forall t \in L(T_D).\ \rangeset(\botw(t)) = \emptyset$. \label{lemma:version1-bot}
-\end{lemma}
-
-\noindent Since version~1 has no markings there are no required parentheses, thus the bottom word has no parentheses.
-
-Our approach centers around finding a pair of lattices, such that one is entirely contained in the other.
-Since the former shares \emph{all} partitions with the latter, it has only ambiguous words, thus any of those words will be unresolvably ambiguous.
-
-\begin{theorem}
-  In versions~1 and 2, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, the existence of a tree $t' \in L(T_D)$ such that $\basic(w_\bot) = \basic(\botw(t')) \land \rangeset(w_\bot) \subseteq \rangeset(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$ implies $\neg \exists w.\ \parse(w) = \{t\}$.
-\label{lemma:lattice-containment-implies-unresolvable}
-\end{theorem}
-
-\noindent For version~1, \emph{every} unresolvable ambiguity has this form:
-
-\begin{theorem}
-  In version 1, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\botw(t')) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(\topw(t')) <=> \neg \exists w.\ \parse(w) = \{t\}$.
-\end{theorem}
-
-\begin{proof}
-  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(\botw(t'))$ implies that $w_\bot = \botw(t')$.
-We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$.
-
-  If all words in $\words(t)$ are ambiguous, then $\exists t' \in L(T_D).\ t' \neq t \land t' \in \parse(w_\top)$.
-Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(\botw(t'))$.
-The definition of $\topw(t')$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(\topw(t'))$, but since the rangebag of $\topw(t')$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(\topw(t'))$.
-\end{proof}
-
-\subsection{A Lattice as a Word, and an Algorithm} \label{sec:lattice-vpl}
-
-This section introduces a linear encoding of the lattices of the previous chapter as words.
-Lemmas~\ref{lemma:same-basic} and \ref{lemma:top-bottom-determine} imply that a lattice encoding requires three things: a basic word, a set of \emph{required} parentheses, and a set of \emph{possible} parentheses.
-We encode required parentheses with ``$\reqp{}$'' and possible parentheses with ``$\posp{}$''.
-For example, the lattice in Fig.~\ref{fig:lattice2} is represented by ``$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ``$(1 + 2) * 3$'' has a lattice represented by ``$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
-With this encoding we can apply the formidable body of knowledge available for word languages, albeit with some extra care; the linear encoding of a lattice is not unique.
-
-For example, ``$\posp{\posp{1}}$'' and ``$\posp{1}$'' represent the same lattice, as do ``$\reqp{\posp{2}}$'' and ``$\reqp{2}$''.
-To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses, e.g., ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
-
-Lattice equality is now equvialent with equality of the linear encoding.
-To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
-For example, ``$\reqp{1}2$'' is entirely contained in ``$\reqp{1}\posp{2}$'', which is entirely contained in ``$\posp{1}\posp{2}$''.
-
-The centerpiece of our algorithm is a visibly pushdown automaton constructed in such a way that:
-
-\begin{itemize}
-\item There is a bijection between trees in $L(T_D)$ and successful runs.
-\item The word recognized by a successful run is the linear encoding of the corresponding tree's lattice.
-\end{itemize}
-
-\noindent Two distinct successful runs through this automaton that recognize the same word thus imply two distinct trees that have exactly the same lattice.
-To detect lattices contained in each other, we create a modified copy, where the copy may add arbitrary (balanced, well-nested) possible parentheses, and replace any required pair of parentheses with a possible pair.
-The copy maintains much of the structure of the original, and in particular, keeps the connection to trees in $L(T_D)$ (though it is no longer a bijection: there are multiple successful runs per tree since there are multiple larger lattices).
-Two distinct successful runs, one in each automaton, that recognize the same word then imply two distinct trees where the lattice of one is entirely contained in the other.
-
-We will now walk through the construction of this automaton, using the following language definition:
-
-\begin{center}
-\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-  \toprule
-  $E$ & $s$ & $E_{\{s\}}$ \verb|';'| $E$ \\
-  $E$ & $v$ & \verb|'e'| \\
-  \bottomrule
-\end{tabular}
-\end{center}
-
-\noindent We first present a simplified construction that assumes no production can match a single non-terminal, i.e., for all right-hand side regular expressions $r$, $\neg \exists N_m.\ N_m \in L(r)$, to make the base idea easier to follow.
-Such cases introduce duplicated parentheses if handled in a naive way, and the extra book-keeping required to correctly handle them complicate this presentation, and will so be introduced after the more naive method.
-
-Conceptually, each non-terminal represents a choice of which production should replace it, and whether that production requires parentheses around it.
-We now make this explicit, to ensure that standard operations on finite automata produce the correct result.
-We replace each $N_m$ with a sum of the labels of productions in $N$, where each label has a subscript $\top$ or $\bot$ if parentheses are required or possible, respectively.
-For example, $E_{\{s\}}$ is replaced with $s_\top + v_\bot$.
-
-\begin{center}
-\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-  \toprule
-  $E$ & $s$ & $(s_\top + v_\bot)$ \verb|';'| $(s_\bot + v_\bot)$ \\
-  $E$ & $v$ & \verb|'e'| \\
-  \bottomrule
-\end{tabular}
-\end{center}
-
-\noindent Next we construct a DFA per production. This can be done in the standard way by constructing an NFA, then determinizing it, and optionally minimizing it.
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-  \node[state,accepting] (4) [right of=3] {$4$};
-
-  \node[state,initial] (5) [below of=2] {$5$};
-  \node[state,accepting] (6) [right of=5] {$6$};
-
-  \path (1) edge[bend left] node {$s_\top$} (2)
-  (1) edge[bend right] node {$v_\bot$} (2)
-  (2) edge node {$';'$} (3)
-  (3) edge[bend left] node {$s_\bot$} (4)
-  (3) edge[bend right] node {$v_\bot$} (4);
-  \path (5) edge node {$'e'$} (6);
-
-\end{tikzpicture}
-\end{center}
-
-\noindent We then combine them into a single visibly pushdown automaton.
-Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we go back.
-We use $Q \times Q$ as our stack alphabet, to record the transition we are replacing in the parent.
-We thus replace every edge $p \xrightarrow{l_\top} q$, where $l \in \Labels$, with:
-
-\begin{itemize}
-  \item an edge $p \xrightarrow{'\reqpl', +(p, q)} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
-  \item an edge $q' \xrightarrow{'\reqpr', -(p, q)} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
-\end{itemize}
-
-\noindent Similarly, for every edge $p \xrightarrow{l_\bot} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\pospl', +(p, q)} p'$ and $q' \xrightarrow{'\pospr', -(p, q)} q$.
-Note the difference: for $l_\top$ we add a required parenthesis ``$\reqpl$'', while $l_\bot$ adds a possible parenthesis ``$\pospl$''.
-
-Intuitively, we parse a child node with parentheses around it, then return. This is where our simplification is used, without it we might introduce double parentheses here.
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-  \node[state,accepting] (4) [right of=3] {$4$};
-
-  \node[state,initial] (5) [below of=2] {$5$};
-  \node[state,accepting] (6) [right of=5] {$6$};
-
-  \path (2) edge[below] node {$';'$} (3);
-  \path (5) edge[below] node {$'e'$} (6);
-
-  \path (1) edge[loop above] node{$'\reqpl', +(1, 2)$} (1)
-  (4) edge[bend right, above] node{$'\reqpr', -(1, 2)$} (2);
-  \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
-  (6) edge[right] node[xshift=5mm,yshift=-3mm]{$'\pospr', -(1, 2)$} (2);
-  \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
-  (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
-  \path (3) edge[right] node[xshift=5mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
-  (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
-
-\end{tikzpicture}
-\end{center}
-
-\noindent Finally, we add a new initial state and a new final state, and connect them with the initial and final states belonging to the starting non-terminal:
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (7) [left of=5, xshift=-1.5cm] {$7$};
-  \node[state,accepting] (8) [right of=6, xshift=1.5cm] {$8$};
-
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-  \node[state] (4) [right of=3] {$4$};
-
-  \node[state] (5) [below of=2] {$5$};
-  \node[state] (6) [right of=5] {$6$};
-
-  \path (2) edge[below] node {$';'$} (3);
-  \path (5) edge[below] node {$'e'$} (6);
-
-  \path (1) edge[loop above] node{$'\reqpl', +(1, 2)$} (1)
-  (4) edge[bend right, above] node{$'\reqpr', -(1, 2)$} (2);
-  \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
-  (6) edge[right] node[xshift=5mm,yshift=-3mm]{$'\pospr', -(1, 2)$} (2);
-  \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
-  (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
-  \path (3) edge[right] node[xshift=5mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
-  (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
-
-  \path (7) edge[bend left, left] node{$'\pospl', +(7, 8)$} (1)
-  (7) edge[below] node{$'\pospl', +(7, 8)$} (5);
-  \path (4) edge[bend left, right] node{$'\pospr', -(7, 8)$} (8)
-  (6) edge[below] node{$'\pospr', -(7, 8)$} (8);
-
-\end{tikzpicture}
-\end{center}
-
-\noindent For example, the linearly encoded lattice $\posp{\posp{e};\posp{e}}$ is recognized by the run
-
-\begin{minipage}{.7\textwidth}
-$$
-\begin{array}{l@{\enspace}l@{\;}r@{\enspace}l@{\;}r@{\enspace}l@{\;}r}
-  (7, \epsilon) &\xrightarrow{\pospl} (1, &(7,8)) &\xrightarrow{\pospl} (5, &(1,2)\cdot(7,8)) &\xrightarrow{e} (6, &(1,2)\cdot(7,8)) \\
-  &\xrightarrow{\pospr} (2, &(7,8)) &\xrightarrow{;} (3, &(7,8)) &\xrightarrow{\pospl} (5, &(3,4)\cdot(7,8)) \\
-  &\xrightarrow{e} (6, &(3,4)\cdot(7,8)) &\xrightarrow{\pospr} (4, &(7, 8)) &\xrightarrow{\pospr} (8, &\epsilon) \\
-\end{array}
-$$
-\end{minipage}
-\begin{minipage}{.3\textwidth}
-\begin{center}
-{
-\newcommand{\terminal}[1]{\ \underline{#1}\ }
-\Tree [.{$s$}
-    [.{$v$} \terminal{e} ]
-    \terminal{;}
-    [.{$v$} \terminal{e} ] ]
-}
-\end{center}
-\end{minipage}
-
-\noindent and corresponds to the tree on the right.
-
-We now illustrate the issue that our previous simplification avoids, and then modify the construction such that the simplification is no longer required. Consider the following language definition:
-
-\begin{center}
-\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-  \toprule
-  $A$ & $a$ & $B$ \\
-  $B$ & $b$ & $C$ $(\verb|';'| + \epsilon)$ \\
-  $C$ & $c$ & \verb|'e'| \\
-  \bottomrule
-\end{tabular}
-\end{center}
-
-\noindent Following the simplified construction we produce the following automaton:
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (8) [left of=1, xshift=-1cm] {$8$};
-  \node[state,accepting] (9) [right of=2, xshift=1cm] {$9$};
-
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-
-  \node[state] (3) [below of=1,yshift=3mm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=3,yshift=3mm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \path (4) edge[below] node {$';'$} (5);
-  \path (6) edge[below] node {$'e'$} (7);
-
-  \path (1) edge[left] node {$'\pospl', +(1,2)$} (3)
-  (4) edge node {$'\pospr', -(1,2)$} (2)
-  (5) edge[right] node {$'\pospr', -(1,2)$} (2);
-  \path (3) edge[left] node {$'\pospl', +(3, 4)$} (6)
-  (7) edge[left] node {$'\pospr', -(3, 4)$} (4);
-
-  \path (8) edge node {$'\pospl', +(8, 9)$} (1)
-  (2) edge node {$'\pospr', -(8, 9)$} (9);
-
-\end{tikzpicture}
-\end{center}
-
-\noindent Unfortunately, this automaton recognizes the word $(((e)))$, which is not the unique linear encoding we require; it should be $(e)$.
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (8) [left of=1, xshift=-1cm] {$8$};
-  \node[state,accepting] (9) [right of=2, xshift=1cm] {$9$};
-
-  \node[state] (1) {$1$};
-
-  \node[state] (3) [below of=1,yshift=3mm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=3,yshift=3mm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \path (4) edge[below] node {$';'$} (5);
-  \path (6) edge[below] node {$'e'$} (7);
-
-  \path (3) edge[left] node {$'\pospl', +(3, 4, \epsilon)$} (6)
-  (7) edge[left] node {$'\pospr', -(3, 4, \epsilon)$} (4);
-
-  \path (8) edge node {$'\pospl', +(8, 9, \epsilon)$} (1)
-  (8) edge[right] node[xshift=1mm,yshift=1mm] {$'\pospl', +(8, 9, a)$} (3)
-  (5) edge[left] node {$'\pospr', -(8,9,a)$} (9)
-  (8) edge[left,out=270,in=180] node {$'\pospl', +(8, 9, ab)$} (6)
-  (7) edge[right,out=0,in=270] node {$'\pospr', -(8, 9, ab)$} (9);
-
-\end{tikzpicture}
-\end{center}
-
-UNFINISHED
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-% FORMAL STUFF BARRIER %
-%%%%%%%%%%%%%%%%%%%%%%%%
-
-Given a language definition $D$, with terminals $\T$, non-terminals $\NT$, labels $\Labels$, and starting non-terminal $S$.
-We use $\labelnt(l)$ and $\labelregex(l)$ as the non-terminal and regex, respectively, of the production labelled $l$.
-We define the set $T : \Labels \times \Labels \times \{\top, \bot\} \times \Labels^{*}$ of transitions between productions inductively through:
-
-\begin{itemize}
-\item $(l, l, \bot, \epsilon) \in T$ for all $l \in \Labels$.
-\item $(l, l', l' \in m, l) \in T$ for all $N_m \in L(\labelregex(l))$ where $\labelnt(l') = N$.
-\item $(l_1, l_3, r_1 \lor r_2, w_1 \cdot w_2) \in T$ for all $(l_1, l_2, r_1, w_1), (l_2, l_3, r_2, w_2) \in T$.
-\end{itemize}
-
-\noindent The third component of each element denotes whether that transition introduces a required pair of parentheses ($\top$) or a possible pair ($\bot$).
-$T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
-Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
-
-We define the translation from production to DFA through a function $\dfa$ from a label to a DFA with the alphabet $\T \cup (\Labels \times \{\top, \bot\})$:
-
-\begin{description}
-\item[$\dfa(l)$:] Compute the regular language $L' := L(\labelregex(l)) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
-Then replace every marked non-terminal $N_m$ with the regular expression $(l_1, l_1 \in m) + (l_2, l_2 \in m) + \ldots + (l_n, l_n \in m)$, where $l_i$ are all the labels of productions $N -> l_i : r_i$.
-For example, $E_{\{a\}}$ in the running example would be replaced by $(l, \bot) + (a, \top) + (m, \bot) + (n, \bot)$.
-Then construct a DFA for this language.
-\end{description}
-
-\noindent We now define the visibly pushdown automaton $A_{\posp{}}$.
-Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
-
-\begin{itemize}
-\item $s$ and $f$ are two new distinct states.
-\item $Q = \{s, f\} \cup \bigsqcup_l Q_l$.
-\item $\T' = \T \cup \{\reqpl, \pospl, \pospr\}$.
-\item $\Gamma = Q \times Q \times \Labels^{*}$.
-\item $\delta$ is defined by (in hindsight this should probably be written as constructing a set $\delta$, i.e., instead of a function $\delta: Q \times \T' \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ we have a set $\delta \subseteq Q \times \T' \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, would probably be easier to read):
-  $$
-  \begin{array}{r@{\;=\;}l@{\;\mid\;}l}
-    \delta(p, a, \lambda)
-    & \multicolumn{2}{@{}l@{}}{\{(\delta_l(p, a), \lambda)\} \qquad \text{ where } p \in Q_l \text{ and } a \in \T} \\
-
-    \delta(p, \text{``}\reqpl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, r_1 \lor r_2 \} \\
-
-    \delta(p, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, \neg (r_1 \lor r_2) \} \\
-
-    \delta(f_{l_3}, \text{``}\reqpr\text{''}, (p, q, w))
-    & \{ (q, \lambda)
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, f_{l_3} \in F_{l_3} \} \\
-
-    \delta(s, \text{``}\reqpl\text{''}, \lambda)
-    & \{ (s_{l_2}, (s, f, w))
-    & \labelnt(l_1) = S, (l_1, l_2, \top, w) \in P \} \\
-
-    \delta(s, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l_2}, (s, f, w))
-    & \labelnt(l_1) = S, (l_1, l_2, \bot, w) \in P \} \\
-
-    \delta(f_{l_2}, \text{``}\reqpr\text{''}, (s, f, w))
-    & \{ (f, \lambda)
-    & nt(l_1) = S, (l_1, l_2, r, w) \in P, f_{l_2} \in F_{l_2}\} \\
-  \end{array}
-  $$
-\end{itemize}
-
-\noindent We also construct a modified VPDA $A'_{\posp{}} = (Q, \T', \Gamma \cup \{\gamma\}, \delta', s, \{f\})$ where:
-
-\begin{itemize}
-\item $\gamma$ is a new distinct stack symbol.
-\item $\delta'(p, a, g) = \delta(p, a, g) \cup \delta''(p, a, g)$, where $\delta''$ is given by:
-  $$
-  \begin{array}{r@{\;=\;}l@{}}
-    \delta''(p, \text{``}\pospl\text{''}, \lambda)
-    & \{ (p, \gamma) \} \\
-
-    \delta''(p, \text{``}\pospr\text{''}, \gamma)
-    & \{ (p, \lambda) \} \\
-
-    \delta''(p, \text{``}\pospl\text{''}, \lambda)
-    & \delta(p, \text{``}\reqpl\text{''}, \lambda) \\
-  \end{array}
-  $$
-\end{itemize}
-
-\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is given in Section~\ref{sec:preliminaries-vpls} (using the partitions $\T_c = \{ \reqpl, \pospl \}$, $\T_i = \T$, and $\T_r = \{\reqpr\}$).
-If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
-We check for the existence of such a run by trimming the product automaton and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
-
-Performing this procedure directly on a language definition $D$ gives a sound decision procedure where the presence of two distinct runs implies unresolvably ambiguous.
-If we instead do it on a modified $D$ with no marks on the right hand side of any production (i.e., all non-terminals have the form $N_\emptyset$) we get a sound decision procedure where the absence of two distinct runs implies resolvably ambiguous.
-
-There are thus cases where we can answer resolvably ambiguous or unresolvably ambiguous with certainty, and cases where we cannot answer either way with certainty.
+\section{Static Proofs}
 
@@
 \begin{proof}
-  (These are two attempts at showing the same thing, next attempt starts at ``Proof sketch''. I would strongly suggest reading ``Formal stuff'' first.)
   $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
@@
 
-\begin{lemma}
-  The word recognized by a successful run in $A_{\posp{}}$ is the unique linear encoding of the lattice for the corresponding $t \in L(T_D)$.
-\end{lemma}
-
-\begin{lemma}
-  For every successful run $R'$ in $A'_{\posp{}}$ there is one (or more) successful runs $R$ in $A_{\posp{}}$ such that the lattice encoded in the word recognized by $R'$ is larger than the lattice for the tree corresponding to $R$.
-\end{lemma}
-
-\begin{theorem}
-  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
-\end{theorem}
-
-\section{Dynamic Resolvability Analysis} \label{sec:dynamic}
-
-The dynamic resolvability problem is as follows: for a given $w' \in L(G'_D)$ determine whether $\forall t \in \parse(w').\ \exists w'_2.\ \parse(w'_2) = \{t\}$.
-Furthermore, for practical reasons, if the word is resolvably ambiguous we wish to produce a (minimal) witness for each tree.
-Additionally, while Section~\ref{sec:static} requires a language definition $D$ to not contain parentheses, this section merely requires parentheses to be balanced.
-
-Our approach centers around around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
-Each such language turns out to be a visibly pushdown language.
-VPLs are closed under difference, which enables us to create a visibly pushdown automaton for the language $\words(t_1) \setminus \words(t_2)$ for any pair of trees $t_1, t_2 \in L(T_D)$.
-This is useful, since $\words(t_1)$ is the set of possible resolutions for $t_1$ in an ambiguity, and $\words(t_1) \cap \words(t_2)$ is the set of ambiguous words that can parse as both $t_1$ and $t_2$, i.e., words that \emph{cannot} resolve an ambiguity.
-
-Intuitively we then need to, for each tree $t \in \parse(w)$, determine if $\words(t) \setminus \bigcup_{t' \in L(T_D) \land t' \neq t} \words(t)$ is the empty set.
-Unfortunately, $L(T_D)$ tends to be infinite.
-However, for any particular $t$, most trees $t' \in L(T_D)$ are irrelevant: their words do not overlap at all; $\words(t) \cap \words(t') = \emptyset$.
-Hence we construct a (finite) subset of $L(T_D)$ such that all other trees are irrelevant.
-As a bonus, this approach constructs a VPDA per tree that recognizes valid resolutions for that tree; a minimal witness is thus a shortest word recognized by a VPDA, and the problem of finding such a word is a known one.
-
-For this section, we assume the presence of a single language definition $D$ with terminals in $\T$, labels in $\Labels$, and non-terminals in $\NT$.
-
-\begin{lemma}
-  Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
-\end{lemma}
-
-\noindent We construct this automaton in a bottom-up fashion.
-A sequence of terminals in $\T$ is easily recognized by a sequence of transitions that do not interact with the stack.
-Each subtree allows surrounding grouping parentheses, which can be represented by transitions that push and pop stack symbols (one unique stack symbol per subtree), to ensure that they are balanced.
-With that, there are two remaining complications: a subtree $l(w)$ recognized from a marked non-terminal $N_m$ where $l \in m$ must have one or more surrounding parentheses instead of zero or more, and we must maintain the visibly pushdown property, i.e., all transitions labelled ``$($'' or ``$)$'' must push or pop stack symbols, respectively.
-The former can be solved by either tracking the subtree's origin when parsing, or reparsing with the right hand side regular expression, while the latter is solved by introducing a new stack symbol $\gamma$ that is pushed and popped by all non-grouping parenthesis terminals.
-We now present the function $f : L(T_D) -> VPDA$ that performs this construction.
-Writing the transition function $\delta : Q \times \T \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ as a subset of $Q \times \T \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, and using $\sqcup$ for disjoint union:
-
-$$
-\begin{array}{r@{\;}l}
-  f(a)
-  & = \begin{cases}
-    (\{p, q\}, \T, \emptyset, \{(p, a, \lambda, q, \lambda)\}, p, \{q\})
-    & \text{if } a \notin \{\text{``}(\text{''}, \text{``})\text{''}\} \\
-
-    (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``}(\text{''}, \lambda, q, \gamma)\}, p, \{q\})
-    & \text{if } a = \text{``}(\text{''} \\
-
-    (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``})\text{''}, \lambda, q, \lambda))\}, p, \{q\})
-    & \text{if } $a = \text{``})\text{''}$ \\
-  \end{cases} \\
-
-  f(w_1 \cdot w_2)
-  & = (Q_1 \sqcup Q_2, \T, \Gamma', \delta_1 \cup \delta_2 \cup \{(f_1, \lambda, \lambda, s_2, \lambda)\}, s_1, \{f_2\}) \\
-  & \text{ where } f(w_i) = (Q_i, \T, \Gamma_i, \delta_i, s_i, \{f_i\}) \\
-  & \hphantom{\text{ where }} \Gamma' = (\Gamma_1 \setminus \{\gamma\}) \sqcup (\Gamma_2 \setminus \{\gamma\}) \cup \{\gamma\} \\
-
-  f(l(w))
-  & = \begin{cases}
-    (Q \sqcup \{p, q\}, \T, \Gamma', \delta \cup \delta_1, p, \{q\}) & \text{if $l$ is marked in the introducing non-terminal.} \\
-    (Q, \T, \Gamma', \delta \cup \delta_2, s, \{f\}) & \text{if $l$ is unmarked in the introducing non-terminal.} \\
-  \end{cases} \\
-  & \text{ where } f(w) = (Q, \T, \Gamma, \delta, s, \{f\}) \\
-  & \hphantom{\text{ where }} \Gamma' = (\Gamma \setminus \{\gamma\}) \sqcup \{g, \gamma\} \\
-  & \hphantom{\text{ where }} \delta_1 = \{(p, \text{``}(\text{''}, \lambda, p, g), (p, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, q, \lambda), (q, \text{``})\text{''}, g, q, \lambda)\} \\
-  & \hphantom{\text{ where }} \delta_2 = \{(s, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, f, \lambda)\} \\
-\end{array}
-$$
-
-\begin{lemma}
-  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
-\end{lemma}
-
-UNFINISHED
-
-\section{Case Studies} \label{sec:evaluation}
-
-% TODO: Make sure that forbids are explained
-
-We have implemented a tool and a small DSL for syntax definition
-that generates a language definition in the style of
-Section~\ref{sec:parse-time-disambiguation}, which the tool then
-uses to construct a parser and do dynamic resolvability analysis.
-The tool is implemented in Haskell and uses an off-the-shelf
-implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
-of the Earley parsing algorithm
-\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
-case studies is to showcase possible use cases of resolvable
-ambiguity in the context of composing separately defined DSLs (a
-subset of Orc, Section~\ref{sec:evaluation-orc}), and of defining
-a pre-existing general purpose programming language (OCaml,
-Section~\ref{sec:evaluation-ocaml}.
-
-The syntax definition DSL builds on the ``syncons'' introduced by
-\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
-production is defined separately from each other (one
-\syncon{syncon} each).
-%
-The main difference between the DSL and the formalism in
-Section~\ref{sec:parse-time-disambiguation} is that marks are
-introduced as separate \syncon{forbid} declarations rather than
-inline in productions. This allows supporting high-level
-convenience constructs, e.g., specifying precedence between
-previously defined operators instead of manually inserting marks,
-but is also important for composability as ambiguities can be
-statically resolved without changing the original definitions.
-Other convenience constructs include defining prefix, infix and
-postfix operators with a given associativity.
-
-The running example used in
-Section~\ref{sec:parse-time-disambiguation}
-(Fig.~\ref{fig:running-example-definition} on
-page~\pageref{fig:running-example-definition}) can be defined as
-follows:
-
-\begin{tabular}{ll}
-\begin{lstlisting}[language=syncon,boxpos=t]
-type Exp
-grouping "(" Exp ")"
-precedence {
-  mul; // higher in precedence list
-  add; // means higher precedence
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[language=syncon,boxpos=t]
-token Integer = "[0-9]+"
-syncon literal: Exp = n:Integer
-syncon list: Exp =
-  "[" (head:Exp (";" tail:Exp)*)? "]"
-infix add: Exp = "+"
-infix mul: Exp = "*"
-\end{lstlisting}
-\end{tabular}\smallskip
-
-\noindent Here, we define a syntax type \syncon{Exp} for
-expressions, and declare that parentheses can be used to group
-expressions. We define a lexical token for integers, and a
-\syncon{syncon} for integer literals using this token. A list is
-defined as a sequence of zero or more expressions separated by
-semi-colons wrapped in square braces. Finally, addition and
-multiplication is defined as infix syncons with the expected
-precedence rules. Note that we could just as well have replaced
-the precedence list with explicit \syncon{forbid} declarations,
-``\syncon{forbid mul.left = add}'' and ``\syncon{forbid mul.right = add}''
-(cf. the mark $\{a\}$ in the production of $m$ in
-Fig.~\ref{fig:running-example-definition}).
-
-% If marks are needed in non-precedence related
-% situations, e.g., adding sequential composition to the running
-% example, they can be introduced with ``forbids'':
-
-% \begin{lstlisting}[language=syncon]
-% infix seqComp: Exp = ";"
-% forbid list.head = seqComp
-% forbid list.tail = seqComp
-% \end{lstlisting}
-
-\subsection{Composing Language Definitions} \label{sec:evaluation-orc}
-
-In this section, we consider the use case of defining a language
-and composing it with another previously defined language, showing
-how we can deal with the resulting ambiguities that arise from the
-composition.
-%
-The language being defined is a subset of
-Orc~\cite{kitchinOrc2009}, a functional programming language which
-includes a number of special-purpose combinators for coordinating
-concurrent workflows. On their own, these combinators act as a DSL
-for concurrency.
-
-In Orc, every expression can ``publish'' zero or more values. Orc
-defines four combinators for orchestrating these published values:
-the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
-pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
-combinators.
-%
-The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
-$e_2$ in parallel, publishing any value published by either of
-them.
-%
-The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ executes
-$[x\mapsto v]e_2$ for \emph{each} value $v$ published by $e_1$,
-building a concurrent pipeline.
-%
-The expression $e_1$ \ocaml{<}$x$\ocaml{<} $e_2$ executes
-$[x\mapsto v]e_1$ for \emph{first} value $v$ published by $e_2$
-(discarding any remaining values published by $e_2$).
-%
-Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if running $e_1$ does
-not publish any values.
-
-The syncon definition of the Orc combinators is very simple (the
-precedence rules follow the original
-definition~\cite{kitchinOrc2009}):
-
-\begin{tabular}{ll}
-\begin{lstlisting}[language=syncon,boxpos=t]
-type Exp
-grouping "(" Exp ")"
-precedence {
-  seq; par; prune; otherwise;
-}
-\end{lstlisting}
-&
-\begin{lstlisting}[language=syncon,boxpos=t]
-token Ident = "[[:lower:]][[:word:]]*"
-infix par:Exp = "|"
-infix seq:Exp = ">" x:Ident ">"
-infix prune:Exp = "<" x:Ident "<"
-infix otherwise:Exp = ";"
-\end{lstlisting}
-\end{tabular}\smallskip
-
-\noindent
-Na\"{i}vely composing this definition with another, separately
-defined language is likely to introduce ambiguities. For example,
-if that language also contains infix operators, the precedence
-between these and the Orc combinators will be undefined. With
-support for resolvable ambiguity however, we can allow this
-ambiguity and let programmers disambiguate expressions using
-parentheses. After composing the above definition with a simple
-language supporting addition, variables, and function calls
-(language definition omitted for brevity), parsing the expression
-%
-``\ocaml{1 + 2 >x> f(x)}''
-%
-results in the following error message from our tool:
-
-\begin{lstlisting}[language={[objective]caml}]
-Ambiguity error with 2 alternatives:
-  ( 1 + 2 ) > x > f ( x )
-  1 + ( 2 > x > f ( x ) )
-\end{lstlisting}
-
-\noindent
-Because there is no precedence specified between \ocaml{+} and
-\ocaml{>x>}, disambiguation is required to specify the order of
-operations.
-%
-In this case, it is likely that the preferred semantics is to have
-all operators in the base language bind tighter than the Orc
-combinators, and these precedence rules can be added after the
-composition without changing the original definitions. Importantly
-though, we are not \emph{required} to resolve this ambiguity at
-time of composition.
-
-Because of how we defined the syntax of the combinators, and since
-parsing with the tool is currently whitespace insensitive, the
-multi-character combinators (sequencing and pruning) are parsed as
-three separate lexical tokens (this is visible in how whitespace
-is inserted in the error message above). This means we can also
-run into \emph{unresolvable} ambiguities, for example if our base
-language includes comparison of numbers with \ocaml{<} and
-\ocaml{>}. With such a base language (assuming no associativity
-for \ocaml{>}), parsing the expression
-%
-``\ocaml{42 >x> f(x)}''
-%
-will result in the following error message:
-
-\begin{lstlisting}[language={[objective]caml}]
-Unresolvable ambiguity error with 2 alternatives.
-\end{lstlisting}
-
-\begin{minipage}{.3\textwidth}
-\begin{lstlisting}[language={[objective]caml}]
-Resolvable alternatives:
-  ( 42 > x ) > f ( x )
-  42 > ( x > f ( x ) )
-\end{lstlisting}
-\end{minipage}
-\hfill
-\begin{minipage}{.5\textwidth}
-\begin{lstlisting}[language={[objective]caml}]
-Unresolvable alternatives:
-  seq
-   - int                  gt.orc:1:1-3
-   - call                 gt.orc:1:8-12
-\end{lstlisting}
-\end{minipage}
-
-\noindent
-By adding parentheses, a programmer can disambiguate the
-expression as two comparisons. However, there is no way for a
-programmer to specify that what they want is a sequential
-combinator with left and right children being an integer and a
-function call respectively. In this case we have at least three
-choices to make as language designers:
-
-\begin{itemize}
-\item We could decide to forbid the case where two comparisons are
-  right next to each other, e.g., \syncon{forbid gt.left = gt} and
-  \syncon{forbid gt.right = gt} (where \syncon{gt} is the syncon
-  for the greater-than operator).
-\item We could change the definition of the parallel combinators
-  and define separate lexical tokens for the combinators, e.g.,
-  \syncon{token Seq = ">[[:lower:]][[:word:]]*>"}. This would add
-  just enough whitespace sensitivity to allow separating the
-  different cases.
-\item We could change the syntax of the combinators completely to
-  avoid clashes with other operators.
-\end{itemize}
-
-\noindent
-The change in the first alternative is somewhat ad-hoc, but can be done after composition
-and would allow us to reuse both language definitions without
-modifications. In the latter two alternatives, we lose reuse of
-the definitions of Orc combinators, but place no additional
-restrictions on the base language. The preferred resolution
-strategy is likely to differ between different compositions and
-different languages.
-%
-For example, another unresolvable ambiguity is going to show up if
-the base language uses semi-colons, e.g. for sequencing
-expressions (this would clash with the ``otherwise'' combinator).
-In this case, there is no reasonable way to disambiguate
-``\ocaml{e1; e2}'' without changing the syntax of one of the
-operations.
-
-The main takeaway from this case study is that resolvable
-ambiguity makes composition of languages less restrictive than if
-all ambiguity is completely banned. The approach is strictly more
-general since dynamic resolvability analysis allows deferring
-disambiguation to the programmer, while removing ambiguities from
-the resulting grammar is also possible.
-
-\subsection{Applying Resolvable Ambiguity to a Full Language} \label{sec:evaluation-ocaml}
-
-The syncon definition of the OCaml language includes most syntax described in chapters 7 and 8 in the OCaml reference manual (the base language and extensions, respectively) and is sufficient to parse roughly 75\% (1012 out of 1334 files) of the \verb|.ml| files present in the OCaml compiler itself.
-The definition is just under 700 lines, excluding comments and empty lines (just over 1000 lines otherwise).
-It has been written to be as close to the grammar presented in the reference manual as possible, but with additions to conform to how the canonical compiler behaves on cases that are underspecified in the manual.
-
-The failing cases stem from three different causes:
-
-\begin{itemize}
-\item As of yet unspecified syntax. Due to a lack of time there are still parts of the language that we have not yet implemented. These appear solvable, but we can of course not be sure until we do.
-\item Our system does not support specifying ``longest match'' on a production, which is required to handle the pattern matching constructs correctly.
-We can handle some of these cases with marks, but not all.
-\item Different definitions of precedence.
-Our translation from precedence to marks is shallow, it only considers direct children, but OCaml has deep precedence.
-For example, addition binds stronger than \ocaml{if} (which syntactically functions as a prefix operator in OCaml), thus
-  \begin{lstlisting}[language={[objective]caml}]
-  1 + let x = 1 in x + 2
-  \end{lstlisting}
-  should be parsed as
-  \begin{lstlisting}[language={[objective]caml}]
-  1 + (let x = 1 in (x + 2))
-  \end{lstlisting}
-  while our tool presents the following ambiguity error message:
-  \begin{lstlisting}[language={[objective]caml}]
-Ambiguity error with 2 alternatives:
-
-  1 + ( let x = 1 in x ) + 2
-  1 + let x = 1 in ( x + 2 )
-  \end{lstlisting}
-  The former alternative is perhaps more easily understood as
-  \begin{lstlisting}[language={[objective]caml}]
-  (1 + (let x = 1 in x)) + 2
-  \end{lstlisting}
-  Looking at this tree, two levels at a time, we see ``\ocaml{(_ + _) + 2}'', ``\ocaml{1 + let _ in _}'', and ``\ocaml{let x = 1 in x}'', all of which are individually precedence correct.
-\end{itemize}
-
-The various definitions in the syncon OCaml definition are organized by the chapter in the reference manual that introduces them, plus an appendix with additions and changes required to mimic the behavior of the canonical compiler.
-Many of these are strict additions, requiring no changes to previous syncons, but not all.
-For example, OCaml allows placing ``attributes'' on most syntactical constructs in the language, extra information visible to tools but otherwise without semantic meaning.
-These can be written with an ``infix'' notation right after the keyword that introduces the syntactical construct, e.g., \ocaml{let[@foo] x = 1 in x}.
-This syntax is not fully specified; the manual merely contains seven examples, while the testsuite uses this syntax in roughly 50 different constructs.
-All these cases require changes to the original syncon definitions, making them slightly more complicated than the manual suggests them to be upon their introduction.
-
-% \subsection{Cyphym} \label{sec:evaluation-cyphym}
-
-\section{Related Work}
-
-\citet{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
-However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
-This approach thus suggests an interesting direction for future work: to integrate it into a system designed around resolvable ambiguity, and extending the algorithms presented in this paper to cover it.
-
-\citet{danielssonParsingMixfixOperators2011} give a method for specifying grammars for expressions containing mixfix operators.
-They allow non-transitive, non-total precedence, and ``feel that it is overly restrictive to require the grammar to be unambiguous.'' Similar to our approach, they do not reject ambiguous grammars, only ambiguous parses.
-They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must be related by precedence (children must have higher precedence than parents).
-This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'-'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.
-
-Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
-For this paper, only the syntactic component \cite{schwerdfegerVerifiableCompositionDeterministic2009} of this analysis is relevant.
-This analysis guarantees that the composition of a base language and any number of extensions that have passed it will compose to a grammar in LALR(1).
-This is a far more restricted class of languages than unambiguous languages, not to mention resolvably ambiguous languages, and thus places far more restrictions on language designers than our approach.
-
-The detection of classical ambiguity in context-free grammars is undecidable in general \cite{cantorAmbiguityProblemBackus1962}, yet numerous heuristic approaches exist.
-Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using sat-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of \citet{bastenAmbiguityDetectionProgramming2011}.
-
-Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language \cite{heeringSyntaxDefinitionFormalism1989}.
-These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but also assume that the composed language is unambiguous, which in practice precludes the composition of languages constructed independently of each other.
-
-UNFINISHED
-
-\section{Conclusion}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% %% Acknowledgments
-%% \begin{acks}                            %% acks environment is optional
-%%                                         %% contents suppressed with 'anonymous'
-%%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-%%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-%%   %% acknowledge financial support and will be used by metadata
-%%   %% extraction tools.
-%%   This material is based upon work supported by the
-%%   \grantsponsor{GS100000001}{National Science
-%%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%%   conclusions or recommendations expressed in this material are those
-%%   of the author and do not necessarily reflect the views of the
-%%   National Science Foundation.
-%% \end{acks}
-
 %% Bibliography

### 2019-07-10 paper/paper.tex
Updated the abstract and the intro.

@@
 \newcommand{\labelregex}{\mathit{regex}} % Go from label the regex of the production with that label
-\newcommand{\treetorun}{\mathit{run}}
-\newcommand{\runtotree}{\mathit{tree}}
-\newcommand{\runtoword}{\mathit{word}}
-\newcommand{\runprimetorun}{\mathit{normalrun}}
 
@@
 \begin{proof}
-  By defining two functions $\treetorun$ and $\runtotree$, the former from a tree to a run, the latter from run to tree, then showing these two functions to be inverses of each other.
+  (These are two attempts at showing the same thing, next attempt starts at ``Proof sketch''. I would strongly suggest reading ``Formal stuff'' first.)
+  $A_{\posp{}}$ has a single initial state, and a single final state, call them $s$ and $f$ respectively.
+The two are distinct, and $f$ is reachable only through transitions that pop a stack symbol with $s$ and $f$ as the two first components.
+Transitions that push such symbols originate in $s$, and no other transitions originate in $s$.
+Thus all successful runs start in configuration $(s, \epsilon)$ and end in $(f, \epsilon)$, whereby every pushing transition in the run can be paired with a corresponding popping transition.
+We will now construct a tree $t \in L(T_D)$ in a bottom-up fashion.
+
+  Select a subsequence of the run $R = (s, \epsilon) \; R_1 \; (p, s) \xrightarrow{a, +(p, q, w)} (p', (p, q, w) \cdot s) \; R_2 \; (q', (p, q, w) \cdot s \xrightarrow{b, -(p, q, w)} (q, s) \; R_2 \; (f, \epsilon)$ such that $R_2$ contains no pushing or popping operations.
+The only transitions that leave the stack unchanged in $A_{\posp{}}$ come from DFAs generated from productions in $D$.
+There is no way to transition between two such DFAs without using a pushing or popping transition, thus $R_2$ must be a successful run in the DFA for some labelled production $N -> l : r$ in $D$.
+$T_D$ must contain a corresponding production $N -> l(r)$.
+Now consider the sequence of input symbols $w'$ along the transitions in $R_2$.
+We have $w' \in L(r)$, since $R_2$ is a successful run in a DFA constructed from $r$.
+Thus $l(w') \in L_N(T_D)$.
+Now add parent nodes to this subtree for every label in $w$, e.g., if $w = l_1 \cdot l_2 \cdot l_3$ construct $l_1(l_2(l_3(l(w'))))$.
+We can then replace the subsequence of $R$ with this subtree, and then repeat the process until we have a tree with root $l'$, where $D$ has a labelled production $S -> l': r'$ for some $r'$ and the starting symbol $S$.
+
+  To construct a successful run from a tree $t \in L(T_D)$ we can traverse the tree in a preorder, while at the same time constructing a run in $A_{\posp{}}$ starting from $(s, \epsilon)$.
+If the next terminal is in $\T$, i.e., a strictly zero-arity terminal, we take a transition originating in a DFA, which must be unique.
+If the next terminal is in $\Labels$, i.e., an unranked terminal, we take one of the added edges.
+However, this choice of edge might not be unique: there may be multiple edges from the current state to an appropriate DFA initial state.
+These stem from the original DFA having two transitions from the same state with the same label but different subscripts, e.g., $p \xrightarrow{l_\bot} q$ and $p \xrightarrow{l_\top} q'$.
+However, if both of these transitions were valid choices for the current tree that would imply that the DFA for that level can recognize two distinct words that differ only in subscripts.
+But distinctly subscripted terminals must originate from distinct occurances of marked non-terminals (since a $\top$ subscript corresponds to a mark, and a $\bot$ subscript corresponds to the absence of a mark), which would imply that the original regular expression is ambiguous if the marks are removed, which we explicitly ruled out in Section~\ref{sec:parse-time-disambiguation}.
+
+% TODO: technically, we should perhaps mention that these two constructions are inverses of each other, but that's relatively easy to see once you understand them.
 \end{proof}
 
-TODO: define a \emph{well-formed} linear encoding as a linear encoding without double parens. Ensure $\runtoword$ and $\lattice$ are defined.
-
 \begin{lemma}
-  All words $w \in L(A_{\posp{}})$ are well-formed linear encodings.
+  The word recognized by a successful run in $A_{\posp{}}$ is the unique linear encoding of the lattice for the corresponding $t \in L(T_D)$.
 \end{lemma}
@@
 \begin{lemma}
-  $\forall t \in L(T_D).\ \lattice(\runtoword(\treetorun(t))) = \lattice(t)$.
+  For every successful run $R'$ in $A'_{\posp{}}$ there is one (or more) successful runs $R$ in $A_{\posp{}}$ such that the lattice encoded in the word recognized by $R'$ is larger than the lattice for the tree corresponding to $R$.
 \end{lemma}
 
-\begin{lemma}
-  For all $t \in L(T_D)$ and every well-formed lattice encoding $w$ such that $\lattice(t) \subseteq \lattice(w)$, $w \in L(A'_{\posp{}})$.
-\end{lemma}
-
-\begin{lemma}
-  We can define a function $\runprimetorun$ from runs in $A'_{\posp{}}$ to runs in $A_{\posp{}}$ such that $\forall R.\ \lattice(\runtoword(\runprimetorun(R))) \subseteq \lattice(\runtoword(R))$.
-\end{lemma}
-
-\begin{proof}
-  By removing every transition that pushes or pops $\gamma$ and replacing every used transition $p \xrightarrow{'\pospl', +g} q$ that does not exist in $A_{\posp{}}$ with a transition $p \xrightarrow{'\reqpl', +g} q$.
-\end{proof}
-
 \begin{theorem}
-  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, and $\runtoword(R) = \runtoword(R')$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
+  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
 \end{theorem}
-
-\begin{proof}
-  Assume $\runprimetorun(R') = R$.
-$\runprimetorun$ either leaves its argument unchanged, or changes the recognized word.
-If $R'$ and $R$ recognize the same word, and $\runprimetorun(R') = R$, then $R' = R$, but that is a contradiction.
-
-  Since $\runprimetorun(R') \neq R$ we have two distinct trees $t_1 = \runtotree(\runprimetorun(R'))$ and $t_2 = \runtotree(R)$.
-Since $\lattice(\runtoword(\runprimetorun(R'))) \subseteq \lattice(\runtoword(R'))$ and $\lattice(\runtoword(R')) = \lattice(\runtoword(R))$ we have that $\lattice(t_1) \subseteq \lattice(t_2)$.
-\end{proof}
 
@@
   Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
-  \label{lemma:vpda-from-t}
 \end{lemma}
@@
   Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
-\end{lemma}
-
-\begin{proof}
-  Follows trivially from Lemma~\ref{lemma:vpda-from-t} and VPDAs being closed under union and difference.
-\end{proof}
-
-\begin{lemma}
 \end{lemma}

### 2019-07-10 paper/paper.tex
Almost finished first draft of Section 8

--- 2019-07-10 paper/paper.tex
@@
 
-% TODO: Make sure that forbids are explained
-
+In this section, we showcase two possible use-cases of dynamic
+resolvability analysis: composing separately defined DSLs
+(Section~\ref{sec:evaluation-orc}), and finding ambiguities in
+grammars through testing (Section~\ref{sec:evaluation-ocaml}).
+%
 We have implemented a tool and a small DSL for syntax definition
@@
 of the Earley parsing algorithm
-\cite{earleyEfficientContextfreeParsing1970}. The purpose of these
-case studies is to showcase possible use cases of resolvable
-ambiguity in the context of composing separately defined DSLs (a
-subset of Orc, Section~\ref{sec:evaluation-orc}), and of defining
-a pre-existing general purpose programming language (OCaml,
-Section~\ref{sec:evaluation-ocaml}.
+\cite{earleyEfficientContextfreeParsing1970}.
 
@@
 introduced as separate \syncon{forbid} declarations rather than
-inline in productions. This allows supporting high-level
+being inlined in productions. This allows supporting high-level
 convenience constructs, e.g., specifying precedence between
@@
 \begin{tabular}{ll}
+\small
 \begin{lstlisting}[language=syncon,boxpos=t]
@@
 &
+\small
 \begin{lstlisting}[language=syncon,boxpos=t]
@@
 
+{\small
 \begin{lstlisting}[language={[objective]caml}]
@@
 \end{lstlisting}
+}
 
@@
 \end{minipage}
+}
 
@@
 
-\subsection{Applying Resolvable Ambiguity to a Full Language} \label{sec:evaluation-ocaml}
-
-The syncon definition of the OCaml language includes most syntax described in chapters 7 and 8 in the OCaml reference manual (the base language and extensions, respectively) and is sufficient to parse roughly 75\% (1012 out of 1334 files) of the \verb|.ml| files present in the OCaml compiler itself.
-The definition is just under 700 lines, excluding comments and empty lines (just over 1000 lines otherwise).
-It has been written to be as close to the grammar presented in the reference manual as possible, but with additions to conform to how the canonical compiler behaves on cases that are underspecified in the manual.
-
-The failing cases stem from three different causes:
-
-\begin{itemize}
-\item As of yet unspecified syntax. Due to a lack of time there are still parts of the language that we have not yet implemented. These appear solvable, but we can of course not be sure until we do.
-\item Our system does not support specifying ``longest match'' on a production, which is required to handle the pattern matching constructs correctly.
-We can handle some of these cases with marks, but not all.
-\item Different definitions of precedence.
-Our translation from precedence to marks is shallow, it only considers direct children, but OCaml has deep precedence.
-For example, addition binds stronger than \ocaml{if} (which syntactically functions as a prefix operator in OCaml), thus
-  \begin{lstlisting}[language={[objective]caml}]
-  1 + let x = 1 in x + 2
-  \end{lstlisting}
-  should be parsed as
-  \begin{lstlisting}[language={[objective]caml}]
-  1 + (let x = 1 in (x + 2))
-  \end{lstlisting}
-  while our tool presents the following ambiguity error message:
-  \begin{lstlisting}[language={[objective]caml}]
-Ambiguity error with 2 alternatives:
-
-  1 + ( let x = 1 in x ) + 2
-  1 + let x = 1 in ( x + 2 )
-  \end{lstlisting}
-  The former alternative is perhaps more easily understood as
-  \begin{lstlisting}[language={[objective]caml}]
-  (1 + (let x = 1 in x)) + 2
-  \end{lstlisting}
-  Looking at this tree, two levels at a time, we see ``\ocaml{(_ + _) + 2}'', ``\ocaml{1 + let _ in _}'', and ``\ocaml{let x = 1 in x}'', all of which are individually precedence correct.
-\end{itemize}
-
-The various definitions in the syncon OCaml definition are organized by the chapter in the reference manual that introduces them, plus an appendix with additions and changes required to mimic the behavior of the canonical compiler.
+\subsection{Finding Ambiguities with Dynamic Resolvability Analysis} \label{sec:evaluation-ocaml}
+
+In this section, we investigate how to apply dynamic resolvability
+analysis to identify ambiguities in the grammar of a real-world
+language.
+%
+Using syncons, we have specified a substantial subset of OCaml's
+syntax as described in chapters 7 and 8 (base language and
+extensions) of the OCaml reference manual~\cite{OCamlReference}.
+The syncon definition is just under 700 lines of code (excluding
+comments), and is currently complete enough to parse roughly 75\%
+(1012 out of 1334 files) of the \verb|.ml| files present in the
+OCaml compiler itself (we discuss limitations of our tool in the
+end of this section).
+
+By using our implementation of the OCaml syntax to parse
+real-world OCaml code, we can dynamically identify ambiguities in
+the grammar presented in the reference manual.
+%
+For example, TODO: INSERT EXAMPLES.
+
+As this case study shows, implementing a parser that reports
+ambiguities lets us identify ambiguities in a grammar through
+testing. Even though detecting ambiguities in a grammar is
+undecidable in general, dynamic resolvability analysis over a
+large corpus of code lets us find (and fix) many cases of
+ambiguity in practice.
+%
+In this case study, out of the 700 lines of syncon code, $\sim$200
+lines are additions to conform to how the canonical compiler
+behaves on cases that are under-specified in the manual.
+
+\paragraph{Limitations of Syncons}
+%
+Other than syntactic constructs that are not yet specified, there
+are two sources of failure in the parts of the OCaml compiler that
+we cannot parse.
+%
+First, our parsing tool does not support specifying ``longest
+match'' on a production, which is required to handle the pattern
+matching constructs correctly. Some cases can be handled using
+\syncon{forbid} declarations, but it does not get us all the way.
+%
+Second, our system uses a different definition of precedence than
+the OCaml language. Our translation from precedence to implicit
+\syncon{forbid} declarations is shallow (it only considers direct
+children), while the OCaml has deep precedence. For example,
+addition binds stronger than \ocaml{let} (which syntactically
+functions as a prefix operator in OCaml), and thus
+%
+``\ocaml{1 + let x = 1 in x + 2}'' should be parsed as
+%
+``\ocaml{1 + (let x = 1 in (x + 2))}''. Our tool, however, reports
+the expression as ambiguous, additionally suggesting the
+alternative interpretation
+%
+``\ocaml{(1 + (let x = 1 in x)) + 2}''.
+%
+This interpretation is indeed valid if we only look at the direct
+children of each operator, since ``\ocaml{(_ + _) + 2}'',
+``\ocaml{1 + let _ in _}'', and ``\ocaml{let x = 1 in x}'' are all
+individually correct with regards to precedence.
+
+% The various definitions in the syncon OCaml definition are organized by the chapter in the reference manual that introduces them, plus an appendix with additions and changes required to mimic the behavior of the canonical compiler.
 Many of these are strict additions, requiring no changes to previous syncons, but not all.

### 2019-07-10 paper/paper.tex
Small addition to Section 8.2

@@
 real-world OCaml code, we can dynamically identify ambiguities in
-the grammar presented in the reference manual.
+the grammar presented in the reference manual, instead of silently
+resolving these in the implementation of an unambiguous parser.
 %
@@
 behaves on cases that are under-specified in the manual.
+
+% TODO: State that we don't report exact numbers since we don't
+% have the full language specified.
 

### 2019-07-11 paper/paper.tex
Pass over beginning of 6.3

@@
 
-This section introduces a linear encoding of the lattices of the previous chapter as words.
+This section introduces a linear encoding of the lattices of the previous section as words.
 Lemmas~\ref{lemma:same-basic} and \ref{lemma:top-bottom-determine} imply that a lattice encoding requires three things: a basic word, a set of \emph{required} parentheses, and a set of \emph{possible} parentheses.
@@
 For example, ``$\posp{\posp{1}}$'' and ``$\posp{1}$'' represent the same lattice, as do ``$\reqp{\posp{2}}$'' and ``$\reqp{2}$''.
-To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses, e.g., ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
-
-Lattice equality is now equvialent with equality of the linear encoding.
+To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses.
+For example, ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
+
+Lattice equality is now equivalent with equality of the linear encoding.
 To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
@@
 
-The centerpiece of our algorithm is a visibly pushdown automaton constructed in such a way that:
+The centerpiece of our algorithm is a \emph{visibly pushdown automaton} constructed in such a way that
 
 \begin{itemize}
-\item There is a bijection between trees in $L(T_D)$ and successful runs.
-\item The word recognized by a successful run is the linear encoding of the corresponding tree's lattice.
+\item there is a bijection between trees in $L(T_D)$ and successful runs; and
+\item the word recognized by a successful run is the linear encoding of the corresponding tree's lattice.
 \end{itemize}

### 2019-07-11 paper/paper.tex
Updated the motivating example.

@@
 A common standpoint when designing the syntax of programming languages is that
-the grammar definition has to be unambiguous.  However, up front unambiguous grammars
-can force language designers to make arbitrary choice to disambiguate
+the grammar definition has to be unambiguous.  However, requiring up front unambiguous grammars
+can force language designers to make more or less arbitrary choices to disambiguate
 the language. In this paper, we depart from the traditional view of
-unambiguous grammar design, and allow ambiguities to be delayed until
-compile time, allowing the user to perform the disambiguation. A
+unambiguous grammar design, and allow detection of ambiguities to be delayed until
+parse time, allowing the user of the language to perform the disambiguation. A
 natural decision problem follows: given a language definition, can a
@@
   ambiguity problem}---and divide it into separate static and dynamic
-resolvability problems. We prove soundness and completeness for a
-restricted language in the static case, and soundness in the general
-case for dynamic resolvability. The approach is evaluated through
-two separate case studies, covering both a large existing
-programming language, and the composability of domain-specific
-languages.
+resolvability problems. We provide solutions to the static problem
+for two restricted classes of languages and prove them sound and,
+depending on the restrictions, complete. We also provide a sound
+and complete solution to the dynamic problem for a much less
+restricted class of languages.
+%
+The approach is evaluated through two separate case studies,
+covering both a large existing programming language, and the
+composability of domain-specific languages.
 \end{abstract}
@@
 \label{sec:motivation}
-This section walks through an example from OCaml where an ambiguity error can be clearer than what the actual compiler produces, suggesting that (resolvable) ambiguity has merit even in the context of a general purpose programming language.
-
-Consider the following nested match expression (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
+In this section, we motivate the need for ambiguous language definitions, where the decision of how to disambiguate a program is taken by the end-user (the programmer) and not the language designer.
+We motivate our new methodology both for engineering of new domain-specific languages, as well as for the design of existing general-purpose programming languages.
+
+\subsection{Domain-Specific Modeling of Components}
+Suppose we are defining a new textual modeling language, where components are composed in series using an infix operator \verb~--~, and in parallel using an infix operator \verb~||~.
+For instance, an expression \verb~C1 -- C2~ puts components \verb~C1~ and \verb~C2~ in series, whereas \verb~C1 || C2~ composes them in parallel.
+In such a case, what is then the meaning of the following expression?
+\begin{verbatim}
+  C1 -- C2 || C3 || C4
+\end{verbatim}
+
+\noindent Are there natural associativity and precedence rules for these new operators?
+If there are no predefined rules of how to disambiguate this expression within the language definition, it is an ambiguous expression, and a parser generates a set of parse trees.
+Consider Fig~\ref{fig:circuits}, which depicts four different alternatives, each with a different meaning, depending on how the ambiguity has been resolved.
+Clearly, the expression has totally different meanings depending on how the end-user places the parenthesis.
+However, if a language designer is forced to make the grammar of the syntax definition unambiguous, a specific choice has to be made for precedence and associativity.
+For instance, assume that the designer makes the arbitrary choice that serial composition has higher precedence than parallel composition, and that both operators are left associative.
+In such a case, the expression without parenthesis is parsed as Figure~\ref{fig:circuits}(d).
+The question is why such an arbitrary choice---which is forced by the traditional design of unambiguous grammars---is the correct way to interpret a domain-specific expression.
+The alternative, as argued for in this paper, is to postpone the decision, and instead give an error message to the end user (programmer or modeler), and expose different alternatives that disambiguate the expression.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%&
+\begin{figure*}[!t]
+\center
+\includegraphics[width=1.0\textwidth]{circuits.pdf}
+\caption{The figure shows four different alternatives for disambiguating the expression $\texttt{C1 -- C2 || C3 || C4}$.
+Note that there is a fifth alternative $\texttt{C1 -- ((C2 || C3) || C4)}$.
+However, the meaning of this expression is the same as (c), presupposed that the parallel operator is associative.
+In such a case, this expression and the expression in (c) both means that components $\texttt{C2}$, $\texttt{C3}$, and $\texttt{C4}$ are composed in parallel.
+}
+\label{fig:circuits}
+\end{figure*}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%&
+
+% ((C1 -- C2) || C3) || C4     --> (d)
+% (C1 -- (C2 || C3)) || C4
+% C1 -- (C2 || (C3 || C4))     --> (c)
+% C1 -- ((C2 || C3) || C4)     actually same as (c), just a different layout
+% (C1 -- C2) || (C3 || C4)     --> (b)
+
+\subsection{Match Cases in OCaml}
+
+Explicit ambiguity is highly relevant also for the design of new and existing general purpose programming languages.
+The following example shows how an ambiguity error can be clearer than what an existing compiler produces today with the traditional approach.
+
+%Consider the following nested match expression (from \cite{palmkvistCreatingDomainSpecificLanguages2019}), and the resulting error message:
+
+Consider the following OCaml example of nested \verb|match| expressions, as stated by \cite{palmkvistCreatingDomainSpecificLanguages2019}).
 
@@
 
-\noindent The compiler sees the last line as belonging to the inner \ocaml{match} rather than the outer, as was intended. The solution is simple; we put parentheses around the inner match:
+\noindent The OCaml compiler output is listed to the right.
+The compiler sees the last line as belonging to the inner \ocaml{match} rather than the outer, as was intended.
+The solution is simple; we put parentheses around the inner match:
 
@@
 
-\noindent However, the connection between error message and solution is not particularly clear; surrounding an expression with parentheses does not change its type.
-
-Instead, we look to the OCaml manual for inspiration.
+\noindent However, the connection between the error message and the solution is not particularly clear; surrounding an expression with parentheses does not change its type.
+%
+%
+%Instead, we look to the OCaml manual for inspiration.
 It contains an informal description of the syntax of the language\footnote{\url{https://caml.inria.fr/pub/docs/manual-ocaml/language.html}}, in the form of an EBNF-like grammar.
-Below is an excerpt of the productions for expressions, written in a more standard variant of EBNF:
-
-\setlength{\grammarindent}{9.5em}
-\begin{grammar}
-<expr> ::= \verb|'match'| <expr> \verb|'with'| <pattern-matching>
-
-<pattern-matching> ::= \verb{'|'{$^{?}$ <pattern> (\verb|'when'| <expr>)$^{?}$ \\
-(\verb{'|'{ <pattern> (\verb|'when'| <expr>)$^{?}$ \verb|'->'| <expr>)*
-\end{grammar}
-
-\noindent If we use this grammar to parse the nested match we find an ambiguity: the last match arm can belong to either the inner match or the outer match.
+Below is an excerpt of the productions for expressions, written in a more standard variant of EBNF:%
+%
+%\setlength{\grammarindent}{9.5em}
+%\begin{grammar}
+%<expr> ::= \verb|'match'| <expr> \verb|'with'| <pattern-matching>
+%
+%<pattern-matching> ::= \verb{'|'{$^{?}$ <pattern> (\verb|'when'| <expr>)$^{?}$ \\
+%(\verb{'|'{ <pattern> (\verb|'when'| <expr>)$^{?}$ \verb|'->'| <expr>)*
+%\end{grammar}
+%
+%\noindent If we use this grammar to parse the nested match we find an ambiguity: the last match arm ca%n belong to either the inner match or the outer match.
+%
 The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended.
-Presenting the two alternatives as parse trees would here be informative, and useful to a language designer, but would also expose implementation details to an end-user.
+Instead, our approach is to postpone the disambiguation to the end user, and present the two alternatives as parse trees, as part of the error message.
+
+%would here be informative, and useful to a language designer, but would also expose implementation details to an end-user.
 Instead, we can present the alternatives in the form of modified code that parses unambiguously as each corresponding alternative:
 
-\begin{center}
-\begin{tabular}{l|l}
-\begin{lstlisting}[language={[objective]caml}]
-match 1 with
-  | 1 -> (match "one" with
-          | str -> str)
-  | 2 -> "two"
-\end{lstlisting} &
-\begin{lstlisting}[language={[objective]caml}]
-match 1 with
-  | 1 -> (match "one" with
-          | str -> str
-  | 2 -> "two")
-\end{lstlisting} \\
-\end{tabular}
-\end{center}
+%\begin{center}
+%\begin{tabular}{l|l}
+%\begin{lstlisting}[language={[objective]caml}]
+%match 1 with
+%  | 1 -> (match "one" with
+%          | str -> str)
+%  | 2 -> "two"
+%\end{lstlisting} &
+%\begin{lstlisting}[language={[objective]caml}]
+%match 1 with
+%  | 1 -> (match "one" with
+%          | str -> str
+%  | 2 -> "two")
+%\end{lstlisting} \\
+%\end{tabular}
+%\end{center}
+
+\subsection{Resolvable Ambiguity}
+The two motivating examples above motivate briefly why grammar
+and language definitions should not always be unambiguous. However, an
+ambiguous grammar that cannot be resolved by the end user is clearly
+undesirable. In the rest of the paper, we formalize the approach of
+resolvable ambiguity, and show how we can give guarantees for
+resolvability.
 

### 2019-07-11 paper/paper.tex
Added a new conclusion.

--- 2019-07-11 paper/paper.tex
@@
 \section{Conclusion}
+In this paper, we introduce the concept of \emph{resolvable ambiguity}.
+This means that a language syntax definition can be ambiguous, but the ambiguity is resolvable later by the end user when compiling a program.
+This approach departs from the common standpoint that grammars and syntax definitions of languages must be unambiguous.
+We motivate our work both for general-purpose languages in general, but for domain-specific languages in particular.
+As part of the new concept, we formalize the fundamental \emph{resolvable ambiguity problem}, and divide it into static and dynamic parts.
+We prove soundness and completeness of static resolvablity for a restricted class of languages, and soundness and completeness of dynamic resolvability for general context-free languages with balanced parenthesis.
+The approach is evaluated on two case studies, including the definition of a large subset of OCaml, and on the composition of domain-specific languages.
 

### 2019-07-11 paper/paper.tex
More work on 6.3

--- 2019-07-11 paper/paper.tex
@@
 
+\subsubsection{Example for Simplified Construction}
+
 We will now walk through the construction of this automaton, using the following language definition:
@@
 
-\noindent We then combine them into a single visibly pushdown automaton.
-Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we go back.
+\noindent We then combine the DFAs into a single visibly pushdown automaton.
+Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we return.
 We use $Q \times Q$ as our stack alphabet, to record the transition we are replacing in the parent.
@@
 
-\noindent Finally, we add a new initial state and a new final state, and connect them with the initial and final states belonging to the starting non-terminal:
+\noindent Finally, we add a new initial state and a new final state, and connect them to the initial and final states belonging to the starting non-terminal:
 
@@
 \noindent and corresponds to the tree on the right.
+
+\subsubsection{Example for Full Construction}
 

### 2019-07-11 paper/paper.tex
Be more positive!

@@
 
-\noindent Unfortunately, this automaton recognizes the word $(((e)))$, which is not the unique linear encoding we require; it should be $(e)$.
+\noindent Unfortunately, this automaton recognizes the word $(((e)))$,
+which is not the unique linear encoding we require; it should be
+$(e)$. The reason for this mismatch is that each of the productions
+with labels $a$ and $b$ introduces an additional pair of parentheses
+in the word recognized by the automaton. To ensure that the automaton
+instead recognizes the required unique linear encoding $(e)$, we
+extend the construction of the automaton such that ``passing through''
+a non-terminal without recognizing any terminals does \emph{not} add
+parentheses. The idea is to skip the DFAs of such non-terminals, but
+still keep track of the skipped production(s) in the symbols pushed to
+the stack; thus, runs which differ in skipped non-terminals are still
+distinguished.
+
+For our example, the resulting automaton looks as follows:
 
@@
 
-UNFINISHED
+Note that the transition from state 8 to state 6 pushes the symbol
+$(8, 9, ab)$ to the stack. This means (a) that we started the
+``replacement'' for the transition from 8 to 9, and (b) that we passed
+through the productions with labels $a$ and $b$ without consuming
+terminals, thus ending up in the starting state of the DFA
+corresponding to production $c$.
 
@@
 
-\paragraph{Limitations of Syncons}
+\paragraph{Current Limitations of Syncons}
 %

### 2019-07-11 paper/paper.tex
Add explanations to informal part in 6.3
Pass over Introduction

@@
 % Motivate why unambiguous parsing is a good (and important) idea.
-Ever since the early 60s, it is known to be undecidable to determine whether a context-free
-grammar is ambiguous~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
+Ever since the early 60s, it has been known that determining whether a context-free
+grammar is ambiguous is undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
 a large number of restricted grammars have been developed to guarantee
-that a language definition is unambiguous. This traditional view of
+that language definitions are unambiguous. This traditional view of
 only allowing unambiguous grammars has been taken for granted as the
@@
 definitions. Specifically, instead of forcing language designers to
-make arbitrary choice to disambiguate the language, a more natural
-choice may be to postpone the decision to the programmer.
+make more or less arbitrary choices to disambiguate the language, a more natural
+solution may be to postpone the decision to the programmer.
 
 % Motivation II, compising DSLs
-Another area---where ambiguous grammars naturally arises---is
-domain-specific language development.  An important
-objective when designing domain-specific languages is to construct
+Another area where ambiguous grammars naturally arise is
+in domain-specific language development.  An important
+objective when designing domain-specific languages is to be able construct
 languages by composing and extending existing
 languages.
-Unfortunately, the composition of restricted
+Unfortunately, the composition of
 %grammars easily results in ambiguous grammars, which again are
@@
 grammars easily produces
-an ambiguous grammar, even if all of the original grammars are
-unambiguous. Approaches for solving this problem can be broadly split
+an ambiguous grammar, even if the original grammars are all
+unambiguous on their own. Approaches for solving this problem can be broadly split
 into two categories: those that handle ambiguities on the
@@
 is not guaranteed to be unambiguous. Concretely, if parsing a program
-results in several parse trees, the grammar is obviously
+could result in several correct parse trees, the grammar is obviously
 ambiguous. The programmer then needs to disambiguate the program by,
-for instance, inserting extra parenthesis. We say that a program has a
-\emph{resolvable ambiguity} if the ambiguity can be solved by
+for instance, inserting extra parentheses. We say that a program is
+\emph{resolvably ambiguous} if the ambiguity can be resolved by
 modifying the program, such that each of the possible parse trees can
 be selected using different modifications.
-We can divide the problem
-related to resolvable ambiguity into several different decision
-problems.
-The \emph{dynamic resolvable ambiguity problem} is as follows: given a parse function for a language, and a specific program that is parsed into a set of parse trees, decided whether it possible to rewrite the program, such that each of the parse trees can be generated unambiguously.
-This problem can be generalized into the \emph{static resolvable ambiguity problem}: decide whether all possible parsable programs are resolvably ambiguous.
-Static resolvablity means that a language designer can be confident that the end-user (the program designer) always can resolve an ambiguity by modifying the program.
-By contrast, dynamic resolvability means that a language designed knows if a \emph{specific} program is resolvable.
-Hence, static resolvability gives formally verified guarantees on the syntax definition, whereas dynamic resolvability enables the language designer to test the resolvability of the syntax definition, program by program.
+%
+We can divide the problem of resolvable ambiguity into two
+different decision problems.
+%
+The \emph{dynamic resolvable ambiguity problem} is as follows:
+given a parse function for a language, and a specific program that
+is parsed into a set of parse trees, decide whether it possible to
+rewrite the program, such that each parse tree can be generated
+unambiguously.
+%
+This problem can be generalized into the \emph{static resolvable
+  ambiguity problem}: given a parse function, decide whether all
+parsable programs are resolvably ambiguous.
+%
+Static resolvablity means that a language designer can be
+confident that the end-user (the programmer) can always resolve an
+ambiguity by modifying the program. Identifying such ambiguities
+for a \emph{specific} program and suggesting disambiguations is
+enabled by dynamic resolvability.
+%
+Static resolvability gives formally verified guarantees on a
+syntax definition, whereas dynamic resolvability enables the
+language designer to test the resolvability of the syntax
+definition, program by program.
 
@@
 
-In this paper, we solve the static resolvability problems for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when designing language syntaxes.
+In this paper, we solve the static resolvability problems for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when defining language syntax.
 More specifically, we make the following contributions:
@@
 \begin{itemize}
-\item We formally define the term \emph{resolvable ambiguity}, as well as giving precise formalizations of the static and dynamic \emph{resolvable ambiguity problems} (Section~\ref{sec:resolvable-definition}).
-\item We formulate the \emph{static resolvability problem} of increasing difficulty, by introducing two different restricted language versions.
-We formalize a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
-\item We define a decidable algorithm that solves the \emph{dynamic resolvability problem} for general context-free languages with well-balanced parentheses (Section~\ref{sec:dynamic}).
-\item The dynamic resolvability methodology is implemented as a tool and a small DSL for defining syntax definitions.
-Two different cases studies are evaluated using this methodology and tool (Section~\ref{sec:evaluation}): (i) an implementation and evaluation of a large subset of OCaml's syntax, where we study the effect of implementing an underspecified language definition, and (ii) a domain-specific language for orchestrating parallel computations, where we investigate the effects of composing language fragments.
+\item We formally define the term \emph{resolvable ambiguity}, as
+  well as giving precise formalizations of the static and dynamic
+  \emph{resolvable ambiguity problems}
+  (Section~\ref{sec:resolvable-definition}).
+\item We look at versions of the \emph{static resolvability
+    problem} with increasing difficulty by considering two
+  differently restricted classes of languages. We formalize a
+  decidable algorithm that is sound and complete for the first
+  version, and sound for the second (Section~\ref{sec:static}).
+\item We define a decidable algorithm that solves the
+  \emph{dynamic resolvability problem} for general context-free
+  languages with well-balanced parentheses
+  (Section~\ref{sec:dynamic}).
+\item We evaluate the dynamic resolvability methodology in two
+  different case studies using a tool and a small DSL for defining
+  syntax definitions (Section~\ref{sec:evaluation}): (i) a
+  domain-specific language for orchestrating parallel
+  computations, where we investigate the effects of composing
+  language fragments, and (ii) an implementation and evaluation of
+  a large subset of OCaml's syntax, where we study the effect of
+  implementing an underspecified language definition.
 \end{itemize}

### 2019-07-11 paper/paper.tex
Spelling

@@
 \begin{document}
-% TODO: Decide on British or American English.
-
 % Submission may be up to 25 pages, excluding references
@@
 %% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followe by
+%% Each author should be introduced by \author, followed by
 %% \authornote (optional), \orcid (optional), \affiliation, and
@@
 
-% Motivation II, compising DSLs
+% Motivation II, composing DSLs
 Another area where ambiguous grammars naturally arise is
@@
 %
-Static resolvablity means that a language designer can be
+Static resolvability means that a language designer can be
 confident that the end-user (the programmer) can always resolve an
@@
 
-% TODO: I feel like there should be a clear transition between the more pracctical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.
+% TODO: I feel like there should be a clear transition between the more practical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.
 
@@
   a large subset of OCaml's syntax, where we study the effect of
-  implementing an underspecified language definition.
+  implementing an under-specified language definition.
 \end{itemize}

### 2019-07-11 paper/paper.tex
Update and polish Section 5

@@
 
-The primary purpose of this formalism is, as described in the previous section, to produce a $\parse$ function, i.e., to describe a word language and assign one or more interpretations to each word.
-The interpretations will be unranked trees, intended to be somewhat reminiscent of the abstract syntax trees used in most compilers.
-Section~\ref{sec:resolvable-definition} suggests that $\parse$ discard some information, to enable the resolution of some ambiguities; we here choose to discard grouping parentheses.
+The primary purpose of this formalism is, as described in the previous section, to produce a $\parse$ function, i.e., to describe a word language and assign one or more parse trees to each word.
+Parse trees are unranked trees; Section~\ref{sec:resolvable-definition} suggests that $\parse$ discards some information, to enable the resolution of some ambiguities; here, we choose to discard grouping parentheses.
 
@@
 This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}.
-In practice, this is likely desireable regardless, since a user is likely to attach different semantic meaning to each occurrence of a non-terminal.
-
-Each non-terminal appearing in the regular expression of a production carry a \emph{mark} $m$, which is a set of labels whose productions may \emph{not} replace that non-terminal.
-To lessen clutter, we will write $E_\emptyset$ as $E$.
-As an example, consider the language definition in Fig.~\ref{fig:running-example-definition}, which will be used as a running example.
+In practice, this is likely desirable regardless, since a user is likely to attach different semantic meanings to each occurrence of a non-terminal.
+
+Each non-terminal appearing in the regular expression of a production carries a \emph{mark} $m$ which is a set of labels whose productions may \emph{not} replace that non-terminal.
+To lessen clutter, we write $E_\emptyset$ as $E$.
+Consider the language definition shown in Fig.~\ref{fig:running-example-definition} which we use as a running example.
 In the production describing multiplication ($m$) both non-terminals are marked with $\{a\}$, which thus forbids addition from being a direct child of a multiplication.
-By ``direct child'' we mean ``without an intermediate node'', most commonly grouping parentheses, thus this enforces conventional precedence.
+By ``direct child'' we mean ``without an intermediate node'', most commonly grouping parentheses; thus, this enforces conventional precedence.
 
@@
 From $D$ we then generate four grammars: $G_D$, $T_D$, $G'_D$, and $T'_D$.
-Technically, only $G'_D$ and $T_D$ are required, $G'_D$ is used as the defined word language and $T_D$ as the interpretations, but the remaining two grammars help the presentation.
+Technically, only $G'_D$ and $T_D$ are required, $G'_D$ is used as the defined word language and $T_D$ as the parse trees, but the remaining two grammars help the presentation.
 
@@
 
-\noindent The latter is mostly useful in later sections, but $\parse$ allow us to consider some concrete examples of resolvable and unresolvable ambiguities.
-For example, in our running example (Fig.~\ref{fig:running-example-definition}), the word \texttt|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
+\noindent The latter is mostly useful in later sections, but $\parse$ allows us to consider some concrete examples of resolvable and unresolvable ambiguities.
+For example, in our running example (Fig.~\ref{fig:running-example-definition}), the word \verb|'1 + 2 + 3'| is ambiguous, since $\parse($\verb|'1 + 2 + 3'|$) = \{t_1, t_2\}$ where
 
@@
 \noindent In this case, $t_4$ has an unambigous word (namely \verb|'[(1 ; 2)]'|), but $t_3$ does not.
-The solution is to forbid list elements from being sequences by modify the language definition in Fig.~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.
+The solution is to forbid list elements from being sequences by modifying the language definition in Fig.~\ref{fig:running-example-definition} so that both non-terminals in the production $l$ are marked with $s$ (i.e., they look like $E_{\{s\}}$), at which point $\parse($\verb|'[1 ; 2]'|$) = \{t_3\}$.
 

### 2019-07-11 paper/paper.tex
Sketchy section 6

@@
 This lattice is bounded, with top and bottom elements $\topw(t)$ and $\botw(t)$.
+We denote this lattice by $\lattice(t)$.
 Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
@@
 For example, the lattice in Fig.~\ref{fig:lattice2} is represented by ``$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ``$(1 + 2) * 3$'' has a lattice represented by ``$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
+We denote the lattice encoded by such a word by $\lattice(w)$.
 With this encoding we can apply the formidable body of knowledge available for word languages, albeit with some extra care; the linear encoding of a lattice is not unique.
@@
 For example, ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
+We define a linear encoding of this form to be \emph{well-formed}.
 
 Lattice equality is now equivalent with equality of the linear encoding.
-To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
+To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a subset) by replacing a required pair with a possible pair.
 For example, ``$\reqp{1}2$'' is entirely contained in ``$\reqp{1}\posp{2}$'', which is entirely contained in ``$\posp{1}\posp{2}$''.
@@
 
-TODO: define a \emph{well-formed} linear encoding as a linear encoding without double parens. Ensure $\runtoword$ and $\lattice$ are defined.
-
 \begin{lemma}
@@
 \end{lemma}
+
+We denote the word recognized by a run $R$ by $\runtoword(R)$.
 
@@
 \begin{lemma}
-  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
+  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can, in finite time, construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
 \end{lemma}

### 2019-07-11 paper/paper.tex
Commented out some text to make section 3 shorter.

@@
 This section briefly describes the theoretical foundations we build upon.
-Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation to make later sections easier to read.
+
+%Sections~\ref{sec:preliminaries-cfgs} and \ref{sec:preliminaries-automata} describe context-free grammars and various forms of automata, the latter with some non-standard notation to make later sections easier to read.
 Section~\ref{sec:preliminaries-vpls} then describes visibly pushdown languages, which enable the analyses described in Sections~\ref{sec:static} and \ref{sec:dynamic}.
@@
 
-\subsection{Bags} \label{sec:bags}
-
-A \emph{bag}, or \emph{multiset}, is a generalization of a set; it is an unordered collection where each element may appear multiple times.
+%\subsection{Bags} \label{sec:bags}
+
+%A \emph{bag}, or \emph{multiset}, is a generalization of a set; it is an unordered collection where each element may appear multiple times.
 The number of times an element $a$ appears in a bag $B$ is called the \emph{multiplicity} of $a$ in $B$ and is written $m_B(a)$.
@@
 
-\subsection{Regular Expressions} \label{sec:preliminaries-regexes}
-
-A regular expression $r$ over alphabet $\T$ is defined inductively:
-
-$$r ::= \epsilon \mid a \mid r \cdot r \mid r + r \mid r^{*}$$
-
-\noindent where $a \in \T$. The language of a regular expression $r$ is given by $L(r)$:
-
-$$
-\begin{array}{r@{\ =\ }l}
-  L(t) & \{a\} \\
-  L(\epsilon) & \{\epsilon\} \\
-  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \} \\
-  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
-  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
-\end{array}
-$$
-
-\noindent We denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
+%\subsection{Regular Expressions} \label{sec:preliminaries-regexes}
+
+%A regular expression $r$ over alphabet $\T$ is defined inductively:
+
+%$$r ::= \epsilon \mid a \mid r \cdot r \mid r + r \mid r^{*}$$
+
+%\noindent where $a \in \T$. The language of a regular expression $r$ is given by $L(r)$:
+
+%$$
+%\begin{array}{r@{\ =\ }l}
+%  L(t) & \{a\} \\
+%  L(\epsilon) & \{\epsilon\} \\
+%  L(r_1 \cdot r_2) & \{ w_1 \cdot w_2 \mid w_1 \in L(r_1), w_2 \in L(r_2) \} \\
+%  L(r_1 + r_2) & L(r_1) \cup L(r_2) \\
+%  L(r^{*}) & L(\epsilon + r \cdot r^{*}) \\
+%\end{array}
+%$$
+
+%\noindent We denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
 
@@
 
-A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$\footnote{Where $^{*}$ is the Kleene star.}, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
-
+A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
+
+%\footnote{Where $^{*}$ is the Kleene star.}
+%
 A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
 Such a sequence is called a \emph{derivation}.
-The set of words recognized by $G$ is written $L(G)$\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.
-
-\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
+The set of words recognized by $G$ is written $L(G)$
+%\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.
+
+%\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
 
@@
 \noindent i.e., the terminal recognized by a transition fully determines the change to the stack height.
-
+%
 The names of the partitions stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
-
+%
 This partitioning gives us some useful properties. Of particular relevance to this paper are the following two points:
@@
   \begin{array}{r@{\ =\ }ll}
-    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \text{where } c \in \T_c \\
-    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \text{where } i \in \T_i \\
-    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \text{where } r \in \T_r \\
+    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \!\!\!\!\text{where } c \in \T_c \\
+    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \!\!\!\!\text{where } i \in \T_i \\
+    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \!\!\!\!\text{where } r \in \T_r \\
   \end{array}

### 2019-07-11 paper/paper.tex
First complete draft of Section 8

@@
 Because of how we defined the syntax of the combinators, and since
-parsing with the tool is currently whitespace insensitive, the
+the tool is currently whitespace insensitive, the
 multi-character combinators (sequencing and pruning) are parsed as
@@
   different cases.
-\item We could change the syntax of the combinators completely to
-  avoid clashes with other operators.
+\item We could change the syntax of the combinators to avoid
+  clashes with other operators.
 \end{itemize}
@@
 In this section, we investigate how to apply dynamic resolvability
-analysis to identify ambiguities in the grammar of a real-world
-language.
+analysis to identify ambiguities in the specification of a
+real-world language.
 %
@@
 extensions) of the OCaml reference manual~\cite{OCamlReference}.
-The syncon definition is just under 700 lines of code (excluding
-comments), and is currently complete enough to parse roughly 75\%
-(1012 out of 1334 files) of the \verb|.ml| files present in the
-OCaml compiler itself (we discuss limitations of our tool in the
-end of this section).
+The syncon definition is just under 700 lines of code, and is
+currently complete enough to parse roughly 75\% (1012 out of 1334
+files) of the \verb|.ml| files present in the OCaml compiler
+itself (we discuss limitations of our tool in the end of this
+section).
 
@@
 %
-For example, TODO: INSERT EXAMPLES.
+For example, an OCaml expression can be a function application
+$\mathit{expr}~\{\mathit{argument}\}^{+}$ or a constructor
+application $\mathit{constr}~\mathit{expr}$. Since an expression
+can also be a constructor, however, the language definition allows
+``\ocaml{Foo 42}'' to be parsed both as a function application and
+as a constructor application (only the latter is well-typed, but
+both are \emph{syntactically} well-formed).
+%
+Similarly, the expression ``\ocaml{Foo.f}'' can be parsed as a
+field access targeting either a constructor or a module, since
+both must start with capital letters (again, only the latter is
+well-typed).
+%
+Another example, that was already mentioned in
+Section~\ref{sec:parse-time-disambiguation}, is the fact that
+semi-colons are used both for sequencing and for separating
+elements in lists, arrays and records. Thus, ``\ocaml{[1; 2]}''
+can be parsed both as a list of two elements and a singleton list
+equivalent to ``\ocaml{[(1; 2)]}''.
 
@@
 lines are additions to conform to how the canonical compiler
-behaves on cases that are under-specified in the manual.
+behaves on cases that are under-specified in the
+manual, including the examples listed above.
 
@@
 %
-Other than syntactic constructs that are not yet specified, there
-are two sources of failure in the parts of the OCaml compiler that
-we cannot parse.
+Other than syntactic OCaml constructs that are not yet specified,
+there are two sources of failure in the parts of the OCaml
+compiler that we cannot parse.
 %

### 2019-07-11 paper/paper.tex
Polish abstract and intro

--- 2019-07-11 paper/paper.tex
@@
 the language. In this paper, we depart from the traditional view of
-unambiguous grammar design, and allow detection of ambiguities to be delayed until
+unambiguous grammar design, and enable the detection of ambiguities to be delayed until
 parse time, allowing the user of the language to perform the disambiguation. A
@@
 make more or less arbitrary choices to disambiguate the language, a more natural
-solution may be to postpone the decision to the programmer.
+solution may be to leave the decision up to the programmer.
 
@@
 given a parse function for a language, and a specific program that
-is parsed into a set of parse trees, decide whether it possible to
+is parsed into a set of parse trees, decide whether it is possible to
 rewrite the program, such that each parse tree can be generated
@@
 
-In this paper, we solve the static resolvability problems for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when defining language syntax.
+In this paper, we solve the static resolvability problem for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when defining language syntax.
 More specifically, we make the following contributions:
@@
 \begin{itemize}
-\item We formally define the term \emph{resolvable ambiguity}, as
-  well as giving precise formalizations of the static and dynamic
+\item We formally define the term \emph{resolvable ambiguity}, and
+  provide precise formalizations of the static and dynamic
   \emph{resolvable ambiguity problems}
   (Section~\ref{sec:resolvable-definition}).
-\item We look at versions of the \emph{static resolvability
+\item We consider versions of the \emph{static resolvability
     problem} with increasing difficulty by considering two
@@
   version, and sound for the second (Section~\ref{sec:static}).
-\item We define a decidable algorithm that solves the
+\item We devise a decidable algorithm that solves the
   \emph{dynamic resolvability problem} for general context-free
@@
   (Section~\ref{sec:dynamic}).
-\item We evaluate the dynamic resolvability methodology in two
+\item We evaluate the dynamic resolvability methodology in the context of two
   different case studies using a tool and a small DSL for defining
@@
 \noindent As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF), where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
-Before the above contributions are introduced, the paper starts with motivating examples (Section~\ref{sec:motivation}), and a section with preliminaries (Section~\ref{sec:prel}).
+Before presenting the above contributions, the paper starts with motivating examples (Section~\ref{sec:motivation}), as well as preliminaries (Section~\ref{sec:prel}).
 
@@
 This lattice is bounded, with top and bottom elements $\topw(t)$ and $\botw(t)$.
-We denote this lattice by $\lattice(t)$.
 Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
@@
 For example, the lattice in Fig.~\ref{fig:lattice2} is represented by ``$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ``$(1 + 2) * 3$'' has a lattice represented by ``$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
-We denote the lattice encoded by such a word by $\lattice(w)$.
 With this encoding we can apply the formidable body of knowledge available for word languages, albeit with some extra care; the linear encoding of a lattice is not unique.
@@
 For example, ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
-We define a linear encoding of this form to be \emph{well-formed}.
 
 Lattice equality is now equivalent with equality of the linear encoding.
-To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a subset) by replacing a required pair with a possible pair.
+To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a strict superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a strict subset) by replacing a required pair with a possible pair.
 For example, ``$\reqp{1}2$'' is entirely contained in ``$\reqp{1}\posp{2}$'', which is entirely contained in ``$\posp{1}\posp{2}$''.
@@
 
+TODO: define a \emph{well-formed} linear encoding as a linear encoding without double parens. Ensure $\runtoword$ and $\lattice$ are defined.
+
 \begin{lemma}
@@
 \end{lemma}
-
-We denote the word recognized by a run $R$ by $\runtoword(R)$.
 
@@
 \begin{lemma}
-  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can, in finite time, construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
+  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
 \end{lemma}

### 2019-07-11 paper/paper.tex
Minor fixes in Section 2

@@
 The question is why such an arbitrary choice---which is forced by the traditional design of unambiguous grammars---is the correct way to interpret a domain-specific expression.
-The alternative, as argued for in this paper, is to postpone the decision, and instead give an error message to the end user (programmer or modeler), and expose different alternatives that disambiguate the expression.
+The alternative, as argued for in this paper, is to postpone the decision, and instead give an error message to the end-user (programmer or modeler), and expose different alternatives that disambiguate the expression.
 
@@
 The OCaml compiler makes an arbitrary choice to remove the ambiguity, which may or may not be the alternative the user intended.
-Instead, our approach is to postpone the disambiguation to the end user, and present the two alternatives as parse trees, as part of the error message.
+With a parser that is aware of possible ambiguities, the disambiguation can be left to the end-user, with the alternatives listed as part of an error message.
 
@@
 The two motivating examples above motivate briefly why grammar
-and language definitions should not always be unambiguous. However, an
-ambiguous grammar that cannot be resolved by the end user is clearly
+and language definitions do not always need to be unambiguous. However, an
+ambiguous program that cannot be resolved by the end-user is clearly
 undesirable. In the rest of the paper, we formalize the approach of

### 2019-07-11 paper/paper.tex
Minor updates in section 4.

@@
 A formal language is defined as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$.
-For defining \emph{resolvable ambiguity} we additionally have to consider the results of parsing words.
+To be able to define \emph{resolvable ambiguity}, we additionally have to consider the results of parsing words.
 In order to stay as general as possible, we first define the notion of an \emph{abstract parser}.
@@
 
-This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}
+\noindent This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}
 
@@
 \noindent For example, consider a simple arithmetic language without precedence and parentheses.
-In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two parse trees shown in Fig.~\ref{fig:arith-example}.
-
-\begin{figure}[t]
-\begin{tabular}{cc}
-  \Tree [.{$+$}
-    1
-    [.{$\cdot$}
-      2
-      3 ] ] &
-  \Tree [.{$\cdot$}
-    [.{$+$}
-      1
-      2 ]
-    3 ] \\
-\end{tabular}
-\caption{The two allowable parse trees of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
-\label{fig:arith-example}
-\end{figure}
-
+In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing two parse trees.
+%the two parse trees shown in Fig.~\ref{fig:arith-example}.
+%
+%\begin{figure}[t]
+%\begin{tabular}{cc}
+%  \Tree [.{$+$}
+%    1
+%    [.{$\cdot$}
+%      2
+%      3 ] ] &
+%  \Tree [.{$\cdot$}
+%    [.{$+$}
+%      1
+%      2 ]
+%    3 ] \\
+%\end{tabular}
+%\caption{The two allowable parse trees of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
+%\label{fig:arith-example}
+%\end{figure}
+%
 The ambiguity of a word $w \in \T^{*}$ is defined in terms of $\parse$:
@@
   is ambiguous, written $\amb_P(w)$, iff
-
+%
   $\exists t_1, t_2 \in T.\ t_1 \neq t_2 \land \{t_1, t_2\} \subseteq \parse(w)$
@@
 
-The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way.
-However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously parsed as the first and second tree, respectively.
+The example $\parse(1 + 2 \cdot 3)$
+%in Fig.~\ref{fig:arith-example}
+is not resolvably ambiguous, since neither of the two resulting trees can be written in any other way.
+However, if we add grouping parentheses to the language, we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously parsed as the first and second tree, respectively.
 
@@
 
-Before introducing this formalism however, we introduce the two central problems we consider in this paper:
+Before introducing this formalism, however, we introduce the two central problems we consider in this paper:
 
@@
 \begin{definition}[Soundness of Static Resolvability]\label{def:static-procedure-sound}
-  A decision procedure $f$ solving the static resolvability problem is sound iff
+  A decision procedure $f$, solving the static resolvability problem, is sound iff
 
@@
 \begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
-  A decision procedure $f$ solving the static resolvability problem is \emph{complete} iff
+  A decision procedure $f$, solving the static resolvability problem, is \emph{complete} iff
 
@@
 \begin{definition}[Soundness of Dynamic Resolvability]\label{def:dynamic-procedure-sound}
-  A decision procedure $f$ solving the dynamic resolvability problem is sound iff
+  A decision procedure $f$, solving the dynamic resolvability problem, is sound iff
 
@@
 
-Analogous to the static case, for the parse languages studied in the
-following sections we can also provide a decision procedure $g$ such
+\noindent Analogous to the static case, for the parse languages studied in the
+following sections, we can also provide a decision procedure $g$ such
 that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
@@
 \begin{definition}[Completeness of Dynamic Resolvability]\label{def:dynamic-procedure-complete}
-  A decision procedure $f$ solving the dynamic resolvability problem is complete iff
+  A decision procedure $f$, solving the dynamic resolvability problem, is complete iff
 

### 2019-07-11 paper/paper.tex
Improve formulation of theorem 6.1 and minor polish

@@
 
-For this section, we will use an alternative formulation of resolvable ambiguity for languages, stated in terms of interpretations instead of words:
+For this section, we will use an alternative formulation of resolvable ambiguity for languages, stated in terms of parse trees instead of words:
 
 \begin{theorem}
-  A language given by $\parse : \T^{*} -> 2^T$ is resolvably ambiguous iff\\$\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
+  Given an abstract parser $P = (\T, T, \parse)$, the language $L(P)$ is resolvably ambiguous iff $\quad\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
 \end{theorem}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsubsection{Formalization}
 

### 2019-07-11 paper/paper.tex
Basically complete dynamic section

@@
 Furthermore, for practical reasons, if the word is resolvably ambiguous we wish to produce a (minimal) witness for each tree.
+Similarly to Section~\ref{sec:static} we place some restrictions on $D$, though the restrictions are less severe:
+
 Additionally, while Section~\ref{sec:static} requires a language definition $D$ to not contain parentheses, this section merely requires parentheses to be balanced.
 
+\begin{enumerate}
+\item Each right-hand side regular expression must only recognize words with balanced parentheses, e.g., ``$\verb|(| \verb|)|$'' is permissible, but ``$\verb|(|^{*} \verb|)|^{*}$'' is not.
+\item $G_D$, but with parentheses removed, must not be infinitely ambiguous.
+\end{enumerate}
+
+\noindent Both of these restrictions can be checked statically.
+
 Our approach centers around around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
-Each such language turns out to be a visibly pushdown language.
+Given restriction (1), each such language turns out to be a visibly pushdown language.
 VPLs are closed under difference, which enables us to create a visibly pushdown automaton for the language $\words(t_1) \setminus \words(t_2)$ for any pair of trees $t_1, t_2 \in L(T_D)$.
@@
 \begin{lemma}
+  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$, the set of trees $t' \in L(T_D)$ such that $\words(t) \cap \words(t') \neq \emptyset$ is finite.
 \end{lemma}
 
-UNFINISHED
+\begin{proof}
+  Lemma~\ref{lemma:same-basic}, presented in Section~\ref{sec:word-view}, holds since restriction (1) implies only well-balanced parentheses, thus all $t'$ must share a basic word with all $t_i$.
+$G_D$ with parentheses removed recognizes these basic words, and has a one-to-one correspondence between leftmost derivations and trees in $L(T_D)$, by construction.
+Restriction (2) states that $G_D$ without parentheses is not infinitely ambiguous, thus there must be a finite number of trees that have the same basic word as all $t_i$.
+\end{proof}
+
+Restriction (2) also implies that $G'_D$ is not infinitely ambiguous, whereby $\parse(w)$ must be finite.
+The dynamic analysis thus parses $\basic(w)$ using $G_D$ with parentheses removed to produce a set $R$ of relevant trees.
+For every $t \in parse(w)$, we construct a VPDA recognizing $\words(t) \setminus \bigcup_{t' \in R, t \neq t'}\words(t')$.
+If none of these VPDAs recognize the empty language, then the word is resolvably ambiguous, otherwise it is unresolvably ambiguous.
 

### 2019-07-11 paper/paper.tex
Pass over Conclusion, adding future work

@@
 \section{Conclusion}
-In this paper, we introduce the concept of \emph{resolvable ambiguity}.
-This means that a language syntax definition can be ambiguous, but the ambiguity is resolvable later by the end user when compiling a program.
-This approach departs from the common standpoint that grammars and syntax definitions of languages must be unambiguous.
-We motivate our work both for general-purpose languages in general, but for domain-specific languages in particular.
-As part of the new concept, we formalize the fundamental \emph{resolvable ambiguity problem}, and divide it into static and dynamic parts.
-We prove soundness and completeness of static resolvablity for a restricted class of languages, and soundness and completeness of dynamic resolvability for general context-free languages with balanced parenthesis.
-The approach is evaluated on two case studies, including the definition of a large subset of OCaml, and on the composition of domain-specific languages.
+In this paper, we have introduced the concept of \emph{resolvable
+  ambiguity}. A language grammar is resolvably ambiguous if all
+ambiguities can be resolved by the end-user at parse time. This
+approach departs from the common standpoint that grammars and
+syntax definitions of languages must be unambiguous.
+%
+As part of the new concept, we have formalized the fundamental
+\emph{resolvable ambiguity problem}, divided it into static and
+dynamic parts, and provided solutions for both variants for a
+restricted class of languages. Through case studies, we have
+showed practical applicability of the approach, both for building
+new domain-specific languages and for reasoning about existing
+general-purpose languages.
+
+In future work, we will investigate weakning the restrictions
+currently needed for the static and dynamic resolvability
+algorithms to work. We also intend to develop our theory and tools
+to support handling deep precedence ambiguities and ambiguities
+based on ``longest match''. A long term goal is to be able to
+suggest changes to the grammar of a language, based on ambiguities
+found through dynamic analysis.
 

### 2019-07-11 paper/paper.tex
Deciding Related Work is finished

@@
 \item $\Gamma = Q \times Q \times \Labels^{*}$.
-\item $\delta$ is defined by (in hindsight this should probably be written as constructing a set $\delta$, i.e., instead of a function $\delta: Q \times \T' \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ we have a set $\delta \subseteq Q \times \T' \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, would probably be easier to read):
+\item $\delta$ is defined by the following equations (unspecified cases produce $\emptyset$):
   $$
@@
 Furthermore, for practical reasons, if the word is resolvably ambiguous we wish to produce a (minimal) witness for each tree.
-Similarly to Section~\ref{sec:static} we place some restrictions on $D$, though the restrictions are less severe:
-
-Additionally, while Section~\ref{sec:static} requires a language definition $D$ to not contain parentheses, this section merely requires parentheses to be balanced.
+Similarly to Section~\ref{sec:static}, we place some restrictions on $D$, though the restrictions are less severe.
+While Section~\ref{sec:static} requires a language definition $D$ to not contain parentheses, this section merely requires parentheses to be balanced.
+Concretely, the restrictions are as follows:
 
 \begin{enumerate}
-\item Each right-hand side regular expression must only recognize words with balanced parentheses, e.g., ``$\verb|(| \verb|)|$'' is permissible, but ``$\verb|(|^{*} \verb|)|^{*}$'' is not.
+\item Each right-hand side regular expression must only recognize words with balanced parentheses. For example, ``$\verb|(| \verb|)|$'' is permissible, but ``$\verb|(|^{*} \verb|)|^{*}$'' is not.
 \item $G_D$, but with parentheses removed, must not be infinitely ambiguous.
@@
 
-Our approach centers around around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
+Our approach centers around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
 Given restriction (1), each such language turns out to be a visibly pushdown language.
@@
 
-Intuitively we then need to, for each tree $t \in \parse(w)$, determine if $\words(t) \setminus \bigcup_{t' \in L(T_D) \land t' \neq t} \words(t)$ is the empty set.
+Intuitively, we then need to, for each tree $t \in \parse(w)$, determine if $\words(t) \setminus \bigcup_{t' \in L(T_D) \land t' \neq t} \words(t)$ is the empty set.
 Unfortunately, $L(T_D)$ tends to be infinite.
@@
 With that, there are two remaining complications: a subtree $l(w)$ recognized from a marked non-terminal $N_m$ where $l \in m$ must have one or more surrounding parentheses instead of zero or more, and we must maintain the visibly pushdown property, i.e., all transitions labelled ``$($'' or ``$)$'' must push or pop stack symbols, respectively.
-The former can be solved by either tracking the subtree's origin when parsing, or reparsing with the right hand side regular expression, while the latter is solved by introducing a new stack symbol $\gamma$ that is pushed and popped by all non-grouping parenthesis terminals.
+The former can be solved by either tracking the subtree's origin when parsing, or reparsing with the right hand side regular expression, while the latter is solved by introducing a new stack symbol $\gamma$ that is pushed and popped by all non-grouping-parenthesis terminals.
+
 We now present the function $f : L(T_D) -> VPDA$ that performs this construction.
@@
 
-The syntax definition DSL builds on the ``syncons'' introduced by
+The syntax definition DSL builds on \emph{syncons}, introduced by
 \citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
@@
 statically resolved without changing the original definitions.
-Other convenience constructs include defining prefix, infix and
+Other convenience constructs include defining prefix, infix, and
 postfix operators with a given associativity.
@@
 semi-colons wrapped in square braces. Finally, addition and
-multiplication is defined as infix syncons with the expected
+multiplication are defined as infix syncons with the expected
 precedence rules. Note that we could just as well have replaced
@@
 between these and the Orc combinators will be undefined. With
-support for resolvable ambiguity however, we can allow this
+support for resolvable ambiguity, however, we can allow this
 ambiguity and let programmers disambiguate expressions using
@@
 %
-In this case, it is likely that the preferred semantics is to have
+In this case, it is likely that the preferred semantics are to have
 all operators in the base language bind tighter than the Orc
@@
 combinator with left and right children being an integer and a
-function call respectively. In this case we have at least three
+function call, respectively. In this case we have at least three
 choices to make as language designers:
@@
 
-UNFINISHED
+% UNFINISHED
 

### 2019-07-11 paper/paper.tex
Minor fixes in section 8.

--- 2019-07-11 paper/paper.tex
@@
 \item $\Gamma = Q \times Q \times \Labels^{*}$.
-\item $\delta$ is defined by the following equations (unspecified cases produce $\emptyset$):
+\item $\delta$ is defined by (in hindsight this should probably be written as constructing a set $\delta$, i.e., instead of a function $\delta: Q \times \T' \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ we have a set $\delta \subseteq Q \times \T' \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, would probably be easier to read):
   $$
@@
 
-% UNFINISHED
+UNFINISHED
 

### 2019-07-11 paper/paper.tex
Added missing reference

@@
 
-Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
+Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{vanwykCopper2007} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
 For this paper, only the syntactic component \cite{schwerdfegerVerifiableCompositionDeterministic2009} of this analysis is relevant.

### 2019-07-11 paper/paper.tex
Weaken abstract

@@
 resolvability problems. We provide solutions to the static problem
-for two restricted classes of languages and prove them sound and,
-depending on the restrictions, complete. We also provide a sound
+for two restricted classes of languages and sketch proofs of soundness and,
+depending on the restrictions, completeness. We also provide a sound
 and complete solution to the dynamic problem for a much less

### 2019-07-11 paper/paper.tex
Pass over Section 9

@@
 \citet{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
-However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case discussed in Section~\ref{sec:evaluation-ocaml}.
+However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case involving deep precedence discussed in Section~\ref{sec:evaluation-ocaml}.
 This approach thus suggests an interesting direction for future work: to integrate it into a system designed around resolvable ambiguity, and extending the algorithms presented in this paper to cover it.
@@
 For this paper, only the syntactic component \cite{schwerdfegerVerifiableCompositionDeterministic2009} of this analysis is relevant.
-This analysis guarantees that the composition of a base language and any number of extensions that have passed it will compose to a grammar in LALR(1).
+This analysis guarantees that the composition of a base language and any number of extensions that have passed the analysis will compose to a grammar in LALR(1).
 This is a far more restricted class of languages than unambiguous languages, not to mention resolvably ambiguous languages, and thus places far more restrictions on language designers than our approach.
@@
 The detection of classical ambiguity in context-free grammars is undecidable in general \cite{cantorAmbiguityProblemBackus1962}, yet numerous heuristic approaches exist.
-Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using sat-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of \citet{bastenAmbiguityDetectionProgramming2011}.
-
-Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language \cite{heeringSyntaxDefinitionFormalism1989}.
+Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using SAT-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of \citet{bastenAmbiguityDetectionProgramming2011}.
+
+Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989}).
 These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but also assume that the composed language is unambiguous, which in practice precludes the composition of languages constructed independently of each other.

### 2019-07-11 paper/paper.tex
Fixing tempus

@@
 \section{Conclusion}
-In this paper, we have introduced the concept of \emph{resolvable
+In this paper, we introduce the concept of \emph{resolvable
   ambiguity}. A language grammar is resolvably ambiguous if all
@@
 %
-As part of the new concept, we have formalized the fundamental
-\emph{resolvable ambiguity problem}, divided it into static and
-dynamic parts, and provided solutions for both variants for a
-restricted class of languages. Through case studies, we have
-showed practical applicability of the approach, both for building
+As part of the new concept, we formalize the fundamental
+\emph{resolvable ambiguity problem}, divide it into static and
+dynamic parts, and provide solutions for both variants for a
+restricted class of languages. Through case studies, we
+showe practical applicability of the approach, both for building
 new domain-specific languages and for reasoning about existing

### 2019-07-11 paper/paper.tex
Speling

@@
 dynamic parts, and provide solutions for both variants for a
-restricted class of languages. Through case studies, we
-showe practical applicability of the approach, both for building
-new domain-specific languages and for reasoning about existing
+restricted class of languages. Through case studies, we show
+practical applicability of the approach, both for building new
+domain-specific languages and for reasoning about existing
 general-purpose languages.

### 2019-07-11 paper/paper.tex
added contrib

--- 2019-07-11 paper/paper.tex
@@
   (Section~\ref{sec:resolvable-definition}).
+\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF), where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
 \item We consider versions of the \emph{static resolvability
@@
 
-\noindent As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF), where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
-Before presenting the above contributions, the paper starts with motivating examples (Section~\ref{sec:motivation}), as well as preliminaries (Section~\ref{sec:prel}).
+\noindent Before presenting the above contributions, the paper starts with motivating examples (Section~\ref{sec:motivation}), as well as preliminaries (Section~\ref{sec:prel}).
 

### 2019-07-11 syncon-parser/src/ErrorMessage.hs
Display detail range info in error message
Fix #13

@@ -64,7 +64,8 @@ formatErrors sources messages = sortedMessages <&> formatSingle & Text.intercala
     formatRange (r, "") = "\n" <> formatPath r <> extractRange r
     formatRange (r, t) = "\n" <> ensureLn t <> formatPath r <> extractRange r
 
-    formatPath (Range f _ _) = " " <> f <> ":\n"
+    formatPath (Range f Position{line=l1,column=c1} Position{line=l2,column=c2}) =
+      " " <> f <> " " <> show l1 <> ":" <> show c1 <> "-" <> show l2 <> ":" <> show c2 <> "\n"
     formatPath Nowhere = ""
 
     extractRange :: Range -> Text

### 2019-07-11 syncon-parser/src/P4Parsing/Parser.hs
Expected end of file will now be printed when there are no expected tokens.
Fix #15

@@ -41,8 +41,9 @@ data Error l
 instance Show l => FormatError (Error l) where
   formatError _ (LexingError e) = formatError () e
   formatError _ (UnexpectedToken expected tok) = simpleErrorMessage (range tok) $
-    "Unexpected token " <> textualToken tok <> ", expected one of:\n"
-    <> (toList expected & sort & foldMap (<> "\n"))
+    "Unexpected token " <> textualToken tok <> ", expected " <> if S.null expected
+    then "end of file.\n"
+    else "one of:\n" <> (toList expected & sort & foldMap (<> "\n"))
   formatError _ (UnexpectedEOF expected) = simpleErrorMessage mempty $
     "Unexpected end of file, expected one of:\n"
     <> (toList expected & sort & foldMap (<> "\n"))

### 2019-07-11 syncon-parser/src/ErrorMessage.hs
Remove end information in error range

@@ -64,8 +64,8 @@ formatErrors sources messages = sortedMessages <&> formatSingle & Text.intercala
     formatRange (r, "") = "\n" <> formatPath r <> extractRange r
     formatRange (r, t) = "\n" <> ensureLn t <> formatPath r <> extractRange r
 
-    formatPath (Range f Position{line=l1,column=c1} Position{line=l2,column=c2}) =
-      " " <> f <> " " <> show l1 <> ":" <> show c1 <> "-" <> show l2 <> ":" <> show c2 <> "\n"
+    formatPath (Range f Position{line=l,column=c} _) =
+      " " <> f <> "#" <> show l <> ":" <> show c <> ":\n"
     formatPath Nowhere = ""
 
     extractRange :: Range -> Text

### 2019-09-17 paper/rebuttal.txt
Initial version of rebuttal

The beginning of this file contains my notes on all the concerns raised by the reviewers. At the end, after "================", is the actual rebuttal.

# Review A (knowledgeable)

line 209: done
line 301: done, but check that the same isn't done elsewher, and that it's clear enough to read without different font or quoting
line 912: slightly simplified, if a language doesn't *need* forbids to be resolvably ambiguous, then we can definitely say resolvably ambiguous. If forbids are needed we cannot answer "resolvably ambiguous" with certainty. The "unresolvably ambiguous" answer is slightly less clear: we can answer this way with certainty if we only need to examine two trees, i.e., if \words(t_1) \subseteq \words(t_2).

Why is operator precedence often mentioned but not really part of any formal constraints, rather assumed as additional meta-level constraints?
- The intent was that forbids would be a strictly more general construct than precedence, whereby the formalities would use forbids, while users would most frequently interact with precedence, which is a more commonly understood thing. The need for deep precedence puts something of a damper on this however.

Why is Theorem 6.1 a theorem, isn't the proof trivial?
- It is indeed, but we figured it'd be better to be overly rigorous where it was easy to do so.

The classical example of if-then-else statement ambiguity is not mentioned in the paper, but it seems to fall into the same category, does it?
- It does indeed. We decided against it since the solution is widely known and accepted, though it does seem a bit dumb to not include right now...

# Review B (no familiarity)
Grammar/spelling issues addressed.
"Overall, nicely written paper and a very novel approach to a rather old problem. My issue is with regards to the practicality of this approach."

# Review C (expert)
- the results/algorithms cover a very limited class of grammars
- it is hard to see whether the general framework proposed in sec 4 is needed
  For this particular language formalism, no, but the general idea could be used for other formalisms with other forms of disambiguation, e.g., semicolons and the like.
- missing a comparison of alternative words and nested words
  TODO: consider this more carefully. You could consider a pair of parens as a nesting level in a nested word.
- the overall approach is not very convincing: It seems that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user, which doesn't make sense to me.
  The idea is that the end-user already runs into these ambiguity issues, it's just that the compiler doesnt' care and picks a semi-arbitrary interpretation. The user's understanding of the grammar is likely ambiguous, though they might not realize it, but the compiler doesn't know this, leading the user to debug ambiguity errors through type errors or runtime errors. E.g. `1 & 3 == 1`, `match` in OCaml, if-then-else ambiguity.

line 70: done
line 94: I assume this means "set both operators at the same relative precedence, and arbitrarily pick left- or right-associativity" (since each operation is likely to be associative, in the mathematical sense, it shouldn't matter which associativity is chosen). The difference is that our approach makes the compiler highlight to the user that there are multiple distinct alternatives, all of which might be what the user intends.
line 144: 2/3 reviewers have now said we should discuss "dangling else", so we probably should. Maybe state that the two problems are essentially the same, but `match` is more likely to become an issue to the user simply by how the languages are used?
line 161: "would it make sense to just speak of derivation trees? The linear notion of derivation doesn't matter at all for this paper." It would indeed.
line 190,192: "why mix \varepsilon and \lambda?" We were following the definition in a particular book, but yes, this should probably just be \epsilon.
line 197: "AFAIK this restricted form is equivalent to the general form. Unless you restrict to deterministic systems, of course." We never use $a = \epsilon$, otherwise true.
line 222: agreed
line 268: added a sentence
line 271: the central required notion is the connection between words and trees, the latter is not present in the classical definition of a language (a set of words).
line 314: true, and done
line 351, 358: done
line 371: this likely means a slight reordering of things, basically a sentence or two, TODO
line 531: "version 3 apparently, nothing can be said about version 3. Version 1 (where soundness and completeness holds) seems a rather draconic and unrealistic restriction. But the authors do not comment on this pragmatic issue." This is essentially just straight up true. We should probably address this. Essentially: the static problem turns out to be harder to solve than one might think, and this is a first attempt at it.
line 551: the alternative word formulation additionally permits multiple nesting edges to share a position, e.g., ((1 + 2)) + 3 corresponds to 1 + 2 + 3 with two nesting levels around 1 + 2, which is not permitted by nesting words.
line 582: intuitively it seems trivial, but actually mechanically proving it is not, I believe (slight weakness here since we haven't actually done it). Quite happy / impressed that this is "obvious". Additional note: it's not just double parens, but also optional parens.
line 932: this could certainly use something at the end that just ties it together. I believe the reasoning has been presented earlier, and this is "just" filling in the blanks, but the conclusion should be present at the end. TODO: look more carefully at this
line 1001: The property is the possibility of constructing this set in finite time, i.e., that it's decidable.
line 1005: True, it should be "[...] the set of trees $t' \in L(T_D)$ such that $\words(t') \cap \bigcup_{i} \words(t_i) \neq \emptyset$ is finite". We should also mention that we use 7.3 to discover the full set of relevant trees, that's its purpose.

sec 8:
It would be good to discuss whether the results achieved in the case study are artifacts of the particular approach to defining the syntax.
- I interpret this to be a request to actually discuss the differences between the formalization version and the syncon version of the specification language, instead of just mentioning it and treating it as a (the) reasonable implementation of it.

--------------------------------

Reviewer A
- can you be more precise when either case can be decided with certainty? (static)

Reviewer C explicit questions:
- what's the relation between alternative words and nested words?
- is there hope for any results for version 3 grammars?

Reviewer C
- very limited class of grammars
- the overall approach is not very convincing: It seems that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user, which doesn't make sense to me.

Reviewer A + C
- if-then-else should be there and/or it's a more well-known version of the match example

===================================================

We thank the reviewers for their thoughtful comments!

# Overall motivation / Reviewer C on the overall approach
Reviewer C comments that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user. We argue that the user *already* has to deal with ambiguity issues in common general-purpose languages, but without compiler assistance. The programmer's understanding of the syntax of the language is likely ambiguous, possibly without them being aware, e.g., the match example in OCaml or infix operators rarely used together (try predicting the result of evaluating `1 & 3 == 1` in, e.g., C and Python). Resolvable ambiguity allows the compiler to assist in such cases by using an ambiguous grammar.

That said, giving the compiler this capability does not absolve the language designer of the responsibility of making sure their language is convenient to use, it merely provides an alternative way to deal with this sort of edge cases.

# Reviewers A and C, dangling else
Reviewer A notes the absence of dangling else, and Reviewer C further notes that nested match is essentially the same issue, only the former is more well-known. We decided to use the match example instead since, anecdotally, nested matches in OCaml occur more often than nested if-statements without braces and only a single else, and is thus more likely to become an issue. We will add a comparison to dangling else.

# Reviewer A, results of the static analysis
Reviewer A asks when the static analysis can give definitive answers. For the static analysis we can achieve a positive result (resolvably ambiguous) for a language that does not need forbids (essentially precedence and associativity) to be resolvable. For example, OCaml does not fit here, since [1; 2] needs to forbid sequential composition as a direct child of a list to be resolvable, while arithmetic expressions do fit, since precedence and associativity are merely conveniences, all possible expressions can be written merely by putting grouping parentheses around all subexpressions.

The negative result is less intuitive, but it arises when some tree can never be disambiguated from some particular other tree, i.e., $\exists t_1, t_2.\ \forall w.\ t_1 \in \parse(w) \rightarrow t_2 \in \parse(w)$.

# Reviewer C, limited class of grammars and hope for version 3
Reviewer C notes that the class of grammars supported by the static analysis is quite limited, which is indeed true (though slightly larger than is immediately obvious, see the previous answer). The contribution here is largely the concept, and the introduction of a problem that is more difficult than it first appears, and some initial steps towards solving it. We thus do not yet know if the problem is decidable for version 3.

The dynamic analysis on the other hand is only limited to languages with balanced parentheses.

# Reviewer B, practicality
Reviewer B wonders about the practicality of this approach. The OCaml case-study contributes the most to this, in that we can parse 75% of the `.ml` files in the OCaml compiler (including test cases, i.e., many uncommonly used language features). As for efficiency and similar concerns, we leave those for future work.

# Reviewer C, nested words and our alternative word representation
Reviewer C asks for the relation between nested words and our alternative word representation. They are quite similar, but nested words disallow nesting edges that share a left and / or right point. This makes them unsuitable for our usecase, since that would disallow double parentheses and parentheses of the form "(a(b))". We will add a comparison to the paper.

### 2019-09-18 paper/rebuttal.txt
Minor punctuation and markdown fixes

@@
 
-# Overall motivation / Reviewer C on the overall approach
+#### Overall motivation / Reviewer C on the overall approach
 Reviewer C comments that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user.
@@
 
-That said, giving the compiler this capability does not absolve the language designer of the responsibility of making sure their language is convenient to use, it merely provides an alternative way to deal with this sort of edge cases.
+That said, giving the compiler this capability does not absolve the language designer of the responsibility of making sure their language is convenient to use $-$ it merely provides an alternative way to deal with this sort of edge cases.
 
-# Reviewers A and C, dangling else
+#### Reviewers A and C, dangling else
 Reviewer A notes the absence of dangling else, and Reviewer C further notes that nested match is essentially the same issue, only the former is more well-known.
@@
 
-# Reviewer A, results of the static analysis
+#### Reviewer A, results of the static analysis
 Reviewer A asks when the static analysis can give definitive answers.
 For the static analysis we can achieve a positive result (resolvably ambiguous) for a language that does not need forbids (essentially precedence and associativity) to be resolvable.
-For example, OCaml does not fit here, since [1; 2] needs to forbid sequential composition as a direct child of a list to be resolvable, while arithmetic expressions do fit, since precedence and associativity are merely conveniences, all possible expressions can be written merely by putting grouping parentheses around all subexpressions.
+For example, OCaml does not fit here, since [1; 2] needs to forbid sequential composition as a direct child of a list to be resolvable, while arithmetic expressions do fit, since precedence and associativity are merely conveniences $-$ all possible expressions can be written merely by putting grouping parentheses around all subexpressions.
 
-The negative result is less intuitive, but it arises when some tree can never be disambiguated from some particular other tree, i.e., $\exists t_1, t_2.\ \forall w.\ t_1 \in \parse(w) \rightarrow t_2 \in \parse(w)$.
+The negative result is less intuitive, but it arises when some tree can never be disambiguated from some particular other tree, i.e., $\exists t_1, t_2.\ \forall w.\ t_1 \in \mathit{parse}(w) \Rightarrow t_2 \in \mathit{parse}(w)$.
 
-# Reviewer C, limited class of grammars and hope for version 3
+#### Reviewer C, limited class of grammars and hope for version 3
 Reviewer C notes that the class of grammars supported by the static analysis is quite limited, which is indeed true (though slightly larger than is immediately obvious, see the previous answer).
@@
 
-# Reviewer B, practicality
+#### Reviewer B, practicality
 Reviewer B wonders about the practicality of this approach.
@@
 
-# Reviewer C, nested words and our alternative word representation
+#### Reviewer C, nested words and our alternative word representation
 Reviewer C asks for the relation between nested words and our alternative word representation.
-They are quite similar, but nested words disallow nesting edges that share a left and / or right point.
-This makes them unsuitable for our usecase, since that would disallow double parentheses and parentheses of the form "(a(b))".
+They are quite similar, but nested words disallow nesting edges that share a left and/or right point.
+This makes them unsuitable for our usecase, since that would disallow double parentheses and parentheses of the form `(a(b))`.
 We will add a comparison to the paper.

### 2019-09-19 syncon-parser/earley-forest/src/Text/Earley/Forest/Parser.hs
Optimization pass

@@ -33,7 +33,7 @@ import qualified Data.HashMap.Lazy as M
 import qualified Data.Vector as Vector
 
 import Text.Earley.Forest.Grammar (TokKind, Parseable, getKind)
-import Text.Earley.Forest.Transformations (EpsNT(..), NT(..), Rule(..), Sym(..), NullStatus(..), NtKind(..))
+import Text.Earley.Forest.Transformations (EpsNT(..), NT(..), Rule(..), Sym(..), NullStatus(..), NtKind(..), rightRecursive)
 import Text.Earley.Forest.SplitEpsDFA (EpsDFA(..), mkDFA, renumberStates, completedNT, isCompleted, DotProd(..))
 
 -- | Several Earley items, in the form of an EpsDFA state and origin
@@ -128,6 +128,9 @@ parse nnf@(s1, s2, rules) = case s1 <|> s2 of
         single _ = Nothing
         haveSuccessors = M.filter (M.null >>> not) transitions
 
+    limInteresting :: HashSet NT
+    limInteresting = rightRecursive rules
+
     -- | The transition (if any) from a given DFA state with an epsilon label.
     epsMap :: HashMap Int Int
     epsMap = M.lookup Nothing `M.mapMaybe` transitions
@@ -173,7 +176,7 @@ parse nnf@(s1, s2, rules) = case s1 <|> s2 of
               let completions = toList completed
                     & mapMaybe (\nt -> M.lookup nt posts <&> (nt,))
               forM_ completions $ \case
-                (nt, Right eims) -> forM_ eims $ mkLink nt  >>> stepAndAddEim set
+                (nt, Right eims) -> forM_ eims $ mkLink nt >>> stepAndAddEim set
                 (nt, Left (_, Lim predIdx eim' nt')) -> stepAndAddEim set LimCause{pred = (predIdx, eim'), limCause = (nt, eim), causeNt = nt'}
             -- Compute postdot map
             modifySTRef' currPostDot $ M.unionWith S.union $ computePostDot eim
@@ -235,23 +238,24 @@ parse nnf@(s1, s2, rules) = case s1 <|> s2 of
         -- most importantly, compute the postDot map, including 'Lim's.
         mkInactiveSet :: ActiveSet s -> ParseM s InactiveSet
         mkInactiveSet ActiveSet{currPostDot, currIdx, currLinks} = do
-          postDots <- readSTRef currPostDot <&> fmap splitSingles
+          postDots <- readSTRef currPostDot
[…]

### 2019-09-19 syncon-parser/earley-forest/src/Text/Earley/Forest/SplitEpsDFA.hs

@@ -124,6 +124,14 @@ mkDFA (s1, s2, rules) = case s1 <|> s2 of
           | S.null new = prev
           | otherwise = go prev' $ (`S.difference` prev') $ foldMap f new
           where prev' = prev <> new
+    -- iterateInductively :: (Eq a, Hashable a) => (a -> HashSet a) -> HashSet a -> HashSet a
+    -- iterateInductively f start = go start $ (`S.difference` start) $ foldMap f start
+    --   where
+    --     go prev new
+    --       | S.null new = prev
+    --       | otherwise = go (S.union prev new) new'
+    --       where
+    --         new' = foldMap (f >>> (`S.difference` prev)) new
 
 completedNT :: Seq (Rule EpsNT (Sym EpsNT t) nodeF) -> DotProd -> Maybe NT
 completedNT rules (DotProd ruleIdx dotIdx) = Seq.index rules ruleIdx

### 2019-09-19 syncon-parser/earley-forest/src/Text/Earley/Forest/Transformations.hs

@@ -9,6 +9,7 @@ module Text.Earley.Forest.Transformations
 , NT(..)
 , NtKind(..)
 , EpsNT(..)
+, rightRecursive
 ) where
 
 import Prelude
@@ -22,7 +23,8 @@ import Control.Monad.Trans.Reader (ReaderT, runReaderT, ask, asks)
 import Control.Monad.ST (runST, ST)
 import Control.Monad.Trans.Class (lift)
 import Data.Bifunctor (first)
-import Data.Foldable (forM_, toList)
+import Data.Maybe (mapMaybe)
+import Data.Foldable (forM_, toList, fold)
 import Data.Function ((&))
 import Data.Functor ((<&>), void)
 import Data.HashSet (HashSet)
@@ -30,6 +32,7 @@ import Data.Hashable (Hashable)
 import Data.List (partition)
 import Data.STRef (STRef, newSTRef, readSTRef, modifySTRef')
 import Data.Sequence (Seq, (<|), (|>))
+import qualified Data.HashMap.Strict as M
 import qualified Data.HashSet as S
 import qualified Data.Sequence as Seq
 
@@ -214,6 +217,27 @@ nullableSet nullStatus = toList
       where
         (nullProds, nextProds) = partition (snd >>> isInteresting prev) prods
         next = nullProds <&> fst & S.fromList & S.union prev
+
+-- | Find the right-recursive non-terminals in the given NNF grammar
+rightRecursive :: Seq (Rule EpsNT (Sym EpsNT t) nodeF) -> HashSet NT
+rightRecursive = toList
+  >>> mapMaybe rightNt
+  >>> M.fromListWith S.union
+  >>> closeTrans
+  >>> M.mapMaybeWithKey isRecursive
+  >>> S.fromMap
+  where
+    rightNt (Rule (EpsNT nt NonNullable) syms _) = do
+      Nt (EpsNT nt' _) <- Seq.findIndexR isNonNullable syms <&> Seq.index syms
+      return (nt, S.singleton nt')
+    rightNt _ = Nothing
+    isNonNullable (Nt (EpsNT _ NonNullable)) = True
+    isNonNullable (Sym _) = True
+    isNonNullable _ = False
+    isRecursive nt nts
+      | nt `S.member` nts = Just ()
+      | otherwise = Nothing
+    closeTrans prev = M.map (\here -> prev `M.intersection` S.toMap here & fold & S.union here) prev
 
 toAny :: a -> Any
 toAny = unsafeCoerce

### 2019-09-19 paper/rebuttal.txt
Rebuttal with extra material and more final formatting

@@
 
+C sec 8:
+It would be good to discuss whether the results achieved in the case study are artifacts of the particular approach to defining the syntax.
+- I interpret this to be a request to actually discuss the differences between the formalization version and the syncon version of the specification language, instead of just mentioning it and treating it as a (the) reasonable implementation of it.
+
+--------------------------------
+
 Reviewer A
@@
 
+---------------------------------
+
 ===================================================
@@
 We thank the reviewers for their thoughtful comments!
+We initially address the larger/more common concerns, then smaller issues in an optional section after the line.
+Issues not mentioned we agree with, and will address.
 
-#### Overall motivation / Reviewer C on the overall approach
-Reviewer C comments that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user.
+**Reviewer C** comments that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user.
 We argue that the user *already* has to deal with ambiguity issues in common general-purpose languages, but without compiler assistance.
@@
 
-#### Reviewers A and C, dangling else
-Reviewer A notes the absence of dangling else, and Reviewer C further notes that nested match is essentially the same issue, only the former is more well-known.
+**Reviewer A** notes the absence of dangling else, and **Reviewer C** further notes that nested match is essentially the same issue, only the former is more well-known.
 We decided to use the match example instead since, anecdotally, nested matches in OCaml occur more often than nested if-statements without braces and only a single else, and is thus more likely to become an issue.
@@
 
-#### Reviewer A, results of the static analysis
-Reviewer A asks when the static analysis can give definitive answers.
+**Reviewer A** asks when the static analysis can give definitive answers.
 For the static analysis we can achieve a positive result (resolvably ambiguous) for a language that does not need forbids (essentially precedence and associativity) to be resolvable.
@@
 
-#### Reviewer C, limited class of grammars and hope for version 3
-Reviewer C notes that the class of grammars supported by the static analysis is quite limited, which is indeed true (though slightly larger than is immediately obvious, see the previous answer).
-The contribution here is largely the concept, and the introduction of a problem that is more difficult than it first appears, and some initial steps towards solving it.
-We thus do not yet know if the problem is decidable for version 3.
+**Reviewer C** notes that the class of grammars supported by the static analysis is quite limited, which is indeed true (though slightly larger than is immediately obvious, see the previous answer).
+The contribution here is largely the concept, and the introduction of a problem that is more difficult than it first appears.
+We thus do not yet know if the problem is decidable for version 3, and leave that for future research.
+However, we will add more discussion of these restrictions.
 
@@
 
-#### Reviewer B, practicality
-Reviewer B wonders about the practicality of this approach.
+**Reviewer B** wonders about the practicality of this approach.
 The OCaml case-study contributes the most to this, in that we can parse 75% of the `.ml` files in the OCaml compiler (including test cases, i.e., many uncommonly used language features).
@@
 
-#### Reviewer C, nested words and our alternative word representation
-Reviewer C asks for the relation between nested words and our alternative word representation.
+**Reviewer C** asks for the relation between nested words and our alternative word representation.
 They are quite similar, but nested words disallow nesting edges that share a left and/or right point.
@@
 We will add a comparison to the paper.
+
+------------------------------------------------------------
+
+#### Reviewer A
+
+> Why is operator precedence often mentioned but not really part of any formal constraints, rather assumed as additional meta-level constraints?
+
+The intent was that forbids would be a strictly more general construct than precedence, whereby the formalities would use forbids, while users would most frequently interact with precedence, seeing as it is a more generally understood concept.
+The need for deep precedence, as discussed in the evaluation, suggests a flaw in this however.
+
+#### Reviewer C
+
+> It is hard to see whether the general framework proposed in sec 4 is needed
+
+For this particular language formalism, no, but the general idea can be used for other formalisms with other forms of disambiguation, e.g., optional semicolons and the like.
+
+> line 94: A different approach would be to design the grammar so that the infix operators have associativity.
+That would force the user to put parentheses and avoid the problem.
+I don't quite see why this is more awkward than what the authors propose.
+
+We assume this to mean "set both operators at the same relative precedence, and arbitrarily pick left- or right-associativity" (since each operation is likely to be associative, in the mathematical sense, it shouldn't matter which associativity is chosen).
+The difference is that our approach makes the compiler highlight to the user that there are multiple distinct alternatives, all of which might be what the user intends, instead of just picking one that the user might not expect.
+
+> line 271: *parse languages* It seems to me that any decidable language can be a parse language, so there's not much point in defining that concept.
+
+The central notion that we require is the connection between words and trees, the latter is not present in the classical definition of a language (a set of words).
+
+> line 582 Can you explain why lemma 6.3 is nontrivial? it obviously corresponds to removing double parentheses.
+
+It corresponds to removing both double and optional parentheses, leaving only those required by forbids, making it slightly less obvious.
+
+> line 932 theorem 6.15. why is this a decidable property and how does it yield the proofs of soundness and completeness of version 1? what about version 2?
+
+The property is decidable since the construction of $A_{()}$ and $A'_{()}$, and their trimmed product, is decidable.
+The procedure discovers unresolvable ambiguities where the lattice of one tree is entirely contained in the lattice of another tree.
+It is thus sound, since it always produces a concrete tree that has no unambiguous word.
+In version 1 there are no other forms of unresolvable ambiguities (Lemma 6.9, line 683-685), thus it is there complete.
+We will clarify the surrounding text to make these connections more clear.
+
+> line 1001 lemma 7.2. the property that is to be proved seems to be missing.
+
+The property is the possibility of constructing this set in finite time, i.e., that it is decidable.
+
+> line 1005 lemma 7.3. the trees $t_1, \dots, t_n$ are not connected to the statement.
+
+Indeed, it should be "[...] the set of trees $t' \in L(T_D)$ such that $\mathit{words}(t') \cap \bigcup_i \mathit{words}(t_i) \neq \emptyset$ is finite".

### 2019-09-19 paper/rebuttal.txt

@@
 
-We thank the reviewers for their thoughtful comments!
-We initially address the larger/more common concerns, then smaller issues in an optional section after the line.
-Issues not mentioned we agree with, and will address.
+We thank the reviewers for their thoughtful comments! We start by addressing major concerns and leave minor comments for the end of the rebuttal.
 
-**Reviewer C** comments that the overall approach shifts the burden of thinking about ambiguity arising from language composition from the language designer to the user.
+**Reviewer C** comments that the overall approach shifts the burden of thinking about ambiguity from the language designer to the user.
 We argue that the user *already* has to deal with ambiguity issues in common general-purpose languages, but without compiler assistance.
-The programmer's understanding of the syntax of the language is likely ambiguous, possibly without them being aware, e.g., the match example in OCaml or infix operators rarely used together (try predicting the result of evaluating `1 & 3 == 1` in, e.g., C and Python).
+The programmer's understanding of syntax is likely (unknowingly) ambiguous, e.g., the match example in OCaml or combinations of infix operators (what is the result of  `1 & 3 == 1` in, e.g., C and Python?).
 Resolvable ambiguity allows the compiler to assist in such cases by using an ambiguous grammar.
 
-That said, giving the compiler this capability does not absolve the language designer of the responsibility of making sure their language is convenient to use $-$ it merely provides an alternative way to deal with this sort of edge cases.
+However, enhancing the compiler does not absolve the language designer of the responsibility of making sure their language is convenient to use $-$ it merely provides an alternative way to deal with this sort of edge cases.
 
-**Reviewer A** notes the absence of dangling else, and **Reviewer C** further notes that nested match is essentially the same issue, only the former is more well-known.
-We decided to use the match example instead since, anecdotally, nested matches in OCaml occur more often than nested if-statements without braces and only a single else, and is thus more likely to become an issue.
+**Reviewers A and B** note the connection between nested matches and dangling else.
+We use the match example since, anecdotally, nested matches in OCaml occur more often than nested if-statements without braces and only a single `else`, and is thus more likely to become an issue.
 We will add a comparison to dangling else.
@@
 **Reviewer A** asks when the static analysis can give definitive answers.
-For the static analysis we can achieve a positive result (resolvably ambiguous) for a language that does not need forbids (essentially precedence and associativity) to be resolvable.
-For example, OCaml does not fit here, since [1; 2] needs to forbid sequential composition as a direct child of a list to be resolvable, while arithmetic expressions do fit, since precedence and associativity are merely conveniences $-$ all possible expressions can be written merely by putting grouping parentheses around all subexpressions.
+We can achieve a positive result (resolvably ambiguous) for a language that does not need forbids (precedence and associativity) to be resolvable.
+For example, OCaml does not fit here, since `[1; 2]` needs to forbid sequential composition as a direct child of a list to be resolvable, while arithmetic expressions do fit, since precedence and associativity are merely conveniences $-$ all possible expressions can disambiguated using parentheses.
 
-The negative result is less intuitive, but it arises when some tree can never be disambiguated from some particular other tree, i.e., $\exists t_1, t_2.\ \forall w.\ t_1 \in \mathit{parse}(w) \Rightarrow t_2 \in \mathit{parse}(w)$.
+The negative result is less intuitive, but arises when some tree can never be disambiguated from another particular tree, i.e., $\exists t_1, t_2.\ \forall w.\ t_1 \in \mathit{parse}(w) \Rightarrow t_2 \in \mathit{parse}(w)$.
 
-**Reviewer C** notes that the class of grammars supported by the static analysis is quite limited, which is indeed true (though slightly larger than is immediately obvious, see the previous answer).
-The contribution here is largely the concept, and the introduction of a problem that is more difficult than it first appears.
-We thus do not yet know if the problem is decidable for version 3, and leave that for future research.
-However, we will add more discussion of these restrictions.
+**Reviewer C** notes that the class of grammars supported by the static analysis is quite limited, which is true (though slightly larger than is immediately obvious, see above).
+Our contribution is largely the concept, and the introduction of the resolvable ambiguity problem.
+We do not yet know if the problem is decidable for version 3, and leave that for future research.
+We will add more discussion of these restrictions.
 
-The dynamic analysis on the other hand is only limited to languages with balanced parentheses.
+The dynamic analysis, however, is only limited to languages with balanced parentheses.
 
 **Reviewer B** wonders about the practicality of this approach.
-The OCaml case-study contributes the most to this, in that we can parse 75% of the `.ml` files in the OCaml compiler (including test cases, i.e., many uncommonly used language features).
-As for efficiency and similar concerns, we leave those for future work.
+As stated in the paper, we can parse 75% of the `.ml` files in the OCaml compiler (including test cases, i.e., many uncommonly used language features).
+We leave efficiency and similar concerns for future work.
 
 **Reviewer C** asks for the relation between nested words and our alternative word representation.
-They are quite similar, but nested words disallow nesting edges that share a left and/or right point.
+They are quite similar, but nested words disallow nesting edges that share a left or right point.
 This makes them unsuitable for our usecase, since that would disallow double parentheses and parentheses of the form `(a(b))`.

### 2019-09-19 paper/rebuttal.txt
More rebuttal edits

@@
 
-We thank the reviewers for their thoughtful comments! We start by addressing major concerns and leave minor comments for the end of the rebuttal.
+We thank the reviewers for their thoughtful comments! We start by addressing major concerns and leave other comments for the end of the rebuttal.
 
@@
 
-**Reviewers A and B** note the connection between nested matches and dangling else.
+**Reviewers A and C** note the connection between nested matches and dangling else.
 We use the match example since, anecdotally, nested matches in OCaml occur more often than nested if-statements without braces and only a single `else`, and is thus more likely to become an issue.
@@
 **Reviewer C** notes that the class of grammars supported by the static analysis is quite limited, which is true (though slightly larger than is immediately obvious, see above).
-Our contribution is largely the concept, and the introduction of the resolvable ambiguity problem.
+Our contribution is largely the concept, and the introduction of the resolvable ambiguity problem, which is more difficult than it initially appears.
 We do not yet know if the problem is decidable for version 3, and leave that for future research.
@@
 This makes them unsuitable for our usecase, since that would disallow double parentheses and parentheses of the form `(a(b))`.
-We will add a comparison to the paper.
+We will add a comparison.
 
@@
 Indeed, it should be "[...] the set of trees $t' \in L(T_D)$ such that $\mathit{words}(t') \cap \bigcup_i \mathit{words}(t_i) \neq \emptyset$ is finite".
+
+> sec 8:
+> It would be good to discuss whether the results achieved in the case study are artifacts of the particular approach to defining the syntax.
+
+Most results would be the same with some other more monolithic approach, assuming it supported ambiguity and a dynamic resolvability analysis, except of course for the compositional examples.
+The compositional examples are interesting since separately defined language fragments cannot be defined to compose unambiguously in general, thus we cannot simply forbid ambiguity.

### 2019-09-27 syncon-parser/bench/Main.hs
Remove gll

@@ -17,7 +17,6 @@ import P2LanguageDefinition.Types (TypeName(..))
 import qualified P2LanguageDefinition.Parser as LD
 import qualified P2LanguageDefinition.BasicChecker as LD
 import qualified P4Parsing.Parser as Parser
-import qualified P4Parsing.Parser2 as Parser2
 import P4Parsing.Types (SingleLanguage(..))
 
 import Criterion.Main
@@ -32,10 +31,9 @@ langToBench :: (FilePath, ([FilePath], ([FilePath], [FilePath]))) -> Benchmark
 langToBench (lang, (defFiles, uncurry mappend -> srcFiles)) = env (mkEnv lang defFiles) $ \env ->
   bgroup lang $ benchSrc env <$> srcFiles
   where
-    benchSrc ~(lexFile, parseTokens, precomputed) srcFile = env (doLexFile lexFile srcFile) $ \toks ->
+    benchSrc ~(lexFile, precomputed) srcFile = env (doLexFile lexFile srcFile) $ \toks ->
       bgroup (takeFileName srcFile)
-      [ bench "gll" $ whnf parseTokens toks
-      , bench "earley-forest" $ whnf (Parser2.parseTokens precomputed) toks ]
+      [ bench "earley-forest" $ whnf (Parser.parseTokens precomputed) toks ]
     doLexFile lexFile path = lexFile path >>= \case
       Error _ -> die $ "Could not lex file " <> toS path
       Data toks -> return toks
@@ -51,7 +49,7 @@ getLanguages languageDirectory = do
   where
     listDir path = listDirectory path <&> fmap (path </>)
 
-mkEnv lang defFiles = mkDf >>= \df -> (,,) <$> mkLexer df <*> mkGllParser df <*> mkEarleyParser df
+mkEnv lang defFiles = mkDf >>= \df -> (,) <$> mkLexer df <*> mkEarleyParser df
   where
     mkDf = do
       foldMapM LD.parseFile defFiles >>= \case
@@ -62,9 +60,6 @@ mkEnv lang defFiles = mkDf >>= \df -> (,,) <$> mkLexer df <*> mkGllParser df <*>
     mkLexer df = case Parser.dfToLanguageTokens df & Lexer.allOneLanguage SingleLanguage of
       Error _ -> die $ "Could not generate a parser for lanugage " <> toS lang
       Data lexFile -> return lexFile
-    mkGllParser df = case Parser.parseTokens @SingleLanguage @[] df of
[…]

### 2019-09-27 syncon-parser/exe/Main.hs

@@ -41,7 +41,7 @@ import qualified P2LanguageDefinition.BasicChecker as LD
 import qualified P2LanguageDefinition.Elaborator as LD
 
 import qualified P4Parsing.Types as Parser
-import qualified P4Parsing.Parser2 as Parser2
+import qualified P4Parsing.Parser as Parser
 
 import qualified P5DynamicAmbiguity.Types as DynAmb
 import qualified P5DynamicAmbiguity.TreeLanguage as DynAmb
@@ -133,10 +133,10 @@ compile = Opt.command "compile" (Opt.info compileCmd $ Opt.progDesc "Compile a l
         tops <- M.traverseWithKey (\defFile _ -> LD.parseFile $ toS defFile) sources
           >>= (fold >>> dataOrError sources ())
         df <- LD.mkDefinitionFile tops & dataOrError sources ()
-        preParse <- Parser2.precomputeSingleLanguage @(Lexer.Token Parser.SingleLanguage LD.TypeName) df & dataOrError sources ()
+        preParse <- Parser.precomputeSingleLanguage @(Lexer.Token Parser.SingleLanguage LD.TypeName) df & dataOrError sources ()
         let pl = DynAmb.precompute @() df
 
-        let serialisable = (Parser2.precomputeToSerialisable preParse, pl)
+        let serialisable = (Parser.precomputeToSerialisable preParse, pl)
         writeFileSerialise outputFile serialisable
 
 parse :: Opt.Mod Opt.CommandFields (IO ())
@@ -177,7 +177,7 @@ parse = Opt.command "parse" (Opt.info parseCmd $ Opt.progDesc "Parse a list of f
 
       pure $ do
         (preParseSerialisable, pl) <- readFileDeserialise synconc
-        preParse <- Parser2.serialisableToPrecompute @(Lexer.Token Parser.SingleLanguage LD.TypeName) preParseSerialisable & dataOrError mempty ()
+        preParse <- Parser.serialisableToPrecompute @(Lexer.Token Parser.SingleLanguage LD.TypeName) preParseSerialisable & dataOrError mempty ()
 
         sources <- files & S.fromList & S.toMap
           & M.traverseWithKey (\path _ -> readFile $ toS path)
@@ -190,12 +190,12 @@ parse = Opt.command "parse" (Opt.info parseCmd $ Opt.progDesc "Parse a list of f
           putStrLn @Text $ "Parsing \"" <> path <> "\""
[…]

### 2019-09-27 syncon-parser/gll/src/GLL/Flags.hs

@@ -1,66 +1,0 @@
-
-module GLL.Flags where
-
-import Prelude
-
--- | Flags to influence the behaviour of the parser.
-data Flags   = Flags    { symbol_nodes          :: Bool
-                        , intermediate_nodes    :: Bool
-                        , edges                 :: Bool
-                        , flexible_binarisation :: Bool
-                        , max_errors            :: Int
-                        , do_select_test        :: Bool
-                        }
-
--- | The default flags:
--- * Do not add symbol nodes to the 'SPPF'.
--- * Do not add intermediate nodes to the 'SPPF'.
--- * Do not add edges to the 'SPPF'.
--- * Flexible binarisation.
--- * The three furthest discoveries of a token mismatch are reported.
--- * Select tests are performed.
-defaultFlags :: Flags
-defaultFlags = Flags False False False True 3 True
-
--- | Execute the given 'Options' in left-to-right order on 'defaultFlags'.
-runOptions :: ParseOptions -> Flags
-runOptions = foldr ($) defaultFlags
-
--- | An option updates the current set of 'Flags'.
-type ParseOption = Flags -> Flags
-
--- | A list of 'ParserOption's
-type ParseOptions = [ParseOption]
-
--- |
--- Create the 'SPPF' with all nodes and edges, not necessarily strictly binarised.
-fullSPPF :: ParseOption
-fullSPPF flags = flags{symbol_nodes = True, intermediate_nodes = True, edges = True}
-
--- |
--- Create all nodes, but no edges between nodes.
-allNodes :: ParseOption
-allNodes flags = flags{symbol_nodes = True, intermediate_nodes = True}
-
--- |
--- Create packed-nodes only.
-packedNodesOnly :: ParseOption
-packedNodesOnly flags = flags{symbol_nodes = False, intermediate_nodes = False, edges = False}
-
--- |
--- Fully binarise the SPPF, resulting in a larger 'SPPF' and possibly slower runtimes.
--- When this flag is on, packed nodes can only have a single symbol node child
--- or one intermediate node child and one symbol node child.
--- With the flag disabled a packed node can have two symbol node children.
-strictBinarisation :: ParseOption
-strictBinarisation flags = flags{flexible_binarisation = False}
-
--- |
--- Set the maximum number of errors shown in case of an unsuccessful parse.
-maximumErrors :: Int -> ParseOption
-maximumErrors n flags = flags {max_errors = n}
-
--- |
[…]

### 2019-09-27 syncon-parser/gll/src/GLL/Parser.hs

@@ -1,577 +1,0 @@
-{-# OPTIONS_GHC -fno-warn-dodgy-exports -fno-warn-unused-top-binds -fno-warn-missing-signatures -fno-warn-unused-matches -fno-warn-orphans -fno-warn-name-shadowing -fno-warn-unused-local-binds -fno-warn-incomplete-uni-patterns #-}
-{-# LANGUAGE NamedFieldPuns #-}
-{-|
-Implementation of the GLL parsing algorithm [Scott and Johnstone 2010,2013,2016]
-with the grammar as an explicit parameter.
-
-Function 'parse' receives a 'Grammar' as input together with a
-list of tokens (the input string).
-
-The type of token is chosen arbitrarily, but the type should be 'Parseable' and 'Ord'erable.
-To be 'Parseable' a type must have two distinct values, 'eos' (end-of-string)
-and 'eps' (epsilon). The user must ensure that these two values will never occur
-as part of the input string.
-
-== GLL Parsing
-=== Recursive Descent
-GLL parsing is a generalisation of recursive descent parsing (RD parsing).
-A RD parser (RDP), for some grammar 'G' , consists of a set of parse
-functions 'f_X', one for every nonterminal 'X', and a main function which
-calls 'f_S', where 'S' is the start symbol.
-The parse function 'f_X' take an integer 'l' as an argument and produces an
-integer 'r', indicating that nonterminal 'X' derives 's_l_r',
-where 's_i_j' is the substring of the input string 's' ranging from
-'i' to 'j'. We call 'l' and 'r' the right- and left-extent, respectively.
-
-The parse function 'f_X'
-has a branch for every production X ::= s_1 ... s_k in 'G', guarded
-by a look-ahead test, and every
-branch has 'k' code fragments, one for every symbol 's_i',
-with 1 <= i <= k.
-A RDP matches grammar positions, represented by /slots/ of the form
-X ::= a . b,  with (input) string positions.
-The dot in a slot tells us how much of the production's symbols have been
-matched (the symbols before the dot) and which symbols still need to
-be matched (the symbols after the dot). The symbol immediately after the dot
-is the next symbol to be match and is either:
-
-* A terminal token, matched directly with the token at the current
-        string position.
-* A nonterminal 'Y', for which 'f_Y' is called. In the case of
-        LL(1) deterministic parsing, only one (or none) of the productions
-        of 'Y' passes the lookahead-test, say "Y ::= c", and its branch
[…]

### 2019-09-27 syncon-parser/gll/src/GLL/Types/Derivations.hs

@@ -1,324 +1,0 @@
-{-# OPTIONS_GHC -fno-warn-name-shadowing #-}
-{-# LANGUAGE StandaloneDeriving #-}
-
-module GLL.Types.Derivations where
-
-import Prelude
-
-import qualified    Data.Map as M
-import qualified    Data.IntMap as IM
-import qualified    Data.Set as S
-import qualified    Data.IntSet as IS
-import              Data.List (findIndices, partition, tails)
-import Data.Function ((&))
-import Data.Functor ((<&>))
-import Control.Arrow ((>>>), (&&&))
-import Data.Foldable (fold)
-import GLL.Types.Grammar
-
--- make sure that tokens are equal independent of their character level value
-type SlotL t    = (Slot t, Int)                   -- slot with left extent
-type PrL t      = (Prod t, Int)                     -- Production rule with left extent
-type NtL        = (Nt, Int)                     -- Nonterminal with left extent
-
--- SPPF
-
--- |
--- An 'SPPF' contains symbol nodes, intermediate nodes, packed nodes and edges between them.
--- See Scott and Johnstone (2013) for an explanation of the 'SPPF'.
-type SPPF t     =   (SymbMap t, ImdMap t, PackMap t, EdgeMap t)
-
--- |
--- Stores packed nodes using nested "Data.IntMap"s, nesting is as follows:
---
--- * left extent
--- * right extent
--- * dot position (from left to right)
--- * mapping from productions to set of pivots
-type PackMap t  =   IM.IntMap (IM.IntMap (IM.IntMap (M.Map (Prod t) IS.IntSet)))
-
--- |
--- Stores symbol nodes using nested "Data.IntMap"s, nesting is as follows:
---
--- * left extent
--- * right extent
--- * set of symbols
-type SymbMap t  =   IM.IntMap (IM.IntMap (S.Set (Symbol t)))
-
--- |
--- Stores intermediate nodes using nested "Data.IntMap"s, nesting is as follows:
---
--- * left extent
--- * right extent
--- * set of slots
-type ImdMap t   =   IM.IntMap (IM.IntMap (S.Set (Slot t)))
-
--- |
--- Stores edges, potentially costly.
-type EdgeMap t  =   M.Map (SPPFNode t) (S.Set (SPPFNode t))
-
--- |
[…]

### 2019-09-27 syncon-parser/gll/src/GLL/Types/Grammar.hs

@@ -1,191 +1,0 @@
-{-# OPTIONS_GHC -fno-warn-name-shadowing #-}
-{-# LANGUAGE StandaloneDeriving #-}
-
-
--- UUAGC 0.9.52.1 (src/GLL/Types/Abstract.ag)
-module GLL.Types.Grammar where
-
-import Prelude
-
-import Data.Text
-
--- | Identifier for nonterminals.
-type Nt  = Text
--- Prod ---------------------------------------------------------
-
--- |
--- A production binds a nonterminal identifier (left-hand side) to a list of symbols
---(the right-hand side of the production).
-data Prod t = Prod (Nt) (Symbols t)
--- Prods --------------------------------------------------------
--- | A list of 'Prod's.
-type Prods t = [Prod t]
--- Grammar -----------------------------------------------------
--- |
--- A grammar is a start symbol and a list of productions.
-type Grammar t = (Nt, Prods t)
--- Slot --------------------------------------------------------
--- |
--- A grammar slot acts as a label to identify progress of matching a production.
--- As such, a slot is a "Prod" with its right-hand side split in two:
--- a part before and a part after 'the dot'.
--- The dot indicates which part of the right-hand side has been processed thus far.
-data Slot t = Slot (Nt) (([Symbol t])) (([Symbol t]))
--- Symbol ------------------------------------------------------
-
--- |
--- A 'Symbol' is either a nonterminal or a terminal,
--- where a terminal contains some arbitrary token.
-data Symbol t   = Nt Nt
-                | Term t
---                | Error (Token) (Token)
--- Symbols -----------------------------------------------------
-
--- |
--- A list of 'Symbol's
-type Symbols t = [Symbol t]
--- Token -------------------------------------------------------
-
--- |
--- A datatype for representing tokens with some builtins
--- and an aribitrary Token constructor.
--- This datatype stores (optional) lexemes.
-data Token = Char       Char
-           | Keyword    String
-           | EOS
-           | Epsilon
-           | IntLit     (Maybe Int)
-           | FloatLit   (Maybe Double)
-           | BoolLit    (Maybe Bool)
-           | StringLit  (Maybe String)
-           | CharLit    (Maybe Char)
-           | IDLit      (Maybe String)
[…]

### 2019-09-27 syncon-parser/src/P1Lexing/Types.hs

@@ -7,8 +7,6 @@ import Codec.Serialise (Serialise)
 
 import Text.Earley.Forest.Grammar as Forest
 
-import P4Parsing.ForestParser (Unlexable(..))
-
 data Range = Nowhere | Range !Text !Position !Position deriving (Show, Eq, Ord, Data, Typeable, Generic)
 data Position = Position { line :: !Int, column :: !Int } deriving (Show, Eq, Ord, Data, Typeable, Generic)
 
@@ -79,10 +77,6 @@ instance Ranged (a, Range) where
 instance Ranged (Token l n) where
   range (LitTok r _ _) = r
   range (OtherTok r _ _ _) = r
-
-instance Unlexable (Token l n) where
-  unlex (LitTok _ _ t) = t
-  unlex (OtherTok _ _ _ t) = t
 
 instance Hashable Range
 instance Hashable Position

### 2019-09-27 syncon-parser/src/P4Parsing/ForestParser.hs

@@ -1,64 +1,0 @@
-{-# LANGUAGE ViewPatterns #-}
-
--- module P4Parsing.ForestParser where
-module P4Parsing.ForestParser
-( Prod
-, Grammar
-, rule
-, ranged
-, terminal
-, ambig
-, alts
-, parse
-, Node
-, forestToDot
-, Unlexable(..)
-) where
-
-import Pre hiding (from)
-
-import qualified Data.Text as Text
-import qualified Data.HashMap.Lazy as M
-
-import P4Parsing.ForestParser.Grammar (Prod, Grammar, rule, alts, terminal, ambig, ranged, Unlexable(..))
-import P4Parsing.ForestParser.GLL (parse, Node)
-
-forestToDot :: forall nodeF n. (Foldable nodeF, Functor nodeF, Eq n, Hashable n)
-            => (nodeF () -> Text) -> (HashMap n (nodeF (HashSet n)), HashSet n) -> Text
-forestToDot showNode (nodeMap, roots) = "digraph {\n"
-  <> "  startState [shape=point, label=\"\"];\n"
-  <> foldMap (\n -> "  startState -> " <> show (keyToI n) <> ";\n") roots
-  <> foldMap nodeDesc (M.toList nodeMap)
-  <> nodeEdgeDescs
-  <> "}\n"
-  where
-    keyToIMap = M.keys nodeMap `zip` [(0::Int)..] & M.fromList
-    keyToI key = M.lookup key keyToIMap & compFromJust "P4Parsing.ForestParser.forestToDot" "Missing key in keyToIMap"
-    nodeDesc (key, node) = "  " <> show (keyToI key) <> " [label = \"" <> escape (showNode $ void node) <> "\"];\n"
-    firstAmbigNode = maximum keyToIMap + 1
-
-    genAlt :: HashSet n -> State (HashMap (HashSet n) Int, Int) (Int, Text)
-    genAlt alt = do
-      (prevs, id) <- get
-      case M.lookup alt prevs of
-        Just prevId -> return (prevId, "")
-        Nothing -> do
-          let altDesc = "  " <> show id <> " [shape=point, label=\"\"];\n"
-              edges = "  " <> show id <> " -> {" <> (toList alt <&> keyToI <&> show & intersperse ", " & Text.concat) <> "};\n"
-          modify $ M.insert alt id *** (+1)
-          return (id, altDesc <> edges)
-
[…]

### 2019-09-27 syncon-parser/src/P4Parsing/ForestParser/GLL.hs

@@ -1,367 +1,0 @@
-{-# OPTIONS_GHC -fno-warn-orphans #-}
-
-module P4Parsing.ForestParser.GLL (parse, Node) where
-
-import Pre hiding (Symbol, Any, State)
-
-import qualified Prelude as Prelude
-import Unsafe.Coerce (unsafeCoerce)
-import GHC.Exts (Any)
-
-import Data.Array (Array, listArray, bounds)
-import qualified Data.Array as Array
-import Data.STRef (STRef, newSTRef, readSTRef, modifySTRef')
-import qualified Data.Text as Text
-import qualified Data.HashMap.Lazy as M
-import qualified Data.HashSet as S
-import qualified Data.IntMap.Strict as IM
-import qualified Data.IntSet as IS
-import qualified Data.Map.Strict as OM
-import qualified Data.Sequence as Seq
-
-import GLL.Parser (Parseable(..), Symbol(..), ParseResult(ParseResult, res_success, error_message, sppf_result))
-import qualified GLL.Parser as GLL
-
-import P4Parsing.ForestParser.Grammar
-import P4Parsing.ForestParser.Grammar as Grammar
-
--- interface
-
-newtype Node = Node Int deriving (Eq, Hashable, Show, NFData)
-
--- | Parse a sequence given a grammar. The underlying parser doesn't give an easily inspectable error,
--- hence we just pass that on, for the moment. In the success case you get a tuple containing the nodes
--- in the parse forest and a sequence of root nodes.
-parse :: forall t nodeF. (Show t, Ord t, Hashable t, Unlexable t)
-      => (forall r. Grammar r t nodeF (Prod r t nodeF r))
-      -> (forall f. Foldable f => f t -> Either Text (HashMap Node (nodeF (HashSet Node)), HashSet Node))  -- TODO: might want to use a 'ghosts of departed proofs' approach to the map and the nodes
-parse unfixedGrammar = go
-  where
-    go :: Foldable f => f t -> Either Text (HashMap Node (nodeF (HashSet Node)), HashSet Node)
-    go foldable = toList input <&> Concrete & gll <&> repack input <&> buildDAG
-      where
-        input = listArray (0, length foldable - 1) (toList foldable)
-
[…]

### 2019-09-27 syncon-parser/src/P4Parsing/ForestParser/Grammar.hs

@@ -1,123 +1,0 @@
-module P4Parsing.ForestParser.Grammar
-( Prod(..)
-, Grammar(..)
-, rule
-, terminal
-, ranged
-, ambig
-, alts
-, runGrammar
-, Unlexable(..)
-) where
-
-import Pre
-
-import Control.Monad.Fix (MonadFix(..))
-
--- This will form a sort of "linked list" of symbols in the right hand side of a single production
-data Prod r t nodeF a where
-  Terminal :: !Text -> !(t -> Bool) -> !(Prod r t nodeF (t -> b)) -> Prod r t nodeF b
-  NonTerminal :: !r -> !(Prod r t nodeF (r -> a)) -> Prod r t nodeF a
-  Ranged :: !(Prod r t nodeF (Maybe (t, t) -> a)) -> !(Prod r t nodeF (a -> b)) -> Prod r t nodeF b
-  Pure :: a -> Prod r t nodeF a
-  Alts :: ![Prod r t nodeF a] -> !(Prod r t nodeF (a -> b)) -> Prod r t nodeF b
-  Many :: !(Prod r t nodeF a) -> !(Prod r t nodeF ([a] -> b)) -> Prod r t nodeF b
-
--- | Match a token for which the given predicate returns @Just a@,
--- and return the @a@.
-terminal :: Text -> (t -> Bool) -> Prod r t nodeF t
-terminal label p = Terminal label p $ Pure identity
-
-instance Functor (Prod r t nodeF) where
-  {-# INLINE fmap #-}
-  fmap f (Terminal label t build) = Terminal label t $ (>>> f) <$> build
-  fmap f (NonTerminal r build) = NonTerminal r $ (>>> f) <$> build
-  fmap f (Ranged r build) = Ranged r $ (>>> f) <$> build
-  fmap f (Pure a) = Pure $ f a
-  fmap f (Alts ps build) = Alts ps $ (>>> f) <$> build
-  fmap f (Many p build) = Many p $ (>>> f) <$> build
-
-alts :: [Prod r t nodeF a] -> Prod r t nodeF (a -> b) -> Prod r t nodeF b
-alts as build = case as >>= go of
-  [] -> empty
-  [a] -> a <**> build
-  as' -> Alts as' build
-  where
-    go (Alts [] _) = []
-    go (Alts as' (Pure f)) = fmap f <$> as'
-    go a = [a]
-
-ranged :: Prod r t nodeF (Maybe (t, t) -> a) -> Prod r t nodeF a
[…]

### 2019-09-27 syncon-parser/src/P4Parsing/Parser.hs

@@ -1,73 +1,91 @@
-{-# LANGUAGE RecordWildCards #-}
+{-# LANGUAGE RecordWildCards, ViewPatterns, UndecidableInstances #-}
 
-module P4Parsing.Parser (Error(..), parseSingleLanguage, parseTokens, dfToLanguageTokens) where
+module P4Parsing.Parser (Error(..), precomputeSingleLanguage, dfToLanguageTokens, parseTokens, parseFile, forestToDot, precomputeToSerialisable, serialisableToPrecompute) where
 
-import Pre
+import Pre hiding (from, some, many, optional)
 import Result (Result(..))
 
 import Control.Monad.Fix (mfix)
-import qualified Data.Sequence as Seq
 import qualified Data.HashMap.Strict as M
 import qualified Data.HashSet as S
+import qualified Data.Text as Text
 
 import Data.Generics.Uniplate.Data (universe)
 import Data.Functor.Foldable (para)
 
-import ErrorMessage (FormatError(..), simpleErrorMessage)
-import qualified P4Parsing.ForestParser as Forest
-import P4Parsing.ForestParser (Node, terminal, ambig, rule, parse)
+import Text.Earley.Forest.Grammar (terminal, ambig, node, NodeKind(..), Parseable(..), (<--), (-->), noparse, nodeLeaf, label, alts, many, some, optional)
+import Text.Earley.Forest.Transformations (mkGrammar, mkNNFGrammar)
+import Text.Earley.Forest.Parser (precompute, recognize, mkDerivation, Node)
+import qualified Text.Earley.Forest.Grammar as Forest
+import qualified Text.Earley.Forest.Parser as Forest
 
-import P1Lexing.Types (Token(..), Range(Nowhere), range)
+import ErrorMessage (FormatError(..), simpleErrorMessage)
+
+import P1Lexing.Types (Token(..), TokenKind(..), Ranged(..))
 import P1Lexing.Lexer (LanguageTokens(..))
 import qualified P1Lexing.Lexer as Lexer
 import P2LanguageDefinition.Types (DefinitionFile(..), TypeName(..), Name(..), Syncon(..), Comment(..), Elaboration, SyntaxDescription(..), SyntaxDescriptionF(..), TokenType(..), SDName, SyntaxType, Repetition(..), Rec(..))
 import P2LanguageDefinition.Elaborator (elaborate)
[…]

### 2019-09-27 syncon-parser/src/P4Parsing/Parser2.hs

@@ -1,266 +1,0 @@
-{-# LANGUAGE RecordWildCards, ViewPatterns, UndecidableInstances #-}
-
-module P4Parsing.Parser2 (Error(..), precomputeSingleLanguage, dfToLanguageTokens, parseTokens, parseFile, forestToDot, precomputeToSerialisable, serialisableToPrecompute) where
-
-import Pre hiding (from, some, many, optional)
-import Result (Result(..))
-
-import Control.Monad.Fix (mfix)
-import qualified Data.HashMap.Strict as M
-import qualified Data.HashSet as S
-import qualified Data.Text as Text
-
-import Data.Generics.Uniplate.Data (universe)
-import Data.Functor.Foldable (para)
-
-import Text.Earley.Forest.Grammar (terminal, ambig, node, NodeKind(..), Parseable(..), (<--), (-->), noparse, nodeLeaf, label, alts, many, some, optional)
-import Text.Earley.Forest.Transformations (mkGrammar, mkNNFGrammar)
-import Text.Earley.Forest.Parser (precompute, recognize, mkDerivation, Node)
-import qualified Text.Earley.Forest.Grammar as Forest
-import qualified Text.Earley.Forest.Parser as Forest
-
-import ErrorMessage (FormatError(..), simpleErrorMessage)
-
-import P1Lexing.Types (Token(..), TokenKind(..), Ranged(..))
-import P1Lexing.Lexer (LanguageTokens(..))
-import qualified P1Lexing.Lexer as Lexer
-import P2LanguageDefinition.Types (DefinitionFile(..), TypeName(..), Name(..), Syncon(..), Comment(..), Elaboration, SyntaxDescription(..), SyntaxDescriptionF(..), TokenType(..), SDName, SyntaxType, Repetition(..), Rec(..))
-import P2LanguageDefinition.Elaborator (elaborate)
-import P4Parsing.Types (NodeF, SL, SingleLanguage(..))
-
-data Error l tok
-  = LexingError (Lexer.Error l TypeName)
-  | ParseError (HashSet (Maybe (TokKind tok))) (Maybe tok)
-  | MissingTop
-
-instance (Show l, Ranged tok, Parseable tok) => FormatError (Error l tok) where
-  formatError _ (LexingError e) = formatError () e
-  formatError _ (ParseError expected found) = simpleErrorMessage (foldMap range found) $
-    "Parse error, found " <> maybe "<end-of-file>" unlex found <> " expected one of:\n"
[…]

### 2019-09-27 syncon-parser/test/Main.hs

@@ -13,11 +13,12 @@ import Data.IORef (newIORef, readIORef, IORef, writeIORef)
 import qualified Data.HashMap.Strict as M
 import Data.List (partition)
 
+import P1Lexing.Types (Token)
 import qualified P1Lexing.Lexer as Lexer
+import P2LanguageDefinition.Types (TypeName(..))
 import qualified P2LanguageDefinition.Parser as LD
 import qualified P2LanguageDefinition.BasicChecker as LD
 import qualified P4Parsing.Parser as Parser
-import qualified P4Parsing.Parser2 as Parser2
 import P4Parsing.Types (SingleLanguage(..))
 
 main :: IO ()
@@ -36,15 +37,13 @@ langToTest failRef (lang, (defFiles, (successFiles, failFiles))) = do
   mapM_ (mkTest env True) successFiles
   mapM_ (mkTest env False) failFiles
   where
-    mkTest (lexFile, parseTokens, parseTokens2) expected srcFile = do
+    mkTest (lexFile, precomputed) expected srcFile = do
       toks <- doLexFile lexFile srcFile
-      let gllResult = parseTokens toks & isData & (== expected)
-          earleyResult = parseTokens2 toks & isData & (== expected)
+      let earleyResult = Parser.parseTokens precomputed toks & isData & (== expected)
           showRes True  = "    "
           showRes False = "FAIL"
-      putStrLn $ showRes gllResult <> " parser/" <> lang <> "/" <> takeFileName srcFile <> "/gll"
       putStrLn $ showRes earleyResult <> " parser/" <> lang <> "/" <> takeFileName srcFile <> "/earley-forest"
-      when (not gllResult || not earleyResult) $ writeIORef failRef True
+      when (not earleyResult) $ writeIORef failRef True
     doLexFile lexFile path = lexFile path >>= \case
       Error _ -> die $ "Could not lex file " <> toS path
       Data toks -> return toks
@@ -62,7 +61,7 @@ getLanguages languageDirectory = do
   where
     listDir path = listDirectory path <&> fmap (path </>)
 
-mkEnv lang defFiles = mkDf >>= \df -> (,,) <$> mkLexer df <*> mkGllParser df <*> mkEarleyParser df
+mkEnv lang defFiles = mkDf >>= \df -> (,) <$> mkLexer df <*> mkEarleyParser df
   where
     mkDf = do
       foldMapM LD.parseFile defFiles >>= \case
[…]

### 2019-09-27 syncon-parser/earley-forest/src/Text/Earley/Forest/Parser.hs
Removed todo

@@ -1,4 +1,3 @@
-{-# OPTIONS_GHC -Wno-unused-imports #-} -- TODO: remove
 {-# LANGUAGE NamedFieldPuns, FlexibleContexts, ScopedTypeVariables, DeriveGeneric, RankNTypes, ViewPatterns, LambdaCase, GeneralizedNewtypeDeriving, TypeApplications, TupleSections, RecordWildCards #-}
 
 module Text.Earley.Forest.Parser
@@ -15,8 +14,7 @@ import Prelude hiding (pred)
 import GHC.Generics (Generic)
 
 import Codec.Serialise (Serialise)
-import Control.Applicative ((<|>))
-import Control.Arrow ((>>>), first)
+import Control.Arrow ((>>>))
 import Control.DeepSeq (NFData)
 import Control.Monad (unless, join, (>=>))
 import Control.Monad.ST (ST, runST)
@@ -28,7 +26,7 @@ import Data.Functor ((<&>), void)
 import Data.HashMap.Lazy (HashMap)
 import Data.HashSet (HashSet)
 import Data.Hashable (Hashable)
-import Data.Maybe (catMaybes, mapMaybe, isJust)
+import Data.Maybe (catMaybes, mapMaybe)
 import Data.STRef (STRef, newSTRef)
 import Data.Sequence (Seq((:|>)), (|>))
 import Data.Traversable (forM)
@@ -41,9 +39,8 @@ import qualified Data.Vector as Vector
 
 import Text.Earley.Forest.Grammar (TokKind, Parseable, getKind)
 import Text.Earley.Forest.ParseTree (ParseTree(..))
-import Text.Earley.Forest.SplitEpsDFA (EpsDFA(..), mkDFA, renumberStates, completedNT, isCompleted, DotProd(..), dotToRule)
+import Text.Earley.Forest.SplitEpsDFA (EpsDFA(..), mkDFA, renumberStates, completedNT, dotToRule)
 import Text.Earley.Forest.Transformations (NT(..), Rule(..), Sym(..), NtKind(..), rightRecursive, NtKind(..), ruleAsTuple)
-import qualified Text.Earley.Forest.ParseTree as PT
 
 -- | Several Earley items, in the form of an EpsDFA state and origin
 data Eim = Eim

### 2019-09-30 paper/paper.tex
Pre-llncs switch

@@
 in domain-specific language development.  An important
-objective when designing domain-specific languages is to be able construct
+objective when designing domain-specific languages is to be able to construct
 languages by composing and extending existing
@@
   decidable algorithm that is sound and complete for the first
-  version, and sound for the second (Section~\ref{sec:static}).
+  class, and sound for the second (Section~\ref{sec:static}).
 \item We devise a decidable algorithm that solves the
@@
 A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
-A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton has one of the following three forms:
+A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
 
@@
 
-\noindent This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}
+\noindent Note that we do not require each tree to have a unique word, i.e., there may exist two distinct $w_1$ and $w_2$ such that $t \in \parse(w_1)$ and $t \in \parse(w_2)$.
+This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}
 
@@
 
-The example $\parse(1 + 2 \cdot 3)$
+The words $w$ and $w'$ will typically be related, e.g., $w'$ has a few added pairs of parentheses but is otherwise identical.
+The example $1 + 2 \cdot 3$
 %in Fig.~\ref{fig:arith-example}
@@
 \item If a given parse tree $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous.
-In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
+In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert = 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
 \end{itemize}
@@
 
-  $f(\parse, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
+  $f(P, w) \implies \rho_P(w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
 \end{definition}
@@
 
-  $\rho_P(w) \implies f(\parse, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
+  $\rho_P(w) \implies f(P, w) \text{ for all } w \in \Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$
 \end{definition}
@@
 
-In future work, we will investigate weakning the restrictions
+In future work, we will investigate weakening the restrictions
 currently needed for the static and dynamic resolvability

### 2019-10-15 paper/paper.tex
EOD

@@
-%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
-\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For double-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
-%% For single-blind review submission, w/ CCS and ACM Reference
-%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission, w/ required CCS and ACM Reference
-%\documentclass[acmsmall]{acmart}\settopmatter{}
-
-%% Journal information
-%% Supplied to authors by publisher for camera-ready submission;
-%% use defaults for review submission.
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
-\acmArticle{1}
-\acmYear{2018}
-\acmMonth{1}
-\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission;
-%% use 'none' for review submission.
-\setcopyright{none}
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2018}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-\citestyle{acmauthoryear}  %% For author/year citations
-%\citestyle{acmnumeric}     %% For numeric citations
-%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
-                            %% sorting of references within a single citation;
-                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
-                            %% rendered as [14,5,2] rather than [2,5,14].
-%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
-                            %% compression of sequential references within a
-                            %% single citation;
-                            %% e.g., \cite{Baker12,Baker14,Baker16}
-                            %% rendered as [2,3,4] rather than [2-4].
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Note: Authors migrating a paper from traditional SIGPLAN
-%% proceedings format to PACMPL format must update the
-%% '\documentclass' and topmatter commands above; see
-%% 'acmart-pacmpl-template.tex'.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% Some recommended packages.
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.20 of 2017/10/04
+%
+\documentclass[runningheads]{llncs}
+%
+\usepackage{cite}
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+% If you use the hyperref package, please uncomment the following line
+% to display URLs in blue roman font according to Springer's eBook style:
+% \renewcommand\UrlFont{\color{blue}\rmfamily}
+
 \usepackage{booktabs}   %% For formal tables:
                         %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-
+%% \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+%%                         %% http://ctan.org/pkg/subcaption
+\usepackage{subfig}
+\usepackage{amsmath}
+\usepackage{amsfonts}
 \usepackage{tikz}
@@
 \newcommand{\regex}{\mathit{Reg}}
-\newcommand{\reqpl}{\underline{(}}
+\newcommand{\reqpl}{(}
 \newcommand{\reqpr}{)}
 \newcommand{\reqp}[1]{\reqpl#1\reqpr}
-\newcommand{\pospl}{(}
-\newcommand{\pospr}{)}
+\newcommand{\pospl}{[}
+\newcommand{\pospr}{]}
 \newcommand{\posp}[1]{\pospl#1\pospr}
@@
 
-%% Title information
-\title{Resolvable Ambiguity}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{Viktor Palmkvist}
-\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position1}
-  \department{Department1}              %% \department is recommended
-  \institution{KTH Royal Institute of Technology}            %% \institution is required
-  \streetaddress{Street1 Address1}
-  \city{Stockholm}
-  \state{State1}
-  \postcode{Post-Code1}
-  \country{Sweden}                    %% \country is recommended
-}
-\email{vipa@kth.se}          %% \email is recommended
-
-%% Author with two affiliations and emails.
-\author{First2 Last2}
-\authornote{with author2 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-  \position{Position2a}
-  \department{Department2a}             %% \department is recommended
-  \institution{Institution2a}           %% \institution is required
-  \streetaddress{Street2a Address2a}
-  \city{City2a}
-  \state{State2a}
-  \postcode{Post-Code2a}
-  \country{Country2a}                   %% \country is recommended
-}
-\email{first2.last2@inst2a.com}         %% \email is recommended
-\affiliation{
-  \position{Position2b}
-  \department{Department2b}             %% \department is recommended
-  \institution{Institution2b}           %% \institution is required
-  \streetaddress{Street3b Address2b}
-  \city{City2b}
-  \state{State2b}
-  \postcode{Post-Code2b}
-  \country{Country2b}                   %% \country is recommended
-}
-\email{first2.last2@inst2b.org}         %% \email is recommended
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
+\title{Contribution Title\thanks{Supported by organization x.}}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
+Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+%
+\authorrunning{F. Author et al.}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{Princeton University, Princeton NJ 08544, USA \and
+Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
+\email{lncs@springer.com}\\
+\url{http://www.springer.com/gp/computer-science/lncs} \and
+ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+\email{\{abc,lncs\}@uni-heidelberg.de}}
+%
+\maketitle              % typeset the header of the contribution
+%
 \begin{abstract}
@@
 composability of domain-specific languages.
+\keywords{First keyword  \and Second keyword \and Another keyword.}
 \end{abstract}
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{Language Composition, Context Free Grammars, Domain-Specific Languages}  %% \keywords are mandatory in final camera-ready submission
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
+%
+%
+%
 
@@
 However, recent work by
-~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
+Palmkvist and Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019} shows why
 carefully designed ambiguous syntax definitions are in some cases
@@
 For proof-technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense, when marks are removed.
-This can be checked using, e.g., the approach described by \citet{brabrandTypedUnambiguousPattern2010}.
+This can be checked using, e.g., the approach described by Brabrand and Thomsen~\cite{brabrandTypedUnambiguousPattern2010}.
 In practice, this is likely desirable regardless, since a user is likely to attach different semantic meanings to each occurrence of a non-terminal.
@@
 \begin{figure}
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
+  \subfloat[$T_D$, the parse trees of $G_D$.\label{fig:running-example-generated:t}]{%
     \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
       \toprule
-      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
-      $E$ & $m$ & $E$ \verb|'*'| $E$ $)$ \\
+      $E$ & $l$ & \texttt{'['} $(\epsilon + E($\texttt{';'} $E)^{*})$ \texttt{']'} $)$ \\
+      $E$ & $a$ & $E$ \texttt{'+'} $E$ $)$ \\
+      $E$ & $m$ & $E$ \texttt{'*'} $E$ $)$ \\
       $E$ & $n$ & $N$ $)$ \\
@@
     \end{tabular}
-    \caption{$T_D$, the parse trees of $G_D$.}
-    \label{fig:running-example-generated:t}
-  \end{subfigure}%
-%
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
+  }
+  \subfloat[$T'_D$, the parse trees of $G'_D$.\label{fig:running-example-generated:t-prime}]{%
     \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
       \toprule
-      $E$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E$ & $a$ & $E$ \verb|'+'| $E$ $)$ \\
-      $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
+      $E$ & $l$ & \texttt{'['} $(\epsilon + E($\texttt{';'} $E)^{*})$ \texttt{']'} $)$ \\
+      $E$ & $a$ & $E$ \texttt{'+'} $E$ $)$ \\
+      $E$ & $m$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ $)$ \\
       $E$ & $n$ & $N$ $)$ \\
-      $E$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
+      $E$ & $g$ & \texttt{'('} $E$ \texttt{')'} $)$ \\
       \midrule
-      $E_{\{a\}}$ & $l$ & \verb|'['| $(\epsilon + E($\verb|';'| $E)^{*})$ \verb|']'| $)$ \\
-      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ $)$ \\
+      $E_{\{a\}}$ & $l$ & \texttt{'['} $(\epsilon + E($\texttt{';'} $E)^{*})$ \texttt{']'} $)$ \\
+      $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ $)$ \\
       $E_{\{a\}}$ & $n$ & $N$ $)$ \\
-      $E_{\{a\}}$ & $g$ & \verb|'('| $E$ \verb|')'| $)$ \\
+      $E_{\{a\}}$ & $g$ & \texttt{'('} $E$ \texttt{')'} $)$ \\
       \bottomrule
     \end{tabular}
-    \caption{$T'_D$, the parse trees of $G'_D$.}
-    \label{fig:running-example-generated:t-prime}
-  \end{subfigure}
-
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
+  }
+
+  \subfloat[$G_D$, the generated abstract syntax.\label{fig:running-example-generated:w}]{%
     \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
       \toprule
-      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
-      $E$ & $E$ \verb|'+'| $E$ \\
-      $E$ & $E$ \verb|'*'| $E$ \\
+      $E$ & \texttt{'['} $E_{l1}$ \texttt{']'} \\
+      $E$ & $E$ \texttt{'+'} $E$ \\
+      $E$ & $E$ \texttt{'*'} $E$ \\
       $E$ & $N$ \\
@@
       $E_{l2}$ & $\epsilon$ \\
-      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
+      $E_{l2}$ & \texttt{';'} $E$ $E_{l2}$ \\
       \bottomrule
     \end{tabular}
-    \caption{$G_D$, the generated abstract syntax.}
-    \label{fig:running-example-generated:w}
-  \end{subfigure}%
-%
-  \begin{subfigure}[t]{.45\linewidth}
-    \centering
+  }
+  \subfloat[$G'_D$, the generated concrete syntax.\label{fig:running-example-generated:w-prime}]{%
     \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
       \toprule
-      $E$ & \verb|'['| $E_{l1}$ \verb|']'| \\
-      $E$ & $E$ \verb|'+'| $E$ \\
-      $E$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
+      $E$ & \texttt{'['} $E_{l1}$ \texttt{']'} \\
+      $E$ & $E$ \texttt{'+'} $E$ \\
+      $E$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ \\
       $E$ & $N$ \\
-      $E$ & \verb|'('| $E$ \verb|')'| \\
+      $E$ & \texttt{'('} $E$ \texttt{')'} \\
       \midrule
-      $E_{\{a\}}$ & \verb|'['| $E_{l1}$ \verb|']'| \\
-      $E_{\{a\}}$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
+      $E_{\{a\}}$ & \texttt{'['} $E_{l1}$ \texttt{']'} \\
+      $E_{\{a\}}$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ \\
       $E_{\{a\}}$ & $N$ \\
-      $E_{\{a\}}$ & \verb|'('| $E$ \verb|')'| \\
+      $E_{\{a\}}$ & \texttt{'('} $E$ \texttt{')'} \\
       \midrule
@@
       $E_{l2}$ & $\epsilon$ \\
-      $E_{l2}$ & \verb|';'| $E$ $E_{l2}$ \\
+      $E_{l2}$ & \texttt{';'} $E$ $E_{l2}$ \\
       \bottomrule
     \end{tabular}
-    \caption{$G'_D$, the generated concrete syntax.}
-    \label{fig:running-example-generated:w-prime}
-  \end{subfigure}
+  }
   \caption{The generated grammars.}
@@
 
+% TODO: (re)make these figures
 \begin{figure}
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{center}
-    \Tree [.{$m$}
-        [.{$a$}
-          [.{$n$} \terminal{1} ]
-          \terminal{+}
-          [.{$n$} \terminal{2} ] ]
-        \terminal{*}
-        [.{$n$} \terminal{3} ] ]
-    \end{center}
-    \[m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))\]
-    \caption{Example tree in $L(T_D)$ and visualization.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.45\linewidth}
-    \begin{center}
-    \Tree [.{$m$}
-        [.{$g$}
-          \terminal{(}
-          [.{$a$}
-            [.{$n$} \terminal{1} ]
-            \terminal{+}
-            [.{$n$} \terminal{2} ] ]
-          \terminal{)} ]
-        \terminal{*}
-        [.{$n$} \terminal{3} ] ]
-    \end{center}
-    \[m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))\]
-    \caption{Example tree in $L(T'_D)$ and visualization.}
-  \end{subfigure}
-
-  \begin{subfigure}[b]{.40\linewidth}
-    \[1 + 2 * 3\]
-    \caption{Example word in $L(G_D)$.}
-  \end{subfigure}
-  \begin{subfigure}[b]{.40\linewidth}
-    \[(1 + 2) * 3\]
-    \caption{Example word in $L(G'_D)$.}
-  \end{subfigure}
+  \subfloat[Example tree in $L(T_D)$ and visualization.\label{fig:sum-add-tree}]{%
+    %% \begin{center}
+    %% \Tree [.{$m$}
+    %%     [.{$a$}
+    %%       [.{$n$} \terminal{1} ]
+    %%       \terminal{+}
+    %%       [.{$n$} \terminal{2} ] ]
+    %%     \terminal{*}
+    %%     [.{$n$} \terminal{3} ] ]
+    %% \end{center}
+    $m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))$
+  }
+  \subfloat[Example tree in $L(T'_D)$ and visualization.]{%
+    %% \begin{center}
+    %% \Tree [.{$m$}
+    %%     [.{$g$}
+    %%       \terminal{(}
+    %%       [.{$a$}
+    %%         [.{$n$} \terminal{1} ]
+    %%         \terminal{+}
+    %%         [.{$n$} \terminal{2} ] ]
+    %%       \terminal{)} ]
+    %%     \terminal{*}
+    %%     [.{$n$} \terminal{3} ] ]
+    %% \end{center}
+    $m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))$
+  }
+
+  \subfloat[Example word in $L(G_D)$.]{%
+    $1 + 2 * 3$
+  }
+  \subfloat[Example word in $L(G'_D)$.]{%
+    $(1 + 2) * 3$
+  }
   \caption{Example with elements from each generated language that correspond to each other. The leaf terminals in the tree languages appear underlined to distinguish the two kinds of parentheses.}
@@
 We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}.
-Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem with one caveat: we only consider languages with balanced parentheses.
+Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem, with one caveat: we only consider languages with balanced parentheses.
 
@@
 
-\begin{theorem}
+\begin{lemma}
   Given an abstract parser $P = (\T, T, \parse)$, the language $L(P)$ is resolvably ambiguous iff $\quad\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
-\end{theorem}
-
-\begin{proof}
-  The quantifier $\forall t \in T$ is equivalent with $\forall w \in \T^{*}. \forall t \in \parse(w)$, since $T = \bigcup_{w \in \T^{*}} \parse(w)$ (by definition).
-\end{proof}
+\end{lemma}
 
@@
 
-Before an actual algorithm, we split the problem into three different versions of increasing difficulty.
-We then lay the groundwork for our correctness proofs, eventually culminating in the algorithm, which we show to be sound and complete for version 1 and sound for version 2.
-
-The three versions are as follows:
+Our approach finds this counterexample by finding a pair of trees $t_1$ and $t_2$ such that $\words(t_1) \subseteq \words(t_2)$.
+Such a pair trivially implies that $t_1$ has no unambiguous words, since $\forall w.\ t_1 \in \parse(w) -> t_2 \in \parse(w)$, which thus implies that the examined language is unresolvably ambiguous.
+The converse is less obvious, in fact, the absence of such a pair does not guarantee a resolvably ambiguous language in general.
+We shall however show it to be sufficient for a limited class of languages.
+
+Central to our approach is the shape of $\words(t)$.
+Consider the tree corresponding to \verb|'(1 + 2) * 3'| (visualized in Figure~\ref{fig:sum-add-tree}).
+Each node in the tree may be surrounded by zero or more parentheses, except \verb|'1 + 2'|, which requires at least one pair.
+Its language is thus represented by $\{(^{n_1}(^{n_2}(^{n_3}1)^{n_3} + (^{n_4}2)^{n_4})^{n_2} * (^{n_5}2)^{n_5})^{n_1} \mid n_i \in \{0, 1, \ldots\}, n_2 \geq 1\}$.
+We can encode this language as a word, if we take ``$\reqp{}$'' to mean ``exactly one pair of parentheses'' and ``$\posp{}$'' to mean ``zero or more pairs of parentheses'': ``$\posp{\posp{\reqp{\posp{1} + \posp{2}}} * \posp{3}}$''.
+This encoding lets us reduce comparisons between languages to comparisons between words, with one caveat: the encoding is not unique.
+For example, ``$\reqp{\posp{}}$'' encodes the same language as ``$\posp{\reqp{}}$'', and ``$\posp{\posp{}}$'' encodes the same language as ``$\posp{}$''.
+We rectify this by repeatedly swapping ``$\reqp{\posp{\ldots}}$'' with ``$\posp{\reqp{\ldots}}$'' and ``$\posp{\posp{\ldots}}$'' with ``$\posp{\ldots}$'' until we get stuck.
+We call the result the canonical linear encoding of the (word) language of a tree in $L(T_D)$, which \emph{is} unique.
+
+We now divide the possible languages along two axes: whether a language has (non-grouping) parentheses, and whether a language has marks.
+The parentheses present in the canonical encoding will have the following shapes:
+
+\begin{center}
+\begin{tabular}{r@{\quad}c@{\quad}c}
+  & Parentheses & No parentheses \\
+  \addlinespace
+  %% \midrule
+  Marks & $\pospl^1\reqpl^{\mathbb{N}}$ or $\pospl^0\reqpl^{\mathbb{Z}_{+}}$ & $\pospl^1\reqpl^{\mathbb{N}}$ \\
+  No marks & $\pospl^1\reqpl^{\mathbb{N}}$ or $\pospl^0\reqpl^{\mathbb{Z}_{+}}$ & $\pospl^1\reqpl^0$ \\
+\end{tabular}
+\end{center}
+
+\noindent The absence of both non-grouping parentheses and marks mean that every permissible occurrence of a pair of parentheses must permit zero or more pairs.
+Adding marks adds required pairs for some trees, turning each occurrence to one of $k$ or more pairs, for some non-negative integer $k$ determined by the tree.
+The left column is more complicated since non-grouping parentheses can introduce occurrences that require an exact number of parentheses, as well as all the complications of marks.
+
+The remainder of this section will be devoted to first showing that $\neg \exists t_1, t_2 \in L(T_D).\ \words(t_1) \subseteq \words(t_2)$ implies that $D$ is resolvably ambiguous if $D$ is in the lower-right quadrant, i.e., if it has no marks and no non-grouping parentheses (in Section~\ref{sec:static-proofs}), and then describing the automaton that forms the basis for our algorithm (Section~\ref{sec:lattice-vpl}).
+Section~\ref{sec:static-proofs} also shows that adding marks to a resolvably ambiguous language in the lower-right quadrant preserves resolvability, which gives us a conservative approach for languages in the top-right quadrant.
+Intuitively, if a language has marks, but they are just conveniences, then we can still detect that the language is resolvable.
+For example, most expression languages fall in this category (we can write any expression unambiguously by adding parentheses everywhere, even without any precedence or associativity), while lists in OCaml do not (no amount of parentheses will make \verb|'[1;2]'| look like a list of two elements without a mark).
+
+\subsection{The Road to Correctness} \label{sec:static-proofs} % TODO: probably want a better title here
+
+This section shows that $\neg \exists t_1, t_2 \in L(T_D).\ \words(t_1) \subseteq \words(t_2)$ implies that $D$ is resolvably ambiguous if $D$ has no non-grouping parentheses and no marks.
+These requirements can be written more formally as follows:
+
+\begin{enumerate}
+  \item $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.
+  \item Every non-terminal $N_m$ on the right hand side of a productin in $D$ has $m = \emptyset$.
+\end{enumerate}
+
+\noindent Together, these imply that the canonical linear encoding of a tree has no ``$\reqp{}$'' (required parentheses), only ``$\posp{}$'' (optional parentheses).
+
+Consider an arbitrary $t \in L(T_D)$ with canonical linear encoding $c$.
+Now replace each pair ``$\posp{}$'' with exactly one pair of parentheses, calling the result $w_\top$.
+We have $w_\top \in \words(t)$ by construction.
+Now consider the two possibilities of ambiguity for $w_\top$:
 
 \begin{description}
-\item[Version 1] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$ and no non-terminals appear marked, i.e., for all non-terminals $N_m$ appearing on the right-hand side of the labelled productions in $D$, we have $m = \emptyset$.
-\item[Version 2] $\T \cap \{$ \verb|'('|$,$ \verb|')'| $\} = \emptyset$.
-\item[Version 3] There are no constraints on $D$.
+  \item[$w_\top$ is ambiguous.] In this case, $\exists t'.\ t \neq t' \land t' \in \parse(w_\top)$.
+This means that we can take the canonical encoding of $t'$ (call it $c'$) and replace every ``$[]$'' with either one or zero parentheses and produce $w_\top$.
+But that also means that every word we construct from $c$ (by choosing some non-negative integer of parentheses for each ``$[]$'') we can construct the same word from $c'$ (by choosing the same number for corresponding parentheses, and zero for the others), i.e., $\words(t) \subseteq \words(t')$.
+  \item[$w_\top$ is unambiguous.] In this case $\words(t)$ has at least one unambiguous word ($w_\top$) which thus cannot be part of $\words(t')$ for some other $t' \in L(T_D)$, thus $\neg \exists t'.\ t \neq t' \land \words(t) \subseteq \words(t')$.
 \end{description}
 
-\noindent The first restriction in version 1 states that $D$ cannot contain parentheses.
-This implies that all parentheses present in $G'_D$ are grouping parentheses.
-The second restriction implies that no parentheses are required (as a counterexample, assuming normal precedence, the parentheses in ``$(1 + 2) * 3$'' are required and removing them would produce a different interpretation).
-
-The reason for this split stems from the following insight: double grouping parentheses do not matter, in the sense that they do not change the interpretation of the word, e.g., $\parse($\verb|'((1 + 2)) * 3'|$) = \parse($\verb|'(1 + 2) * 3'|$)$.
-Versions~1 and 2 have only grouping parentheses, meaning that no double parentheses matter.
-What follows is a brief outline of the remainder of this section, which details our static analysis for versions~1 and 2, building from that observation.
+\noindent We thus arrive at the first of two central theorems for our approach:
+
+\begin{theorem}
+  Given a language with productions $D$ and terminals $\T$, where $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$ and all non-terminals $N_m$ appearing in a right-hand side of a production in $D$ having $m = \emptyset$, the following holds: $\rho_P(D) <-> \neg \exists t_1, t_2 \in L(T_D).\ t_1 \neq t_2 \land \words(t_1) \subseteq \words(t_2)$.
+\end{theorem}
+
+Adding marks to such a language preserves resolvability:
 
 \begin{itemize}
-\item We can consider an alternate representation of words where parentheses are not explicitly part of the string of terminals, but are represented by an accompanying bag of ranges denoting which terminals are covered by parentheses.
-(Section~\ref{sec:word-view}).
-\item This alternate representation gives rise to a lattice per tree in $L(T_D)$, where a tree has only ambiguous words iff its lattice is entirely covered by the lattices of other trees (Section~\ref{sec:lattice}).
-\item These lattices can be encoded as words, leading to the construction of a visibly pushdown automaton that we can examine to determine the existence of a tree whose lattice is covered by the lattice of a \emph{single} other tree, which is sufficient for completeness in version 1 (Section~\ref{sec:lattice-vpl}).
+  \item Marks can introduce at most one required pair of parentheses per node.
+  \item Adding marks to a language can only ever shrink $\words(t)$ for any particular $t \in L(T_D)$.
+  \item Given a tree $t$, the word produced by putting one pair of parentheses around each node in $t$ (potentially adding double parentheses), call it $w$, is unambiguous in the language without marks.
+It cannot be excluded from $\words(t)$ by adding marks (since it already has a pair of parentheses around each node), and it also cannot be ambiguous, since that would require $\words(t')$ for some other tree $t'$ to have grown.
+Thus, $\parse(w) = \{t\}$ even when we add marks to $D$.
 \end{itemize}
 
-\subsection{An Alternative Word View} \label{sec:word-view}
-
-This section introduces an alternative (isomorphic) definition of a word, heavily used in sections~\ref{sec:lattice} and \ref{sec:lattice-vpl}.
-The method by which we generate $G_D$ and $G'_D$ limits the possible differences between them significantly.
-In particular, no new terminals are introduced, except '(' and ')', and they are always introduced in a well-balanced fashion.
-
-If we thus delimit ourselves to only consider languages where words have no unbalanced parentheses\footnote{I.e., the vast, vast majority of programming languages currently in use.} we can give the following alternative definition of a word: a word is a two-tuple containing a sequence of non-parenthesis terminals and a bag (or multiset) of ranges covered by parentheses.
-For example, the word ``\verb|(1 + 2) * 3|'' is equivalent to $(\text{``}\verb|1 + 2 * 3|\text{''}, \{\range{0}{3}\})$, while ``\verb|((1 + 2)) * 3|'' is equivalent to $(\text{``}\verb|1 + 2 * 3|\text{''}, \{\range{0}{3}, \range{0}{3}\})$.
-We will refer to the first component of the tuple as the \emph{basic word} and the second as the \emph{range bag}.
-
-However, there is a case where this alternate representation is not fully isomorphic; namely when a pair of parentheses would yield a zero-length range.
-As an example, consider the two words ``\verb|(())|'' and ``\verb|()()|''.
-Both of these have an alternate representation of $(\text{``}\text{''}, \{\range{0}{0}, \range{0}{0}\})$.
-In versions~1 and 2 this corresponds to grouping parentheses surrounding zero-length productions.
-We can avoid this case by finding every production $N -> l : r$ where $\epsilon \in L(r)$, replacing it with $N -> l : r'$ where $L(r') = L(r) \setminus \{\epsilon\}$, and replacing every use of $N$ with $N + \epsilon$.
-The one exception is if the start symbol has a nullable production, but that can be trivially special cased.
-Thus, we will assume no zero-length grouping parentheses in the following.
-
-We denote the functions producing the two components by $\basic$ and $\rangebag$, i.e., \\$\basic(\text{``}\verb|((1 + 2)) * 3|\text{''}) = \text{``}\verb|1 + 2 * 3|\text{''}$ and $\rangebag(\text{``}\verb|((1 + 2)) * 3|\text{''}) = \{\range{0}{3}, \range{0}{3}\}$.
-
-The next four lemmas, all concerning the words in $\words(t)$ for some given $t \in L(G_t)$, form the basis of the lattices mentioned at the end of the previous section.
-
-\begin{lemma}
-  $\forall w_1, w_2 \in \words(t).\ \basic(w_1) = \basic(w_2)$.
-  \label{lemma:same-basic}
-\end{lemma}
-\noindent More informally, all $w \in \words(t)$ have the same basic word.
-Intuitively, $\words$ first applies the ``inverse'' of $\semantic$, i.e., adding some number of grouping parentheses nodes between pre-existing nodes, then $\yield$, which flattens the tree to a word.
-The only terminals changed in this process are parentheses, i.e., different $w$ can only differ in their range bags.
-This also implies that two trees that share a word (i.e., trees $t_1$ and $t_2$ that have a word $w$ such that $\parse(w) \supseteq \{t_1, t_2\}$) must also share a basic word.
-
-\begin{lemma}
-  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ R \subseteq \support(\rangebag(w))$
-  \label{lemma:required-parentheses}
-\end{lemma}
-\noindent Recall that $\support(B)$ is the support of the bag $B$, i.e., the underlying set of elements in $B$.
-More informally, some parentheses are required.
-For example, removing the parentheses in ``\verb|(1 + 2) * 3|'' changes the produced parse trees.
-Each required range is a direct consequence of a mark on a non-terminal in $D$.
-We will write $\botw(t)$ for the word implied by the first quantified expression.
-It is unique, since its basic word is fixed by $t$, and its rangebag is a set.
-
-\begin{lemma}
-  There exists a set $R$ such that $\exists w \in \words(t).\ \rangebag(w) = R$ and $\forall w \in \words(t).\\ \support(\rangebag(w)) \subseteq R$.
-  \label{lemma:possible-parentheses}
-\end{lemma}
-\noindent More informally, there is a finite set of possible parentheses.
-As mentioned in Section~\ref{sec:parse-time-disambiguation}, grouping parentheses can only be added if they exactly cover a production, which amounts to an internal node in $t$, and each tree has a finite amount of nodes.
-We will write $\topw(t)$ for the word implied by the first quantified expression, which is also unique.
-
-\begin{lemma}
-  $\forall w_1, w_2 \in \words(t).\ \support(\rangebag(w_1)) = \support(\rangebag(w_2)) => \parse(w_1) = \parse(w_2)$.
-  \label{lemma:rangeset-equality}
-\end{lemma}
-\noindent More informally, duplicated parentheses do not matter.
-We first note that in version 1 and 2 there are no parentheses in $D$, i.e., all parentheses in $G'_D$ are introduced as grouping parentheses.
-Each pair of parentheses thus corresponds to a single grouping node $g$ in a tree in $L(T'_D)$.
-Duplicated grouping parentheses correspond to nested grouping nodes, all of which are removed by $\semantic$, thus producing identical sets of trees in $L(T_D)$.
-For example, \verb|((1 + 2)) * 3| has the same interpretations as \verb|(1 + 2) * 3|.
-
-\subsection{A Lattice of Word Partitions} \label{sec:lattice}
-
-\begin{figure}[t]
-  \begin{tikzpicture}[every node/.style={rectangle,draw}]
-    \node (a) [draw=white] at (0,5) {$((1) + (2)) + (3)$};
-
-    \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
-    \node (b2) [draw=white] at (-1.2,4.2) {$((1) + (2)) + 3$};
-    \node (b3) [draw=white] at ( 1.2,4.2) {$((1) + 2) + (3)$};
-    \node (b4) [draw=white] at ( 3.6,4.2) {$(1 + (2)) + (3)$};
-
-    \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
-    \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
-    \node (c3) [draw=white] at (-1.1,3) {$((1) + 2) + 3$};
-    \node (c4) at (1.1,3) {$1 + (2) + (3)$};
-    \node (c5) [draw=white] at (3.3,3) {$(1 + (2)) + 3$};
-    \node (c6) [draw=white] at (5.5,3) {$(1 + 2) + (3)$};
-
-    \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
-    \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
-    \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
-    \node (d4) [draw=white] at ( 3.6,1.8) {$(1 + 2) + 3$};
-
-    \node (e) at (0, 1) {$1 + 2 + 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1.90) -- (c1.270);
-  \draw (d1.90) -- (c2.270);
-  \draw (d1.90) -- (c3.270);
-  \draw (d2.90) -- (c1.270);
-  \draw (d2.90) -- (c4.270);
-  \draw (d2.90) -- (c5.270);
-  \draw (d3.90) -- (c2.270);
-  \draw (d3.90) -- (c4.270);
-  \draw (d3.90) -- (c6.270);
-  \draw (d4.90) -- (c3.270);
-  \draw (d4.90) -- (c5.270);
-  \draw (d4.90) -- (c6.270);
-
-  \draw (c1.90) -- (b1.270);
-  \draw (c1.90) -- (b2.270);
-  \draw (c2.90) -- (b1.270);
-  \draw (c2.90) -- (b3.270);
-  \draw (c3.90) -- (b2.270);
-  \draw (c3.90) -- (b3.270);
-  \draw (c4.90) -- (b1.270);
-  \draw (c4.90) -- (b4.270);
-  \draw (c5.90) -- (b2.270);
-  \draw (c5.90) -- (b4.270);
-  \draw (c6.90) -- (b3.270);
-  \draw (c6.90) -- (b4.270);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-    %%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-  \end{tikzpicture}
-  \caption{The lattice of words for the tree $a(a(n(1)+n(2))+n(3))$. The boxed words are shared with the lattice in Fig.~\ref{fig:lattice2}.}
-  \label{fig:lattice1}
-\end{figure}
-
-Lemma~\ref{lemma:rangeset-equality} suggests a partition of the words in $\words(t)$ for any given $t \in L(T_D)$; group words $w$ by their \emph{rangeset}, where $\rangeset(w) = \support(\rangebag(w))$.
-These partitions can be partially ordered by subset on the rangeset, resulting in a lattice of word partitions per tree.
-This lattice is bounded, with top and bottom elements $\topw(t)$ and $\botw(t)$.
-We denote this lattice by $\lattice(t)$.
-Lemma~\ref{lemma:same-basic} further states that all words in $\words(t)$ share the same basic word.
-For example, Fig.~\ref{fig:lattice1} contains the lattice for the tree $a(a(n(1)+n(2))+n(3))$ (from our running example, defined in Fig.~\ref{fig:running-example-definition} on page~\pageref{fig:running-example-definition}).
-Each partition is represented by the word whose rangebag is a set.
-To reduce clutter, we do not draw the partitions that have grouping parentheses around the entire word.
-This outermost possible pair is uninteresting since it is always allowed, and would double the size of the figure if it was included.
-
-\begin{figure}[t]
-
-\begin{tikzpicture}[every node/.style={rectangle,draw}]
-  \node (a) [draw=white] at (0,5) {$(1) + ((2) + (3))$};
-
-  \node (b1) at (-3.6,4.2) {$(1) + (2) + (3)$};
-  \node (b2) [draw=white] at (-1.2,4.2) {$(1) + ((2) + 3)$};
-  \node (b3) [draw=white] at ( 1.2,4.2) {$(1) + (2 + (3))$};
-  \node (b4) [draw=white] at ( 3.6,4.2) {$1 + ((2) + (3))$};
-
-  \node (c1) at (-5.5,3) {$(1) + (2) + 3$};
-  \node (c2) at (-3.3,3) {$(1) + 2 + (3)$};
-  \node (c3) [draw=white] at (-1.1,3) {$(1) + (2 + 3)$};
-  \node (c4) at (1.1,3) {$1 + (2) + (3)$};
-  \node (c5) [draw=white] at (3.3,3) {$1 + ((2) + 3)$};
-  \node (c6) [draw=white] at (5.5,3) {$1 + (2 + (3))$};
-
-  \node (d1) at (-3.6,1.8) {$(1) + 2 + 3$};
-  \node (d2) at (-1.2,1.8) {$1 + (2) + 3$};
-  \node (d3) at ( 1.2,1.8) {$1 + 2 + (3)$};
-  \node (d4) [draw=white] at ( 3.6,1.8) {$1 + (2 + 3)$};
-
-  \node (e) at (0, 1) {$1 + 2 + 3$};
-
-  \draw (e) -- (d1);
-  \draw (e) -- (d2);
-  \draw (e) -- (d3);
-  \draw (e) -- (d4);
-
-  \draw (d1.90) -- (c1.270);
-  \draw (d1.90) -- (c2.270);
-  \draw (d1.90) -- (c3.270);
-  \draw (d2.90) -- (c1.270);
-  \draw (d2.90) -- (c4.270);
-  \draw (d2.90) -- (c5.270);
-  \draw (d3.90) -- (c2.270);
-  \draw (d3.90) -- (c4.270);
-  \draw (d3.90) -- (c6.270);
-  \draw (d4.90) -- (c3.270);
-  \draw (d4.90) -- (c5.270);
-  \draw (d4.90) -- (c6.270);
-
-  \draw (c1.90) -- (b1.270);
-  \draw (c1.90) -- (b2.270);
-  \draw (c2.90) -- (b1.270);
-  \draw (c2.90) -- (b3.270);
-  \draw (c3.90) -- (b2.270);
-  \draw (c3.90) -- (b3.270);
-  \draw (c4.90) -- (b1.270);
-  \draw (c4.90) -- (b4.270);
-  \draw (c5.90) -- (b2.270);
-  \draw (c5.90) -- (b4.270);
-  \draw (c6.90) -- (b3.270);
-  \draw (c6.90) -- (b4.270);
-
-  \draw (b1) -- (a);
-  \draw (b2) -- (a);
-  \draw (b3) -- (a);
-  \draw (b4) -- (a);
-%%   \draw[preaction={draw=white, -,line width=6pt}] (a) -- (e) -- (c);
-\end{tikzpicture}
-\caption{The lattice of words for the tree $a(n(1)+a(n(2)+n(3)))$. The boxed words are shared with the lattice in Fig.~\ref{fig:lattice1}.}
-\label{fig:lattice2}
-\end{figure}
-
-To show the connection between resolvable ambiguity and these lattices, consider the word ``\verb|1 + 2 + 3|''.
-It has two interpretations in our running example, $a(a(n(1)+n(2))+n(3))$, which we would normally write as \verb|(1 + 2) + 3|, and $a(n(1)+a(n(2)+n(3)))$, which we would normally write \verb|1 + (2 + 3)|.
-The lattices for these two trees are given in Figures~\ref{fig:lattice1} and \ref{fig:lattice2} respectively.
-The partitions that appear in both lattices are represented as boxed words, the others are unboxed.
-These shared partitions represent words that are ambiguous between these particular trees.
-Finding an unambiguous word is thus the same as finding a partition that is not shared with any other tree.
-In this particular case, there is no ambiguity with any other tree at all, and so the unboxed words are all valid resolutions of the ambiguity.
-
-At this point it is clear that a given tree has an unambiguous word iff its lattice has at least one partition that is not shared with any other tree.
-
-Next, we note that each lattice is uniquely determined by its top and bottom elements; it contains all elements between them:
-
-\begin{lemma}
-  Given $t \in L(T_D)$, we have that:\\
-  $\forall w.\ \rangebag(\botw(t)) \subseteq \rangeset(w) \subseteq \rangebag(\topw(t)) => w \in \words(t)$. \label{lemma:top-bottom-determine}
-\end{lemma}
-
-\noindent Recall that $\rangebag = \rangeset$ for $\botw(t)$ and $\topw(t)$.
-Finally, we show the difference between versions~1 and 2 in the lattice setting: in version~1, the rangeset of the bottom word is the empty set:
-
-\begin{lemma}
-  In version~1, $\forall t \in L(T_D).\ \rangeset(\botw(t)) = \emptyset$. \label{lemma:version1-bot}
-\end{lemma}
-
-\noindent Since version~1 has no markings there are no required parentheses, thus the bottom word has no parentheses.
-
-Our approach centers around finding a pair of lattices, such that one is entirely contained in the other.
-Since the former shares \emph{all} partitions with the latter, it has only ambiguous words, thus any of those words will be unresolvably ambiguous.
-
 \begin{theorem}
-  In versions~1 and 2, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, the existence of a tree $t' \in L(T_D)$ such that $\basic(w_\bot) = \basic(\botw(t')) \land \rangeset(w_\bot) \subseteq \rangeset(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$ implies $\neg \exists w.\ \parse(w) = \{t\}$.
-\label{lemma:lattice-containment-implies-unresolvable}
+  Adding marks to a resolvably ambiguous language $D$ preserves resolvability, if $D$ had no parentheses and no marks.
 \end{theorem}
 
-\noindent For version~1, \emph{every} unresolvable ambiguity has this form:
-
-\begin{theorem}
-  In version 1, given a $t \in L(T_D)$, $w_\bot = \botw(t)$, and $w_\top = \topw(t)$, $\exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\botw(t')) \subseteq \rangeset(w_\bot) \land \rangeset(w_\top) \subseteq \rangeset(\topw(t')) <=> \neg \exists w.\ \parse(w) = \{t\}$.
-\end{theorem}
-
-\begin{proof}
-  The left-to-right implication is already given in Theorem~\ref{lemma:lattice-containment-implies-unresolvable}, and Lemma~\ref{lemma:version1-bot} together with $\basic(w_\bot) = \basic(\botw(t'))$ implies that $w_\bot = \botw(t')$.
-We thus need to show that $\neg \exists w.\ \parse(w) = \{t\} => \exists t' \in L(T_D).\ \basic(w_\bot) = \basic(\botw(t')) \land \rangeset(\topw(t')) \subseteq \rangeset(w_\top)$.
-
-  If all words in $\words(t)$ are ambiguous, then $\exists t' \in L(T_D).\ t' \neq t \land t' \in \parse(w_\top)$.
-Lemma~\ref{lemma:same-basic} gives $\basic(w_\bot) = \basic(\botw(t'))$.
-The definition of $\topw(t')$ (Lemma~\ref{lemma:possible-parentheses}) gives $\rangeset(w_\top) \subseteq \rangebag(\topw(t'))$, but since the rangebag of $\topw(t')$ is a set, this also implies that $\rangeset(w_\top) \subseteq \rangeset(\topw(t'))$.
-\end{proof}
-
-\subsection{A Lattice as a Word, and an Algorithm} \label{sec:lattice-vpl}
-
-This section introduces a linear encoding of the lattices of the previous section as words.
-Lemmas~\ref{lemma:same-basic} and \ref{lemma:top-bottom-determine} imply that a lattice encoding requires three things: a basic word, a set of \emph{required} parentheses, and a set of \emph{possible} parentheses.
-We encode required parentheses with ``$\reqp{}$'' and possible parentheses with ``$\posp{}$''.
-For example, the lattice in Fig.~\ref{fig:lattice2} is represented by ``$\posp{\posp{1} + \posp{\posp{2} + \posp{3}}}$'', while the tree for ``$(1 + 2) * 3$'' has a lattice represented by ``$\posp{\reqp{\posp{1} + \posp{2}} * \posp{3}}$''.
-We denote the lattice encoded by such a word by $\lattice(w)$.
-With this encoding we can apply the formidable body of knowledge available for word languages, albeit with some extra care; the linear encoding of a lattice is not unique.
-
-For example, ``$\posp{\posp{1}}$'' and ``$\posp{1}$'' represent the same lattice, as do ``$\reqp{\posp{2}}$'' and ``$\reqp{2}$''.
-To gain uniqueness we forbid duplicated parentheses and prioritize required parentheses over possible parentheses.
-For example, ``$\posp{\posp{1}}$'' is uniquely represented as ``$\posp{1}$'' while ``$\reqp{\posp{2}}$'' is uniquely represented as ``$\reqp{2}$''.
-We define a linear encoding of this form to be \emph{well-formed}.
-
-Lattice equality is now equivalent with equality of the linear encoding.
-To determine if a lattice is entirely contained in another, we make the following observation: we can move the top of the lattice ``up'' (new top rangeset is a superset) by adding a pair of possible parentheses, and move the bottom ``down'' (new bottom rangeset is a subset) by replacing a required pair with a possible pair.
-For example, ``$\reqp{1}2$'' is entirely contained in ``$\reqp{1}\posp{2}$'', which is entirely contained in ``$\posp{1}\posp{2}$''.
-
-The centerpiece of our algorithm is a \emph{visibly pushdown automaton} constructed in such a way that
-
-\begin{itemize}
-\item there is a bijection between trees in $L(T_D)$ and successful runs; and
-\item the word recognized by a successful run is the linear encoding of the corresponding tree's lattice.
-\end{itemize}
-
-\noindent Two distinct successful runs through this automaton that recognize the same word thus imply two distinct trees that have exactly the same lattice.
-To detect lattices contained in each other, we create a modified copy, where the copy may add arbitrary (balanced, well-nested) possible parentheses, and replace any required pair of parentheses with a possible pair.
-The copy maintains much of the structure of the original, and in particular, keeps the connection to trees in $L(T_D)$ (though it is no longer a bijection: there are multiple successful runs per tree since there are multiple larger lattices).
-Two distinct successful runs, one in each automaton, that recognize the same word then imply two distinct trees where the lattice of one is entirely contained in the other.
-
-\subsubsection{Example for Simplified Construction}
-
-We will now walk through the construction of this automaton, using the following language definition:
+\subsection{The Algorithm} \label{sec:lattice-vpl}
+
+This section describes the construction of a visibly pushdown automaton that we examine to determine if a language has a pair of trees where the language of one entirely contains the other.
+We begin by outlining the approach and then describe it in more detail.
+
+The automaton we examine recognizes canonical linear encodings and has a one-to-one correspondence between runs and trees.
+Two distinct runs that recognize the same word thus corresponds to two distinct trees that have the \emph{same} word language.
+This is essentially the question of whether a visibly pushdown automaton is ambiguous, which we can detect by constructing a product automaton and trimming it.
+
+To detect subsumption we make the following observation: given two trees $t_1$ and $t_2$ with canonical encodings $c_1$ and $c_2$, respectively, we have $\words(t_1) \subset \words(t_2)$ iff we can add arbitrary well-nested ``$\posp{}$'' pairs to $c_1$, producing a $c_1'$ such that $c_1' = c_2$ (Note that we do not need to consider ``$\reqp{}$'' since no such pairs appear in this language class).
+We thus make a product automaton of two slightly different automata, where the second may add arbitrary ``$\posp{}$'', but is otherwise identical.
+
+\subsubsection{Illustration by Example}
+
+We begin by constructing a visibly pushdown automaton that recognizes \emph{a} linear encoding for each tree, then modify it to only recognize the \emph{canonical} linear encoding.
+To illustrate the approach we consider the following language:
 
 \begin{center}
-\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
+  \begin{tabular}{@{}r@{$\;->\;$}c@{\,:\;}l}
     \toprule
-  $E$ & $s$ & $E_{\{s\}}$ \verb|';'| $E$ \\
-  $E$ & $v$ & \verb|'e'| \\
+    $S$ & $a$ & $E(\verb|';'| + \epsilon)$ \\
+    $E$ & $b$ & $E\verb|'s'|$ \\
+    $E$ & $c$ & \verb|'z'| \\
     \bottomrule
@@
 
-\noindent We first present a simplified construction that assumes no production can match a single non-terminal, i.e., for all right-hand side regular expressions $r$, $\neg \exists N_m.\ N_m \in L(r)$, to make the base idea easier to follow.
-Such cases introduce duplicated parentheses if handled in a naive way, and the extra book-keeping required to correctly handle them complicate this presentation, and will so be introduced after the more naive method.
-
-Conceptually, each non-terminal represents a choice of which production should replace it, and whether that production requires parentheses around it.
-We now make this explicit, to ensure that standard operations on finite automata produce the correct result.
-We replace each $N_m$ with a sum of the labels of productions in $N$, where each label has a subscript $\top$ or $\bot$ if parentheses are required or possible, respectively.
-For example, $E_{\{s\}}$ is replaced with $s_\top + v_\bot$.
-
-\begin{center}
-\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-  \toprule
-  $E$ & $s$ & $(s_\top + v_\bot)$ \verb|';'| $(s_\bot + v_\bot)$ \\
-  $E$ & $v$ & \verb|'e'| \\
-  \bottomrule
-\end{tabular}
-\end{center}
-
-\noindent Next we construct a DFA per production. This can be done in the standard way by constructing an NFA, then determinizing it, and optionally minimizing it.
+\noindent This language has no non-grouping parentheses and no marks and is thus in the language class we consider.
+We begin by constructing a DFA per production (using standard methods, since the right hand side is a regular expression).
 
@@
     scale = 1,transform shape]
-
-  \node[state,initial] (1) {$1$};
+  \node[state] (1) {$1$};
   \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-  \node[state,accepting] (4) [right of=3] {$4$};
-
-  \node[state,initial] (5) [below of=2] {$5$};
-  \node[state,accepting] (6) [right of=5] {$6$};
-
-  \path (1) edge[bend left] node {$s_\top$} (2)
-  (1) edge[bend right] node {$v_\bot$} (2)
-  (2) edge node {$';'$} (3)
-  (3) edge[bend left] node {$s_\bot$} (4)
-  (3) edge[bend right] node {$v_\bot$} (4);
-  \path (5) edge node {$'e'$} (6);
+
+  \node[state] (3) [below of=1,xshift=-1.2cm] {$3$};
+  \node[state] (4) [right of=3] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+
+  \node[state] (6) [below of=1,yshift=-2cm] {$6$};
+  \node[state] (7) [right of=6] {$7$};
+
+  \node[state] (8) [below of=6] {$8$};
+  \node[state] (9) [right of=8] {$9$};
+
+  \node[state,initial] (s) [left of=1,xshift=-1cm] {$s$};
+  \node[state,accepting] (f) [right of=2,xshift=1cm] {$f$};
+
+  \path (1) edge node[left]{$'\pospl', +(1, 2)$} (3)
+  (5) edge[right] node{$'\pospr', -(1, 2)$} (2);
+  \path (1) edge[out=180,in=180,looseness=2,left] node{$'\pospl', +(1, 2)$} (6)
+  (7) edge[out=0,in=0,looseness=2,right] node{$'\pospr', -(1, 2)$} (2);
+
+  \path (3) edge node{$'+'$} (4);
+  \path (4) edge[bend right,above,out=270,in=225] node[xshift=5mm]{$'\pospl', +(4, 5)$} (3)
+  (5) edge[loop right,below] node[yshift=-3.5mm]{$'\pospr', -(4, 5)$} (5);
+  \path (4) edge[left] node{$'\pospl', +(4, 5)$} (6)
+  (7) edge node{$'\pospr', -(4, 5)$} (5);
+
+  \path (6) edge[left] node{$'\pospl', +(6, 7)$} (8)
+  (9) edge[right] node{$'\pospr', -(6, 7)$} (7);
+
+  \path (8) edge[above] node{$'n'$} (9);
+
+  \path (s) edge[bend left] node{$'\pospl', +(s, f)$} (1)
+  (2) edge[bend left] node{$'\pospr', -(s, f)$} (f);
 
@@
 
-\noindent We then combine the DFAs into a single visibly pushdown automaton.
-Every time we transition from a production to a child we push a symbol to the stack, which we then pop when we return.
-We use $Q \times Q$ as our stack alphabet, to record the transition we are replacing in the parent.
-We thus replace every edge $p \xrightarrow{l_\top} q$, where $l \in \Labels$, with:
-
+\noindent We then combine these disparate DFAs to produce a single visibly pushdown automaton in three steps:
+
+\begin{enumerate}
+\item Introduce two new distinct states (call them $s$ and $f$) and a transition $s \xrightarrow{S} f$, where $S$ is the starting symbol of the language being examined.
+\item Replace each transition $p \xrightarrow{N} q$ with:
   \begin{itemize}
-  \item an edge $p \xrightarrow{'\reqpl', +(p, q)} p'$, where $p'$ is the initial state in the DFA corresponding to the production with label $l$, and
-  \item an edge $q' \xrightarrow{'\reqpr', -(p, q)} q$ for every final state $q'$ in the DFA corresponding to the production with label $l$.
+  \item A transition $p \xrightarrow{'\pospl', +(p,q)} q_0$ for every initial state $q_0$ in a DFA corresponding to a production with lefthand side $N$.
+  \item A transition $q_f \xrightarrow{'\pospr', -(p,q)} q$ for every accepting state $q_f$ in a DFA corresponding to a production with lefthand side $N$.
   \end{itemize}
-
-\noindent Similarly, for every edge $p \xrightarrow{l_\bot} q$, where $l \in \Labels$, we add edges $p \xrightarrow{'\pospl', +(p, q)} p'$ and $q' \xrightarrow{'\pospr', -(p, q)} q$.
-Note the difference: for $l_\top$ we add a required parenthesis ``$\reqpl$'', while $l_\bot$ adds a possible parenthesis ``$\pospl$''.
-
-Intuitively, we parse a child node with parentheses around it, then return. This is where our simplification is used, without it we might introduce double parentheses here.
+\item Make $s$ the only initial state, and $f$ the only accepting state.
+\end{enumerate}
+
+\noindent To make the figure less cluttered we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')}$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')}$.
 
@@
     scale = 1,transform shape]
-
-  \node[state,initial] (1) {$1$};
+  \node[state] (1) {$1$};
   \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-  \node[state,accepting] (4) [right of=3] {$4$};
-
-  \node[state,initial] (5) [below of=2] {$5$};
-  \node[state,accepting] (6) [right of=5] {$6$};
-
-  \path (2) edge[below] node {$';'$} (3);
-  \path (5) edge[below] node {$'e'$} (6);
-
-  \path (1) edge[loop above] node{$'\reqpl', +(1, 2)$} (1)
-  (4) edge[bend right, above] node{$'\reqpr', -(1, 2)$} (2);
-  \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
-  (6) edge[right] node[xshift=5mm,yshift=-3mm]{$'\pospr', -(1, 2)$} (2);
-  \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
-  (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
-  \path (3) edge[right] node[xshift=5mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
-  (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
+
+  \node[state] (3) [below of=1,xshift=-1.2cm] {$3$};
+  \node[state] (4) [right of=3] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+
+  \node[state] (6) [below of=1,yshift=-2cm] {$6$};
+  \node[state] (7) [right of=6] {$7$};
+
+  \node[state] (8) [below of=6] {$8$};
+  \node[state] (9) [right of=8] {$9$};
+
+  \node[state,initial] (s) [left of=1,xshift=-1cm] {$s$};
+  \node[state,accepting] (f) [right of=2,xshift=1cm] {$f$};
+
+  \path (1) edge node[left]{$'\pospl, +(1, 2)$} (3)
+  (5) edge[right] node{$'\pospr', -(1, 2)$} (2);
+  \path (1) edge[out=180,in=180,looseness=2,left] node{$'\pospl', +(1, 2)$} (6)
+  (7) edge[out=0,in=0,looseness=2,right] node{$'\pospr', -(1, 2)$} (2);
+
+  \path (3) edge node{$'+'$} (4);
+  \path (4) edge[bend right,above,out=270,in=225] node[xshift=5mm]{$'\pospl', +(4, 5)$} (3)
+  (5) edge[loop right,below] node[yshift=-3.5mm]{$'\pospr', -(4, 5)$} (5);
+  \path (4) edge[left] node{$'\pospl', +(4, 5)$} (6)
+  (7) edge node{$'\pospr', -(4, 5)$} (5);
+
+  \path (6) edge[left] node{$'\pospl', +(6, 7)$} (8)
+  (9) edge[right] node{$'\pospr', -(6, 7)$} (7);
+
+  \path (8) edge[above] node{$'n'$} (9);
+
+  \path (s) edge[bend left] node{$'\pospl', +(s, f)$} (1)
+  (2) edge[bend left] node{$'\pospr', -(s, f)$} (f);
 
@@
 
-\noindent Finally, we add a new initial state and a new final state, and connect them to the initial and final states belonging to the starting non-terminal:
+\noindent This automaton recognizes a linear encoding by simply putting a pair $'\posp{}'$ around each production.
+That encoding will be canonical in many cases (e.g.
+``$\posp{\posp{\posp{z}s};}$''), but not all; ``$\posp{\posp{z}}$'', which is recognized for the tree $a(c(z))$, is not canonical, it should be ``$\posp{z}$''.
+The extra $'\posp{}'$ pair stems from $S -> a: E(\verb|';'| + \epsilon)$, its righthand side matches $E$ (i.e., a word consisting of a single non-terminal).
+
+We solve this by first recording which productions have righthand sides that match single non-terminal words (and which non-terminal) and then compute DFAs that do \emph{not} match such words.
+For our running example we note that the production $a$ matches the word $E$ (we record this as a tuple $(S, a, E)$) and replace its automaton with one that does not match $E$.
+In this case it is sufficient to mark state $2$ as non-accepting.
+
+The recorded tuple $(S, a, E)$ can be read as ``starting from non-terminal $S$ we can match the word $E$ by choosing the production with label $a$.'' If we consider the middle element as a sequence of labels we can build a relation by adding $(N, \epsilon, N)$ for all non-terminals $N$ and computing the closure of $(A, w, B) \cdot (B, w', C) = (A, w \cdot w', C)$.
+Call the resulting set $T$.
+
+Finally, we change step 2 in our construction to the following:
+
+\begin{enumerate}
+  \setcounter{enumi}{1}
+\item Replace each transition $p \xrightarrow{N} q$ with:
+  \begin{itemize}
+  \item A transition $p \xrightarrow{'\pospl', +(p,q,w)} q_0$, for every initial state $q_0$ in a DFA corresponding to a production with lefthand side $N'$, for every $(N, w, N') \in T$.
+  \item A transition $q_f \xrightarrow{'\pospr', -(p,q,w)} q$, for every accepting state $q_f$ in a DFA corresponding to a production with lefthand side $N'$, for every $(N, w, N') \in T$.
+  \end{itemize}
+\end{enumerate}
 
@@
     scale = 1,transform shape]
-
-  \node[state,initial] (7) [left of=5, xshift=-1.5cm] {$7$};
-  \node[state,accepting] (8) [right of=6, xshift=1.5cm] {$8$};
-
   \node[state] (1) {$1$};
   \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
+
+  \node[state] (3) [below of=1,xshift=-1.2cm] {$3$};
   \node[state] (4) [right of=3] {$4$};
-
-  \node[state] (5) [below of=2] {$5$};
-  \node[state] (6) [right of=5] {$6$};
-
-  \path (2) edge[below] node {$';'$} (3);
-  \path (5) edge[below] node {$'e'$} (6);
-
-  \path (1) edge[loop above] node{$'\reqpl', +(1, 2)$} (1)
-  (4) edge[bend right, above] node{$'\reqpr', -(1, 2)$} (2);
-  \path (1) edge[bend right, left] node{$'\pospl', +(1, 2)$} (5)
-  (6) edge[right] node[xshift=5mm,yshift=-3mm]{$'\pospr', -(1, 2)$} (2);
-  \path (3) edge[bend right, above] node[xshift=-1mm]{$'\pospl', +(3, 4)$} (1)
-  (4) edge[loop above] node{$'\pospr', -(3, 4)$} (4);
-  \path (3) edge[right] node[xshift=5mm,yshift=3mm]{$'\pospl', +(3, 4)$} (5)
-  (6) edge[bend right, right] node[xshift=1mm]{$'\pospr', -(3, 4)$} (4);
-
-  \path (7) edge[bend left, left] node{$'\pospl', +(7, 8)$} (1)
-  (7) edge[below] node{$'\pospl', +(7, 8)$} (5);
-  \path (4) edge[bend left, right] node{$'\pospr', -(7, 8)$} (8)
-  (6) edge[below] node{$'\pospr', -(7, 8)$} (8);
+  \node[state] (5) [right of=4] {$5$};
+
+  \node[state] (6) [below of=1,yshift=-2cm] {$6$};
+  \node[state] (7) [right of=6] {$7$};
+
+  \node[state] (8) [below of=6] {$8$};
+  \node[state] (9) [right of=8] {$9$};
+
+  \node[state,initial] (s) [left of=1,xshift=-1cm] {$s$};
+  \node[state,accepting] (f) [right of=2,xshift=1cm] {$f$};
+
+  \path (1) edge node[left]{$'\pospl, +(1, 2)$} (3)
+  (5) edge[right] node{$'\pospr', -(1, 2)$} (2);
+  \path (1) edge[out=180,in=180,looseness=2,left] node{$'\pospl', +(1, 2)$} (6)
+  (7) edge[out=0,in=0,looseness=2,right] node{$'\pospr', -(1, 2)$} (2);
+
+  \path (3) edge node{$'+'$} (4);
+  \path (4) edge[bend right,above,out=270,in=225] node[xshift=5mm]{$'\pospl', +(4, 5)$} (3)
+  (5) edge[loop right,below] node[yshift=-3.5mm]{$'\pospr', -(4, 5)$} (5);
+  \path (4) edge[left] node{$'\pospl', +(4, 5)$} (6)
+  (7) edge node{$'\pospr', -(4, 5)$} (5);
+
+  \path (6) edge[left] node{$'\pospl', +(6, 7)$} (8)
+  (9) edge[right] node{$'\pospr', -(6, 7)$} (7);
+
+  \path (8) edge[above] node{$'n'$} (9);
+
+  \path (s) edge[bend left] node{$'\pospl', +(s, f)$} (1)
+  (2) edge[bend left] node{$'\pospr', -(s, f)$} (f);
 
@@
 
-\noindent For example, the linearly encoded lattice $\posp{\posp{e};\posp{e}}$ is recognized by the run
-
-\begin{minipage}{.7\textwidth}
-$$
-\begin{array}{l@{\enspace}l@{\;}r@{\enspace}l@{\;}r@{\enspace}l@{\;}r}
-  (7, \epsilon) &\xrightarrow{\pospl} (1, &(7,8)) &\xrightarrow{\pospl} (5, &(1,2)\cdot(7,8)) &\xrightarrow{e} (6, &(1,2)\cdot(7,8)) \\
-  &\xrightarrow{\pospr} (2, &(7,8)) &\xrightarrow{;} (3, &(7,8)) &\xrightarrow{\pospl} (5, &(3,4)\cdot(7,8)) \\
-  &\xrightarrow{e} (6, &(3,4)\cdot(7,8)) &\xrightarrow{\pospr} (4, &(7, 8)) &\xrightarrow{\pospr} (8, &\epsilon) \\
-\end{array}
-$$
-\end{minipage}
-\begin{minipage}{.3\textwidth}
-\begin{center}
-{
-\newcommand{\terminal}[1]{\ \underline{#1}\ }
-\Tree [.{$s$}
-    [.{$v$} \terminal{e} ]
-    \terminal{;}
-    [.{$v$} \terminal{e} ] ]
-}
-\end{center}
-\end{minipage}
-
-\noindent and corresponds to the tree on the right.
-
-\subsubsection{Example for Full Construction}
-
-We now illustrate the issue that our previous simplification avoids, and then modify the construction such that the simplification is no longer required. Consider the following language definition:
-
-\begin{center}
-\begin{tabular}{@{}l@{\quad$->$\quad}l@{ $:$\quad}l@{}}
-  \toprule
-  $A$ & $a$ & $B$ \\
-  $B$ & $b$ & $C$ $(\verb|';'| + \epsilon)$ \\
-  $C$ & $c$ & \verb|'e'| \\
-  \bottomrule
-\end{tabular}
-\end{center}
-
-\noindent Following the simplified construction we produce the following automaton:
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (8) [left of=1, xshift=-1cm] {$8$};
-  \node[state,accepting] (9) [right of=2, xshift=1cm] {$9$};
-
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-
-  \node[state] (3) [below of=1,yshift=3mm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=3,yshift=3mm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \path (4) edge[below] node {$';'$} (5);
-  \path (6) edge[below] node {$'e'$} (7);
-
-  \path (1) edge[left] node {$'\pospl', +(1,2)$} (3)
-  (4) edge node {$'\pospr', -(1,2)$} (2)
-  (5) edge[right] node {$'\pospr', -(1,2)$} (2);
-  \path (3) edge[left] node {$'\pospl', +(3, 4)$} (6)
-  (7) edge[left] node {$'\pospr', -(3, 4)$} (4);
-
-  \path (8) edge node {$'\pospl', +(8, 9)$} (1)
-  (2) edge node {$'\pospr', -(8, 9)$} (9);
-
-\end{tikzpicture}
-\end{center}
-
-\noindent Unfortunately, this automaton recognizes the word $(((e)))$,
-which is not the unique linear encoding we require; it should be
-$(e)$. The reason for this mismatch is that each of the productions
-with labels $a$ and $b$ introduces an additional pair of parentheses
-in the word recognized by the automaton. To ensure that the automaton
-instead recognizes the required unique linear encoding $(e)$, we
-extend the construction of the automaton such that ``passing through''
-a non-terminal without recognizing any terminals does \emph{not} add
-parentheses. The idea is to skip the DFAs of such non-terminals, but
-still keep track of the skipped production(s) in the symbols pushed to
-the stack; thus, runs which differ in skipped non-terminals are still
-distinguished.
-
-For our example, the resulting automaton looks as follows:
-
-\begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-
-  \node[state,initial] (8) [left of=1, xshift=-1cm] {$8$};
-  \node[state,accepting] (9) [right of=2, xshift=1cm] {$9$};
-
-  \node[state] (1) {$1$};
-
-  \node[state] (3) [below of=1,yshift=3mm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=3,yshift=3mm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \path (4) edge[below] node {$';'$} (5);
-  \path (6) edge[below] node {$'e'$} (7);
-
-  \path (3) edge[left] node {$'\pospl', +(3, 4, \epsilon)$} (6)
-  (7) edge[left] node {$'\pospr', -(3, 4, \epsilon)$} (4);
-
-  \path (8) edge node {$'\pospl', +(8, 9, \epsilon)$} (1)
-  (8) edge[right] node[xshift=1mm,yshift=1mm] {$'\pospl', +(8, 9, a)$} (3)
-  (5) edge[left] node {$'\pospr', -(8,9,a)$} (9)
-  (8) edge[left,out=270,in=180] node {$'\pospl', +(8, 9, ab)$} (6)
-  (7) edge[right,out=0,in=270] node {$'\pospr', -(8, 9, ab)$} (9);
-
-\end{tikzpicture}
-\end{center}
-
-Note that the transition from state 8 to state 6 pushes the symbol
-$(8, 9, ab)$ to the stack. This means (a) that we started the
-``replacement'' for the transition from 8 to 9, and (b) that we passed
-through the productions with labels $a$ and $b$ without consuming
-terminals, thus ending up in the starting state of the DFA
-corresponding to production $c$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-% FORMAL STUFF BARRIER %
-%%%%%%%%%%%%%%%%%%%%%%%%
+\noindent We call this automaton $A_{\posp{}}$.
+We also construct a modified automaton that recognizes linear encodings of larger word languages.
+We call this automaton $A'_{\posp{}}$ and obtain it by adding transitions $p \xrightarrow{'\pospl', +\gamma} p$ and $p \xrightarrow{'\pospr', -\gamma} p$ for every state $p$, where $\gamma$ is a new distinct stack symbol.
+Note that all these new transition push and pop the \emph{same} stack symbol, meaning that a new pair ``$\posp{}$'' can be started in any state and then closed in any state (not just the same state) as long as intermediate transitions leave the stack unchanged, i.e., only introduce zero or more well-balanced pairs.
 
@@
 We use $\labelnt(l)$ and $\labelregex(l)$ as the non-terminal and regex, respectively, of the production labelled $l$.
-We define the set $T : \Labels \times \Labels \times \{\top, \bot\} \times \Labels^{*}$ of transitions between productions inductively through:
+We define the set $T : \NT \times \Labels^{*} \times \NT$ of transitions between productions inductively through:
 
 \begin{itemize}
-\item $(l, l, \bot, \epsilon) \in T$ for all $l \in \Labels$.
-\item $(l, l', l' \in m, l) \in T$ for all $N_m \in L(\labelregex(l))$ where $\labelnt(l') = N$.
-\item $(l_1, l_3, r_1 \lor r_2, w_1 \cdot w_2) \in T$ for all $(l_1, l_2, r_1, w_1), (l_2, l_3, r_2, w_2) \in T$.
+\item $(N, \epsilon, N) \in T$ for all $N \in \NT$.
+\item $(N, l, N') \in T$ for all $N' \in L(\labelregex(l))$ where $\labelnt(l) = N$.
+\item $(N_1, w_1 \cdot w_2, N_3) \in T$ for all $(N_1, w_1, N_2), (N_2, w_2, N_3) \in T$.
 \end{itemize}
 
-\noindent The third component of each element denotes whether that transition introduces a required pair of parentheses ($\top$) or a possible pair ($\bot$).
-$T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
-Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces infinitely ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
-
-We define the translation from production to DFA through a function $\dfa$ from a label to a DFA with the alphabet $\T \cup (\Labels \times \{\top, \bot\})$:
+% TODO: footnote should probably be removed
+\noindent $T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
+Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces (infinitely) ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
+
+We define the translation from production to DFA through a function $\dfa$ from a label to a DFA with the alphabet $\T \cup \Labels$:
 
 \begin{description}
-\item[$\dfa(l)$:] Compute the regular language $L' := L(\labelregex(l)) \setminus \{N'_m \mid N' \in \NT, m \subseteq \Labels\}$.
-Then replace every marked non-terminal $N_m$ with the regular expression $(l_1, l_1 \in m) + (l_2, l_2 \in m) + \ldots + (l_n, l_n \in m)$, where $l_i$ are all the labels of productions $N -> l_i : r_i$.
-For example, $E_{\{a\}}$ in the running example would be replaced by $(l, \bot) + (a, \top) + (m, \bot) + (n, \bot)$.
-Then construct a DFA for this language.
+\item[$\dfa(l)$:] Compute the regular language $L' := L(\labelregex(l)) \setminus \NT$, then construct a DFA for this language.
 \end{description}
@@
 \noindent We now define the visibly pushdown automaton $A_{\posp{}}$.
-Given $\dfa(l) = (Q_l, \T \cup (\Labels \times \{\top, \bot\}), \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
+Given $\dfa(l) = (Q_l, \T \cup \NT, \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
 
@@
 \item $Q = \{s, f\} \cup \bigsqcup_l Q_l$.
-\item $\T' = \T \cup \{\reqpl, \pospl, \pospr\}$.
+\item $\T' = \T \cup \{'\pospl', '\pospr'\}$.
 \item $\Gamma = Q \times Q \times \Labels^{*}$.
@@
 
-    \delta(p, \text{``}\reqpl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, r_1 \lor r_2 \} \\
-
     \delta(p, \text{``}\pospl\text{''}, \lambda)
     & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, \neg (r_1 \lor r_2) \} \\
-
-    \delta(f_{l_3}, \text{``}\reqpr\text{''}, (p, q, w))
-    & \{ (q, \lambda)
-    & p \in Q_{l_1}, \delta_{l_1}(p, (l_2, r_1)) = q, (l_2, l_3, r_2, w) \in P, f_{l_3} \in F_{l_3} \} \\
-
-    \delta(s, \text{``}\reqpl\text{''}, \lambda)
-    & \{ (s_{l_2}, (s, f, w))
-    & \labelnt(l_1) = S, (l_1, l_2, \top, w) \in P \} \\
+    & p \in Q_{l_1}, \delta_{l_1}(p, l_2) = q, (l_2, l_3, w) \in T \} \\
 
@@
     & \{ (s_{l_2}, (s, f, w))
-    & \labelnt(l_1) = S, (l_1, l_2, \bot, w) \in P \} \\
-
-    \delta(f_{l_2}, \text{``}\reqpr\text{''}, (s, f, w))
-    & \{ (f, \lambda)
-    & nt(l_1) = S, (l_1, l_2, r, w) \in P, f_{l_2} \in F_{l_2}\} \\
+    & \labelnt(l_1) = S, (l_1, l_2, w) \in T \} \\
   \end{array}
@@
     & \{ (p, \lambda) \} \\
-
-    \delta''(p, \text{``}\pospl\text{''}, \lambda)
-    & \delta(p, \text{``}\reqpl\text{''}, \lambda) \\
   \end{array}
@@
 
-\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is given in Section~\ref{sec:preliminaries-vpls} (using the partitions $\T_c = \{ \reqpl, \pospl \}$, $\T_i = \T$, and $\T_r = \{\reqpr\}$).
+\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is given in Section~\ref{sec:preliminaries-vpls} (using the partitions $\T_c = \{ \pospl \}$, $\T_i = \T$, and $\T_r = \{\pospr\}$).
 If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
@@
 The syntax definition DSL builds on \emph{syncons}, introduced by
-\citet{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
+Palmkvist and Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
 production is defined separately from each other (one
@@
 syntax as described in chapters 7 and 8 (base language and
-extensions) of the OCaml reference manual~\cite{OCamlReference}.
+extensions) of the OCaml reference manual~\cite{leroyOCamlSystemRelease2018}.
 The syncon definition is just under 700 lines of code, and is
@@
 
-\citet{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
+Afroozeh et al.~\cite{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
 However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case involving deep precedence discussed in Section~\ref{sec:evaluation-ocaml}.
@@
 
-\citet{danielssonParsingMixfixOperators2011} give a method for specifying grammars for expressions containing mixfix operators.
+Danielsson and Norell~\cite{danielssonParsingMixfixOperators2011} give a method for specifying grammars for expressions containing mixfix operators.
 They allow non-transitive, non-total precedence, and ``feel that it is overly restrictive to require the grammar to be unambiguous.'' Similar to our approach, they do not reject ambiguous grammars, only ambiguous parses.
@@
 
-Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{vanwykCopper2007} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
+Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{vanwykContextawareScanningParsing2007} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
 For this paper, only the syntactic component \cite{schwerdfegerVerifiableCompositionDeterministic2009} of this analysis is relevant.
@@
 The detection of classical ambiguity in context-free grammars is undecidable in general \cite{cantorAmbiguityProblemBackus1962}, yet numerous heuristic approaches exist.
-Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using SAT-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of \citet{bastenAmbiguityDetectionProgramming2011}.
+Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using SAT-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of Basten~\cite{bastenAmbiguityDetectionProgramming2011}.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% %% Acknowledgments
-%% \begin{acks}                            %% acks environment is optional
-%%                                         %% contents suppressed with 'anonymous'
-%%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-%%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-%%   %% acknowledge financial support and will be used by metadata
-%%   %% extraction tools.
-%%   This material is based upon work supported by the
-%%   \grantsponsor{GS100000001}{National Science
-%%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%%   conclusions or recommendations expressed in this material are those
-%%   of the author and do not necessarily reflect the views of the
-%%   National Science Foundation.
-%% \end{acks}
-
 %% Bibliography
+\bibliographystyle{splncs04}
+
 \bibliography{All}
 
-%% %% Appendix
-%% \appendix
-%% \section{Appendix}
-
-%% Text of appendix \ldots
-
 \end{document}

### 2019-10-20 paper/paper.tex
EOD

@@
 \usepackage{tikz}
-\usetikzlibrary{arrows,automata}
+\usetikzlibrary{arrows.meta,automata,calc}
 \usepackage{tikz-qtree} % Used for syntax trees
@@
 
-\noindent The absence of both non-grouping parentheses and marks mean that every permissible occurrence of a pair of parentheses must permit zero or more pairs.
+\noindent The absence of both non-grouping parentheses and marks mean that every occurrence of a pair of parentheses must permit zero or more pairs.
 Adding marks adds required pairs for some trees, turning each occurrence to one of $k$ or more pairs, for some non-negative integer $k$ determined by the tree.
@@
 This means that we can take the canonical encoding of $t'$ (call it $c'$) and replace every ``$[]$'' with either one or zero parentheses and produce $w_\top$.
-But that also means that every word we construct from $c$ (by choosing some non-negative integer of parentheses for each ``$[]$'') we can construct the same word from $c'$ (by choosing the same number for corresponding parentheses, and zero for the others), i.e., $\words(t) \subseteq \words(t')$.
-  \item[$w_\top$ is unambiguous.] In this case $\words(t)$ has at least one unambiguous word ($w_\top$) which thus cannot be part of $\words(t')$ for some other $t' \in L(T_D)$, thus $\neg \exists t'.\ t \neq t' \land \words(t) \subseteq \words(t')$.
+But that also means that every word we construct from $c$ (by choosing some non-negative integer of parentheses for each ``$[]$'') can also be constructed from $c'$ (by choosing the same number for corresponding parentheses, and zero for the others), i.e., $\words(t) \subseteq \words(t')$.
+  \item[$w_\top$ is unambiguous.] In this case $\words(t)$ has at least one unambiguous word ($w_\top$) which thus cannot be part of $\words(t')$ for any other $t' \in L(T_D)$, thus $\neg \exists t'.\ t \neq t' \land \words(t) \subseteq \words(t')$.
 \end{description}
@@
 
-Adding marks to such a language preserves resolvability:
+Adding marks to such a language preserves resolvability, since:
 
@@
   \item Adding marks to a language can only ever shrink $\words(t)$ for any particular $t \in L(T_D)$.
-  \item Given a tree $t$, the word produced by putting one pair of parentheses around each node in $t$ (potentially adding double parentheses), call it $w$, is unambiguous in the language without marks.
+  \item Given a tree $t$, the word produced by putting one pair of parentheses around each node in $t$ (thus potentially adding double parentheses), call it $w$, is unambiguous in the language without marks.
 It cannot be excluded from $\words(t)$ by adding marks (since it already has a pair of parentheses around each node), and it also cannot be ambiguous, since that would require $\words(t')$ for some other tree $t'$ to have grown.
@@
 \begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
     scale = 1,transform shape]
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-
-  \node[state] (3) [below of=1,xshift=-1.2cm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
+  \node[state,initial] (1) {$1$};
+  \node[state,accepting] (2) [right of=1] {$2$};
+  \node[state,accepting] (3) [right of=2] {$3$};
+
+  \node[state,initial] (4) [below of=1,yshift=7mm] {$4$};
   \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=1,yshift=-2cm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \node[state] (8) [below of=6] {$8$};
-  \node[state] (9) [right of=8] {$9$};
-
-  \node[state,initial] (s) [left of=1,xshift=-1cm] {$s$};
-  \node[state,accepting] (f) [right of=2,xshift=1cm] {$f$};
-
-  \path (1) edge node[left]{$'\pospl', +(1, 2)$} (3)
-  (5) edge[right] node{$'\pospr', -(1, 2)$} (2);
-  \path (1) edge[out=180,in=180,looseness=2,left] node{$'\pospl', +(1, 2)$} (6)
-  (7) edge[out=0,in=0,looseness=2,right] node{$'\pospr', -(1, 2)$} (2);
-
-  \path (3) edge node{$'+'$} (4);
-  \path (4) edge[bend right,above,out=270,in=225] node[xshift=5mm]{$'\pospl', +(4, 5)$} (3)
-  (5) edge[loop right,below] node[yshift=-3.5mm]{$'\pospr', -(4, 5)$} (5);
-  \path (4) edge[left] node{$'\pospl', +(4, 5)$} (6)
-  (7) edge node{$'\pospr', -(4, 5)$} (5);
-
-  \path (6) edge[left] node{$'\pospl', +(6, 7)$} (8)
-  (9) edge[right] node{$'\pospr', -(6, 7)$} (7);
-
-  \path (8) edge[above] node{$'n'$} (9);
-
-  \path (s) edge[bend left] node{$'\pospl', +(s, f)$} (1)
-  (2) edge[bend left] node{$'\pospr', -(s, f)$} (f);
-
+  \node[state,accepting] (6) [right of=5] {$6$};
+
+  \node[state,initial] (7) [below of=4,yshift=7mm] {$7$};
+  \node[state,accepting] (8) [right of=7] {$8$};
+
+  \path (1) edge node{$E$} (2)
+  (2) edge node{$';'$} (3);
+
+  \path (4) edge node{$E$} (5)
+  (5) edge node{$'s'$} (6);
+
+  \path (7) edge node{$'z'$} (8);
 \end{tikzpicture}
@@
 
-\noindent We then combine these disparate DFAs to produce a single visibly pushdown automaton in three steps:
+\noindent We then combine these separate DFAs to produce a single visibly pushdown automaton in three steps:
 
@@
 
-\noindent To make the figure less cluttered we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')}$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')}$.
+\noindent To make the figure less cluttered we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')} q$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')} q$.
 
 \begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
     scale = 1,transform shape]
@@
   \node[state] (2) [right of=1] {$2$};
-
-  \node[state] (3) [below of=1,xshift=-1.2cm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
+  \node[state] (3) [right of=2] {$3$};
+
+  \node[state] (4) [below of=1] {$4$};
   \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=1,yshift=-2cm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \node[state] (8) [below of=6] {$8$};
-  \node[state] (9) [right of=8] {$9$};
-
-  \node[state,initial] (s) [left of=1,xshift=-1cm] {$s$};
-  \node[state,accepting] (f) [right of=2,xshift=1cm] {$f$};
-
-  \path (1) edge node[left]{$'\pospl, +(1, 2)$} (3)
-  (5) edge[right] node{$'\pospr', -(1, 2)$} (2);
-  \path (1) edge[out=180,in=180,looseness=2,left] node{$'\pospl', +(1, 2)$} (6)
-  (7) edge[out=0,in=0,looseness=2,right] node{$'\pospr', -(1, 2)$} (2);
-
-  \path (3) edge node{$'+'$} (4);
-  \path (4) edge[bend right,above,out=270,in=225] node[xshift=5mm]{$'\pospl', +(4, 5)$} (3)
-  (5) edge[loop right,below] node[yshift=-3.5mm]{$'\pospr', -(4, 5)$} (5);
-  \path (4) edge[left] node{$'\pospl', +(4, 5)$} (6)
-  (7) edge node{$'\pospr', -(4, 5)$} (5);
-
-  \path (6) edge[left] node{$'\pospl', +(6, 7)$} (8)
-  (9) edge[right] node{$'\pospr', -(6, 7)$} (7);
-
-  \path (8) edge[above] node{$'n'$} (9);
-
-  \path (s) edge[bend left] node{$'\pospl', +(s, f)$} (1)
-  (2) edge[bend left] node{$'\pospr', -(s, f)$} (f);
-
+  \node[state] (6) [right of=5] {$6$};
+
+  \node[state] (7) [below of=4] {$7$};
+  \node[state] (8) [right of=7] {$8$};
+
+  \node[state,initial] (s) [left of=1,xshift=-0.9cm] {$s$};
+  \node[state,accepting] (f) [right of=3,xshift=0.9cm] {$f$};
+
+  \path (2) edge node{$';'$} (3);
+
+  \path (5) edge node{$'s'$} (6);
+
+  \path (7) edge node{$'z'$} (8);
+
+  \path (s) edge node{$+(s, f)$} (1);
+  \draw (2.north) -- ([yshift=4mm]2.north) -| (f.north);
+  \draw (3.north) -- ([yshift=4mm]3.north) -| (f.north);
+  \node at ([yshift=5mm]$(3)!0.5!(f)$) {$-(s,f)$};
+
+  % 1 -E-> 2
+  \draw (1.east) -| ([xshift=4mm]4.east) |- ([xshift=4mm,yshift=10mm]7.east) -- (7);
+  \draw ([xshift=4mm]4.east) -- (4);
+  \draw (6.north) |- ([xshift=-8mm]$(2)!0.5!(5)$) |- (2.west);
+  \draw (8) -- ([xshift=-3.5mm,yshift=10mm]8.west) |- (2.west);
+  \node at ([xshift=2mm,yshift=2mm]$(1)!0.5!(4)$) {$+(1,2)$};
+  \node at ([xshift=-2mm,yshift=2mm]$(2)!0.5!(5)$) {$-(1,2)$};
+
+  % 4 -E-> 5
+  \draw (4.south) -- (7.north);
+  \path (4) edge[out=270,in=225,looseness=5] node[xshift=4mm]{$+(4,5)$} (4);
+  \draw (6.south) -- ([yshift=-5mm]6.south) -| (5.south);
+  \path (8) edge[right] node[yshift=-3mm]{$-(4,5)$} (5);
 \end{tikzpicture}
@@
 \begin{center}
-\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
     scale = 1,transform shape]
@@
   \node[state] (2) [right of=1] {$2$};
-
-  \node[state] (3) [below of=1,xshift=-1.2cm] {$3$};
-  \node[state] (4) [right of=3] {$4$};
+  \node[state] (3) [right of=2] {$3$};
+
+  \node[state] (4) [below of=1] {$4$};
   \node[state] (5) [right of=4] {$5$};
-
-  \node[state] (6) [below of=1,yshift=-2cm] {$6$};
-  \node[state] (7) [right of=6] {$7$};
-
-  \node[state] (8) [below of=6] {$8$};
-  \node[state] (9) [right of=8] {$9$};
-
-  \node[state,initial] (s) [left of=1,xshift=-1cm] {$s$};
-  \node[state,accepting] (f) [right of=2,xshift=1cm] {$f$};
-
-  \path (1) edge node[left]{$'\pospl, +(1, 2)$} (3)
-  (5) edge[right] node{$'\pospr', -(1, 2)$} (2);
-  \path (1) edge[out=180,in=180,looseness=2,left] node{$'\pospl', +(1, 2)$} (6)
-  (7) edge[out=0,in=0,looseness=2,right] node{$'\pospr', -(1, 2)$} (2);
-
-  \path (3) edge node{$'+'$} (4);
-  \path (4) edge[bend right,above,out=270,in=225] node[xshift=5mm]{$'\pospl', +(4, 5)$} (3)
-  (5) edge[loop right,below] node[yshift=-3.5mm]{$'\pospr', -(4, 5)$} (5);
-  \path (4) edge[left] node{$'\pospl', +(4, 5)$} (6)
-  (7) edge node{$'\pospr', -(4, 5)$} (5);
-
-  \path (6) edge[left] node{$'\pospl', +(6, 7)$} (8)
-  (9) edge[right] node{$'\pospr', -(6, 7)$} (7);
-
-  \path (8) edge[above] node{$'n'$} (9);
-
-  \path (s) edge[bend left] node{$'\pospl', +(s, f)$} (1)
-  (2) edge[bend left] node{$'\pospr', -(s, f)$} (f);
-
+  \node[state] (6) [right of=5] {$6$};
+
+  \node[state] (7) [below of=4] {$7$};
+  \node[state] (8) [right of=7] {$8$};
+
+  \node[state,initial] (s) [left of=1,xshift=-0.9cm] {$s$};
+  \node[state,accepting] (f) [right of=3,xshift=0.9cm] {$f$};
+
+  \path (2) edge node{$';'$} (3);
+
+  \path (5) edge node{$'s'$} (6);
+
+  \path (7) edge node{$'z'$} (8);
+
+  \path (s) edge node{$+(s, f, \epsilon)$} (1)
+  (3) edge node{$-(s, f, \epsilon)$} (f);
+
+  % 1 -E-> 2
+  \draw (1.east) -| ([xshift=4mm]4.east) |- ([xshift=4mm,yshift=10mm]7.east) -- (7);
+  \draw ([xshift=4mm]4.east) -- (4);
+  \draw (6.north) |- ([xshift=-8mm]$(2)!0.5!(5)$) |- (2.west);
+  \draw (8) -- ([xshift=-3.5mm,yshift=10mm]8.west) |- (2.west);
+  \node at ([xshift=1mm,yshift=2mm]$(1)!0.5!(4)$) {$+(1,2,\epsilon)$};
+  \node at ([xshift=-1mm,yshift=2mm]$(2)!0.5!(5)$) {$-(1,2,\epsilon)$};
+
+  % 4 -E-> 5
+  \draw (4.south) -- (7.north);
+  \path (4) edge[out=270,in=225,looseness=5] node[xshift=4mm]{$+(4,5,\epsilon)$} (4);
+  \draw (6.south) -- ([yshift=-5mm]6.south) -| (5.south);
+  \path (8) edge[right] node[yshift=-3mm]{$-(4,5,\epsilon)$} (5);
+
+  % s -E(t)-> f
+  \draw (s.south) |- (7.west);
+  \draw (s.south) |- (4.west);
+  \node at ([xshift=8mm,yshift=-10mm]s) {$+(s, f, a)$};
+  \draw (8.east) -| (f.south);
+  \draw (6.east) -| (f.south);
+  \node at ([xshift=-8mm,yshift=-10mm]f) {$-(s, f, a)$};
 \end{tikzpicture}
@@
 
+    \delta(s, \text{``}\pospl\text{''}, \lambda)
+    & \{ (s_{l}, (s, f, w))
+    & (S, w, N) \in T, N = \labelnt(l) \} \\
+
     \delta(p, \text{``}\pospl\text{''}, \lambda)
     & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, l_2) = q, (l_2, l_3, w) \in T \} \\
-
-    \delta(s, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l_2}, (s, f, w))
-    & \labelnt(l_1) = S, (l_1, l_2, w) \in T \} \\
+    & p \in Q_{l_1}, \delta_{l_1}(p, N) = q, (N, w, N') \in T, N' = \labelnt(l_3) \} \\
+
+    \delta(p, \text{``}\pospr\text{''}, (s, f, w))
+    & \{ (f, \lambda)
+    & (S, w, N) \in T, N = \labelnt(l), p \in F_l \} \\
+
+    \delta(q', \text{``}\pospr\text{''}, (p, q, w))
+    & \{ (q, \lambda)
+    & q' \in F_{l_1}, \labelnt(l_1) = N', (N, w, N') \in T, p \in Q_{l_2}, \delta_{l_2}(p, N) = q \} \\
+
   \end{array}
@@
 
-\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is given in Section~\ref{sec:preliminaries-vpls} (using the partitions $\T_c = \{ \pospl \}$, $\T_i = \T$, and $\T_r = \{\pospr\}$).
-If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
-We check for the existence of such a run by trimming the product automaton and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
-
-Performing this procedure directly on a language definition $D$ gives a sound decision procedure where the presence of two distinct runs implies unresolvably ambiguous.
-If we instead do it on a modified $D$ with no marks on the right hand side of any production (i.e., all non-terminals have the form $N_\emptyset$) we get a sound decision procedure where the absence of two distinct runs implies resolvably ambiguous.
-
-There are thus cases where we can answer resolvably ambiguous or unresolvably ambiguous with certainty, and cases where we cannot answer either way with certainty.
+\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is described in \cite{alurVisiblyPushdownLanguages2004} (using the partitions $\T_c = \{ \pospl \}$, $\T_i = \T$, and $\T_r = \{\pospr\}$).
+If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct successful runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
+We check for the existence of such a run by trimming the product automaton (as described in \cite{caralpTrimmingVisiblyPushdown2015}) and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
+
+Performing this procedure on a language definition $D$ without non-grouping parentheses and marks gives a sound and complete decision procedure for determining if $D$ is resolvably ambiguous.
+If we instead take a language definition $D$ with marks but no non-grouping parentheses, then remove the marks and perform the procedure, then the absence of two such runs implies that $D$ is resolvably ambiguous, while their presence only implies the possibility of an unresolvable ambiguity; the marks might remove it.
 
@@
 
+We denote the word recognized by a run $R$ by $\runtoword(R)$.
+
 \begin{lemma}
-  All words $w \in L(A_{\posp{}})$ are well-formed linear encodings.
+  $\forall t \in L(T_D).\ \runtoword(\treetorun(t))$ is the canonical linear encoding of $\words(t)$.
 \end{lemma}
 
-We denote the word recognized by a run $R$ by $\runtoword(R)$.
-
 \begin{lemma}
-  $\forall t \in L(T_D).\ \lattice(\runtoword(\treetorun(t))) = \lattice(t)$.
+  For all $t \in L(T_D)$ and every linear encoding $c$ such that $\words(t) \subseteq L(c)$, $c \in L(A'_{\posp{}})$.
 \end{lemma}
@@
 \begin{lemma}
-  For all $t \in L(T_D)$ and every well-formed lattice encoding $w$ such that $\lattice(t) \subseteq \lattice(w)$, $w \in L(A'_{\posp{}})$.
+  We can define a function $\runprimetorun$ from runs in $A'_{\posp{}}$ to runs in $A_{\posp{}}$ such that $\forall R.\ L(\runtoword(\runprimetorun(R))) \subseteq L(\runtoword(R))$.
 \end{lemma}
 
-\begin{lemma}
-  We can define a function $\runprimetorun$ from runs in $A'_{\posp{}}$ to runs in $A_{\posp{}}$ such that $\forall R.\ \lattice(\runtoword(\runprimetorun(R))) \subseteq \lattice(\runtoword(R))$.
-\end{lemma}
-
 \begin{proof}
-  By removing every transition that pushes or pops $\gamma$ and replacing every used transition $p \xrightarrow{'\pospl', +g} q$ that does not exist in $A_{\posp{}}$ with a transition $p \xrightarrow{'\reqpl', +g} q$.
+  By removing every transition that pushes or pops $\gamma$.
 \end{proof}
@@
 \begin{theorem}
-  The existence of two distinct runs $R$ and $R'$, where $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, and $\runtoword(R) = \runtoword(R')$, implies the existence of two distinct trees such that the lattice of one is entirely contained in the other.
+  The existence of two distinct runs $R$ and $R'$ such that $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, and $\runtoword(R) = \runtoword(R')$, implies the existence of two distinct trees such that the word language of one is entirely contained in the other.
 \end{theorem}
@@
   Since $\runprimetorun(R') \neq R$ we have two distinct trees $t_1 = \runtotree(\runprimetorun(R'))$ and $t_2 = \runtotree(R)$.
-Since $\lattice(\runtoword(\runprimetorun(R'))) \subseteq \lattice(\runtoword(R'))$ and $\lattice(\runtoword(R')) = \lattice(\runtoword(R))$ we have that $\lattice(t_1) \subseteq \lattice(t_2)$.
+Since $L(\runtoword(\runprimetorun(R'))) \subseteq L(\runtoword(R'))$ and $L(\runtoword(R')) = L(\runtoword(R))$ we have that $L(t_1) \subseteq L(t_2)$.
 \end{proof}
+
+We thus have a way to detect the presence or absence of a pair of trees $t_1, t_2 \in L(T_D)$ such that $\words(t_1) \subseteq \words(t_2)$, which completely determines the resolvability of $D$ when $D$ has no non-grouping parentheses or marks.
 
@@
 Similarly to Section~\ref{sec:static}, we place some restrictions on $D$, though the restrictions are less severe.
-While Section~\ref{sec:static} requires a language definition $D$ to not contain parentheses, this section merely requires parentheses to be balanced.
-Concretely, the restrictions are as follows:
+While Section~\ref{sec:static} requires $D$ to have no non-grouping parentheses and no marks, this section has two much less stringent requirements:
 
@@
 \noindent Both of these restrictions can be checked statically.
+Additionally, if we already know statically that $D$ is resolvably ambiguous\footnote{While the \emph{decision} part of this procedure is uninteresting in this case (it will always answer ``resolvable''), the minimal witness \emph{is} interesting: it shows the user how to resolve the ambiguity.} we can drop the second requirement.
+
+Similarly to Section~\ref{sec:static} the approach centers around the word language $\words(t)$.
+We can easily construct a visibly pushdown automaton for this language by going via the canonical linear encoding: let every pair ``$\reqp{}$'' push and pop the same stack symbol (call it $\gamma$) and let every ``$\posp{}$'' push and pop a unique stack symbol.
+For example, ``$\posp{\reqp{a}\posp{b}}$'' produces the following automaton:
+
+\begin{center}
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+  \node[state,initial] (1) {};
+  \node[state] (2) [right of=1] {};
+  \node[state] (3) [right of=2] {};
+  \node[state] (4) [right of=3] {};
+  \node[state,accepting] (5) [right of=4] {};
+
+  \path (1) edge[loop above] node{$'(', +1$} (1)
+  (1) edge node{$'(', +\gamma$} (2)
+  (2) edge node{$'a'$} (3)
+  (3) edge node{$')', -\gamma$} (4)
+  (4) edge[loop above] node{$'(', +2$} (4)
+  (4) edge node{$'b'$} (5)
+  (5) edge[loop above] node{$')', -2$} (5)
+  (5) edge[loop right] node{$')', -1$} (5);
+\end{tikzpicture}
+\end{center}
+
+\noindent This is useful to us since visibly pushdown automata are closed under difference.
+Given a pair of trees $t_1$ and $t_2$ we can thus produce a pair of automata recognizing $\words(t_1)$ and $\words(t_2)$ respectively, then construct a pair of automata recognizing $\words(t_1) \setminus \words(t_2)$ and $\words(t_2) \setminus \words(t_1)$, respectively.
+These new automata recognize the words that are not ambiguous between these two particular trees.
+Our algorithm thus takes the following shape:
+
+\begin{verbatim}
+Proc(T)
+  let R = \emptyset
+  let U = \emptyset
+  for t \in T
+    let T' = T \setminus \{t\}
+    let A = \words(t)
+    loop
+      A = A \setminus \bigcup_{t' \in T'} \words(t')
+      if L(A) = \emptyset
+        U = U \cup \{t\}
+        break loop
+      if \parse(\shortest(A)) = \{t\}
+        R = R \cup \{(t, \shortest(A))\}
+        break loop
+      T' = \parse(\shortest(A)) \setminus \{t\}
+  return (R, U)
+\end{verbatim}
+
+%% TODO: CONTINUE
 

### 2019-10-21 paper/paper.tex
Most major parts of dynamic are done

@@
 \usepackage{tikz-cd} % Used for commutative diagrams
+\usepackage{algorithm}
+\usepackage[noend]{algpseudocode}
 
@@
 \newcommand{\runprimetorun}{\mathit{normalrun}}
+\newcommand{\shortest}{\mathit{shortest}}
 
@@
 
-\begin{verbatim}
-Proc(T)
-  let R = \emptyset
-  let U = \emptyset
-  for t \in T
-    let T' = T \setminus \{t\}
-    let A = \words(t)
-    loop
-      A = A \setminus \bigcup_{t' \in T'} \words(t')
-      if L(A) = \emptyset
-        U = U \cup \{t\}
-        break loop
-      if \parse(\shortest(A)) = \{t\}
-        R = R \cup \{(t, \shortest(A))\}
-        break loop
-      T' = \parse(\shortest(A)) \setminus \{t\}
-  return (R, U)
-\end{verbatim}
+\begin{algorithm}
+\caption{The Dynamic Algorithm}
+\begin{algorithmic}[1]
+\Procedure{Dynamic}{$T$}
+  \State $R <- \emptyset$
+  \State $U <- \emptyset$
+  \ForAll{$t \in T$}
+    \State $T' <- T \setminus \{t\}$
+    \State $A <- \words(t)$ \Comment{$A$ is a visibly pushdown automaton.}
+    \Loop
+      \State $A <- A \setminus \bigcup_{t' \in T'} \words(t')$ \Comment{VPDAs closed under difference and union.}
+      \If{$L(A) = \emptyset$}
+        \State $U <- U \cup \{t\}$
+        \State $\textbf{break loop}$
+      \EndIf
+      \If{$\parse(\shortest(A)) = \{t\}$} \Comment{$\shortest(A)$ denotes a shortest}
+        \State $R <- R \cup \{(t, \shortest(A))\}$ \Comment{word recognized by $A$.}
+        \State $\textbf{break loop}$
+      \EndIf
+      \State $T' <- \parse(\shortest(A)) \setminus \{t\}$
+    \EndLoop
+  \EndFor
+  \State $\textbf{return} (R, U)$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
 

### 2019-10-21 paper/paper.tex
First pass at added motivation section

@@
 \newcommand{\syncon}{\lstinline[language=syncon]}
+\newcommand{\haskell}{\lstinline[language=haskell]}
 
@@
 
-\section{Motivating Example}
+\section{Motivating Examples}
 \label{sec:motivation}
@@
 %\end{center}
+
+\subsection{Language Composition}
+
+% TODO: more language composition references
+The possibility of composing different languages at will is a prevalent idea in research \cite{vanwykSilverExtensibleAttribute2010,lorenzenSoundTypedependentSyntactic2016,danielssonParsingMixfixOperators2011}.
+Particularly relevant to this paper is the behavior of composed \emph{syntax}; composing unambiguous syntaxes does not necessarily produce an unambiguous composed syntax.
+Pre-existing systems commonly solve this either by strict requirements on the individual languages \cite{vanwykSilverExtensibleAttribute2010} or by largely ignoring the problem and merely presenting the multiple syntax trees in case of an ambiguity \cite{lorenzenSoundTypedependentSyntactic2016,danielssonParsingMixfixOperators2011}.
+The former case gives good guarantees for the end-user experience (no ambiguities) while the former gives more freedom to the language designer.
+Resolvable ambiguity, as presented in this paper, gives a middle ground with more language designer freedom than the former and a better user experience than the latter.
+
+Language composition has limited prevalence outside of research, but it does exist in a simplified form: custom operators in libraries.
+For example, Haskell allows a programmer to define new operators, with new syntax and custom precedence level and associativity.
+A user can then import these operators from any number of libraries and use them together in one program.
+Precedence in Haskell is defined as a number between 0 and 9, meaning that operators from different libraries have a defined relative precedence.
+If the libraries were designed by different authors this precedence is essentially by accident and it may or may not be sensible.
+Leaving the relative precedence undefined would not affect expressivity, every possible combination of operators can be written by simply adding parentheses, and would avoid surprising interpretations.
+This does in fact happen in some cases in Haskell, namely when two operators have the same precedence, but incompatible associativity.
+Both \haskell{>>=} and \haskell{=<<} have precedence 1, but the former is left associative while the latter is right associative.
+\haskell{a >>= b =<< c} then produces the following error:
+
+\begin{lstlisting}
+Precedence parsing error
+  cannot mix '>>=' [infixl 1] and '=<<' [infixr 1]
+  in the same infix expression
+\end{lstlisting}
+
+\noindent The error message shows the problem but suggests no solution.
+Our approach can suggest solutions, allow arbitrary operators to have undefined relative precedence (intransitive precedence), and is general; it handles non-operator ambiguities as well.
 

### 2019-10-23 paper/paper.tex
PEGs in related work

@@
 
-Our approach centers around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
+%% Our approach centers around $\words(t)$, which, being a set of words, can be considered a language in the classical sense.
 Given restriction (1), each such language turns out to be a visibly pushdown language.
@@
 
-Intuitively, we then need to, for each tree $t \in \parse(w)$, determine if $\words(t) \setminus \bigcup_{t' \in L(T_D) \land t' \neq t} \words(t)$ is the empty set.
+%% Intuitively, we then need to, for each tree $t \in \parse(w)$, determine if $\words(t) \setminus \bigcup_{t' \in L(T_D) \land t' \neq t} \words(t)$ is the empty set.
 Unfortunately, $L(T_D)$ tends to be infinite.
@@
 
-For this section, we assume the presence of a single language definition $D$ with terminals in $\T$, labels in $\Labels$, and non-terminals in $\NT$.
-
-\begin{lemma}
-  Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
-  \label{lemma:vpda-from-t}
-\end{lemma}
-
-\noindent We construct this automaton in a bottom-up fashion.
+%% For this section, we assume the presence of a single language definition $D$ with terminals in $\T$, labels in $\Labels$, and non-terminals in $\NT$.
+
+%% \begin{lemma}
+%%   Given a $t \in L(T_D)$, we can construct a visibly pushdown automaton that accepts exactly $\words(t)$.
+%%   \label{lemma:vpda-from-t}
+%% \end{lemma}
+
+%% \noindent We construct this automaton in a bottom-up fashion.
 A sequence of terminals in $\T$ is easily recognized by a sequence of transitions that do not interact with the stack.
@@
 
-We now present the function $f : L(T_D) -> VPDA$ that performs this construction.
+%% We now present the function $f : L(T_D) -> VPDA$ that performs this construction.
 Writing the transition function $\delta : Q \times \T \times (\Gamma \cup \{\lambda\}) -> 2^{Q \times (\Gamma \cup \{\lambda\})}$ as a subset of $Q \times \T \times (\Gamma \cup \{\lambda\}) \times Q \times (\Gamma \cup \{\lambda\})$, and using $\sqcup$ for disjoint union:
 
-$$
-\begin{array}{r@{\;}l}
-  f(a)
-  & = \begin{cases}
-    (\{p, q\}, \T, \emptyset, \{(p, a, \lambda, q, \lambda)\}, p, \{q\})
-    & \text{if } a \notin \{\text{``}(\text{''}, \text{``})\text{''}\} \\
-
-    (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``}(\text{''}, \lambda, q, \gamma)\}, p, \{q\})
-    & \text{if } a = \text{``}(\text{''} \\
-
-    (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``})\text{''}, \lambda, q, \lambda))\}, p, \{q\})
-    & \text{if } $a = \text{``})\text{''}$ \\
-  \end{cases} \\
-
-  f(w_1 \cdot w_2)
-  & = (Q_1 \sqcup Q_2, \T, \Gamma', \delta_1 \cup \delta_2 \cup \{(f_1, \lambda, \lambda, s_2, \lambda)\}, s_1, \{f_2\}) \\
-  & \text{ where } f(w_i) = (Q_i, \T, \Gamma_i, \delta_i, s_i, \{f_i\}) \\
-  & \hphantom{\text{ where }} \Gamma' = (\Gamma_1 \setminus \{\gamma\}) \sqcup (\Gamma_2 \setminus \{\gamma\}) \cup \{\gamma\} \\
-
-  f(l(w))
-  & = \begin{cases}
-    (Q \sqcup \{p, q\}, \T, \Gamma', \delta \cup \delta_1, p, \{q\}) & \text{if $l$ is marked in the introducing non-terminal.} \\
-    (Q, \T, \Gamma', \delta \cup \delta_2, s, \{f\}) & \text{if $l$ is unmarked in the introducing non-terminal.} \\
-  \end{cases} \\
-  & \text{ where } f(w) = (Q, \T, \Gamma, \delta, s, \{f\}) \\
-  & \hphantom{\text{ where }} \Gamma' = (\Gamma \setminus \{\gamma\}) \sqcup \{g, \gamma\} \\
-  & \hphantom{\text{ where }} \delta_1 = \{(p, \text{``}(\text{''}, \lambda, p, g), (p, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, q, \lambda), (q, \text{``})\text{''}, g, q, \lambda)\} \\
-  & \hphantom{\text{ where }} \delta_2 = \{(s, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, f, \lambda)\} \\
-\end{array}
-$$
-
-\begin{lemma}
-  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can, in finite time, construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
-\end{lemma}
-
-\begin{proof}
-  Follows trivially from Lemma~\ref{lemma:vpda-from-t} and VPDAs being closed under union and difference.
-\end{proof}
-
-\begin{lemma}
-  Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$, the set of trees $t' \in L(T_D)$ such that $\words(t) \cap \words(t') \neq \emptyset$ is finite.
-\end{lemma}
-
-\begin{proof}
-  Lemma~\ref{lemma:same-basic}, presented in Section~\ref{sec:word-view}, holds since restriction (1) implies only well-balanced parentheses, thus all $t'$ must share a basic word with all $t_i$.
+%% $$
+%% \begin{array}{r@{\;}l}
+%%   f(a)
+%%   & = \begin{cases}
+%%     (\{p, q\}, \T, \emptyset, \{(p, a, \lambda, q, \lambda)\}, p, \{q\})
+%%     & \text{if } a \notin \{\text{``}(\text{''}, \text{``})\text{''}\} \\
+
+%%     (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``}(\text{''}, \lambda, q, \gamma)\}, p, \{q\})
+%%     & \text{if } a = \text{``}(\text{''} \\
+
+%%     (\{p, q\}, \T, \{\gamma\}, \{(p, \text{``})\text{''}, \lambda, q, \lambda))\}, p, \{q\})
+%%     & \text{if } $a = \text{``})\text{''}$ \\
+%%   \end{cases} \\
+
+%%   f(w_1 \cdot w_2)
+%%   & = (Q_1 \sqcup Q_2, \T, \Gamma', \delta_1 \cup \delta_2 \cup \{(f_1, \lambda, \lambda, s_2, \lambda)\}, s_1, \{f_2\}) \\
+%%   & \text{ where } f(w_i) = (Q_i, \T, \Gamma_i, \delta_i, s_i, \{f_i\}) \\
+%%   & \hphantom{\text{ where }} \Gamma' = (\Gamma_1 \setminus \{\gamma\}) \sqcup (\Gamma_2 \setminus \{\gamma\}) \cup \{\gamma\} \\
+
+%%   f(l(w))
+%%   & = \begin{cases}
+%%     (Q \sqcup \{p, q\}, \T, \Gamma', \delta \cup \delta_1, p, \{q\}) & \text{if $l$ is marked in the introducing non-terminal.} \\
+%%     (Q, \T, \Gamma', \delta \cup \delta_2, s, \{f\}) & \text{if $l$ is unmarked in the introducing non-terminal.} \\
+%%   \end{cases} \\
+%%   & \text{ where } f(w) = (Q, \T, \Gamma, \delta, s, \{f\}) \\
+%%   & \hphantom{\text{ where }} \Gamma' = (\Gamma \setminus \{\gamma\}) \sqcup \{g, \gamma\} \\
+%%   & \hphantom{\text{ where }} \delta_1 = \{(p, \text{``}(\text{''}, \lambda, p, g), (p, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, q, \lambda), (q, \text{``})\text{''}, g, q, \lambda)\} \\
+%%   & \hphantom{\text{ where }} \delta_2 = \{(s, \text{``}(\text{''}, \lambda, s, g), (f, \text{``})\text{''}, g, f, \lambda)\} \\
+%% \end{array}
+%% $$
+
+%% \begin{lemma}
+%%   Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$ we can, in finite time, construct a set of VPDAs $\{A_1, A_2, \ldots, A_n\}$ such that $\forall i.\ L(A_i) \setminus \bigcup_{i \neq i'} \words(t_{i'})$.
+%% \end{lemma}
+
+%% \begin{proof}
+%%   Follows trivially from Lemma~\ref{lemma:vpda-from-t} and VPDAs being closed under union and difference.
+%% \end{proof}
+
+%% \begin{lemma}
+%%   Given a finite set $\{t_1, t_2, \ldots, t_n\} \subseteq L(T_D)$, the set of trees $t' \in L(T_D)$ such that $\words(t) \cap \words(t') \neq \emptyset$ is finite.
+%% \end{lemma}
+
+%% \begin{proof}
+%%   Lemma~\ref{lemma:same-basic}, presented in Section~\ref{sec:word-view}, holds since restriction (1) implies only well-balanced parentheses, thus all $t'$ must share a basic word with all $t_i$.
 $G_D$ with parentheses removed recognizes these basic words, and has a one-to-one correspondence between leftmost derivations and trees in $L(T_D)$, by construction.
 Restriction (2) states that $G_D$ without parentheses is not infinitely ambiguous, thus there must be a finite number of trees that have the same basic word as all $t_i$.
-\end{proof}
-
-Restriction (2) also implies that $G'_D$ is not infinitely ambiguous, whereby $\parse(w)$ must be finite.
+%% \end{proof}
+
+%% Restriction (2) also implies that $G'_D$ is not infinitely ambiguous, whereby $\parse(w)$ must be finite.
 The dynamic analysis thus parses $\basic(w)$ using $G_D$ with parentheses removed to produce a set $R$ of relevant trees.
@@
 
-Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989}).
+Parsing expression grammars \cite{fordParsingExpressionGrammars2004} sidestep the issue of ambiguity by not introducing it in the first place.
+However, this also loses the potential gains of leaving certain ambiguities.
+Additionally, since the ordering of productions matter, composition of languages must be ordered, and the interactions between composed languages becomes non-obvious, e.g., merely adding productions may remove previously recognized words from the language, depending on where they are added.
+
+Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989,lorenzenSoundTypedependentSyntactic2016}).
 These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but also assume that the composed language is unambiguous, which in practice precludes the composition of languages constructed independently of each other.

### 2019-10-23 paper/paper.tex
Revisions for Sections 1 through 3.4

@@
   (Section~\ref{sec:resolvable-definition}).
-\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF), where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
+\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF) where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
 \item We consider versions of the \emph{static resolvability
@@
 \noindent Before presenting the above contributions, the paper starts with motivating examples (Section~\ref{sec:motivation}), as well as preliminaries (Section~\ref{sec:prel}).
+Finally, following a discussion of related work (Section~\ref{sec:related-work}), the paper provides some conclusions (Section~\ref{sec:conclusion}).
 
@@
 If there are no predefined rules of how to disambiguate this expression within the language definition, it is an ambiguous expression, and a parser generates a set of parse trees.
-Consider Fig~\ref{fig:circuits}, which depicts four different alternatives, each with a different meaning, depending on how the ambiguity has been resolved.
-Clearly, the expression has totally different meanings depending on how the end-user places the parenthesis.
+Consider Figure~\ref{fig:circuits} which depicts four different alternatives, each with a different meaning, depending on how the ambiguity has been resolved.
+Clearly, the expression has totally different meanings depending on how the end-user places the parentheses.
 However, if a language designer is forced to make the grammar of the syntax definition unambiguous, a specific choice has to be made for precedence and associativity.
-For instance, assume that the designer makes the arbitrary choice that serial composition has higher precedence than parallel composition, and that both operators are left associative.
-In such a case, the expression without parenthesis is parsed as Figure~\ref{fig:circuits}(d).
+For instance, assume that the designer makes the arbitrary choice that serial composition has higher precedence than parallel composition, and that both operators are left-associative.
+In such a case, the expression without parentheses is parsed as Figure~\ref{fig:circuits}(d).
 The question is why such an arbitrary choice---which is forced by the traditional design of unambiguous grammars---is the correct way to interpret a domain-specific expression.
@@
 Note that there is a fifth alternative $\texttt{C1 -- ((C2 || C3) || C4)}$.
-However, the meaning of this expression is the same as (c), presupposed that the parallel operator is associative.
-In such a case, this expression and the expression in (c) both means that components $\texttt{C2}$, $\texttt{C3}$, and $\texttt{C4}$ are composed in parallel.
+However, the meaning of this expression is the same as (c) assuming the parallel operator is associative.
+In such a case, this expression and the expression in (c) both mean that components $\texttt{C2}$, $\texttt{C3}$, and $\texttt{C4}$ are composed in parallel.
 }
@@
 
-Explicit ambiguity is highly relevant also for the design of new and existing general purpose programming languages.
+Explicit ambiguity is highly relevant also for the design of new and existing general-purpose programming languages.
 The following example shows how an ambiguity error can be clearer than what an existing compiler produces today with the traditional approach.
@@
 The former case gives good guarantees for the end-user experience (no ambiguities) while the former gives more freedom to the language designer.
-Resolvable ambiguity, as presented in this paper, gives a middle ground with more language designer freedom than the former and a better user experience than the latter.
+Resolvable ambiguity, as presented in this paper, provides a middle ground with more language designer freedom than the former and a better user experience than the latter.
 
@@
 Leaving the relative precedence undefined would not affect expressivity, every possible combination of operators can be written by simply adding parentheses, and would avoid surprising interpretations.
-This does in fact happen in some cases in Haskell, namely when two operators have the same precedence, but incompatible associativity.
-Both \haskell{>>=} and \haskell{=<<} have precedence 1, but the former is left associative while the latter is right associative.
+This does in fact happen in some cases in Haskell, namely when two operators have the same precedence but incompatible associativity.
+Both \haskell{>>=} and \haskell{=<<} have precedence 1, but the former is left-associative while the latter is right-associative.
 \haskell{a >>= b =<< c} then produces the following error:
@@
 \begin{definition}
-A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of that word.
+A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of $w$.
 \end{definition}
@@
 
-A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$ ; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
-
-A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that:
-
-\begin{itemize}
-\item $r_0 = q_0$.
-\item $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$.
-\item $r_n \in F$.
-\end{itemize}
-
-\noindent We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a succesful run.
+A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
+A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that $r_0 = q_0$, $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$ and $r_n \in F$.
+We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a successful run.
 
@@
 
-A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols.
+A pushdown automaton extends a finite automaton with a stack to/from which transitions can push/pop symbols.
 Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
@@
 
-However, in this paper we will only consider pushdown automata with relatively limited stack manipulation, and will thus use some convenient shorthand:
+However, in this paper we only consider pushdown automata with relatively limited stack manipulation, and thus use some convenient shorthands:
 
@@
 A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
-A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
+A visibly pushdown automaton (VPDA) is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
 
@@
 %
-This partitioning gives us some useful properties. Of particular relevance to this paper are the following two points:
+This partitioning gives us the following particularly relevant properties:
 
@@
 \item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}.
-Intersection in particular is given by a product automaton, i.e., given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
+Intersection is given by a product automaton.
+Given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
   $$
@@
 
-\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
+\item A VPDA can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
 Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
@@
 The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.
-
 Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree.
@@
 
-\section{Related Work}
+\section{Related Work}\label{sec:related-work}
 
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{sec:conclusion}
 In this paper, we introduce the concept of \emph{resolvable

### 2019-10-23 paper/paper.tex
Fixed messed up figures

@@
 
+\newcommand{\RED}[1]{\textcolor{red}{#1}}
+
 \synctex=1
@@
 
-Consider the following OCaml example of nested \verb|match| expressions, as stated by \cite{palmkvistCreatingDomainSpecificLanguages2019}).
-
-\begin{minipage}{.35\textwidth}
-\begin{lstlisting}[language={[objective]caml},numbers=left]
+Consider this OCaml example of nested \verb|match| expressions, as stated by \cite{palmkvistCreatingDomainSpecificLanguages2019}:
+
+\begin{minipage}{.34\textwidth}
+\begin{lstlisting}[language={[objective]caml},numbers=left,basicstyle=\small\tt]
 match 1 with
@@
 \begin{minipage}{\textwidth}
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\small\tt]
  File "./nestmatch.ml", line 4, characters 4-5:
@@
 \begin{minipage}{\textwidth}
-\begin{lstlisting}[language={[objective]caml},numbers=left]
+\begin{lstlisting}[language={[objective]caml},numbers=left,basicstyle=\small\tt]
 match 1 with
@@
 \begin{itemize}
-\item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}.
-Intersection is given by a product automaton.
-Given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
-  $$
-  \begin{array}{r@{\ =\ }ll}
-    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \!\!\!\!\text{where } c \in \T_c \\
-    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \!\!\!\!\text{where } i \in \T_i \\
-    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \!\!\!\!\text{where } r \in \T_r \\
+\item Visibly pushdown languages with the same input partitions
+  are closed under intersection, complement, and union
+  \cite{alurVisiblyPushdownLanguages2004}. Intersection is given
+  by a product automaton. Given a pair of VPDAs
+  $(Q_1, \T, \delta_1, q_0, F_1)$ and
+  $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the
+  form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0),$
+  $ F_1 \times F_2)$ where, when $c \in \T_c$, $i \in \T_i$ and
+  $r \in \T_r$:
+  \[
+  \begin{array}{r@{\ =\ }l}
+    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \}\\
+    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
+    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
   \end{array}
-  $$
+  \]
 
@@
 \begin{figure}
+
 \centering
@@
   \subfloat[$T_D$, the parse trees of $G_D$.\label{fig:running-example-generated:t}]{%
+    \begin{minipage}{0.5\textwidth}
+    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
@@
     \end{tabular}
+    \end{minipage}
   }
   \subfloat[$T'_D$, the parse trees of $G'_D$.\label{fig:running-example-generated:t-prime}]{%
+    \begin{minipage}{0.5\textwidth}
+    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
@@
     \end{tabular}
+    \end{minipage}
   }
@@
   \subfloat[$G_D$, the generated abstract syntax.\label{fig:running-example-generated:w}]{%
+    \begin{minipage}{0.5\textwidth}
+    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
@@
     \end{tabular}
+    \end{minipage}
   }
   \subfloat[$G'_D$, the generated concrete syntax.\label{fig:running-example-generated:w-prime}]{%
+    \begin{minipage}{0.5\textwidth}
+    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
@@
 
-Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
+Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, \RED{along with visualizations of the syntax trees}.
 Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$.
@@
 \begin{figure}
-  \subfloat[Example tree in $L(T_D)$ and visualization.\label{fig:sum-add-tree}]{%
+  \subfloat[Example tree in $L(T_D)$.\label{fig:sum-add-tree}]{% and visualization.]{%
     %% \begin{center}
@@
     %% \end{center}
+    \begin{minipage}{.5\linewidth}
+      \centering
       $m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))$
+    \end{minipage}
   }
-  \subfloat[Example tree in $L(T'_D)$ and visualization.]{%
+  \subfloat[Example tree in $L(T'_D)$.]{% and visualization.]{%
     %% \begin{center}
@@
     %%   \end{center}
+    \begin{minipage}{.5\linewidth}
+      \centering
       $m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))$
+    \end{minipage}
   }
@@
   \subfloat[Example word in $L(G_D)$.]{%
+    \begin{minipage}{.5\linewidth}
+      \centering
       $1 + 2 * 3$
+    \end{minipage}
   }
   \subfloat[Example word in $L(G'_D)$.]{%
+    \begin{minipage}{.5\linewidth}
+      \centering
       $(1 + 2) * 3$
+    \end{minipage}
   }
@@
 \item $(N, \epsilon, N) \in T$ for all $N \in \NT$.
-\item $(N, l, N') \in T$ for all $N' \in L(\labelregex(l))$ where $\labelnt(l) = N$.
+\item $(N, l, N') \in T$ for all $N' \in \NT \cap L(\labelregex(l))$ where $\labelnt(l) = N$.
 \item $(N_1, w_1 \cdot w_2, N_3) \in T$ for all $(N_1, w_1, N_2), (N_2, w_2, N_3) \in T$.
@@
 These new automata recognize the words that are not ambiguous between these two particular trees.
-Our algorithm thus takes the following shape:
+Our algorithm can thus be seen as Algorithm~\ref{alg:dynamic}.
 
@@
 \caption{The Dynamic Algorithm}
+\label{alg:dynamic}
 \begin{algorithmic}[1]

### 2019-10-23 paper/paper.tex
Turned figure 6 into a wrapfigure
Moved preliminaries to the appendix
Move wrapfig
Added short notations section

@@
 \usepackage{subfig}
+\usepackage{wrapfig}
+\usepackage{multicol}
 \usepackage{amsmath}
@@
 
-\section{Preliminaries}
+\section{Notation}
 \label{sec:prel}
 
-This section briefly describes the theoretical foundations we build upon.
+This section briefly explains the notations used in this paper.
+%
+Given a context-free grammar $G$, we write $L(G)$ to denote the
+set of all words recognized by $G$.
+%
+The standard definition of ambiguity states that a word
+$w \in L(G)$ is ambiguous if there are two distinct leftmost
+derivations of that word.
+%
+An automata is a 5-tuple $(Q,\T,\delta,q_0,F)$, where $Q$ is a
+finite set of states; $\T$ a finite set of terminals; $\delta$ a
+transition function from $Q \times \T$ to finite subsets of $Q$
+(or single states of $Q$ for a deterministic automaton);
+$q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final
+states.
+%
+A pushdown automata further adds a set of stack symbols $\Gamma$
+and equips the transition function $\delta$ with information on
+which symbols are pushed, popped or just read from the stack.
+
+Appendix~\ref{app:prel} contains a more detailed
+description of the preliminaries (for reviewers).
 
@@
 %\noindent We denote the set of regular expressions with alphabet $\T$ as $\regex(\T)$.
-
-\subsection{Context-Free Grammars} \label{sec:preliminaries-cfgs}
-
-A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
-
-%\footnote{Where $^{*}$ is the Kleene star.}
-%
-A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
-Such a sequence is called a \emph{derivation}.
-The set of words recognized by $G$ is written $L(G)$
-%\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.
-
-%\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
-
-The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{leftmost derivations}.
-A leftmost derivation is a derivation where the non-terminal being replaced is always the leftmost one.
-
-\begin{definition}
-A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of $w$.
-\end{definition}
-
-\subsection{Automata} \label{sec:preliminaries-automata}
-
-A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
-A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that $r_0 = q_0$, $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$ and $r_n \in F$.
-We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a successful run.
-
-A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to.
-NFAs and DFAs have the same expressive power as regular expressions, i.e., for every regular expression there is an NFA and a DFA both reconizing the same language, and vice-versa.
-% TODO: probably reference for this
-
-A pushdown automaton extends a finite automaton with a stack to/from which transitions can push/pop symbols.
-Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
-$\lambda$ essentially means ``ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).
-
-A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma^{*}$.
-
-However, in this paper we only consider pushdown automata with relatively limited stack manipulation, and thus use some convenient shorthands:
-
-\begin{itemize}
-\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all, i.e., $\delta(p, a, \lambda) \supseteq \{(q, \lambda)\}$.
-\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack, i.e., $\delta(p, a, \lambda) \supseteq \{(q, g)\}$.
-\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack, i.e., $\delta(p, a, g) \supseteq \{(q, \lambda)\}$.
-\end{itemize}
-
-\subsection{Visibly Pushdown Languages} \label{sec:preliminaries-vpls}
-
-A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
-A visibly pushdown automaton (VPDA) is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
-
-\begin{itemize}
-\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$; or
-\item $p \xrightarrow{i} q$, where $i \in \T_i$; or
-\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$,
-\end{itemize}
-
-\noindent i.e., the terminal recognized by a transition fully determines the change to the stack height.
-%
-The names of the partitions stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
-%
-This partitioning gives us the following particularly relevant properties:
-
-\begin{itemize}
-\item Visibly pushdown languages with the same input partitions
-  are closed under intersection, complement, and union
-  \cite{alurVisiblyPushdownLanguages2004}. Intersection is given
-  by a product automaton. Given a pair of VPDAs
-  $(Q_1, \T, \delta_1, q_0, F_1)$ and
-  $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the
-  form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0),$
-  $ F_1 \times F_2)$ where, when $c \in \T_c$, $i \in \T_i$ and
-  $r \in \T_r$:
-  \[
-  \begin{array}{r@{\ =\ }l}
-    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \}\\
-    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
-    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
-  \end{array}
-  \]
-
-\item A VPDA can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
-Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
-\end{itemize}
-
-\subsection{Unranked Regular Tree Grammars} \label{sec:preliminaries-trees}
-
-Trees generalize words by allowing each terminal to have multiple ordered successors, instead of just zero or one.
-Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors.
-This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed.
-The sequence of successors to a single terminal in an unranked tree tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.
-
-The results and properties presented in this paper are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
-We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves.
-Since we will use unranked trees to represent parse trees, the former will represent terminals from the parsed word, while the latter represent terminals introduced as internal nodes.
-
-An unranked tree grammar $T$ is a tuple $(\NT, \T, X, P, S)$ where:
-
-\begin{itemize}
-\item $\NT$ is a set of (zero-arity) non-terminals.
-\item $\T$ is a set of zero-arity terminals, used as leaves.
-\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
-\item $P$ is a set of productions, a finite subset of $\NT \times X \times \regex(\T \cup X)$. We will write a production $(N, x, r)$ as $N -> x(r)$.
-\end{itemize}
-
-\noindent A tree $t$ (containing only terminals from $\T$ and $X$) is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
-The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.
-Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree.
-Informally, it is the flattening of a tree after all internal nodes have been removed.
 
@@
 
-At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses; they are allowed exactly where they would surround a complete production.
-For example, $G_D$ in Fig.~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
-However, starting with that language definition would allow the expression ``$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ``$;2$'' is not a valid expression.
-
-Finally, we require a function $\semantic : L(T'_D) -> L(T_D)$ that removes grouping parentheses from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
-The relation between the four grammars in terms of $\yield$ and $\semantic$ can be seen in Fig.~\ref{fig:grammar-square}.
-With this we can define $\parse : L(G'_D) -> 2^{L(T_D)}$, along with its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
-
-\begin{figure}
+\begin{wrapfigure}{l}{0.3\textwidth}
   \begin{tikzcd}
@@
   \label{fig:grammar-square}
-\end{figure}
+\end{wrapfigure}
+
+At this point we also note that the shape of $D$ determines where the final concrete syntax permits grouping parentheses; they are allowed exactly where they would surround a complete production.
+For example, $G_D$ in Fig.~\ref{fig:running-example-generated:w} can be seen as a valid language definition (if we generate new unique labels for each of the productions).
+However, starting with that language definition would allow the expression ``$[1(;2)]$'', which makes no intuitive sense; grouping parentheses should only be allowed around complete expressions, but ``$;2$'' is not a valid expression.
+
+Finally, we require a function $\semantic : L(T'_D) -> L(T_D)$ that removes grouping parentheses from a parse tree, i.e., it replaces every subtree $g($ \verb|'('| $t$ \verb|')'| $)$ with $t$.
+The relation between the four grammars in terms of $\yield$ and $\semantic$ can be seen in Fig.~\ref{fig:grammar-square}.
+With this we can define $\parse : L(G'_D) -> 2^{L(T_D)}$, along with its inverse $\words : L(T_D) -> 2^{L(G'_D)}$:
 
@@
 
+\appendix
+
+\section{Preliminaries}
+\label{app:prel}
+
+This appendix briefly describes some of the theoretical
+foundations we build upon.
+
+\subsection{Context-Free Grammars} \label{sec:preliminaries-cfgs}
+
+A context-free grammar (CFG) $G$ is a 4-tuple $(\NT, \T, P, S)$ where $\NT$ is a set of non-terminals; $\T$ a set of terminals, disjoint from $\NT$; $P$ a finite subset of $\NT \times (\NT \cup \T)^{*}$, i.e., a set of productions; and $S \in \NT$ the starting non-terminal.
+
+%\footnote{Where $^{*}$ is the Kleene star.}
+%
+A word $w \in \T^{*}$ is recognized by $G$ if there is a sequence of steps starting with $S$ and ending with $w$, where each step replaces a single non-terminal using a production in $P$.
+Such a sequence is called a \emph{derivation}.
+The set of words recognized by $G$ is written $L(G)$
+%\footnote{We will always name a regular expression $r$ (possibly with a subscript) and a CFG $G$ (possibly with a subscript), to lessen the risk of confusion}.
+
+%\subsection{Ambiguity} \label{sec:preliminaries-ambiguity}
+
+The standard definition of ambiguity, given a context-free grammar $G$, is expressed in terms of \emph{leftmost derivations}.
+A leftmost derivation is a derivation where the non-terminal being replaced is always the leftmost one.
+
+\begin{definition}
+A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of $w$.
+\end{definition}
+
+\subsection{Automata} \label{sec:preliminaries-automata}
+
+A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
+A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that $r_0 = q_0$, $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$ and $r_n \in F$.
+We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a successful run.
+
+A deterministic finite automaton (DFA) has the same definition, except $\delta : Q \times \Sigma -> Q$, i.e., given a state and a symbol there is always a single state we can transition to.
+NFAs and DFAs have the same expressive power as regular expressions, i.e., for every regular expression there is an NFA and a DFA both reconizing the same language, and vice-versa.
+% TODO: probably reference for this
+
+A pushdown automaton extends a finite automaton with a stack to/from which transitions can push/pop symbols.
+Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
+$\lambda$ essentially means ``ignore'', i.e., $\delta(q_1, \lambda, \lambda) = \{(q_2, \lambda)\}$ means: transition from state $q_1$ without consuming an input symbol (the first $\lambda$) and without examining or popping from the current stack (the second $\lambda$), to state $q_2$ without pushing a new symbol on the stack (the third $\lambda$).
+
+A successful run is now a sequence of \emph{configurations}, elements of $Q \times \Gamma^{*}$, starting with $(q_0, \epsilon)$, ending with $(f, \gamma)$ for some $f \in F$ and $\gamma \in \Gamma^{*}$.
+
+However, in this paper we only consider pushdown automata with relatively limited stack manipulation, and thus use some convenient shorthands:
+
+\begin{itemize}
+\item $p \xrightarrow{a} q$, a transition that recognizes the terminal $a$ and does not interact with the stack at all, i.e., $\delta(p, a, \lambda) \supseteq \{(q, \lambda)\}$.
+\item $p \xrightarrow{a, +g} q$, a transition that recognizes the terminal $a$ and pushes the symbol $g$ on the stack, i.e., $\delta(p, a, \lambda) \supseteq \{(q, g)\}$.
+\item $p \xrightarrow{a, -g} q$, a transition that recognizes the terminal $a$ and pops the symbol $g$ from the stack, i.e., $\delta(p, a, g) \supseteq \{(q, \lambda)\}$.
+\end{itemize}
+
+\subsection{Visibly Pushdown Languages} \label{sec:preliminaries-vpls}
+
+A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
+A visibly pushdown automaton (VPDA) is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
+
+\begin{itemize}
+\item $p \xrightarrow{c, +s} q$, where $c \in \T_c$ and $s \in \Gamma$; or
+\item $p \xrightarrow{i} q$, where $i \in \T_i$; or
+\item $p \xrightarrow{r, -s} q$, where $r \in \T_r$ and $s \in \Gamma$,
+\end{itemize}
+
+\noindent i.e., the terminal recognized by a transition fully determines the change to the stack height.
+%
+The names of the partitions stem from their original use in program analysis, $c$ is for \emph{call}, $i$ for \emph{internal}, and $r$ for \emph{return}.
+%
+This partitioning gives us the following particularly relevant properties:
+
+\begin{itemize}
+\item Visibly pushdown languages with the same input partitions
+  are closed under intersection, complement, and union
+  \cite{alurVisiblyPushdownLanguages2004}. Intersection is given
+  by a product automaton. Given a pair of VPDAs
+  $(Q_1, \T, \delta_1, q_0, F_1)$ and
+  $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the
+  form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0),$
+  $ F_1 \times F_2)$ where, when $c \in \T_c$, $i \in \T_i$ and
+  $r \in \T_r$:
+  \[
+  \begin{array}{r@{\ =\ }l}
+    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \}\\
+    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
+    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
+  \end{array}
+  \]
+
+\item A VPDA can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
+Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
+\end{itemize}
+
+\subsection{Unranked Regular Tree Grammars} \label{sec:preliminaries-trees}
+
+Trees generalize words by allowing each terminal to have multiple ordered successors, instead of just zero or one.
+Most literature considers \emph{ranked} tree languages, where each terminal has a fixed arity, i.e., the same terminal must always have the same number of successors.
+This is as opposed to \emph{unranked} tree languages, where the arity of a terminal is not fixed.
+The sequence of successors to a single terminal in an unranked tree tends to be described by a word language (referred to as a horizontal language in \cite{comonTreeAutomataTechniques2007}), often a regular language.
+
+The results and properties presented in this paper are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
+We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves.
+Since we will use unranked trees to represent parse trees, the former will represent terminals from the parsed word, while the latter represent terminals introduced as internal nodes.
+
+An unranked tree grammar $T$ is a tuple $(\NT, \T, X, P, S)$ where:
+
+\begin{itemize}
+\item $\NT$ is a set of (zero-arity) non-terminals.
+\item $\T$ is a set of zero-arity terminals, used as leaves.
+\item $X$ is a set of terminals without fixed arity, used as inner nodes or leaves.
+\item $P$ is a set of productions, a finite subset of $\NT \times X \times \regex(\T \cup X)$. We will write a production $(N, x, r)$ as $N -> x(r)$.
+\end{itemize}
+
+\noindent A tree $t$ (containing only terminals from $\T$ and $X$) is recognized by $T$ if there is a sequence of steps starting with $S$ and ending with $t$, where each step either replaces a single non-terminal using a production in $P$, or replaces a regular expression $r$ with a sequence in $L(r)$.
+The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.
+Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree.
+Informally, it is the flattening of a tree after all internal nodes have been removed.
+
 \end{document}

### 2019-10-24 paper/paper.tex
Minor fixes

--- 2019-10-23 paper/paper.tex
@@
   (Section~\ref{sec:resolvable-definition}).
-\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF) where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
+\item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF), where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
 \item We consider versions of the \emph{static resolvability
@@
 \noindent Before presenting the above contributions, the paper starts with motivating examples (Section~\ref{sec:motivation}), as well as preliminaries (Section~\ref{sec:prel}).
-Finally, following a discussion of related work (Section~\ref{sec:related-work}), the paper provides some conclusions (Section~\ref{sec:conclusion}).
 
@@
 If there are no predefined rules of how to disambiguate this expression within the language definition, it is an ambiguous expression, and a parser generates a set of parse trees.
-Consider Figure~\ref{fig:circuits} which depicts four different alternatives, each with a different meaning, depending on how the ambiguity has been resolved.
-Clearly, the expression has totally different meanings depending on how the end-user places the parentheses.
+Consider Fig~\ref{fig:circuits}, which depicts four different alternatives, each with a different meaning, depending on how the ambiguity has been resolved.
+Clearly, the expression has totally different meanings depending on how the end-user places the parenthesis.
 However, if a language designer is forced to make the grammar of the syntax definition unambiguous, a specific choice has to be made for precedence and associativity.
-For instance, assume that the designer makes the arbitrary choice that serial composition has higher precedence than parallel composition, and that both operators are left-associative.
-In such a case, the expression without parentheses is parsed as Figure~\ref{fig:circuits}(d).
+For instance, assume that the designer makes the arbitrary choice that serial composition has higher precedence than parallel composition, and that both operators are left associative.
+In such a case, the expression without parenthesis is parsed as Figure~\ref{fig:circuits}(d).
 The question is why such an arbitrary choice---which is forced by the traditional design of unambiguous grammars---is the correct way to interpret a domain-specific expression.
@@
 Note that there is a fifth alternative $\texttt{C1 -- ((C2 || C3) || C4)}$.
-However, the meaning of this expression is the same as (c) assuming the parallel operator is associative.
-In such a case, this expression and the expression in (c) both mean that components $\texttt{C2}$, $\texttt{C3}$, and $\texttt{C4}$ are composed in parallel.
+However, the meaning of this expression is the same as (c), presupposed that the parallel operator is associative.
+In such a case, this expression and the expression in (c) both means that components $\texttt{C2}$, $\texttt{C3}$, and $\texttt{C4}$ are composed in parallel.
 }
@@
 
-Explicit ambiguity is highly relevant also for the design of new and existing general-purpose programming languages.
+Explicit ambiguity is highly relevant also for the design of new and existing general purpose programming languages.
 The following example shows how an ambiguity error can be clearer than what an existing compiler produces today with the traditional approach.
@@
 The former case gives good guarantees for the end-user experience (no ambiguities) while the former gives more freedom to the language designer.
-Resolvable ambiguity, as presented in this paper, provides a middle ground with more language designer freedom than the former and a better user experience than the latter.
+Resolvable ambiguity, as presented in this paper, gives a middle ground with more language designer freedom than the former and a better user experience than the latter.
 
@@
 Leaving the relative precedence undefined would not affect expressivity, every possible combination of operators can be written by simply adding parentheses, and would avoid surprising interpretations.
-This does in fact happen in some cases in Haskell, namely when two operators have the same precedence but incompatible associativity.
-Both \haskell{>>=} and \haskell{=<<} have precedence 1, but the former is left-associative while the latter is right-associative.
+This does in fact happen in some cases in Haskell, namely when two operators have the same precedence, but incompatible associativity.
+Both \haskell{>>=} and \haskell{=<<} have precedence 1, but the former is left associative while the latter is right associative.
 \haskell{a >>= b =<< c} then produces the following error:
@@
 \begin{definition}
-A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of $w$.
+A word $w \in L(G)$ is ambiguous if there are two distinct leftmost derivations of that word.
 \end{definition}
@@
 
-A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
-A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that $r_0 = q_0$, $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$ and $r_n \in F$.
-We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a successful run.
+A nondeterministic finite automaton (NFA) is a 5-tuple $(Q, \T, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of terminals; $\delta$ a transition function from $Q \times \T$ to finite subsets of $Q$ ; $q_0 \in Q$ an initial state; and $F \subseteq Q$ a set of final states.
+
+A successful run is a sequence of states $r_0, \ldots, r_n$ and a word $a_0\cdots a_n$ such that:
+
+\begin{itemize}
+\item $r_0 = q_0$.
+\item $\forall i \in \{0, 1, \ldots, n-1\}.\ r_{i+1} \in \delta(r_i, a_i)$.
+\item $r_n \in F$.
+\end{itemize}
+
+\noindent We say that the automaton accepts the word $a_0a_1\cdots a_n$ iff there is such a succesful run.
 
@@
 
-A pushdown automaton extends a finite automaton with a stack to/from which transitions can push/pop symbols.
+A pushdown automaton extends a finite automaton with a stack from which transitions can push or pop symbols.
 Formally, a nondeterministic pushdown automaton is a 6-tuple $(Q, \T, \Gamma, \delta, q_0, F)$ where $Q$ is a finite set of states; $\T$ a finite set of input symbols, i.e., an input alphabet; $\Gamma$ a finite set of stack symbols, i.e., a stack alphabet; $\delta$ a transition function from $Q \times (\T \cup \{\lambda\}) \times (\Gamma \cup \{\lambda\}$ to finite subsets of $Q \times (\Gamma \cup \{\lambda\}$; $q_0 \in Q$ the initial state; and $F \subseteq Q$ a set of final states.
@@
 
-However, in this paper we only consider pushdown automata with relatively limited stack manipulation, and thus use some convenient shorthands:
+However, in this paper we will only consider pushdown automata with relatively limited stack manipulation, and will thus use some convenient shorthand:
 
@@
 A visibly pushdown language \cite{alurVisiblyPushdownLanguages2004} is a language that can be recognized by a visibly pushdown automaton.
-A visibly pushdown automaton (VPDA) is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
+A visibly pushdown automaton is a pushdown automaton where the input alphabet $\T$ can be partitioned into three disjoint sets $\T_c$, $\T_i$, and $\T_r$, such that all transitions in the automaton have one of the following three forms:
 
@@
 %
-This partitioning gives us the following particularly relevant properties:
+This partitioning gives us some useful properties. Of particular relevance to this paper are the following two points:
 
@@
 \item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}.
-Intersection is given by a product automaton.
-Given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
+Intersection in particular is given by a product automaton, i.e., given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
   $$
@@
 
-\item A VPDA can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
+\item A visibly pushdown automaton can be trimmed \cite{caralpTrimmingVisiblyPushdown2015}, i.e., modified in such a way that all remaining states and transitions are part of at least one successful run; none are redundant.
 Furthermore, a successful run in the trimmed automaton corresponds to exactly one successful run in the original automaton, and vice-versa.
@@
 The set of trees recognized by $T$ is written $L(T)$\footnote{Again, to distinguish from regular expressions and context-free languages, all trees will be named $T$, possibly with a subscript.}.
+
 Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals $a \in \T$ obtained by a left-to-right\footnote{Preorder, postorder, or inorder does not matter since terminals in $\T$ only appear as leaves} traversal of a tree.
@@
 
-\section{Related Work}\label{sec:related-work}
+\section{Related Work}
 
@@
 
-\section{Conclusion}\label{sec:conclusion}
+\section{Conclusion}
 In this paper, we introduce the concept of \emph{resolvable

### 2019-10-24 paper/paper.tex

--- 2019-10-23 paper/paper.tex
@@
 
-\newcommand{\RED}[1]{\textcolor{red}{#1}}
-
 \synctex=1
@@
 
-Consider this OCaml example of nested \verb|match| expressions, as stated by \cite{palmkvistCreatingDomainSpecificLanguages2019}:
-
-\begin{minipage}{.34\textwidth}
-\begin{lstlisting}[language={[objective]caml},numbers=left,basicstyle=\small\tt]
+Consider the following OCaml example of nested \verb|match| expressions, as stated by \cite{palmkvistCreatingDomainSpecificLanguages2019}).
+
+\begin{minipage}{.35\textwidth}
+\begin{lstlisting}[language={[objective]caml},numbers=left]
 match 1 with
@@
 \begin{minipage}{\textwidth}
-\begin{lstlisting}[basicstyle=\small\tt]
+\begin{lstlisting}
   File "./nestmatch.ml", line 4, characters 4-5:
@@
 \begin{minipage}{\textwidth}
-\begin{lstlisting}[language={[objective]caml},numbers=left,basicstyle=\small\tt]
+\begin{lstlisting}[language={[objective]caml},numbers=left]
 match 1 with
@@
 \begin{itemize}
-\item Visibly pushdown languages with the same input partitions
-  are closed under intersection, complement, and union
-  \cite{alurVisiblyPushdownLanguages2004}. Intersection is given
-  by a product automaton. Given a pair of VPDAs
-  $(Q_1, \T, \delta_1, q_0, F_1)$ and
-  $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the
-  form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0),$
-  $ F_1 \times F_2)$ where, when $c \in \T_c$, $i \in \T_i$ and
-  $r \in \T_r$:
-  \[
-  \begin{array}{r@{\ =\ }l}
-    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \}\\
-    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
-    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
+\item Visibly pushdown languages with the same input partitions are closed under intersection, complement, and union \cite{alurVisiblyPushdownLanguages2004}.
+Intersection is given by a product automaton.
+Given a pair of VPDAs $(Q_1, \T, \delta_1, q_0, F_1)$ and $(Q_2, \T, \delta_2, q'_0, F_2)$ their product automaton has the form $(Q_1 \times Q_2, \T, \delta', (q_0, q'_0), F_1 \times F_2)$ where:
+  $$
+  \begin{array}{r@{\ =\ }ll}
+    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \} & \!\!\!\!\text{where } c \in \T_c \\
+    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \} & \!\!\!\!\text{where } i \in \T_i \\
+    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \} & \!\!\!\!\text{where } r \in \T_r \\
   \end{array}
-  \]
+  $$
 
@@
 \begin{figure}
-
 \centering
@@
   \subfloat[$T_D$, the parse trees of $G_D$.\label{fig:running-example-generated:t}]{%
-    \begin{minipage}{0.5\textwidth}
-    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
@@
     \end{tabular}
-    \end{minipage}
   }
   \subfloat[$T'_D$, the parse trees of $G'_D$.\label{fig:running-example-generated:t-prime}]{%
-    \begin{minipage}{0.5\textwidth}
-    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}c@{$($ }l@{}}
@@
     \end{tabular}
-    \end{minipage}
   }
@@
   \subfloat[$G_D$, the generated abstract syntax.\label{fig:running-example-generated:w}]{%
-    \begin{minipage}{0.5\textwidth}
-    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
@@
     \end{tabular}
-    \end{minipage}
   }
   \subfloat[$G'_D$, the generated concrete syntax.\label{fig:running-example-generated:w-prime}]{%
-    \begin{minipage}{0.5\textwidth}
-    \centering
     \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
@@
 
-Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, \RED{along with visualizations of the syntax trees}.
+Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, along with visualizations of the syntax trees.
 Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$.
@@
 \begin{figure}
-  \subfloat[Example tree in $L(T_D)$.\label{fig:sum-add-tree}]{% and visualization.]{%
+  \subfloat[Example tree in $L(T_D)$ and visualization.\label{fig:sum-add-tree}]{%
     %% \begin{center}
@@
     %% \end{center}
-    \begin{minipage}{.5\linewidth}
-      \centering
     $m(a(n(\terminal{1}) \terminal{+} n(\terminal{2})) \terminal{*} n(\terminal{3}))$
-    \end{minipage}
   }
-  \subfloat[Example tree in $L(T'_D)$.]{% and visualization.]{%
+  \subfloat[Example tree in $L(T'_D)$ and visualization.]{%
     %% \begin{center}
@@
     %% \end{center}
-    \begin{minipage}{.5\linewidth}
-      \centering
     $m(g( \terminal{(} a(n( \terminal{1} ) \terminal{+} n( \terminal{2} )) \terminal{)} ) \terminal{*} n( \terminal{3} ))$
-    \end{minipage}
   }
@@
   \subfloat[Example word in $L(G_D)$.]{%
-    \begin{minipage}{.5\linewidth}
-      \centering
     $1 + 2 * 3$
-    \end{minipage}
   }
   \subfloat[Example word in $L(G'_D)$.]{%
-    \begin{minipage}{.5\linewidth}
-      \centering
     $(1 + 2) * 3$
-    \end{minipage}
   }

### 2019-10-24 paper/paper.tex
Minor

--- 2019-10-23 paper/paper.tex
@@
 This section briefly explains the notations used in this paper.
-%
+Appendix~\ref{app:prel} contains a more detailed description of
+the preliminaries (for reviewers).
+
+\paragraph{Grammars}
 Given a context-free grammar $G$, we write $L(G)$ to denote the
@@
 derivations of that word.
-%
+
+\paragraph{Automata}
 An automata is a 5-tuple $(Q,\T,\delta,q_0,F)$, where $Q$ is a
@@
 which symbols are pushed, popped or just read from the stack.
-
-Appendix~\ref{app:prel} contains a more detailed
-description of the preliminaries (for reviewers).
 
@@
 
+\pagebreak
 \appendix

### 2019-10-24 paper/paper.tex
Updated intro and motivation

@@
 preferable compared to completely unambiguous
-definitions. Specifically, instead of forcing language designers to
-make more or less arbitrary choices to disambiguate the language, a more natural
-solution may be to leave the decision up to the programmer.
+definitions.
 
@@
 \emph{resolvably ambiguous} if the ambiguity can be resolved by
-modifying the program, such that each of the possible parse trees can
+an end-user such that each of the possible parse trees can
 be selected using different modifications.
@@
 different decision problems.
-%
-The \emph{dynamic resolvable ambiguity problem} is as follows:
-given a parse function for a language, and a specific program that
-is parsed into a set of parse trees, decide whether it is possible to
-rewrite the program, such that each parse tree can be generated
-unambiguously.
-%
-This problem can be generalized into the \emph{static resolvable
-  ambiguity problem}: given a parse function, decide whether all
-parsable programs are resolvably ambiguous.
-%
-Static resolvability means that a language designer can be
-confident that the end-user (the programmer) can always resolve an
-ambiguity by modifying the program. Identifying such ambiguities
-for a \emph{specific} program and suggesting disambiguations is
-enabled by dynamic resolvability.
-%
-Static resolvability gives formally verified guarantees on a
-syntax definition, whereas dynamic resolvability enables the
-language designer to test the resolvability of the syntax
-definition, program by program.
-
-%This presents a problem, since such an ambiguity is an error an end-user might encounter, yet it can only be solved by the language designer, since it requires a change to the language grammar.
-We thus have a concept of ambiguities that can be resolved by an end-user, and those that cannot.
-We call an instance of the former a \emph{resolvable ambiguity}, and the latter an \emph{unresolvable ambiguity}.
-This paper introduces these concepts in the context of language theory, thus we henceforth use terms such as ``word'' instead of ``program''.
-
-% TODO: I feel like there should be a clear transition between the more practical setting of programs and code, and the theoretical setting with words and terminals, but I'm not convince the above sentence is very good at that.
-
-% TODO: resolvable ambiguity cannot be defined solely in terms of a grammar, we require interpretations of words as well, thus we have to introduce a new formalism that does this, the formalism is an intuitive extension of EBNF grammars.
-this is probably written as a contrast with classical ambiguity, which should be described briefly for contrast
-
-%Resolvable ambiguity, like classical ambiguity, gives rise to two questions: can we determine if a particular word is resolvably ambiguous (dynamic resolvability), and can we determine if all ambiguous words in a language are resolvably ambiguous (static resolvability)?
-For classical ambiguity in context-free grammars, the former problem is solved by any general parser, while the latter is undecidable \cite{cantorAmbiguityProblemBackus1962}.
-For resolvable ambiguity, in a formalism that associates trees with EBNF-grammars, the dynamic problem is similarly solvable, and the static problem is similarly difficult.
-We do not yet know if the latter is undecidable in general, but we have found a subclass of languages for which it is decidable.
-
-In this paper, we solve the static resolvability problem for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when defining language syntax.
-More specifically, we make the following contributions:
+For a particular language, the \emph{dynamic resolvable ambiguity problem} asks the following question: can every parse tree of \emph{one particular} program be written unambiguously, while the \emph{static resolvability problem} asks if every parse tree of \emph{every} program can be written unambiguously.
+Furthermore we must be able to produce an unambiguous program for a given parse tree to be able to give good error messages to an end-user encountering an ambiguity.
+
+In this paper we describe a syntax definition formalism where ambiguities can be resolved by grouping parentheses, and precedence and associativity is specified through \emph{marks} (c.f.
+Section~\ref{sec:parse-time-disambiguation}).
+As a delimitation we only consider languages with balanced parentheses.
+Our approach centers around the idea of the language of a parse tree; since a parse tree can be written in multiple ways we can usefully describe it as a language.
+This language turns out to be visibly pushdown \cite{alurVisiblyPushdownLanguages2004}, which gives us the tools solve the dynamic problem and present unambiguous programs.
+For the static problem we divide languages along two axes: whether they have non-grouping parentheses, and whether they have marks.
+We then present a novel encoding of the language of a tree that allows us to create an automaton that recognizes the language of languages of trees.
+Examining this automaton solves the static problem for a language subclass: languages with neither non-grouping parentheses nor marks.
+We also show that adding marks to a resolvably ambiguous language in this subclass preserves resolvability.
+
+A language designer can thus either use our solution to the static problem (if their language is in the appropriate subclass) or use an approach similar to unit testing through our solution to the dynamic problem.
+Our solution to the dynamic problem also gives suggested fixes to end-users encountering ambiguities.
+
+More concretely, we make the following contributions:
 
@@
 \item As part of the formalization of the dynamic and static resolvability problems, we define a syntax definition formalism based on Extended Backus–Naur Form (EBNF) where ambiguities can be resolved with grouping parentheses (Section~\ref{sec:parse-time-disambiguation}).
-\item We consider versions of the \emph{static resolvability
-    problem} with increasing difficulty by considering two
-  differently restricted classes of languages. We formalize a
-  decidable algorithm that is sound and complete for the first
-  class, and sound for the second (Section~\ref{sec:static}).
-\item We devise a decidable algorithm that solves the
-  \emph{dynamic resolvability problem} for general context-free
-  languages with well-balanced parentheses
-  (Section~\ref{sec:dynamic}).
+\item We describe the language of a single, arbitrary parse tree and present a novel linear encoding of the same (Section~\ref{sec:linear-encoding}).
+\item We devise a decidable algorithm that solves the \emph{dynamic resolvability problem} for languages with balanced parentheses (Section~\ref{sec:dynamic}).
+Additionally, this algorithm also produces a minimal unambiguous word for each resolvable parse tree.
+\item We solve the \emph{static resolvability problem} for a language subclass.
+We further show that a resolvable result can be preserved through certain modifications that bring the language into a larger subclass (Section~\ref{sec:static}).
 \item We evaluate the dynamic resolvability methodology in the context of two
@@
 Our approach can suggest solutions, allow arbitrary operators to have undefined relative precedence (intransitive precedence), and is general; it handles non-operator ambiguities as well.
-
-\subsection{Resolvable Ambiguity}
-The two motivating examples above motivate briefly why grammar
-and language definitions do not always need to be unambiguous. However, an
-ambiguous program that cannot be resolved by the end-user is clearly
-undesirable. In the rest of the paper, we formalize the approach of
-resolvable ambiguity, and show how we can give guarantees for
-resolvability.
 
@@
 
+\section{The Word Language of a Parse Tree} \label{sec:linear-encoding}
+
 \section{Static Resolvability Analysis} \label{sec:static}

### 2019-10-24 paper/paper.tex
Updated intro + broke out words(t) into separate section + swapped order of static and dynamic

@@
 For a particular language, the \emph{dynamic resolvable ambiguity problem} asks the following question: can every parse tree of \emph{one particular} program be written unambiguously, while the \emph{static resolvability problem} asks if every parse tree of \emph{every} program can be written unambiguously.
+The latter problem in particular is difficult, since it in general involves examining an infinite set (one per parse tree) of infinite sets (words that can parse as that parse tree) to find one without a unique element (i.e., an unambiguous word).
 Furthermore we must be able to produce an unambiguous program for a given parse tree to be able to give good error messages to an end-user encountering an ambiguity.
@@
 As a delimitation we only consider languages with balanced parentheses.
-Our approach centers around the idea of the language of a parse tree; since a parse tree can be written in multiple ways we can usefully describe it as a language.
-This language turns out to be visibly pushdown \cite{alurVisiblyPushdownLanguages2004}, which gives us the tools solve the dynamic problem and present unambiguous programs.
-For the static problem we divide languages along two axes: whether they have non-grouping parentheses, and whether they have marks.
-We then present a novel encoding of the language of a tree that allows us to create an automaton that recognizes the language of languages of trees.
-Examining this automaton solves the static problem for a language subclass: languages with neither non-grouping parentheses nor marks.
-We also show that adding marks to a resolvably ambiguous language in this subclass preserves resolvability.
+Our approach builds on the following key insights: i) the language of a parse tree is visibly pushdown \cite{alurVisiblyPushdownLanguages2004}, ii) these languages can be encoded as words, and iii) the language of the word encodings of all trees for a language is also visibly pushdown.
+The last point allows the construction of a visibly pushdown automaton (VPDA) that we examine to solve the static problem for a particular language subclass.
+Our solution to the dynamic problem, which builds on the first point, is general.
 
@@
 Pre-existing systems commonly solve this either by strict requirements on the individual languages \cite{vanwykSilverExtensibleAttribute2010} or by largely ignoring the problem and merely presenting the multiple syntax trees in case of an ambiguity \cite{lorenzenSoundTypedependentSyntactic2016,danielssonParsingMixfixOperators2011}.
-The former case gives good guarantees for the end-user experience (no ambiguities) while the former gives more freedom to the language designer.
+The former case gives good guarantees for the end-user experience (no ambiguities) while the latter gives more freedom to the language designer.
 Resolvable ambiguity, as presented in this paper, provides a middle ground with more language designer freedom than the former and a better user experience than the latter.
@@
 Precedence in Haskell is defined as a number between 0 and 9, meaning that operators from different libraries have a defined relative precedence.
-If the libraries were designed by different authors this precedence is essentially by accident and it may or may not be sensible.
+If the libraries are designed by different authors this precedence is essentially by accident and it may or may not be sensible.
 Leaving the relative precedence undefined would not affect expressivity, every possible combination of operators can be written by simply adding parentheses, and would avoid surprising interpretations.
@@
 
-\noindent We can now state a few things:
+\noindent We can now make the following observations:
 
@@
 Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace.
-In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the parse tree.
+In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguous program.
 
@@
 
-\section{Static Resolvability Analysis} \label{sec:static}
-
-For this section, we will use an alternative formulation of resolvable ambiguity for languages, stated in terms of parse trees instead of words:
-
-\begin{lemma}
-  Given an abstract parser $P = (\T, T, \parse)$, the language $L(P)$ is resolvably ambiguous iff $\quad\forall t \in T.\ \exists w \in \T^{*}.\ \parse(w) = \{t\}$.
-\end{lemma}
-
-\noindent To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w \in L(G'_D)$ for which $\parse(w) = \{t\}$, or prove that no such tree exists.
-Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.
-
-Our approach finds this counterexample by finding a pair of trees $t_1$ and $t_2$ such that $\words(t_1) \subseteq \words(t_2)$.
-Such a pair trivially implies that $t_1$ has no unambiguous words, since $\forall w.\ t_1 \in \parse(w) -> t_2 \in \parse(w)$, which thus implies that the examined language is unresolvably ambiguous.
-The converse is less obvious, in fact, the absence of such a pair does not guarantee a resolvably ambiguous language in general.
-We shall however show it to be sufficient for a limited class of languages.
-
 Central to our approach is the shape of $\words(t)$.
@@
 Its language is thus represented by $\{(^{n_1}(^{n_2}(^{n_3}1)^{n_3} + (^{n_4}2)^{n_4})^{n_2} * (^{n_5}2)^{n_5})^{n_1} \mid n_i \in \{0, 1, \ldots\}, n_2 \geq 1\}$.
-We can encode this language as a word, if we take ``$\reqp{}$'' to mean ``exactly one pair of parentheses'' and ``$\posp{}$'' to mean ``zero or more pairs of parentheses'': ``$\posp{\posp{\reqp{\posp{1} + \posp{2}}} * \posp{3}}$''.
+To have something more manageable we introduce an alternative representation: a linear encoding as a word.
+This representation is convenient for Section~\ref{sec:dynamic} and essential for Section~\ref{sec:static}.
+
+Continuing with the example above, we can encode this language as a word, if we take ``$\reqp{}$'' to mean ``exactly one pair of parentheses'' and ``$\posp{}$'' to mean ``zero or more pairs of parentheses'': ``$\posp{\posp{\reqp{\posp{1} + \posp{2}}} * \posp{3}}$''.
 This encoding lets us reduce comparisons between languages to comparisons between words, with one caveat: the encoding is not unique.
 For example, ``$\reqp{\posp{}}$'' encodes the same language as ``$\posp{\reqp{}}$'', and ``$\posp{\posp{}}$'' encodes the same language as ``$\posp{}$''.
-We rectify this by repeatedly swapping ``$\reqp{\posp{\ldots}}$'' with ``$\posp{\reqp{\ldots}}$'' and ``$\posp{\posp{\ldots}}$'' with ``$\posp{\ldots}$'' until we get stuck.
-We call the result the canonical linear encoding of the (word) language of a tree in $L(T_D)$, which \emph{is} unique.
-
-We now divide the possible languages along two axes: whether a language has (non-grouping) parentheses, and whether a language has marks.
-The parentheses present in the canonical encoding will have the following shapes:
-
-\begin{center}
-\begin{tabular}{r@{\quad}c@{\quad}c}
-  & Parentheses & No parentheses \\
-  \addlinespace
-  %% \midrule
-  Marks & $\pospl^1\reqpl^{\mathbb{N}}$ or $\pospl^0\reqpl^{\mathbb{Z}_{+}}$ & $\pospl^1\reqpl^{\mathbb{N}}$ \\
-  No marks & $\pospl^1\reqpl^{\mathbb{N}}$ or $\pospl^0\reqpl^{\mathbb{Z}_{+}}$ & $\pospl^1\reqpl^0$ \\
-\end{tabular}
-\end{center}
-
-\noindent The absence of both non-grouping parentheses and marks mean that every occurrence of a pair of parentheses must permit zero or more pairs.
-Adding marks adds required pairs for some trees, turning each occurrence to one of $k$ or more pairs, for some non-negative integer $k$ determined by the tree.
-The left column is more complicated since non-grouping parentheses can introduce occurrences that require an exact number of parentheses, as well as all the complications of marks.
-
-The remainder of this section will be devoted to first showing that $\neg \exists t_1, t_2 \in L(T_D).\ \words(t_1) \subseteq \words(t_2)$ implies that $D$ is resolvably ambiguous if $D$ is in the lower-right quadrant, i.e., if it has no marks and no non-grouping parentheses (in Section~\ref{sec:static-proofs}), and then describing the automaton that forms the basis for our algorithm (Section~\ref{sec:lattice-vpl}).
-Section~\ref{sec:static-proofs} also shows that adding marks to a resolvably ambiguous language in the lower-right quadrant preserves resolvability, which gives us a conservative approach for languages in the top-right quadrant.
-Intuitively, if a language has marks, but they are just conveniences, then we can still detect that the language is resolvable.
-For example, most expression languages fall in this category (we can write any expression unambiguously by adding parentheses everywhere, even without any precedence or associativity), while lists in OCaml do not (no amount of parentheses will make \verb|'[1;2]'| look like a list of two elements without a mark).
-
-\subsection{The Road to Correctness} \label{sec:static-proofs} % TODO: probably want a better title here
-
-This section shows that $\neg \exists t_1, t_2 \in L(T_D).\ \words(t_1) \subseteq \words(t_2)$ implies that $D$ is resolvably ambiguous if $D$ has no non-grouping parentheses and no marks.
-These requirements can be written more formally as follows:
-
-\begin{enumerate}
-  \item $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.
-  \item Every non-terminal $N_m$ on the right hand side of a productin in $D$ has $m = \emptyset$.
-\end{enumerate}
-
-\noindent Together, these imply that the canonical linear encoding of a tree has no ``$\reqp{}$'' (required parentheses), only ``$\posp{}$'' (optional parentheses).
-
-Consider an arbitrary $t \in L(T_D)$ with canonical linear encoding $c$.
-Now replace each pair ``$\posp{}$'' with exactly one pair of parentheses, calling the result $w_\top$.
-We have $w_\top \in \words(t)$ by construction.
-Now consider the two possibilities of ambiguity for $w_\top$:
-
-\begin{description}
-  \item[$w_\top$ is ambiguous.] In this case, $\exists t'.\ t \neq t' \land t' \in \parse(w_\top)$.
-This means that we can take the canonical encoding of $t'$ (call it $c'$) and replace every ``$[]$'' with either one or zero parentheses and produce $w_\top$.
-But that also means that every word we construct from $c$ (by choosing some non-negative integer of parentheses for each ``$[]$'') can also be constructed from $c'$ (by choosing the same number for corresponding parentheses, and zero for the others), i.e., $\words(t) \subseteq \words(t')$.
-  \item[$w_\top$ is unambiguous.] In this case $\words(t)$ has at least one unambiguous word ($w_\top$) which thus cannot be part of $\words(t')$ for any other $t' \in L(T_D)$, thus $\neg \exists t'.\ t \neq t' \land \words(t) \subseteq \words(t')$.
-\end{description}
-
-\noindent We thus arrive at the first of two central theorems for our approach:
-
-\begin{theorem}
-  Given a language with productions $D$ and terminals $\T$, where $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$ and all non-terminals $N_m$ appearing in a right-hand side of a production in $D$ having $m = \emptyset$, the following holds: $\rho_P(D) <-> \neg \exists t_1, t_2 \in L(T_D).\ t_1 \neq t_2 \land \words(t_1) \subseteq \words(t_2)$.
-\end{theorem}
-
-Adding marks to such a language preserves resolvability, since:
-
-\begin{itemize}
-  \item Marks can introduce at most one required pair of parentheses per node.
-  \item Adding marks to a language can only ever shrink $\words(t)$ for any particular $t \in L(T_D)$.
-  \item Given a tree $t$, the word produced by putting one pair of parentheses around each node in $t$ (thus potentially adding double parentheses), call it $w$, is unambiguous in the language without marks.
-It cannot be excluded from $\words(t)$ by adding marks (since it already has a pair of parentheses around each node), and it also cannot be ambiguous, since that would require $\words(t')$ for some other tree $t'$ to have grown.
-Thus, $\parse(w) = \{t\}$ even when we add marks to $D$.
-\end{itemize}
-
-\begin{theorem}
-  Adding marks to a resolvably ambiguous language $D$ preserves resolvability, if $D$ had no parentheses and no marks.
-\end{theorem}
-
-\subsection{The Algorithm} \label{sec:lattice-vpl}
-
-This section describes the construction of a visibly pushdown automaton that we examine to determine if a language has a pair of trees where the language of one entirely contains the other.
-We begin by outlining the approach and then describe it in more detail.
-
-The automaton we examine recognizes canonical linear encodings and has a one-to-one correspondence between runs and trees.
-Two distinct runs that recognize the same word thus corresponds to two distinct trees that have the \emph{same} word language.
-This is essentially the question of whether a visibly pushdown automaton is ambiguous, which we can detect by constructing a product automaton and trimming it.
-
-To detect subsumption we make the following observation: given two trees $t_1$ and $t_2$ with canonical encodings $c_1$ and $c_2$, respectively, we have $\words(t_1) \subset \words(t_2)$ iff we can add arbitrary well-nested ``$\posp{}$'' pairs to $c_1$, producing a $c_1'$ such that $c_1' = c_2$ (Note that we do not need to consider ``$\reqp{}$'' since no such pairs appear in this language class).
-We thus make a product automaton of two slightly different automata, where the second may add arbitrary ``$\posp{}$'', but is otherwise identical.
-
-\subsubsection{Illustration by Example}
-
-We begin by constructing a visibly pushdown automaton that recognizes \emph{a} linear encoding for each tree, then modify it to only recognize the \emph{canonical} linear encoding.
-To illustrate the approach we consider the following language:
-
-\begin{center}
-  \begin{tabular}{@{}r@{$\;->\;$}c@{\,:\;}l}
-    \toprule
-    $S$ & $a$ & $E(\verb|';'| + \epsilon)$ \\
-    $E$ & $b$ & $E\verb|'s'|$ \\
-    $E$ & $c$ & \verb|'z'| \\
-    \bottomrule
-  \end{tabular}
-\end{center}
-
-\noindent This language has no non-grouping parentheses and no marks and is thus in the language class we consider.
-We begin by constructing a DFA per production (using standard methods, since the right hand side is a regular expression).
-
-\begin{center}
-\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-  \node[state,initial] (1) {$1$};
-  \node[state,accepting] (2) [right of=1] {$2$};
-  \node[state,accepting] (3) [right of=2] {$3$};
-
-  \node[state,initial] (4) [below of=1,yshift=7mm] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-  \node[state,accepting] (6) [right of=5] {$6$};
-
-  \node[state,initial] (7) [below of=4,yshift=7mm] {$7$};
-  \node[state,accepting] (8) [right of=7] {$8$};
-
-  \path (1) edge node{$E$} (2)
-  (2) edge node{$';'$} (3);
-
-  \path (4) edge node{$E$} (5)
-  (5) edge node{$'s'$} (6);
-
-  \path (7) edge node{$'z'$} (8);
-\end{tikzpicture}
-\end{center}
-
-\noindent We then combine these separate DFAs to produce a single visibly pushdown automaton in three steps:
-
-\begin{enumerate}
-\item Introduce two new distinct states (call them $s$ and $f$) and a transition $s \xrightarrow{S} f$, where $S$ is the starting symbol of the language being examined.
-\item Replace each transition $p \xrightarrow{N} q$ with:
-  \begin{itemize}
-  \item A transition $p \xrightarrow{'\pospl', +(p,q)} q_0$ for every initial state $q_0$ in a DFA corresponding to a production with lefthand side $N$.
-  \item A transition $q_f \xrightarrow{'\pospr', -(p,q)} q$ for every accepting state $q_f$ in a DFA corresponding to a production with lefthand side $N$.
-  \end{itemize}
-\item Make $s$ the only initial state, and $f$ the only accepting state.
-\end{enumerate}
-
-\noindent To make the figure less cluttered we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')} q$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')} q$.
-
-\begin{center}
-\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-
-  \node[state] (4) [below of=1] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-  \node[state] (6) [right of=5] {$6$};
-
-  \node[state] (7) [below of=4] {$7$};
-  \node[state] (8) [right of=7] {$8$};
-
-  \node[state,initial] (s) [left of=1,xshift=-0.9cm] {$s$};
-  \node[state,accepting] (f) [right of=3,xshift=0.9cm] {$f$};
-
-  \path (2) edge node{$';'$} (3);
-
-  \path (5) edge node{$'s'$} (6);
-
-  \path (7) edge node{$'z'$} (8);
-
-  \path (s) edge node{$+(s, f)$} (1);
-  \draw (2.north) -- ([yshift=4mm]2.north) -| (f.north);
-  \draw (3.north) -- ([yshift=4mm]3.north) -| (f.north);
-  \node at ([yshift=5mm]$(3)!0.5!(f)$) {$-(s,f)$};
-
-  % 1 -E-> 2
-  \draw (1.east) -| ([xshift=4mm]4.east) |- ([xshift=4mm,yshift=10mm]7.east) -- (7);
-  \draw ([xshift=4mm]4.east) -- (4);
-  \draw (6.north) |- ([xshift=-8mm]$(2)!0.5!(5)$) |- (2.west);
-  \draw (8) -- ([xshift=-3.5mm,yshift=10mm]8.west) |- (2.west);
-  \node at ([xshift=2mm,yshift=2mm]$(1)!0.5!(4)$) {$+(1,2)$};
-  \node at ([xshift=-2mm,yshift=2mm]$(2)!0.5!(5)$) {$-(1,2)$};
-
-  % 4 -E-> 5
-  \draw (4.south) -- (7.north);
-  \path (4) edge[out=270,in=225,looseness=5] node[xshift=4mm]{$+(4,5)$} (4);
-  \draw (6.south) -- ([yshift=-5mm]6.south) -| (5.south);
-  \path (8) edge[right] node[yshift=-3mm]{$-(4,5)$} (5);
-\end{tikzpicture}
-\end{center}
-
-\noindent This automaton recognizes a linear encoding by simply putting a pair $'\posp{}'$ around each production.
-That encoding will be canonical in many cases (e.g.
-``$\posp{\posp{\posp{z}s};}$''), but not all; ``$\posp{\posp{z}}$'', which is recognized for the tree $a(c(z))$, is not canonical, it should be ``$\posp{z}$''.
-The extra $'\posp{}'$ pair stems from $S -> a: E(\verb|';'| + \epsilon)$, its righthand side matches $E$ (i.e., a word consisting of a single non-terminal).
-
-We solve this by first recording which productions have righthand sides that match single non-terminal words (and which non-terminal) and then compute DFAs that do \emph{not} match such words.
-For our running example we note that the production $a$ matches the word $E$ (we record this as a tuple $(S, a, E)$) and replace its automaton with one that does not match $E$.
-In this case it is sufficient to mark state $2$ as non-accepting.
-
-The recorded tuple $(S, a, E)$ can be read as ``starting from non-terminal $S$ we can match the word $E$ by choosing the production with label $a$.'' If we consider the middle element as a sequence of labels we can build a relation by adding $(N, \epsilon, N)$ for all non-terminals $N$ and computing the closure of $(A, w, B) \cdot (B, w', C) = (A, w \cdot w', C)$.
-Call the resulting set $T$.
-
-Finally, we change step 2 in our construction to the following:
-
-\begin{enumerate}
-  \setcounter{enumi}{1}
-\item Replace each transition $p \xrightarrow{N} q$ with:
-  \begin{itemize}
-  \item A transition $p \xrightarrow{'\pospl', +(p,q,w)} q_0$, for every initial state $q_0$ in a DFA corresponding to a production with lefthand side $N'$, for every $(N, w, N') \in T$.
-  \item A transition $q_f \xrightarrow{'\pospr', -(p,q,w)} q$, for every accepting state $q_f$ in a DFA corresponding to a production with lefthand side $N'$, for every $(N, w, N') \in T$.
-  \end{itemize}
-\end{enumerate}
-
-\begin{center}
-\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
-  \node[state] (1) {$1$};
-  \node[state] (2) [right of=1] {$2$};
-  \node[state] (3) [right of=2] {$3$};
-
-  \node[state] (4) [below of=1] {$4$};
-  \node[state] (5) [right of=4] {$5$};
-  \node[state] (6) [right of=5] {$6$};
-
-  \node[state] (7) [below of=4] {$7$};
-  \node[state] (8) [right of=7] {$8$};
-
-  \node[state,initial] (s) [left of=1,xshift=-0.9cm] {$s$};
-  \node[state,accepting] (f) [right of=3,xshift=0.9cm] {$f$};
-
-  \path (2) edge node{$';'$} (3);
-
-  \path (5) edge node{$'s'$} (6);
-
-  \path (7) edge node{$'z'$} (8);
-
-  \path (s) edge node{$+(s, f, \epsilon)$} (1)
-  (3) edge node{$-(s, f, \epsilon)$} (f);
-
-  % 1 -E-> 2
-  \draw (1.east) -| ([xshift=4mm]4.east) |- ([xshift=4mm,yshift=10mm]7.east) -- (7);
-  \draw ([xshift=4mm]4.east) -- (4);
-  \draw (6.north) |- ([xshift=-8mm]$(2)!0.5!(5)$) |- (2.west);
-  \draw (8) -- ([xshift=-3.5mm,yshift=10mm]8.west) |- (2.west);
-  \node at ([xshift=1mm,yshift=2mm]$(1)!0.5!(4)$) {$+(1,2,\epsilon)$};
-  \node at ([xshift=-1mm,yshift=2mm]$(2)!0.5!(5)$) {$-(1,2,\epsilon)$};
-
-  % 4 -E-> 5
-  \draw (4.south) -- (7.north);
-  \path (4) edge[out=270,in=225,looseness=5] node[xshift=4mm]{$+(4,5,\epsilon)$} (4);
-  \draw (6.south) -- ([yshift=-5mm]6.south) -| (5.south);
-  \path (8) edge[right] node[yshift=-3mm]{$-(4,5,\epsilon)$} (5);
-
-  % s -E(t)-> f
-  \draw (s.south) |- (7.west);
-  \draw (s.south) |- (4.west);
-  \node at ([xshift=8mm,yshift=-10mm]s) {$+(s, f, a)$};
-  \draw (8.east) -| (f.south);
-  \draw (6.east) -| (f.south);
-  \node at ([xshift=-8mm,yshift=-10mm]f) {$-(s, f, a)$};
-\end{tikzpicture}
-\end{center}
-
-\noindent We call this automaton $A_{\posp{}}$.
-We also construct a modified automaton that recognizes linear encodings of larger word languages.
-We call this automaton $A'_{\posp{}}$ and obtain it by adding transitions $p \xrightarrow{'\pospl', +\gamma} p$ and $p \xrightarrow{'\pospr', -\gamma} p$ for every state $p$, where $\gamma$ is a new distinct stack symbol.
-Note that all these new transition push and pop the \emph{same} stack symbol, meaning that a new pair ``$\posp{}$'' can be started in any state and then closed in any state (not just the same state) as long as intermediate transitions leave the stack unchanged, i.e., only introduce zero or more well-balanced pairs.
-
-\subsubsection{Formalization}
-
-Given a language definition $D$, with terminals $\T$, non-terminals $\NT$, labels $\Labels$, and starting non-terminal $S$.
-We use $\labelnt(l)$ and $\labelregex(l)$ as the non-terminal and regex, respectively, of the production labelled $l$.
-We define the set $T : \NT \times \Labels^{*} \times \NT$ of transitions between productions inductively through:
-
-\begin{itemize}
-\item $(N, \epsilon, N) \in T$ for all $N \in \NT$.
-\item $(N, l, N') \in T$ for all $N' \in \NT \cap L(\labelregex(l))$ where $\labelnt(l) = N$.
-\item $(N_1, w_1 \cdot w_2, N_3) \in T$ for all $(N_1, w_1, N_2), (N_2, w_2, N_3) \in T$.
-\end{itemize}
-
-% TODO: footnote should probably be removed
-\noindent $T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
-Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces (infinitely) ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
-
-We define the translation from production to DFA through a function $\dfa$ from a label to a DFA with the alphabet $\T \cup \Labels$:
-
-\begin{description}
-\item[$\dfa(l)$:] Compute the regular language $L' := L(\labelregex(l)) \setminus \NT$, then construct a DFA for this language.
-\end{description}
-
-\noindent We now define the visibly pushdown automaton $A_{\posp{}}$.
-Given $\dfa(l) = (Q_l, \T \cup \NT, \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
-
-\begin{itemize}
-\item $s$ and $f$ are two new distinct states.
-\item $Q = \{s, f\} \cup \bigsqcup_l Q_l$.
-\item $\T' = \T \cup \{'\pospl', '\pospr'\}$.
-\item $\Gamma = Q \times Q \times \Labels^{*}$.
-\item $\delta$ is defined by the following equations (unspecified cases produce $\emptyset$):
-  $$
-  \begin{array}{r@{\;=\;}l@{\;\mid\;}l}
-    \delta(p, a, \lambda)
-    & \multicolumn{2}{@{}l@{}}{\{(\delta_l(p, a), \lambda)\} \qquad \text{ where } p \in Q_l \text{ and } a \in \T} \\
-
-    \delta(s, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l}, (s, f, w))
-    & (S, w, N) \in T, N = \labelnt(l) \} \\
-
-    \delta(p, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, N) = q, (N, w, N') \in T, N' = \labelnt(l_3) \} \\
-
-    \delta(p, \text{``}\pospr\text{''}, (s, f, w))
-    & \{ (f, \lambda)
-    & (S, w, N) \in T, N = \labelnt(l), p \in F_l \} \\
-
-    \delta(q', \text{``}\pospr\text{''}, (p, q, w))
-    & \{ (q, \lambda)
-    & q' \in F_{l_1}, \labelnt(l_1) = N', (N, w, N') \in T, p \in Q_{l_2}, \delta_{l_2}(p, N) = q \} \\
-
-  \end{array}
-  $$
-\end{itemize}
-
-\noindent We also construct a modified VPDA $A'_{\posp{}} = (Q, \T', \Gamma \cup \{\gamma\}, \delta', s, \{f\})$ where:
-
-\begin{itemize}
-\item $\gamma$ is a new distinct stack symbol.
-\item $\delta'(p, a, g) = \delta(p, a, g) \cup \delta''(p, a, g)$, where $\delta''$ is given by:
-  $$
-  \begin{array}{r@{\;=\;}l@{}}
-    \delta''(p, \text{``}\pospl\text{''}, \lambda)
-    & \{ (p, \gamma) \} \\
-
-    \delta''(p, \text{``}\pospr\text{''}, \gamma)
-    & \{ (p, \lambda) \} \\
-  \end{array}
-  $$
-\end{itemize}
-
-\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is described in \cite{alurVisiblyPushdownLanguages2004} (using the partitions $\T_c = \{ \pospl \}$, $\T_i = \T$, and $\T_r = \{\pospr\}$).
-If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct successful runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
-We check for the existence of such a run by trimming the product automaton (as described in \cite{caralpTrimmingVisiblyPushdown2015}) and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
-
-Performing this procedure on a language definition $D$ without non-grouping parentheses and marks gives a sound and complete decision procedure for determining if $D$ is resolvably ambiguous.
-If we instead take a language definition $D$ with marks but no non-grouping parentheses, then remove the marks and perform the procedure, then the absence of two such runs implies that $D$ is resolvably ambiguous, while their presence only implies the possibility of an unresolvable ambiguity; the marks might remove it.
-
-\begin{lemma}
-  There is a bijection between successful runs in $A_{\posp{}}$ and trees $t \in L(T_D)$.
-\end{lemma}
-
-\begin{proof}
-  By defining two functions $\treetorun$ and $\runtotree$, the former from a tree to a run, the latter from run to tree, then showing these two functions to be inverses of each other.
-\end{proof}
-
-We denote the word recognized by a run $R$ by $\runtoword(R)$.
-
-\begin{lemma}
-  $\forall t \in L(T_D).\ \runtoword(\treetorun(t))$ is the canonical linear encoding of $\words(t)$.
-\end{lemma}
-
-\begin{lemma}
-  For all $t \in L(T_D)$ and every linear encoding $c$ such that $\words(t) \subseteq L(c)$, $c \in L(A'_{\posp{}})$.
-\end{lemma}
-
-\begin{lemma}
-  We can define a function $\runprimetorun$ from runs in $A'_{\posp{}}$ to runs in $A_{\posp{}}$ such that $\forall R.\ L(\runtoword(\runprimetorun(R))) \subseteq L(\runtoword(R))$.
-\end{lemma}
-
-\begin{proof}
-  By removing every transition that pushes or pops $\gamma$.
-\end{proof}
-
-\begin{theorem}
-  The existence of two distinct runs $R$ and $R'$ such that $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, and $\runtoword(R) = \runtoword(R')$, implies the existence of two distinct trees such that the word language of one is entirely contained in the other.
-\end{theorem}
-
-\begin{proof}
-  Assume $\runprimetorun(R') = R$.
-$\runprimetorun$ either leaves its argument unchanged, or changes the recognized word.
-If $R'$ and $R$ recognize the same word, and $\runprimetorun(R') = R$, then $R' = R$, but that is a contradiction.
-
-  Since $\runprimetorun(R') \neq R$ we have two distinct trees $t_1 = \runtotree(\runprimetorun(R'))$ and $t_2 = \runtotree(R)$.
-Since $L(\runtoword(\runprimetorun(R'))) \subseteq L(\runtoword(R'))$ and $L(\runtoword(R')) = L(\runtoword(R))$ we have that $L(t_1) \subseteq L(t_2)$.
-\end{proof}
-
-We thus have a way to detect the presence or absence of a pair of trees $t_1, t_2 \in L(T_D)$ such that $\words(t_1) \subseteq \words(t_2)$, which completely determines the resolvability of $D$ when $D$ has no non-grouping parentheses or marks.
+We rectify this by repeatedly swapping ``$\reqp{\posp{\ldots}}$'' with ``$\posp{\reqp{\ldots}}$'' and ``$\posp{\posp{\ldots}}$'' with ``$\posp{\ldots}$'' until we reach a fixpoint.
+We call the result the \emph{canonical linear encoding} of the (word) language of a tree in $L(T_D)$, which \emph{is} unique.
 
@@
 Furthermore, for practical reasons, if the word is resolvably ambiguous we wish to produce a (minimal) witness for each tree.
-Similarly to Section~\ref{sec:static}, we place some restrictions on $D$, though the restrictions are less severe.
-While Section~\ref{sec:static} requires $D$ to have no non-grouping parentheses and no marks, this section has two much less stringent requirements:
+We place two restriction on languages $D$ we consider:
 
 \begin{enumerate}
-\item Each right-hand side regular expression must only recognize words with balanced parentheses. For example, ``$\verb|(| \verb|)|$'' is permissible, but ``$\verb|(|^{*} \verb|)|^{*}$'' is not.
+\item Stemming from our initial delimitation, each right-hand side regular expression must only recognize words with balanced parentheses.
+For example, ``$\verb|(| \verb|)|$'' is permissible, but ``$\verb|(|^{*} \verb|)|^{*}$'' is not.
 \item $G_D$, but with parentheses removed, must not be infinitely ambiguous.
@@
 
-Similarly to Section~\ref{sec:static} the approach centers around the word language $\words(t)$.
-We can easily construct a visibly pushdown automaton for this language by going via the canonical linear encoding: let every pair ``$\reqp{}$'' push and pop the same stack symbol (call it $\gamma$) and let every ``$\posp{}$'' push and pop a unique stack symbol.
+Our approach centers around the construction of a visibly pushdown automaton recognizing $\words(t)$ for any particular $t \in L(T_D)$.
+We can easily construct this automaton via the canonical linear encoding: let every pair ``$\reqp{}$'' push and pop the same stack symbol (call it $\gamma$) and let every ``$\posp{}$'' push and pop a unique stack symbol.
 For example, ``$\posp{\reqp{a}\posp{b}}$'' produces the following automaton:
@@
 
+\section{Static Resolvability Analysis} \label{sec:static}
+
+To determine if a given language definition $D$ is resolvably ambiguous we attempt to find a counterexample: a tree $t \in L(T_D)$ such that there is no $w \in L(G'_D)$ for which $\parse(w) = \{t\}$, or prove that no such tree exists.
+Or, more briefly put: find a tree that has only ambiguous words or show that no such tree exists.
+
+Our approach finds this counterexample by finding a pair of trees $t_1$ and $t_2$ such that $\words(t_1) \subseteq \words(t_2)$.
+Such a pair trivially implies that $t_1$ has no unambiguous words, since $\forall w.\ t_1 \in \parse(w) -> t_2 \in \parse(w)$, which thus implies that the examined language is unresolvably ambiguous.
+The converse is less obvious, in fact, the absence of such a pair does not guarantee a resolvably ambiguous language in general.
+We shall however show it to be sufficient for a limited class of languages.
+
+We now divide the possible languages along two axes: whether a language has (non-grouping) parentheses, and whether a language has marks.
+The parentheses present in the canonical encoding will have the following shapes:
+
+\begin{center}
+\begin{tabular}{r@{\quad}c@{\quad}c}
+  & Parentheses & No parentheses \\
+  \addlinespace
+  Marks & $\pospl^1\reqpl^{\mathbb{N}}$ or $\pospl^0\reqpl^{\mathbb{Z}_{+}}$ & $\pospl^1\reqpl^{\mathbb{N}}$ \\
+  No marks & $\pospl^1\reqpl^{\mathbb{N}}$ or $\pospl^0\reqpl^{\mathbb{Z}_{+}}$ & $\pospl^1\reqpl^0$ \\
+\end{tabular}
+\end{center}
+
+\noindent The absence of both non-grouping parentheses and marks mean that every occurrence of a pair of parentheses must permit zero or more pairs.
+Adding marks adds required pairs for some trees, turning each occurrence to one of $k$ or more pairs, for some non-negative integer $k$ determined by the tree.
+The left column is more complicated since non-grouping parentheses can introduce occurrences that require an exact number of parentheses, as well as all the complications of marks.
+
+The remainder of this section will be devoted to first showing that $\neg \exists t_1, t_2 \in L(T_D).\ \words(t_1) \subseteq \words(t_2)$ implies that $D$ is resolvably ambiguous if $D$ is in the lower-right quadrant, i.e., if it has no marks and no non-grouping parentheses (in Section~\ref{sec:static-proofs}), and then describing the automaton that forms the basis for our algorithm (Section~\ref{sec:lattice-vpl}).
+Section~\ref{sec:static-proofs} additionally shows that adding marks to a resolvably ambiguous language in the lower-right quadrant preserves resolvability, which gives us a conservative approach for languages in the top-right quadrant.
+Intuitively, if a language has marks, but they are just conveniences, then we can still detect that the language is resolvable.
+For example, most expression languages fall in this category (we can write any expression unambiguously by adding parentheses everywhere, even without any precedence or associativity), while lists in OCaml do not (no amount of parentheses will make \verb|'[1;2]'| look like a list of two elements without a mark).
+
+\subsection{The Road to Correctness} \label{sec:static-proofs} % TODO: probably want a better title here
+
+This section shows that $\neg \exists t_1, t_2 \in L(T_D).\ \words(t_1) \subseteq \words(t_2)$ implies that $D$ is resolvably ambiguous if $D$ has no non-grouping parentheses and no marks.
+These requirements can be written more formally as follows:
+
+\begin{enumerate}
+  \item $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$.
+  \item Every non-terminal $N_m$ on the right hand side of a productin in $D$ has $m = \emptyset$.
+\end{enumerate}
+
+\noindent Together, these imply that the canonical linear encoding of a tree has no ``$\reqp{}$'' (required parentheses), only ``$\posp{}$'' (optional parentheses).
+
+Consider an arbitrary $t \in L(T_D)$ with canonical linear encoding $c$.
+Now replace each pair ``$\posp{}$'' with exactly one pair of parentheses, calling the result $w_\top$.
+We have $w_\top \in \words(t)$ by construction.
+Now consider the two possibilities of ambiguity for $w_\top$:
+
+\begin{description}
+  \item[$w_\top$ is ambiguous.] In this case, $\exists t'.\ t \neq t' \land t' \in \parse(w_\top)$.
+This means that we can take the canonical encoding of $t'$ (call it $c'$) and replace every ``$[]$'' with either one or zero parentheses and produce $w_\top$.
+But that also means that every word we construct from $c$ (by choosing some non-negative integer of parentheses for each ``$[]$'') can also be constructed from $c'$ (by choosing the same number for corresponding parentheses, and zero for the others), i.e., $\words(t) \subseteq \words(t')$.
+  \item[$w_\top$ is unambiguous.] In this case $\words(t)$ has at least one unambiguous word ($w_\top$) which thus cannot be part of $\words(t')$ for any other $t' \in L(T_D)$, thus $\neg \exists t'.\ t \neq t' \land \words(t) \subseteq \words(t')$.
+\end{description}
+
+\noindent We thus arrive at the first of two central theorems for our approach:
+
+\begin{theorem}
+  Given a language with productions $D$ and terminals $\T$, where $\T \cap \{\verb|'('|, \verb|')'|\} = \emptyset$ and all non-terminals $N_m$ appearing in a right-hand side of a production in $D$ having $m = \emptyset$, the following holds: $\rho_P(D) <-> \neg \exists t_1, t_2 \in L(T_D).\ t_1 \neq t_2 \land \words(t_1) \subseteq \words(t_2)$.
+\end{theorem}
+
+Adding marks to such a language preserves resolvability, since:
+
+\begin{itemize}
+  \item Marks can introduce at most one required pair of parentheses per node.
+  \item Adding marks to a language can only ever shrink $\words(t)$ for any particular $t \in L(T_D)$.
+  \item Given a tree $t$, the word produced by putting one pair of parentheses around each node in $t$ (thus potentially adding double parentheses), call it $w$, is unambiguous in the language without marks.
+It cannot be excluded from $\words(t)$ by adding marks (since it already has a pair of parentheses around each node), and it also cannot be ambiguous, since that would require $\words(t')$ for some other tree $t'$ to have grown.
+Thus, $\parse(w) = \{t\}$ even when we add marks to $D$.
+\end{itemize}
+
+\begin{theorem}
+  Adding marks to a resolvably ambiguous language $D$ preserves resolvability, if $D$ had no parentheses and no marks.
+\end{theorem}
+
+\subsection{The Algorithm} \label{sec:lattice-vpl}
+
+This section describes the construction of a visibly pushdown automaton that we examine to determine if a language has a pair of trees where the language of one entirely contains the other.
+We begin by outlining the approach and then describe it in more detail.
+
+The automaton we examine recognizes canonical linear encodings and has a one-to-one correspondence between runs and trees.
+Two distinct runs that recognize the same word thus corresponds to two distinct trees that have the \emph{same} word language.
+This is essentially the question of whether a visibly pushdown automaton is ambiguous, which we can detect by constructing a product automaton and trimming it.
+
+To detect subsumption we make the following observation: given two trees $t_1$ and $t_2$ with canonical encodings $c_1$ and $c_2$, respectively, we have $\words(t_1) \subset \words(t_2)$ iff we can add arbitrary well-nested ``$\posp{}$'' pairs to $c_1$, producing a $c_1'$ such that $c_1' = c_2$ (Note that we do not need to consider ``$\reqp{}$'' since no such pairs appear in this language class).
+We thus make a product automaton of two slightly different automata, where the second may add arbitrary ``$\posp{}$'', but is otherwise identical.
+
+\subsubsection{Illustration by Example}
+
+We begin by constructing a visibly pushdown automaton that recognizes \emph{a} linear encoding for each tree, then modify it to only recognize the \emph{canonical} linear encoding.
+To illustrate the approach we consider the following language:
+
+\begin{center}
+  \begin{tabular}{@{}r@{$\;->\;$}c@{\,:\;}l}
+    \toprule
+    $S$ & $a$ & $E(\verb|';'| + \epsilon)$ \\
+    $E$ & $b$ & $E\verb|'s'|$ \\
+    $E$ & $c$ & \verb|'z'| \\
+    \bottomrule
+  \end{tabular}
+\end{center}
+
+\noindent This language has no non-grouping parentheses and no marks and is thus in the language class we consider.
+We begin by constructing a DFA per production (using standard methods, since the right hand side is a regular expression).
+
+\begin{center}
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+  \node[state,initial] (1) {$1$};
+  \node[state,accepting] (2) [right of=1] {$2$};
+  \node[state,accepting] (3) [right of=2] {$3$};
+
+  \node[state,initial] (4) [below of=1,yshift=7mm] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+  \node[state,accepting] (6) [right of=5] {$6$};
+
+  \node[state,initial] (7) [below of=4,yshift=7mm] {$7$};
+  \node[state,accepting] (8) [right of=7] {$8$};
+
+  \path (1) edge node{$E$} (2)
+  (2) edge node{$';'$} (3);
+
+  \path (4) edge node{$E$} (5)
+  (5) edge node{$'s'$} (6);
+
+  \path (7) edge node{$'z'$} (8);
+\end{tikzpicture}
+\end{center}
+
+\noindent We then combine these separate DFAs to produce a single visibly pushdown automaton in three steps:
+
+\begin{enumerate}
+\item Introduce two new distinct states (call them $s$ and $f$) and a transition $s \xrightarrow{S} f$, where $S$ is the starting symbol of the language being examined.
+\item Replace each transition $p \xrightarrow{N} q$ with:
+  \begin{itemize}
+  \item A transition $p \xrightarrow{'\pospl', +(p,q)} q_0$ for every initial state $q_0$ in a DFA corresponding to a production with lefthand side $N$.
+  \item A transition $q_f \xrightarrow{'\pospr', -(p,q)} q$ for every accepting state $q_f$ in a DFA corresponding to a production with lefthand side $N$.
+  \end{itemize}
+\item Make $s$ the only initial state, and $f$ the only accepting state.
+\end{enumerate}
+
+\noindent To make the figure less cluttered we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')} q$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')} q$.
+
+\begin{center}
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+  \node[state] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+  \node[state] (3) [right of=2] {$3$};
+
+  \node[state] (4) [below of=1] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+  \node[state] (6) [right of=5] {$6$};
+
+  \node[state] (7) [below of=4] {$7$};
+  \node[state] (8) [right of=7] {$8$};
+
+  \node[state,initial] (s) [left of=1,xshift=-0.9cm] {$s$};
+  \node[state,accepting] (f) [right of=3,xshift=0.9cm] {$f$};
+
+  \path (2) edge node{$';'$} (3);
+
+  \path (5) edge node{$'s'$} (6);
+
+  \path (7) edge node{$'z'$} (8);
+
+  \path (s) edge node{$+(s, f)$} (1);
+  \draw (2.north) -- ([yshift=4mm]2.north) -| (f.north);
+  \draw (3.north) -- ([yshift=4mm]3.north) -| (f.north);
+  \node at ([yshift=5mm]$(3)!0.5!(f)$) {$-(s,f)$};
+
+  % 1 -E-> 2
+  \draw (1.east) -| ([xshift=4mm]4.east) |- ([xshift=4mm,yshift=10mm]7.east) -- (7);
+  \draw ([xshift=4mm]4.east) -- (4);
+  \draw (6.north) |- ([xshift=-8mm]$(2)!0.5!(5)$) |- (2.west);
+  \draw (8) -- ([xshift=-3.5mm,yshift=10mm]8.west) |- (2.west);
+  \node at ([xshift=2mm,yshift=2mm]$(1)!0.5!(4)$) {$+(1,2)$};
+  \node at ([xshift=-2mm,yshift=2mm]$(2)!0.5!(5)$) {$-(1,2)$};
+
+  % 4 -E-> 5
+  \draw (4.south) -- (7.north);
+  \path (4) edge[out=270,in=225,looseness=5] node[xshift=4mm]{$+(4,5)$} (4);
+  \draw (6.south) -- ([yshift=-5mm]6.south) -| (5.south);
+  \path (8) edge[right] node[yshift=-3mm]{$-(4,5)$} (5);
+\end{tikzpicture}
+\end{center}
+
+\noindent This automaton recognizes a linear encoding by simply putting a pair $'\posp{}'$ around each production.
+That encoding will be canonical in many cases (e.g.
+``$\posp{\posp{\posp{z}s};}$''), but not all; ``$\posp{\posp{z}}$'', which is recognized for the tree $a(c(z))$, is not canonical, it should be ``$\posp{z}$''.
+The extra $'\posp{}'$ pair stems from $S -> a: E(\verb|';'| + \epsilon)$, its righthand side matches $E$ (i.e., a word consisting of a single non-terminal).
+
+We solve this by first recording which productions have righthand sides that match single non-terminal words (and which non-terminal) and then compute DFAs that do \emph{not} match such words.
+For our running example we note that the production $a$ matches the word $E$ (we record this as a tuple $(S, a, E)$) and replace its automaton with one that does not match $E$.
+In this case it is sufficient to mark state $2$ as non-accepting.
+
+The recorded tuple $(S, a, E)$ can be read as ``starting from non-terminal $S$ we can match the word $E$ by choosing the production with label $a$.'' If we consider the middle element as a sequence of labels we can build a relation by adding $(N, \epsilon, N)$ for all non-terminals $N$ and computing the closure of $(A, w, B) \cdot (B, w', C) = (A, w \cdot w', C)$.
+Call the resulting set $T$.
+
+Finally, we change step 2 in our construction to the following:
+
+\begin{enumerate}
+  \setcounter{enumi}{1}
+\item Replace each transition $p \xrightarrow{N} q$ with:
+  \begin{itemize}
+  \item A transition $p \xrightarrow{'\pospl', +(p,q,w)} q_0$, for every initial state $q_0$ in a DFA corresponding to a production with lefthand side $N'$, for every $(N, w, N') \in T$.
+  \item A transition $q_f \xrightarrow{'\pospr', -(p,q,w)} q$, for every accepting state $q_f$ in a DFA corresponding to a production with lefthand side $N'$, for every $(N, w, N') \in T$.
+  \end{itemize}
+\end{enumerate}
+
+\begin{center}
+\begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
+    scale = 1,transform shape]
+  \node[state] (1) {$1$};
+  \node[state] (2) [right of=1] {$2$};
+  \node[state] (3) [right of=2] {$3$};
+
+  \node[state] (4) [below of=1] {$4$};
+  \node[state] (5) [right of=4] {$5$};
+  \node[state] (6) [right of=5] {$6$};
+
+  \node[state] (7) [below of=4] {$7$};
+  \node[state] (8) [right of=7] {$8$};
+
+  \node[state,initial] (s) [left of=1,xshift=-0.9cm] {$s$};
+  \node[state,accepting] (f) [right of=3,xshift=0.9cm] {$f$};
+
+  \path (2) edge node{$';'$} (3);
+
+  \path (5) edge node{$'s'$} (6);
+
+  \path (7) edge node{$'z'$} (8);
+
+  \path (s) edge node{$+(s, f, \epsilon)$} (1)
+  (3) edge node{$-(s, f, \epsilon)$} (f);
+
+  % 1 -E-> 2
+  \draw (1.east) -| ([xshift=4mm]4.east) |- ([xshift=4mm,yshift=10mm]7.east) -- (7);
+  \draw ([xshift=4mm]4.east) -- (4);
+  \draw (6.north) |- ([xshift=-8mm]$(2)!0.5!(5)$) |- (2.west);
+  \draw (8) -- ([xshift=-3.5mm,yshift=10mm]8.west) |- (2.west);
+  \node at ([xshift=1mm,yshift=2mm]$(1)!0.5!(4)$) {$+(1,2,\epsilon)$};
+  \node at ([xshift=-1mm,yshift=2mm]$(2)!0.5!(5)$) {$-(1,2,\epsilon)$};
+
+  % 4 -E-> 5
+  \draw (4.south) -- (7.north);
+  \path (4) edge[out=270,in=225,looseness=5] node[xshift=4mm]{$+(4,5,\epsilon)$} (4);
+  \draw (6.south) -- ([yshift=-5mm]6.south) -| (5.south);
+  \path (8) edge[right] node[yshift=-3mm]{$-(4,5,\epsilon)$} (5);
+
+  % s -E(t)-> f
+  \draw (s.south) |- (7.west);
+  \draw (s.south) |- (4.west);
+  \node at ([xshift=8mm,yshift=-10mm]s) {$+(s, f, a)$};
+  \draw (8.east) -| (f.south);
+  \draw (6.east) -| (f.south);
+  \node at ([xshift=-8mm,yshift=-10mm]f) {$-(s, f, a)$};
+\end{tikzpicture}
+\end{center}
+
+\noindent We call this automaton $A_{\posp{}}$.
+We also construct a modified automaton that recognizes linear encodings of larger word languages.
+We call this automaton $A'_{\posp{}}$ and obtain it by adding transitions $p \xrightarrow{'\pospl', +\gamma} p$ and $p \xrightarrow{'\pospr', -\gamma} p$ for every state $p$, where $\gamma$ is a new distinct stack symbol.
+Note that all these new transition push and pop the \emph{same} stack symbol, meaning that a new pair ``$\posp{}$'' can be started in any state and then closed in any state (not just the same state) as long as intermediate transitions leave the stack unchanged, i.e., only introduce zero or more well-balanced pairs.
+
+\subsubsection{Formalization}
+
+Given a language definition $D$, with terminals $\T$, non-terminals $\NT$, labels $\Labels$, and starting non-terminal $S$.
+We use $\labelnt(l)$ and $\labelregex(l)$ as the non-terminal and regex, respectively, of the production labelled $l$.
+We define the set $T : \NT \times \Labels^{*} \times \NT$ of transitions between productions inductively through:
+
+\begin{itemize}
+\item $(N, \epsilon, N) \in T$ for all $N \in \NT$.
+\item $(N, l, N') \in T$ for all $N' \in \NT \cap L(\labelregex(l))$ where $\labelnt(l) = N$.
+\item $(N_1, w_1 \cdot w_2, N_3) \in T$ for all $(N_1, w_1, N_2), (N_2, w_2, N_3) \in T$.
+\end{itemize}
+
+% TODO: footnote should probably be removed
+\noindent $T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
+Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces (infinitely) ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
+
+We define the translation from production to DFA through a function $\dfa$ from a label to a DFA with the alphabet $\T \cup \Labels$:
+
+\begin{description}
+\item[$\dfa(l)$:] Compute the regular language $L' := L(\labelregex(l)) \setminus \NT$, then construct a DFA for this language.
+\end{description}
+
+\noindent We now define the visibly pushdown automaton $A_{\posp{}}$.
+Given $\dfa(l) = (Q_l, \T \cup \NT, \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
+
+\begin{itemize}
+\item $s$ and $f$ are two new distinct states.
+\item $Q = \{s, f\} \cup \bigsqcup_l Q_l$.
+\item $\T' = \T \cup \{'\pospl', '\pospr'\}$.
+\item $\Gamma = Q \times Q \times \Labels^{*}$.
+\item $\delta$ is defined by the following equations (unspecified cases produce $\emptyset$):
+  $$
+  \begin{array}{r@{\;=\;}l@{\;\mid\;}l}
+    \delta(p, a, \lambda)
+    & \multicolumn{2}{@{}l@{}}{\{(\delta_l(p, a), \lambda)\} \qquad \text{ where } p \in Q_l \text{ and } a \in \T} \\
+
+    \delta(s, \text{``}\pospl\text{''}, \lambda)
+    & \{ (s_{l}, (s, f, w))
+    & (S, w, N) \in T, N = \labelnt(l) \} \\
+
+    \delta(p, \text{``}\pospl\text{''}, \lambda)
+    & \{ (s_{l_3}, (p, q, w))
+    & p \in Q_{l_1}, \delta_{l_1}(p, N) = q, (N, w, N') \in T, N' = \labelnt(l_3) \} \\
+
+    \delta(p, \text{``}\pospr\text{''}, (s, f, w))
+    & \{ (f, \lambda)
+    & (S, w, N) \in T, N = \labelnt(l), p \in F_l \} \\
+
+    \delta(q', \text{``}\pospr\text{''}, (p, q, w))
+    & \{ (q, \lambda)
+    & q' \in F_{l_1}, \labelnt(l_1) = N', (N, w, N') \in T, p \in Q_{l_2}, \delta_{l_2}(p, N) = q \} \\
+
+  \end{array}
+  $$
+\end{itemize}
+
+\noindent We also construct a modified VPDA $A'_{\posp{}} = (Q, \T', \Gamma \cup \{\gamma\}, \delta', s, \{f\})$ where:
+
+\begin{itemize}
+\item $\gamma$ is a new distinct stack symbol.
+\item $\delta'(p, a, g) = \delta(p, a, g) \cup \delta''(p, a, g)$, where $\delta''$ is given by:
+  $$
+  \begin{array}{r@{\;=\;}l@{}}
+    \delta''(p, \text{``}\pospl\text{''}, \lambda)
+    & \{ (p, \gamma) \} \\
+
+    \delta''(p, \text{``}\pospr\text{''}, \gamma)
+    & \{ (p, \lambda) \} \\
+  \end{array}
+  $$
+\end{itemize}
+
+\noindent To find two distinct successful runs, one in each automaton, that recognize the same word we construct the product automaton $A_{\posp{}} \times A'_{\posp{}} = (Q \times Q, \T', \Gamma \times (\Gamma \cup \{\gamma\}), \delta_{\times}, (s, s), \{(f, f)\})$, where $\delta_{\times}$ is described in \cite{alurVisiblyPushdownLanguages2004} (using the partitions $\T_c = \{ \pospl \}$, $\T_i = \T$, and $\T_r = \{\pospr\}$).
+If the product automaton has a successful run that passes through at least one configuration $((p, p'), (g, g')\cdot w)$ such that $p \neq p' \lor g \neq g'$ (distinct states, or distinct stack symbols, respectively), then that run corresponds to two distinct successful runs in $A_{\posp{}}$ and $A'_{\posp{}}$.
+We check for the existence of such a run by trimming the product automaton (as described in \cite{caralpTrimmingVisiblyPushdown2015}) and looking for a transition that pushes $(g, g')$ where $g \neq g'$, or transitions to $(q, q')$ where $q \neq q'$.
+
+Performing this procedure on a language definition $D$ without non-grouping parentheses and marks gives a sound and complete decision procedure for determining if $D$ is resolvably ambiguous.
+If we instead take a language definition $D$ with marks but no non-grouping parentheses, then remove the marks and perform the procedure, then the absence of two such runs implies that $D$ is resolvably ambiguous, while their presence only implies the possibility of an unresolvable ambiguity; the marks might remove it.
+
+\begin{lemma}
+  There is a bijection between successful runs in $A_{\posp{}}$ and trees $t \in L(T_D)$.
+\end{lemma}
+
+\begin{proof}
+  By defining two functions $\treetorun$ and $\runtotree$, the former from a tree to a run, the latter from run to tree, then showing these two functions to be inverses of each other.
+\end{proof}
+
+We denote the word recognized by a run $R$ by $\runtoword(R)$.
+
+\begin{lemma}
+  $\forall t \in L(T_D).\ \runtoword(\treetorun(t))$ is the canonical linear encoding of $\words(t)$.
+\end{lemma}
+
+\begin{lemma}
+  For all $t \in L(T_D)$ and every linear encoding $c$ such that $\words(t) \subseteq L(c)$, $c \in L(A'_{\posp{}})$.
+\end{lemma}
+
+\begin{lemma}
+  We can define a function $\runprimetorun$ from runs in $A'_{\posp{}}$ to runs in $A_{\posp{}}$ such that $\forall R.\ L(\runtoword(\runprimetorun(R))) \subseteq L(\runtoword(R))$.
+\end{lemma}
+
+\begin{proof}
+  By removing every transition that pushes or pops $\gamma$.
+\end{proof}
+
+\begin{theorem}
+  The existence of two distinct runs $R$ and $R'$ such that $R$ is a successful run in $A_{\posp{}}$ and $R'$ in $A'_{\posp{}}$, and $\runtoword(R) = \runtoword(R')$, implies the existence of two distinct trees such that the word language of one is entirely contained in the other.
+\end{theorem}
+
+\begin{proof}
+  Assume $\runprimetorun(R') = R$.
+$\runprimetorun$ either leaves its argument unchanged, or changes the recognized word.
+If $R'$ and $R$ recognize the same word, and $\runprimetorun(R') = R$, then $R' = R$, but that is a contradiction.
+
+  Since $\runprimetorun(R') \neq R$ we have two distinct trees $t_1 = \runtotree(\runprimetorun(R'))$ and $t_2 = \runtotree(R)$.
+Since $L(\runtoword(\runprimetorun(R'))) \subseteq L(\runtoword(R'))$ and $L(\runtoword(R')) = L(\runtoword(R))$ we have that $L(t_1) \subseteq L(t_2)$.
+\end{proof}
+
+We thus have a way to detect the presence or absence of a pair of trees $t_1, t_2 \in L(T_D)$ such that $\words(t_1) \subseteq \words(t_2)$, which completely determines the resolvability of $D$ when $D$ has no non-grouping parentheses or marks.
+
 %% TODO: CONTINUE

### 2019-10-24 paper/paper.tex
Unranked Tree Grammars

@@
 which symbols are pushed, popped or just read from the stack.
+%
+% TODO: Something on VPDAs?
+
+\paragraph{Unranked Regular Tree Grammars}
+Trees generalize words by allowing each terminal to have multiple
+ordered successors, instead of just zero or one. In this paper, we
+are only concerned with \emph{unranked} trees, where the arity of
+a terminal is not fixed.
+%
+Each terminal with a non-zero rank has a label, allowing a
+\emph{horizontal}
+representation~\cite{comonTreeAutomataTechniques2007}. For
+example, $a(n(\verb|'1'|) \;\verb|'+'|\; n(\verb|'2'|))$ represents
+the parse tree of the string ``$1+2$''.
+%
+Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals
+obtained by a left-to-right traversal of a tree. Informally, it is
+the flattening of a tree after all internal nodes have been
+removed.
 
@@
 
-The results and properties presented in this paper are more naturally described through unranked trees, thus all further references to trees are to unranked trees, despite ranked being more common in the literature.
+The results and properties presented in this paper are more naturally described through unranked trees, thus all references to trees here are to unranked trees, despite ranked being more common in the literature.
 We further distinguish terminals used solely as leaves from terminals that may be either nodes or leaves.

### 2019-10-24 paper/paper.tex
Structured the intro of Section 8

@@
 
-In this section, we showcase two possible use-cases of dynamic
-resolvability analysis: composing separately defined DSLs
-(Section~\ref{sec:evaluation-orc}), and finding ambiguities in
-grammars through testing (Section~\ref{sec:evaluation-ocaml}).
-%
 We have implemented a tool and a small DSL for syntax definition
-that generates a language definition in the style of
+(Section~\ref{sec:evaluation-syncon}) that generates a language
+definition in the style of
 Section~\ref{sec:parse-time-disambiguation}, which the tool then
 uses to construct a parser and do dynamic resolvability analysis.
-The tool is implemented in Haskell and uses an off-the-shelf
+%
+In this section, we showcase two possible use-cases of dynamic
+resolvability analysis using this tool: composing separately
+defined DSLs (Section~\ref{sec:evaluation-orc}), and finding
+ambiguities in grammars through testing
+(Section~\ref{sec:evaluation-ocaml}).
+
+\subsection{A DSL for Syntax Definition}
+\label{sec:evaluation-syncon}
+
+We write our syntax definitions in a DSL building on
+\emph{syncons}, introduced by Palmkvist and
+Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019},
+wherein each production is defined separately from each other (one
+\syncon{syncon} each).
+%
+The tool which performs the parsing is implemented in Haskell and
+uses an off-the-shelf
 implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
@@
 
-The syntax definition DSL builds on \emph{syncons}, introduced by
-Palmkvist and Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019}, wherein each
-production is defined separately from each other (one
-\syncon{syncon} each).
-%
 The main difference between the DSL and the formalism in

### 2019-10-24 paper/paper.tex
Save the hboxes!

@@
 \begin{lstlisting}[basicstyle=\small\tt]
- File "./nestmatch.ml", line 4, characters 4-5:
- Error: This pattern matches values of type int
-        but a pattern was expected which
+ File "./nest.ml", line 4, characters 4-5:
+ Error: This pattern matches values of type
+        int but a pattern was expected which
         matches values of type string
@@
 
-\noindent This is a resolvable ambiguity, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and $\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
+\noindent This is a resolvable ambiguity, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and \\$\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
 To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
@@
 \item $\delta$ is defined by the following equations (unspecified cases produce $\emptyset$):
+  % TODO: This could be made prettier and more aligned
   $$
-  \begin{array}{r@{\;=\;}l@{\;\mid\;}l}
-    \delta(p, a, \lambda)
-    & \multicolumn{2}{@{}l@{}}{\{(\delta_l(p, a), \lambda)\} \qquad \text{ where } p \in Q_l \text{ and } a \in \T} \\
-
-    \delta(s, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l}, (s, f, w))
-    & (S, w, N) \in T, N = \labelnt(l) \} \\
-
-    \delta(p, \text{``}\pospl\text{''}, \lambda)
-    & \{ (s_{l_3}, (p, q, w))
-    & p \in Q_{l_1}, \delta_{l_1}(p, N) = q, (N, w, N') \in T, N' = \labelnt(l_3) \} \\
-
-    \delta(p, \text{``}\pospr\text{''}, (s, f, w))
-    & \{ (f, \lambda)
-    & (S, w, N) \in T, N = \labelnt(l), p \in F_l \} \\
-
-    \delta(q', \text{``}\pospr\text{''}, (p, q, w))
-    & \{ (q, \lambda)
-    & q' \in F_{l_1}, \labelnt(l_1) = N', (N, w, N') \in T, p \in Q_{l_2}, \delta_{l_2}(p, N) = q \} \\
+  \begin{array}{l}
+    \delta(p, a, \lambda) = \{(\delta_l(p, a), \lambda)\} \qquad \text{ where } p \in Q_l \text{ and } a \in \T \\
+
+    \delta(s, \text{``}\pospl\text{''}, \lambda) =
+    \{ (s_{l}, (s, f, w)) \mid
+    (S, w, N) \in T, N = \labelnt(l) \} \\
+
+    \delta(p, \text{``}\pospl\text{''}, \lambda) = \\
+    \qquad\{ (s_{l_3}, (p, q, w)) \mid
+    p \in Q_{l_1}, \delta_{l_1}(p, N) = q, (N, w, N') \in T, N' = \labelnt(l_3) \} \\
+
+    \delta(p, \text{``}\pospr\text{''}, (s, f, w)) =
+    \{ (f, \lambda) \mid
+    (S, w, N) \in T, N = \labelnt(l), p \in F_l \} \\
+
+    \delta(q', \text{``}\pospr\text{''}, (p, q, w)) = \\
+    \qquad \{ (q, \lambda) \mid
+    q' \in F_{l_1}, \labelnt(l_1) = N', (N, w, N') \in T, p \in Q_{l_2}, \delta_{l_2}(p, N) = q \} \\
 
@@
 precedence {
-  mul; // higher in precedence list
-  add; // means higher precedence
+  mul; // higher in list means
+  add; // higher precedence
 }
@@
 \item We could change the definition of the parallel combinators
-  and define separate lexical tokens for the combinators, e.g.,
-  \syncon{token Seq = ">[[:lower:]][[:word:]]*>"}. This would add
-  just enough whitespace sensitivity to allow separating the
-  different cases.
+  and define separate lexical tokens for the combinators, e.g,
+  \syncon{token Seq = ">[[:lower:]]}\\\syncon{[[:word:]]*>"}. This
+  would add just enough whitespace sensitivity to allow separating
+  the different cases.
 \item We could change the syntax of the combinators to avoid
@@
 %
-``\ocaml{(1 + (let x = 1 in x)) + 2}''.
+``\ocaml{(1 + (let x}\\\ocaml{= 1 in x)) + 2}''.
 %
@@
   $ F_1 \times F_2)$ where, when $c \in \T_c$, $i \in \T_i$ and
-  $r \in \T_r$:
+  $r \in \T_r$: \RED{TODO: Is this definition correct? $g_1$ and
+    $g_2$ are not used in the third case}
   \[
-  \begin{array}{r@{\ =\ }l}
-    \delta'((p_1, p_2), c, \lambda) & \{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \}\\
-    \delta'((p_1, p_2), i, \lambda) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
-    \delta'((p_1, p_2), r, (g_1, g_2)) & \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
+  \begin{array}{l}
+    \delta'((p_1, p_2), c, \lambda) =\\ \qquad\{((q_1, q_2), (g_1, g_2)) \mid (q_1, g_1) \in \delta_1(p_1, c, \lambda), (q_2, g_2) \in \delta_2(p_2, c, \lambda) \}\\
+    \delta'((p_1, p_2), i, \lambda) =\\ \qquad \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
+    \delta'((p_1, p_2), r, (g_1, g_2)) =\\ \qquad \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
   \end{array}

### 2019-10-25 paper/paper.tex
Chase the whitespace! Don't let your dreams be dreams!

@@
 \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
+    scale = .9,transform shape]
   \node[state,initial] (1) {};
@@
 \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
+    scale = .9,transform shape]
   \node[state,initial] (1) {$1$};
@@
 \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=2cm,
-    scale = 1,transform shape]
+    scale = .9,transform shape]
   \node[state] (1) {$1$};
@@
 %
-The tool which performs the parsing is implemented in Haskell and
-uses an off-the-shelf
+The tool is implemented in Haskell and uses an off-the-shelf
 implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
@@
 Section~\ref{sec:parse-time-disambiguation} is that marks are
-introduced as separate \syncon{forbid} declarations rather than
-being inlined in productions. This allows supporting high-level
+introduced as separate \syncon{forbid} declarations instead of
+being inlined in productions. This allows supporting
 convenience constructs, e.g., specifying precedence between
@@
 statically resolved without changing the original definitions.
-Other convenience constructs include defining prefix, infix, and
+Other convenience constructs include prefix, infix, and
 postfix operators with a given associativity.
-
+%
 The running example used in
@@
 \syncon{syncon} for integer literals using this token. A list is
-defined as a sequence of zero or more expressions separated by
-semi-colons wrapped in square braces. Finally, addition and
+defined as a bracketed sequence of zero or more expressions separated by
+semi-colons. Finally, addition and
 multiplication are defined as infix syncons with the expected
-precedence rules. Note that we could just as well have replaced
+precedence rules. Note that we could also have replaced
 the precedence list with explicit \syncon{forbid} declarations,
@@
 support for resolvable ambiguity, however, we can allow this
-ambiguity and let programmers disambiguate expressions using
-parentheses. After composing the above definition with a simple
-language supporting addition, variables, and function calls
-(language definition omitted for brevity), parsing the expression
+ambiguity and let programmers use parentheses for disambiguation.
+After composing the above definition with a simple language
+supporting addition, variables, and function calls (full
+definition omitted for brevity), parsing the expression
 %
@@
 \item We could change the syntax of the combinators to avoid
-  clashes with other operators.
+  clashes.
 \end{itemize}
@@
 \noindent
-The change in the first alternative is somewhat ad-hoc, but can be done after composition
+The first alternative is somewhat ad-hoc, but can be done after composition
 and would allow us to reuse both language definitions without
@@
 In this section, we investigate how to apply dynamic resolvability
-analysis to identify ambiguities in the specification of a
+analysis to find ambiguities in the specification of a
 real-world language.
@@
 extensions) of the OCaml reference manual~\cite{leroyOCamlSystemRelease2018}.
-The syncon definition is just under 700 lines of code, and is
-currently complete enough to parse roughly 75\% (1012 out of 1334
+The syncon definition is just under 700 lines long, and can
+currently parse roughly 75\% (1012 out of 1334
 files) of the \verb|.ml| files present in the OCaml compiler
@@
 They allow non-transitive, non-total precedence, and ``feel that it is overly restrictive to require the grammar to be unambiguous.'' Similar to our approach, they do not reject ambiguous grammars, only ambiguous parses.
-They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must be related by precedence (children must have higher precedence than parents).
+They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must have higher precedence than parents.
 This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'-'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.
@@
 
-Parsing expression grammars \cite{fordParsingExpressionGrammars2004} sidestep the issue of ambiguity by not introducing it in the first place.
+Parsing expression grammars \cite{fordParsingExpressionGrammars2004} sidestep the issue of ambiguity by not introducing it at all.
 However, this also loses the potential gains of leaving certain ambiguities.

### 2019-10-25 paper/paper.tex
Removed some RED
Some updates

@@
 
-\title{Contribution Title\thanks{Supported by organization x.}}
+\title{Resolvable Ambiguity\thanks{Supported by organization x.}}
 %
@@
 %
-\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-%
-\authorrunning{F. Author et al.}
+\author{Viktor Palmkvist\inst{1}\orcidID{0000-0003-0669-4085} \and
+Elias Castegren\inst{1}\orcidID{0000-0003-4918-6582} \and
+Philipp Haller\inst{1}\orcidID{0000-0002-2659-5271} \and
+David Broman\inst{1}\orcidID{0000-0001-8457-4105}}
+%
+\authorrunning{V. Palmkvist et al.}
 % First names are abbreviated in the running head.
@@
 %
-\institute{Princeton University, Princeton NJ 08544, USA \and
-Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-\email{lncs@springer.com}\\
-\url{http://www.springer.com/gp/computer-science/lncs} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{\{abc,lncs\}@uni-heidelberg.de}}
+\institute{KTH Royal Institute of Technology, 100 44 Stockholm, Sweden
+\email{\{vipa,eliasca,dbro,phaller\}@kth.se}%\and
+%% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
+%% \email{lncs@springer.com}\\
+%% \url{http://www.springer.com/gp/computer-science/lncs} \and
+%% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+%% \email{\{abc,lncs\}@uni-heidelberg.de}
+}
 %
@@
 composability of domain-specific languages.
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+\keywords{Syntax \and Ambiguity \and Grammars \and Parsing \and Domain-specific languages}
 \end{abstract}
@@
 % TODO: more language composition references
-The possibility of composing different languages at will is a prevalent idea in research \cite{vanwykSilverExtensibleAttribute2010,lorenzenSoundTypedependentSyntactic2016,danielssonParsingMixfixOperators2011}.
+The possibility of composing different languages is a prevalent idea in research \cite{vanwykSilverExtensibleAttribute2010,lorenzenSoundTypedependentSyntactic2016,danielssonParsingMixfixOperators2011}.
 Particularly relevant to this paper is the behavior of composed \emph{syntax}; composing unambiguous syntaxes does not necessarily produce an unambiguous composed syntax.
@@
 The former case gives good guarantees for the end-user experience (no ambiguities) while the latter gives more freedom to the language designer.
-Resolvable ambiguity, as presented in this paper, provides a middle ground with more language designer freedom than the former and a better user experience than the latter.
+Resolvable ambiguity, as presented in this paper, provides a middle ground with more language designer freedom than the former and a different guarantee for the end-user: no \emph{unresolvable} ambiguities.
 
@@
 Precedence in Haskell is defined as a number between 0 and 9, meaning that operators from different libraries have a defined relative precedence.
-If the libraries are designed by different authors this precedence is essentially by accident and it may or may not be sensible.
+If the libraries are designed by different authors then this precedence is unlikely to have been considered and may or may not be sensible.
 Leaving the relative precedence undefined would not affect expressivity, every possible combination of operators can be written by simply adding parentheses, and would avoid surprising interpretations.
@@
 %
-Each terminal with a non-zero rank has a label, allowing a
-\emph{horizontal}
-representation~\cite{comonTreeAutomataTechniques2007}. For
+The allowed sequence of ordered successors is described by a regular language~\cite{comonTreeAutomataTechniques2007}. For
 example, $a(n(\verb|'1'|) \;\verb|'+'|\; n(\verb|'2'|))$ represents
@@
 Finally, $yield : L(T) -> \T^{*}$ is the sequence of terminals
-obtained by a left-to-right traversal of a tree. Informally, it is
+obtained by a left-to-right traversal of the leaves of a tree. Informally, it is
 the flattening of a tree after all internal nodes have been
@@
 
-\noindent A \emph{resolvably ambiguous} word is a word where all its parse trees can be written in an unambiguous way, formally:
+\noindent For \emph{resolvable ambiguity} the definition instead centers around parse trees:
+
+\begin{definition}\label{def:resolvable-tree}
+  Given an abstract parser $P = (\T, T, \parse)$, a tree $t \in T$ is \emph{resolvably ambiguous}, written $\rho_P(t)$, iff
+
+  $\exists w \in \T^{*}.\ \parse(w') = \{t\}$.
+\end{definition}
+
+\noindent A word is then resolvably ambiguous if all its parse trees are resolvably ambiguous:
 
@@
 
-  $\forall t \in \parse(w).\ \exists w' \in \T^{*}.\ \parse(w') = \{t\}$.
+  $\forall t \in \parse(w).\ \rho_P(t)$.
 \end{definition}
 
-The words $w$ and $w'$ will typically be related, e.g., $w'$ has a few added pairs of parentheses but is otherwise identical.
-The example $1 + 2 \cdot 3$
-%in Fig.~\ref{fig:arith-example}
-is not resolvably ambiguous, since neither of the two resulting trees can be written in any other way.
-However, if we add grouping parentheses to the language, we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously parsed as the first and second tree, respectively.
-
-Additionally, we define an \emph{abstract parser} to be resolvably ambiguous if all words in its corresponding parse language $L(P)$ are resolvably ambiguous, formally:
+\noindent Additionally, we define an \emph{abstract parser} to be resolvably ambiguous if all its parse trees are resolvably ambiguous, formally:
 
@@
 
-  $\forall w \in L(P).\ \rho_P(w)$.
+  $\forall t \in T.\ \rho_P(t)$.
 \end{definition}
@@
 The primary purpose of this formalism is, as described in the previous section, to produce a $\parse$ function, i.e., to describe a word language and assign one or more parse trees to each word.
-Parse trees are unranked trees; Section~\ref{sec:resolvable-definition} suggests that $\parse$ discards some information, to enable the resolution of some ambiguities; here, we choose to discard grouping parentheses.
+Furthermore, Section~\ref{sec:resolvable-definition} suggests that disambiguation is made possible by letting $\parse$ discard information.
+We use unranked trees as parse trees and have $\parse$ discard grouping parentheses.
 
@@
 Also note that the right-hand side of a production is a regular expression, rather than the theoretically simpler sequence used in a context-free grammar.
-For proof-technical reasons (see Section~\ref{sec:lattice-vpl}) we require each regular expression to be unambiguous, in the classical sense, when marks are removed.
-This can be checked using, e.g., the approach described by Brabrand and Thomsen~\cite{brabrandTypedUnambiguousPattern2010}.
-In practice, this is likely desirable regardless, since a user is likely to attach different semantic meanings to each occurrence of a non-terminal.
 
@@
     $E$ & $m$ & $E_{\{a\}}$ \verb|'*'| $E_{\{a\}}$ \\
-    $E$ & $n$ & $N$ \\
+    $E$ & $n$ & $I$ \\
   \end{tabular}
   \captionof{figure}{The input language definition used as a running example, an expression language with lists, addition, and multiplication, with precedence defined, but not associativity.
-Assumes that $N$ matches a numeric terminal.}
+Assumes that $I$ matches a numeric terminal.}
   \label{fig:running-example-definition}
@@
       $E$ & $m$ & $E$ \texttt{'*'} $E$ $)$ \\
-      $E$ & $n$ & $N$ $)$ \\
+      $E$ & $n$ & $I$ $)$ \\
       \bottomrule
@@
       $E$ & $m$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ $)$ \\
-      $E$ & $n$ & $N$ $)$ \\
+      $E$ & $n$ & $I$ $)$ \\
       $E$ & $g$ & \texttt{'('} $E$ \texttt{')'} $)$ \\
@@
       $E_{\{a\}}$ & $m$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ $)$ \\
-      $E_{\{a\}}$ & $n$ & $N$ $)$ \\
+      $E_{\{a\}}$ & $n$ & $I$ $)$ \\
       $E_{\{a\}}$ & $g$ & \texttt{'('} $E$ \texttt{')'} $)$ \\
@@
       $E$ & $E$ \texttt{'*'} $E$ \\
-      $E$ & $N$ \\
+      $E$ & $I$ \\
       \midrule
@@
       $E$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ \\
-      $E$ & $N$ \\
+      $E$ & $I$ \\
       $E$ & \texttt{'('} $E$ \texttt{')'} \\
@@
       $E_{\{a\}}$ & $E_{\{a\}}$ \texttt{'*'} $E_{\{a\}}$ \\
-      $E_{\{a\}}$ & $N$ \\
+      $E_{\{a\}}$ & $I$ \\
       $E_{\{a\}}$ & \texttt{'('} $E$ \texttt{')'} \\
@@
 
-Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}, \RED{along with visualizations of the syntax trees}.
+Examples of elements in each of these four languages can be seen in Fig.~\ref{fig:example-words-in-square}.
 Each element corresponds to the word ``$(1 + 2) * 3$'' in $L(G'_D)$.
@@
 
-We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}.
-Section~\ref{sec:static} gives a partial solution to the static problem, while Section~\ref{sec:dynamic} fully solves the dynamic problem, with one caveat: we only consider languages with balanced parentheses.
-
-\section{The Word Language of a Parse Tree} \label{sec:linear-encoding}
+\subsection{The Word Language of a Parse Tree} \label{sec:linear-encoding}
 
@@
 
+We are now ready to construct decision procedures for the static and dynamic resolvability problems, as given in Definitions~\ref{def:static-procedure-sound}, \ref{def:static-procedure-complete}, \ref{def:dynamic-procedure-sound}, and \ref{def:dynamic-procedure-complete}.
+Section~\ref{sec:static} solves the static problem for a language subclass, while Section~\ref{sec:dynamic} fully solves the dynamic problem.
+
 \section{Dynamic Resolvability Analysis} \label{sec:dynamic}
@@
 
-Our approach centers around the construction of a visibly pushdown automaton recognizing $\words(t)$ for any particular $t \in L(T_D)$.
+Our approach centers around the construction of a VPDA recognizing $\words(t)$ for any particular $t \in L(T_D)$.
 We can easily construct this automaton via the canonical linear encoding: let every pair ``$\reqp{}$'' push and pop the same stack symbol (call it $\gamma$) and let every ``$\posp{}$'' push and pop a unique stack symbol.
@@
 
-\noindent This is useful to us since visibly pushdown automata are closed under difference.
+\noindent This is useful to us since VPDAs are closed under difference.
 Given a pair of trees $t_1$ and $t_2$ we can thus produce a pair of automata recognizing $\words(t_1)$ and $\words(t_2)$ respectively, then construct a pair of automata recognizing $\words(t_1) \setminus \words(t_2)$ and $\words(t_2) \setminus \words(t_1)$, respectively.
@@
     \State $T' <- T \setminus \{t\}$
-    \State $A <- \words(t)$ \Comment{$A$ is a visibly pushdown automaton.}
+    \State $A <- \words(t)$ \Comment{$A$ is a VPDA.}
     \Loop
@@
 
+Note that the shortest word we produce might not be unique, in the general case there may be more than one shortest word that disambiguates a tree.
+
+We require two things to ensure termination: that $T'$ is finite, and that the inner loop terminates.
+$T'$ is finite if no words are infinitely ambiguous, which is ensured by requirement 2.
+It is also ensured if $D$ is resolvably ambiguous, since an infinite ambiguity requires that $G_D$ contains a cycle $N =>^{+} N$, which would imply that any parse tree containing a production from $N$ only has (infinitely) ambiguous words, i.e., contradiction.
+Requirement 2 also ensures that the number of trees that share an ambiguous word with $t$ is finite (since each tree corresponds to a left-most derivation in $G_D$ and must share exactly the same non-parenthesis terminals).
+
 \section{Static Resolvability Analysis} \label{sec:static}
@@
 
-This section describes the construction of a visibly pushdown automaton that we examine to determine if a language has a pair of trees where the language of one entirely contains the other.
+This section describes the construction of a VPDA that we examine to determine if a language has a pair of trees where the language of one entirely contains the other.
 We begin by outlining the approach and then describe it in more detail.
@@
 Two distinct runs that recognize the same word thus corresponds to two distinct trees that have the \emph{same} word language.
-This is essentially the question of whether a visibly pushdown automaton is ambiguous, which we can detect by constructing a product automaton and trimming it.
+This is essentially the question of whether a VPDA is ambiguous, which we can detect by constructing a product automaton and trimming it.
 
@@
 
-We begin by constructing a visibly pushdown automaton that recognizes \emph{a} linear encoding for each tree, then modify it to only recognize the \emph{canonical} linear encoding.
+We begin by constructing a VPDA that recognizes \emph{a} linear encoding for each tree, then modify it to only recognize the \emph{canonical} linear encoding.
 To illustrate the approach we consider the following language:
@@
 
-\noindent We then combine these separate DFAs to produce a single visibly pushdown automaton in three steps:
+\noindent We then combine these separate DFAs to produce a single VPDA in three steps:
 
@@
 
-\noindent To make the figure less cluttered we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')} q$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')} q$.
+\noindent To reduce clutter we omit $'\pospl'$ and $'\pospr'$, i.e., we abbreviate $p \xrightarrow{'\pospl', +(p',q')} q$ as $p \xrightarrow{+(p', q')} q$ and $p \xrightarrow{'\pospr', -(p',q')} q$ as $p \xrightarrow{-(p', q')} q$.
 
@@
 % TODO: footnote should probably be removed
-\noindent $T$ is finite if $D$ has no cycles $N =>^{+} N$\footnote{This is notation from CFGs, which is semi-obvious what it means, but technically not defined here.}.
+\noindent $T$ is finite if $D$ has no cycles $N =>^{+} N$.
 Such a cycle is easy to detect (depth-first search), and means that any tree that contains a production from the non-terminal $N$ only ever produces (infinitely) ambiguous words, i.e., all words for such a tree are unresolvably ambiguous.
@@
 
-\noindent We now define the visibly pushdown automaton $A_{\posp{}}$.
-Given $\dfa(l) = (Q_l, \T \cup \NT, \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
+\noindent We now define the VPDA $A_{\posp{}}$. Given $\dfa(l) = (Q_l, \T \cup \NT, \delta_l, s_l, F_l)$ for all $l \in Labels$, $A_{\posp{}} = (Q, \T', \Gamma, \delta, s, \{f\})$ where:
 
@@
   $ F_1 \times F_2)$ where, when $c \in \T_c$, $i \in \T_i$ and
-  $r \in \T_r$: \RED{TODO: Is this definition correct? $g_1$ and
-    $g_2$ are not used in the third case}
+  $r \in \T_r$:
   \[
@@
     \delta'((p_1, p_2), i, \lambda) =\\ \qquad \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, i, \lambda), (q_2, \lambda) \in \delta_2(p_2, i, \lambda) \}\\
-    \delta'((p_1, p_2), r, (g_1, g_2)) =\\ \qquad \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, \lambda), (q_2, \lambda) \in \delta_2(p_2, r, \lambda) \}\\
+    \delta'((p_1, p_2), r, (g_1, g_2)) =\\ \qquad \{((q_1, q_2), \lambda) \mid (q_1, \lambda) \in \delta_1(p_1, r, g_1), (q_2, \lambda) \in \delta_2(p_2, r, g_2) \}\\
   \end{array}

### 2019-10-25 paper/paper.tex
Pre read through

@@
 
-\title{Resolvable Ambiguity\thanks{Supported by organization x.}}
+\title{Resolvable Ambiguity\thanks{This project is financially supported by the Swedish Foundation for Strategic Research (FFL15-0032)}}
 %
@@
 %
-\author{Viktor Palmkvist\inst{1}\orcidID{0000-0003-0669-4085} \and
-Elias Castegren\inst{1}\orcidID{0000-0003-4918-6582} \and
-Philipp Haller\inst{1}\orcidID{0000-0002-2659-5271} \and
-David Broman\inst{1}\orcidID{0000-0001-8457-4105}}
+\author{Viktor Palmkvist\inst{1} \and
+Elias Castegren\inst{1} \and
+Philipp Haller\inst{1} \and
+David Broman\inst{1}}
+%% \author{Viktor Palmkvist\inst{1}\orcidID{0000-0003-0669-4085} \and
+%% Elias Castegren\inst{1}\orcidID{0000-0003-4918-6582} \and
+%% Philipp Haller\inst{1}\orcidID{0000-0002-2659-5271} \and
+%% David Broman\inst{1}\orcidID{0000-0001-8457-4105}}
 %
@@
 \noindent This is a resolvable ambiguity, since $\parse($\verb|'1 + (2 + 3)'|$) = \{t_1\}$ and \\$\parse($\verb|'(1 + 2) + 3'|$) = \{t_2\}$.
-To demonstrate the unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$, at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
+To demonstrate an unresolvable case, we add the production $E -> s: E$ \verb|';'| $E$ (common syntax for sequential composition), at which point we find that the word \verb|'[1 ; 2]'| is unresolvably ambiguous; $\parse($\verb|'[1 ; 2]'|$) = \{t_3, t_4\}$ where:
 
@@
 Central to our approach is the shape of $\words(t)$.
-Consider the tree corresponding to \verb|'(1 + 2) * 3'| (visualized in Figure~\ref{fig:sum-add-tree}).
+Consider the tree corresponding to \verb|'(1 + 2) * 3'|.
 Each node in the tree may be surrounded by zero or more parentheses, except \verb|'1 + 2'|, which requires at least one pair.
@@
 We now divide the possible languages along two axes: whether a language has (non-grouping) parentheses, and whether a language has marks.
-The parentheses present in the canonical encoding will have the following shapes:
+The parentheses present in the canonical encoding will have the following shapes ($\mathbb{N} = \{0,1,\ldots\}$):
 
@@
 Section~\ref{sec:static-proofs} additionally shows that adding marks to a resolvably ambiguous language in the lower-right quadrant preserves resolvability, which gives us a conservative approach for languages in the top-right quadrant.
-Intuitively, if a language has marks, but they are just conveniences, then we can still detect that the language is resolvable.
+Intuitively, if a language has marks, but they are not required for resolvability, then we can still detect that the language is resolvable.
 For example, most expression languages fall in this category (we can write any expression unambiguously by adding parentheses everywhere, even without any precedence or associativity), while lists in OCaml do not (no amount of parentheses will make \verb|'[1;2]'| look like a list of two elements without a mark).
@@
 
-\begin{proof}
+\begin{proof}[sketch]
   By defining two functions $\treetorun$ and $\runtotree$, the former from a tree to a run, the latter from run to tree, then showing these two functions to be inverses of each other.
@@
 
-\begin{proof}
+\begin{proof}[sketch]
   By removing every transition that pushes or pops $\gamma$.
@@
 
-\begin{proof}
+\begin{proof}[sketch]
   Assume $\runprimetorun(R') = R$.
@@
 %
-The tool is implemented in Haskell and uses an off-the-shelf
-implementation\footnote{\url{http://hackage.haskell.org/package/Earley}}
-of the Earley parsing algorithm
+The tool is implemented in Haskell and uses a custom implementation of the Earley parsing algorithm
 \cite{earleyEfficientContextfreeParsing1970}.
@@
 
+Our related work falls in three categories: syntax definition formalisms (including subclasses of CFGs), language frameworks, and other approaches to ambiguity.
+
 Afroozeh et al.~\cite{afroozehSafeSpecificationOperator2013}'s operator ambiguity removal patterns bear a striking resemblance to the marks presented in this paper.
 However, in special-casing (what in this paper would be) marks on left and right-recursions in productions they correctly cover the edge case involving deep precedence discussed in Section~\ref{sec:evaluation-ocaml}.
-This approach thus suggests an interesting direction for future work: to integrate it into a system designed around resolvable ambiguity, and extending the algorithms presented in this paper to cover it.
+This approach thus suggests an interesting direction for future work: extend the algorithms presented in this paper to cover it.
 
@@
 They also introduce a concept of \emph{precedence correct} expressions; expressions where direct children must have higher precedence than parents.
-This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'-'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.
-
-Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{vanwykContextawareScanningParsing2007} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, with which extension language developers can check that their extension will compose well with other extensions, without any knowledge of any particular other extension.
-For this paper, only the syntactic component \cite{schwerdfegerVerifiableCompositionDeterministic2009} of this analysis is relevant.
-This analysis guarantees that the composition of a base language and any number of extensions that have passed the analysis will compose to a grammar in LALR(1).
-This is a far more restricted class of languages than unambiguous languages, not to mention resolvably ambiguous languages, and thus places far more restrictions on language designers than our approach.
-
-The detection of classical ambiguity in context-free grammars is undecidable in general \cite{cantorAmbiguityProblemBackus1962}, yet numerous heuristic approaches exist.
-Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using SAT-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of Basten~\cite{bastenAmbiguityDetectionProgramming2011}.
+This is more restrictive than our approach, e.g., in a language where \verb|'+'| and \verb|'*'| have no defined relative precedence they reject \verb|'1 + 2 * 3'| as syntactically invalid, while we parse it as an ambiguous expression.
 
@@
 
-Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989,lorenzenSoundTypedependentSyntactic2016}).
+Most commonly used parser generators are based on unambiguous CFG subclasses, e.g., LL(k), LR(k), or LR(*).
+Others do not fit neatly in the Chomsky hierarchy, but still produce a single parse tree per parse, e.g., LL(*) \cite{parrLLFoundationANTLR2011} and ALL(*) \cite{parrAdaptiveLLParsing2014}.
+Yet others produce multiple parse trees or other forms of parse forests, e.g., GLR \cite{langDeterministicTechniquesEfficient1974}, GLL \cite{scottGLLParsing2010}, and Earley \cite{earleyEfficientContextfreeParsing1970}.
+
+Silver \cite{vanwykSilverExtensibleAttribute2010}, a system for defining extensible languages using attribute grammars, and its associated parser Copper \cite{vanwykContextawareScanningParsing2007} have a ``Modular Well-Definedness Analysis'' \cite{kaminskiModularWellDefinednessAnalysis2013}, the syntactic component of which can be found in \cite{schwerdfegerVerifiableCompositionDeterministic2009}.
+This analysis guarantees that the composition of a base language and any number of extensions that have passed the analysis will compose to a grammar in LALR(1).
+This language class is more restrictive than both unambiguous and resolvably ambiguous languages, though somewhat comparable to the subclass our static analysis supports.
+
+The detection of classical ambiguity in context-free grammars is undecidable in general \cite{cantorAmbiguityProblemBackus1962}, yet numerous heuristic approaches exist.
+Examples include linguistic characterizations and regular language approximations \cite{brabrandAnalyzingAmbiguityContextFree2007}, using SAT-solvers \cite{axelssonAnalyzingContextFreeGrammars2008}, and other conservative approaches \cite{schmitzConservativeAmbiguityDetection2007}, For an overview, and additional approaches, see the PhD thesis of Basten~\cite{bastenAmbiguityDetectionProgramming2011}.
+
+Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989,lorenzenSoundTypedependentSyntactic2016,erdwegLayoutSensitiveGeneralizedParsing2013}).
 These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but also assume that the composed language is unambiguous, which in practice precludes the composition of languages constructed independently of each other.

### 2019-10-25 paper/paper.tex
Submitted version (ESOP 2020)

@@
 resolvability problems. We provide solutions to the static problem
-for two restricted classes of languages and sketch proofs of soundness and,
-depending on the restrictions, completeness. We also provide a sound
+for a restricted language class and sketch proofs of soundness and completeness. We also provide a sound
 and complete solution to the dynamic problem for a much less
@@
 %
-However, recent work by
-Palmkvist and Broman~\cite{palmkvistCreatingDomainSpecificLanguages2019} shows why
-carefully designed ambiguous syntax definitions are in some cases
+However, our recent work~\cite{palmkvistCreatingDomainSpecificLanguages2019} suggests that
+carefully designed ambiguous syntax definitions can be
 preferable compared to completely unambiguous
@@
 For a particular language, the \emph{dynamic resolvable ambiguity problem} asks the following question: can every parse tree of \emph{one particular} program be written unambiguously, while the \emph{static resolvability problem} asks if every parse tree of \emph{every} program can be written unambiguously.
-The latter problem in particular is difficult, since it in general involves examining an infinite set (one per parse tree) of infinite sets (words that can parse as that parse tree) to find one without a unique element (i.e., an unambiguous word).
+In particular, the latter problem is difficult, since it in general involves examining an infinite set (one per parse tree) of infinite sets (words that can parse as that parse tree) to find one without a unique element (i.e., an unambiguous word).
 Furthermore we must be able to produce an unambiguous program for a given parse tree to be able to give good error messages to an end-user encountering an ambiguity.
@@
 \paragraph{Automata}
-An automata is a 5-tuple $(Q,\T,\delta,q_0,F)$, where $Q$ is a
+An automaton is a 5-tuple $(Q,\T,\delta,q_0,F)$, where $Q$ is a
 finite set of states; $\T$ a finite set of terminals; $\delta$ a
@@
 
-In practice, we may additionally be interested in a decision procedure
-$g$ such that $g(P) \implies \lnot \rho(P) \text{ for all abstract
-  parsers } P$. It turns out, we can give $g$ for the parse languages
-studied in the following sections.
-
 \begin{definition}[Completeness of Static Resolvability]\label{def:static-procedure-complete}
@@
 \end{definition}
-
-\noindent Analogous to the static case, for the parse languages studied in the
-following sections, we can also provide a decision procedure $g$ such
-that $g(\parse, w) \implies \lnot \rho_P(w) \text{ for all } w \in
-\Sigma^{*} \text{ and abstract parsers } P = (\T, T, \parse)$.
 
@@
 
-\begin{figure}
+\begin{figure}[b]
 
@@
 
-\begin{figure}
+\begin{figure}[t]
   \subfloat[$T_D$, the parse trees of $G_D$.\label{fig:running-example-generated:t}]{%
@@
 
-Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989,lorenzenSoundTypedependentSyntactic2016,erdwegLayoutSensitiveGeneralizedParsing2013}).
-These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but also assume that the composed language is unambiguous, which in practice precludes the composition of languages constructed independently of each other.
+Numerous language development frameworks and libraries support syntactic language composition \emph{without} any guarantees on the resulting language (e.g., \cite{heeringSyntaxDefinitionFormalism1989,lorenzenSoundTypedependentSyntactic2016,erdwegLayoutSensitiveGeneralizedParsing2013,katsSpoofaxLanguageWorkbench2010}).
+These systems tend to have some form of general parser, so that they can handle arbitrary context-free grammars, but mention no handling of ambiguities encountered by an end-user.
 

### 2019-11-14 syncon-parser/exe/Main.hs
Add a 'dev' command
Fixes #3

@@ -10,7 +10,7 @@ import FileAnnotation (annotate, putInTextTemplate)
 import ErrorMessage (FormatError, formatErrors, formatError, ErrorOpts)
 
 import System.Environment (withArgs)
-import System.FilePath ((</>), (<.>), takeDirectory)
+import System.FilePath ((</>), (<.>), takeDirectory, isExtensionOf)
 import System.Timeout (timeout)
 import System.Directory (createDirectoryIfMissing)
 
@@ -97,12 +97,12 @@ test :: IO ()
 -- test = withArgs ["languages/simple-right-recursion/def.syncon", "languages/simple-right-recursion/test.test", "--two-level"] main
 -- test = withArgs ["languages/atest/atest.syncon", "languages/atest/atest.test", "--two-level"] main
 -- test = withArgs ["languages/ocaml/ocaml.syncon", "languages/ocaml/a_ustring.ml", "--html=out.html", "--two-level"] main
-test = withArgs ["languages/ocaml/ocaml.syncon", "languages/ocaml/ustring.ml", "--html=out.html", "--two-level"] main
+-- test = withArgs ["languages/ocaml/ocaml.syncon", "languages/ocaml/ustring.ml", "--html=out.html", "--two-level"] main
 -- test = withArgs ["case-studies/ocaml.syncon", "case-studies/ocaml/fizzbuzz.ml", "--html=out.html", "--two-level"] main
 -- test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--dot=out"] main
 -- test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--two-level"] main
 -- test = withArgs ["examples/bootstrap.syncon", "--source=examples/bootstrap.syncon", "--json=out.json"] main
--- test = withArgs ["--help"] main
+test = withArgs ["--help"] main
 -- test = GLL.test
 
 getArgsSeq :: IO (Seq [Char])
@@ -115,129 +115,20 @@ die' :: Text -> IO a
 die' t = do
   throwIO $ SourceFileException t
 
-compile :: Opt.Mod Opt.CommandFields (IO ())
-compile = Opt.command "compile" (Opt.info compileCmd $ Opt.progDesc "Compile a list of '.syncon' files to a single '.synconc' file, to be used for parsing.")
-  where
-    compileCmd = do
-      files <- some $ Opt.argument Opt.str $
-        Opt.metavar "FILES..."
[…]

### 2019-11-14 syncon-parser/src/P4Parsing/Parser.hs

@@ -1,6 +1,6 @@
 {-# LANGUAGE RecordWildCards, ViewPatterns, UndecidableInstances #-}
 
-module P4Parsing.Parser (Error(..), precomputeSingleLanguage, dfToLanguageTokens, parseTokens, parseFile, forestToDot, precomputeToSerialisable, serialisableToPrecompute) where
+module P4Parsing.Parser (Error(..), precomputeSingleLanguage, dfToLanguageTokens, parseTokens, parseFile, forestToDot, precomputeToSerialisable, serialisableToPrecompute, Precomputed) where
 
 import Pre hiding (from, some, many, optional)
 import Result (Result(..))

### 2019-11-14 syncon-parser/src/P5DynamicAmbiguity/Isolation.hs

@@ -1,4 +1,4 @@
-module P5DynamicAmbiguity.Isolation (isolate, getElidable, showElidable) where
+module P5DynamicAmbiguity.Isolation (isolate, getElidable, showElidable, Elidable) where
 
 import Pre hiding (reduce, State, state, orElse)
 import Result (Result(..))

### 2019-11-14 syncon-parser/exe/Main.hs
Catch deserialisation error when reading .synconc
Fix #2

@@ -27,7 +27,7 @@ import Data.FileEmbed (embedFile)
 import Data.Generics.Uniplate.Data (universe, universeBi)
 import Data.Functor.Foldable (project, cata)
 import Text.Show.Pretty (pPrint)
-import Codec.Serialise (writeFileSerialise, readFileDeserialise)
+import Codec.Serialise (writeFileSerialise, readFileDeserialise, DeserialiseFailure(..))
 
 import qualified Options.Applicative as Opt
 
@@ -102,7 +102,8 @@ test :: IO ()
 -- test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--dot=out"] main
 -- test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--two-level"] main
 -- test = withArgs ["examples/bootstrap.syncon", "--source=examples/bootstrap.syncon", "--json=out.json"] main
-test = withArgs ["--help"] main
+-- test = withArgs ["--help"] main
+test = withArgs ["parse", "README.md", "source.test"] main
 -- test = GLL.test
 
 getArgsSeq :: IO (Seq [Char])
@@ -241,10 +242,12 @@ parseCommand = Opt.command "parse" (Opt.info parseCmd $ Opt.progDesc "Parse a li
       files <- some $ Opt.argument Opt.str $
         Opt.metavar "FILES..."
 
-      pure $ do
-        (preParseSerialisable, pl) <- readFileDeserialise synconc
-        preParse <- Parser.serialisableToPrecompute @(Lexer.Token Parser.SingleLanguage LD.TypeName) preParseSerialisable & dataOrError mempty ()
-        parseAction' (preParse, pl) files
+      pure $
+        try (readFileDeserialise synconc) >>= \case
+          Left DeserialiseFailure{} -> die $ "Could not parse '" <> toS synconc <> "' as a '.synconc' file, did you supply the right file first?"
+          Right (preParseSerialisable, pl) -> do
+            preParse <- Parser.serialisableToPrecompute @(Lexer.Token Parser.SingleLanguage LD.TypeName) preParseSerialisable & dataOrError mempty ()
+            parseAction' (preParse, pl) files
 
 devCommand :: Opt.Mod Opt.CommandFields (IO ())
 devCommand = Opt.command "dev" (Opt.info devCmd $ Opt.progDesc "Compile and parse a language in one command, for use during language development.")

### 2019-12-11 paper/esop2020rebuttal.txt
Started on rebuttal for ESOP2020

----------------------- REVIEW 1 ---------------------
SUBMISSION: 79
TITLE: Resolvable Ambiguity
AUTHORS: Viktor Palmkvist, Elias Castegren, Philipp Haller and David Broman

----------- Overall evaluation -----------
SCORE: 1 (weak accept)
----- TEXT:
The paper identifies the notion of "resolvable ambiguity" (RA) in
  context-free grammars. A word in a language is resolvable ambiguous
  if for any of its parsing trees there is another word, for which
  this tree is unique. That is, the initial "ambiguous" word can be
  disambiguated in a certain way, e.g., by putting the necessary
  parentheses or other precedence operators. The paper provides formal
  definitions of RA and presents two algorithms determining RA: (a)
  dynamically (for a given word) and (b) statically (for a given
  restricted language). The paper presents a DSL for defining grammars
  amenable to the above mentioned analyses and evaluates it on two
  case studies.

  Overall, as a paper about parsing, this work presents a well-scoped
  problem, backed with a sufficient amount of theory and some
  evaluation. As an outsider to this area, I'm mostly concerned about
  the practical aspects of using the idea of RA for developing
  programming languages. That is, the case studies, besides mentioning
  some large language developments (a subset of OCaml) didn't give me
  a good sense of what is it like to define an ambiguous grammar in a
  general work flow of building a compiler. In this regard, I'm also
  not entirely convinced that having a fully unambiguous grammar from
  the beginning is a Bad Thing. It would be nice if the paper
  elaborated more on it, showing making a convincing case why the
  flexibility afforded by RA is worthy of practical consideration.

  The writing of the paper, while mostly coherent, is a bit uneven at
  some places. For instance, it mentions "parse tree being visibly
  pushdown" on page two without any definition of what visibly
  pushdown automaton is (I had to look it up).

  Small nits:

  * Definition 4: $w'$ should be just $w$.
----------- Reviewer's confidence -----------
SCORE: 1 ((no familiarity))
----------- Questions for author response -----------
* Could you please explain the intended mode of interaction with
  your tool for dynamic and static analysis of ambiguity, when
  developing a new language? Specifically, assuming a dynamic analysis
  determines a certain program to be resolvably ambiguous, what is the
  programmer expected to do next? What about static analysis?

----------------------- REVIEW 2 ---------------------
SUBMISSION: 79
TITLE: Resolvable Ambiguity
AUTHORS: Viktor Palmkvist, Elias Castegren, Philipp Haller and David Broman

----------- Overall evaluation -----------
SCORE: -1 (weak reject)
----- TEXT:
The paper describes disambiguation mechanisms for context-free languages.
Although that parsing is considered more or less a finished topic it is
good that there are still a number of researchers that work on this topici
and produce interesting results. The creation of domain specific languages
and the composition of grammars poses interesting new research challenges,
certainly in the area of ambiguity resolution. The description of legacy
languages also imposes interesting challenges to ambiguity resolution.
The paper contributes to this interesting research area by introducting
two different disambiguation mechanisms for context-free languages.

Although that the paper starts strong, I got less excited during
reading the entire paper. Overall the paper is well structured but
not always very readable.

Section 1 gives an overall introduction to parsing, grammars and ambiguities.

Section 2 introduces a number of motivating examples of grammars and
possible ambiguities, related to binary operators, "dangling-else" problem,
and language composition.

Section 3 introduces the some notation. I wonder why a separate section is
needed for this.

Section 4 introduces the concept of ambiguity resolution. A number of
definitions are presented. I had some trouble in understand definition 4,
what is the relation between w and w', is w' a modified version (with
inserted brackets) of w?
In definition 5 you use \rho_P with w as argument as well as t, is this
what you mean?
It may be possible that after disambiguation all obtained trees are
removed, how do your definitions take care of this?

Section 5 introduces the parse-time disambiguation. I miss in figure 3
the production of the brackets, in figure 4(d) the bracket production
is explicitly introduced. In figure 3 you use the + operator in the right
hand side of the production rule for E (with label l), I assume this is
reads as alternative.
I am totally confused by the last paragraph on page 10 where you show
2 alternative representations of the sentence (1+2)*3 of which you claim
that these representations are ambiguous, why?
At the bottom of page 11 you introduce a new for sequential composition
which leads to ambiguities for the grammar G_D of page 10, or not? I fail
to see the point that you try to make here. The explanation at the top
of page 12 does not help me in understanding your point.
The example in the first paragraph of 5.1 contains a typo, the second "2"
should be a "3".

Section 6 introduces dynamic ambiguity resolution. One question that remains
after reading this section is whether your approach is applicable to
ambiguities that cannot be resolved by introducing paired brackets?
Please explain the automaton at the top of page 13 better, are you able
to recognize [(((a))(b))], I do not think so? Please add a number of
small examples to this section to clarify your approach.

Section 7 introduced the static ambiguity resolution. This section is
unfortunately very hard to understand. Specially in section 7.1 the
presentation is "cluttered", the example does not contribute to understanding
the algorithm. The step from the 3 DFAs on page 16 to the automata presented
on pages 17 and 18 is too big. The added explanation does not clarify the
resulting automata and how to use them. Given this observation the rest
of this section is very hard to understand.

Section 8 introduces 2 case studies, together with the introduction of the
grammar formalism. The used notion in the grammar formalism is a bit
unusual: 'infix add: Exp = "+"', I assume that in a traditional way
this would have been written as 'E = E "+" E'? The "forbid" functionality
is introduced, but none of the examples show explicitly how it is used.
The next section introduces compositionality of grammars and corresponding
ambiguity problems. You have a strong claim at the end of section 8.2, but
I miss the actual motivation for this claim. Section 8.3 has in my opinion
not much added value in its current form. It does not make a strong case
for the approach presented in this paper.

Section 9 presents related work, it is fairly complete, although that
some interesting papers have been published related to disambiguation
and (S)GLR parsing. In a number of these papers the "longest match"
problem have been addressed, which as described int Section 10 (Conclusions)
is future work.
----------- Reviewer's confidence -----------
SCORE: 3 ((expert))
----------- Questions for author response -----------
I do not have specific questions.

----------------------- REVIEW 3 ---------------------
SUBMISSION: 79
TITLE: Resolvable Ambiguity
AUTHORS: Viktor Palmkvist, Elias Castegren, Philipp Haller and David Broman

----------- Overall evaluation -----------
SCORE: -1 (weak reject)
----- TEXT:
The authors argue (following others) that the syntax of a language
does not necessarily have to be unambiguous. They define the concept
of "resolvable ambiguity": ambiguity that can be resolved by
changing the string that is parsed. They also present an algorithm
(for a certain class of languages) that, for a given word, produces
unambiguous minimal words for each of the possible parse trees for
that word, or states that there is no such unambiguous word. Such an
algorithm can be used to aid programmers who write ambiguous code.
Finally the authors present an algorithm, for a seemingly rather
restrictive class of languages, that decides whether every parse
tree is the result of parsing an unambiguous word.

I liked the main ideas of the text. However, lots of details, and
also some high-level explanation, were omitted. For instance, I had
a hard time understanding the first algorithm. For this reason I do
not vouch for the correctness of what is presented in the paper.
This problem could have been mitigated somewhat if the paper had
been accompanied by machine-checked proofs, but there is no
indication that such proofs exist.

My suggestion for the authors is to either write a longer paper, or
to remove the part about the second algorithm and instead expand the
explanation of the first one (and other parts). I did not see the
point of the second algorithm, because the class of languages that
it supports seemed rather restrictive to me. The authors discussed
how the first algorithm could be used (and I liked that discussion),
but there was no similar discussion of the second algorithm.

Another omission: There was no discussion of the time complexity of
the algorithms, or even some indication of whether the authors'
implementations run quickly in practice.

Overall: I liked the ideas, especially the definition of resolvable
ambiguity and the (concept of the) first algorithm. However, given
that I had a hard time understanding the first algorithm, I do not
think that the paper should be published in its current form.

Detailed comments for the authors:

§1: Please state right away that you have not solved the static
resolvability problem for languages where parentheses are used both
for grouping and for other things (like procedure/method
applications). Perhaps you could also indicate what kind of
languages you have solved the problem for.

Definition 4: w' should be w.

Definitions 9 and 10: Swap the two conjuncts of the "for all" part
so that Σ is "in scope" when it is used.

Figure 2: Regular expressions typically include an expression for
the empty language. Why is this expression not included?

§5: Is there no formal definition of how the four grammars are
generated? How do these processes compare to similar processes
discussed by other authors?

§5: In what sense is words an inverse of parse?

§5: You did not state how you turn a language definition D into a
parser P. Does this involve making parse return the empty set for
words in Σ^*-L(G'_D)?

§5.1: Have you proved that the canonical linear encoding is unique?
I would have liked to see more details, including a proof that this
encoding can be computed, efficiently and correctly.

§5.1: You claim that "Section 6 fully solves the dynamic problem",
but in §6 you introduce two restrictions.

§6: It seems as if you only intended to handle languages with
balanced parentheses. In that case you could have moved the first
restriction to §5, in which case it would no longer be a restriction
of the algorithm.

§6: What does "but with parentheses removed" mean? Please be more
precise.

§6: You write that the restrictions can be checked statically. I
believe you, but I think you should explain how. I'd also like to
see some discussion of the time complexity of these checks.

§6: Please explain what a VPDA is. I see that you included some
information in an appendix, but I guess that information would have
to be cut if the paper is accepted for publication.

Algorithm 1: The dynamic resolvability problem takes a word as
input, not a set of parse trees. Is the intention that the algorithm
should be run with parse(w) as its input?

Algorithm 1: Please give a high-level explanation of what the
algorithm does. What is the purpose of the two sets returned from
the algorithm? ("Resolvable"/"Unresolvable"?) What is the purpose of
the final assignment in the inner loop?

My reading: Input: All parse trees for a certain word. For each
parse tree t we try to see whether this tree is resolvably
ambiguous. We do this by finding the words which generate t, but not
any other parse tree in T. If there is no such word, then t is not
resolvably ambiguous. If there is such a word, then t might still
not be resolvably ambiguous, so we check whether t is the unique
parse tree for one of the words, w (a minimal one). If t is unique,
then t is resolvably ambiguous. Otherwise we repeat the process for
the parse trees for w? No, not quite, A is not reset to words(t).
I'm not sure what invariants are supposed to hold in the algorithm,
and I don't understand why it must terminate. Perhaps I could figure
this out if I spent more time on it, but I think the explanation in
the paper should be better.

Algorithm 1: Is shortest(A) always (efficiently) computable for a
VPDA A?

Algorithm 1: Please discuss the algorithm's time complexity.

Algorithm 1, discussion of termination: When you write "that T' is
finite", do you mean "T"? You write that "T' is finite if no words
are infinitely ambiguous", but that sounds like a statement about
"parse(w)", not a statement about an almost unrestricted set.

Algorithm 1, discussion of termination, "T' is finite if no words
are infinitely ambiguous": With respect to what grammar or parser?

Algorithm 1, discussion of termination: What if there are no parse
trees "containing a production from N"?

Algorithm 1, discussion of termination: I did not understand exactly
how requirement 2, stated for "G_D without parentheses", ensures
termination. Please expand the explanation.

§7, "a pair of trees": /Distinct/ trees? The statement
¬∃t₁,t₂∈L(T_D). words(t₁) ⊆ words(t₂) is false whenever L(T_D) is
non-empty.

§7: I do not understand the table. Are the two entries in the
left-most column supposed to be identical?

§7: I did not understand the example with the list [1;2], perhaps
because I am not familiar with the syntax of OCaml. Does OCaml use ;
both to separate list elements and statements?

Proof of Theorem 2, "Given a tree t": A parse tree for the language
with marks? Please explain exactly how you patch the proof of
Theorem 1 to arrive at a proof of Theorem 2, I think the current
text is too vague.

§7.2: Is ambiguity decidable for VPDAs? What does "trimming" mean?

§7.2: When you write ⊂, do you mean "is a /proper/ subset of"?
Presumably not. In that case, why do you use two different symbols
meaning the same thing?

§7.2, "computing the closure": What do you mean by "computing"? Is
this relation finite?

§7.2, "Formalization": Why do you repeat the definitions of the
relation and the VPDAs?

§8.3, limitations: I think that the expression involving let would
be seen as unambiguous if the approach described by Aasa in
"Precedences in specifications and implementations of programming
languages" (TCS, 1995) was used.

§9, "though somewhat comparable to the subclass our static analysis
supports". In what sense? One can have LALR(1) parsers for at least
some languages where parentheses are used for both grouping and
other things.
----------- Reviewer's confidence -----------
SCORE: 3 ((expert))

----------------------- REVIEW 4 ---------------------
SUBMISSION: 79
TITLE: Resolvable Ambiguity
AUTHORS: Viktor Palmkvist, Elias Castegren, Philipp Haller and David Broman

----------- Overall evaluation -----------
SCORE: 2 (strong accept)
----- TEXT:
Summary
-------
The paper proposes to depart from the traditional choice of designing programming languages with unambiguous grammars, which imposes precedence and associativity rules to the user. Instead, the proposal is to let the user directly choose the disambiguation where necessary. The paper formalizes the problem of deciding whether this is always possible for a language definition, in two variants: a dynamic variant considering a particular program, and a static variant considering all programs written in a language. Solutions are given for (less and more) restricted language classes. Finally, the paper demonstrate the practical usefulness of the dynamic solution on two case studies.

Evaluation
----------
In my opinion, the paper goes in a very interesting direction. The motivating examples chosen by the authors are very convincing and struck a chord with me. Indeed, something similar to the first example (Section 2.1) also happens in Python with comparison operators chaining (and it is often very confusing for beginners). The remark made about not clear error messages Section 2.2 is also very valid: anybody who has programmed in OCaml at some point has certainly encountered something similar and suffered because of it.

Unfortunately, I found the core of the paper (Section 6 and 7) rather hard to follow for a reader not familiar with pushdown automata. The illustration by example in Section 7.2 helps only a little. I think that more intuitive explanations (and, ideally, further examples) would go a long way in improving the presentation. Minor remarks on previous sections are below.

Finally, I liked the case studies and, in particular, how the proposed approach can be also applied in the traditional language design settings to find ambiguities in real-world language specifications.

Minor remarks:
- Page 7: Can it happen in practice that two distinct words have the same parse tree? How does it happen?
- In Definition 3 the expression '\exists t_1, t_2...' can be simply replaced by '|parse(w)| >= 2'
- Similarly, below Definition 3, the expression '\exists t...' can be replace by '|parse(w)| = 1'
- In Definition 4 there is a w' that should just be w (or, vice-versa, both should be w')
- It would help to give an example to clearly show the difference between 'unambiguous' and 'resolvably ambiguous' already on page 7 (the example on page 11 could be moved here)
----------- Reviewer's confidence -----------
SCORE: 1 ((no familiarity))
----------- Questions for author response -----------
Can you give a summary of Section 6 and 7 that is more intuitive for an outsider?

----------------------- REVIEW 5 ---------------------
SUBMISSION: 79
TITLE: Resolvable Ambiguity
AUTHORS: Viktor Palmkvist, Elias Castegren, Philipp Haller and David Broman

----------- Overall evaluation -----------
SCORE: -1 (weak reject)
----- TEXT:
## Summary

This paper tackles the problem of ambiguity on grammars. This is
particularly a problem when composing grammars, since the composition
of two unambiguous grammars may be ambiguious, but appears in other
settings, as with the introduction of new operators with unknown
precedence.

The paper defines two notions of how easy it is to fix ambiguities
given an ambiguious grammar. In particular, a program is "resolvably
ambiguious" -- their new term -- if it _can be_ disambiguated, perhaps
by adding parentheses for grouping. This allows us to ask if _every_
program allowed by a grammar is resolvably ambiguious (this is the
static variant of the problem) or if one particular program is
resolvably ambiguious (this is the dynamic variant).

Further, the show that for a particular EBNF-like grammar formalism,
these questions are decidable and they give algorithms for them. The
algorithms rely on translating grammar definitions to visibly pushdown
automata.

They implement the system as a small Haskell library, and show how
resolvability can improve error messages. They also implement an OCaml
parser and use it to detect ambiguities in the grammar given by the
OCaml reference.

## Evaluation

This paper contributes several things:
  - a problem definition -- when do grammars support resolution of
    ambiguity?
  - algorithms to decide this problem for a particular formalism
  - some uses of those algorithms for reporting errors to users and
    managing grammar ambiguity

These all aim at real contributions, but it's not clear how well they
really fit together, and how significant they are.

First, the central motivation of the paper is that ambigious grammars
are useful but need better tooling and that can be provided by solving
these problems. Why is "resolvable ambiguity" the key to making
ambiguity palatable? On the one had, grammar composition does not
ensure that grammars are not unresolvably ambigious (see the Orc
example). On the other hand, there are many other strategies for
disambiguating. The paper discusses OCaml's use of typing for a simple
case, but other systems such as Isabelle tolerate many ambiguities in
the grammar as long as typing only accepts one of them.

Second, if the motivation is to provide the very nice error messages
of page 22, is a sound and complete algorithm needed? Note that no
implementation of static resolvability analysis is presented.

Finally, the OCaml grammar discussion was underwhelming. Simply
writing down the grammar and using any full-CFG parser would produce
the same results -- resolvable ambiguity isn't used at all.

One additional question is whether the new grammar formalism (dubbed
"syncons") is intended as an important new contribution, or merely a
technical device for demonstrating the issues of ambiguity. If the
former, more discussion of why it's an improvement would be
welcome. If the latter, its differences from existing systems should
be reduced.

### Smaller issues

- page 21: the number of new features in this grammar compared to the
  formalisms earlier makes this hard to follow.

- It's possible to avoid the problems described in 8.2 in much simpler
  ways -- simply reject programs that combine operators from multiple
  grammars without parentheses, rather than assuming the precedence
  should be a total order. This is, for example, how the Fortress
  language worked.

- In the discussion of parsers that produce parse forests, some
  discussion of how they handle these cases would be nice.
----------- Reviewer's confidence -----------
SCORE: 1 ((no familiarity))

------------------------------------------------------

==============================================

Reviewer 1:
For a language designer, resolvable ambiguity provides a smoother way
to explore syntax. You can postpone disambiguation decisions until a
later point of development, when you have actually written some programs
using the syntax (or decide to leave the ambiguity). Note that changing
a language to make a previously ambiguous program unambiguous (via
precedence and the like) is a backwards compatible change. The static
analysis ensures that no unresolvable ambiguity can appear (with the
caveat that the current static analysis is limited).
For a programmer, the dynamic analysis provides explicit solutions
to a resolvable ambiguity error, i.e., the programmer is expected
to replace the ambiguous section with one of the suggested solutions.

Reviewer 2:
It may be possible that after disambiguation all obtained trees are
removed, how do your definitions take care of this?
  Assuming disambiguation by programmer: that would imply that the
  new word $w'$ has $\mathit{parse}(w') = \emptyset$, which means it
  is not part of the language, whereby it is not a valid disambiguation.
  Assuming disambiguation by language designer: this just removes the
  word from the language, which is again not a valid disambiguation.
I am totally confused by the last paragraph on page 10 where you show
2 alternative representations of the sentence (1+2)*3 of which you claim
that these representations are ambiguous, why?
  The ambiguous word is the corresponding one in $L(G_D)$ (note the absence
  of a prime), i.e., "1+2*3".
At the bottom of page 11 you introduce a new for sequential composition
which leads to ambiguities for the grammar G_D of page 10, or not? I fail
to see the point that you try to make here. The explanation at the top
of page 12 does not help me in understanding your point.
  $G_D$ already was ambiguous, but $G'_D$ was not. Having an ambiguous $G'_D$
  does not guarantee an unresolvable ambiguity, but it does cause one in this
  case. The intended point is to show an example of an unresolvable
  ambiguity, after which we mention how one might solve this particular one
[...] whether your approach is applicable to ambiguities that cannot be
resolved by introducing paired brackets?
  It cannot, since the design in Section 5 only discards grouping parentheses.
  One could formulate a different formalism, still following the definitions
  in Section 4, that discards more information, which can then be used for
  disambiguation, which naturally poses more challenges for the analyses.
Please explain the automaton at the top of page 13 better, are you able
to recognize [(((a))(b))], I do not think so?
  This automaton does not recognize any word containing '[' or ']', but it
  does indeed recognize (((a))(b)).

Reviewer 3:
Really good review, but too many questions to try to answer here.
The static analysis is present because it gives a desireable guarantee
for a language designer: a user cannot encounter an unresolvable
ambiguity (with the limitations as caveats). Additionally, we
suspected that the dynamic analysis alone would be considered too weak
a contribution.

Reviewer 4:
Can you give a summary of Section 6 and 7 that is more intuitive for an outsider?
  Section 6 takes an ambiguous word, creates an automaton for each parse tree
  (that recognizes the (possibly ambiguous) words that can parse as that tree),
  then combines these to create automata that recognize unambiguous words for each
  tree.
  Section 7 describes a relation between trees that corresponds to an unresolvable
  ambiguity, then describes a way to check for this relation using VPDAs. The
  equivalence only holds for a limited language subclass.

Reviewer 5:
Why is "resolvable ambiguity" the key to making ambiguity palatable?
  The reasoning is largely this: if the language is ambiguous then a user
  can write an ambiguous program. If they do, we need to present it in
  an understandable way that is helpful for solving the ambiguity. The user
  never sees an AST otherwise, whereby we would like to present clarifying
  source code. This is possible exactly when the ambiguity is resolvable,
  by definition.
Finally, the OCaml grammar discussion was underwhelming. Simply
writing down the grammar and using any full-CFG parser would produce
the same results -- resolvable ambiguity isn't used at all.
  This is correct, with the small caveat that the errors produced
  by our methods were quite helpful in finding the underspecified
  places in the grammar.
One additional question is whether the new grammar formalism (dubbed
"syncons") is intended as an important new contribution, or merely a
technical device for demonstrating the issues of ambiguity.
  It is a tool built using the ideas presented here, intended for
  other developments in the research group, thus its syntax is
  designed for that.
It's possible to avoid the problems described in 8.2 in much simpler
ways -- simply reject programs that combine operators from multiple
grammars without parentheses, rather than assuming the precedence
should be a total order. This is, for example, how the Fortress
language worked.
  This is also how our tool works, except we use the dynamic analysis
  to provide example resolutions, and we support this also for
  cases caused by something other than precedence or associativity.

### 2019-12-11 paper/esop2020rebuttal.txt
Updated rebuttal

@@
 
+Thank you for the insightful comments! They will help improve future
+versions of this paper. We begin by answering the explicit "Questions
+for author response", then address some other comments. The latter exceed
+the space limit, and are thus intended for an interested reviewer, or
+the one posing that particular question.
+
 Reviewer 1:
@@
 
-Reviewer 2:
-It may be possible that after disambiguation all obtained trees are
-removed, how do your definitions take care of this?
-  Assuming disambiguation by programmer: that would imply that the
-  new word $w'$ has $\mathit{parse}(w') = \emptyset$, which means it
-  is not part of the language, whereby it is not a valid disambiguation.
-  Assuming disambiguation by language designer: this just removes the
-  word from the language, which is again not a valid disambiguation.
-I am totally confused by the last paragraph on page 10 where you show
-2 alternative representations of the sentence (1+2)*3 of which you claim
-that these representations are ambiguous, why?
-  The ambiguous word is the corresponding one in $L(G_D)$ (note the absence
-  of a prime), i.e., "1+2*3".
-At the bottom of page 11 you introduce a new for sequential composition
-which leads to ambiguities for the grammar G_D of page 10, or not? I fail
-to see the point that you try to make here. The explanation at the top
-of page 12 does not help me in understanding your point.
-  $G_D$ already was ambiguous, but $G'_D$ was not. Having an ambiguous $G'_D$
-  does not guarantee an unresolvable ambiguity, but it does cause one in this
-  case. The intended point is to show an example of an unresolvable
-  ambiguity, after which we mention how one might solve this particular one
-[...] whether your approach is applicable to ambiguities that cannot be
-resolved by introducing paired brackets?
-  It cannot, since the design in Section 5 only discards grouping parentheses.
-  One could formulate a different formalism, still following the definitions
-  in Section 4, that discards more information, which can then be used for
-  disambiguation, which naturally poses more challenges for the analyses.
-Please explain the automaton at the top of page 13 better, are you able
-to recognize [(((a))(b))], I do not think so?
-  This automaton does not recognize any word containing '[' or ']', but it
-  does indeed recognize (((a))(b)).
-
-Reviewer 3:
-Really good review, but too many questions to try to answer here.
-The static analysis is present because it gives a desireable guarantee
-for a language designer: a user cannot encounter an unresolvable
-ambiguity (with the limitations as caveats). Additionally, we
-suspected that the dynamic analysis alone would be considered too weak
-a contribution.
-
 Reviewer 4:
-Can you give a summary of Section 6 and 7 that is more intuitive for an outsider?
 Section 6 takes an ambiguous word, creates an automaton for each parse tree
@@
 
-Reviewer 5:
-Why is "resolvable ambiguity" the key to making ambiguity palatable?
+# Other Comments
+
+## Reviewer 2:
+### It may be possible that after disambiguation all obtained trees are
+removed, how do your definitions take care of this?
+  Assuming disambiguation by programmer: that would imply that the
+  new word $w'$ has $\mathit{parse}(w') = \emptyset$, which means it
+  is not part of the language, whereby it is not a valid disambiguation.
+  Assuming disambiguation by language designer: this just removes the
+  word from the language, which is again not a valid disambiguation.
+### I am totally confused by the last paragraph on page 10 where you show 2 alternative representations of the sentence (1+2)*3 of which you claim
+that these representations are ambiguous, why?
+  The ambiguous word is the corresponding one in $L(G_D)$ (note the absence
+  of a prime), i.e., "1+2*3".
+### At the bottom of page 11 you introduce a new for sequential composition which leads to ambiguities for the grammar G_D of page 10, or not?
+I fail to see the point that you try to make here.
+The explanation at the top of page 12 does not help me in understanding your point.
+  $G_D$ already was ambiguous, but $G'_D$ was not. Having an ambiguous $G'_D$
+  does not guarantee an unresolvable ambiguity, but it does cause one in this
+  case. The intended point is to show an example of an unresolvable
+  ambiguity, after which we mention how one might solve this particular one
+### [...] whether your approach is applicable to ambiguities that cannot be resolved by introducing paired brackets?
+  It cannot, since the design in Section 5 only discards grouping parentheses.
+  One could formulate a different formalism, still following the definitions
+  in Section 4, that discards more information, which can then be used for
+  disambiguation, which naturally poses more challenges for the analyses.
+### Please explain the automaton at the top of page 13 better, are you able to recognize [(((a))(b))], I do not think so?
+  This automaton does not recognize any word containing '[' or ']', but it
+  does indeed recognize (((a))(b)).
+
+## Reviewer 3:
+Thank you especially for the detailed comments on the more technical parts
+of the paper. To avoid being long-winded we will not address all of them
+here, but they will most certainly be considered for future versions.
+### My suggestion for the authors is [...] to remove the part about the second algorithm [...]
+  The static analysis is present because it gives a desireable guarantee
+  for a language designer: a user cannot encounter an unresolvable
+  ambiguity (with the limitations as caveats). However, using the
+  extra space to explore complexity and performance concerns is an
+  interesting idea.
+### I'm not sure what invariants are supposed to hold in the algorithm
+  Roughly, the invariants are the following:
+  - A recognizes a subset of words(t), such that words(t) \ L(A) contains
+    only ambiguous words.
+  - T' contains trees that are known to be ambiguous with t for some word
+    in L(A).
+
+## Reviewer 5:
+### Why is "resolvable ambiguity" the key to making ambiguity palatable?
   The reasoning is largely this: if the language is ambiguous then a user
@@
   by definition.
-Finally, the OCaml grammar discussion was underwhelming. Simply
-writing down the grammar and using any full-CFG parser would produce
-the same results -- resolvable ambiguity isn't used at all.
+### Finally, the OCaml grammar discussion was underwhelming. Simply writing down the grammar and using any full-CFG parser would produce the same results -- resolvable ambiguity isn't used at all.
   This is correct, with the small caveat that the errors produced
@@
   places in the grammar.
-One additional question is whether the new grammar formalism (dubbed
-"syncons") is intended as an important new contribution, or merely a
-technical device for demonstrating the issues of ambiguity.
+### One additional question is whether the new grammar formalism (dubbed "syncons") is intended as an important new contribution, or merely a technical device for demonstrating the issues of ambiguity.
   It is a tool built using the ideas presented here, intended for
@@
   designed for that.
-It's possible to avoid the problems described in 8.2 in much simpler
-ways -- simply reject programs that combine operators from multiple
-grammars without parentheses, rather than assuming the precedence
-should be a total order. This is, for example, how the Fortress
-language worked.
-  This is also how our tool works, except we use the dynamic analysis
-  to provide example resolutions, and we support this also for
-  cases caused by something other than precedence or associativity.
+### It's possible to avoid the problems described in 8.2 in much simpler ways -- simply reject programs that combine operators from multiple grammars without parentheses, rather than assuming the precedenced should be a total order.
+This is, for example, how the Fortress language worked.
+  Our tool also rejects programs combining operators without defined
+  relative precedence, except we use the dynamic analysis to provide
+  example resolutions, and we support this also for cases caused by
+  something other than precedence or associativity.

### 2019-12-11 paper/esop2020rebuttal.txt
Minor rebuttal update

@@
 Reviewer 1:
+For a programmer, the dynamic analysis provides explicit solutions
+to a resolvable ambiguity error, i.e., the programmer is expected
+to replace the ambiguous section with one of the suggested solutions.
+The static analysis should never be run by the programmer.
 For a language designer, resolvable ambiguity provides a smoother way
@@
 analysis ensures that no unresolvable ambiguity can appear (with the
-caveat that the current static analysis is limited).
-For a programmer, the dynamic analysis provides explicit solutions
-to a resolvable ambiguity error, i.e., the programmer is expected
-to replace the ambiguous section with one of the suggested solutions.
+caveat that the current static analysis is limited). The language designer
+is thus expected to change the grammar such that the unresolvable
+ambiguity is no longer possible.
 

### 2020-06-01 paper/esop2020rebuttal.txt
Minor changes to paper and rebuttal

@@
 
-Thank you for the insightful comments! They will help improve future
-versions of this paper. We begin by answering the explicit "Questions
-for author response", then address some other comments. The latter exceed
-the space limit, and are thus intended for an interested reviewer, or
-the one posing that particular question.
+We thank the reviewers for their insightful comments which will
+help improve future versions of the paper. Due to the word limit,
+we begin by answering the explicit "Questions for author
+response". At the end of our response, we allow ourselves to go
+over the word limit addressing other comments raised by reviewers.
+This part is intended for the interested reviewer.
 
 Reviewer 1:
-For a programmer, the dynamic analysis provides explicit solutions
-to a resolvable ambiguity error, i.e., the programmer is expected
-to replace the ambiguous section with one of the suggested solutions.
-The static analysis should never be run by the programmer.
-For a language designer, resolvable ambiguity provides a smoother way
-to explore syntax. You can postpone disambiguation decisions until a
-later point of development, when you have actually written some programs
-using the syntax (or decide to leave the ambiguity). Note that changing
-a language to make a previously ambiguous program unambiguous (via
-precedence and the like) is a backwards compatible change. The static
-analysis ensures that no unresolvable ambiguity can appear (with the
-caveat that the current static analysis is limited). The language designer
-is thus expected to change the grammar such that the unresolvable
-ambiguity is no longer possible.
+For a programmer, the dynamic analysis provides explicit solutions to
+a resolvable ambiguity error, i.e., the programmer is expected to
+replace the ambiguous section with one of the suggested solutions.
+The static analysis should never be run by the programmer.  For a
+language designer, resolvable ambiguity provides a smoother way to
+explore syntax. You can postpone disambiguation decisions until a
+later point of development, when you have actually written some
+programs using the syntax (or decide to leave the ambiguity). Note
+that changing a language to make a previously ambiguous program
+unambiguous (via precedence and the like) is a backwards compatible
+change. The static analysis ensures that no unresolvable ambiguity can
+appear. The language designer is thus expected to change the grammar
+such that the unresolvable ambiguity is no longer possible.
 
@@
 ## Reviewer 2:
-### It may be possible that after disambiguation all obtained trees are
+> It may be possible that after disambiguation all obtained trees are
 removed, how do your definitions take care of this?
@@
   word from the language, which is again not a valid disambiguation.
-### I am totally confused by the last paragraph on page 10 where you show 2 alternative representations of the sentence (1+2)*3 of which you claim
-that these representations are ambiguous, why?
+> I am totally confused by the last paragraph on page 10 where you show 2 alternative representations of the sentence (1+2)*3 of which you claim that these representations are ambiguous, why?
   The ambiguous word is the corresponding one in $L(G_D)$ (note the absence
   of a prime), i.e., "1+2*3".
-### At the bottom of page 11 you introduce a new for sequential composition which leads to ambiguities for the grammar G_D of page 10, or not?
+> At the bottom of page 11 you introduce a new for sequential composition which leads to ambiguities for the grammar G_D of page 10, or not?
 I fail to see the point that you try to make here.
@@
   ambiguity, after which we mention how one might solve this particular one
-### [...] whether your approach is applicable to ambiguities that cannot be resolved by introducing paired brackets?
+> [...] whether your approach is applicable to ambiguities that cannot be resolved by introducing paired brackets?
   It cannot, since the design in Section 5 only discards grouping parentheses.
@@
   disambiguation, which naturally poses more challenges for the analyses.
-### Please explain the automaton at the top of page 13 better, are you able to recognize [(((a))(b))], I do not think so?
+> Please explain the automaton at the top of page 13 better, are you able to recognize [(((a))(b))], I do not think so?
   This automaton does not recognize any word containing '[' or ']', but it
@@
 of the paper. To avoid being long-winded we will not address all of them
-here, but they will most certainly be considered for future versions.
-### My suggestion for the authors is [...] to remove the part about the second algorithm [...]
-  The static analysis is present because it gives a desireable guarantee
+here, but they will most certainly be addressed in future versions.
+> The static analysis is present because it gives a desirable guarantee
   for a language designer: a user cannot encounter an unresolvable
-  ambiguity (with the limitations as caveats). However, using the
-  extra space to explore complexity and performance concerns is an
-  interesting idea.
-### I'm not sure what invariants are supposed to hold in the algorithm
+  ambiguity (with the limitations as caveats). On the other hand, a paper
+  focusing only on the dynamic analysis could use the extra space to discuss
+  complexity and performance concerns. This is something we will consider for
+  future versions of the paper.
+> I'm not sure what invariants are supposed to hold in the algorithm
   Roughly, the invariants are the following:
@@
 ## Reviewer 5:
-### Why is "resolvable ambiguity" the key to making ambiguity palatable?
+> Why is "resolvable ambiguity" the key to making ambiguity palatable?
   The reasoning is largely this: if the language is ambiguous then a user
@@
   by definition.
-### Finally, the OCaml grammar discussion was underwhelming. Simply writing down the grammar and using any full-CFG parser would produce the same results -- resolvable ambiguity isn't used at all.
+> Finally, the OCaml grammar discussion was underwhelming. Simply writing down the grammar and using any full-CFG parser would produce the same results -- resolvable ambiguity isn't used at all.
   This is correct, with the small caveat that the errors produced
@@
   places in the grammar.
-### One additional question is whether the new grammar formalism (dubbed "syncons") is intended as an important new contribution, or merely a technical device for demonstrating the issues of ambiguity.
+> One additional question is whether the new grammar formalism (dubbed "syncons") is intended as an important new contribution, or merely a technical device for demonstrating the issues of ambiguity.
   It is a tool built using the ideas presented here, intended for
@@
   designed for that.
-### It's possible to avoid the problems described in 8.2 in much simpler ways -- simply reject programs that combine operators from multiple grammars without parentheses, rather than assuming the precedenced should be a total order.
+> It's possible to avoid the problems described in 8.2 in much simpler ways -- simply reject programs that combine operators from multiple grammars without parentheses, rather than assuming the precedence should be a total order.
 This is, for example, how the Fortress language worked.

### 2020-06-01 paper/paper.tex
Remove reference to reviewers
Minor changes to paper and rebuttal

@@
 Appendix~\ref{app:prel} contains a more detailed description of
-the preliminaries (for reviewers).
+the preliminaries.
 
@@
 %
-The tool is implemented in Haskell and uses a custom implementation of the Earley parsing algorithm
-\cite{earleyEfficientContextfreeParsing1970}.
+The tool is implemented in Haskell and uses a custom implementation of
+the Earley parsing algorithm \cite{earleyEfficientContextfreeParsing1970}
+and a different version of the dynamic analysis that is faster in the
+common case, but not guaranteed to terminate on an unresolvable
+ambiguity. We bound the execution time with a timeout, and report
+``unresolvable'' if this timeout is triggered.
 

### 2020-08-12 syncon-parser/exe/Main.hs
Small tool to randomly compose fragments

@@ -57,6 +57,8 @@ import qualified P5DynamicAmbiguity.Analysis as DynAmb
 
 import qualified P6Output.JsonV1 as Output
 
+import qualified RandomCompose as RC
+
 import qualified Data.Automaton.NVA as NVA
 import qualified Data.Automaton.GraphViz as GraphViz
 
@@ -69,15 +71,15 @@ nodeAnnotation n = (range n, Parser.n_name n & coerce)
     tokToAnno (Lexer.LitTok r _ t) = (r, "literal " <> show t)
     tokToAnno (Lexer.OtherTok r _ (LD.TypeName tyn) t) = (r, "token " <> show tyn <> " " <> show t)
 
-dataOrError :: (Functor f, Foldable f, FormatError e) => HashMap Text Text -> ErrorOpts e -> Result (f e) a -> IO a
+dataOrError :: (Foldable f, FormatError e) => HashMap Text Text -> ErrorOpts e -> Result (f e) a -> IO a
 dataOrError _ _ (Data a) = return a
 dataOrError source opts (Error e) =
-  die $ formatErrors source $ formatError opts <$> e
+  die $ formatErrors source $ formatError opts <$> toList e
 
-dataOrError' :: (Functor f, Foldable f, FormatError e) => HashMap Text Text -> ErrorOpts e -> Result (f e) a -> IO a
+dataOrError' :: (Foldable f, FormatError e) => HashMap Text Text -> ErrorOpts e -> Result (f e) a -> IO a
 dataOrError' _ _ (Data a) = return a
 dataOrError' source opts (Error e) =
-  die' $ formatErrors source $ formatError opts <$> e
+  die' $ formatErrors source $ formatError opts <$> toList e
 
 testReduce :: IO ()
 testReduce = do
@@ -112,6 +114,7 @@ test :: IO ()
 -- test = withArgs ["--help"] main
 -- test = withArgs ["parse", "README.md", "source.test"] main
 test = withArgs ["pbt", "examples/ambig.syncon"] main
+-- test = withArgs ["compose", "examples/ambig.syncon"] main
 -- test = GLL.test
 
 getArgsSeq :: IO (Seq [Char])
@@ -398,8 +401,38 @@ pbtCommand = Opt.command "pbt" (Opt.info pbtCmd $ Opt.progDesc "Explore the ambi
         Left _ -> Hedgehog.discard
         Right a -> pure a
 
+composeCommand :: Opt.Mod Opt.CommandFields (IO ())
[…]

### 2020-08-12 syncon-parser/exe/RandomCompose.hs

@@ -1,0 +1,173 @@
+module RandomCompose
+( ID
+, mkID
+, computeInfo
+, writeAllFragmentsToDir
+, generateComposition
+) where
+
+import Pre hiding (state, (<.>))
+import Result (Result(..))
+import ErrorMessage (FormatError(..), simpleErrorMessage)
+
+import System.FilePath (dropExtension, (</>), (<.>))
+import Data.Char (isSpace)
+import System.Directory (createDirectoryIfMissing)
+import qualified Text.Printf as Printf
+import qualified Data.Text as Text
+import qualified Data.HashSet as S
+import qualified Data.HashMap.Strict as M
+import System.Random (randomRIO)
+import Data.Array.IO (IOArray, readArray, writeArray, newListArray)
+
+newtype ID = ID { idToText :: Text } deriving (Eq, Ord, Hashable)
+
+data FileInfo = FileInfo
+  { id :: !ID
+  , required :: !(HashSet ID)
+  , forbidden :: !(HashSet ID)
+  , contents :: !Text
+  } deriving (Eq, Generic)
+instance Hashable FileInfo
+
+data Error
+  = UnknownID ID
+  | NotEnoughFragments (HashMap ID FileInfo)  -- ^ The IDs picked so far, since it may be their requirements that limit the number of available fragments
+  deriving (Eq, Generic)
+instance Hashable Error
+
+instance FormatError Error where
+  type ErrorOpts Error = ()
+  formatError () (UnknownID id) = simpleErrorMessage mempty $
+    "Unknown fragment " <> idToText id
+  formatError () (NotEnoughFragments picked) = simpleErrorMessage mempty $
+    "Couldn't pick enough fragments. Here are the fragments that were picked, and which ones they each forbid:\n"
+    <> formatted
+    where
+      formatted = toList picked
+        & sortBy (comparing id)
+        <&> (\FileInfo{id, forbidden} ->
+               (toS :: [Char] -> Text) $ Printf.printf "%- 20s %s" (idToText id) $
+               if S.null forbidden then "" else
+                 toList forbidden <&> idToText & Text.intercalate ", ")
+        & Text.unlines
+
+mkID :: FilePath -> ID
+mkID = dropExtension >>> toS >>> ID
+
+mkFileInfo :: FilePath -> IO FileInfo
+mkFileInfo path = do
+  contents <- readFile path
+  let (normal, deps) = contents
+        & Text.lines
[…]

### 2020-08-12 syncon-parser/src/P5DynamicAmbiguity/Analysis.hs

@@ -125,9 +125,9 @@ analyze timeout pl mkToken getElided showElided alts = do
         & return
       where
         Max len = foldMap (shortestWord >>> fmap (length >>> Max) >>> fold) nvas
-        langsWithCounts = toList languages  -- TODO: OPTIMIZE: I believe this is acutally quite expensive, each NVA is quite large
+        langsWithCounts = toList languages
           <&> (, Sum @Int 1)
-          & M.fromListWith (<>)
+          & M.fromListWith (<>)  -- TODO: OPTIMIZE: I believe this is acutally quite expensive, each NVA is quite large
         nvas = M.keysSet langsWithCounts
         duplicateLangs = langsWithCounts
           & M.filter (getSum >>> (> 1))

### 2020-08-12 paper/paper.tex
Pre repository split snapshot

@@
 %
-\documentclass[runningheads]{llncs}
-%
-\usepackage{cite}
+\documentclass[sigplan,anonymous,review]{acmart}
+%
+%% \usepackage{cite}
 \usepackage{graphicx}
@@
 
-\usepackage{booktabs}   %% For formal tables:
+%% \usepackage{booktabs}   %% For formal tables:
                         %% http://ctan.org/pkg/booktabs
-%% \usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
 %%                         %% http://ctan.org/pkg/subcaption
-\usepackage{subfig}
+%% \usepackage{subfig}
 \usepackage{wrapfig}
@@
 
-\title{Resolvable Ambiguity\thanks{This project is financially supported by the Swedish Foundation for Strategic Research (FFL15-0032)}}
-%
-%\titlerunning{Abbreviated paper title}
-% If the paper title is too long for the running head, you can set
-% an abbreviated paper title here
-%
-\author{Viktor Palmkvist\inst{1} \and
-Elias Castegren\inst{1} \and
-Philipp Haller\inst{1} \and
-David Broman\inst{1}}
-%% \author{Viktor Palmkvist\inst{1}\orcidID{0000-0003-0669-4085} \and
-%% Elias Castegren\inst{1}\orcidID{0000-0003-4918-6582} \and
-%% Philipp Haller\inst{1}\orcidID{0000-0002-2659-5271} \and
-%% David Broman\inst{1}\orcidID{0000-0001-8457-4105}}
-%
-\authorrunning{V. Palmkvist et al.}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
-\institute{KTH Royal Institute of Technology, 100 44 Stockholm, Sweden
-\email{\{vipa,eliasca,dbro,phaller\}@kth.se}%\and
-%% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-%% \email{lncs@springer.com}\\
-%% \url{http://www.springer.com/gp/computer-science/lncs} \and
-%% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-%% \email{\{abc,lncs\}@uni-heidelberg.de}
+\title{Resolvable Ambiguity}
+%% \thanks{This project is financially supported by the Swedish Foundation for Strategic Research (FFL15-0032)} TODO: figure out where to include this
+\author{Viktor Palmkvist}
+\email{vipa@kth.se}
+\orcid{0000-0003-0669-4085}
+
+\author{Elias Castegren}
+\email{eliasca@kth.se}
+\orcid{0000-0003-4918-6582}
+
+\author{Philipp Haller}
+\email{phaller@kth.se}
+\orcid{0000-0002-2659-5271}
+
+\author{David Broman}
+\email{dbro@kth.se}
+\orcid{0000-0001-8457-4105}
+
+\affiliation{%
+  \institution{KTH Royal Institute of Technology}
+  \streetaddress{100 44}
+  \city{Stockholm}
+  \state{Sweden}
   }
-%
-\maketitle              % typeset the header of the contribution
+
+\renewcommand{\shortauthors}{V. Palmkvist et al.}
+
+%
 %
@@
 composability of domain-specific languages.
-\keywords{Syntax \and Ambiguity \and Grammars \and Parsing \and Domain-specific languages}
 \end{abstract}
@@
 %
+\keywords{Syntax, Ambiguity, Grammars, Parsing, Domain-specific languages}
+
+\maketitle              % typeset the header of the contribution
 
@@
 only truth: \emph{the} way of how the syntax of a language must be
-defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
+defined~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
 %
@@
 As a delimitation we only consider languages with balanced parentheses.
-Our approach builds on the following key insights: i) the language of a parse tree is visibly pushdown \cite{alurVisiblyPushdownLanguages2004}, ii) these languages can be encoded as words, and iii) the language of the word encodings of all trees for a language is also visibly pushdown.
+Our approach builds on the following key insights: i) the language of a parse tree is visibly pushdown~\cite{alurVisiblyPushdownLanguages2004}, ii) these languages can be encoded as words, and iii) the language of the word encodings of all trees for a language is also visibly pushdown.
 The last point allows the construction of a visibly pushdown automaton (VPDA) that we examine to solve the static problem for a particular language subclass.
@@
 
-\section{Motivating Examples}
-\label{sec:motivation}
+\section{Motivating Examples}\label{sec:motivation}
 In this section, we motivate the need for ambiguous language definitions, where the decision of how to disambiguate a program is taken by the end-user (the programmer) and not the language designer.
@@
 
-\section{Notation}
-\label{sec:prel}
+\section{Notation}\label{sec:prel}
 
@@
 
-\section{Resolvable Ambiguity} \label{sec:resolvable-definition}
-
-This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to standard concepts in formal languages.
+% TODO: this actually refers to ``Concrete Syntax'' and ``Abstract Syntax Trees'', i.e., not to ``Concrete Syntax Trees''. Not sure how to clarify that
+\section{Describing Concrete Syntax and Abstract Syntax Trees}\label{sec:grammars}
+
+This section describes the grammars we use to describe the concrete
+syntax and corresponding abstract syntax trees (ASTs) of programs. We
+start with context-free grammars and modify them step by step. The
+purpose of these modifications is two-fold: we require explicit
+descriptions of ASTs, and we require these ASTs to discard some
+information from the original parsed program. The discarded
+information can then be used for disambiguation, which we describe in
+more detail in Section~\ref{sec:tree-language}.
+
+We use the following small language with addition, multiplication,
+lists, let-expressions, and numeric literals as a running example:
+
+\begin{center}
+  \begin{minipage}[t]{.5\columnwidth}
+    \begin{tabular}{@{}l@{\quad}l@{}}
+      $E$\quad$->$ & $E$ \texttt{'+'} $E$ \\
+      $E$\quad$->$ & $E$ \texttt{'*'} $E$ \\
+      $E$\quad$->$ & \texttt{'['} $E_l$ \texttt{']'} \\
+      $E$\quad$->$ & \texttt{'let'} \texttt{Ident} \\ % TODO: Ident stands for any identifier, which means that 'Ident' is confusing, but I'm not entirely sure how to present it
+      & \texttt{'='} $E$ \texttt{'in'} $E$ \\
+      $E$\quad$->$ & \texttt{Int} \\
+    \end{tabular}
+  \end{minipage}
+  \begin{minipage}[t]{.3\columnwidth}
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      $E_l$ & $\epsilon$ \\
+      $E_l$ & $E$ $E_{l'}$ \\
+      \addlinespace
+      $E_{l'}$ & $\epsilon$ \\
+      $E_{l'}$ & \texttt{','} $E$ $E_{l'}$ \\
+    \end{tabular}
+  \end{minipage}
+\end{center}
+
+\noindent Monospace font, \texttt{'quoted'} or \texttt{otherwise},
+represents terminals, $\epsilon$ represents the empty sequence.
+
+First we change the right-hand side of each production to be a regular
+expression instead of a sequence (i.e., our grammars are in Extended
+Backus-Naur Form, or EBNF) and add (unique) labels to each production.
+We use parentheses for grouping, $+$ for union, and $^{*}$ for Kleene
+star.
+
+\begin{center}
+  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
+    $E$ & $a$ & $E$ \texttt{'+'} $E$ \\
+    $E$ & $m$ & $E$ \texttt{'*'} $E$ \\
+    $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+    $E$ & $x$ & \texttt{'let'} \texttt{Ident} \texttt{'='} $E$ \texttt{'in'} $E$ \\
+    $E$ & $n$ & \texttt{Int} \\
+  \end{tabular}
+\end{center}
+
+\noindent Both changes serve to shape our ASTs. For example,
+consider the string \texttt{'[1,2]'} with EBNF (left) and without (right):
+
+\begin{center}
+  \begin{tabular}{c@{\qquad}c}
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=20pt}}
+      \tikzset{frontier/.style={distance from root=40pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-4pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        \texttt{[}
+          [.{$n$} \texttt{1} ]
+          \texttt{,}
+          [.{$n$} \texttt{2} ]
+          \texttt{]}
+      ]
+    \end{tikzpicture} &
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=20pt}}
+      \tikzset{frontier/.style={distance from root=80pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-4pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        \texttt{[}
+          [.{$l'$}
+            [.{$n$} \texttt{1} ]
+            [.{$l''$}
+              \texttt{,}
+              [.{$n$} \texttt{2} ]
+              [.{$l'''$} ] ] ]
+          \texttt{]} ]
+    \end{tikzpicture}\\
+  \end{tabular}
+\end{center}
+
+\noindent The extra internal nodes in the right tree would present
+problems for disambiguation later.
+
+Next we add grouping parentheses in the form of a pair of terminals
+per non-terminal. Syntactically speaking these are equivalent with
+adding a production $E \rightarrow \texttt{'('} E \texttt{')'}$,
+except they are discarded when constructing the AST. Note the
+difference between $()$ and \texttt{'('}\texttt{')'}, the former is
+grouping in a regular expression, the latter is a sequence of two
+terminals. We also allow discarding terminals that carry no useful
+semantic information once the production has been identified. Since
+most terminals do \emph{not} carry useful information we denote the
+ones to be kept by underlining them:
+
+\begin{center}
+  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
+    \multicolumn{3}{l}{Productions} \\
+    \midrule
+    $E$ & $a$ & $E$ \texttt{'+'} $E$ \\
+    $E$ & $m$ & $E$ \texttt{'*'} $E$ \\
+    $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+    $E$ & $x$ & \texttt{'let'} \underline{\texttt{Ident}} \texttt{'='} $E$ \texttt{'in'} $E$ \\
+    $E$ & $n$ & \underline{\texttt{Int}} \\
+  \end{tabular}
+  \begin{tabular}{c} % TODO: figure out how to top align these two
+    Grouping \\
+    \midrule
+    \texttt{'('}$E$\texttt{')'} \\
+  \end{tabular}
+\end{center}
+
+\noindent We can thus use grouping parentheses and discarded terminals
+for disambiguation. The former is often used for operator expressions,
+while the latter can be used to, e.g., disambiguate with optional
+semicolons. For example, the strings \texttt{'(1+2)*3'} and
+\texttt{'[1,2]'} produce these ASTs:
+
+\begin{center}
+  \begin{tabular}{c@{\qquad}c}
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=16pt}}
+      \tikzset{frontier/.style={distance from root=48pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
+      \Tree [.{$m$}
+        [.{$a$}
+          [.{$n$} \texttt{1} ]
+          [.{$n$} \texttt{2} ] ]
+        [.{$n$} \texttt{3} ] ]
+    \end{tikzpicture} &
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=16pt}}
+      \tikzset{frontier/.style={distance from root=32pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        [.{$n$} \texttt{1} ]
+        [.{$n$} \texttt{2} ] ]
+    \end{tikzpicture} \\
+  \end{tabular}
+
+\end{center}
+
+\noindent Note that most terminals have been discarded from the
+ASTs. Note also that we need to keep track of where each node came
+from, i.e., which terminal or non-terminal in the grammar it
+represents. For example, the $a$ node in the left AST came from the
+first $E$ in the right-hand side of the $m$ production. This could,
+e.g., be achieved by attaching unique labels to each non-terminal and
+kept terminal in a right-hand side, then additionally annotating each
+AST node with the corresponding label. However, in most examples this
+connection will be obvious, thus we omit these to avoid cluttering the
+presentation, and clarify the connection in text if necessary.
+
+Finally we consider precedence and associativity. A common approach is
+to create a \emph{precedence ladder} by splitting each non-terminal
+into several non-terminals, one per precedence level. Explicitly doing
+this is undesireable in our setting, for two reasons:
+
+\begin{enumerate}
+\item It requires total precedence, but we wish to allow, e.g.,
+  leaving the precedence between operators from different languages
+  undefined.
+\item Grouping parentheses as described above recurse to the
+  \emph{same} non-terminal, which would then not contain productions
+  representing lower precedence operators.
+\end{enumerate}
+
+\noindent The former suggests that we cannot use precedence ladders
+unchanged, and the latter suggests that we need some explicit support
+in our grammars; we cannot simply rewrite the grammar without breaking
+grouping parentheses.
+
+We thus adopt an approach that coincides with precedence ladders if
+precedence is total and transitive, but allows the absence of both
+these properties. To simplify later analysis we treat precedence and
+associativity as user-facing conveniences to be translated to a
+simpler form. We describe the simpler form first, then the
+translation, then illustrate both with examples.
+
+We introduce \emph{marks}, annotations on the non-terminals appearing
+on the right-hand side of a production. Each mark is a set of
+production labels denoting the productions that may \emph{not} appear
+in that position. For example, $E_{\{a, x\}}$ may parse as neither the
+production labelled $a$, nor the production labelled $x$. It may,
+however, parse as a pair of grouping parentheses that contain either of
+these productions. We write $E_\emptyset$ as $E$ to reduce clutter.
+
+Translation is then a process of examining the relative precedence of
+every pair of productions and the associativity of each production and
+adding marks as appropriate. For example, if $E -> m: E \texttt{'*'}
+E$ has higher precedence than $E -> a: E \texttt{'+'} E$ we update $m$
+to $E -> m: E_{\{a\}} \texttt{'*'} E_{\{a\}}$. If $m$ is additionally
+left-associative we update it to $E -> m: E_{\{a\}} \texttt{'*'}
+E_{\{a, m\}}$.
+
+% TODO: figure out how to lay these out nicely. The first and third subfigure should fit vertically and be roughly the same height as the second.
+\begin{figure*}[t]
+  \subcaptionbox{The completed running example.\label{fig:grammar-example}}[.48\textwidth]{
+    \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
+      \multicolumn{3}{l}{Productions} \\
+      \midrule
+      $E$ & $a$ & $E_{\{x\}}$ \texttt{'+'} $E_{\{x,a\}}$ \\
+      $E$ & $m$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
+      $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+      $E$ & $x$ & \texttt{'let'} \underline{\texttt{Ident}} \texttt{'='} $E$ \texttt{'in'} $E$ \\
+      $E$ & $n$ & \underline{\texttt{Int}} \\
+    \end{tabular}
+    \begin{tabular}{c} % TODO: figure out how to top align these two
+      Grouping \\
+      \midrule
+      \texttt{'('}$E$\texttt{')'} \\
+    \end{tabular}
+  }
+  \subcaptionbox{The running example, translated to EBNF.\label{fig:grammar-example-ebnf}}[.48\textwidth]{
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      $E$ & $E_{\{x\}}$ \texttt{'+'} $E_{\{x,a\}}$ \\
+      $E$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
+      $E$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+      $E$ & \texttt{'let'} \texttt{Ident} \texttt{'='} $E$ \texttt{'in'} $E$ \\
+      $E$ & \texttt{Int} \\
+      $E$ & \texttt{'('} $E$ \texttt{')'} \\
+      \addlinespace
+      $E_{\{x\}}$ & $E_{\{x\}}$ \texttt{'+'} $E_{\{x,a\}}$ \\
+      $E_{\{x\}}$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
+      $E_{\{x\}}$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+      $E_{\{x\}}$ & \texttt{Int} \\
+      $E_{\{x\}}$ & \texttt{'('} $E$ \texttt{')'} \\
+      \addlinespace
+      $E_{\{x,a\}}$ & $E_{\{x,a\}}$ \texttt{'*'} $E_{\{x,a,m\}}$ \\
+      $E_{\{x,a\}}$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+      $E_{\{x,a\}}$ & \texttt{Int} \\
+      $E_{\{x,a\}}$ & \texttt{'('} $E$ \texttt{')'} \\
+      \addlinespace
+      $E_{\{x,a,m\}}$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+      $E_{\{x,a,m\}}$ & \texttt{Int} \\
+      $E_{\{x,a,m\}}$ & \texttt{'('} $E$ \texttt{')'} \\
+    \end{tabular}
+  }
+  \subcaptionbox{The running example, written in EBNF with a precedence ladder.\label{fig:grammar-example-ebnf-ladder}}[.48\textwidth]{
+    \begin{tabular}{@{}l@{\quad$->$\quad}l@{}}
+      $E$ & \texttt{'let'} \texttt{Ident} \texttt{'='} $E$ \texttt{'in'} $E$ \\
+      $E$ & $E_1$ \\
+      \addlinespace
+      $E_1$ & $E_1$ \texttt{'+'} $E_2$ \\
+      $E_1$ & $E_2$ \\
+      \addlinespace
+      $E_2$ & $E_2$ \texttt{'*'} $E_3$ \\
+      $E_2$ & $E_3$ \\
+      \addlinespace
+      $E_3$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) \texttt{']'} \\
+      $E_3$ & \texttt{Int} \\
+      $E_3$ & \texttt{'('} $E$ \texttt{')'} \\
+    \end{tabular}
+  }
+  \caption{Completed grammars of the running example.}
+\end{figure*}
+
+% TODO: consistency between ``left associative'' and ``left-associative''
+Applying the usual precedences to our running example (let $<$
+addition $<$ multiplication) as well as associativities
+(left-associative multiplication and addition) yields the grammar in
+Figure~\ref{fig:grammar-example}\footnote{The observant reader may
+  notice that this grammar no longer recognizes, e.g., \texttt{1 + let
+    x = 2 in 3}. We discuss this in Section~\ref{sec:low-prec-unary}}.
+Note that production $a$ has marks containing $x$ due to precedence
+and $a$ due to associativity, and correspondingly $m$ has marks
+containing both $x$ and $a$ due to precedence, and $m$ due to
+associativity.
+
+To illustrate the concrete syntax in more familiar terms we also show
+a translation to normal EBNF grammars in
+Figure~\ref{fig:grammar-example-ebnf}. The translation is done by
+generating a new non-terminal per uniquely marked non-terminal,
+duplicating all productions that are \emph{not} in the mark, and
+adding grouping parentheses. This can be compared with
+Figure~\ref{fig:grammar-example-ebnf-ladder} which is the same
+language, but written more directly as a precedence ladder. Note that
+Figure~\ref{fig:grammar-example-ebnf} is equivalent with
+Figure~\ref{fig:grammar-example-ebnf-ladder} after inlining all
+passthrough productions (e.g. $E -> E_1$).
+
+When describing grammar formalisms a natural question is that of
+expressiveness: how expressive are the grammars we present here?
+Technically speaking they retain the full expressiveness of
+context-free languages; we can trivially translate a CFG by simply
+generating unique labels for each production and retaining all
+terminals. However, such a translation would benefit from none of our
+work as no information is discarded during parsing. We explore the
+implications of this in the evaluation, in
+Section~\ref{sec:expressiveness}.
+
+We are now ready to describe the language of an AST.
+
+\section{The Language of an Abstract Syntax Tree}\label{sec:tree-language}
+
+Central to our approach is the idea that a single AST can be written
+in multiple ways. This means that an AST corresponds to a set of
+written programs, i.e., a language. We thus discuss two kinds of
+languages: those described by the grammars in
+Section~\ref{sec:grammars}, and those consisting of programs that may
+parse as a given AST. We refer to the latter as
+\emph{AST-languages}. The grammars described in
+Section~\ref{sec:grammars} are designed to keep these AST-languages in
+the class of visibly pushdown languages
+\cite{alurVisiblyPushdownLanguages2004} which have many useful closure
+properties. This section informally describes the process of obtaining
+an AST-language from an AST and a grammar. We follow the same process
+of explanation as Section~\ref{sec:grammars}, this time beginning with
+EBNF grammars, then handling each additional modification. Note that
+we do not discuss precedence or associativity, we instead handle marks
+directly. As a running example we consider the AST obtained by parsing
+\texttt{[1, !(2?)]} using the following grammar:
+
+\begin{center}
+  \begin{tabular}{@{}l@{\quad$->$\quad}c@{ $:$\quad}l@{}}
+    \multicolumn{3}{l}{Productions} \\
+    \midrule
+    $E$ & $q$ & $E_{\{c\}}$ \texttt{'?'} \\
+    $E$ & $c$ & \texttt{'!'} $E$ \\
+    $E$ & $l$ & \texttt{'['} ($E$ (\texttt{','} $E$)$^{*}$ $+$ $\epsilon$) (\texttt{','} $+$ $\epsilon$) \texttt{']'} \\
+    $E$ & $n$ & \underline{\texttt{Int}} \\
+  \end{tabular}
+  \begin{tabular}{c} % TODO: figure out how to top align these two
+    Grouping \\
+    \midrule
+    \texttt{'('}$E$\texttt{')'} \\
+  \end{tabular}
+\end{center}
+
+When parsing using a normal EBNF grammar, without discarding any
+information, a given AST can only be parsed in exactly one way, namely
+the flattening of the tree. In this case, the AST (which is then more
+like a parse tree) looks as on the left (using $g$ for grouping
+parentheses), and the AST-language is recognized by the pushdown
+automaton on the right (though we have no need of the stack yet):
+
+\begin{center}
+  \begin{tabular}{cc}
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=16pt}}
+      \tikzset{frontier/.style={distance from root=80pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        \texttt{[}
+        [.{$n$} \texttt{1} ]
+        \texttt{,}
+        [.{$c$}
+          \texttt{!}
+          [.{$g$}
+            \texttt{(}
+            [.{$q$} [.{$n$} \texttt{2} ] \texttt{?} ]
+            \texttt{)} ] ]
+        \texttt{]} ]
+    \end{tikzpicture} &
+    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
+      \tikzset{every state/.style={minimum size=12pt}}
+      \node[state,initial] (1) {};
+      \node[state] (2) [right of=1] {};
+      \node[state] (3) [below of=2] {};
+      \node[state] (4) [left of=3] {};
+      \node[state] (5) [left of=4] {};
+      \node[state] (6) [below of=5] {};
+      \node[state] (7) [right of=6] {};
+      \node[state] (8) [right of=7] {};
+      \node[state] (9) [below of=8] {};
+      \node[state,accepting] (10) [left of=9] {};
+
+      \path (1) edge node{\texttt{[}} (2);
+      \path (2) edge node{\texttt{1}} (3);
+      \path (3) edge[above] node{\texttt{,}} (4);
+      \path (4) edge[above] node{\texttt{!}} (5);
+      \path (5) edge[left] node{\texttt{(}} (6);
+      \path (6) edge node{\texttt{2}} (7);
+      \path (7) edge node{\texttt{?}} (8);
+      \path (8) edge node{\texttt{)}} (9);
+      \path (9) edge[above] node{\texttt{]}} (10);
+    \end{tikzpicture}
+  \end{tabular}
+\end{center}
+
+\noindent When ASTs additionally discard some terminals the
+AST-language is no longer merely a flattening of the tree; instead it
+instead has to represent all possible discarded sequences of
+terminals. In some cases this yields no difference, e.g., production
+$q$ discards exactly one terminal \texttt{'?'}. In other cases there
+is a small difference, e.g., the end of production $l$ may discard
+either \texttt{',]'} or merely \texttt{']'}. Our running example thus
+yields:
+
+\begin{center}
+  \begin{tabular}{c@{\qquad}c}
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=16pt}}
+      \tikzset{frontier/.style={distance from root=80pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        [.{$n$} \texttt{1} ]
+        [.{$c$}
+          [.{$g$}
+            [.{$q$} [.{$n$} \texttt{2} ] ] ] ] ]
+    \end{tikzpicture} &
+    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
+      \tikzset{every state/.style={minimum size=12pt}}
+      \node[state,initial] (1) {};
+      \node[state] (2) [right of=1] {};
+      \node[state] (3) [right of=2] {};
+      \node[state] (4) [right of=3] {};
+      \node[state] (5) [right of=4] {};
+      \node[state] (6) [below of=5] {};
+      \node[state] (7) [left of=6] {};
+      \node[state] (8) [left of=7] {};
+      \node[state] (9) [left of=8] {};
+      % TODO: perform the first line break one step later, then have only the accepting state on the lowest line.
+      \node[state] (11) [left of=9] {};
+      \node[state,accepting] (10) [below of=11] {};
+
+      \path (1) edge node{\texttt{[}} (2);
+      \path (2) edge node{\texttt{1}} (3);
+      \path (3) edge node{\texttt{,}} (4);
+      \path (4) edge node{\texttt{!}} (5);
+      \path (5) edge node{\texttt{(}} (6);
+      \path (6) edge[above] node{\texttt{2}} (7);
+      \path (7) edge[above] node{\texttt{?}} (8);
+      \path (8) edge[above] node{\texttt{)}} (9);
+      \path (9) edge node{\texttt{]}} (10);
+      \path (9) edge[above] node{\texttt{,}} (11);
+      \path (11) edge[left] node{\texttt{]}} (10);
+    \end{tikzpicture}
+  \end{tabular}
+\end{center}
+
+% TODO: I question the clarity of this paragraph. The idea is simple, but I'm not sure I'm conveying it properly
+\noindent In general, the discarded sequence of terminals between two
+retained points in a production must be described by a regular
+language since all non-terminals are retained. For example, the AST
+obtained by parsing \texttt{'[]'} consists of a single node $l$ with
+no children. Its AST-language is described by the regular expression
+$\texttt{'['} (\texttt{','} + \epsilon) \texttt{']'}$ since the other
+sub-expressions in the production contain non-terminals (which must be
+retained) that are not present in the AST.
+
+Next we take discarded grouping parentheses into account. The AST now
+discards the $g$ node, and the automaton must allow inserting an
+arbitrary (balanced) amount of parentheses around each node. We ensure
+this by pushing to the stack when recognizing a \texttt{'('} and
+popping on \texttt{')'}. To ensure that parentheses pair correctly
+around each node we use a distinct stack symbol for each node. These
+stack symbols are represented by letters below. To reduce clutter we
+write $\circ \xrightarrow{+a} \circ$ as shorthand for an edge that
+recognizes \texttt{'('} and pushes $a$ on the stack, while $\circ
+\xrightarrow{-a} \circ$ correspondingly recognizes \texttt{')'} and
+pops $a$ from the stack\footnote{This works for our example since we
+  have only a single form of grouping parentheses, otherwise we would
+  have to be explicit.}.
+
+\begin{center}
+  \begin{tabular}{c@{\qquad}c}
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=16pt}}
+      \tikzset{frontier/.style={distance from root=64pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        [.{$n$} \texttt{1} ]
+        [.{$c$}
+          [.{$q$} [.{$n$} \texttt{2} ] ] ] ]
+    \end{tikzpicture} &
+    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
+      \tikzset{every state/.style={minimum size=12pt}}
+      \node[state,initial] (1) {};
+      \node[state] (2) [right of=1] {};
+      \node[state] (3) [right of=2] {};
+      \node[state] (4) [right of=3] {};
+      \node[state] (5) [right of=4] {};
+      \node[state] (7) [below of=4] {};
+      \node[state] (8) [left of=7] {};
+      % TODO: perform the first line break one step later, then have only the accepting state on the lowest line.
+      \node[state] (11) [below of=1] {};
+      \node[state,accepting] (10) [below of=11] {};
+
+      \path (1) edge node{\texttt{[}} (2);
+      \path (2) edge node{\texttt{1}} (3);
+      \path (3) edge node{\texttt{,}} (4);
+      \path (4) edge node{\texttt{!}} (5);
+      \path (5) edge node{\texttt{2}} (7);
+      \path (7) edge[above] node{\texttt{?}} (8);
+      \path (8) edge node{\texttt{]}} (10);
+      \path (8) edge[above] node{\texttt{,}} (11);
+      \path (11) edge[left] node{\texttt{]}} (10);
+
+      \path (1) edge[loop above] node{$+a$} (1);
+      \path (10) edge[loop left] node{$-a$} (1);
+
+      \path (2) edge[loop above] node{$+b$} (2);
+      \path (3) edge[loop above] node{$-b$} (3);
+
+      \path (4) edge[loop above] node{$+c$} (4);
+      \path (5) edge[loop above,align=left] node{$+d$\\$+e$} (5);
+      \path (8) edge[loop below,align=left] node{$-c$\\$-d$} (8);
+      \path (7) edge[loop below] node{$-e$} (8);
+    \end{tikzpicture}
+  \end{tabular}
+\end{center}
+
+\noindent Finally we account for marks. A mark has the effect of
+forbidding a particular node from being a direct child without at
+least one intermediate pair of grouping parentheses. The AST-language
+must thus require at least one pair in such a location. In this case
+we have only one relevant mark: $c$ may not be a direct child of
+$q$. The final AST-language is thus:
+
+\begin{center}
+  \begin{tabular}{c@{\qquad}c}
+    \begin{tikzpicture}
+      \tikzset{level 1+/.style={level distance=16pt}}
+      \tikzset{frontier/.style={distance from root=64pt}}
+      \tikzset{edge from parent/.style={draw,edge from parent path={(\tikzparentnode.south) -- +(0,-2pt) -| (\tikzchildnode)}}}
+      \Tree [.{$l$}
+        [.{$n$} \texttt{1} ]
+        [.{$c$}
+          [.{$q$} [.{$n$} \texttt{2} ] ] ] ]
+    \end{tikzpicture} &
+    \begin{tikzpicture}[->,>=Latex,shorten >=1pt,auto,node distance=1cm,transform shape]
+      \tikzset{every state/.style={minimum size=12pt}}
+      \node[state,initial] (1) {};
+      \node[state] (2) [right of=1] {};
+      \node[state] (3) [right of=2] {};
+      \node[state] (4) [right of=3] {};
+      \node[state] (5) [right of=4] {};
+      \node[state] (6) [below of=5] {};
+      \node[state] (7) [left of=6] {};
+      \node[state] (8) [left of=7] {};
+      \node[state] (9) [left of=8] {};
+      % TODO: perform the first line break one step later, then have only the accepting state on the lowest line.
+      \node[state] (11) [left of=9] {};
+      \node[state,accepting] (10) [below of=11] {};
+
+      \path (1) edge node{\texttt{[}} (2);
+      \path (2) edge node{\texttt{1}} (3);
+      \path (3) edge node{\texttt{,}} (4);
+      \path (4) edge node{\texttt{!}} (5);
+      \path (5) edge node{$+d$} (6);
+      \path (6) edge[above] node{\texttt{2}} (7);
+      \path (7) edge[above] node{\texttt{?}} (8);
+      \path (8) edge[above] node{$-d$} (9);
+      \path (9) edge node{\texttt{]}} (10);
+      \path (9) edge[above] node{\texttt{,}} (11);
+      \path (11) edge[left] node{\texttt{]}} (10);
+
+      \path (1) edge[loop above] node{$+a$} (1);
+      \path (10) edge[loop left] node{$-a$} (1);
+
+      \path (2) edge[loop above] node{$+b$} (2);
+      \path (3) edge[loop above] node{$-b$} (3);
+
+      \path (4) edge[loop above] node{$+c$} (4);
+      \path (5) edge[loop above,align=left] node{$+d$\\$+e$} (5);
+      \path (8) edge[loop below,align=left] node{$-c$\\$-d$} (8);
+      \path (7) edge[loop below] node{$-e$} (8);
+    \end{tikzpicture}
+  \end{tabular}
+\end{center}
+
+\noindent Note that there are now two required transitions $\circ
+\xrightarrow{+d} \circ$ and $\circ \xrightarrow{-d} \circ$, i.e.,
+there is a required pair of parentheses around the $q$ node.
+
+This automaton is clearly visibly pushdown: it pushes on \texttt{'('}
+and pops on \texttt{')'}, and does not examine the stack on any other
+terminal. However, to ensure that this holds in general we require two
+additional restrictions on our grammars:
+
+\begin{itemize}
+\item The set of opening brackets and closing brackets must be disjoint.
+\item No production may recognize an unbalanced sequence of terminals,
+  i.e., each recognized opening bracket has exactly one later closing
+  bracket.
+\end{itemize}
+
+\noindent Note that the latter requirement is slightly less
+restrictive than it may seem: an opening bracket can be closed with
+\emph{any} closing bracket, not just the one it was defined with.
+
+We are now ready to start resolving ambiguities.
+
+\section{Resolvable Ambiguity}\label{sec:resolvable-definition}
+
+This section introduces our definition of \emph{resolvable ambiguity},
+and then relates it to standard concepts in formal languages.
 
@@
 
+\section{Evaluation}
+
+\subsection{Expressiveness of Our Grammars}\label{sec:expressiveness}
+
+% TODO: Write this section properly once the evaluation is done.
+
+Technically speaking the grammars presented in this paper retain the
+full expressibility of context-free languages; we can trivially
+translate a CFG by simply generating unique labels for each
+production. However, such a translation would benefit from none of our
+work as no information is discarded during parsing. This section thus
+discusses the expressibility limitations we have encountered while
+designing grammars that do discard information appropriately.
+
+\subsubsection{Longest Match}
+
+Some commonly occurring language constructs require longest match,
+perhaps most commonly if-then-else, commonly referred to as the
+``dangling else'' problem.
+
+\subsubsection{Low Precedence Unary Operators}\label{sec:low-prec-unary}
+It can change the recognized language in the presence of
+low-precedence unary
+operators~\cite{aasaPrecedencesSpecificationsImplementations1995}. For
+example, using normal CFGs, a language containing addition,
+let-expressions, and numeric literals could be written naively (left)
+or with a precedence ladder (right):
+
+\begin{center}
+  \begin{tabular}{cc}
+    \begin{tabular}{@{}l@{\quad}l@{}}
+      $E$\quad$->$ & \texttt{'let'} \texttt{Ident} \\
+      & \texttt{'='} $E$ \texttt{'in'} $E$ \\
+      $E$\quad$->$ & $E$ \texttt{'+'} $E$ \\
+      $E$\quad$->$ & \texttt{Int} \\
+    \end{tabular} &
+    \begin{tabular}{@{}l@{\quad}l@{}}
+      $E_1$\quad$->$ & \texttt{'let'} \texttt{Ident} \\
+      & \texttt{'='} $E_1$ \texttt{'in'} $E_1$ \\
+      $E_1$\quad$->$ & $E_2$ \\
+      \addlinespace
+      $E_2$\quad$->$ & $E_2$ \texttt{'+'} $E_3$ \\
+      $E_2$\quad$->$ & $E_3$ \\
+      \addlinespace
+      $E_3$\quad$->$ & \texttt{Int} \\
+    \end{tabular} \\
+  \end{tabular}
+\end{center}
+
+\noindent The left language recognizes \texttt{1 + let x = 2 in 3}
+while the right does not.
+
 \section{Case Studies} \label{sec:evaluation}
@@
 %% Bibliography
-\bibliographystyle{splncs04}
+\bibliographystyle{ACM-Reference-Format}
 

### 2020-08-12 pandoc-doc/graphviz.py
Remove paper and informal writings

@@ -1,46 +1,0 @@
-#!/usr/bin/env python
-
-"""
-Pandoc filter to process code blocks with class "graph" into
-graphviz-generated images.
-Requires pygraphviz, pandocfilters
-"""
-
-import os
-import sys
-
-import pygraphviz
-
-from pandocfilters import toJSONFilter, Para, Image, get_filename4code
-from pandocfilters import get_caption, get_extension, get_value
-
-def document_name():
-    if "PANDOC_INPUT_FILE" in os.environ:
-        return os.environ["PANDOC_INPUT_FILE"]
-    else:
-        return "graph"
-
-def graphviz(key, value, format, _):
-    if key == 'CodeBlock':
-        [[ident, classes, keyvals], code] = value
-        if "graph" in classes:
-            caption, typef, keyvals = get_caption(keyvals)
-            prog, keyvals = get_value(keyvals, u"prog", u"dot")
-            filetype = get_extension(format, "svg", html="svg", latex="pdf")
-            dest = get_filename4code(document_name(), code, filetype)
-
-            if not os.path.isfile(dest):
-                g = pygraphviz.AGraph(string=code)
-                g.layout()
-                g.draw(dest, prog=prog)
-                sys.stderr.write('Created image ' + dest + '\n')
-
-            image = Image([ident, classes, keyvals], 
-                          caption, 
-                          [dest, typef])
-
-            return Para([image])
-
-if __name__ == "__main__":
-
-    toJSONFilter(graphviz)

### 2020-08-13 syncon-parser/src/P5DynamicAmbiguity/TreeLanguage.hs
Make TreeLanguage a bit more strict on grouping parens

@@ -191,12 +191,12 @@ mkLanguage pl@PreLanguage{..} mkToken getElidable (Node n@NodeF{n_nameF}) =
     genPoint (NodePoint sdname _ node) = case node of
       Node NodeF{n_nameF=innerName}
         | isForbidden pl n_nameF sdname innerName -> withOpt
-          { EpsNVA.initial = S.singleton (-1)
-          , EpsNVA.final = S.singleton (-2)
-          , EpsNVA.openTransitions = mkMandEdge (-1) <$> toList (EpsNVA.initial withOpt) <*> (second fst <$> toList groupings)
+          { EpsNVA.initial = S.singleton (-3)
+          , EpsNVA.final = S.singleton (-4)
+          , EpsNVA.openTransitions = mkMandEdge (-3) (-1) <$> (second fst <$> toList groupings)
             & EpsNVA.fromTriples
             & addTransitions (EpsNVA.openTransitions withOpt)
-          , EpsNVA.closeTransitions = mkMandEdge <$> toList (EpsNVA.final withOpt) <*> pure (-2) <*> (second snd <$> toList groupings)
+          , EpsNVA.closeTransitions = mkMandEdge (-2) (-4) <$> (second snd <$> toList groupings)
             & EpsNVA.fromTriples
             & addTransitions (EpsNVA.closeTransitions withOpt) }
         | otherwise -> withOpt
@@ -211,12 +211,21 @@ mkLanguage pl@PreLanguage{..} mkToken getElidable (Node n@NodeF{n_nameF}) =
           <&> first Right
           & Seq.fromList
         withOpt = innerLang
-          { EpsNVA.openTransitions = addTransitions (EpsNVA.openTransitions innerLang) optOpens
-          , EpsNVA.closeTransitions = addTransitions (EpsNVA.closeTransitions innerLang) optCloses }
-        optOpens = mkOptEdge <$> toList (EpsNVA.initial innerLang) <*> (second fst <$> toList groupings)
+          { EpsNVA.initial = S.singleton (-1)
+          , EpsNVA.final = S.singleton (-2)
+          , EpsNVA.openTransitions = addTransitions (EpsNVA.openTransitions innerLang) optOpens
+          , EpsNVA.closeTransitions = addTransitions (EpsNVA.closeTransitions innerLang) optCloses
+          , EpsNVA.innerTransitions = M.unionWith (M.unionWith S.union)
[…]

### 2020-08-13 syncon-parser/exe/Main.hs
Add option to run complete dynamb (though it's broken atm)

@@ -111,9 +111,9 @@ test :: IO ()
 -- test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--dot=out"] main
 -- test = withArgs ["examples/ambig.syncon", "examples/ambig.test", "--two-level"] main
 -- test = withArgs ["examples/bootstrap.syncon", "--source=examples/bootstrap.syncon", "--json=out.json"] main
--- test = withArgs ["--help"] main
+test = withArgs ["--help"] main
 -- test = withArgs ["parse", "README.md", "source.test"] main
-test = withArgs ["pbt", "examples/ambig.syncon"] main
+-- test = withArgs ["pbt", "examples/ambig.syncon"] main
 -- test = withArgs ["compose", "examples/ambig.syncon"] main
 -- test = GLL.test
 
@@ -161,10 +161,11 @@ parseAction = do
     <> Opt.metavar "S"
     <> Opt.help "Timeout for attempting to parse a single source file, in seconds. A negative value means 'wait forever'."
     <> Opt.value (-1)
+  dynAmbKind <- dynOptFlag
   dynAmbTimeout <- fmap (*1_000) $ Opt.option Opt.auto
     $ Opt.long "dynamic-resolvability-timeout"
     <> Opt.metavar "MS"
-    <> Opt.help "Timeout for determining if a single ambiguity is resolvable, in milliseconds. A negative value means 'wait forever'."
+    <> Opt.help "Timeout for determining if a single ambiguity is resolvable, in milliseconds. A negative value means 'wait forever'. Only used by '--fast'."
     <> Opt.value 1_000
   noIsolation <- Opt.switch
     $ Opt.long "no-isolation"
@@ -184,6 +185,9 @@ parseAction = do
         isolate = if noIsolation
                   then DynAmb.dummyIsolate >>> first Seq.singleton
                   else DynAmb.isolate
+        analyze = case fromMaybe FastDyn dynAmbKind of
+          FastDyn -> DynAmb.analyze dynAmbTimeout pl DynAmb.convertToken
+          CompleteDyn -> \a b -> DynAmb.completeAnalyze pl DynAmb.convertToken a b >>> return
     srcNodes <- flip M.traverseWithKey sources $ \path _ -> do
       putStrLn $ "Parsing \"" <> path <> "\""
       handle (\(SourceFileException t) -> modifyIORef' failureFiles (+1) >> sourceFailureHandler t) $ do
@@ -206,7 +210,7 @@ parseAction = do
[…]

### 2020-08-13 syncon-parser/src/Data/Automaton/DVA.hs

@@ -345,12 +345,6 @@ renumber = renumberStates >>> renumberStack
 
 mapKeys :: (Eq k2, Hashable k2) => (k1 -> k2) -> HashMap k1 v -> HashMap k2 v
 mapKeys convert = M.toList >>> fmap (first convert) >>> M.fromList
--- data NVA s sta i o c = NVA
---   { initial :: HashSet s
---   , innerTransitions :: HashMap s (HashMap i (HashSet s))
---   , openTransitions :: HashMap s (HashMap o (HashSet (sta, s)))
---   , closeTransitions :: HashMap s (HashMap c (HashSet (sta, s)))
---   , final :: HashSet s }
 
 asNVA :: ( Eq s, Hashable s
          , Eq sta, Hashable sta)

### 2020-08-13 syncon-parser/src/Data/Automaton/NVA.hs

@@ -198,6 +198,21 @@ product nva1 nva2 = NVA
     cartesianProduct' as bs = (\(sta, a) (stb, b) -> ((sta, stb), (a, b)))
       <$> S.toList as <*> S.toList bs & S.fromList
 
+unions :: forall f s sta i o c. (Foldable f, Eq s, Hashable s, Eq i, Hashable i, Eq o, Hashable o, Eq sta, Hashable sta, Eq c, Hashable c)
+       => f (NVA s sta i o c) -> NVA Int sta i o c
+unions originals = NVA
+  { initial = S.unions $ initial <$> nvas
+  , innerTransitions = foldl' (M.unionWith $ M.unionWith S.union) mempty $ innerTransitions <$> nvas
+  , openTransitions = foldl' (M.unionWith $ M.unionWith S.union) mempty $ openTransitions <$> nvas
+  , closeTransitions = foldl' (M.unionWith $ M.unionWith S.union) mempty $ closeTransitions <$> nvas
+  , final = S.unions $ final <$> nvas
+  } & renumberStates
+  where
+    addStateN :: Int -> NVA s sta i o c -> NVA (Int, s) sta i o c
+    addStateN n = mapStates (n,)
+
+    nvas = zipWith addStateN [0..] $ toList originals
+
 -- |
 -- = Reducing an NVA
 
@@ -435,6 +450,15 @@ fromEpsNVA EpsNVA{..} = NVA
       return newStates
 
     addTransitions transitions = modify (mappend transitions)
+
+asEpsNVA :: (Eq i, Hashable i) => NVA s sta i o c -> EpsNVA s sta i o c
+asEpsNVA NVA{..} = EpsNVA
+  { initial
+  , innerTransitions = mapKeys Just <$> innerTransitions
+  , openTransitions
+  , closeTransitions
+  , final
+  }
 
 renumberStates :: forall s sta i o c.
                   ( Eq s, Hashable s

### 2020-08-13 syncon-parser/src/P5DynamicAmbiguity/Analysis.hs

@@ -1,6 +1,6 @@
-module P5DynamicAmbiguity.Analysis (analyze, Error, ErrorOptions(..), ambiguityStyle, AmbiguityStyle(..), didTimeout) where
+module P5DynamicAmbiguity.Analysis (analyze, completeAnalyze, Error, ErrorOptions(..), ambiguityStyle, AmbiguityStyle(..), didTimeout) where
 
-import Pre
+import Pre hiding (reduce)
 
 import Text.Printf (printf)
 import qualified Data.Text as Text
@@ -8,8 +8,9 @@ import qualified Data.HashSet as S
 import qualified Data.HashMap.Lazy as M
 import Data.Semigroup (Max(..))
 
-import Data.Automaton.NVA (NVA, shortestUniqueWord, shortestWord, TimeoutInfo(..))
+import Data.Automaton.NVA (NVA, shortestUniqueWord, shortestWord, unions, reduce, TimeoutInfo(..))
 import Data.Automaton.EpsilonNVA (untag, TaggedTerminal(..))
+import qualified Data.Automaton.DVA as DVA
 
 import ErrorMessage (FormatError(..), simpleErrorMessage)
 
@@ -83,6 +84,7 @@ data ErrorOptions elidable tok = EO
   , tokRange :: tok -> Range }
 
 type ResLang elidable = NVA Int Int (Token elidable) (Token elidable) (Token elidable)
+type ResStr elidable = [TaggedTerminal (Token elidable) (Token elidable) (Token elidable)]
 
 analyze :: forall elidable tok. (Eq elidable, Hashable elidable, Show elidable, Show tok, Ranged tok)
         => Int  -- ^ Timeout in microseconds. Negative to never timeout.
@@ -116,7 +118,7 @@ analyze timeout pl mkToken getElided showElided alts = do
     languages :: HashMap (NodeOrElide elidable tok) (ResLang elidable)
     languages = S.toMap alts & M.mapWithKey (\node _ -> mkLanguage node)
 
-    shortest :: IO (TimeoutInfo, HashMap (ResLang elidable) [TaggedTerminal (Token elidable) (Token elidable) (Token elidable)])
+    shortest :: IO (TimeoutInfo, HashMap (ResLang elidable) (ResStr elidable))
     shortest = do
       (timeoutInfo, result) <- shortestUniqueWord timeout (len + 10) nvas
       S.toMap nvas
[…]

### 2020-08-13 syncon-parser/test/Main.hs
Updated test suite to run properly

@@ -4,74 +4,12 @@ module Main where
 
 import Prelude ()
 import Pre
-import Result (Result(..))
 
-import System.FilePath (takeFileName, isExtensionOf, (</>))
-import System.Directory (listDirectory, doesDirectoryExist, doesFileExist)
+import Hedgehog.Main (defaultMain)
 
-import Data.IORef (newIORef, readIORef, IORef, writeIORef)
-import qualified Data.HashMap.Strict as M
-import Data.List (partition)
-
-import P1Lexing.Types (Token)
-import qualified P1Lexing.Lexer as Lexer
-import P2LanguageDefinition.Types (TypeName(..))
-import qualified P2LanguageDefinition.Parser as LD
-import qualified P2LanguageDefinition.BasicChecker as LD
-import qualified P4Parsing.Parser as Parser
-import P4Parsing.Types (SingleLanguage(..))
+import qualified ParseTest
 
 main :: IO ()
-main = do
-  failRef <- newIORef False
-  langFiles <- getLanguages "languages"
-  mapM_ (langToTest failRef) langFiles
-  failed <- readIORef failRef
-  if failed
-    then exitFailure
-    else exitSuccess
-
-langToTest :: IORef Bool -> (FilePath, ([FilePath], ([FilePath], [FilePath]))) -> IO ()
-langToTest failRef (lang, (defFiles, (successFiles, failFiles))) = do
-  env <- mkEnv lang defFiles
-  mapM_ (mkTest env True) successFiles
-  mapM_ (mkTest env False) failFiles
-  where
-    mkTest (lexFile, precomputed) expected srcFile = do
-      toks <- doLexFile lexFile srcFile
-      let earleyResult = Parser.parseTokens precomputed toks & isData & (== expected)
-          showRes True  = "    "
-          showRes False = "FAIL"
-      putStrLn $ showRes earleyResult <> " parser/" <> lang <> "/" <> takeFileName srcFile <> "/earley-forest"
-      when (not earleyResult) $ writeIORef failRef True
-    doLexFile lexFile path = lexFile path >>= \case
-      Error _ -> die $ "Could not lex file " <> toS path
-      Data toks -> return toks
-    isData (Data (a, b)) = a `deepseq` b `deepseq` True
-    isData _ = False
-
-getLanguages :: FilePath -> IO [(FilePath, ([FilePath], ([FilePath], [FilePath])))]
-getLanguages languageDirectory = do
-  languages <- listDir languageDirectory >>= filterM doesDirectoryExist
[…]

### 2020-08-14 syncon-parser/src/Data/Automaton/EpsilonNVA.hs
Added a few automata tests

@@ -11,6 +11,8 @@ import qualified Data.HashMap.Lazy as M
 import Data.Automaton (EpsNFA(EpsNFA))
 import qualified Data.Automaton as EpsNFA
 
+import Util (iterateInductively)
+
 -- | A visibly pushdown automaton with
 -- * 's' as the states
 -- * 'sta' as the stack alphabet
@@ -31,6 +33,34 @@ data TaggedTerminal i o c
   | Close c
   deriving (Eq, Generic, Show)
 instance (Hashable i, Hashable o, Hashable c) => Hashable (TaggedTerminal i o c)
+
+recognizes :: forall s sta i o c. (Eq s, Hashable s, Eq sta, Hashable sta, Eq i, Hashable i, Eq o, Hashable o, Eq c, Hashable c)
+           => EpsNVA s sta i o c -> [TaggedTerminal i o c] -> Bool
+recognizes EpsNVA{..} = recur (S.map (,[]) $ epsClose initial)
+  where
+    recur :: HashSet (s, [sta]) -> [TaggedTerminal i o c] -> Bool
+    recur configs [] = S.filter (snd >>> null) configs & S.map fst & any (`S.member` final)
+    recur configs _
+      | S.null configs = False
+    recur configs (c : str) = recur (foldMap (step c) configs) str
+
+    epsTrans s = M.lookup s innerTransitions >>= M.lookup Nothing & fold
+    epsClose = iterateInductively epsTrans
+
+    step :: TaggedTerminal i o c -> (s, [sta]) -> HashSet (s, [sta])
+    step (Inner i) (s, stack) = M.lookup s innerTransitions
+      >>= M.lookup (Just i)
+      & foldMap (epsClose >>> S.map (, stack))
+    step (Open o) (s, stack) = M.lookup s openTransitions
+      >>= M.lookup o
+      & fold
+      & foldMap (\(sta, s') -> S.map (, sta : stack) $ epsClose $ S.singleton s')
+    step (Close _) (_, []) = S.empty
+    step (Close c) (s, sta : stack) = M.lookup s closeTransitions
+      >>= M.lookup c
+      & fold
+      & S.filter (fst >>> (== sta))
+      & foldMap (\(_, s') -> S.map (, stack) $ epsClose $ S.singleton s')
 
 untag :: (o -> a) -> (i -> a) -> (c -> a) -> TaggedTerminal i o c -> a
 untag f _ _ (Open o) = f o

### 2020-08-14 syncon-parser/src/Data/Automaton/NVA.hs

@@ -34,6 +34,32 @@ data NVA s sta i o c = NVA
   deriving (Show, Eq, Generic)
 instance (Hashable s, Hashable sta, Hashable i, Hashable o, Hashable c) => Hashable (NVA s sta i o c)
 
+recognizes :: forall s sta i o c. (Eq s, Hashable s, Eq sta, Hashable sta, Eq i, Hashable i, Eq o, Hashable o, Eq c, Hashable c)
+           => NVA s sta i o c -> [TaggedTerminal i o c] -> Bool
+recognizes NVA{..} = recur (S.map (,[]) initial)
+  where
+    recur :: HashSet (s, [sta]) -> [TaggedTerminal i o c] -> Bool
+    recur configs [] = S.filter (snd >>> null) configs & S.map fst & any (`S.member` final)
+    recur configs _
+      | S.null configs = False
+    recur configs (c : str) = recur (foldMap (step c) configs) str
+
+    step :: TaggedTerminal i o c -> (s, [sta]) -> HashSet (s, [sta])
+    step (Inner i) (s, stack) = M.lookup s innerTransitions
+      >>= M.lookup i
+      & foldMap (S.map (, stack))
+    step (Open o) (s, stack) = M.lookup s openTransitions
+      >>= M.lookup o
+      & fold
+      & S.map (\(sta, s') -> (s', sta : stack))
+    step (Close _) (_, []) = S.empty
+    step (Close c) (s, sta : stack) = M.lookup s closeTransitions
+      >>= M.lookup c
+      & fold
+      & S.filter (fst >>> (== sta))
+      & S.map (\(_, s') -> (s', stack))
+
+
 -- | Retrieve all the states mentioned in an NVA
 states :: (Eq s, Hashable s) => NVA s sta i o c -> HashSet s
 states NVA{..} = initial `S.union` final `S.union` S.fromList (inners ++ opens ++ closes)

### 2020-08-14 syncon-parser/test/AutomataTest.hs

@@ -1,0 +1,136 @@
+{-# LANGUAGE TemplateHaskell, RecordWildCards #-}
+
+module AutomataTest (test) where
+
+import Pre hiding (State)
+
+import Data.Automaton.EpsilonNVA (EpsNVA(EpsNVA), TaggedTerminal(..))
+import qualified Data.Automaton.EpsilonNVA as EpsNVA
+import Data.Automaton.NVA (NVA(NVA))
+import qualified Data.Automaton.NVA as NVA
+
+import qualified Data.HashSet as S
+import qualified Data.HashMap.Strict as M
+
+import Hedgehog hiding (test)
+import qualified Hedgehog.Gen as G
+import qualified Hedgehog.Range as R
+
+test :: IO Bool
+test = checkParallel $$(discover)
+
+data Sigma = A | B | C deriving (Eq, Show, Enum, Bounded, Generic)
+instance Hashable Sigma
+
+data State = P | Q | R | S | T deriving (Eq, Show, Enum, Bounded, Generic)
+instance Hashable State
+
+showStr :: (Show i, Show o, Show c) => [TaggedTerminal i o c] -> [Char]
+showStr = fmap fmt >>> intercalate " " >>> \case
+  [] -> "<empty>"
+  x -> x
+  where
+    fmt (Open o) = "<" <> show o
+    fmt (Close c) = show c <> ">"
+    fmt (Inner i) = show i
+
+gTaggedTerminal :: Gen (TaggedTerminal Sigma Sigma Sigma)
+gTaggedTerminal = G.choice
+  [ Inner <$> G.enumBounded
+  , Open <$> G.enumBounded
+  , Close <$> G.enumBounded
+  ]
+
+gStr :: Gen [TaggedTerminal Sigma Sigma Sigma]
+gStr = G.list (R.linear 0 10) gTaggedTerminal
+
+showEpsNVA :: (Show s, Show sta, Show i, Show o, Show c) => EpsNVA s sta i o c -> [Char]
+showEpsNVA EpsNVA{..} =
+  "initial: " <> intercalate ", " (show <$> toList initial) <> "\n"
+  <> "final: " <> intercalate ", " (show <$> toList final) <> "\n"
+  <> "trans:\n" <> intercalate "\n" (sort $ inners <> opens <> closes)
+  where
+    inners = EpsNVA.toTriples innerTransitions
[…]

### 2020-08-14 syncon-parser/test/Main.hs

@@ -8,8 +8,10 @@ import Pre
 import Hedgehog.Main (defaultMain)
 
 import qualified ParseTest
+import qualified AutomataTest
 
 main :: IO ()
 main = defaultMain
-  [ ParseTest.test
+  [ AutomataTest.test
+  , ParseTest.test
   ]

### 2020-08-14 syncon-parser/test/AutomataTest.hs
Some more tests

@@ -2,17 +2,22 @@
 
 module AutomataTest (test) where
 
-import Pre hiding (State)
+import Pre hiding (State, race, diff)
+
+import Control.Concurrent.Async.Lifted (race)
 
 import Data.Automaton.EpsilonNVA (EpsNVA(EpsNVA), TaggedTerminal(..))
 import qualified Data.Automaton.EpsilonNVA as EpsNVA
 import Data.Automaton.NVA (NVA(NVA))
 import qualified Data.Automaton.NVA as NVA
+import Data.Automaton.DVA (DVA(DVA))
+import qualified Data.Automaton.DVA as DVA
 
 import qualified Data.HashSet as S
 import qualified Data.HashMap.Strict as M
 
 import Hedgehog hiding (test)
+import qualified Hedgehog
 import qualified Hedgehog.Gen as G
 import qualified Hedgehog.Range as R
 
@@ -24,6 +29,20 @@ instance Hashable Sigma
 
 data State = P | Q | R | S | T deriving (Eq, Show, Enum, Bounded, Generic)
 instance Hashable State
+
+discardSlow :: Int -> Hedgehog.TestT IO a -> Hedgehog.PropertyT IO a
+discardSlow timelimit v = do
+  result <- Hedgehog.test $ race (liftIO $ threadDelay timelimit) v
+  case result of
+    Left _ -> discard
+    Right a -> pure a
+
+failSlow :: Int -> Hedgehog.TestT IO a -> Hedgehog.PropertyT IO a
+failSlow timelimit v = do
+  result <- Hedgehog.test $ race (liftIO $ threadDelay timelimit) v
+  case result of
+    Left _ -> failure
+    Right a -> pure a
 
 showStr :: (Show i, Show o, Show c) => [TaggedTerminal i o c] -> [Char]
 showStr = fmap fmt >>> intercalate " " >>> \case
@@ -57,19 +76,6 @@ showEpsNVA EpsNVA{..} =
     closes = EpsNVA.toTriples closeTransitions
       <&> \(p, l, (sta, q)) -> "  " <> show p <> " -- >" <> show l <> ", -" <> show sta <> " --> " <> show q
 
-showNVA :: (Show s, Show sta, Show i, Show o, Show c) => NVA s sta i o c -> [Char]
-showNVA NVA{..} =
-  "initial: " <> intercalate ", " (show <$> toList initial) <> "\n"
[…]

### 2020-08-14 syncon-parser/src/Data/Automaton/DVA.hs
Fix bug in dvaOp

@@ -13,6 +13,7 @@ import Data.These (fromThese)
 
 import Util (iterateInductivelyOptM, repeatUntilStableM)
 
+import Data.Automaton.EpsilonNVA (TaggedTerminal(..))
 import Data.Automaton.NVA (NVA(NVA))
 import qualified Data.Automaton.NVA as NVA
 
@@ -49,6 +50,28 @@ instance
   => Monoid (DeterminizeState s sta i o c) where
   mappend = (<>)
   mempty = DeterminizeState mempty mempty mempty
+
+recognizes :: forall s sta i o c. (Eq s, Hashable s, Eq sta, Hashable sta, Eq i, Hashable i, Eq o, Hashable o, Eq c, Hashable c)
+           => DVA s sta i o c -> [TaggedTerminal i o c] -> Bool
+recognizes DVA{..} = recur (initial, [])
+  where
+    recur :: (s, [sta]) -> [TaggedTerminal i o c] -> Bool
+    recur (s, []) [] = s `S.member` final
+    recur _ [] = False
+    recur config (c : cs) = step c config & maybe False (\conf -> recur conf cs)
+
+    step :: TaggedTerminal i o c -> (s, [sta]) -> Maybe (s, [sta])
+    step (Inner i) (s, stack) = M.lookup s innerTransitions
+      >>= M.lookup i
+      <&> (, stack)
+    step (Open o) (s, stack) = M.lookup s openTransitions
+      >>= M.lookup o
+      <&> \(sta, s') -> (s', sta : stack)
+    step (Close _) (_, []) = Nothing
+    step (Close c) (s, sta : stack) = M.lookup s closeTransitions
+      >>= M.lookup c
+      >>= M.lookup sta
+      <&> (, stack)
 
 dstates :: (Eq s, Hashable s) => DeterminizeState s sta i o c -> HashSet s
 dstates DeterminizeState{..} = S.fromList $ is <> os <> cs
@@ -267,8 +290,8 @@ dvaOp op dva1 dva2 = DVA
       (iterateInductivelyOptM findTransitions $ S.singleton newInitial)
       mempty
 
-    dva1close = stackSymbols dva1 & S.map Just & S.toMap & fmap (const Nothing)
-    dva2close = stackSymbols dva2 & S.map Just & S.toMap & fmap (const Nothing)
+    dva1close = stackSymbols dva1 & S.map Just & S.insert Nothing & S.toMap <&> const Nothing
[…]

### 2020-08-14 syncon-parser/exe/Main.hs
Allow running both ambiguity algorithms and using the fastest

@@ -185,9 +185,12 @@ parseAction = do
         isolate = if noIsolation
                   then DynAmb.dummyIsolate >>> first Seq.singleton
                   else DynAmb.isolate
+        fastAnalyze = DynAmb.analyze dynAmbTimeout pl DynAmb.convertToken
+        completeAnalyze = \a b -> DynAmb.completeAnalyze pl DynAmb.convertToken a b >>> return
         analyze = case fromMaybe FastDyn dynAmbKind of
           FastDyn -> DynAmb.analyze dynAmbTimeout pl DynAmb.convertToken
           CompleteDyn -> \a b -> DynAmb.completeAnalyze pl DynAmb.convertToken a b >>> return
+          RaceDyn -> \a b c -> race (fastAnalyze a b c) (completeAnalyze a b c) <&> either identity identity
     srcNodes <- flip M.traverseWithKey sources $ \path _ -> do
       putStrLn $ "Parsing \"" <> path <> "\""
       handle (\(SourceFileException t) -> modifyIORef' failureFiles (+1) >> sourceFailureHandler t) $ do
@@ -265,16 +268,19 @@ compileCommand = Opt.command "compile" (Opt.info compileCmd $ Opt.progDesc "Comp
         let serialisable = (Parser.precomputeToSerialisable preParse, pl)
         writeFileSerialise outputFile serialisable
 
-data DynAnalysisKind = FastDyn | CompleteDyn
+data DynAnalysisKind = FastDyn | CompleteDyn | RaceDyn
 dynOptFlag :: Opt.Parser (Maybe DynAnalysisKind)
-dynOptFlag = optional $ fast <|> complete
+dynOptFlag = optional $ fast <|> complete <|> race'
   where
     fast = Opt.flag' FastDyn
-      $ Opt.long "fast"
+      $ Opt.long "fast-dyn"
       <> Opt.help "Use the fast dynamic analysis that might not terminate (except for timeout) on unresolvable input"
     complete = Opt.flag' CompleteDyn
-      $ Opt.long "complete"
-      <> Opt.help "Use the complete dynamic analsis that will always terminate, but that might be very slow"
+      $ Opt.long "complete-dyn"
+      <> Opt.help "Use the complete dynamic analysis that will always terminate, but that might be very slow"
+    race' = Opt.flag' RaceDyn
+      $ Opt.long "race-dyn"
+      <> Opt.help "Run both dynamic analyses and use the one that terminates first."
 
[…]

### 2020-08-14 syncon-parser/exe/Main.hs
Default to --race-dyn when looking for unresolvable ambiguity

@@ -380,7 +380,7 @@ pbtCommand = Opt.command "pbt" (Opt.info pbtCmd $ Opt.progDesc "Explore the ambi
                       then DynAmb.dummyIsolate >>> first Seq.singleton
                       else DynAmb.isolate
             defDynAmbKind = case target of
-              UnresolvableAmbiguity -> CompleteDyn
+              UnresolvableAmbiguity -> RaceDyn
               Ambiguity -> FastDyn
             fastAnalyze = DynAmb.analyze (-1) pl DynAmb.convertToken
             completeAnalyze = \a b -> DynAmb.completeAnalyze pl DynAmb.convertToken a b >>> return
